### 2024-09-06 docs/hello/hello.tex
vimtex stuff working

\documentclass{article}

% These are only here for the sake of this tutorial.
\usepackage{amsmath}
\newcommand{\lt}{\textbackslash{}lt }
\newcommand{\lk}{\textbackslash{}lk }
\newcommand{\lv}{\textbackslash{}lv }
\newcommand{\lc}{\textbackslash{}lc }
\renewcommand{\ll}{\textbackslash{}ll }
\renewcommand{\le}{\textbackslash{}le }

\begin{document}

\section{Compiling}

Press \ll to start (or stop) compiling the document.
This turns on automatic compiling if supported by your compiler, meaning your
document will be compiled each time you save.

Press \lk to stop the compilation process

Press \lc to clear auxiliary files.

\section{Forward and Backward Search}

Press \lv to forward search. This will open the compiled PDF.

Some PDF viewers support the ability to jump to the current location in the PDF.

Some also support backward search (jumping from PDF to source).

\section{Errors, Warnings and the QuickFix window}

VimTeX will populate the QuickFix menu with any errors and warnings it finds in
the log file.

This is done after (attempting to) compile or pressing \le.

\le also closes the QuickFix menu when it is open.

\emph{Example.} Let's put some undefined commands here and compile:

Moving to the QuickFix window and pressing enter on the error entry moves the
cursor to the line containing the error. We can fix the error and continue.

\section{Document Overview}

Press \lt to show a window with a table of contents for your document.

Also shows labels, references and TODOs.

Can be used to jump to sections, labels, references and TODOs.

% TODO: Show table of contents.

\section{Motions and Text Objects}

VimTeX adds various LaTeX specific motions and text objects.

\emph{Example motion.} move between section boundaries with [[, [], ][ and ]].

\emph{Example text object.} ic refers to a LaTeX command excluding the
backslash. ac includes the backslash.

cic: \( \alpha \)

dac: \(  \)

\section{Other Cool Stuff}

cse changes the surrounding environment:

\begin{equation}
  v \cdot \nabla_x f_1 = -\sigma_a f_1, \quad f_1 \rvert_{\Gamma_-} = \phi(x,v).
\end{equation}

tsd toggles between () and \( \left( \right) \):

\begin{equation*}
  \frac{d}{ds} \left[ e^{\int_k^s \sigma_a(x + tv)\,dt} F_1(s) \right] = 0
\end{equation*}

tse toggles the * in environments:

\begin{equation}
  f_1(x, v) = e^{-\int_0^{\tau_-(x, v)} \sigma_a(x - sv)\,dt} \phi(x_-, v),
\end{equation}

This is only a tiny sliver of what VimTeX offers.
Read the README for an overview, and read :h vimtex for a full description of
all things VimTeX.

\end{document}

### 2024-09-06 docs/proposal/proposal.tex
vimtex stuff working

\documentclass{article}
\usepackage{hyperref}

\begin{document}

This is the repository of \href{https://github.com/zslwyuan/AMF-Placer}{AMF-Placer}.

\end{document}

### 2024-09-06 java/src/main/java/AnalyticalPlacer.java
minor fomatting

@@ -17,7 +17,6 @@ import com.xilinx.rapidwright.design.ModuleInst;
 import com.xilinx.rapidwright.design.ModuleImpls;
 import com.xilinx.rapidwright.design.Cell;
 
-
 import com.xilinx.rapidwright.edif.EDIFLibrary;
 import com.xilinx.rapidwright.edif.EDIFNetlist;
 import com.xilinx.rapidwright.edif.EDIFCell;
@@ -49,6 +48,5 @@ public class AnalyticalPlacer extends Placer {
 
         return design;
     }
-
 
 }

### 2024-09-06 java/src/main/java/Main.java

@@ -7,7 +7,6 @@ import java.io.FileWriter;
 import java.io.BufferedWriter;
 import java.io.IOException;
 
-
 public class Main {
 
     private static final Logger logger = Logger.getLogger(Main.class.getName());
@@ -16,7 +15,7 @@ public class Main {
     public static void main(String[] args) {
         try {
             // Logger to keep track of execution progress.
-            FileHandler fileHandler = new FileHandler(rootDir+"outputs/logger.txt", true); // 'true' appends to file
+            FileHandler fileHandler = new FileHandler(rootDir + "outputs/logger.txt", true); // 'true' appends to file
             fileHandler.setFormatter(new SimpleFormatter());
             logger.addHandler(fileHandler);
             logger.setLevel(Level.ALL); // Set logging level to record all messages
@@ -25,7 +24,7 @@ public class Main {
             SimulatedAnnealingPlacer SAPlacer = new SimulatedAnnealingPlacer();
             SAPlacer.run();
             logger.log(Level.INFO, "Data writing complete. Check 'output.txt'");
-            
+
         } catch (IOException e) {
             logger.log(Level.SEVERE, "An IOException occurred while configuring the logger.", e);
         }

### 2024-09-06 java/src/main/java/SimulatedAnnealingPlacer.java

@@ -17,7 +17,6 @@ import com.xilinx.rapidwright.design.ModuleInst;
 import com.xilinx.rapidwright.design.ModuleImpls;
 import com.xilinx.rapidwright.design.Cell;
 
-
 import com.xilinx.rapidwright.edif.EDIFLibrary;
 import com.xilinx.rapidwright.edif.EDIFNetlist;
 import com.xilinx.rapidwright.edif.EDIFCell;
@@ -49,6 +48,5 @@ public class SimulatedAnnealingPlacer extends Placer {
 
         return design;
     }
-
 
 }

### 2024-09-06 run.sh
update run.sh to use gradle

@@ -32,21 +32,42 @@ if [ "$start_stage" == "synth" ] || [ "$start_stage" == "all" ]; then
     echo "Vivado synthesis completed. Check 'synthesized.dcp'."
 fi
 
-# Java Compilation Stage
+# # Java Compilation Stage
+# if [ "$start_stage" == "compile" ] || [ "$start_stage" == "all" ]; then
+#     echo "Compiling Java files..."
+#     javac src/*.java
+#     check_exit_status "Java compilation"
+#     echo "Java compilation completed."
+# fi
+#
+# # Java Placement Stage
+# if [ "$start_stage" == "place" ] || [ "$start_stage" == "all" ]; then
+#     echo "Running Java placement..."
+#     java -cp "$CLASSPATH:." src.Main
+#     check_exit_status "Java placement execution"
+#     echo "Java placement executed. Check 'logger.txt' for output."
+# fi
+
+# Gradle Build Stage (replaces Java Compilation)
 if [ "$start_stage" == "compile" ] || [ "$start_stage" == "all" ]; then
-    echo "Compiling Java files..."
-    javac src/*.java
-    check_exit_status "Java compilation"
-    echo "Java compilation completed."
+    echo "Building Java project with Gradle..."
+    cd $PROJ_DIR/java
+    gradle build
+    check_exit_status "Gradle build"
+    echo "Gradle build completed."
 fi
 
-# Java Placement Stage
+# Gradle Run Stage (replaces Java Placement)
 if [ "$start_stage" == "place" ] || [ "$start_stage" == "all" ]; then
-    echo "Running Java placement..."
-    java -cp "$CLASSPATH:." src.Main
-    check_exit_status "Java placement execution"
+    echo "Running Java placement with Gradle..."
+    cd $PROJ_DIR/java
+    gradle run
+    check_exit_status "Gradle run"
     echo "Java placement executed. Check 'logger.txt' for output."
 fi
+
+# Return to outer dir
+cd $PROJ_DIR
 
 # Vivado Route Stage
 if [ "$start_stage" == "route" ] || [ "$start_stage" == "all" ]; then

### 2024-09-08 docs/proposal/proposal.tex
begin migration to latex

@@
 \usepackage{hyperref}
+\usepackage{enumitem}
 
 \begin{document}
+\title{Technical Paper Proposal: \\ Placement Algorithms for Heterogenous FPGAs}
+\author{Brian Cheng \\ Department of Electrical and Computer Engineering}
 
-This is the repository of \href{https://github.com/zslwyuan/AMF-Placer}{AMF-Placer}.
+\date{}
+\maketitle{}
+
+\section{Abbreviations}
+\begin{itemize}[label={--}, left=0.25cm] % Adjust label and left indent as desired
+    \item FPGA: Field Programmable Gate Array
+    \item VLSI: Very Large Scale Integration
+    \item EDA: Electronic Design Automation
+    \item VHSIC: Very High Speed Integrated Circuits
+    \item HDL: Hardware Description Language
+    \item VHDL: VHSIC HDL
+    \item HLS: High Level Synthesis
+    \item PL-PS: Programmable Logic - Processing System
+\end{itemize}
+
+\section{Ideas}
+\begin{itemize}[label={\textbullet}, left=0.25cm]
+    \item \textbf{FPGA}: Field Programmable Gate Array
+    \begin{itemize}[label={--}, left=0.25cm]
+        \item FPGA Vendors:
+        \begin{itemize}[label={$\cdot$}, left=0.25cm]
+            \item AMD-Xilinx (~50\% FPGA vendor market share)
+            \item Intel-Altera (~35\% share)
+            \item Lattice
+            \item Microsemi
+        \end{itemize}
+    \end{itemize}
+
+    \item \textbf{EDA}: Electronic Design Automation
+    \begin{itemize}[label={--}, left=0.25cm]
+        \item Proprietary software for FPGA and VLSI development:
+        \begin{itemize}[label={$\cdot$}, left=0.25cm]
+            \item Xilinx - Vivado (Design + Simulation) + Vitis (HLS + PL-PS codesign)
+            \item Altera - Quartus (Design) + ModelSim (Simulation)
+            \item Synopsis (VLSI)
+            \item Cadence (VLSI)
+        \end{itemize}
+        \item Open source software for FPGA development:
+        \begin{itemize}[label={$\cdot$}, left=0.25cm]
+            \item VTR: a PNR tool popular amongst researchers who study placement techniques
+            \item OSS-CAD: a full-flow software suite that includes ABC synthesis, Yosys synthesis, Yosys nextpnr.
+            \item AMF-Placer (Analytical Placer)
+            \item DREAMPlace (VLSI) + DREAMPlaceFPGA
+            \item RapidLayout (Hard Block Placer)
+            \item RapidStream (Systolic Array Placer)
+            \item RapidWright: Semi-open source API that provides backend access to Xilinx Vivado EDA using design checkpoints.
+
+        \end{itemize}
+    \end{itemize}
+
+    \item \textbf{Synthesis}
+    \begin{itemize}[label={--}, left=0.25cm]
+        \item 
+        \item 
+        \item 
+        \item 
+    \end{itemize}
+\end{itemize}
 

### 2024-09-08 docs/proposal/proposal.tex
minor corrections
more tex

@@
 \date{}
-\maketitle{}
+\maketitle
 
@@
     \item PL-PS: Programmable Logic - Processing System
+    \item EDIF: Electronic Design Interchange Format
 \end{itemize}
@@
             \item DREAMPlace (VLSI) + DREAMPlaceFPGA
-            \item RapidLayout (Hard Block Placer)
-            \item RapidStream (Systolic Array Placer)
+            \item RapidLayout (Hard Block Placer for Systolic Arrays)
+            \item RapidStream (HLS Placer)
             \item RapidWright: Semi-open source API that provides backend access to Xilinx Vivado EDA using design checkpoints.
-
         \end{itemize}
@@
     \begin{itemize}[label={--}, left=0.25cm]
-        \item 
-        \item 
-        \item 
-        \item 
-    \end{itemize}
+        \item Takes a design written in a high-level HDL like VHDL or Verilog and "synthesizes" a \textbf{logical netlist} out of it.
+This logical netlist is usually generated as an EDIF, JSON, or a low-level Verilog file.
+This netlist describes the necessary basic elements of logic (BELs) and the wired connections between them that are necessary to implement the design.
     \end{itemize}
 
+    \item \textbf{Placement}
+    \begin{itemize}[label={--}, left=0.25cm]
+        \item Takes the \textbf{logical netlist} and produces a \textbf{physical netlist}.
+    \end{itemize}
+
+    \item \textbf{Routing}
+    \begin{itemize}[label={--}, left=0.25cm]
+        \item Takes the \textbf{physical netlist} and maps the connections between BELs onto physical switchboxes.
+    \end{itemize}
+
+\end{itemize}
+
+This is a citation for AMFPlacer. \cite{AMFPlacer}
+
+\bibliographystyle{ieeetr}
+\bibliography{references}
+
 \end{document}

### 2024-09-09 docs/proposal/proposal.tex
prepare references and figures

@@
 \usepackage{enumitem}
+\usepackage{graphicx}
 
@@
     \item EDIF: Electronic Design Interchange Format
+    \item HPWL: Half Perimeter Wire Length
+\end{itemize}
+
+\section{Keywords}
+\begin{itemize}
+    \item FPGA, EDA, Synthesis, Placement, Routing, Parallel, Optimization
 \end{itemize}
@@
         \item Takes a design written in a high-level HDL like VHDL or Verilog and "synthesizes" a \textbf{logical netlist} out of it. 
-This logical netlist is usually generated as an EDIF, JSON, or a low-level Verilog file.
-This netlist describes the necessary basic elements of logic (BELs) and the wired connections between them that are necessary to implement the design.
+        \item The logical netlist is usually generated as an EDIF, JSON, or a low-level Verilog file. 
+        \item The netlist describes the necessary basic elements of logic (BELs) and the wired connections between them that are necessary to implement the design.
     \end{itemize}
@@
         \item Takes the \textbf{logical netlist} and produces a \textbf{physical netlist}.
+        \item For each BEL in the netlist, assign the BEL to a Cell, Site, and Tile on the physical FPGA device.
+        \item History of the landscape of placement algorithms:
     \end{itemize}
@@
     \begin{itemize}[label={--}, left=0.25cm]
-        \item Takes the \textbf{physical netlist} and maps the connections between BELs onto physical switchboxes.
+        \item Takes the \textbf{physical netlist} and maps the connections between BELs onto wires, interconnects, and switchboxes on the FPGA.
+        \begin{figure}
+            \begin{center}
+                \includegraphics[width=0.75\textwidth]{figures/kraftwerk2.png}
+            \end{center}
+            \caption{Landscape of VLSI placement techniques (Spindler) \cite{kraftwerk2} }
+            \label{fig:}
+        \end{figure}
+        \begin{figure}
+            \begin{center}
+                \includegraphics[width=0.95\textwidth]{figures/ProCha.png}
+            \end{center}
+            \caption{Historical timeline of VLSI placement techniques (Markov) \cite{ProCha} }
+            \label{fig:}
+        \end{figure}
     \end{itemize}
@@
 
+\newpage
 \bibliographystyle{ieeetr}
-\bibliography{references}
+\nocite{*}
+\bibliography{
+    references/surveys,
+    references/fpga_placement,
+    references/vlsi_placement,
+}
 

### 2024-09-09 docs/proposal/proposal.tex
more doc work

@@
 \maketitle
+
+\section{Proposal}
 
@@
             \item DREAMPlace (VLSI) + DREAMPlaceFPGA
-            \item RapidLayout (Hard Block Placer for Systolic Arrays)
-            \item RapidStream (HLS Placer)
             \item RapidWright: Semi-open source API that provides backend access to Xilinx Vivado EDA using design checkpoints.
+            \item RapidLayout: Hard Block Placer for Systolic Arrays. Built with RapidWright.
+            \item RapidStream: HLS Placer. Built with RapidWright.
         \end{itemize}
@@
     references/surveys,
+    references/rapidwright,
     references/fpga_placement,

### 2024-09-10 docs/proposal/proposal.tex
start actual proposal

@@
 \title{Technical Paper Proposal: \\ Placement Algorithms for Heterogenous FPGAs}
-\author{Brian Cheng \\ Department of Electrical and Computer Engineering}
+\author{Brian B Cheng \\ Department of Electrical and Computer Engineering}
 
@@
 \section{Proposal}
+In the software world, an engineer writes a program in a high-level langauge like C, Java, or Python. 
+Then the engineer compiles the program using some compilation software like gcc or clang. 
+First, the compiler takes the engineer's high-level program, and through multiple stages of lexing, parsing, and optimizations, generates a low-level program in an assembly language like x86 or MIPS. 
+Then, the compiler assembles the assembly code into machine code. 
+The produced machine code is often referred to as a binary and is ready to execute on the machine. 
+Different compilers may have different strategies for this process. 
+For example, LLVM and clang skip the assembly stage and go directly from high-level code to machine code, while gcc sticks to the traditional two-step process. 
+The Java and Python are more complex in that they compile high-level code into an intermediate language as bytecode. 
+In Java, the compiler assembles the bytecode into machine code at runtime in a process called Just-in-Time compilation which allows it to be platform independent. 
+Python offers the choice to assemble the machine code before or at runtime. 
+Simplify this downto 2 sentences.
+
+In the FPGA world, the design entry to product pipeline is despairingly complex by comparison.
+First, the design entry.
+The engineer describes a digital system using a high-level HDL like Verilog or VHDL.
+Then the engineer submits the design entry to an EDA like Vivado or Quartus to perform three stages of automated design.
+These are: synthesis, placement, and routing.
+and then yap about NP-hardness and massive search space and hellish runtime.
 
@@
 \begin{itemize}[label={--}, left=0.25cm] % Adjust label and left indent as desired
-    \item FPGA: Field Programmable Gate Array
-    \item VLSI: Very Large Scale Integration
-    \item EDA: Electronic Design Automation
-    \item VHSIC: Very High Speed Integrated Circuits
-    \item HDL: Hardware Description Language
-    \item VHDL: VHSIC HDL
-    \item HLS: High Level Synthesis
-    \item PL-PS: Programmable Logic - Processing System
-    \item EDIF: Electronic Design Interchange Format
-    \item HPWL: Half Perimeter Wire Length
+    \item \textbf{FPGA}: Field Programmable Gate Array
+    \item \textbf{VLSI}: Very Large Scale Integration
+    \item \textbf{EDA}: Electronic Design Automation
+    \item \textbf{VHSIC}: Very High Speed Integrated Circuits
+    \item \textbf{HDL}: Hardware Description Language
+    \item \textbf{VHDL}: VHSIC HDL
+    \item \textbf{HLS}: High Level Synthesis: Generating synthesizable HDL from high-level software languages.
+A company might want to have a software engineer write C or C++ code and have a program translate it into synthesizable Verilog.
+This can can boost productivity and save the company the need to hire a hardware engineer.
+    \item \textbf{IP}: Intellectual Property: In FPGA context, this means pre-built modules or subsystems like a hardened microprocessor or Ethernet controller. These are usually proprietary.
+    \item \textbf{SoC}: System on Chip: An FPGA device (chip) that features hardened IP in addition to the programmable logic fabric.
+    \item \textbf{PL-PS}: Programmable Logic - Processing System: A design that utilizes the on-chip hard microprocessor in conjunction with the programmable logic fabric.
+    \item \textbf{EDIF}: Electronic Design Interchange Format
+    \item \textbf{HPWL}: Half Perimeter Wire Length
 \end{itemize}
@@
         \begin{itemize}[label={$\cdot$}, left=0.25cm]
-            \item AMD-Xilinx (~50\% FPGA vendor market share)
-            \item Intel-Altera (~35\% share)
+            \item AMD-Xilinx ($\sim$50\% FPGA vendor market share)
+            \item Intel-Altera ($\sim$35\% share)
             \item Lattice
@@
         \begin{itemize}[label={$\cdot$}, left=0.25cm]
-            \item VTR: a PNR tool popular amongst researchers who study placement techniques
-            \item OSS-CAD: a full-flow software suite that includes ABC synthesis, Yosys synthesis, Yosys nextpnr.
-            \item AMF-Placer (Analytical Placer)
-            \item DREAMPlace (VLSI) + DREAMPlaceFPGA
-            \item RapidWright: Semi-open source API that provides backend access to Xilinx Vivado EDA using design checkpoints.
-            \item RapidLayout: Hard Block Placer for Systolic Arrays. Built with RapidWright.
-            \item RapidStream: HLS Placer. Built with RapidWright.
+            \item \textbf{VTR}: Simulated Annealing placer for FPGAs. Popular among researchers who study placement techniques. Commonly referred to as an "academic placer".
+            \item \textbf{OSS-CAD}: a full-flow software suite that includes ABC synthesis, Yosys synthesis, Yosys nextpnr.
+            \item \textbf{AMF-Placer}: Analytical Placer for FPGAs
+            \item \textbf{RapidWright}: Semi-open source API that provides backend access to Xilinx Vivado EDA using design checkpoints.
+            \item \textbf{RapidLayout}: Hard Block Placer for Systolic Arrays. Built with RapidWright.
+            \item \textbf{RapidStream}: HLS Placer. Built with RapidWright.
+            \item \textbf{DREAMPlace}: GPU-powered deep learning placement for VLSI.
+            \item \textbf{DREAMPlaceFPGA}: DREAMPlace, adapted to FPGAs via the RapidWright API.
         \end{itemize}
@@
         \item For each BEL in the netlist, assign the BEL to a Cell, Site, and Tile on the physical FPGA device.
-        \item History of the landscape of placement algorithms:
     \end{itemize}
@@
         \item Takes the \textbf{physical netlist} and maps the connections between BELs onto wires, interconnects, and switchboxes on the FPGA.
+    \end{itemize}
+
+\end{itemize}
+
+\newpage
 \begin{figure}
@@
     \caption{Landscape of VLSI placement techniques (Spindler) \cite{kraftwerk2} }
-            \label{fig:}
+    \label{fig:kraftwerk2}
 \end{figure}
@@
     \caption{Historical timeline of VLSI placement techniques (Markov) \cite{ProCha} }
-            \label{fig:}
+    \label{fig:ProCha}
 \end{figure}
-    \end{itemize}
-
-\end{itemize}
 

### 2024-09-10 docs/proposal/proposal.tex
more proposal work

@@
 \section{Proposal}
-In the software world, an engineer writes a program in a high-level langauge like C, Java, or Python. 
-Then the engineer compiles the program using some compilation software like gcc or clang. 
-First, the compiler takes the engineer's high-level program, and through multiple stages of lexing, parsing, and optimizations, generates a low-level program in an assembly language like x86 or MIPS. 
-Then, the compiler assembles the assembly code into machine code. 
-The produced machine code is often referred to as a binary and is ready to execute on the machine. 
-Different compilers may have different strategies for this process. 
-For example, LLVM and clang skip the assembly stage and go directly from high-level code to machine code, while gcc sticks to the traditional two-step process. 
-The Java and Python are more complex in that they compile high-level code into an intermediate language as bytecode. 
-In Java, the compiler assembles the bytecode into machine code at runtime in a process called Just-in-Time compilation which allows it to be platform independent. 
-Python offers the choice to assemble the machine code before or at runtime. 
-Simplify this downto 2 sentences.
+
+% In the software world, an engineer writes a program in a high-level langauge like C, Java, or Python. 
+% Then the engineer compiles the program using some compilation software like gcc or clang. 
+% First, the compiler takes the engineer's high-level program, and through multiple stages of lexing, parsing, 
+% and optimizations, generates a low-level program in an assembly language like x86 or MIPS. 
+% Then, the compiler assembles the assembly code into machine code. 
+% The produced machine code is often referred to as a binary and is ready to execute on the machine. 
+% Different compilers may have different strategies for this process. 
+% For example, LLVM and clang skip the assembly stage and go directly from high-level code to machine code, while gcc sticks to the traditional two-step process. 
+% The Java and Python are more complex in that they compile high-level code into an intermediate language as bytecode. 
+% In Java, the compiler assembles the bytecode into machine code at runtime in a process called Just-in-Time compilation which allows it to be platform independent. 
+% Python offers the choice to assemble the machine code before or at runtime. 
+% Simplify this downto 2 sentences.
+
+In modern software development, the process of compiling high-level languages like C++ or Java into executable machine code is highly efficient and streamlined. 
+The software toolchains only consist of compilers and scripts and are mostly open source and easy to learn. 
+The compilation is where most of the magic happens.
+The developer can influence a few optimization flags if needed, but for the most part, the developer simply executes the default compiler command.
+Through multiple stages of lexing, parsing, and optimizations, the compiler takes the developer's program and produces working machine code.
+All of this happens within milliseconds to seconds, even for large projects.
+
+In the same way that a compiler takes your high-level C++ file and assembles machine code ready to be executed on a CPU, an Electronic Design Automation (EDA) tool takes your high-level Verilog file and produces a hardware configuration in the form of a bitstream, ready to be deployed onto an FPGA or ASIC.
+In a superficial way, they perform the same task but for different industries.
+One for software, the other for hardware.
+
+Dramatic transition.
 
 In the FPGA world, the design entry to product pipeline is despairingly complex by comparison. 
-First, the design entry.
-The engineer describes a digital system using a high-level HDL like Verilog or VHDL.
+First, the design entry. The engineer describes a digital system using a high-level HDL like Verilog or VHDL. 
 Then the engineer submits the design entry to an EDA like Vivado or Quartus to perform three stages of automated design. 
-These are: synthesis, placement, and routing.
-and then yap about NP-hardness and massive search space and hellish runtime.
+These are: synthesis, placement, and routing. and then yap about NP-hardness and massive search space and hellish runtime.
 
@@
         \begin{itemize}[label={$\cdot$}, left=0.25cm]
-            \item \textbf{VTR}: Simulated Annealing placer for FPGAs. Popular among researchers who study placement techniques. Commonly referred to as an "academic placer".
+            \item \textbf{VTR}: Simulated Annealing placer for FPGAs. Popular among researchers who study placement techniques. 
+                Commonly referred to as an "academic placer".
             \item \textbf{OSS-CAD}: a full-flow software suite that includes ABC synthesis, Yosys synthesis, Yosys nextpnr.

### 2024-09-10 docs/proposal/proposal.tex
more prop

@@
 The compilation is where most of the magic happens.
-The developer can influence a few optimization flags if needed, but for the most part, the developer simply executes the default compiler command.
 Through multiple stages of lexing, parsing, and optimizations, the compiler takes the developer's program and produces working machine code.
+The developer can influence a few optimization flags if needed, but for the most part, the developer simply runs the default compiler command and is presented with the final product.
 All of this happens within milliseconds to seconds, even for large projects.
 
-In the same way that a compiler takes your high-level C++ file and assembles machine code ready to be executed on a CPU, an Electronic Design Automation (EDA) tool takes your high-level Verilog file and produces a hardware configuration in the form of a bitstream, ready to be deployed onto an FPGA or ASIC.
+In the same way that a compiler takes a high-level C++ file and assembles machine code ready to be executed on a CPU, an Electronic Design Automation (EDA) tool takes a high-level Verilog file and produces a hardware configuration in the form of a bitstream, ready to be deployed onto an FPGA or ASIC.
 In a superficial way, they perform the same task but for different industries.
@@
 Then the engineer submits the design entry to an EDA like Vivado or Quartus to perform three stages of automated design. 
-These are: synthesis, placement, and routing. and then yap about NP-hardness and massive search space and hellish runtime.
+These are: synthesis, placement, and routing. 
+All three are NP-hard problems.
+For even modest designs like binary counters or 8-bit CPUs that utilize a small percentage of the available resources on the FPGA, the EDA will spend a minimum of a couple minutes building the design.
+The EDA build time increases exponentially with the scale of the project. 
+Projects that utilize 80\% or more of the FPGA's resources may build for hours, and for the high-end devices like Xilinx's Kintex which boast millions of logical elements, can take days.
+Sometimes the EDA cannot fit the design onto the device.
+Sometimes you step out of the doll-house and the world seems like a heironymous bosch painting.
 
@@
 }
-
 \end{document}

### 2024-09-11 docs/proposal/proposal.tex
more prop
remove comments
more waffling

@@
 
+\section{Keywords}
+\begin{itemize}
+    \item FPGA, EDA, Synthesis, Placement, Routing, Parallel, Optimization
+\end{itemize}
+
 \section{Proposal}
-
-% In the software world, an engineer writes a program in a high-level langauge like C, Java, or Python. 
-% Then the engineer compiles the program using some compilation software like gcc or clang. 
-% First, the compiler takes the engineer's high-level program, and through multiple stages of lexing, parsing, 
-% and optimizations, generates a low-level program in an assembly language like x86 or MIPS. 
-% Then, the compiler assembles the assembly code into machine code. 
-% The produced machine code is often referred to as a binary and is ready to execute on the machine. 
-% Different compilers may have different strategies for this process. 
-% For example, LLVM and clang skip the assembly stage and go directly from high-level code to machine code, while gcc sticks to the traditional two-step process. 
-% The Java and Python are more complex in that they compile high-level code into an intermediate language as bytecode. 
-% In Java, the compiler assembles the bytecode into machine code at runtime in a process called Just-in-Time compilation which allows it to be platform independent. 
-% Python offers the choice to assemble the machine code before or at runtime. 
-% Simplify this downto 2 sentences.
-
 In modern software development, the process of compiling high-level languages like C++ or Java into executable machine code is highly efficient and streamlined. 
@@
 
-Dramatic transition.
-
 In the FPGA world, the design entry to product pipeline is despairingly complex by comparison. 
@@
 All three are NP-hard problems.
-For even modest designs like binary counters or 8-bit CPUs that utilize a small percentage of the available resources on the FPGA, the EDA will spend a minimum of a couple minutes building the design.
-The EDA build time increases exponentially with the scale of the project. 
-Projects that utilize 80\% or more of the FPGA's resources may build for hours, and for the high-end devices like Xilinx's Kintex which boast millions of logical elements, can take days.
-Sometimes the EDA cannot fit the design onto the device.
-Sometimes you step out of the doll-house and the world seems like a heironymous bosch painting.
+For even modest designs like binary counters or 8-bit CPUs that utilize a small percentage of the available resources on the FPGA, the EDA will spend a minimum of a couple minutes synthesizing and implementing the design.
+The EDA run time increases exponentially with the scale of the project. 
+Projects that utilize 80\% or more of the FPGA's resources may run for hours, and for the high-end devices like Xilinx's Kintex FPGAs which have millions of logical elements, can run for days.
+At high utilization, sometimes the design EDA cannot fit the design onto the device due to routing congestion even when it has enough logical elements to implement the design, essentially wasting hours or days of runtime.
+
+One of the most common complaints from those in the FPGA design industry, especially from those in software development, is that the FPGA toolchains are overall slow and heavy. 
+Unlike in software development where an error in the program can be patched and recompiled in a matter of minutes, EDA runtimes lasting several hours means that an engineer can only go through a couple design cycles per day.
+A Verilog developer must be very precise with their coding and practice thorough verification of their design before submitting it to the EDA.
+This overall heightens the barrier of entry into FPGA development and contributes to a shortage of qualified FPGA engineers.
+
+This paper aims to attack one of the sections of this wall: the placement stage. Like synthesis and routing, it is an NP-hard problem. The search space for a placed design is massive.
 
 \section{Abbreviations}
-\begin{itemize}[label={--}, left=0.25cm] % Adjust label and left indent as desired
+\begin{itemize}[label={--}, left=0.25cm]
     \item \textbf{FPGA}: Field Programmable Gate Array
@@
     \item \textbf{HPWL}: Half Perimeter Wire Length
-\end{itemize}
-
-\section{Keywords}
-\begin{itemize}
-    \item FPGA, EDA, Synthesis, Placement, Routing, Parallel, Optimization
 \end{itemize}

### 2024-09-11 docs/proposal/proposal.tex
waffles
more proposal

@@
 \section{Proposal}
-In modern software development, the process of compiling high-level languages like C++ or Java into executable machine code is highly efficient and streamlined. 
-The software toolchains only consist of compilers and scripts and are mostly open source and easy to learn. 
-The compilation is where most of the magic happens.
-Through multiple stages of lexing, parsing, and optimizations, the compiler takes the developer's program and produces working machine code.
-The developer can influence a few optimization flags if needed, but for the most part, the developer simply runs the default compiler command and is presented with the final product.
-All of this happens within milliseconds to seconds, even for large projects.
+A compiler takes a high-level program like a Java or Python file and assembles a corresponding machine code which is ready to be executed on a CPU.
+In similar fashion, an Electronic Design Automation tool (EDA) takes a high-level description of a digital system written in Verilog or VHDL and produces a bitstream which is ready to be deployed to an FPGA.
+In a superficial way, compilers and EDAs perform the same task for engineers but in different industries. 
+One for software, the latter for hardware.
 
-In the same way that a compiler takes a high-level C++ file and assembles machine code ready to be executed on a CPU, an Electronic Design Automation (EDA) tool takes a high-level Verilog file and produces a hardware configuration in the form of a bitstream, ready to be deployed onto an FPGA or ASIC.
-In a superficial way, they perform the same task but for different industries.
-One for software, the other for hardware.
+% Software compilers have become incredibly optimized over the decades and have allowed software engineers to be very productive. 
+% The speed and user-friendliness that compilers offer allows engineers to push dozens, even hundreds of design cycles per day.
+% Bugs and design changes can be patched and implemented into the program and recompiled in a matter of milliseconds to minutes. 
+% Modern compilers are so refined that compilation can be regarded as where the "magic" happens in the software design cycle.
+% Software engineers typically do not concern themselves with how their program gets compiled.
+% Through multiple stages of lexing, parsing, and optimizations, the compiler takes the developer's program and produces working machine code.
+% The developer can influence a few optimization flags if needed, but for the most part, the developer simply runs the default compiler command and is presented with the final product.
+% All of this happens within milliseconds to seconds, even for large projects. 
+% monotone, sentence length
 
-In the FPGA world, the design entry to product pipeline is despairingly complex by comparison. 
-First, the design entry. The engineer describes a digital system using a high-level HDL like Verilog or VHDL. 
+Software compilers have evolved to become highly optimized, significantly boosting the productivity of software developers. 
+The compilers are so robust that developers rarely need to worry about the correctness or efficiency of their machine code, and are so refined that it all happens in a matter of milliseconds to minutes, even for large projects.
+% break this up?
+The speed of compilation enables rapid debugging and implementation of design changes and allows developers to iterate through dozens or even hundreds of design cycles per day.
+Instead of tinkering around with assembly code, software engineers can comfortably focus on the high-level abstractions of their design.
+
+% Compilation, which takes mere milliseconds to minutes, enables rapid debugging and implementation of design changes. 
+% The compilation process has become so streamlined that it can be considered the "magic" of the software development cycle. 
+% They can influence a few optimization flags when needed, but for the most part, the developer simply runs the default compiler command.
+% The compiler handles the lexing, parsing, and optimization automatically, with minimal input from the developer who typically relies on default settings unless specific optimization flags are needed.
+% Even for large-scale projects, this process is fast and efficient, allowing engineers to focus on higher-level design rather than low-level details.
+
+The compilation stages mainly consists of lexing, parsing, and instruction optimization.
+While these stages can be complex from a design standpoint, they are, mathematically, trivial.
+The EDA stages are despairingly complex by comparison. 
+First, the FPGA engineer creates and describes a digital design using a hardware description language (HDL) like Verilog or VHDL.     
 Then the engineer submits the design entry to an EDA like Vivado or Quartus to perform three stages of automated design. 
@@
 All three are NP-hard problems.
+
 For even modest designs like binary counters or 8-bit CPUs that utilize a small percentage of the available resources on the FPGA, the EDA will spend a minimum of a couple minutes synthesizing and implementing the design.
@@
 
-One of the most common complaints from those in the FPGA design industry, especially from those in software development, is that the FPGA toolchains are overall slow and heavy. 
+One of the most common complaints from new FPGA engineers, especially those coming from software development, is that the FPGA toolchains are overall slow and heavy.
+This is because the problems that the EDA must solve are inherently complex.
 Unlike in software development where an error in the program can be patched and recompiled in a matter of minutes, EDA runtimes lasting several hours means that an engineer can only go through a couple design cycles per day.
-A Verilog developer must be very precise with their coding and practice thorough verification of their design before submitting it to the EDA.
+An FPGA engineer must be very precise with their coding and practice thorough verification of their design before submitting it to the EDA.
 This overall heightens the barrier of entry into FPGA development and contributes to a shortage of qualified FPGA engineers.
 
-This paper aims to attack one of the sections of this wall: the placement stage. Like synthesis and routing, it is an NP-hard problem. The search space for a placed design is massive.
+This paper aims to attack one of the sections of this wall: the placement stage.
+First, it will survey the current landscape of placement algorithms and strategies.
+Like synthesis and routing, it is an NP-hard problem.
+The search space for a well placed design is massive.
 

### 2024-09-12 docs/proposal/proposal.tex
start talking about basic technicals

@@
 \usepackage{graphicx}
+\usepackage{amsmath}
 
@@
 
-For even modest designs like binary counters or 8-bit CPUs that utilize a small percentage of the available resources on the FPGA, the EDA will spend a minimum of a couple minutes synthesizing and implementing the design.
+For even modest designs that utilize a small percentage of the available resources on the FPGA, the EDA will spend a minimum of a couple minutes synthesizing and implementing the design.
 The EDA run time increases exponentially with the scale of the project. 
 Projects that utilize 80\% or more of the FPGA's resources may run for hours, and for the high-end devices like Xilinx's Kintex FPGAs which have millions of logical elements, can run for days.
-At high utilization, sometimes the design EDA cannot fit the design onto the device due to routing congestion even when it has enough logical elements to implement the design, essentially wasting hours or days of runtime.
+At high utilization, there is a possibility that the EDA cannot fit the design onto the device due to routing congestion even when it has enough logical elements to implement the design. 
+The EDA might only reach this conclusion after attempting to fit the design for several hours.
 
@@
 An FPGA engineer must be very precise with their coding and practice thorough verification of their design before submitting it to the EDA.
-This overall heightens the barrier of entry into FPGA development and contributes to a shortage of qualified FPGA engineers.
+This overall heightens the barrier of entry into FPGA development and contributes to a shortage of qualified FPGA engineers and a limitation of their productivity.
 
-This paper aims to attack one of the sections of this wall: the placement stage.
-First, it will survey the current landscape of placement algorithms and strategies.
-Like synthesis and routing, it is an NP-hard problem.
-The search space for a well placed design is massive.
+This paper aims to study one particular pillar of this barrier: the placement stage. 
+First, it will review prerequisite knowledge for placement algorithms. 
+This includes graph theory to represent logical elements of a digital design and the wired connections between them, and a corresponding optimization problem with a cost function which we seek to minimize.
+
+In mathematics, a graph is usually modeled as \( G = (V, E) \), where \( V \) represents the set of vertices and \( E \) represents the set of edges between them.
+A hypergraph is denoted as \( G_{H} = (V_{H}, E_{H}) \), where \( V_{H} \) represents the set of vertices, and \( E_{H} \) represents the set of hyperedges, which are edges that can connect more than two vertices.
+Electronic circuits are typically represented as hypergraphs where % FINISH ME
+In FPGA design, the graph is often modeled using a "netlist".
+A netlist is a graph \( G = (V, E) \), where \( V \) represents the components (e.g., logic blocks or gates), and \( E \) represents the connections or nets between them.
+
+The vast majority of existing placers use the Half Perimeter Wire Length (HPWL) cost function or one of its various flavors - manhattan distance, euclidian distance, beta-regularized, etc..
+The basic HPWL models the problem as the summation of the manhattan lengths of all nets in the netlist.
+
+\begin{equation}
+    \text{HPWL} = \sum_{i=1}^{n} \left( \max_{j \in \text{net}_i} (x_j) - \min_{j \in \text{net}_i} (x_j) + \max_{j \in \text{net}_i} (y_j) - \min_{j \in \text{net}_i} (y_j) \right)
+\end{equation}
+
+Then, the paper will survey the current landscape of placement algorithms and strategies. 
+This includes Simulated Annealing (SA), Analytical Placement (AP),
 
@@
             \item \textbf{AMF-Placer}: Analytical Placer for FPGAs
-            \item \textbf{RapidWright}: Semi-open source API that provides backend access to Xilinx Vivado EDA using design checkpoints.
+            \item \textbf{RapidWright}: Semi-open source Java API that provides backend access to Xilinx Vivado EDA using design checkpoints.
             \item \textbf{RapidLayout}: Hard Block Placer for Systolic Arrays. Built with RapidWright.

### 2024-09-12 docs/proposal/proposal.tex
minor work

@@
 
-In mathematics, a graph is usually modeled as \( G = (V, E) \), where \( V \) represents the set of vertices and \( E \) represents the set of edges between them.
+In mathematics, a graph is modeled as \( G = (V, E) \), where \( V \) represents the set of vertices and \( E \) represents the set of edges between them.
 A hypergraph is denoted as \( G_{H} = (V_{H}, E_{H}) \), where \( V_{H} \) represents the set of vertices, and \( E_{H} \) represents the set of hyperedges, which are edges that can connect more than two vertices.
-Electronic circuits are typically represented as hypergraphs where % FINISH ME
-In FPGA design, the graph is often modeled using a "netlist".
+Electronic circuits are represented as hypergraphs as because a voltage source can have one or many sinks. 
+In FPGA design, the hypergraph is often transform into something called a "netlist", which is a hypergraph that has been flattened down into a graph. 
+For every hyperedge in the hypergraph, transform the hyperedge into a collection of 2-pin edges - one source to one sink.
+There are different strategies for this reduction - unoriented star model, clique model, etc.. 
 A netlist is a graph \( G = (V, E) \), where \( V \) represents the components (e.g., logic blocks or gates), and \( E \) represents the connections or nets between them.
@@
 The vast majority of existing placers use the Half Perimeter Wire Length (HPWL) cost function or one of its various flavors - manhattan distance, euclidian distance, beta-regularized, etc..
-The basic HPWL models the problem as the summation of the manhattan lengths of all nets in the netlist.
+The basic HPWL models the cost function as the summation of the manhattan lengths of all nets in the netlist.
+
+This paper will primarily use Xilinx's open-source API, RapidWright to implement various custom placement algorithms.
+RapidWright is a Java API that provides backend access to Xilinx's EDA, Vivado.
 
@@
 Then, the paper will survey the current landscape of placement algorithms and strategies. 
-This includes Simulated Annealing (SA), Analytical Placement (AP),
+This includes Simulated Annealing (SA) which is stochastic, but will focus on Analytical Placement (AP).
 

### 2024-09-12 docs/proposal/proposal.tex
lots more prop work

--- 2024-09-12 docs/proposal/proposal.tex
@@
 \section{Proposal}
-A compiler takes a high-level program like a Java or Python file and assembles a corresponding machine code which is ready to be executed on a CPU.
+A compiler takes a program written in high-level programming language like a Java or Python and assembles low-level machine code which is ready to be executed on a CPU.
 In similar fashion, an Electronic Design Automation tool (EDA) takes a high-level description of a digital system written in Verilog or VHDL and produces a bitstream which is ready to be deployed to an FPGA.
@@
 
-% Software compilers have become incredibly optimized over the decades and have allowed software engineers to be very productive. 
-% The speed and user-friendliness that compilers offer allows engineers to push dozens, even hundreds of design cycles per day.
-% Bugs and design changes can be patched and implemented into the program and recompiled in a matter of milliseconds to minutes. 
-% Modern compilers are so refined that compilation can be regarded as where the "magic" happens in the software design cycle.
-% Software engineers typically do not concern themselves with how their program gets compiled.
-% Through multiple stages of lexing, parsing, and optimizations, the compiler takes the developer's program and produces working machine code.
-% The developer can influence a few optimization flags if needed, but for the most part, the developer simply runs the default compiler command and is presented with the final product.
-% All of this happens within milliseconds to seconds, even for large projects. 
-% monotone, sentence length
-
 Software compilers have evolved to become highly optimized, significantly boosting the productivity of software developers. 
@@
 % break this up?
-The speed of compilation enables rapid debugging and implementation of design changes and allows developers to iterate through dozens or even hundreds of design cycles per day.
+The speed of compilation enables rapid debugging and implementation of new features and allows developers to iterate through dozens or even hundreds of design cycles per day.
 Instead of tinkering around with assembly code, software engineers can comfortably focus on the high-level abstractions of their design.
 
-% Compilation, which takes mere milliseconds to minutes, enables rapid debugging and implementation of design changes. 
-% The compilation process has become so streamlined that it can be considered the "magic" of the software development cycle. 
-% They can influence a few optimization flags when needed, but for the most part, the developer simply runs the default compiler command.
-% The compiler handles the lexing, parsing, and optimization automatically, with minimal input from the developer who typically relies on default settings unless specific optimization flags are needed.
-% Even for large-scale projects, this process is fast and efficient, allowing engineers to focus on higher-level design rather than low-level details.
-
 The compilation stages mainly consists of lexing, parsing, and instruction optimization.
-While these stages can be complex from a design standpoint, they are, mathematically, trivial.
-The EDA stages are despairingly complex by comparison. 
+While these stages are complex from a systems standpoint, they are, mathematically, trivial.
+The EDA stages on the other hand are deeply mathematically complex. 
 First, the FPGA engineer creates and describes a digital design using a hardware description language (HDL) like Verilog or VHDL.     
 Then the engineer submits the design entry to an EDA like Vivado or Quartus to perform three stages of automated design. 
-These are: synthesis, placement, and routing. 
+These are mainly: synthesis, placement, and routing. 
 All three are NP-hard problems.
 
-For even modest designs that utilize a small percentage of the available resources on the FPGA, the EDA will spend a minimum of a couple minutes synthesizing and implementing the design.
+For even modest designs that utilize a small percentage of the available resources on the FPGA, the EDA will spend a minimum two to three minutes running synthesis and implementation.
 The EDA run time increases exponentially with the scale of the project. 
@@
 
-One of the most common complaints from new FPGA engineers, especially those coming from software development, is that the FPGA toolchains are overall slow and heavy.
+One of the most common complaints from new FPGA engineers, especially those coming from software development, is that the FPGA toolchains are slow and heavy.
 This is because the problems that the EDA must solve are inherently complex.
@@
 This paper aims to study one particular pillar of this barrier: the placement stage. 
-First, it will review prerequisite knowledge for placement algorithms. 
+First, it will review prerequisite knowledge for placement strategies. 
 This includes graph theory to represent logical elements of a digital design and the wired connections between them, and a corresponding optimization problem with a cost function which we seek to minimize.
 
-In mathematics, a graph is usually modeled as \( G = (V, E) \), where \( V \) represents the set of vertices and \( E \) represents the set of edges between them.
+Second, it will briefly explore the historical progress made on placement algorithms for both VLSI and FPGA and then explore the current trends in placement research.
+
+Then third, the paper will outline my own work on placement strategies, whether they be reproductions of existing strategies, or if I manage to think of a novel strategy. 
+I intend to use Xilinx's open-source API, RapidWright, which is written in Java and provides backend access to Xilinx's FPGA EDA, Vivado.
+
+\section{Foundational Knowledge (so far)}
+In mathematics, a graph is modeled as \( G = (V, E) \), where \( V \) represents the set of vertices and \( E \) represents the set of edges between them.
 A hypergraph is denoted as \( G_{H} = (V_{H}, E_{H}) \), where \( V_{H} \) represents the set of vertices, and \( E_{H} \) represents the set of hyperedges, which are edges that can connect more than two vertices.
-Electronic circuits are typically represented as hypergraphs where % FINISH ME
-In FPGA design, the graph is often modeled using a "netlist".
-A netlist is a graph \( G = (V, E) \), where \( V \) represents the components (e.g., logic blocks or gates), and \( E \) represents the connections or nets between them.
+Electronic circuits are typically represented as hypergraphs because a voltage source pin can have one or many sink pins. 
+In FPGA and VLSI design, the hypergraph is transformed into something called a "netlist", which is a hypergraph that has been flattened down into a graph. 
+For every hyperedge in the hypergraph, transform the hyperedge into a set of 2-pin edges, each having one source to one sink.
+The resulting graph can be represented by \( G = (V, E) \), where \( V \) represnts the ports (pins) of all modules (logic gates, )
+There are different strategies for this reduction - unoriented star model, clique model, etc.. 
+\cite{AP_2000}
 
-The vast majority of existing placers use the Half Perimeter Wire Length (HPWL) cost function or one of its various flavors - manhattan distance, euclidian distance, beta-regularized, etc..
-The basic HPWL models the problem as the summation of the manhattan lengths of all nets in the netlist.
+The vast majority of existing placers use the Half Perimeter Wire Length (HPWL) cost function in one of its various flavors - manhattan distance, euclidian distance, bounding box, etc..
+The manhattan and euclidian HPWL models the cost function as the summation of the lengths of all nets in the netlist. 
+The manhattan HPWL is convex while the euclidian HPWL is strictly convex.
+We can use smooth methods to minimize the euclidian, however, it generally produces worse Quality of Result (QoR) than the manhattan \cite{AP_2012}.
+The bounding box HPWL models the cost function as simply the longest net in the netlist.
+Shown below in (1) and (2) are the manhattan and euclidian distance HPWL objective functions respectively.
+Shown in (3) is the bounding box version.
 
 \begin{equation}
-    \text{HPWL} = \sum_{i=1}^{n} \left( \max_{j \in \text{net}_i} (x_j) - \min_{j \in \text{net}_i} (x_j) + \max_{j \in \text{net}_i} (y_j) - \min_{j \in \text{net}_i} (y_j) \right)
+    \Phi(\vec{x}, \vec{y}) = \sum_{i,j} w_{i,j} \left[ (x_i - x_j)^2 + (y_i - y_j)^2 \right]
+    \label{eq:}
 \end{equation}
 
-Then, the paper will survey the current landscape of placement algorithms and strategies. 
-This includes Simulated Annealing (SA), Analytical Placement (AP),
+\begin{equation}
+    \Phi(\vec{x}, \vec{y}) = \sum_{i,j} w_{i,j} \left( |x_i - x_j| + |y_i - y_j| \right)
+    \label{eq:}
+\end{equation}
+
+\begin{equation}
+    \Phi(\vec{x}, \vec{y}) = \sum_{i=1}^{n} \left( \max_{j \in e_i} (x_j) - \min_{j \in e_i} (x_j) + \max_{j \in e_i} (y_j) - \min_{j \in e_i} (y_j) \right)
+    \label{eq:}
+\end{equation}
+
+\( x = \{ x_{1}, x_{2}, ..., x_{N} \} \) represents the set of physical x-coordinates of the logic block pins on the chip and \( y = \{ y_{1}, y_{2}, ..., y_{N} \} \) represents the set of y-coordinates.
+The units for these coordinates can be nanometers in VLSI design, but for FPGAs where the logic blocks are uniform size and arranged in a grid, the coordinates can simply be positive integers.
+\( N \) is the total number of logic block pins in the design. \( E = \{ e_{1}, e_{2}, ..., e_{M} \} \) represents the netlist with M total nets.
+
+We want to minimize the total wirelength in our design for two reasons: power consumption, and clock skew, and routing.
+FPGAs are commonly used for mission critical applications and edge computing devices that are required to run on low power.
+We can reduce the power consumption of the FPGA by reducing the power dissipation of current running through wires.
+This is particularly important for FPGAs because signal paths must propagate not only through wires but also switchboxes and programmable interconnects which can also dissipate power.
+The second and more important reason is to minimize clock skew. 
+Signal propagation through wires and transistors takes time. 
+The clock signal is crucial for state transition in an FSM.
+If the clock signal drives two logic blocks that were placed too far apart, they may become desynchronized as the same clock edge may reach the two flip-flops at different clock periods.
+All flip-flops operating on the same clock domain must strike at the same nanosecond.
+
+The third and equally relevant reason is the simplify the routing stage which takes place after placement.
+If the logic blocks are all placed close together but in such a way that the wiring between them will become spaghetti, the router will produce an over-congested wiring and potentially risk routing failure because of it.
+The placer ideally must cluster relevant logic blocks into semi-islolated neighborhood such that the wiring in one neighborhood to another is minimized.
 

### 2024-09-12 docs/proposal/proposal.tex
final touches

@@
     In a superficial way, compilers and EDAs perform the same task for engineers but in different industries. 
-One for software, the latter for hardware.
+    One for software, the other for hardware.
 
@@
 
-Second, it will briefly explore the historical progress made on placement algorithms for both VLSI and FPGA and then explore the current trends in placement research.
+    Second, it will briefly explore the historical progress made on placement algorithms for both VLSI and FPGA and then explore the current trends in open-source placement research 
+    \cite{Yosys}, \cite{AMFPlacer2}, \cite{RapidLayout}, \cite{RapidStream}, \cite{DREAMPlace}, \cite{DREAMPlaceFPGA}.
+    I would have liked to explore the current trends in proprietary EDAs but they are unfortunately closed source.
 
-Then third, the paper will outline my own work on placement strategies, whether they be reproductions of existing strategies, or if I manage to think of a novel strategy. 
+    Then third, the paper will outline my own work on placement strategies, whether they be reproductions of existing strategies, or a novel strategy if I manage to think of one (unlikely). 
     I intend to use Xilinx's open-source API, RapidWright, which is written in Java and provides backend access to Xilinx's FPGA EDA, Vivado. 
+    I have already familiarized myself with the API and, over the past month have practiced using the Vivado + RapidWright toolchain to do basic tasks like extracting the netlist of a synthesized VHDL design.
+    You can see my preliminary work in my Github: \url{https://github.com/TotoroTron/place-and-route}.
 
@@
     }
+
 \end{document}

### 2024-09-12 docs/proposal/proposal.tex
finaller touches

@@
     The manhattan HPWL is convex while the euclidian HPWL is strictly convex.
-    We can use smooth methods to minimize the euclidian, however, it generally produces worse Quality of Result (QoR) than the manhattan \cite{AP_2012}.
+    We can use smooth methods to minimize the euclidian, however, it produces worse Quality of Result (QoR) than the manhattan \cite{AP_2012}, which is something to consider.
     The bounding box HPWL models the cost function as simply the longest net in the netlist.
@@
     \begin{equation}
-        \Phi(\vec{x}, \vec{y}) = \sum_{i,j} w_{i,j} \left[ (x_i - x_j)^2 + (y_i - y_j)^2 \right]
-        \label{eq:}
+        \Phi(\vec{x}, \vec{y}) = \sum_{i,j} w_{i,j} \left( |x_i - x_j| + |y_i - y_j| \right)
+        \label{Manhattan}
     \end{equation}
@@
     \begin{equation}
-        \Phi(\vec{x}, \vec{y}) = \sum_{i,j} w_{i,j} \left( |x_i - x_j| + |y_i - y_j| \right)
-        \label{eq:}
+        \Phi(\vec{x}, \vec{y}) = \sum_{i,j} w_{i,j} \left[ (x_i - x_j)^2 + (y_i - y_j)^2 \right]
+        \label{Euclidian}
     \end{equation}
@@
         \Phi(\vec{x}, \vec{y}) = \sum_{i=1}^{n} \left( \max_{j \in e_i} (x_j) - \min_{j \in e_i} (x_j) + \max_{j \in e_i} (y_j) - \min_{j \in e_i} (y_j) \right)
-        \label{eq:}
+        \label{BoundingBox}
     \end{equation}

### 2024-09-13 docs/proposal/proposal.tex
needs more editing

@@
 
-    The compilation stages mainly consists of lexing, parsing, and instruction optimization.
-    While these stages are complex from a systems standpoint, they are, mathematically, trivial.
-    The EDA stages on the other hand are deeply mathematically complex. 
-    First, the FPGA engineer creates and describes a digital design using a hardware description language (HDL) like Verilog or VHDL.     
-    Then the engineer submits the design entry to an EDA like Vivado or Quartus to perform three stages of automated design. 
-    These are mainly: synthesis, placement, and routing. 
-    All three are NP-hard problems.
+    % The compilation stages mainly consists of lexing, parsing, and instruction optimization which is NP-hard.
+    % Mathematically, the optimization is NP-hard.
+    % The search space for optimizations, while large, is typically constrained to the instruction set and a relatively small number of optimization passes like register allocation and instruction scheduling.
+    The compilation stages mainly consists of lexing, parsing, and optimization.
+    The optimization stage is where most of the complexity occurs as it deals with NP-hard problems.
+    The compiler at this stage must optimize the ordering of the machine instructions to take advantage of pipelining, register allocation.
+    While the search space for the best possible machine code is massive, it is typically constrained to the size of the CPU instruction set and architecture which are fixed.
+    Furthermore, compiler makers have developed highly efficient heuristics to approximate the best machine code within a reasonable time.
+
+    On the other hand, the search space for the EDA stages are much more massive. 
+    These are: synthesis, placement, and routing, and all three are NP-hard with massive search spaces.
+
+    % First, the FPGA engineer creates and describes a digital design using a hardware description language (HDL) like Verilog or VHDL.     
+    % Then the engineer submits the design entry to an EDA like Vivado or Quartus to perform three stages of automated design.
 

### 2024-09-13 docs/proposal/proposal.tex
more editing

--- 2024-09-12 docs/proposal/proposal.tex
@@
 
-    The compilation stages mainly consists of lexing, parsing, and instruction optimization.
-    While these stages are complex from a systems standpoint, they are, mathematically, trivial.
-    The EDA stages on the other hand are deeply mathematically complex. 
-    First, the FPGA engineer creates and describes a digital design using a hardware description language (HDL) like Verilog or VHDL.     
-    Then the engineer submits the design entry to an EDA like Vivado or Quartus to perform three stages of automated design. 
-    These are mainly: synthesis, placement, and routing. 
-    All three are NP-hard problems.
+    The compilation stages mainly consists of lexing, parsing, and optimization.
+    The optimization stage is where most of the complexity occurs as it deals with NP-hard problems.
+    The compiler at this stage must optimize the ordering of the machine instructions to maximize pipelining and register allocation.
+    While the search space for the best possible machine code is massive, it is typically constrained to the size of the CPU instruction set and architecture which are fixed.
+    Furthermore, compiler makers have developed highly efficient heuristics to approximate the best machine code within an acceptable runtime.
+
+    On the other hand, the EDA stages consist mainly of synthesis, placement, and routing.
+    All three are NP-hard optimization problems with massive search spaces with massive search spaces.
+    Even with heuristics and approximations to circumvent NP-hardness, the size of their search spaces make the EDA runtimes less than acceptable.
+
+    In a typical FPGA design flow, the engineer creates and describes a digital design using a hardware description language (HDL) like Verilog or VHDL.     
+    Then, the FPGA engineer verifies the design by creating a testbench that wraps around the design and feeds the design input signals and observes the output signals.
+    % The formal verification process is an art of its own and is so matured that companies will often separate their engineers into two departments - one focusing on design and the other on verification.
+    Then the engineer submits the design entry to an EDA like Vivado or Quartus to perform the three stages of automated design.
 

### 2024-09-13 docs/proposal/proposal.tex
minor editing

@@
     The compiler at this stage must optimize the ordering of the machine instructions to maximize pipelining and register allocation.
-    While the search space for the best possible machine code is massive, it is typically constrained to the size of the CPU instruction set and architecture which are fixed.
+    While the search space for the best possible machine code is massive, it is typically constrained by the size of the CPU instruction set and architecture which are fixed.
     Furthermore, compiler makers have developed highly efficient heuristics to approximate the best machine code within an acceptable runtime.
@@
     On the other hand, the EDA stages consist mainly of synthesis, placement, and routing.
-    All three are NP-hard optimization problems with massive search spaces with massive search spaces.
-    Even with heuristics and approximations to circumvent NP-hardness, the size of their search spaces make the EDA runtimes less than acceptable.
+    % Placement and routing is often packaged into one big stage referred to as "implementation" as is the case in Vivado.
+    All three are NP-hard optimization problems with massive search spaces.
+    Even with heuristics and approximations to circumvent NP-hardness, the size of their search spaces make the EDA runtimes less than desirable.
 
@@
     Then the engineer submits the design entry to an EDA like Vivado or Quartus to perform the three stages of automated design. 
-
     For even modest designs that utilize a small percentage of the available resources on the FPGA, the EDA will spend a minimum two to three minutes running synthesis and implementation.
@@
 
-    Then third, the paper will outline my own work on placement strategies, whether they be reproductions of existing strategies, or a novel strategy if I manage to think of one (unlikely). 
+    Then third, the paper will outline my own work on placement strategies, whether they be reproductions of existing strategies, or a novel strategy if I manage to create one in the span of 3 months (unlikely).
     I intend to use Xilinx's open-source API, RapidWright, which is written in Java and provides backend access to Xilinx's FPGA EDA, Vivado. 

### 2024-09-13 docs/proposal/proposal.tex
even more editing

@@
     The compilation stages mainly consists of lexing, parsing, and optimization.
-    The optimization stage is where most of the complexity occurs as it deals with NP-hard problems.
-    The compiler at this stage must optimize the ordering of the machine instructions to maximize pipelining and register allocation.
-    While the search space for the best possible machine code is massive, it is typically constrained by the size of the CPU instruction set and architecture which are fixed.
+    The optimization stage is where most of the complexity occurs as it can deal with NP-complete problems, depending on what optimization flags the developer set when running compile.
+    The compiler at this stage must optimize loop unrolling and code pruning as well as the ordering of the machine instructions to maximize pipelining and register allocation.
+    While the search space for the best possible machine code is large, it is typically constrained by the size of the CPU instruction set and architecture which are fixed.
     Furthermore, compiler makers have developed highly efficient heuristics to approximate the best machine code within an acceptable runtime.

### 2024-09-14 docs/proposal/proposal.tex
final edits

@@
     The speed of compilation enables rapid debugging and implementation of new features and allows developers to iterate through dozens or even hundreds of design cycles per day.
-    Instead of tinkering around with assembly code, software engineers can comfortably focus on the high-level abstractions of their design.
+    Instead of tinkering with assembly code, software engineers can comfortably focus on the high-level abstractions of their design.
 
@@
     The compiler at this stage must optimize loop unrolling and code pruning as well as the ordering of the machine instructions to maximize pipelining and register allocation.
-    While the search space for the best possible machine code is large, it is typically constrained by the size of the CPU instruction set and architecture which are fixed.
+    While the search space for the best possible machine code is large, it is typically constrained by the size of the CPU instruction set and CPU architecture which are fixed.
     Furthermore, compiler makers have developed highly efficient heuristics to approximate the best machine code within an acceptable runtime.
@@
 
-    Second, it will briefly explore the historical progress made on placement algorithms for both VLSI and FPGA and then explore the current trends in open-source placement research 
+    Second, it will briefly explore the historical progress made on placement algorithms for both VLSI and FPGA \cite{ProCha} and then explore the current trends in open-source placement research 
     \cite{Yosys}, \cite{AMFPlacer2}, \cite{RapidLayout}, \cite{RapidStream}, \cite{DREAMPlace}, \cite{DREAMPlaceFPGA}.
@@
 
-    Then third, the paper will outline my own work on placement strategies, whether they be reproductions of existing strategies, or a novel strategy if I manage to create one in the span of 3 months (unlikely).
-    I intend to use Xilinx's open-source API, RapidWright, which is written in Java and provides backend access to Xilinx's FPGA EDA, Vivado. 
-    I have already familiarized myself with the API and, over the past month have practiced using the Vivado + RapidWright toolchain to do basic tasks like extracting the netlist of a synthesized VHDL design.
+    And third, the paper will outline my own technical work on placement strategies, whether they be reproductions of existing strategies, or a novel strategy if I manage to create one in the span of 3 months.
+    This will include performance analysis and comparisons among different placement algorithms and strategies.
+    To achieve this, I will use Xilinx's open-source API, RapidWright which provides backend access to Vivado which is Xilinx's EDA for FPGAs. 
+    I have already familiarized myself with the API and and able to use the Vivado + RapidWright toolchain to do basic tasks like extracting the netlist of a synthesized VHDL design or performing manual placement of logic elements.
     You can see my preliminary work in my Github: \url{https://github.com/TotoroTron/place-and-route}.
@@
 \section{Foundational Knowledge (so far)}
-    In mathematics, a graph is modeled as \( G = (V, E) \), where \( V \) represents the set of vertices and \( E \) represents the set of edges between them.
+    A graph is modeled as \( G = (V, E) \), where \( V \) represents the set of vertices and \( E \) represents the set of edges between them.
     A hypergraph is denoted as \( G_{H} = (V_{H}, E_{H}) \), where \( V_{H} \) represents the set of vertices, and \( E_{H} \) represents the set of hyperedges, which are edges that can connect more than two vertices.
@@
     The manhattan HPWL is convex while the euclidian HPWL is strictly convex.
-    We can use smooth methods to minimize the euclidian, however, it produces worse Quality of Result (QoR) than the manhattan \cite{AP_2012}, which is something to consider.
+    We can use smooth methods to minimize the euclidian, however, it produces worse Quality of Result (QoR) than the manhattan \cite{AP_2012}.
     The bounding box HPWL models the cost function as simply the longest net in the netlist.
@@
     The units for these coordinates can be nanometers in VLSI design, but for FPGAs where the logic blocks are uniform size and arranged in a grid, the coordinates can simply be positive integers.
-    \( N \) is the total number of logic block pins in the design. \( E = \{ e_{1}, e_{2}, ..., e_{M} \} \) represents the netlist with M total nets.
+    \( N \) is the total number of logic block pins in the design. 
+    \( E = \{ e_{1}, e_{2}, ..., e_{M} \} \) represents the netlist with M total nets.
 
-    We want to minimize the total wirelength in our design for two reasons: power consumption, and clock skew, and routing.
+    We want to minimize the total wirelength in our design for three reasons: power, clock skew, and routing.
     FPGAs are commonly used for mission critical applications and edge computing devices that are required to run on low power.
-    We can reduce the power consumption of the FPGA by reducing the power dissipation of current running through wires.
+    Minimizing the wirelength, or more accurately, minimizing the signal path length can help save power because wires have a non-zero resisitance.
     This is particularly important for FPGAs because signal paths must propagate not only through wires but also switchboxes and programmable interconnects which can also dissipate power.
+
     The second and more important reason is to minimize clock skew. 
     Signal propagation through wires and transistors takes time. 
-    The clock signal is crucial for state transition in an FSM.
-    If the clock signal drives two logic blocks that were placed too far apart, they may become desynchronized as the same clock edge may reach the two flip-flops at different clock periods.
+    The clock signal is crucial for state transition in a finite state machine (FSM).
+    If a clock signal drives two logic blocks that were placed too far apart, they may become desynchronized as the same clock edge may reach the two flip-flops at different clock periods.
     All flip-flops operating on the same clock domain must strike at the same nanosecond.
 
-    The third and equally relevant reason is the simplify the routing stage which takes place after placement.
-    If the logic blocks are all placed close together but in such a way that the wiring between them will become spaghetti, the router will produce an over-congested wiring and potentially risk routing failure because of it.
-    The placer ideally must cluster relevant logic blocks into semi-islolated neighborhood such that the wiring in one neighborhood to another is minimized.
+    The third, equally important reason is to simplify the routing stage that follows placement.
+If logic blocks are placed close together in a way that creates chaotic, spaghetti-like wiring between them, certain routing points may become over-congested and risk failure.
+To avoid this, the placer should group related logic blocks into semi-isolated neighborhoods, minimizing inter-neighborhood connections and ensuring more efficient and manageable routing.
 

### 2024-12-26 hdl/verilog/fir_filter/python/sine.py
256 taps 8 pipelines fir working

@@ -16,7 +16,7 @@ def generate(sine_freq_high, sine_freq_low, sample_freq, ampl, offset, bit_depth
     num_samples = int(sample_freq / sine_freq_low)  # number of samples per period
     t = np.arange(num_samples) / sample_freq
     scale_factor = (1 << (bit_depth - 1))  # scale factor for signed fixed-point range [-2^23, 2^23 - 1]
-    sine_wave = ampl * np.sin(2 * np.pi * sine_freq_low * t) + ampl * np.sin(2 * np.pi * sine_freq_high * t) + offset
+    sine_wave = ampl * np.sin(2 * np.pi * sine_freq_low * t) + 0.5 * ampl * np.sin(2 * np.pi * sine_freq_high * t) + offset
     sine_wave_int = np.round(sine_wave * scale_factor).astype(int)
 
     plot(sine_wave, t, "sine")
@@ -45,7 +45,7 @@ def write_verilog(filename, sine_signal, data_width):
 
 def main():
     # Parameters
-    sine_freq_high = 2000  # (Hz)
+    sine_freq_high = 5000  # (Hz)
     sine_freq_low = 200 # (Hz)
     sample_freq = 44000  # (Hz)
     amplitude = 0.2  # (range between -1 and 1)

### 2024-12-26 hdl/verilog/fir_filter/python/weights.py

@@ -64,12 +64,14 @@ def write_verilog(filename, weights, data_width):
 def main():
 
     # Parameters
-    NUM_PIPELINES = int(sys.argv[1])
-    FILTER_DEPTH = int(sys.argv[2])
+    FILTER_DEPTH = int(sys.argv[1])
+    NUM_PIPELINES = int(sys.argv[2])
     DATA_WIDTH = 24          # Number of bits for each coefficient
 
     PIPE_DEPTH = int(FILTER_DEPTH / NUM_PIPELINES)
-    print("weights.py: PIPE_DEPTH = " + str(PIPE_DEPTH))
+    print(f"weights.py: NUM_PIPELINES = {NUM_PIPELINES}")
+    print(f"weights.py: FILTER_DEPTH = {NUM_PIPELINES}")
+    print(f"weights.py: PIPE_DEPTH = {PIPE_DEPTH}")
     SAMPLE_RATE = 44000
     CUTOFF_FREQ = 1000
 

### 2024-12-26 hdl/verilog/fir_filter/sim_functional/simulate.sh

@@ -3,8 +3,8 @@
 DESIGN=fir_filter
 TOP_LEVEL=top_level
 
-NUM_PIPELINES=1
-FILTER_DEPTH=48
+FILTER_DEPTH=256
+NUM_PIPELINES=8
 
 check_exit_status() {
     if [ $? -ne 0 ]; then
@@ -21,7 +21,7 @@ echo "Running Functional Simulation..."
 # generate sine.mem and weights.mem
 cd "$DESIGN_DIR/python"
 python3 sine.py
-python3 weights.py "$NUM_PIPELINES" "$FILTER_DEPTH"
+python3 weights.py "$FILTER_DEPTH" "$NUM_PIPELINES"
 python3 generate_xpm_spram.py "$NUM_PIPELINES"
 
 cd "$DESIGN_DIR/sim_functional"

### 2024-12-26 run.sh

@@ -3,8 +3,8 @@
 DESIGN="fir_filter"
 TOP_LEVEL="top_level"
 
-NUM_PIPELINES=1
-FILTER_DEPTH=48
+FILTER_DEPTH=256
+NUM_PIPELINES=8
 
 export XILINX_VIVADO=/home/bcheng/workspace/tools/Xilinx/Vivado/2023.2
 export PATH="$PATH:$XILINX_VIVADO/bin"
@@ -57,7 +57,7 @@ if [ "$start_stage" == "sim_functional" ]; then
     # generate sine.mem and weights.mem
     cd "$DESIGN_DIR/python"
     python3 sine.py
-    python3 weights.py "$NUM_PIPELINES" "$FILTER_DEPTH"
+    python3 weights.py "$FILTER_DEPTH" "$NUM_PIPELINES"
     python3 generate_xpm_spram.py "$NUM_PIPELINES"
 
     cd "$DESIGN_DIR/sim_functional"

### 2024-12-26 java/src/main/java/placer/PlacerPackingSiteCentric.java
start placing stray cells

@@ -192,9 +192,10 @@ public class PlacerPackingSiteCentric extends Placer {
         int x_min = minmax.get("X_MIN");
         int y_max = minmax.get("Y_MAX");
         int y_min = minmax.get("Y_MIN");
-        writer.write("\n\tselectedSiteType: " + selectedSiteType);
-        writer.write(
-                "\n\tX_MAX: " + x_max + ", X_MIN: " + x_min + ", Y_MAX: " + y_max + ", Y_MIN: " + y_min);
+        // writer.write("\n\tselectedSiteType: " + selectedSiteType);
+        // writer.write(
+        // "\n\tX_MAX: " + x_max + ", X_MIN: " + x_min + ", Y_MAX: " + y_max + ", Y_MIN:
+        // " + y_min);
         Site anchorSite = null;
         int x = 0;
         int y = 0;
@@ -266,7 +267,7 @@ public class PlacerPackingSiteCentric extends Placer {
         writer.write("\n\nPlacing carry chains... (" + EDIFCarryChains.size() + ")");
         // PLACE CARRY CHAINS
         for (List<CarryCellGroup> chain : EDIFCarryChains) {
-            writer.write("\n\tchain size: " + chain.size());
+            writer.write("\n\t\tchain size: " + chain.size());
             Random rand = new Random();
             CarryCellGroup anchorGroup = chain.get(0);
             List<SiteTypeEnum> compatibleSiteTypes = new ArrayList<SiteTypeEnum>();
@@ -369,7 +370,7 @@ public class PlacerPackingSiteCentric extends Placer {
         List<Site> compatibleSites = new ArrayList<Site>(
                 Arrays.asList(device.getAllCompatibleSites(SiteTypeEnum.RAMB18E1)));
         compatibleSites.removeAll(occupiedRAMSites);
-        System.out.println(EDIFCellGroups.get("RAMB18E1").size());
+        // System.out.println(EDIFCellGroups.get("RAMB18E1").size());
 
         while (true) {
             Tile selectedTile = compatibleSites.get(rand.nextInt(compatibleSites.size())).getTile();
@@ -398,9 +399,10 @@ public class PlacerPackingSiteCentric extends Placer {
             occupiedRAMSites.add(ramSites.get(1));
             EDIFCellGroups.get("RAMB18E1").remove(cell1);
 
-            for (Site site : ramSites)
[…]

### 2024-12-28 latex/ece574-final/final_report.tex
start final report for ece574

\documentclass{article}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{mathpazo}

\begin{document}
\title{ECE574 Final Project Report: \\ Reconfigurable Partially Pipelined RAM-based FIR Filter}
\author{Brian B Cheng \\ Rutgers University Department of Electrical and Computer Engineering}

\date{}
\maketitle

\section{Abstract}
    This report presents a Verilog implementation of a reconfigurable Finite Impulse Response (FIR) filter designed for heterogeneous FPGAs.
    The model is highly flexible, with customizable data width, filter depth, and the number of parallel multiply-accumulate (MAC) pipelines, all configurable through top-level parameters.
    The filter modifies the traditional direct-form FIR structure by replacing the input shift register with RAM elements and the stationary weights with ROM elements. 
    Input samples are written to RAM as new samples arrive, with the write address counting downward.
    Concurrently, weights and samples are read from the ROMs and RAMs into their respective MAC units, with read addresses counting upward. 
    This counter-directional address counting achieves the "shift register" effect of a traditional direct-form filter.
    The method is demonstrated with a 256-tap FIR filter featuring a 24-bit data width and eight parallel MAC pipelines. 
    A testbench simulation provides a noisy sine wave as input, demonstrating the filter's performance in producing a clean sine wave output, verified in the waveform diagrams.

\section{Introduction}
    Finite Impulse Response (FIR) filters are commonly used in digital signal processing to remove unwanted frequencies from a noisy signal while preserving the desired frequency bands.
    They are particularly well-suited for implementation on FPGAs due to their unique balance between low latency, parallelism, and hardware flexibility.
    When implementing FIR filters on FPGAs, several key design constraints must be considered.
    Hardware constraints include the number utilization of multipliers (which are costly), adders, and registers used.
    Performance constraints include throughput, output latency, and placement area.
    Efficient design requires careful trade-offs between these constraints to meet the application's performance and resource requirements.
    This report presents an FIR filter that prioritizes placement area by opting for RAM and ROM elements instead of registers, and sacrifices low-latency for higher throughput via extensive usage of parallel pipelining.
    At the same time however, the usage of RAM and ROM instead of registers hinders the higher throughput as memory access throughput is slower than register throughput.
    Thus, the main benefit of this implementation is the placement profile on the FPGA chip.
    A smaller placement profiles can yield fewer wire nets and less congested routing, which may also contribute to lower power usage.

\section{Related Work}
    Related Work.

\section{Data Description}
    Not actually sure what this is upposed to be...
    The filter is highly flexible can be easily reconfigured by the user via the top level parameters DATA\_WIDTH, FIR\_DEPTH, and PIPELINES.
    The parameter DATA\_WIDTH defines the bit-width of the input signal, filter weights, and the output signal.
    The parameter FIR\_DEPTH defines the number of `taps' in the filter, or in other words, the number of multiply-accumulates (MAC) performed per input signal sample.
    The parameter PIPELINES defines the number of parallel hardware MAC units in the filter, or in other words, the number of `time-slices' per input signal sample.
    This paper demonstrates a 256 tap FIR Filter with 8 parallel MAC pipelines on 24-bit data, that is to say, with DATA\_WIDTH=24, FIR\_DEPTH=256, and PIPELINES=8.
    With 8 parallel MAC pipelines, each MAC unit performs 256/8=32 multiply-accumulates per input signal sample.
    The 256 weights are generated via python scipy's firwin() function for a low-pass filter with a cutoff frequency of 2KHz.
    An input signal of 200Hz sine wave plus a 10KHz sine wave to represent noise is fed into the filter.

\section{Method Description}
    Why RAM-based shift register instead of Flip-Flop-based shift register?

    24-bit signals => 24 registers per filter tap.

    For example, 256 taps * 24-bits = 6144 registers just to implement the input shift register.

    On a 7-series Xilinx FPGA, each CLB has only 8 registers.

    6144 / 8 registers => 768 CLBs.

    On the other hand, 3 RAM or ROM elements take up the same space profile of just 5 CLBs.

    8 MAC pipelines => 8 RAMs and 8 ROMs => 16 memory elements / 3 memory elements per BRAM sites => 6 BRAM sites => space profile of 30 CLBs.

\section{Model Description}
    Model Description.

\section{Experimental Procedure and Results}
    Experimental Procedure and Results

\section{Conclusion}
    Conclusion

\section{References}
    References

\end{document}

### 2024-12-28 hdl/verilog/fir_filter/python/weights.py

@@ -73,7 +73,7 @@ def main():
     print(f"weights.py: FILTER_DEPTH = {NUM_PIPELINES}")
     print(f"weights.py: PIPE_DEPTH = {PIPE_DEPTH}")
     SAMPLE_RATE = 44000
-    CUTOFF_FREQ = 1000
+    CUTOFF_FREQ = 2000
 
     # Generate FIR filter weights
     weights = generate(FILTER_DEPTH, DATA_WIDTH, CUTOFF_FREQ, SAMPLE_RATE)

### 2024-12-29 latex/ece574-final/final_report.tex
ece574-final report is good enough

@@
 
-\section{Related Work}
-    Related Work.
+\section{Data Description}
+    As mentioned in the Introduction, the purpose of a FIR Filter is to filter out unwanted frequency bands from an input signal and provide an output signal containing only the desired frequency bands.
+    This project presents a low pass filter with a cutoff frequency of 2KHz.
+    Thus, to verify the filter, we supply a synthetic noisy signal containing frequencies below and above 2KHz and observe the output signal to see if the filter correctly filters out frequencies above 2KHz.
+    To create this synthetic signal, we use Python to generate a Verilog header file containing the sine signal samples in hexadecimal format.
+    We use the \texttt{`include "filename.vh"} compiler directive in our testbench SystemVerilog code to import the Python-generated input samples.
+    All data signals in this implementation are 24-bit signed with fixed point representation Q1.23, meaning 1 sign bit, zero whole bits, and 23 fractional bits.
+    Thus, the range of all data signals, including the weights, is between negative one and positive one.
+    Shown in Equation~\ref{eq:sine} and Figure~\ref{fig:sine} is the formula and graph of the noisy sine wave, with the desired 200Hz sine wave added with an unwanted 5KHz sine wave noise.
+    The sampling frequency is 44\,\text{kHz}, meaning the sampling period is \( 22.273 \times 10^{-6} \) seconds.
 
-\section{Data Description}
-    Not actually sure what this is upposed to be...
-    The filter is highly flexible can be easily reconfigured by the user via the top level parameters DATA\_WIDTH, FIR\_DEPTH, and PIPELINES.
-    The parameter DATA\_WIDTH defines the bit-width of the input signal, filter weights, and the output signal.
-    The parameter FIR\_DEPTH defines the number of `taps' in the filter, or in other words, the number of multiply-accumulates (MAC) performed per input signal sample.
-    The parameter PIPELINES defines the number of parallel hardware MAC units in the filter, or in other words, the number of `time-slices' per input signal sample.
-    This paper demonstrates a 256 tap FIR Filter with 8 parallel MAC pipelines on 24-bit data, that is to say, with DATA\_WIDTH=24, FIR\_DEPTH=256, and PIPELINES=8.
-    With 8 parallel MAC pipelines, each MAC unit performs 256/8=32 multiply-accumulates per input signal sample.
-    The 256 weights are generated via python scipy's firwin() function for a low-pass filter with a cutoff frequency of 2KHz.
-    An input signal of 200Hz sine wave plus a 10KHz sine wave to represent noise is fed into the filter.
+    \begin{equation}
+        \text{input\_signal}(t) = 0.2 \cdot \sin(2 \pi f_{\text{low}} t)
+        + 0.1 \cdot \sin(2 \pi f_{\text{high}} t)
+        \label{eq:sine}
+    \end{equation}
+    \[
+    \begin{aligned}
+        f_{\text{low}} &= 200 \, \text{Hz}, & f_{\text{high}} &= 5 \, \text{KHz} \\
+    \end{aligned}
+    \]
+    \begin{figure}
+        \centering
+        \includegraphics[width=1.00\textwidth]{figures/sine.png}
+        \caption{Noisy input signal over 220 samples}\label{fig:sine}
+    \end{figure}
+
+    In similar fashion, we generate the 256 filter weights in Python using scipy's \texttt{firwin()} function.
+    We store the weights in hexadecimal format in \texttt{.mem} files, one \texttt{.mem} file for each MAC pipeline.
+    Each MAC operates on their unique subvector of the whole weight vector.
+    We use Xilinx's XPM macro to instantiate ROM elements and initialize the memories with their respective \texttt{.mem} files.
+    It is generally good practice to use vendor-specific macros instead of hand written memories to ensure correct memory behavior.
+    Shown in Figure~\ref{fig:weights} below is the function call and plot of filter weights.
+
+    \begin{verbatim}
+        filter_depth = 256
+        cutoff_freq = 2000
+        sample_rate = 44000
+        normalized_cutoff = cutoff_freq / (sample_rate / 2)
+        firwin(filter_depth, normalized_cutoff)
+    \end{verbatim}
+    \begin{figure}
+        \centering
+        \includegraphics[width=1.00\textwidth]{figures/fir_coeff.png}
+        \caption{256 filter weights for 44KHz sampling frequency and 2KHz cutoff frequency}\label{fig:weights}
+    \end{figure}
 
 \section{Method Description}
-    Why RAM-based shift register instead of Flip-Flop-based shift register?
+    \begin{figure}
+        \centering
+        \includegraphics[width=1.00\textwidth]{figures/hierarchy.png}
+        \caption{Xilinx FPGA Architecture Hierarchy}\label{fig:hierarchy}
+    \end{figure}
+    \subsection{Why Choose a RAM-Based Shift Register Over a Flip-Flop-Based Shift Register?}
+    A critical decision in this design is replacing a traditional flip-flop-based shift register with a RAM-based implementation.
+    This choice is motivated mainly by reducing CLB usage in favor of BRAM and reducing the area profile of the shift register.
+    This, I theorize, improves the placement profile on the FPGA chip, and reduces the total number of wire nets which can alleviate routing congestion in the programmable interconnects.
+    These two benefits overall contributes to a better Quality of Result (QoR) given that the performance specifications are met.
+    The usage of BRAMs over registers can also theoretically reduce power usage as vendor-specific memory macros are highly optimized in modern FPGAs.
 
-    24-bit signals => 24 registers per filter tap.
-
-    For example, 256 taps * 24-bits = 6144 registers just to implement the input shift register.
-
-    On a 7-series Xilinx FPGA, each CLB has only 8 registers.
-
-    6144 / 8 registers => 768 CLBs.
-
-    On the other hand, 3 RAM or ROM elements take up the same space profile of just 5 CLBs.
-
-    8 MAC pipelines => 8 RAMs and 8 ROMs => 16 memory elements / 3 memory elements per BRAM sites => 6 BRAM sites => space profile of 30 CLBs.
+    For a 24-bit signal, each filter tap requires 24 registers.
+    For instance, implementing a 256-tap FIR filter with 24-bit signals demands:
+    \[
+        256 \, \text{taps} \times 24 \, \text{bits/tap} = 6144 \, \text{flip-flops}.
+    \]
+    In a typical low-end Xilinx FPGA, such as the xc7z020, the available resources include \textbf{6650 configurable logic block (CLB) tiles}.
+    Each CLB contains two logic sites, which can either be both SLICELs or a combination of one SLICEL and one SLICEM.
+    The key difference between SLICEL and SLICEM is that SLICEM can reconfigure its lookup tables (LUTs) to function as small local RAMs, known as ``distributed RAM.''
+    The xc7z020 provides a total of \textbf{8950 SLICELs} and \textbf{4350 SLICEMs}, summing to \textbf{13,300 logic sites} across all CLBs.
+    Each logic site, regardless of type, contains \textbf{8 flip-flops}, meaning each CLB tile has \textbf{16 flip-flops}.
+    To implement the shift register using flip-flops:
+    \[
+        \frac{6144 \, \text{flip-flops}}{16 \, \text{flip-flops/CLB}} = 384 \, \text{CLBs}.
+    \]
+    On the other hand, block RAM (BRAM) tiles provide a more resource-efficient alternative.
+    Each BRAM tile contains \textbf{3 memory sites} (2 $\times$ 18Kb banks and 1 $\times$ 36Kb bank) that can be configured as RAM, ROM, or FIFO.
+    A single BRAM tile occupies approximately the same space as \textbf{10 CLBs}.
+    For an 8-pipeline FIR filter, the design requires \textbf{8 RAMs} (for input samples) and \textbf{8 ROMs} (for weights), totaling \textbf{16 memory elements}.
+    With \textbf{3 memory sites per BRAM tile}, this requires:
+    \[
+        \frac{16 \, \text{memory elements}}{3 \, \text{memory sites per BRAM}} = 6 \, \text{BRAM tiles}.
+    \]
+    In terms of CLB-equivalent space, these 6 BRAM tiles occupy:
+    \[
+        6 \, \text{BRAM tiles} \times 10 \, \text{CLBs/BRAM} = 60 \, \text{CLBs}.
+    \]
+    By opting for a RAM-based shift register, the design trades the \textbf{384 CLBs} required for a flip-flop-based implementation for just \textbf{60 CLBs}, resulting in a significant reduction in logic site utilization.
+    This trade-off demonstrates the value of leveraging the specialized memory resources available on FPGAs, such as BRAM, to implement resource-intensive designs efficiently.
+    With a smaller floorplanning profile (also known as a PBlock size), we can dedicate the freed up logic sites to other parts of the system design.
+    This also allows us to pack more FIR filters in parallel should the system design operate on multiple signal channels.
+    Refer to Figure~\ref{fig:CLB_BRAM} to see a space profile comparison between BRAMs and CLBs.
+    All of this of course comes with the tradeoffs mentioned earlier, and obviously, if the system is memory intensive elsewhere in the design, the RAM-based FIR filter might not be feasible.
+    \begin{figure}
+        \centering
+        \includegraphics[width=1.25\textwidth]{figures/CLB_BRAM.png}
+        \caption{CLB vs BRAM area profile comparison on the xc7z020. 1 BRAM tile highlighted on the right, 10 CLB tiles highlighted on the left.}
+        \label{fig:CLB_BRAM}
+    \end{figure}
 
 \section{Model Description}
-    Model Description.
+    All Verilog source codes for the implementation can be found in the \texttt{/src/} directory.
+    Refer to Figure~\ref{fig:rtl_top_level} for an RTL elaboration schematic for the top\_level module.
+    The FIR Filter module on its own communicates via multiple 24-bit data buses.
+    Since an FPGA package only has a limited amount of IO pins, we implement serializer-deserializer (SERDES) modules to communicate serially over a simple AXI-like handshake protocol.
+    Observe from the top\_level schematic from left to right: input ports which correspond to physical IO pins on the chip, the deserializer, the fir filter itself, the serializer, and finally, the output ports.
+    \begin{figure}
+        \centering
+        \includegraphics[width=1.4\textwidth]{figures/rtl_top_level.png}
+        \caption{top\_level RTL block diagram (auto-generated in Vivado: \texttt{synth -rtl})}
+        \label{fig:rtl_top_level}
+    \end{figure}
+
+    Shown in Figure~\ref{fig:RAM_ROM_MAC} is a block diagram showing the expandable ROM-RAM-MAC structure for the MAC pipelines.
+    In the figure are only the first 3 pipelines, but the pattern for how to expand it to the necessary number of pipelines via should be clear.
+    The partial sums on the right-side are summed concurrently in an adder chain.
+    Further improvements can be made here in the final summation such as using a binary adder tree.
+    \begin{figure}
+        \centering
+        \includegraphics[width=1.25\textwidth]{figures/RAM_ROM_MAC.png}
+        \caption{
+            Block diagram describing the expandable ROM-RAM-MAC pipeline structure (manually drawn via Draw.io).
+        }
+        \label{fig:RAM_ROM_MAC}
+    \end{figure}
+
+    The full RTL elaborated schematic of the exploded FIR filter module is too large to fit on one screen, but to give an impression,
+    shown in Figure~\ref{fig:rtl_fir_closeup} is a closeup of the filter with one pipeline highlited in dark-blue border.
+    \begin{figure}
+        \centering
+        \includegraphics[width=1.25\textwidth]{figures/rtl_fir_closeup.png}
+        \caption{
+            Closeup of RTL elaboration schematic highlighting one ROM-RAM-MAC pipeline in dark-blue border (auto-generated in Vivado: \texttt{synth -rtl}).
+        }
+        \label{fig:rtl_fir_closeup}
+    \end{figure}
+
+    Shown in Figure~\ref{fig:read_write_operations} is a timing diagram illustrating the write operations for an 8-tap 2-pipeline FIR Filter.
+    Every horizontal section divided by dotted lines represents a slice in time corresponding with each write operation.
+    Note that t=1,2,.. only corresponds to write operations.
+    In between each write operation are \texttt{PIPE\_DEPTH} consecutive read operations, with:
+
+    \[
+        \text{PIPE\_DEPTH} = \frac{\text{FIR\_DEPTH}}{\text{PIPELINES}}
+    \]
+
+    For simple illustration, the diagram describes an 8-tap filter with just 2 MAC pipelines.
+    Thus, PIPE\_DEPTH = 4 in this illustration.
+    Each MAC holds 4 input samples and 4 weights.
+    The arrows between the RAM and ROM show the corresponding pairs of samples and weights that must be read into the MAC.
+    The samples are written into the RAM starting at address = max\_value, in this case, max\_value = 2'b11.
+
+    Then, samples and weights are read consecutively into the MAC with the ROM read address starting at 2'b00 and the RAM read address starting at the write address.
+    With each sample, the addresses increment upwards. The RAM read address resets to 2'b00 when it overflows and continues counting upward.
+    Then, before the next input sample is received by the filter, the sample to be overwritten in each RAM is written into the RAM of the next pipeline.
+    The sample to be overwritten at this point is already on the RAM's data\_out bus and is immediately ready to be written into the next RAM.
+    The newly received input sample is then written into the first pipeline RAM.
+    Shown in Figure~\ref{fig:read_write_operations} at t=4 illustrates this write-into-next-ram-then-overwrite operation.
+    Also note that the RAMs have a separate read address and write address which are multiplexed into its address port.
+
+    This illustration should hopefully suffice in showing how to expand this operation from 8-taps with 2-pipelines to 256-taps with 8-pipelines.
+
+    \begin{figure}
+        \centering
+        \includegraphics[width=1.25\textwidth]{figures/read_write_operations.png}
+        \caption{
+            Timing diagram describing the write operation with each incoming input sample for an 8-tap 2-MAC-pipelines FIR Filter (manually drawn via Draw.io).
+        }
+        \label{fig:read_write_operations}
+    \end{figure}
 
 \section{Experimental Procedure and Results}
-    Experimental Procedure and Results
+    As described in the Data Description section,
+    we design a testbench that feeds a noisy signal containing a 200Hz sine wave + 5KHz noise sine wave into the design and 
+    expect a clean 200Hz sine wave from the output.
+    Shown in Figure~\ref{fig:waveform} is the testbench waveform simulation showing the noisy input signal passed into the FIR filter
+    and the clean signal output.
+    In an industry setting, a more mature verification approach would be to do a spectral analysis on the output signal, likely via Python or Matlab,
+    to see what frequency bands have passed the filter.
+    Verifying an analog signal via waveform diagrams is just a quick-and-dirty way to demonstrate the basic principles of the low-pass filter.
+
+    \begin{figure}
+        \centering
+        \includegraphics[width=1.25\textwidth]{figures/waveform.png}
+        \caption{
+            Testbench waveform simulation showing the noisy input signal iv\_din and clean output signal ov\_dout.
+        }
+        \label{fig:waveform}
+    \end{figure}
+
+    Shown in Figure~\ref{fig:utilization} is the resource utilization report summarizing the number of LUTs, FFs, DSPs, and BRAMs used.
+    Note that BMEM count shows 8, but after inspecting the netlist, I can confirm that 16 BRAM BELs are indeed used.
+    I am not entirely sure what BMEM is or how it is counted.
+    \begin{figure}
+        \centering
+        \includegraphics[width=0.75\textwidth]{figures/utilization.png}
+        \caption{
+            Synthesis resource utilization report.
+        }
+        \label{fig:utilization}
+    \end{figure}
+
+    Shown in Figure~\ref{fig:device} is the design placed and routed onto the xc7z020 using the default Vivado implementation strategy.
+    \begin{figure}
+        \centering
+        \includegraphics[width=1.25\textwidth]{figures/device.png}
+        \caption{
+            The final design placed and routed onto the xc7z020 via Vivado default implementation strategy.
+        }
+        \label{fig:device}
+    \end{figure}
+    Shown in Figure~\ref{fig:timing} is the timing report with the clock constraint set to 10MHz.
+    With a worst negative slack of +87ns which represents the critical path delay of the design,
+    we can likely increase our clock speed without penalty if higher performance is needed.
+    \begin{figure}
+        \centering
+        \includegraphics[width=1.25\textwidth]{figures/timing.png}
+        \caption{
+            Timing report shoing critical path delay (Worst Negative Slack)
+        }
+        \label{fig:timing}
+    \end{figure}
+
+    I have not used the Power report summary in Vivado before and not familiar with the report methodology, 
+    so I cannot offer concrete commentary on the power usage.
+    However, to give some impression, shown in Figure~\ref{fig:power_breakdown} and Figure~\ref{fig:power_environment} 
+    are the power breakdown and environment settings.
+    At a glance, it seems that BRAM usage takes up a lionshare of 44\% of the power usage.
+    Using 16 BRAM sites in our partially pipelined implementation replaced 6144 flip-flops of a 
+    traditional direct-form implementation as discussed in the Method Description section.
+    From the utilization report, 334 FLOP\_LATCH (Flip Flops) were used.
+    From the power report breakdown, I assume flip flop power usage is represented by Clocks and Signals which use 5\% and 16\% respectively, 
+    summing up to 21\%.
+    Had we used the traditional direct-form implementation and added 6144 additional flip-flops to the design, 
+    I suspect they may have used a larger percentage of the power.
+    \begin{figure}
+        \centering
+        \includegraphics[width=1.00\textwidth]{figures/power_breakdown.png}
+        \caption{
+        }
+        \label{fig:power_breakdown}
+    \end{figure}
+    \begin{figure}
+        \centering
+        \includegraphics[width=0.60\textwidth]{figures/power_environment.png}
+        \caption{
+        }
+        \label{fig:power_environment}
+    \end{figure}
 
 \section{Conclusion}
-    Conclusion
+    This paper presented a RAM-based partially pipelined Verilog model of an FIR Filter whose data width, filter depth, and number of pipelines
+    can be easily reconfigured via the top\_level parameters.
+    We have also demonstrated a 24-bit 256th order low-pass filter with 8 parallel MAC pipelines and performed a simple testbench simulation to verify its functionality.
+    There is certainly many improvements that could be done with this Verilog model in the future.
 
+    As mentioned before, the final summation of the partial sums from each MAC pipeline are summed in one clock cycle via adder chain.
+    One could improve upon this by using a binary adder tree and introducing some pipelining along the tree stages.
+    However, the FPGA CARRY4 BELs in modern 7-series FPGAs are very optimized, and the fact that our throughput is likely
+    limited by BRAM memory access suggests that this is probably not the bottleneck.
+    I suspect the limiting factor in terms of throughput is indeed the BRAM memory access.
+    If BRAM access becomes a problem, then the designer will have to increase the number of MAC pipelines to exploit the parallel time-slice processing.
+
+    Another consideration would be to explore the usage of Distributed RAMs mentioned previously to forego expensive BRAMs entirely.
+    The SLICEM logic sites have special LUTs that can be reconfigured as small local LUTRAMs.
+    Some more implementations can be set up to see if this could keep CLB usage low while avoiding BRAM.
+
+    As mentioned in the Experiments and Results section, a more mature verification environment can be employed to really stress-test the filter
+    and to run batches of tests with various random input signals to really see the band of frequencies that passes through the filter.
+    This could also help us in verifying the frequency response and to judge the signal to noise ratio of the signal.
+
+    A more in-depth exploration of the power usage and the QoR of the placement and routing is also due.
+
+\newpage
 \section{References}
-    References
+    \bibliographystyle{ieeetr}
+    \nocite{*}
+    \bibliography{
+        references/references
+    }
 

### 2024-12-29 hdl/verilog/fir_filter/python/sine.py

@@ -15,6 +15,7 @@ def plot(sine_wave, t, fname):
 def generate(sine_freq_high, sine_freq_low, sample_freq, ampl, offset, bit_depth):
     num_samples = int(sample_freq / sine_freq_low)  # number of samples per period
     t = np.arange(num_samples) / sample_freq
+    print(t.size)
     scale_factor = (1 << (bit_depth - 1))  # scale factor for signed fixed-point range [-2^23, 2^23 - 1]
     sine_wave = ampl * np.sin(2 * np.pi * sine_freq_low * t) + 0.5 * ampl * np.sin(2 * np.pi * sine_freq_high * t) + offset
     sine_wave_int = np.round(sine_wave * scale_factor).astype(int)

### 2025-03-04 java/src/main/java/placer/Placer.java
refactor class structure

@@ -3,18 +3,53 @@ package placer;
 import java.io.FileWriter;
 import java.io.IOException;
 
+import java.util.stream.Collectors;
+import java.util.Collection;
+import java.util.HashMap;
+import java.util.List;
+import java.util.ArrayList;
+import java.util.Arrays;
+import java.util.Set;
+import java.util.Map;
+import java.util.Random;
+
 import com.xilinx.rapidwright.design.Design;
+import com.xilinx.rapidwright.design.SiteInst;
+import com.xilinx.rapidwright.design.Net;
+
 import com.xilinx.rapidwright.device.Device;
+import com.xilinx.rapidwright.device.Tile;
+import com.xilinx.rapidwright.device.Site;
+import com.xilinx.rapidwright.device.SiteTypeEnum;
+import com.xilinx.rapidwright.device.ClockRegion;
 
 public abstract class Placer {
     protected String placerName;
     FileWriter writer;
+    protected String rootDir;
+    protected String graphicsDir;
+    protected String placedDcp;
 
     protected final Device device;
     protected final Design design;
 
-    protected String rootDir;
-    protected String placedDcp;
+    protected List<Double> costHistory;
+    protected List<Long> moveTimes;
+    protected List<Long> evalTimes;
+    protected List<Long> renderTimes;
+    protected List<Long> writeTimes;
+    protected int movesLimit;
+    protected List<Double> coolingSchedule;
+    protected double currentTemp;
+
+    protected ClockRegion regionConstraint;
+    protected Set<SiteTypeEnum> uniqueSiteTypes;
+    protected Map<SiteTypeEnum, List<Site>> allSites;
+    protected Map<SiteTypeEnum, Map<Site, SiteInst>> occupiedSites;
+
+    // store chain occupation info for fast site-to-chain access
+    protected Map<SiteTypeEnum, Map<Site, List<SiteInst>>> occupiedSiteChains;
+    protected Random rand;
 
     protected String[] FF5_BELS = new String[] { "A5FF", "B5FF", "C5FF", "D5FF" };
     protected String[] FF_BELS = new String[] { "AFF", "BFF", "CFF", "DFF" };
@@ -45,5 +80,255 @@ public abstract class Placer {
     }
 
     protected abstract void placeDesign(PackedDesign packedDesign) throws IOException;
+
+    protected abstract Site proposeSite(SiteTypeEnum ste, boolean swapEnable);
+
[…]

### 2025-03-04 java/src/main/java/placer/PlacerAnnealMidpoint.java

@@ -1,53 +1,22 @@
 
 package placer;
 
-import java.io.FileWriter;
-import java.io.File;
 import java.io.IOException;
-import java.util.stream.Collectors;
-import java.util.Collection;
 import java.util.Collections;
-import java.util.Comparator;
 import java.util.HashMap;
 import java.util.List;
-import java.util.LinkedList;
 import java.util.ArrayList;
-import java.util.Arrays;
-import java.util.Set;
 import java.util.HashSet;
-import java.util.Map;
-import java.util.Random;
 
-import com.xilinx.rapidwright.edif.EDIFHierCellInst;
-
-import com.xilinx.rapidwright.design.ModuleInst;
 import com.xilinx.rapidwright.design.Design;
 import com.xilinx.rapidwright.design.SiteInst;
-import com.xilinx.rapidwright.design.SitePinInst;
-import com.xilinx.rapidwright.design.Net;
 
 import com.xilinx.rapidwright.device.Device;
-import com.xilinx.rapidwright.device.Tile;
-
 import com.xilinx.rapidwright.device.Site;
 import com.xilinx.rapidwright.device.SiteTypeEnum;
 
 public class PlacerAnnealMidpoint extends Placer {
 
-    private String graphicsDir;
-    private Set<SiteTypeEnum> uniqueSiteTypes;
-    private Map<SiteTypeEnum, List<Site>> allSites;
-    private Map<SiteTypeEnum, Map<Site, SiteInst>> occupiedSites;
-
-    // store chain occupation info for fast access
-    private Map<SiteTypeEnum, Map<Site, List<SiteInst>>> occupiedSiteChains;
-    private Random rand;
-
-    private List<Float> costHistory;
-    private List<Long> moveTimes;
-    private List<Long> evalTimes;
-    private List<Long> renderTimes;
-    private List<Long> writeTimes;
     private List<Pair<Integer, Integer>> spiralPath = SpiralPath.generateDiamondSpiral(1000);
 
     public PlacerAnnealMidpoint(String rootDir, Design design, Device device) throws IOException {
@@ -63,31 +32,31 @@ public class PlacerAnnealMidpoint extends Placer {
         this.evalTimes = new ArrayList<>();
         this.renderTimes = new ArrayList<>();
         this.writeTimes = new ArrayList<>();
-        this.rand = new Random();
     }
 
     public void placeDesign(PackedDesign packedDesign) throws IOException {
         initSites();
[…]

### 2025-03-04 java/src/main/java/placer/PlacerAnnealRandom.java

@@ -1,60 +1,23 @@
 
 package placer;
 
-import java.io.FileWriter;
-import java.io.File;
 import java.io.IOException;
-import java.util.stream.Collectors;
-import java.util.Collection;
 import java.util.Collections;
-import java.util.Comparator;
 import java.util.HashMap;
 import java.util.List;
-import java.util.LinkedList;
 import java.util.ArrayList;
-import java.util.Arrays;
-import java.util.Set;
 import java.util.HashSet;
-import java.util.Map;
 import java.util.Random;
 
-import com.xilinx.rapidwright.edif.EDIFHierCellInst;
-
-import com.xilinx.rapidwright.design.ModuleInst;
 import com.xilinx.rapidwright.design.Design;
 import com.xilinx.rapidwright.design.SiteInst;
-import com.xilinx.rapidwright.design.SitePinInst;
-import com.xilinx.rapidwright.design.Net;
-import com.xilinx.rapidwright.design.Cell;
 
 import com.xilinx.rapidwright.device.Device;
-import com.xilinx.rapidwright.device.Tile;
-
 import com.xilinx.rapidwright.device.Site;
 import com.xilinx.rapidwright.device.SiteTypeEnum;
 import com.xilinx.rapidwright.device.ClockRegion;
 
 public class PlacerAnnealRandom extends Placer {
-
-    private String graphicsDir;
-    private Set<SiteTypeEnum> uniqueSiteTypes;
-    private Map<SiteTypeEnum, List<Site>> allSites;
-    private Map<SiteTypeEnum, Map<Site, SiteInst>> occupiedSites;
-
-    protected ClockRegion regionConstraint;
-
-    // store chain occupation info for fast access
-    private Map<SiteTypeEnum, Map<Site, List<SiteInst>>> occupiedSiteChains;
-    private Random rand;
-
-    private List<Double> costHistory;
-    private List<Long> moveTimes;
-    private List<Long> evalTimes;
-    private List<Long> renderTimes;
-    private List<Long> writeTimes;
-    private int movesLimit;
-    private List<Double> coolingSchedule;
-    private double currentTemp;
 
     public PlacerAnnealRandom(String rootDir, Design design, Device device, ClockRegion region) throws IOException {
         super(rootDir, design, device);
@@ -75,24 +38,20 @@ public class PlacerAnnealRandom extends Placer {
 
     public void placeDesign(PackedDesign packedDesign) throws IOException {
         initSites();
[…]

### 2025-03-04 java/src/main/java/placer/PlacerGreedyRandom.java

@@ -1,648 +1,0 @@
-
-package placer;
-
-import java.io.FileWriter;
-import java.io.File;
-import java.io.IOException;
-import java.util.stream.Collectors;
-import java.util.Collection;
-import java.util.Collections;
-import java.util.Comparator;
-import java.util.HashMap;
-import java.util.List;
-import java.util.LinkedList;
-import java.util.ArrayList;
-import java.util.Arrays;
-import java.util.Set;
-import java.util.HashSet;
-import java.util.Map;
-import java.util.Random;
-
-import com.xilinx.rapidwright.edif.EDIFHierCellInst;
-
-import com.xilinx.rapidwright.design.ModuleInst;
-import com.xilinx.rapidwright.design.Design;
-import com.xilinx.rapidwright.design.SiteInst;
-import com.xilinx.rapidwright.design.SitePinInst;
-import com.xilinx.rapidwright.design.Net;
-
-import com.xilinx.rapidwright.device.Device;
-import com.xilinx.rapidwright.device.Tile;
-
-import com.xilinx.rapidwright.device.Site;
-import com.xilinx.rapidwright.device.SiteTypeEnum;
-
-public class PlacerGreedyRandom extends Placer {
-
-    private String graphicsDir;
-    private Set<SiteTypeEnum> uniqueSiteTypes;
-    private Map<SiteTypeEnum, List<Site>> allSites;
-    private Map<SiteTypeEnum, Map<Site, SiteInst>> occupiedSites;
-
-    // store chain occupation info for fast access
-    private Map<SiteTypeEnum, Map<Site, List<SiteInst>>> occupiedSiteChains;
-    private Random rand;
-
-    private List<Float> costHistory;
-    private List<Long> moveTimes;
-    private List<Long> evalTimes;
-    private List<Long> renderTimes;
-    private List<Long> writeTimes;
-    private List<Pair<Integer, Integer>> spiralPath = SpiralPath.generateDiamondSpiral(1000);
-
-    public PlacerGreedyRandom(String rootDir, Design design, Device device) throws IOException {
-        super(rootDir, design, device);
-        this.placerName = "PlacerGreedyRandom";
-        this.graphicsDir = rootDir + "/outputs/graphics/" + placerName;
-        this.uniqueSiteTypes = new HashSet<>();
-        this.occupiedSites = new HashMap<>();
-        this.occupiedSiteChains = new HashMap<>();
-        this.allSites = new HashMap<>();
[…]

### 2025-03-04 java/src/main/java/placer/Main.java
corrected the mistake of only evaluating output sitepininsts during cost eval. placement QOR is so much better now, though slower

@@ -16,6 +16,7 @@ import com.xilinx.rapidwright.design.Design;
 import com.xilinx.rapidwright.design.SiteInst;
 
 import com.xilinx.rapidwright.device.Device;
+import com.xilinx.rapidwright.device.ClockRegion;
 
 public class Main {
 
@@ -43,8 +44,10 @@ public class Main {
             // Stage 2) Packer:
             // takes the prepackedDesign and packs the EDIFHierCellInst into SiteInsts
             // also provides an initial random placement of SiteInsts onto actual Sites.
+            // ClockRegion regionConstraint = device.getClockRegion("X1Y0");
+            ClockRegion regionConstraint = null;
             PackerBasic1 B1Packer = new PackerBasic1(
-                    rootDir, design, device, device.getClockRegion("X1Y0"));
+                    rootDir, design, device, regionConstraint);
             B1Packer.printUniqueSites();
             B1Packer.printClockBuffers();
             PackedDesign packedDesign = B1Packer.run(prepackedDesign);
@@ -54,7 +57,7 @@ public class Main {
             // Sites via simulated annealing, analytical, electrostatic placement, etc.
             // works entirely on the SiteInst/Site/Tile level.
             PlacerAnnealRandom SARPlacer = new PlacerAnnealRandom(
-                    rootDir, design, device, device.getClockRegion("X1Y0"));
+                    rootDir, design, device, regionConstraint);
             SARPlacer.run(packedDesign, prepackedDesign);
 
         } catch (IOException e) {

### 2025-03-04 java/src/main/java/placer/Placer.java

@@ -124,16 +124,26 @@ public abstract class Placer {
         si.unPlace();
     }
 
-    protected List<Site> findSinkSites(SiteInst si) {
+    protected List<Site> findConnectedSites(SiteInst si) {
         List<Site> sinkSites = si.getSitePinInsts().stream()
-                .filter(spi -> spi.isOutPin())
+                // .map(spi -> {
+                // if (spi.getNet() == null) {
+                // System.out.println("Net is null for SitePinInst: " + spi.getName());
+                // }
+                // return spi.getNet();
+                // })
+                .filter(spi -> {
+                    if (spi.getNet() == null) {
+                        return false;
+                    } else
+                        return true;
+                })
                 .map(spi -> spi.getNet())
-                .map(net -> net.getSinkPins())
+                .map(net -> net.getPins())
                 .map(spis -> spis.stream()
                         .map(spi -> spi.getSite())
                         .collect(Collectors.toList()))
                 .flatMap(List::stream) // List<List<Site>> into List<Site>
-                // .filter(site -> site != null)
                 .collect(Collectors.toList());
         return sinkSites;
     }
@@ -146,6 +156,9 @@ public abstract class Placer {
         // otherwise, accept the move with a probability given by the Boltzmann factor
         double delta = newCost - oldCost;
         double acceptanceProbability = Math.exp(-delta / this.currentTemp);
+        // if (acceptanceProbability > 0.1) {
+        // System.out.println("Acceptance Probability: " + acceptanceProbability);
+        // }
         // decide acceptance based on a uniform random draw in [0, 1)
         return Math.random() < acceptanceProbability;
     }
@@ -153,10 +166,14 @@ public abstract class Placer {
     protected double evaluateSite(List<Site> sinkSites, Site srcSite) throws IOException {
         double cost = 0;
         for (Site sinkSite : sinkSites) {
-            // if (sinkSite == null)
-            // continue; // sink has not been placed yet
-            if (sinkSite.isGlobalClkBuffer() || sinkSite.isGlobalClkPad())
+            if (sinkSite.isGlobalClkBuffer()) {
+                // System.out.println("Skipped global clock buffer evaluation.");
                 continue;
+            }
+            if (sinkSite.isGlobalClkPad()) {
+                // System.out.println("Skipped global clock pad evaluation.");
+                continue;
+            }
[…]

### 2025-03-04 java/src/main/java/placer/PlacerAnnealMidpoint.java

@@ -197,21 +197,21 @@ public class PlacerAnnealMidpoint extends Placer {
     protected void randomMoveSingleSite(List<SiteInst> sites) throws IOException {
         for (SiteInst si : sites) {
             SiteTypeEnum ste = si.getSiteTypeEnum();
-            List<Site> homeSinks = findSinkSites(si);
+            List<Site> homeConns = findConnectedSites(si);
             Site homeSite = si.getSite();
             Site awaySite = proposeSite(ste, true);
             double oldCost = 0;
             double newCost = 0;
             SiteInst awaySi = occupiedSites.get(ste).get(awaySite);
             if (awaySi != null) {
-                List<Site> awaySinks = findSinkSites(awaySi);
-                oldCost += evaluateSite(homeSinks, homeSite);
-                oldCost += evaluateSite(awaySinks, awaySite);
-                newCost += evaluateSite(homeSinks, awaySite);
-                newCost += evaluateSite(awaySinks, homeSite);
+                List<Site> awayConns = findConnectedSites(awaySi);
+                oldCost += evaluateSite(homeConns, homeSite);
+                oldCost += evaluateSite(awayConns, awaySite);
+                newCost += evaluateSite(homeConns, awaySite);
+                newCost += evaluateSite(awayConns, homeSite);
             } else {
-                oldCost += evaluateSite(homeSinks, homeSite);
-                newCost += evaluateSite(homeSinks, awaySite);
+                oldCost += evaluateSite(homeConns, homeSite);
+                newCost += evaluateSite(homeConns, awaySite);
             }
             if (evaluateMoveAcceptance(oldCost, newCost)) {
                 if (awaySi != null) {
@@ -297,26 +297,26 @@ public class PlacerAnnealMidpoint extends Placer {
             double newCost = 0;
             for (int i = 0; i < homeBuffer.size(); i++) {
                 if (siteInstsInHomeBuffer.get(i) != null) {
-                    List<Site> homeSinks = findSinkSites(siteInstsInHomeBuffer.get(i));
-                    for (Site sink : homeSinks) {
-                        if (awayBuffer.contains(sink)) {
+                    List<Site> homeConns = findConnectedSites(siteInstsInHomeBuffer.get(i));
+                    for (Site conn : homeConns) {
+                        if (awayBuffer.contains(conn)) {
                             continue loopThruChains; // skip this chain swap proposal
[…]

### 2025-03-04 java/src/main/java/placer/PlacerAnnealRandom.java

@@ -42,8 +42,8 @@ public class PlacerAnnealRandom extends Placer {
         randomInitialPlacement(packedDesign);
         design.writeCheckpoint(rootDir + "/outputs/checkpoints/init_placed.dcp");
 
-        this.movesLimit = 500;
-        initCoolingSchedule(1.0f, 0.1f);
+        this.movesLimit = 250;
+        initCoolingSchedule(1000.0f, 0.99f);
         int move = 0;
         while (true) {
             if (move >= movesLimit)
@@ -73,7 +73,6 @@ public class PlacerAnnealRandom extends Placer {
 
             move++;
         }
-
         ImageMaker imPlaced = new ImageMaker(design);
         imPlaced.renderAll();
         imPlaced.exportImage(graphicsDir + "/final_placement.png", "png");
@@ -199,21 +198,21 @@ public class PlacerAnnealRandom extends Placer {
     protected void randomMoveSingleSite(List<SiteInst> sites) throws IOException {
         for (SiteInst si : sites) {
             SiteTypeEnum ste = si.getSiteTypeEnum();
-            List<Site> homeSinks = findSinkSites(si);
+            List<Site> homeConns = findConnectedSites(si);
             Site homeSite = si.getSite();
             Site awaySite = proposeSite(ste, true);
             double oldCost = 0;
             double newCost = 0;
             SiteInst awaySi = occupiedSites.get(ste).get(awaySite);
             if (awaySi != null) {
-                List<Site> awaySinks = findSinkSites(awaySi);
-                oldCost += evaluateSite(homeSinks, homeSite);
-                oldCost += evaluateSite(awaySinks, awaySite);
-                newCost += evaluateSite(homeSinks, awaySite);
-                newCost += evaluateSite(awaySinks, homeSite);
+                List<Site> awayConns = findConnectedSites(awaySi);
+                oldCost += evaluateSite(homeConns, homeSite);
+                oldCost += evaluateSite(awayConns, awaySite);
+                newCost += evaluateSite(homeConns, awaySite);
+                newCost += evaluateSite(awayConns, homeSite);
             } else {
-                oldCost += evaluateSite(homeSinks, homeSite);
-                newCost += evaluateSite(homeSinks, awaySite);
+                oldCost += evaluateSite(homeConns, homeSite);
+                newCost += evaluateSite(homeConns, awaySite);
             }
             if (evaluateMoveAcceptance(oldCost, newCost)) {
[…]

### 2025-03-04 java/src/main/java/placer/Placer.java
separate handling for input/output spis. much faster runtime, but generally lower QOR

@@ -16,6 +16,7 @@ import java.util.Random;
 import com.xilinx.rapidwright.design.Design;
 import com.xilinx.rapidwright.design.SiteInst;
 import com.xilinx.rapidwright.design.Net;
+import com.xilinx.rapidwright.design.SitePinInst;
 
 import com.xilinx.rapidwright.device.Device;
 import com.xilinx.rapidwright.device.Tile;
@@ -125,27 +126,30 @@ public abstract class Placer {
     }
 
     protected List<Site> findConnectedSites(SiteInst si) {
-        List<Site> sinkSites = si.getSitePinInsts().stream()
-                // .map(spi -> {
-                // if (spi.getNet() == null) {
-                // System.out.println("Net is null for SitePinInst: " + spi.getName());
-                // }
-                // return spi.getNet();
-                // })
-                .filter(spi -> {
-                    if (spi.getNet() == null) {
-                        return false;
-                    } else
-                        return true;
-                })
+        Collection<SitePinInst> pins = si.getSitePinInsts();
+        // Handle SPI output pins. Get all sinks.
+        List<Site> outputSinks = pins.stream()
+                .filter(spi -> spi.isOutPin())
                 .map(spi -> spi.getNet())
-                .map(net -> net.getPins())
+                .map(net -> net.getSinkPins())
                 .map(spis -> spis.stream()
                         .map(spi -> spi.getSite())
                         .collect(Collectors.toList()))
                 .flatMap(List::stream) // List<List<Site>> into List<Site>
                 .collect(Collectors.toList());
-        return sinkSites;
+        // Handle SPI input pins. Only get the source.
+        List<Site> inputSources = pins.stream()
+                .filter(spi -> !spi.isOutPin())
+                .map(spi -> spi.getNet())
+                .filter(net -> net != null)
+                .map(net -> net.getSource())
+                .filter(spi -> spi != null)
+                .map(spi -> spi.getSite())
+                .collect(Collectors.toList());
+        List<Site> allSites = new ArrayList<>();
+        allSites.addAll(inputSources);
+        allSites.addAll(outputSinks);
+        return allSites;
     }
 
     protected boolean evaluateMoveAcceptance(double oldCost, double newCost) {
@@ -153,27 +157,22 @@ public abstract class Placer {
         if (newCost < oldCost) {
             return true;
         }
-        // otherwise, accept the move with a probability given by the Boltzmann factor
         double delta = newCost - oldCost;
         double acceptanceProbability = Math.exp(-delta / this.currentTemp);
[…]

### 2025-03-04 java/src/main/java/placer/PlacerAnnealRandom.java

@@ -42,8 +42,8 @@ public class PlacerAnnealRandom extends Placer {
         randomInitialPlacement(packedDesign);
         design.writeCheckpoint(rootDir + "/outputs/checkpoints/init_placed.dcp");
 
-        this.movesLimit = 250;
-        initCoolingSchedule(1000.0f, 0.99f);
+        this.movesLimit = 500;
+        initCoolingSchedule(10000.0f, 0.95f);
         int move = 0;
         while (true) {
             if (move >= movesLimit)

### 2025-03-05 java/src/main/java/placer/Placer.java
better QoR, good runtime. DSP cascades can move more freely by ignoring chain self connections

@@ -125,7 +125,7 @@ public abstract class Placer {
         si.unPlace();
     }
 
-    protected List<Site> findConnectedSites(SiteInst si) {
+    protected List<Site> findConnectedSites(SiteInst si, List<Site> selfConns) {
         Collection<SitePinInst> pins = si.getSitePinInsts();
         // Handle SPI output pins. Get all sinks.
         List<Site> outputSinks = pins.stream()
@@ -157,6 +157,7 @@ public abstract class Placer {
         if (newCost < oldCost) {
             return true;
         }
+        // otherwise, evaluate probability to accept higher cost
         double delta = newCost - oldCost;
         double acceptanceProbability = Math.exp(-delta / this.currentTemp);
         return Math.random() < acceptanceProbability;

### 2025-03-05 java/src/main/java/placer/PlacerAnnealMidpoint.java

@@ -197,14 +197,14 @@ public class PlacerAnnealMidpoint extends Placer {
     protected void randomMoveSingleSite(List<SiteInst> sites) throws IOException {
         for (SiteInst si : sites) {
             SiteTypeEnum ste = si.getSiteTypeEnum();
-            List<Site> homeConns = findConnectedSites(si);
+            List<Site> homeConns = findConnectedSites(si, null);
             Site homeSite = si.getSite();
             Site awaySite = proposeSite(ste, true);
             double oldCost = 0;
             double newCost = 0;
             SiteInst awaySi = occupiedSites.get(ste).get(awaySite);
             if (awaySi != null) {
-                List<Site> awayConns = findConnectedSites(awaySi);
+                List<Site> awayConns = findConnectedSites(awaySi, null);
                 oldCost += evaluateSite(homeConns, homeSite);
                 oldCost += evaluateSite(awayConns, awaySite);
                 newCost += evaluateSite(homeConns, awaySite);
@@ -297,7 +297,7 @@ public class PlacerAnnealMidpoint extends Placer {
             double newCost = 0;
             for (int i = 0; i < homeBuffer.size(); i++) {
                 if (siteInstsInHomeBuffer.get(i) != null) {
-                    List<Site> homeConns = findConnectedSites(siteInstsInHomeBuffer.get(i));
+                    List<Site> homeConns = findConnectedSites(siteInstsInHomeBuffer.get(i), homeBuffer);
                     for (Site conn : homeConns) {
                         if (awayBuffer.contains(conn)) {
                             continue loopThruChains; // skip this chain swap proposal
@@ -308,7 +308,7 @@ public class PlacerAnnealMidpoint extends Placer {
                     newCost += evaluateSite(homeConns, awayBuffer.get(i));
                 }
                 if (siteInstsInAwayBuffer.get(i) != null) {
-                    List<Site> awayConns = findConnectedSites(siteInstsInAwayBuffer.get(i));
+                    List<Site> awayConns = findConnectedSites(siteInstsInHomeBuffer.get(i), awayBuffer);
                     for (Site conn : awayConns) {
                         if (homeBuffer.contains(conn)) {
                             continue loopThruChains; // skip this chain swap proposal
@@ -370,7 +370,7 @@ public class PlacerAnnealMidpoint extends Placer {
         SiteTypeEnum ste = chain.get(0).getSiteTypeEnum();
         List<Site> conns = new ArrayList<>();
[…]

### 2025-03-05 java/src/main/java/placer/PlacerAnnealRandom.java

@@ -3,6 +3,8 @@ package placer;
 
 import java.io.IOException;
 import java.util.Collections;
+import java.util.stream.Collectors;
+import java.util.Collection;
 import java.util.HashMap;
 import java.util.List;
 import java.util.ArrayList;
@@ -11,6 +13,7 @@ import java.util.Random;
 
 import com.xilinx.rapidwright.design.Design;
 import com.xilinx.rapidwright.design.SiteInst;
+import com.xilinx.rapidwright.design.SitePinInst;
 
 import com.xilinx.rapidwright.device.Device;
 import com.xilinx.rapidwright.device.Site;
@@ -43,7 +46,7 @@ public class PlacerAnnealRandom extends Placer {
         design.writeCheckpoint(rootDir + "/outputs/checkpoints/init_placed.dcp");
 
         this.movesLimit = 500;
-        initCoolingSchedule(10000.0f, 0.95f);
+        initCoolingSchedule(20000.0f, 0.98f);
         int move = 0;
         while (true) {
             if (move >= movesLimit)
@@ -79,6 +82,25 @@ public class PlacerAnnealRandom extends Placer {
         exportCostHistory(rootDir + "/outputs/printout/convergence.csv");
         printTimingBenchmarks();
         writer.write("\n\nTotal move iterations: " + move);
+    }
+
+    @Override
+    protected List<Site> findConnectedSites(SiteInst si, List<Site> selfConns) {
+        List<Site> connectedSites = si.getSitePinInsts().stream()
+                .map(spi -> spi.getNet())
+                .filter(net -> net != null)
+                .filter(net -> !net.isClockNet() && !net.isStaticNet())
+                .map(net -> net.getPins())
+                .map(spis -> spis.stream()
+                        .map(spi -> spi.getSite())
+                        // for chain swaps, ignore connections within the buffers.
+                        // DSP cascades can have very many self connections.
+                        // Allows DSP cascades to move more freely
+                        .filter(site -> !((selfConns != null) && selfConns.contains(site)))
+                        .collect(Collectors.toList()))
+                .flatMap(List::stream) // List<List<Site>> into List<Site>
+                .collect(Collectors.toList());
+        return connectedSites;
     }
 
     protected void initCoolingSchedule(double initialTemp, double alpha) throws IOException {
@@ -198,14 +220,14 @@ public class PlacerAnnealRandom extends Placer {
     protected void randomMoveSingleSite(List<SiteInst> sites) throws IOException {
[…]

### 2025-03-05 java/src/main/java/placer/Placer.java
passes default routing

@@ -192,7 +192,7 @@ public abstract class Placer {
             // continue;
             // }
             Tile srcTile = net.getSourceTile();
-            if (srcTile == null) // net is null if its' purely intrasite!
+            if (srcTile == null) // tile is null if its' purely intrasite!
                 continue;
             List<Tile> sinkTiles = net.getSinkPins().stream()
                     .map(spi -> spi.getTile())

### 2025-03-05 latex/paper/paper.tex
begin work on midpoint

--- 2024-09-13 docs/proposal/proposal.tex
@@
     A compiler takes a program written in high-level programming language like a Java or Python and assembles low-level machine code which is ready to be executed on a CPU.
-    In similar fashion, an Electronic Design Automation tool (EDA) takes a high-level description of a digital system written in Verilog or VHDL and produces a bitstream which is ready to be deployed to an FPGA.
-    In a superficial way, compilers and EDAs perform the same task for engineers but in different industries. 
-    One for software, the other for hardware.
-
-    Software compilers have evolved to become highly optimized, significantly boosting the productivity of software developers. 
-    The compilers are so robust that developers rarely need to worry about the correctness or efficiency of their machine code, and are so refined that it all happens in a matter of milliseconds to minutes, even for large projects.
-% break this up?
-    The speed of compilation enables rapid debugging and implementation of new features and allows developers to iterate through dozens or even hundreds of design cycles per day.
-    Instead of tinkering around with assembly code, software engineers can comfortably focus on the high-level abstractions of their design.
-
-    The compilation stages mainly consists of lexing, parsing, and optimization.
-    The optimization stage is where most of the complexity occurs as it deals with NP-hard problems.
-    The compiler at this stage must optimize the ordering of the machine instructions to maximize pipelining and register allocation.
-    While the search space for the best possible machine code is massive, it is typically constrained by the size of the CPU instruction set and architecture which are fixed.
-    Furthermore, compiler makers have developed highly efficient heuristics to approximate the best machine code within an acceptable runtime.
-
-    On the other hand, the EDA stages consist mainly of synthesis, placement, and routing.
-    % Placement and routing is often packaged into one big stage referred to as "implementation" as is the case in Vivado.
-    All three are NP-hard optimization problems with massive search spaces.
-    Even with heuristics and approximations to circumvent NP-hardness, the size of their search spaces make the EDA runtimes less than desirable.
-
-    In a typical FPGA design flow, the engineer creates and describes a digital design using a hardware description language (HDL) like Verilog or VHDL.     
-    Then, the FPGA engineer verifies the design by creating a testbench that wraps around the design and feeds the design input signals and observes the output signals.
-    % The formal verification process is an art of its own and is so matured that companies will often separate their engineers into two departments - one focusing on design and the other on verification.
-    Then the engineer submits the design entry to an EDA like Vivado or Quartus to perform the three stages of automated design. 
-    For even modest designs that utilize a small percentage of the available resources on the FPGA, the EDA will spend a minimum two to three minutes running synthesis and implementation.
-    The EDA run time increases exponentially with the scale of the project. 
-    Projects that utilize 80\% or more of the FPGA's resources may run for hours, and for the high-end devices like Xilinx's Kintex FPGAs which have millions of logical elements, can run for days.
-    At high utilization, there is a possibility that the EDA cannot fit the design onto the device due to routing congestion even when it has enough logical elements to implement the design. 
-    The EDA might only reach this conclusion after attempting to fit the design for several hours.
-
-    One of the most common complaints from new FPGA engineers, especially those coming from software development, is that the FPGA toolchains are slow and heavy.
-    This is because the problems that the EDA must solve are inherently complex.
-    Unlike in software development where an error in the program can be patched and recompiled in a matter of minutes, EDA runtimes lasting several hours means that an engineer can only go through a couple design cycles per day.
-    An FPGA engineer must be very precise with their coding and practice thorough verification of their design before submitting it to the EDA.
-    This overall heightens the barrier of entry into FPGA development and contributes to a shortage of qualified FPGA engineers and a limitation of their productivity.
-
-    This paper aims to study one particular pillar of this barrier: the placement stage. 
-    First, it will review prerequisite knowledge for placement strategies. 
-    This includes graph theory to represent logical elements of a digital design and the wired connections between them, and a corresponding optimization problem with a cost function which we seek to minimize.
-
-    Second, it will briefly explore the historical progress made on placement algorithms for both VLSI and FPGA and then explore the current trends in open-source placement research 
-    \cite{Yosys}, \cite{AMFPlacer2}, \cite{RapidLayout}, \cite{RapidStream}, \cite{DREAMPlace}, \cite{DREAMPlaceFPGA}.
-    I would have liked to explore the current trends in proprietary EDAs but they are unfortunately closed source.
-
-    Then third, the paper will outline my own work on placement strategies, whether they be reproductions of existing strategies, or a novel strategy if I manage to create one in the span of 3 months (unlikely).
-    I intend to use Xilinx's open-source API, RapidWright, which is written in Java and provides backend access to Xilinx's FPGA EDA, Vivado. 
-    I have already familiarized myself with the API and, over the past month have practiced using the Vivado + RapidWright toolchain to do basic tasks like extracting the netlist of a synthesized VHDL design.
-    You can see my preliminary work in my Github: \url{https://github.com/TotoroTron/place-and-route}.
-
-\section{Foundational Knowledge (so far)}
-    In mathematics, a graph is modeled as \( G = (V, E) \), where \( V \) represents the set of vertices and \( E \) represents the set of edges between them.
-    A hypergraph is denoted as \( G_{H} = (V_{H}, E_{H}) \), where \( V_{H} \) represents the set of vertices, and \( E_{H} \) represents the set of hyperedges, which are edges that can connect more than two vertices.
-    Electronic circuits are typically represented as hypergraphs because a voltage source pin can have one or many sink pins. 
-    In FPGA and VLSI design, the hypergraph is transformed into something called a "netlist", which is a hypergraph that has been flattened down into a graph. 
-    For every hyperedge in the hypergraph, transform the hyperedge into a set of 2-pin edges, each having one source to one sink.
-    The resulting graph can be represented by \( G = (V, E) \), where \( V \) represnts the ports (pins) of all modules (logic gates, )
-    There are different strategies for this reduction - unoriented star model, clique model, etc.. 
-    \cite{AP_2000}
-
-    The vast majority of existing placers use the Half Perimeter Wire Length (HPWL) cost function in one of its various flavors - manhattan distance, euclidian distance, bounding box, etc..
-    The manhattan and euclidian HPWL models the cost function as the summation of the lengths of all nets in the netlist. 
-    The manhattan HPWL is convex while the euclidian HPWL is strictly convex.
-    We can use smooth methods to minimize the euclidian, however, it produces worse Quality of Result (QoR) than the manhattan \cite{AP_2012}, which is something to consider.
-    The bounding box HPWL models the cost function as simply the longest net in the netlist.
-    Shown below in (1) and (2) are the manhattan and euclidian distance HPWL objective functions respectively.
-    Shown in (3) is the bounding box version.
-
-    \begin{equation}
-        \Phi(\vec{x}, \vec{y}) = \sum_{i,j} w_{i,j} \left( |x_i - x_j| + |y_i - y_j| \right)
-        \label{Manhattan}
-    \end{equation}
-
-    \begin{equation}
-        \Phi(\vec{x}, \vec{y}) = \sum_{i,j} w_{i,j} \left[ (x_i - x_j)^2 + (y_i - y_j)^2 \right]
-        \label{Euclidian}
-    \end{equation}
-
-    \begin{equation}
-        \Phi(\vec{x}, \vec{y}) = \sum_{i=1}^{n} \left( \max_{j \in e_i} (x_j) - \min_{j \in e_i} (x_j) + \max_{j \in e_i} (y_j) - \min_{j \in e_i} (y_j) \right)
-        \label{BoundingBox}
-    \end{equation}
-
-    \( x = \{ x_{1}, x_{2}, ..., x_{N} \} \) represents the set of physical x-coordinates of the logic block pins on the chip and \( y = \{ y_{1}, y_{2}, ..., y_{N} \} \) represents the set of y-coordinates.
-    The units for these coordinates can be nanometers in VLSI design, but for FPGAs where the logic blocks are uniform size and arranged in a grid, the coordinates can simply be positive integers.
-    \( N \) is the total number of logic block pins in the design. \( E = \{ e_{1}, e_{2}, ..., e_{M} \} \) represents the netlist with M total nets.
-
-    We want to minimize the total wirelength in our design for two reasons: power consumption, and clock skew, and routing.
-    FPGAs are commonly used for mission critical applications and edge computing devices that are required to run on low power.
-    We can reduce the power consumption of the FPGA by reducing the power dissipation of current running through wires.
-    This is particularly important for FPGAs because signal paths must propagate not only through wires but also switchboxes and programmable interconnects which can also dissipate power.
-    The second and more important reason is to minimize clock skew. 
-    Signal propagation through wires and transistors takes time. 
-    The clock signal is crucial for state transition in an FSM.
-    If the clock signal drives two logic blocks that were placed too far apart, they may become desynchronized as the same clock edge may reach the two flip-flops at different clock periods.
-    All flip-flops operating on the same clock domain must strike at the same nanosecond.
-
-    The third and equally relevant reason is the simplify the routing stage which takes place after placement.
-    If the logic blocks are all placed close together but in such a way that the wiring between them will become spaghetti, the router will produce an over-congested wiring and potentially risk routing failure because of it.
-    The placer ideally must cluster relevant logic blocks into semi-islolated neighborhood such that the wiring in one neighborhood to another is minimized.
-
-\section{Abbreviations}
-    \begin{itemize}[label={--}, left=0.25cm]
-        \item \textbf{FPGA}: Field Programmable Gate Array
-        \item \textbf{VLSI}: Very Large Scale Integration
-        \item \textbf{EDA}: Electronic Design Automation
-        \item \textbf{VHSIC}: Very High Speed Integrated Circuits
-        \item \textbf{HDL}: Hardware Description Language
-        \item \textbf{VHDL}: VHSIC HDL
-        \item \textbf{HLS}: High Level Synthesis: Generating synthesizable HDL from high-level software languages.
-A company might want to have a software engineer write C or C++ code and have a program translate it into synthesizable Verilog.
-This can can boost productivity and save the company the need to hire a hardware engineer.
-        \item \textbf{IP}: Intellectual Property: In FPGA context, this means pre-built modules or subsystems like a hardened microprocessor or Ethernet controller. These are usually proprietary.
-        \item \textbf{SoC}: System on Chip: An FPGA device (chip) that features hardened IP in addition to the programmable logic fabric.
-        \item \textbf{PL-PS}: Programmable Logic - Processing System: A design that utilizes the on-chip hard microprocessor in conjunction with the programmable logic fabric.
-        \item \textbf{EDIF}: Electronic Design Interchange Format
-        \item \textbf{HPWL}: Half Perimeter Wire Length
-    \end{itemize}
-
-\section{Ideas}
-    \begin{itemize}[label={\textbullet}, left=0.25cm]
-        \item \textbf{FPGA}: Field Programmable Gate Array
-        \begin{itemize}[label={--}, left=0.25cm]
-            \item FPGA Vendors:
-            \begin{itemize}[label={$\cdot$}, left=0.25cm]
-                \item AMD-Xilinx ($\sim$50\% FPGA vendor market share)
-                \item Intel-Altera ($\sim$35\% share)
-                \item Lattice
-                \item Microsemi
-            \end{itemize}
-        \end{itemize}
-
-        \item \textbf{EDA}: Electronic Design Automation
-        \begin{itemize}[label={--}, left=0.25cm]
-            \item Proprietary software for FPGA and VLSI development:
-            \begin{itemize}[label={$\cdot$}, left=0.25cm]
-                \item Xilinx - Vivado (Design + Simulation) + Vitis (HLS + PL-PS codesign)
-                \item Altera - Quartus (Design) + ModelSim (Simulation)
-                \item Synopsis (VLSI)
-                \item Cadence (VLSI)
-            \end{itemize}
-            \item Open source software for FPGA development:
-            \begin{itemize}[label={$\cdot$}, left=0.25cm]
-                \item \textbf{VTR}: Simulated Annealing placer for FPGAs. Popular among researchers who study placement techniques. 
-                    Commonly referred to as an "academic placer".
-                \item \textbf{OSS-CAD}: a full-flow software suite that includes ABC synthesis, Yosys synthesis, Yosys nextpnr.
-                \item \textbf{AMF-Placer}: Analytical Placer for FPGAs
-                \item \textbf{RapidWright}: Semi-open source Java API that provides backend access to Xilinx Vivado EDA using design checkpoints.
-                \item \textbf{RapidLayout}: Hard Block Placer for Systolic Arrays. Built with RapidWright.
-                \item \textbf{RapidStream}: HLS Placer. Built with RapidWright.
-                \item \textbf{DREAMPlace}: GPU-powered deep learning placement for VLSI.
-                \item \textbf{DREAMPlaceFPGA}: DREAMPlace, adapted to FPGAs via the RapidWright API.
-            \end{itemize}
-        \end{itemize}
-
-        \item \textbf{Synthesis}
-        \begin{itemize}[label={--}, left=0.25cm]
-            \item Takes a design written in a high-level HDL like VHDL or Verilog and "synthesizes" a \textbf{logical netlist} out of it. 
-            \item The logical netlist is usually generated as an EDIF, JSON, or a low-level Verilog file. 
-            \item The netlist describes the necessary basic elements of logic (BELs) and the wired connections between them that are necessary to implement the design.
-        \end{itemize}
-
-        \item \textbf{Placement}
-        \begin{itemize}[label={--}, left=0.25cm]
-            \item Takes the \textbf{logical netlist} and produces a \textbf{physical netlist}.
-            \item For each BEL in the netlist, assign the BEL to a Cell, Site, and Tile on the physical FPGA device.
-        \end{itemize}
-
-        \item \textbf{Routing}
-        \begin{itemize}[label={--}, left=0.25cm]
-            \item Takes the \textbf{physical netlist} and maps the connections between BELs onto wires, interconnects, and switchboxes on the FPGA.
-        \end{itemize}
-
-    \end{itemize}
-
-    \newpage
-    \begin{figure}
-        \begin{center}
-            \includegraphics[width=0.75\textwidth]{figures/kraftwerk2.png}
-        \end{center}
-        \caption{Landscape of VLSI placement techniques (Spindler) \cite{kraftwerk2} }
-        \label{fig:kraftwerk2}
-    \end{figure}
-    \begin{figure}
-        \begin{center}
-            \includegraphics[width=0.95\textwidth]{figures/ProCha.png}
-        \end{center}
-        \caption{Historical timeline of VLSI placement techniques (Markov) \cite{ProCha} }
-        \label{fig:ProCha}
-    \end{figure}
-
-    This is a citation for AMFPlacer. \cite{AMFPlacer}
-
-    \newpage
-    \bibliographystyle{ieeetr}
-    \nocite{*}
-    \bibliography{
-        references/surveys,
-        references/rapidwright,
-        references/fpga_placement,
-        references/vlsi_placement,
-    }
-
-\end{document}

### 2025-03-05 java/src/main/java/placer/ImageMaker.java

@@ -46,8 +46,8 @@ public class ImageMaker {
     private static final Map<SiteTypeEnum, Color> SITE_TYPE_COLORS = new HashMap<>();
     static {
         // Dull empty colors
-        SITE_TYPE_COLORS.put(SiteTypeEnum.SLICEL, Color.decode("#3A4F5F")); // Dark Gray
-        SITE_TYPE_COLORS.put(SiteTypeEnum.SLICEM, Color.decode("#3A4F5F")); // Dark Gray
+        SITE_TYPE_COLORS.put(SiteTypeEnum.SLICEL, Color.decode("#3A4F5F")); // Dull Blue
+        SITE_TYPE_COLORS.put(SiteTypeEnum.SLICEM, Color.decode("#5F533A")); // Dull Orange
         SITE_TYPE_COLORS.put(SiteTypeEnum.RAMB18E1, Color.decode("#775555")); // Dull Red
         SITE_TYPE_COLORS.put(SiteTypeEnum.FIFO18E1, Color.decode("#775555")); // Dull Red
         SITE_TYPE_COLORS.put(SiteTypeEnum.RAMB36E1, Color.decode("#775555")); // Dull Red
@@ -59,7 +59,7 @@ public class ImageMaker {
     static {
         // Bright active colors
         PLACEMENT_COLORS.put(SiteTypeEnum.SLICEL, Color.decode("#00FFFF")); // Cyan
-        PLACEMENT_COLORS.put(SiteTypeEnum.SLICEM, Color.decode("#00FFFF")); // Cyan
+        PLACEMENT_COLORS.put(SiteTypeEnum.SLICEM, Color.decode("#FFA300")); // Orange
         PLACEMENT_COLORS.put(SiteTypeEnum.RAMB18E1, Color.decode("#FF00FF")); // Magenta
         PLACEMENT_COLORS.put(SiteTypeEnum.FIFO18E1, Color.decode("#FF00FF")); // Magenta
         PLACEMENT_COLORS.put(SiteTypeEnum.RAMB36E1, Color.decode("#FF00FF")); // Magenta

### 2025-03-05 java/src/main/java/placer/Main.java

@@ -56,9 +56,12 @@ public class Main {
             // takes the packedDesign and figures out an optimal mapping of SiteInsts onto
             // Sites via simulated annealing, analytical, electrostatic placement, etc.
             // works entirely on the SiteInst/Site/Tile level.
-            PlacerAnnealRandom SARPlacer = new PlacerAnnealRandom(
+            // PlacerAnnealRandom SARPlacer = new PlacerAnnealRandom(
+            // rootDir, design, device, regionConstraint);
+            // SARPlacer.run(packedDesign, prepackedDesign);
+            PlacerAnnealMidpoint PAMPlacer = new PlacerAnnealMidpoint(
                     rootDir, design, device, regionConstraint);
-            SARPlacer.run(packedDesign, prepackedDesign);
+            PAMPlacer.run(packedDesign, prepackedDesign);
 
         } catch (IOException e) {
             logger.log(Level.SEVERE, "An IOException occurred while configuring the logger.", e);

### 2025-03-05 java/src/main/java/placer/Placer.java

@@ -82,9 +82,10 @@ public abstract class Placer {
 
     protected abstract void placeDesign(PackedDesign packedDesign) throws IOException;
 
-    protected abstract Site proposeSite(SiteTypeEnum ste, boolean swapEnable);
+    // protected abstract Site proposeSite(SiteTypeEnum ste, boolean swapEnable);
 
-    protected abstract Site proposeAnchorSite(SiteTypeEnum ste, int chainSize, boolean swapEnable);
+    // protected abstract Site proposeAnchorSite(SiteTypeEnum ste, int chainSize,
+    // boolean swapEnable);
 
     protected abstract void randomInitSingleSite(List<SiteInst> siteInsts) throws IOException;
 
@@ -229,6 +230,8 @@ public abstract class Placer {
             siteTypePrefix = "DSP48_";
         else if (siteType == SiteTypeEnum.SLICEL || siteType == SiteTypeEnum.SLICEM)
             siteTypePrefix = "SLICE_";
+        else if (siteType == SiteTypeEnum.RAMB18E1)
+            siteTypePrefix = "RAMB18_";
         else
             throw new IllegalStateException("ERROR: Could not assign a String prefix to  SiteTypeEnum: " + siteType);
         return siteTypePrefix;

### 2025-03-05 java/src/main/java/placer/PlacerAnnealMidpoint.java

@@ -3,26 +3,34 @@ package placer;
 
 import java.io.IOException;
 import java.util.Collections;
+import java.util.stream.Collectors;
+import java.util.Collection;
 import java.util.HashMap;
 import java.util.List;
 import java.util.ArrayList;
 import java.util.HashSet;
+import java.util.Random;
 
 import com.xilinx.rapidwright.design.Design;
 import com.xilinx.rapidwright.design.SiteInst;
+import com.xilinx.rapidwright.design.SitePinInst;
 
 import com.xilinx.rapidwright.device.Device;
 import com.xilinx.rapidwright.device.Site;
 import com.xilinx.rapidwright.device.SiteTypeEnum;
+import com.xilinx.rapidwright.device.ClockRegion;
 
 public class PlacerAnnealMidpoint extends Placer {
 
-    private List<Pair<Integer, Integer>> spiralPath = SpiralPath.generateDiamondSpiral(1000);
+    private int spiralPathMaxRadius = 1000;
+    private List<Pair<Integer, Integer>> spiralPath = SpiralPath.generateDiamondSpiral(spiralPathMaxRadius);
 
-    public PlacerAnnealMidpoint(String rootDir, Design design, Device device) throws IOException {
+    public PlacerAnnealMidpoint(String rootDir, Design design, Device device, ClockRegion region)
+            throws IOException {
         super(rootDir, design, device);
         this.placerName = "PlacerAnnealMidpoint";
-        this.graphicsDir = rootDir + "/outputs/graphics/" + placerName;
+        this.graphicsDir = rootDir + "/outputs/graphics";
+        this.regionConstraint = region;
         this.uniqueSiteTypes = new HashSet<>();
         this.occupiedSites = new HashMap<>();
         this.occupiedSiteChains = new HashMap<>();
@@ -32,6 +40,7 @@ public class PlacerAnnealMidpoint extends Placer {
         this.evalTimes = new ArrayList<>();
         this.renderTimes = new ArrayList<>();
         this.writeTimes = new ArrayList<>();
+        this.rand = new Random();
     }
 
     public void placeDesign(PackedDesign packedDesign) throws IOException {
@@ -41,7 +50,7 @@ public class PlacerAnnealMidpoint extends Placer {
         design.writeCheckpoint(rootDir + "/outputs/checkpoints/init_placed.dcp");
 
         this.movesLimit = 500;
-        initCoolingSchedule(1.0f, 0.1f);
[…]

### 2025-03-06 java/src/main/java/placer/Main.java
need to compare to default vivado placement

@@ -32,6 +32,8 @@ public class Main {
             logger.setLevel(Level.ALL); // Set logging level to record all messages
             logger.log(Level.INFO, "Begin Placer...");
 
+            drawVivadoPlacement();
+
             Design design = Design.readCheckpoint(synthesizedDcp);
             Device device = Device.getDevice("xc7z020clg400-1");
 
@@ -69,6 +71,8 @@ public class Main {
     }
 
     public static void drawVivadoPlacement() throws IOException {
+        Design design = Design.readCheckpoint(rootDir + "/outputs/checkpoints/vivado_placed.dcp");
+        Device device = Device.getDevice("xc7z020clg400-1");
 
     }
 

### 2025-03-06 java/src/main/java/placer/PlacerAnnealMidpoint.java

@@ -55,7 +55,6 @@ public class PlacerAnnealMidpoint extends Placer {
         while (true) {
             if (move >= movesLimit)
                 break;
-            System.out.println("move: " + move);
 
             this.currentTemp = this.coolingSchedule.get(move);
 
@@ -66,6 +65,7 @@ public class PlacerAnnealMidpoint extends Placer {
 
             t0 = System.currentTimeMillis();
             double currCost = evaluateDesign();
+            System.out.println("Move: " + move + ", Design Cost: " + currCost);
             t1 = System.currentTimeMillis();
             evalTimes.add(t1 - t0);
 
@@ -168,15 +168,15 @@ public class PlacerAnnealMidpoint extends Placer {
         // Site awaySite = proposeSite(ste, true);
 
         Pair<Integer, Integer> homeConnsMidpt = findMidpoint(conns);
-        int x = homeConnsMidpt.key();
-        int y = homeConnsMidpt.value();
-        System.out.println("Midpoint: " + x + ", " + y);
+        int mid_x = homeConnsMidpt.key();
+        int mid_y = homeConnsMidpt.value();
+        System.out.println("Midpoint: " + mid_x + ", " + mid_y);
         int spiralPathSize;
 
         if (si.getSiteTypeEnum() == SiteTypeEnum.SLICEL || si.getSiteTypeEnum() == SiteTypeEnum.SLICEM)
-            spiralPathSize = 100000;
+            spiralPathSize = 100;
         else
-            spiralPathSize = 1000000;
+            spiralPathSize = 1000;
 
         // spiral path search for legal site
         Site selectedSite = null;
@@ -184,10 +184,14 @@ public class PlacerAnnealMidpoint extends Placer {
         while (true) {
             if (attempts > spiralPathSize)
                 throw new IllegalStateException(
-                        "ERROR: Could not propose " + ste + " site after  " + spiralPathSize + " attempts!");
+                        "ERROR: Could not propose " + ste + " site after  " + spiralPathSize + " attempts!"
+                                + "\n Site: " + si.getSite());
             int dx = spiralPath.get(attempts).key();
             int dy = spiralPath.get(attempts).value();
-            selectedSite = device.getSite(getSiteTypePrefix(ste) + "X" + (x + dx) + "Y" + (y + dy));
+            int curr_x = mid_x + dx;
+            int curr_y = mid_y + dy;
+            System.out.println("(x, y): (" + curr_x + ", " + curr_y + ")");
[…]

### 2025-03-07 java/src/main/java/placer/ImageMaker.java
more work on midpoint

@@ -127,6 +127,10 @@ public class ImageMaker {
         }
     }
 
+    public Site[][] get2DSiteArray() throws IOException {
+        return this.siteArray;
+    }
+
     public void construct2DPlacementArray() throws IOException {
         this.siteInstArray = new SiteInst[width][height];
         for (SiteInst si : this.design.getSiteInsts()) {

### 2025-03-07 java/src/main/java/placer/PackerBasic2.java

@@ -24,9 +24,15 @@ import com.xilinx.rapidwright.device.SitePIPStatus;
 import com.xilinx.rapidwright.device.SiteTypeEnum;
 import com.xilinx.rapidwright.device.ClockRegion;
 
-// DOES NOT WORK!
+//
+//
+// **************
+// THIS DOES NOT WORK!
+// **************
+//
 // This Packer unPlace()s SiteInsts immediately after creating them. No initial random placement.
 // This causes routeSite() to produce errors regarding null Site.
+
 public class PackerBasic2 extends Packer {
 
     private Random rand;

### 2025-03-07 java/src/main/java/placer/Placer.java

@@ -46,10 +46,11 @@ public abstract class Placer {
     protected ClockRegion regionConstraint;
     protected Set<SiteTypeEnum> uniqueSiteTypes;
     protected Map<SiteTypeEnum, List<Site>> allSites;
+    protected Site[][] rpmGrid;
     protected Map<SiteTypeEnum, Map<Site, SiteInst>> occupiedSites;
-
     // store chain occupation info for fast site-to-chain access
     protected Map<SiteTypeEnum, Map<Site, List<SiteInst>>> occupiedSiteChains;
+
     protected Random rand;
 
     protected String[] FF5_BELS = new String[] { "A5FF", "B5FF", "C5FF", "D5FF" };
@@ -78,6 +79,21 @@ public abstract class Placer {
         placeDesign(packedDesign);
         writer.close();
         design.writeCheckpoint(placedDcp);
+    }
+
+    protected void initRpmGrid() throws IOException {
+        int rpm_x_high = 0, rpm_y_high = 0;
+        int rpm_x_low = Integer.MAX_VALUE, rpm_y_low = Integer.MAX_VALUE;
+        for (Map.Entry<SiteTypeEnum, List<Site>> entry : this.allSites.entrySet()) {
+            for (Site site : entry.getValue()) {
+                int x = site.getRpmX();
+                int y = site.getRpmY();
+                this.rpmGrid[x][y] = site;
+                //
+                // NEED TO INITIALIZE RPM GRID WITH CORRECT DIMENSIONS!
+                //
+            }
+        }
     }
 
     protected abstract void placeDesign(PackedDesign packedDesign) throws IOException;
@@ -264,6 +280,10 @@ public abstract class Placer {
             System.out.println("\tinitAnchor: " + initAnchor.getInstanceY() + ", initTail: " + initTail.getInstanceY());
             System.out.println("\tfinalAnchor: " + finalAnchorInstY + ", finalTail: " + finalTailInstY);
             for (Site site : sites) {
+                if (site == null) {
+                    System.out.println("\t\tnull site.");
+                    continue;
+                }
                 System.out.println("\t\tSiteInst: " + occupiedSites.get(site.getSiteTypeEnum()).get(site));
             }
         }

### 2025-03-07 java/src/main/java/placer/PlacerAnnealMidpoint.java

@@ -45,6 +45,7 @@ public class PlacerAnnealMidpoint extends Placer {
 
     public void placeDesign(PackedDesign packedDesign) throws IOException {
         initSites();
+        initRpmGrid();
         unplaceAllSiteInsts(packedDesign);
         randomInitialPlacement(packedDesign);
         design.writeCheckpoint(rootDir + "/outputs/checkpoints/init_placed.dcp");
@@ -162,6 +163,9 @@ public class PlacerAnnealMidpoint extends Placer {
     } // end proposeRandomSite()
 
     protected Site proposeMidpointSite(SiteInst si, List<Site> conns, boolean swapEnable) {
+
+        // https://adaptivesupport.amd.com/s/question/0D52E00006hpT8KSAU/vivado-coordinate-systems?language=en_US
+
         SiteTypeEnum ste = si.getSiteTypeEnum();
         Site homeSite = si.getSite();
         // how to make this midpoint?
@@ -170,13 +174,13 @@ public class PlacerAnnealMidpoint extends Placer {
         Pair<Integer, Integer> homeConnsMidpt = findMidpoint(conns);
         int mid_x = homeConnsMidpt.key();
         int mid_y = homeConnsMidpt.value();
-        System.out.println("Midpoint: " + mid_x + ", " + mid_y);
+        // System.out.println("Midpoint: " + mid_x + ", " + mid_y);
         int spiralPathSize;
 
         if (si.getSiteTypeEnum() == SiteTypeEnum.SLICEL || si.getSiteTypeEnum() == SiteTypeEnum.SLICEM)
-            spiralPathSize = 100;
+            spiralPathSize = 10000;
         else
-            spiralPathSize = 1000;
+            spiralPathSize = 10000;
 
         // spiral path search for legal site
         Site selectedSite = null;
@@ -190,8 +194,14 @@ public class PlacerAnnealMidpoint extends Placer {
             int dy = spiralPath.get(attempts).value();
             int curr_x = mid_x + dx;
             int curr_y = mid_y + dy;
-            System.out.println("(x, y): (" + curr_x + ", " + curr_y + ")");
-            selectedSite = device.getSite(getSiteTypePrefix(ste) + "X" + curr_x + "Y" + curr_y);
+            // System.out.println("(x, y): (" + curr_x + ", " + curr_y + ")");
+
+            selectedSite = rpmGrid[curr_x][curr_y];
[…]

### 2025-03-07 run.sh

@@ -166,11 +166,16 @@ if [ "$start_stage" == "video" ] || [ "$start_stage" == "all" ]; then
     cd $PROJ_DIR
     python3 python/plot_convergence.py
     FPS=10
-    OUTPUT="../video.mp4"
     INPUT="outputs/graphics/images"
     INPUT_PATTERN="%08d.png"
     cd $INPUT
-    ffmpeg -y -framerate $FPS -i $INPUT_PATTERN -vcodec libx264 "$OUTPUT" -vf "pad=ceil(iw/2)*2:ceil(ih/2)*2"
+
+    # OUTPUT="../video.mp4"
+    # ffmpeg -y -framerate $FPS -i $INPUT_PATTERN -vcodec libx264 "$OUTPUT" -vf "pad=ceil(iw/2)*2:ceil(ih/2)*2"
+
+    OUTPUT="../output.gif"
+    ffmpeg -y -framerate $FPS -i $INPUT_PATTERN -vf "pad=ceil(iw/2)*2:ceil(ih/2)*2" -loop 0 "$OUTPUT"
+
     # height not divisible by 2 error:
     # https://stackoverflow.com/questions/20847674/ffmpeg-libx264-height-not-divisible-by-2
 fi

### 2025-03-10 java/src/main/java/placer/Main.java
lots of reorganizing

@@ -51,9 +51,9 @@ public class Main {
             ClockRegion region = null;
             PackerBasic1 B1Packer = new PackerBasic1(
                     rootDir, design, device, region);
+            PackedDesign packedDesign = B1Packer.run(prepackedDesign);
             B1Packer.printUniqueSites();
             B1Packer.printClockBuffers();
-            PackedDesign packedDesign = B1Packer.run(prepackedDesign);
 
             // Stage 3) Placer:
             // takes the packedDesign and figures out an optimal mapping of SiteInsts onto

### 2025-03-10 java/src/main/java/placer/Packer.java

@@ -12,6 +12,7 @@ import java.util.HashMap;
 import java.util.HashSet;
 import java.util.Set;
 
+import java.io.File;
 import java.io.FileWriter;
 import java.io.BufferedWriter;
 import java.io.IOException;
@@ -57,7 +58,12 @@ public abstract class Packer {
     }
 
     public PackedDesign run(PrepackedDesign prepackedDesign) throws IOException {
-        writer = new FileWriter(rootDir + "/outputs/printout/" + packerName + ".txt");
+        String printoutDir = rootDir + "/outputs/packers";
+        File printoutFile = new File(printoutDir);
+        if (!printoutFile.exists()) {
+            printoutFile.mkdirs();
+        }
+        writer = new FileWriter(printoutDir + "/" + packerName + ".txt");
         writer.write(packerName + ".txt");
         PackedDesign packedDesign = packDesign(prepackedDesign);
         writer.close();
@@ -356,7 +362,7 @@ public abstract class Packer {
 
     public void printClockBuffers() throws IOException {
         BufferedWriter writer = new BufferedWriter(
-                new FileWriter(rootDir + "/outputs/printout/DeviceGlobalClkBuffers.txt"));
+                new FileWriter(rootDir + "/outputs/packers/DeviceGlobalClkBuffers.txt"));
 
         Site[] sites = device.getAllSites();
 
@@ -387,7 +393,7 @@ public abstract class Packer {
     }
 
     public Set<SiteTypeEnum> printUniqueSites() throws IOException {
-        BufferedWriter writer = new BufferedWriter(new FileWriter(rootDir + "/outputs/printout/DeviceUniqueSites.txt"));
+        BufferedWriter writer = new BufferedWriter(new FileWriter(rootDir + "/outputs/packers/DeviceUniqueSites.txt"));
         writer.write("\nPrinting unique sites in the device: ");
         writer.newLine();
         Site[] sites = device.getAllSites();
@@ -434,7 +440,7 @@ public abstract class Packer {
 
     public void printSitesOfType(SiteTypeEnum type) throws IOException {
         BufferedWriter writer = new BufferedWriter(
-                new FileWriter(rootDir + "/outputs/printout/AllSites_" + type + ".txt"));
+                new FileWriter(rootDir + "/outputs/packers/AllSites_" + type + ".txt"));
         Site[] sites = device.getAllSitesOfType(type);
 
         writer.write("Found sites of type " + type + "(" + sites.length + ").");

### 2025-03-10 java/src/main/java/placer/PackerBasic1.java

@@ -122,7 +122,7 @@ public class PackerBasic1 extends Packer {
 
         ImageMaker im = new ImageMaker(design);
         im.renderAll();
-        im.exportImage(rootDir + "/outputs/graphics/init_packing.png");
+        im.exportImage(rootDir + "/outputs/packers/init_packing.png");
         return packedDesign;
 
     } // end placeDesign()

### 2025-03-10 java/src/main/java/placer/Placer.java

@@ -81,7 +81,7 @@ public abstract class Placer {
 
     public void makeOutputDirs(String placerName) throws IOException {
         this.placerName = placerName;
-        String resultsDir = rootDir + "/outputs/results/" + placerName;
+        String resultsDir = rootDir + "/outputs/placers/" + placerName;
         this.placedDcp = resultsDir + "/checkpoints";
         File placedDcp = new File(this.placedDcp);
         if (!placedDcp.exists()) {

### 2025-03-10 java/src/main/java/placer/PlacerAnnealRandom.java

@@ -74,7 +74,7 @@ public class PlacerAnnealRandom extends Placer {
         ImageMaker imPlaced = new ImageMaker(design);
         imPlaced.renderAll();
         imPlaced.exportImage(graphicsDir + "/final_placement.png");
-        exportCostHistory(rootDir + "/outputs/printout/convergence.csv");
+        exportCostHistory(printoutDir + "/cost_history.csv");
         printTimingBenchmarks();
         writer.write("\n\nTotal move iterations: " + move);
     }

### 2025-03-10 python/plot_convergence.py

@@ -26,4 +26,4 @@ def plot_csv(filename, output_file):
     plt.savefig(output_file, dpi=200, bbox_inches='tight')
 
 if __name__ == "__main__":
-    plot_csv("outputs/printout/convergence.csv", "outputs/graphics/convergence.png")
+    plot_csv("outputs/placers/PlacerAnnealRandom/printout/cost_history.csv", "outputs/placers/PlacerAnnealRandom/graphics/cost_history.png")

### 2025-03-10 run.sh

@@ -165,7 +165,7 @@ if [ "$start_stage" == "video" ] || [ "$start_stage" == "all" ]; then
     cd $PROJ_DIR
     python3 python/plot_convergence.py
     FPS=10
-    INPUT="outputs/graphics/images"
+    INPUT="outputs/placers/PlacerAnnealRandom/graphics/images"
     INPUT_PATTERN="%08d.png"
     cd $INPUT
 

### 2025-03-10 java/src/main/java/placer/Main.java
fixed FIFO support for packer init placement. FIFO support still broken in placer

@@ -61,9 +61,9 @@ public class Main {
             // works entirely on the SiteInst/Site/Tile level.
             List<PlacerAnnealRandom> SAPlacers = new ArrayList<PlacerAnnealRandom>();
             SAPlacers.add(new PlacerAnnealRandom(rootDir, design, device, region));
-            SAPlacers.add(new PlacerAnnealMidpoint(rootDir, design, device, region));
-            SAPlacers.add(new PlacerGreedyRandom(rootDir, design, device, region));
-            SAPlacers.add(new PlacerGreedyMidpoint(rootDir, design, device, region));
+            // SAPlacers.add(new PlacerAnnealMidpoint(rootDir, design, device, region));
+            // SAPlacers.add(new PlacerGreedyRandom(rootDir, design, device, region));
+            // SAPlacers.add(new PlacerGreedyMidpoint(rootDir, design, device, region));
 
             for (PlacerAnnealRandom placer : SAPlacers) {
                 placer.makeOutputDirs(placer.getPlacerName());

### 2025-03-10 java/src/main/java/placer/PackerBasic1.java

@@ -212,24 +212,10 @@ public class PackerBasic1 extends Packer {
         List<SiteInst> RAMSiteInsts = new ArrayList<>();
         writer.write("\n\nPacking RAMBCells... (" + RAMCells.size() + ")");
         for (EDIFHierCellInst ehci : RAMCells) {
-            Site selectedSite = selectRAMSite();
-            SiteTypeEnum ste = selectedSite.getSiteTypeEnum();
-            SiteInst si = null;
-            if (ste == SiteTypeEnum.RAMB18E1) {
-                si = new SiteInst(
-                        ehci.getFullHierarchicalInstName(), design, SiteTypeEnum.RAMB18E1, selectedSite);
-                si.createCell(ehci, si.getBEL("RAMB18E1"));
-            }
-            if (ste == SiteTypeEnum.FIFO18E1) {
-                si = new SiteInst(
-                        ehci.getFullHierarchicalInstName(), design, SiteTypeEnum.FIFO18E1, selectedSite);
-                si.createCell(ehci, si.getBEL("RAMB18E1"));
-                System.out.println("SiteInst:" + si);
-                System.out.println("\tCell-BEL Map:");
-                for (Map.Entry<String, Cell> entry : si.getCellMap().entrySet()) {
-                    System.out.println("\t\t<" + entry.getKey() + ", " + entry.getValue() + ">");
-                }
-            }
+            Site selectedSite = selectRAMSite(); // selects uniformly randomly between RAMB18E1 and FIFO18E1 Site
+            SiteInst si = new SiteInst(
+                    ehci.getFullHierarchicalInstName(), design, SiteTypeEnum.RAMB18E1, selectedSite);
+            si.createCell(ehci, si.getBEL("RAMB18E1"));
             si.routeSite();
             RAMSiteInsts.add(si);
         }
@@ -402,8 +388,8 @@ public class PackerBasic1 extends Packer {
                     "ERROR: device or clock region contains no Sites of type FIFO18E1 or RAMB18E1!");
         }
         compatibleSites.sort(
-                Comparator.comparingInt(Site::getInstanceX)
-                        .thenComparingInt(Site::getInstanceY)
+                Comparator.comparingInt(Site::getInstanceY)
+                        .thenComparingInt(Site::getInstanceX)
                         .reversed());
         Site selectedSite = compatibleSites.get(0);
         SiteTypeEnum ste = selectedSite.getSiteTypeEnum();

### 2025-03-10 java/src/main/java/placer/PlacerAnnealRandom.java

@@ -110,7 +110,15 @@ public class PlacerAnnealRandom extends Placer {
     }
 
     protected Site proposeSite(SiteInst si, List<Site> connections, boolean swapEnable) {
-        SiteTypeEnum ste = si.getSiteTypeEnum();
+        SiteTypeEnum ste = null;
+        SiteTypeEnum[] altStes = si.getAlternateSiteTypeEnums();
+        if (altStes.length == 0) {
+            ste = si.getSiteTypeEnum();
+        } else {
+            int randIndex = rand.nextInt(altStes.length);
+            ste = altStes[randIndex];
+            System.out.println(ste);
+        }
         Site selectedSite = null;
         int attempts = 0;
         while (true) {

### 2025-03-10 java/src/main/java/placer/PlacerAnnealRandom.java
explicitly choose between FIFO and RAMB in proposeSite(). not robust but works

@@ -111,14 +111,22 @@ public class PlacerAnnealRandom extends Placer {
 
     protected Site proposeSite(SiteInst si, List<Site> connections, boolean swapEnable) {
         SiteTypeEnum ste = null;
-        SiteTypeEnum[] altStes = si.getAlternateSiteTypeEnums();
-        if (altStes.length == 0) {
-            ste = si.getSiteTypeEnum();
-        } else {
-            int randIndex = rand.nextInt(altStes.length);
-            ste = altStes[randIndex];
+        if (si.getSiteTypeEnum() == SiteTypeEnum.RAMB18E1) {
+            SiteTypeEnum[] compatibleStes = { SiteTypeEnum.RAMB18E1, SiteTypeEnum.FIFO18E1 };
+            int randIndex = rand.nextInt(compatibleStes.length);
+            ste = compatibleStes[randIndex];
             System.out.println(ste);
+        } else {
+            ste = si.getSiteTypeEnum();
         }
+        // SiteTypeEnum[] altStes = si.getAlternateSiteTypeEnums();
+        // if (altStes.length == 0) {
+        // ste = si.getSiteTypeEnum();
+        // } else {
+        // int randIndex = rand.nextInt(altStes.length);
+        // ste = altStes[randIndex];
+        // System.out.println(ste);
+        // }
         Site selectedSite = null;
         int attempts = 0;
         while (true) {

### 2025-03-10 java/src/main/java/placer/ImageMaker.java
update route tcl. price-in the clock net during hpwl eval. show clock net in imagemaker

@@ -203,10 +203,11 @@ public class ImageMaker {
             if (net.isStaticNet())
                 continue;
             SitePinInst src = net.getSource();
-            if (src == null) // SPI is null if the net is purely intrasite
+            if (src == null) { // SPI is null if the net is purely intrasite
                 continue;
-            if (src.getSiteTypeEnum().equals(SiteTypeEnum.BUFGCTRL))
-                continue;
+            }
+            // if (src.getSiteTypeEnum().equals(SiteTypeEnum.BUFGCTRL))
+            // continue;
             List<SitePinInst> sinks = net.getSinkPins();
             Site srcSite = src.getSite();
             if (srcSite == null)

### 2025-03-10 java/src/main/java/placer/Placer.java

@@ -219,10 +219,10 @@ public abstract class Placer {
     protected double evaluateSite(List<Site> sinkSites, Site srcSite) throws IOException {
         double cost = 0;
         for (Site sinkSite : sinkSites) {
-            // if (sinkSite.isGlobalClkBuffer()) {
-            // // System.out.println("Skipped global clock buffer evaluation.");
-            // continue;
-            // }
+            if (sinkSite.isGlobalClkBuffer()) {
+                // System.out.println("Skipped global clock buffer evaluation.");
+                continue;
+            }
             // if (sinkSite.isGlobalClkPad()) {
             // // System.out.println("Skipped global clock pad evaluation.");
             // continue;

### 2025-03-10 java/src/main/java/placer/PlacerAnnealRandom.java

@@ -115,7 +115,7 @@ public class PlacerAnnealRandom extends Placer {
             SiteTypeEnum[] compatibleStes = { SiteTypeEnum.RAMB18E1, SiteTypeEnum.FIFO18E1 };
             int randIndex = rand.nextInt(compatibleStes.length);
             ste = compatibleStes[randIndex];
-            System.out.println(ste);
+            // System.out.println(ste);
         } else {
             ste = si.getSiteTypeEnum();
         }

### 2025-03-11 java/src/main/java/placer/Main.java
hybrid midpoint/random annealer

@@ -61,13 +61,14 @@ public class Main {
             // works entirely on the SiteInst/Site/Tile level.
             List<PlacerAnnealRandom> SAPlacers = new ArrayList<PlacerAnnealRandom>();
             SAPlacers.add(new PlacerAnnealRandom(rootDir, design, device, region));
+            SAPlacers.add(new PlacerAnnealHybrid(rootDir, design, device, region));
             // SAPlacers.add(new PlacerAnnealMidpoint(rootDir, design, device, region));
             // SAPlacers.add(new PlacerGreedyRandom(rootDir, design, device, region));
             // SAPlacers.add(new PlacerGreedyMidpoint(rootDir, design, device, region));
 
             for (PlacerAnnealRandom placer : SAPlacers) {
                 placer.makeOutputDirs(placer.getPlacerName());
-                placer.initCoolingSchedule(10000.0d, 0.95d, 250);
+                placer.initCoolingSchedule(10000.0d, 0.99d, 300);
                 placer.run(packedDesign);
             }
 

### 2025-03-11 java/src/main/java/placer/PlacerAnnealHybrid.java

@@ -1,0 +1,64 @@
+package placer;
+
+import java.io.IOException;
+import java.util.Collections;
+import java.util.stream.Collectors;
+import java.util.Collection;
+import java.util.HashMap;
+import java.util.List;
+import java.util.ArrayList;
+import java.util.HashSet;
+import java.util.Random;
+
+import com.xilinx.rapidwright.design.Design;
+import com.xilinx.rapidwright.design.SiteInst;
+import com.xilinx.rapidwright.design.SitePinInst;
+
+import com.xilinx.rapidwright.device.Device;
+import com.xilinx.rapidwright.device.Site;
+import com.xilinx.rapidwright.device.SiteTypeEnum;
+import com.xilinx.rapidwright.device.ClockRegion;
+
+public class PlacerAnnealHybrid extends PlacerAnnealMidpoint {
+    private String placerName = "PlacerAnnealHybrid";
+
+    public PlacerAnnealHybrid(String rootDir, Design design, Device device, ClockRegion region) throws IOException {
+        super(rootDir, design, device, region);
+    }
+
+    @Override
+    public String getPlacerName() {
+        return this.placerName;
+    }
+
+    @Override
+    protected Site proposeSite(SiteInst si, List<Site> connections, boolean swapEnable) {
+        // for random intial placements
+        if (connections == null || connections.isEmpty()) {
+            return proposeRandomSite(si, connections, swapEnable);
+        }
+        // for movement during general placement
+        // hybrid: 50% chance midpoint, 50% chance random
+        if (rand.nextInt(2) == 0) {
+            return proposeMidpointSite(si, connections, swapEnable);
+        } else {
+            return proposeRandomSite(si, connections, swapEnable);
+        }
+    }
+
+    @Override
+    protected Site proposeAnchorSite(List<SiteInst> chain, List<Site> connections, boolean swapEnable) {
+        // for random intial placements
+        if (connections == null || connections.isEmpty()) {
+            return proposeRandomAnchorSite(chain, connections, swapEnable);
+        }
+        // for movement during general placement
+        // hybrid: 50% chance midpoint, 50% chance random
+        if (rand.nextInt(2) == 0) {
+            return proposeMidpointAnchorSite(chain, connections, swapEnable);
+        } else {
+            return proposeRandomAnchorSite(chain, connections, swapEnable);
+        }
+
+    }
+}

### 2025-03-11 java/src/main/java/placer/PlacerAnnealMidpoint.java

@@ -52,8 +52,23 @@ public class PlacerAnnealMidpoint extends PlacerAnnealRandom {
     @Override
     protected Site proposeSite(SiteInst si, List<Site> connections, boolean swapEnable) {
         if (connections == null || connections.isEmpty()) {
-            return super.proposeSite(si, connections, swapEnable);
+            return proposeRandomSite(si, connections, swapEnable);
+        } else {
+            return proposeMidpointSite(si, connections, swapEnable);
         }
+    } // end proposeSite()
+
+    @Override
+    protected Site proposeAnchorSite(List<SiteInst> chain, List<Site> connections, boolean swapEnable) {
+        if (connections == null || connections.isEmpty()) {
+            return proposeRandomAnchorSite(chain, connections, swapEnable);
+        } else {
+            return proposeMidpointAnchorSite(chain, connections, swapEnable);
+        }
+
+    }
+
+    protected Site proposeMidpointSite(SiteInst si, List<Site> connections, boolean swapEnable) {
         // https://adaptivesupport.amd.com/s/question/0D52E00006hpT8KSAU/vivado-coordinate-systems?language=en_US
         SiteTypeEnum ste = si.getSiteTypeEnum();
         Pair<Integer, Integer> connsMidpt = findMidpoint(connections);
@@ -109,12 +124,12 @@ public class PlacerAnnealMidpoint extends PlacerAnnealRandom {
             break;
         }
         return selectedSite;
-    } // end proposeSite()
 
-    @Override
-    protected Site proposeAnchorSite(List<SiteInst> chain, List<Site> connections, boolean swapEnable) {
+    }
+
+    protected Site proposeMidpointAnchorSite(List<SiteInst> chain, List<Site> connections, boolean swapEnable) {
         if (connections == null || connections.isEmpty()) {
-            return super.proposeAnchorSite(chain, connections, swapEnable);
+            return proposeRandomAnchorSite(chain, connections, swapEnable);
         }
 
         boolean validAnchor = false;
@@ -132,7 +147,11 @@ public class PlacerAnnealMidpoint extends PlacerAnnealRandom {
             int dx = spiralPath.get(attempts).key();
             int dy = spiralPath.get(attempts).value();
             int curr_x = mid_x + dx;
+            if (curr_x < 0)
+                curr_x = 0;
             int curr_y = mid_y + dy;
+            if (curr_y < 0)
+                curr_y = 0;
[…]

### 2025-03-11 java/src/main/java/placer/PlacerAnnealRandom.java

@@ -49,18 +49,18 @@ public class PlacerAnnealRandom extends Placer {
             this.currentTemp = this.coolingSchedule.get(move);
 
             long t0 = System.currentTimeMillis();
-            move(packedDesign);
-            long t1 = System.currentTimeMillis();
-            moveTimes.add(t1 - t0);
-
-            t0 = System.currentTimeMillis();
             double currCost = evaluateDesign();
             if (move % 20 == 0)
-                System.out.println("Move: " + move + ", Design Cost: " + currCost);
-            t1 = System.currentTimeMillis();
+                System.out
+                        .println("Move: " + move + ", Design Cost: " + currCost + ", Temperature: " + this.currentTemp);
+            long t1 = System.currentTimeMillis();
             evalTimes.add(t1 - t0);
-
             this.costHistory.add(currCost);
+
+            t0 = System.currentTimeMillis();
+            move(packedDesign);
+            t1 = System.currentTimeMillis();
+            moveTimes.add(t1 - t0);
 
             t0 = System.currentTimeMillis();
             ImageMaker gifFrame = new ImageMaker(design);
@@ -110,6 +110,14 @@ public class PlacerAnnealRandom extends Placer {
     }
 
     protected Site proposeSite(SiteInst si, List<Site> connections, boolean swapEnable) {
+        return proposeRandomSite(si, connections, swapEnable);
+    }
+
+    protected Site proposeAnchorSite(List<SiteInst> chain, List<Site> connections, boolean swapEnable) {
+        return proposeRandomAnchorSite(chain, connections, swapEnable);
+    }
+
+    protected Site proposeRandomSite(SiteInst si, List<Site> connections, boolean swapEnable) {
         SiteTypeEnum ste = null;
         if (si.getSiteTypeEnum() == SiteTypeEnum.RAMB18E1) {
             SiteTypeEnum[] compatibleStes = { SiteTypeEnum.RAMB18E1, SiteTypeEnum.FIFO18E1 };
@@ -150,7 +158,7 @@ public class PlacerAnnealRandom extends Placer {
         return selectedSite;
     } // end proposeSite()
 
-    protected Site proposeAnchorSite(List<SiteInst> chain, List<Site> connections, boolean swapEnable) {
+    protected Site proposeRandomAnchorSite(List<SiteInst> chain, List<Site> connections, boolean swapEnable) {
         int chainSize = chain.size();
[…]

### 2025-03-11 python/plot_convergence.py

@@ -26,4 +26,5 @@ def plot_csv(filename, output_file):
     plt.savefig(output_file, dpi=200, bbox_inches='tight')
 
 if __name__ == "__main__":
-    plot_csv("outputs/placers/PlacerAnnealRandom/printout/cost_history.csv", "outputs/placers/PlacerAnnealRandom/graphics/cost_history.png")
+    plot_csv("outputs/placers/PlacerAnnealHybrid/printout/cost_history.csv", "outputs/placers/PlacerAnnealHybrid/graphics/cost_history.png")
+    # plot_csv("outputs/placers/PlacerAnnealRandom/printout/cost_history.csv", "outputs/placers/PlacerAnnealRandom/graphics/cost_history.png")

### 2025-03-11 run.sh

@@ -165,7 +165,8 @@ if [ "$start_stage" == "video" ] || [ "$start_stage" == "all" ]; then
     cd $PROJ_DIR
     python3 python/plot_convergence.py
     FPS=10
-    INPUT="outputs/placers/PlacerAnnealRandom/graphics/images"
+    # INPUT="outputs/placers/PlacerAnnealRandom/graphics/images"
+    INPUT="outputs/placers/PlacerAnnealHybrid/graphics/images"
     INPUT_PATTERN="%08d.png"
     cd $INPUT
 

### 2025-03-11 java/src/main/java/placer/ImageMaker.java
comparison between vivado and rapidwright placements. both are equal HPWL cost, but vivado placement is probably much better for routing

@@ -23,6 +23,7 @@ import com.xilinx.rapidwright.design.SitePinInst;
 import com.xilinx.rapidwright.device.Device;
 import com.xilinx.rapidwright.device.Site;
 import com.xilinx.rapidwright.device.SiteTypeEnum;
+import com.xilinx.rapidwright.device.TileTypeEnum;
 
 public class ImageMaker {
 
@@ -212,6 +213,8 @@ public class ImageMaker {
             Site srcSite = src.getSite();
             if (srcSite == null)
                 continue; // sink has not been placed yet
+            if (srcSite.getTile().getTileTypeEnum() == TileTypeEnum.RIOB33)
+                continue;
             float cost = 0;
             for (SitePinInst sink : sinks) {
                 Site sinkSite = sink.getSite();

### 2025-03-11 java/src/main/java/placer/Main.java

@@ -9,8 +9,11 @@ import java.io.IOException;
 
 import java.util.ArrayList;
 import java.util.Collection;
+
+import java.util.stream.Collectors;
 import java.util.List;
 
+import com.xilinx.rapidwright.design.Net;
 import com.xilinx.rapidwright.design.ModuleInst;
 import com.xilinx.rapidwright.design.Module;
 import com.xilinx.rapidwright.design.Design;
@@ -18,6 +21,8 @@ import com.xilinx.rapidwright.design.SiteInst;
 
 import com.xilinx.rapidwright.device.Device;
 import com.xilinx.rapidwright.device.ClockRegion;
+import com.xilinx.rapidwright.device.Tile;
+import com.xilinx.rapidwright.device.TileTypeEnum;
 
 public class Main {
 
@@ -33,7 +38,14 @@ public class Main {
             logger.setLevel(Level.ALL); // Set logging level to record all messages
             logger.log(Level.INFO, "Begin Placer...");
 
-            drawVivadoPlacement();
+            // Design vivado_design = Design.readCheckpoint(rootDir +
+            // "/outputs/checkpoints/vivado_routed.dcp");
+            // drawPlacement(vivado_design, rootDir + "/outputs/placers/vivado_routed.png");
+
+            // Design rapidwright_design = Design
+            // .readCheckpoint(rootDir + "/outputs/checkpoints/routed.dcp");
+            // drawPlacement(rapidwright_design, rootDir +
+            // "/outputs/placers/rapidwright_routed.png");
 
             Design design = Design.readCheckpoint(synthesizedDcp);
             Device device = Device.getDevice("xc7z020clg400-1");
@@ -68,7 +80,7 @@ public class Main {
 
             for (PlacerAnnealRandom placer : SAPlacers) {
                 placer.makeOutputDirs(placer.getPlacerName());
-                placer.initCoolingSchedule(10000.0d, 0.99d, 300);
+                placer.initCoolingSchedule(10000.0d, 0.98d, 300);
                 placer.run(packedDesign);
             }
 
@@ -77,9 +89,53 @@ public class Main {
         }
     }
 
-    public static void drawVivadoPlacement() throws IOException {
-        Design design = Design.readCheckpoint(rootDir + "/outputs/checkpoints/vivado_placed.dcp");
-        Device device = Device.getDevice("xc7z020clg400-1");
+    public static void drawPlacement(Design design, String filePath) throws IOException {
+        ImageMaker im = new ImageMaker(design);
+        im.renderAll();
[…]

### 2025-03-11 java/src/main/java/placer/Main.java
minor cleanup

@@ -112,56 +112,4 @@ public class Main {
         System.out.println("Design Cost: " + cost);
     }
 
-    public static void drawRapidWrightPlacement() throws IOException {
-
-        Design rapidwright_design = Design
-                .readCheckpoint(rootDir + "/outputs/placers/PlacerAnnealRandom/checkpoints/PlacerAnnealRandom.dcp");
-        ImageMaker rim = new ImageMaker(rapidwright_design);
-        rim.renderAll();
-        rim.exportImage(rootDir + "/outputs/placers/vivado_routed.png");
-
-        double cost = 0;
-        Collection<Net> nets = rapidwright_design.getNets();
-        System.out.println("Number of Nets: " + nets.size());
-        for (Net net : nets) {
-            Tile srcTile = net.getSourceTile();
-            if (srcTile == null) // tile is null if its' purely intrasite!
-                continue;
-            List<Tile> sinkTiles = net.getSinkPins().stream()
-                    .map(spi -> spi.getTile())
-                    .collect(Collectors.toList());
-            for (Tile sinkTile : sinkTiles) {
-                cost = cost + srcTile.getTileManhattanDistance(sinkTile);
-            }
-        }
-
-        System.out.println("Design Cost: " + cost);
-
-    }
-
-    public static void testModuleInst() throws IOException {
-        Design design = Design.readCheckpoint(rootDir + "/outputs/vivado_placed.dcp");
-        Collection<ModuleInst> mis = design.getModuleInsts();
-        if (mis.isEmpty()) {
-            System.out.println("Collection<ModuleInst> empty!");
-        } else {
-            for (ModuleInst mi : mis) {
-                System.out.println("\tSites in ModuleImpl: " + mi.getSiteInsts());
-                for (SiteInst si : mi.getSiteInsts()) {
-                    System.out.println("SiteInst: " + si.getName());
-                }
-            }
-        }
-
-        Collection<Module> modules = new ArrayList<>();
-        Module module = new Module(design);
-        modules.add(module);
-        for (Module m : modules) {
-            System.out.println("\tSites in Module: " + m.getName());
-            for (SiteInst si : m.getSiteInsts()) {
-                System.out.println("SiteInst: " + si.getName());
-            }
-
-        }
-    }
 }

### 2025-03-13 latex/paper/paper.tex
hybrid midpoint/random annealer
streamline gif generation. start analytical placer

@@
 \usepackage{amsmath}
+\usepackage{mathpazo}
+\usepackage{multicol}
+\usepackage[a4paper,
+            bindingoffset=0.2in,
+            left=1in,
+            right=1in,
+            top=1in,
+            bottom=1in,
+            footskip=.25in]{geometry}
 
 \begin{document}
-\title{Technical Paper Proposal: \\ Placement Algorithms for Heterogenous FPGAs}
-\author{Brian B Cheng \\ Department of Electrical and Computer Engineering}
+\title{MS Technical Paper: \\ Placement Algorithms for Heterogenous FPGAs}
+\author{Brian B Cheng \\ Rutgers University Department of Electrical and Computer Engineering}
 
@@
 \begin{itemize}
-    \item FPGA, EDA, Synthesis, Placement, Routing, Parallel, Optimization
+    \item FPGA, EDA, Placement, Simulated Annealing, Optimization, RapidWright
 \end{itemize}
 
-\section{Proposal}
-    A compiler takes a program written in high-level programming language like a Java or Python and assembles low-level machine code which is ready to be executed on a CPU.
+\begin{multicols}{2}
+\section{Abstract}
+    fdsafdsafdsa.
+
+\section{Design Flow}
+A typical FPGA design flow goes like the following. First, the developer writes the design entry in a high level Hardware Description Language (HDL) like Verilog or VHDL.
+\begin{itemize}
+\item Synthesis (Vivado default)
+\item Placement (the focus of this paper)
+    \begin{itemize}
+        \item \textbf{Prepacking}:\\
+        Traverse the raw EDIFNetlist from synthesis and identify patterns of cells that should be grouped together.
+        This can be CARRY chains, DSP cascades, LUTFF pairs, etc.
+    \item \textbf{Packing}:\\
+        Take the prepacked cell groups and pack them into SiteInsts.
+        Intra\-Site routing.
+    \item \textbf{Placement}:\\
+        Take the SiteInsts and optimally place them onto physical Sites.
+    \end{itemize}
+\item Routing
+    Inter\-Site routing
+\end{itemize}
+
+\end{multicols}
+\end{document}

### 2025-03-13 java/src/main/java/placer/Main.java

@@ -74,9 +74,9 @@ public class Main {
             List<PlacerAnnealRandom> SAPlacers = new ArrayList<PlacerAnnealRandom>();
             SAPlacers.add(new PlacerAnnealRandom(rootDir, design, device, region));
             SAPlacers.add(new PlacerAnnealHybrid(rootDir, design, device, region));
-            // SAPlacers.add(new PlacerAnnealMidpoint(rootDir, design, device, region));
-            // SAPlacers.add(new PlacerGreedyRandom(rootDir, design, device, region));
-            // SAPlacers.add(new PlacerGreedyMidpoint(rootDir, design, device, region));
+            SAPlacers.add(new PlacerAnnealMidpoint(rootDir, design, device, region));
+            SAPlacers.add(new PlacerGreedyRandom(rootDir, design, device, region));
+            SAPlacers.add(new PlacerGreedyMidpoint(rootDir, design, device, region));
 
             for (PlacerAnnealRandom placer : SAPlacers) {
                 placer.makeOutputDirs(placer.getPlacerName());

### 2025-03-13 java/src/main/java/placer/Placer.java

@@ -100,6 +100,10 @@ public abstract class Placer {
         this.writer = new FileWriter(this.printoutDir + "/" + placerName + ".txt");
     }
 
+    public abstract String getPlacerName();
+
+    public abstract void placeDesign(PackedDesign packedDesign) throws IOException;
+
     public void run(PackedDesign packedDesign) throws IOException {
         placeDesign(packedDesign);
         writer.close();
@@ -107,22 +111,116 @@ public abstract class Placer {
         DesignTools.toCSV(printoutDir + "/" + placerName, design);
     }
 
-    public abstract String getPlacerName();
+    protected void randomInitialPlacement(PackedDesign packedDesign) throws IOException {
+        randomInitSiteChains(packedDesign.DSPSiteInstCascades);
+        randomInitSiteChains(packedDesign.CARRYSiteInstChains);
+        randomInitSingleSite(packedDesign.RAMSiteInsts);
+        randomInitSingleSite(packedDesign.CLBSiteInsts);
+    }
 
-    protected abstract void placeDesign(PackedDesign packedDesign) throws IOException;
+    protected void randomInitSingleSite(List<SiteInst> siteInsts) throws IOException {
+        for (SiteInst si : siteInsts) {
+            Site selectedSite = proposeSite(si, null, false);
+            placeSiteInst(si, selectedSite);
+        }
+    }
 
-    // protected abstract Site proposeSite(SiteTypeEnum ste, boolean swapEnable);
+    protected void randomInitSiteChains(List<List<SiteInst>> chains) throws IOException {
+        for (List<SiteInst> chain : chains) {
+            SiteTypeEnum siteType = chain.get(0).getSiteTypeEnum();
+            Site selectedAnchor = proposeAnchorSite(chain, null, false);
+            for (int i = 0; i < chain.size(); i++) {
+                Site newSite = device.getSite(getSiteTypePrefix(siteType) + "X" + selectedAnchor.getInstanceX() +
+                        "Y" + (selectedAnchor.getInstanceY() + i));
+                occupiedSiteChains.get(siteType).put(newSite, chain);
+                placeSiteInst(chain.get(i), newSite);
+            }
+        }
+    } // end randomInitSiteChains()
 
-    // protected abstract Site proposeAnchorSite(SiteTypeEnum ste, int chainSize,
-    // boolean swapEnable);
+    protected Site proposeSite(SiteInst si, List<Site> connections, boolean swapEnable) {
+        return proposeRandomSite(si, connections, swapEnable);
+    }
 
[…]

### 2025-03-13 java/src/main/java/placer/PlacerAnalytical.java

@@ -1,0 +1,64 @@
+
+package placer;
+
+import java.io.FileWriter;
+import java.io.File;
+import java.io.IOException;
+import java.util.Collections;
+import java.util.stream.Collectors;
+import java.util.Collection;
+import java.util.HashMap;
+import java.util.List;
+import java.util.ArrayList;
+import java.util.HashSet;
+import java.util.Random;
+
+import com.xilinx.rapidwright.design.Design;
+import com.xilinx.rapidwright.design.SiteInst;
+import com.xilinx.rapidwright.design.SitePinInst;
+
+import com.xilinx.rapidwright.device.Device;
+import com.xilinx.rapidwright.device.Site;
+import com.xilinx.rapidwright.device.SiteTypeEnum;
+import com.xilinx.rapidwright.device.ClockRegion;
+
+public class PlacerAnalytical extends Placer {
+
+    private String placerName = "PlacerAnalytical";
+
+    public PlacerAnalytical(String rootDir, Design design, Device device, ClockRegion region) throws IOException {
+        super(rootDir, design, device);
+        this.regionConstraint = region;
+    }
+
+    public String getPlacerName() {
+        return this.placerName;
+    }
+
+    public void placeDesign(PackedDesign packedDesign) throws IOException {
+        initSites();
+        initRpmGrid();
+        unplaceAllSiteInsts(packedDesign);
+        randomInitialPlacement(packedDesign);
+        ImageMaker imInitRandom = new ImageMaker(design);
+        imInitRandom.renderAll();
+        imInitRandom.exportImage(graphicsDir + "/random_placement.png");
+        design.writeCheckpoint(rootDir + "/outputs/checkpoints/init_placed.dcp");
+
+        int move = 0;
+        while (move < 100) {
+            move++;
+            //
+            //
+            //
+            //
+        }
+
+        ImageMaker imPlaced = new ImageMaker(design);
+        imPlaced.renderAll();
+        imPlaced.exportImage(graphicsDir + "/final_placement.png");
+        exportCostHistory(printoutDir + "/cost_history.csv");
+        printTimingBenchmarks();
+        writer.write("\n\nTotal move iterations: " + move);
+    }
+}

### 2025-03-13 java/src/main/java/placer/PlacerAnnealRandom.java

@@ -31,6 +31,22 @@ public class PlacerAnnealRandom extends Placer {
         this.regionConstraint = region;
     }
 
+    public String getPlacerName() {
+        return this.placerName;
+    }
+
+    public void initCoolingSchedule(double initialTemp, double alpha, int movesLimit) throws IOException {
+        this.movesLimit = movesLimit;
+        this.writer.write("\nPrinting Cooling Schedule...");
+        // geometric cooling
+        double currentTemp = initialTemp;
+        for (int i = 0; i < movesLimit; i++) {
+            this.coolingSchedule.add(currentTemp);
+            this.writer.write("\n\t" + currentTemp);
+            currentTemp *= alpha;
+        }
+    }
+
     public void placeDesign(PackedDesign packedDesign) throws IOException {
         initSites();
         initRpmGrid();
@@ -79,139 +95,12 @@ public class PlacerAnnealRandom extends Placer {
         writer.write("\n\nTotal move iterations: " + move);
     }
 
-    public String getPlacerName() {
-        return this.placerName;
-    }
-
-    public void initCoolingSchedule(double initialTemp, double alpha, int movesLimit) throws IOException {
-        this.movesLimit = movesLimit;
-        this.writer.write("\nPrinting Cooling Schedule...");
-        // geometric cooling
-        double currentTemp = initialTemp;
-        for (int i = 0; i < movesLimit; i++) {
-            this.coolingSchedule.add(currentTemp);
-            this.writer.write("\n\t" + currentTemp);
-            currentTemp *= alpha;
-        }
-    }
-
-    protected void randomInitialPlacement(PackedDesign packedDesign) throws IOException {
-        randomInitSiteChains(packedDesign.DSPSiteInstCascades);
-        randomInitSiteChains(packedDesign.CARRYSiteInstChains);
-        randomInitSingleSite(packedDesign.RAMSiteInsts);
-        randomInitSingleSite(packedDesign.CLBSiteInsts);
-    }
-
     protected void move(PackedDesign packedDesign) throws IOException {
         moveSiteChains(packedDesign.DSPSiteInstCascades);
         moveSiteChains(packedDesign.CARRYSiteInstChains);
         moveSingleSite(packedDesign.RAMSiteInsts);
         moveSingleSite(packedDesign.CLBSiteInsts);
     }
-
-    protected Site proposeSite(SiteInst si, List<Site> connections, boolean swapEnable) {
-        return proposeRandomSite(si, connections, swapEnable);
-    }
-
[…]

### 2025-03-13 python/plot_convergence.py

@@ -1,4 +1,5 @@
-
+import sys
+import os
 import csv
 import matplotlib.pyplot as plt
 
@@ -18,7 +19,7 @@ def plot_csv(filename, output_file):
 
     plt.figure(figsize=(8, 4))
     plt.plot(iters, costs, marker=None, linestyle='-')
-    plt.title("HPWL Cost vs Iteration")
+    plt.title(placer_name + ": HPWL Cost vs Iteration")
     plt.xlabel("Iter")
     plt.ylabel("HPWL Cost")
     plt.grid(True)
@@ -26,5 +27,11 @@ def plot_csv(filename, output_file):
     plt.savefig(output_file, dpi=200, bbox_inches='tight')
 
 if __name__ == "__main__":
-    plot_csv("outputs/placers/PlacerAnnealHybrid/printout/cost_history.csv", "outputs/placers/PlacerAnnealHybrid/graphics/cost_history.png")
+    placer_path = sys.argv[1]
+    placer_name = os.path.basename(placer_path)
+    csv_file =  placer_path + "/printout/cost_history.csv"
+    png_file =  placer_path + "/graphics/" + placer_name + "_cost_history.png"
+    plot_csv(csv_file, png_file)
+    print(placer_path + " plot_convergence finished")
+    # plot_csv("outputs/placers/PlacerAnnealHybrid/printout/cost_history.csv", "outputs/placers/PlacerAnnealHybrid/graphics/cost_history.png")
     # plot_csv("outputs/placers/PlacerAnnealRandom/printout/cost_history.csv", "outputs/placers/PlacerAnnealRandom/graphics/cost_history.png")

### 2025-03-13 run.sh

@@ -163,18 +163,48 @@ cd $PROJ_DIR
 # Render placement video
 if [ "$start_stage" == "video" ] || [ "$start_stage" == "all" ]; then
     cd $PROJ_DIR
-    python3 python/plot_convergence.py
     FPS=10
     # INPUT="outputs/placers/PlacerAnnealRandom/graphics/images"
-    INPUT="outputs/placers/PlacerAnnealHybrid/graphics/images"
+    PLACERS_DIR="outputs/placers"
     INPUT_PATTERN="%08d.png"
-    cd $INPUT
+
+    # Loop over all items in PLACERS_DIR
+    for placer_dir in "$PLACERS_DIR"/*; do
+        python3 python/plot_convergence.py "$placer_dir"
+        # Only proceed if it's a directory
+        if [ -d "$placer_dir" ]; then
+            images_path="$placer_dir/graphics/images"
+
+            # Check if the images directory exists and has PNG files
+            if [ -d "$images_path" ] && ls "$images_path"/*.png &>/dev/null; then
+                echo "Generating GIF for: $placer_dir"
+
+                # Go to images directory
+                cd "$images_path" || continue
+
+                # Run ffmpeg to create the GIF
+                ffmpeg -y \
+                    -framerate "$FPS" \
+                    -i "$INPUT_PATTERN" \
+                    -vf "pad=ceil(iw/2)*2:ceil(ih/2)*2" \
+                    -loop 0 \
+                    ../output.gif
+
+                # Return to previous directory
+                cd - &>/dev/null
+            else
+                echo "Skipping $placer_dir (no 'graphics/images' folder or no PNG files found)."
+            fi
+        fi
+    done
+
+    # cd $INPUT
 
     # OUTPUT="../video.mp4"
     # ffmpeg -y -framerate $FPS -i $INPUT_PATTERN -vcodec libx264 "$OUTPUT" -vf "pad=ceil(iw/2)*2:ceil(ih/2)*2"
 
-    OUTPUT="../output.gif"
-    ffmpeg -y -framerate $FPS -i $INPUT_PATTERN -vf "pad=ceil(iw/2)*2:ceil(ih/2)*2" -loop 0 "$OUTPUT"
+    # OUTPUT="../output.gif"
+    # ffmpeg -y -framerate $FPS -i $INPUT_PATTERN -vf "pad=ceil(iw/2)*2:ceil(ih/2)*2" -loop 0 "$OUTPUT"
 
[…]

### 2025-03-13 java/src/main/java/placer/Main.java
add EJML api for linalg

@@ -24,6 +24,9 @@ import com.xilinx.rapidwright.device.ClockRegion;
 import com.xilinx.rapidwright.device.Tile;
 import com.xilinx.rapidwright.device.TileTypeEnum;
 
+import org.ejml.data.DMatrixRMaj;
+import org.ejml.dense.row.CommonOps_DDRM;
+
 public class Main {
 
     private static final Logger logger = Logger.getLogger(Main.class.getName());
@@ -110,6 +113,50 @@ public class Main {
             }
         }
         System.out.println("Design Cost: " + cost);
+    }
+
+    public static void helloEJML() {
+        // 1. Create two 2x2 matrices A and B
+        DMatrixRMaj A = new DMatrixRMaj(new double[][] {
+                { 1, 2 },
+                { 3, 4 }
+        });
+        DMatrixRMaj B = new DMatrixRMaj(new double[][] {
+                { 5, 6 },
+                { 7, 8 }
+        });
+
+        // 2. Matrix addition C = A + B
+        DMatrixRMaj C = new DMatrixRMaj(A.numRows, A.numCols);
+        CommonOps_DDRM.add(A, B, C);
+
+        // 3. Matrix multiplication D = A * B
+        DMatrixRMaj D = new DMatrixRMaj(A.numRows, B.numCols);
+        CommonOps_DDRM.mult(A, B, D);
+
+        // 4. Matrix inversion invA = A^-1
+        DMatrixRMaj invA = new DMatrixRMaj(A.numRows, A.numCols);
+        CommonOps_DDRM.invert(A, invA);
+
+        // 5. Solve Ax = b (for x)
+        // Here, b is a 2x1 column vector
+        DMatrixRMaj b = new DMatrixRMaj(new double[][] { { 1 }, { 2 } });
+        DMatrixRMaj x = new DMatrixRMaj(A.numRows, b.numCols);
+        CommonOps_DDRM.solve(A, b, x);
+
+        // Print results
+        System.out.println("Matrix A:");
+        A.print();
+        System.out.println("Matrix B:");
+        B.print();
+        System.out.println("C = A + B:");
+        C.print();
+        System.out.println("D = A * B:");
+        D.print();
+        System.out.println("invA = A^-1:");
+        invA.print();
+        System.out.println("Solution x for Ax = b:");
+        x.print();
     }
 
 }

### 2025-03-13 run.sh

@@ -14,6 +14,10 @@ export PATH="$PATH:$RAPIDWRIGHT_PATH/bin"
 export CLASSPATH=$RAPIDWRIGHT_PATH/bin:$RAPIDWRIGHT_PATH/jars/*
 export _JAVA_OPTIONS=-Xmx32736m
 
+# EJML
+export EJML_PATH=/home/bcheng/workspace/tools/ejml-v0.42-libs
+export CLASSPATH=$CLASSPATH:$EJML_PATH/*
+
 PROJ_DIR="/home/bcheng/workspace/dev/place-and-route"
 
 SYNTH_TCL="$PROJ_DIR/tcl/synth.tcl"

### 2025-03-24 latex/paper/paper.tex
work on paper

@@
 
-\section{Design Flow}
-A typical FPGA design flow goes like the following. First, the developer writes the design entry in a high level Hardware Description Language (HDL) like Verilog or VHDL.
-\begin{itemize}
-\item Synthesis (Vivado default)
-\item Placement (the focus of this paper)
-    \begin{itemize}
-        \item \textbf{Prepacking}:\\
-        Traverse the raw EDIFNetlist from synthesis and identify patterns of cells that should be grouped together.
-        This can be CARRY chains, DSP cascades, LUTFF pairs, etc.
-    \item \textbf{Packing}:\\
-        Take the prepacked cell groups and pack them into SiteInsts.
-        Intra\-Site routing.
-    \item \textbf{Placement}:\\
-        Take the SiteInsts and optimally place them onto physical Sites.
-    \end{itemize}
-\item Routing
-    Inter\-Site routing
-\end{itemize}
+\section{Introduction}
+
+    Field-Programmable Gate Arrays (FPGAs) have witnessed unprecedented growth in capacity and versatility, driving significant advances in computer-aided design (CAD) and electronic design automation (EDA) methodologies.
+    Since the early-to-mid 2000s, the stagnation of single-processor performance relative to the rapid increase in integrated circuit sizes has led to a design productivity gap, where the computational effort for designing complex chips continues to rise.
+    In FPGA CAD flows—which traditionally encompass logic synthesis, placement, routing, and bitstream generation—the placement stage has emerged as one of the most time-consuming processes. 
+    Inefficiencies in placement not only extend design times from hours to days, thereby elevating cost and reducing engineering productivity, but also limit the broader adoption of FPGAs by software engineers who expect compile times akin to those of conventional software compilers.
+
+    In this paper, we recreate and experiment with established placement algorithms by leveraging the Xilinx's RapidWright, which is a semi-open-source API that offers backend access to Xilinx's industry-standard FPGA environment, Vivado.
+We implement a simulated annealing placer for Xilinx's 7-series FPGAs, with an emphasis on minimizing total wirelength while mitigating runtime.
+Our implementation is organized into three consecutive substages.
+The \textbf{prepacking} stage involves traversing a raw EDIF netlist to identify recurring cell patterns—such as CARRY chains, DSP cascades, and LUT-FF pairs—that are critical for efficient mapping and legalization.
+In the subsequent \textbf{packing} stage, these identified patterns, along with any remaining loose cells, are consolidated into SiteInst objects that encapsulate the FPGA’s discrete resource constraints and architectural nuances.
+Finally, the \textbf{placement} stage employs a simulated annealing (SA) algorithm to optimally assign SiteInst objects to physical sites, aiming to minimize total wirelength while adhering to the constraints of the 7-series architecture.
+
+    Simulated annealing iteratively swaps placement objects, in our case, SiteInsts, guided by a cost function that decides which swaps should be accepted or rejected.
+Hill climbing is permitted by occasionally accepting moves that increase cost, in hope that such swaps may later lead to a better final solution.
+SA remains a popular approach in FPGA placement research due to its simplicity and robustness in handling the discrete architectural constraints of FPGA devices.
+While SA yields surprisingly good results given relatively simple rules, it is ultimately a heuristic and stochastic approach that explores the vast placement space by making random moves.
+Most of these moves will be rejected, meaning that SA must run many iterations, usually hundreds to thousands, to arrive at a desirable solution.
+
+    In the ASIC domain, where placers must handle designs with millions of cells, the SA approach has largely been abandoned in favor of analytical techniques, owing to SA's runtime and poor scalability.
+Modern FPGA placers have also followed suit, as new legalization strategies allow FPGA placers to leverage traditionally ASIC placement algorithms and adapt them to the discrete constraints of FPGA architectures.
+While this paper does not present a working analytical placer, it will explore ways to build upon our existing infrastructure (prepacker and packer) to replace SA with AP.
+
+\section{Xilinx 7-Series Architecture}
+    Before any work can be done on a placer, one must consider what is being placed and upon what it is being placed. \\
+    (Square peg square hole analogy) \\
+    (Show the arch hierarchy). \\
+
+    how to organize this? \\
+
+    Device hierarchical objects: \\
+        Super Logic Region, Clock Region, Tile, Site, BEL \\
+
+    Device connection objects: \\
+        Switchbox, PIP, BELPin, SitePin \\
+
+    Device object instances: \\
+        SitePinInst, SiteInst, CellPin
+
+    EDIF objects: \\
+        EDIFNet, EDIFCell, EDIFPort \\
+        EDIFHierNet, EDIFHierCell, EDIFHierPort \\
+        EDIFHierCellInst, EDIFHierPortInst \\
+
+\section{RapidWright API}
+    RapidWright represents the architecture constructs as faithfully as possible
+
+    The RapidWright API offers a user-friendly toolkit to take pre-implemented designs, that is, designs that have already been placed and/or routed by Vivado, and make custom optimizations to fit complex design criteria.
+    RapidWright has several "BlockPlacers", a general Router, but no general Placer. 
+    We use this API to implement a crude general placer, in our case, a simulated annealer. 
+    (Show what parts of the FPGA toolchain that RapidWright already has and how our placer fits into that toolchain).
+
+\section{Simulated Annealing}
+    (Describe the algorithm, then in context of 7-Series arhictecture). \\
+
+\section{Analytical Placement}
+    Talk about HeAP.
 

### 2025-03-24 java/src/main/java/placer/PlacerAnalytical.java

@@ -50,10 +50,10 @@ public class PlacerAnalytical extends Placer {
         int move = 0;
         while (move < 100) {
             move++;
-            //
-            //
-            //
-            //
+            // 1) create weight matrix
+            // 2) solve linear system
+            // 3) legalize
+            // 4)
         }
 
         ImageMaker imPlaced = new ImageMaker(design);

### 2025-03-24 run.sh

@@ -152,7 +152,7 @@ fi
 
 # Gradle Run Stage (replaces Java Placement)
 if [ "$start_stage" == "place" ] || [ "$start_stage" == "all" ]; then
-    rm $PROJ_DIR/outputs/results/* -r
+    rm $PROJ_DIR/outputs/placers/* -r
     echo "Running Java placement with Gradle..."
     cd $PROJ_DIR/java
     gradle run

### 2025-03-25 latex/paper/paper.tex
git rm ignored latex files
talk about AP

@@
     Device object instances: \\
-        SitePinInst, SiteInst, CellPin
+        SitePinInst, SiteInst, CellPin, Cell
 
@@
 \section{Analytical Placement}
-    Talk about HeAP.
+    Introduce placement with legalization methods. \\
+    Many AP approaches use a Global placement, Legalization, then Detailed placement approach. \\
+    Our SA only considers legal moves via bookkeeping structures, so it has no concept of global placement or legalization. \\
+    It is a detailed placement only approach. \\
+    \begin{itemize}
+        \item Talk about HeAP as an example.
+        \item Global Placement, Legalization, then Detailed Placement approach.
+        \item Global placement using the Bound2Bound net model from SimPL.
+        \item Legalization Step 1: \\
+            Find an area of the device that is overutilized (illegal) for twhich the blocks contained within must spread to a larger area. \\
+            To obtain this overutilized area, adjacent locations (Sites, Tiles, etc.) on the device that are occupied by more than one block (SiteInst, ModuleInst, etc.) are repeatedly clustered together, until all clusters are bordered on all sides by non-overutilized locations.
+\\
+            Then, the area (?) is expanded in both the x and y dimensions (?) until it is large enough to accommodate all blocks contained within. \\
+            Specifically, the area is expanded until its "occupance" \(O_A\) divided by its "capacity" \(C_A\) is less than a maximum "utilization factor" $\beta$, where $\beta$ is less than or equal to 1.
+HeAP uses 0.9.
+        \item Legalization Step 2: \\
+            Two cuts are generated: a "source cut" and a "target cut". \\
+            The source cut pertains to the blocks being placed (SiteInsts in home buffer). \\ 
+            The target cut pertains to the area into which the blocks are placed (away buffer). \\
+            The source cut splits the blocks into 2 partitions, while the target cut splits the area into 2 sub-areas, into which the blocks in each partition are spread. \\
+            Two objectives are minimized during this process: The imbalance between the number of blocks in each partition, and the difference in the utilization of each sub-area. \\
+            Utilization defined as the occupancy divided by the capacity of the sub-area. \\
+            \(U_{sub-area} = \frac{O_{sub-area}}{C_{sub-area}}\) \\
+            To generate the source cut, the cells are first sorted by their x or y location, depending on the orientation of the desired cut. \\
+            Once the cells are sorted, the source cut generation is akin to choosing a pivot in a sorted list, where all blocks to the left of the pivot are assigned to the left/bottom partition and all blocks to the right of the pivot are assigned to the right/top partition.
+\\
+            The target cut is an x or y cut of the area such that all blocks in each partition fit in their respective sub-areas, and such that \( | U_{sub-area_1} 
+             - U_{sub-area_2} | \) is minimized.
+
+    \end{itemize}
 

### 2025-03-26 latex/paper/paper.tex
more outlining

@@
 
-    In this paper, we recreate and experiment with established placement algorithms by leveraging the Xilinx's RapidWright, which is a semi-open-source API that offers backend access to Xilinx's industry-standard FPGA environment, Vivado.
-We implement a simulated annealing placer for Xilinx's 7-series FPGAs, with an emphasis on minimizing total wirelength while mitigating runtime.
+    For these reasons, FPGA placement remains a critical research effort even today. 
+    In this paper, we study and implement established placement methods by leveraging RapidWright, which is a semi-open-source API that offers backend access to Xilinx's industry-standard FPGA environment, Vivado.
+    We implement multiple variations of simulated annealing placers for Xilinx's 7-series FPGAs, with an emphasis on minimizing total wirelength while mitigating runtime. 
     Our implementation is organized into three consecutive substages. 
@@
 
-    Simulated annealing iteratively swaps placement objects, in our case, SiteInsts, guided by a cost function that decides which swaps should be accepted or rejected.
+    Simulated annealing iteratively swaps placement objects guided by a cost function that decides which swaps should be accepted or rejected. 
     Hill climbing is permitted by occasionally accepting moves that increase cost, in hope that such swaps may later lead to a better final solution. 
@@
 
-\section{Xilinx 7-Series Architecture}
-    Before any work can be done on a placer, one must consider what is being placed and upon what it is being placed. \\
+\end{multicols}
+
+\section{The FPGA Toolchain}
+    Engineer design entry (Verilog, VHDL) \\
+    Synthesis \\ 
+    Placement \\ 
+    Routing \\ 
+    Bitstream generation \\
+
     (Square peg square hole analogy) \\
-    (Show the arch hierarchy). \\
+    (Show the architecture hierarchy). \\
 
@@
 
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{figures/hierarchy.png}
+    \caption{Architecture Hierarchy of a Xilinx FPGA}\label{fig:hierarchy}
+\end{figure}
+
+\begin{multicols}{2}
+\section{Xilinx 7-Series Architecture}
+    Before any work can be done on a placer, one must understand the objects being placed and the structure of the medium in which those objects reside. 
+    Figure~\ref{fig:hierarchy} illustrates the architectural hierarchy of a Xilinx FPGA. 
+    At its most basic level, an FPGA is a vast physical array of replicated atomic components known as Basic Elements of Logic (BELs).
+    These BELs, which include programmable lookup tables (LUTs), flip flops (FFs), and programmable interconnects, form the fundamental building blocks for custom digital circuit implementations on FPGAs.
+
+    In modern FPGAs like the Xilinx 7-Series, additional BELs like Block RAM (BRAM) and Digital Processing Slices (DSPs) are also available, creating a more heterogeneous programmable logic fabric. 
+    These dedicated resources allow designers to implement common digital structures, such as multipliers and memories, without having to recreate them from scratch using only LUTs and flip flops.
+
+    The Xilinx architectural hierarchy organizes these BELs into increasingly abstract structures. 
+    \textbf{BELs} are packaged into \textbf{Sites}, which are then embedded into \textbf{Tiles}, which are subsequently consolidated into \textbf{Clock Regions}. 
+    In high-end Xilinx devices, Clock Regions may be further consolidated into \textbf{Super Logic Regions} (SLRs). 
+    However, for the scope of this paper, we focus on Xilinx devices that have only a single SLR.
+
+    Take for example, the xc7z020 chip, a low-end Xilinx device from the 7-Series family of FPGAs, commonly found on hobbyist boards like the Digilent Arty-Z7.
+
 \section{RapidWright API}
@@
     (Show what parts of the FPGA toolchain that RapidWright already has and how our placer fits into that toolchain). 
+    \subsection{Device Objects}
+    \subsection{Design Objects}
 
-\section{Simulated Annealing}
+\section{Placement Implementation}
     (Describe the algorithm, then in context of 7-Series arhictecture). \\
+    \subsection{PrePacking}
+        In the prepacking stage, we traverse the raw EDIF Netlist to identify recurring cell patterns. In particular, we will specifically look for CARRY chains, DSP cascades, and LUT-FF pairs.
+        \subsection{CARRY Chains}
+        \subsection{DSP Cascades}
+        \subsection{LUT-FF Pairs}
+            Identify unique CE-SR net pairs. All FF cells in a Site must share the same CE and SR nets.
+    \subsection{Packing}
+        In the packing stage, we take the identified cell clusters and pack them into SiteInst Design objects which target the Device Site objects.
+        \subsection{CLB Sites}
+            Can support LUT-FF pairs, loose LUTs, loose FFs, CARRY chains.
+            CARRY chains span vertically across multiple Sites.
+    \subsection{Placement}
 

### 2025-03-27 latex/paper/paper.tex
mooooore wooooork

@@
-\documentclass{article}
+\documentclass[twocolumn]{article}
 \usepackage{hyperref}
@@
 \usepackage{mathpazo}
-\usepackage{multicol}
+% \usepackage{multicol}
 \usepackage[a4paper,
@@
 
-\begin{multicols}{2}
+% \begin{multicols}{2}
 \section{Abstract}
@@
 
-\end{multicols}
+% \end{multicols}
 
-\section{The FPGA Toolchain}
+\section{The FPGA Design Flow and Toolchain}
+    \begin{figure}
+        \centering
+        \includegraphics[width=6.5cm]{figures/vivado_dcps.png}
+        \caption{Showing how RapidWright can take a design checkpoint at different stages of the Vivado toolchain, make customizations, then feed it back into the toolchain.}
+        \label{fig:device_carry_chain_routing}
+    \end{figure}
+    A typical FPGA design flow goes like the following:
     Engineer design entry (Verilog, VHDL) \\
-    Synthesis \\ 
-    Placement \\ 
+    Synthesis (Instantiation, substitution, inference, and optimization) \\ 
+    Placement (Prepacking, Packing, Global Placement, Legalization, Detailed Placement)\\ 
     Routing \\ 
@@
 
-    (Square peg square hole analogy) \\
-    (Show the architecture hierarchy). \\
-
-    how to organize this? \\
-
-    Device hierarchical objects: \\
-        Super Logic Region, Clock Region, Tile, Site, BEL \\
-
-    Device connection objects: \\
-        Switchbox, PIP, BELPin, SitePin \\
-
-    Device object instances: \\
-        SitePinInst, SiteInst, CellPin, Cell
-
-    EDIF objects: \\
-        EDIFNet, EDIFCell, EDIFPort \\
-        EDIFHierNet, EDIFHierCell, EDIFHierPort \\
-        EDIFHierCellInst, EDIFHierPortInst \\
-
-\begin{figure}
+\begin{figure*}
     \centering
@@
     \caption{Architecture Hierarchy of a Xilinx FPGA}\label{fig:hierarchy}
-\end{figure}
+\end{figure*}
 
-\begin{multicols}{2}
+% \begin{multicols}{2}
 \section{Xilinx 7-Series Architecture}
-    Before any work can be done on a placer, one must understand the objects being placed and the structure of the medium in which those objects reside. 
+    Before any work can be done on a placer, one must understand the objects being placed and the structure of the medium in which those objects are placed. 
     Figure~\ref{fig:hierarchy} illustrates the architectural hierarchy of a Xilinx FPGA. 
@@
 \section{RapidWright API}
-    RapidWright represents the architecture constructs as faithfully as possible
+
+    DO WE REALLY NEED THIS SECTION? JUST REFERENCE THE RW DOCUMENTATION.
 
@@
     (Show what parts of the FPGA toolchain that RapidWright already has and how our placer fits into that toolchain).
+
+    There are three packages of objects that are most relavent to building our placer: the edif package, design package, and device package.
+
     \subsection{Device Objects}
-    \subsection{Design Objects}
+    The \textbf{device} package contains the classes that correspond to constructs in the hardware and/or silicon devices. 
+    The most prominent class, aptly named Device, represents a specific product family member (like the xc7z020 device of the 7-series family) and provides a reference to all of the logical and routing resources available on the device.
+    There device classes can be separated into two groups: logical and routing.
+    Among the logical classes are the BEL, Site, Tile, and ClockRegion objects discussed before.
+    Among the routing classes are BELPin, SitePin, PIP (Programmable InterConnect Point), Wire, and Node.
+
+    \subsection{Design Package}
+    The \textbf{design} package is the collection of objects used to describe how a logical netlist maps to the device netlist. 
+    The design is also referred to as the "physical netlist" or "implementation".
+    It contains all of the primitive logical cell mappings to hardware, specifically the cells to BEL placements and physical net mapping to programmable interconnect or routing.
+
+    \subsection{EDIF Package}
 
@@
         \subsection{CARRY Chains}
+            Each SLICE Site contains one CARRY4 BEL.
+            Each CLB Tile contains two SLICE Sites.
+            CARRY chains span vertically across multiple SLICE Sites/Tiles. (Show picture)
+            \begin{figure}
+                \centering
+                \includegraphics[width=3.0cm]{figures/carry_chain_routes.png}
+                \includegraphics[width=3.0cm]{figures/dsp_cascade_routes.png}
+                \caption{A Carry Chain of length 6 spanning 6 Tiles. On the left, the cell-only representation. On the right, cells with routing representation.}
+                \label{fig:device_carry_chain_routing}
+            \end{figure}
         \subsection{DSP Cascades}
+            Each DSP Site contains one DSP BEL.
+            Each DSP Tile contains two DSP Sites.
+            DSP chains can span vertically across multiple DSP Sites/Tiles.
+
         \subsection{LUT-FF Pairs}
@@
             Can support LUT-FF pairs, loose LUTs, loose FFs, CARRY chains.
-            CARRY chains span vertically across multiple Sites.
+            Each SLICE has 8 "lanes" of LUT-FFs. 4 LUT5s and 4 LUT6s. 8 FFs.
+            For SLICEMs, LUT6s can be configured as shallow 32-bit LUTRAMs or "RAMS32".
     \subsection{Placement}
+        Create BEL "fields": CLB, DSP, CLB and DSP Chains, RAMs
+
+        \subsection{Detailed Placement}
 
@@
 
-\end{multicols}
+% \end{multicols}
 \end{document}

### 2025-03-27 latex/paper/paper.tex
yea

@@
 
-% \begin{multicols}{2}
 \section{Abstract}
@@
 
-% \end{multicols}
+\section{Brief History of FPGA Architecture}
+    \begin{figure}
+        \centering
+        \includegraphics[width=7.0cm]{figures/pal_2.png}
+        \caption{PAL architecture with 5 inputs, 8 programmable AND gates and 4 fixed OR gates}
+        \label{fig:pla}
+        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/Programmable-Logic-Devices/Programmable-Array-Logic-PAL/
+        % https://www.naukri.com/code360/library/difference-between-pla-and-pal 
+    \end{figure}
+
+    Before any work can be done on a placer, one must understand the objects being placed and the structure of the medium in which those objects are placed.
+
+    Configurable logic devices have undergone significant evolution over the last four decades.
+    The journey began with simple devices like Programmable Logic Array (PLA), which came if the form of logic gates arranged in a "programmable-OR, programmable-AND" plane, used to implement a "sum-of-products" logic equation for each of the outputs in terms of the inputs via programmable fuses.
+    Around the same time came the redundantly named Programmable Array Logic (PAL), which simplified the PLA by fixing the OR gates, resulting in a "fixed-OR, programmable-AND" plane, which sacrificed some flexibility to simplify its manufacture and broaden its applications.
+
+    \begin{figure}
+        \centering
+        \includegraphics[width=7.0cm]{figures/cpld.png}
+        \caption{CPLD architecture with 4 CLBs}
+        \label{fig:cpld}
+        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/CPLD/Complex-Programmable-Logic-Device-CPLDs/
+    \end{figure}
+
+    Then came the Complex Programmable Logic Device (CPLD), which came as an array of Configurable Logic Blocks (CLBs) stitched together by interconnection wires.
+    These CLBs usually took the form of modified PAL blocks, which contained the PAL itself and some macrocells including flip-flops, multiplexers, and tri-state buffers.
+    The CPLD was essentially an array of PALs interconnected by a programmable switch matrix and could be programmed via a hardware description language (HDL) like VHDL.
+
+    \begin{figure}
+        \centering
+        \includegraphics[width=7.0cm]{figures/fpga.png}
+        \caption{Island-style FPGA architecture with 16 CLBs}
+        \label{fig:fpga}
+        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/FPGA/Field-programmable-gate-array-FPGA/
+    \end{figure}
+
+\section{Xilinx 7-Series Architecture}
+
+    \begin{figure*}
+        \centering
+        \includegraphics[width=\textwidth]{figures/hierarchy.png}
+        \caption{Architecture Hierarchy of a Xilinx FPGA}
+        \label{fig:hierarchy}
+    \end{figure*}
+
+    Figure~\ref{fig:hierarchy} illustrates the architectural hierarchy of a Xilinx FPGA. 
+    At its most basic level, an FPGA is a vast physical array of replicated atomic components known as Basic Elements of Logic (BELs).
+    These BELs, which include programmable lookup tables (LUTs), flip flops (FFs), and programmable interconnects, form the fundamental building blocks for custom digital circuit implementations on FPGAs.
+
+    In modern FPGAs like the Xilinx 7-Series, additional BELs like Block RAM (BRAM) and Digital Processing Slices (DSPs) are also available, creating a more heterogeneous programmable logic fabric. 
+    These dedicated resources allow designers to implement common digital structures, such as multipliers and memories, without having to recreate them from scratch using only LUTs and flip flops.
+
+    The Xilinx architectural hierarchy organizes these BELs into increasingly abstract structures. 
+    \textbf{BELs} are packaged into \textbf{Sites}, which are then embedded into \textbf{Tiles}, which are subsequently consolidated into \textbf{Clock Regions}. 
+    In high-end Xilinx devices, Clock Regions may be further consolidated into \textbf{Super Logic Regions} (SLRs). 
+    However, for the scope of this paper, we focus on Xilinx devices that have only a single SLR.
+
+    Take for example, the xc7z020 chip, a low-end Xilinx device from the 7-Series family of FPGAs, commonly found on hobbyist boards like the Digilent Arty-Z7.
 
@@
     Bitstream generation \\
-
-\begin{figure*}
-    \centering
-    \includegraphics[width=\textwidth]{figures/hierarchy.png}
-    \caption{Architecture Hierarchy of a Xilinx FPGA}\label{fig:hierarchy}
-\end{figure*}
-
-% \begin{multicols}{2}
-\section{Xilinx 7-Series Architecture}
-    Before any work can be done on a placer, one must understand the objects being placed and the structure of the medium in which those objects are placed. 
-    Figure~\ref{fig:hierarchy} illustrates the architectural hierarchy of a Xilinx FPGA. 
-    At its most basic level, an FPGA is a vast physical array of replicated atomic components known as Basic Elements of Logic (BELs).
-    These BELs, which include programmable lookup tables (LUTs), flip flops (FFs), and programmable interconnects, form the fundamental building blocks for custom digital circuit implementations on FPGAs.
-
-    In modern FPGAs like the Xilinx 7-Series, additional BELs like Block RAM (BRAM) and Digital Processing Slices (DSPs) are also available, creating a more heterogeneous programmable logic fabric. 
-    These dedicated resources allow designers to implement common digital structures, such as multipliers and memories, without having to recreate them from scratch using only LUTs and flip flops.
-
-    The Xilinx architectural hierarchy organizes these BELs into increasingly abstract structures. 
-    \textbf{BELs} are packaged into \textbf{Sites}, which are then embedded into \textbf{Tiles}, which are subsequently consolidated into \textbf{Clock Regions}. 
-    In high-end Xilinx devices, Clock Regions may be further consolidated into \textbf{Super Logic Regions} (SLRs). 
-    However, for the scope of this paper, we focus on Xilinx devices that have only a single SLR.
-
-    Take for example, the xc7z020 chip, a low-end Xilinx device from the 7-Series family of FPGAs, commonly found on hobbyist boards like the Digilent Arty-Z7.
 
@@
 
-% \end{multicols}
 \end{document}

### 2025-03-27 latex/paper/paper.tex
more

@@
     Configurable logic devices have undergone significant evolution over the last four decades.
-    The journey began with simple devices like Programmable Logic Array (PLA), which came if the form of logic gates arranged in a "programmable-OR, programmable-AND" plane, used to implement a "sum-of-products" logic equation for each of the outputs in terms of the inputs via programmable fuses.
+    The journey began with simple devices like Programmable Logic Array (PLA) in the early 1970s, which came if the form of logic gates arranged in a "programmable-OR, programmable-AND" plane, used to implement a "sum-of-products" logic equation for each of the outputs in terms of the inputs via programmable fuses.
     Around the same time came the redundantly named Programmable Array Logic (PAL), which simplified the PLA by fixing the OR gates, resulting in a "fixed-OR, programmable-AND" plane, which sacrificed some flexibility to simplify its manufacture and broaden its applications.
@@
         \includegraphics[width=7.0cm]{figures/cpld.png}
-        \caption{CPLD architecture with 4 CLBs}
+        \caption{CPLD architecture with 4 CLBs (PAL-like blocks)}
         \label{fig:cpld}
@@
 
-    Then came the Complex Programmable Logic Device (CPLD), which came as an array of Configurable Logic Blocks (CLBs) stitched together by interconnection wires.
+    Later in the same decade came the Complex Programmable Logic Device (CPLD), which came as an array of Configurable Logic Blocks (CLBs).
     These CLBs usually took the form of modified PAL blocks, which contained the PAL itself and some macrocells including flip-flops, multiplexers, and tri-state buffers.
@@
     \end{figure}
+
+    The next milestone in the mid-1980s was the advent of homogenous FPGA, which like the CPLD, took the form of a vast array of CLBs. 
+    Instead of having a central programmable switch matrix like that of a CPLD, FPGAs adopted an island style architecture, where each CLB, or "island" of logic, is surrounded on all sides by programmable routing resources, resulting in a fabric of programmable logic and routing.
 
@@
 \section{The FPGA Design Flow and Toolchain}
-    \begin{figure}
-        \centering
-        \includegraphics[width=6.5cm]{figures/vivado_dcps.png}
-        \caption{Showing how RapidWright can take a design checkpoint at different stages of the Vivado toolchain, make customizations, then feed it back into the toolchain.}
-        \label{fig:device_carry_chain_routing}
-    \end{figure}
+
     A typical FPGA design flow goes like the following: 
@@
 \section{RapidWright API}
+
+    \begin{figure}
+        \centering
+        \includegraphics[width=6.5cm]{figures/vivado_dcps.png}
+        \caption{Showing how RapidWright can take a design checkpoint at different stages of the Vivado toolchain, make customizations, then feed it back into the toolchain.}
+        \label{fig:device_carry_chain_routing}
+    \end{figure}
 
@@
     \subsection{EDIF Package}
+    In Vivado, all designs post synthesis have a logical netlist that can be exported in the Electronic Design Interchange Format (EDIF) netlist format.
+    The synthesis stage of the FPGA toolchain produces an EDIF file, which is a low-level file containing the technology-mapped netlist.
+    The EDIFNetlist is the top level class that contains the netlist and cell libraries. 
+    The EDIFNetlist keeps a reference to the top cell which is wrapped in the EDIFDesign class. 
+    It also maintains a top cell instance reference that is generated when the file is loaded.
 

### 2025-03-28 latex/paper/paper.tex
toolchain

--- 2025-03-27 latex/paper/paper.tex
@@
 
-    Field-Programmable Gate Arrays (FPGAs) have witnessed unprecedented growth in capacity and versatility, driving significant advances in computer-aided design (CAD) and electronic design automation (EDA) methodologies.
+    Field-Programmable Gate Arrays (FPGAs) have witnessed rapid growth in capacity and versatility, driving significant advances in computer-aided design (CAD) and electronic design automation (EDA) methodologies.
     Since the early-to-mid 2000s, the stagnation of single-processor performance relative to the rapid increase in integrated circuit sizes has led to a design productivity gap, where the computational effort for designing complex chips continues to rise.
@@
 
-% \end{multicols}
+\section{A Brief History on FPGA Architecture}
 
-\section{The FPGA Design Flow and Toolchain}
     \begin{figure}
         \centering
-        \includegraphics[width=6.5cm]{figures/vivado_dcps.png}
-        \caption{Showing how RapidWright can take a design checkpoint at different stages of the Vivado toolchain, make customizations, then feed it back into the toolchain.}
-        \label{fig:device_carry_chain_routing}
+        \includegraphics[width=8.5cm]{figures/pal_2.png}
+        \caption{PAL architecture with 5 inputs, 8 programmable AND gates and 4 fixed OR gates}
+        \label{fig:pla}
+        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/Programmable-Logic-Devices/Programmable-Array-Logic-PAL/
+        % https://www.naukri.com/code360/library/difference-between-pla-and-pal 
     \end{figure}
-    A typical FPGA design flow goes like the following:
-    Engineer design entry (Verilog, VHDL) \\
-    Synthesis (Instantiation, substitution, inference, and optimization) \\ 
-    Placement (Prepacking, Packing, Global Placement, Legalization, Detailed Placement)\\ 
-    Routing \\ 
-    Bitstream generation \\
+
+    Before any work can begin on a placer, it is necessary to understand both the objects to be placed and the medium in which they are placed.
+
+    Configurable logic devices have undergone significant evolution over the past four decades. 
+    The journey began with the Programmable Logic Array (PLA) in the early 1970s. 
+    The PLA implemented output logic using a programmable-OR and programmable-AND plane that formed a sum-of-products equation for each output through programmable fuses. 
+    Around the same time, the Programmable Array Logic (PAL) was introduced. 
+    The PAL simplified the PLA by fixing the OR gates, resulting in a fixed-OR, programmable-AND design, which sacrificed some logic flexibility to simplify its manufacture.
+
+    \begin{figure}
+        \centering
+        \includegraphics[width=7.5cm]{figures/cpld.png}
+        \caption{CPLD architecture with 4 CLBs (PAL-like blocks)}
+        \label{fig:cpld}
+        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/CPLD/Complex-Programmable-Logic-Device-CPLDs/
+    \end{figure}
+
+    Later in the same decade came the Complex Programmable Logic Device (CPLD), which took the form of an array of Configurable Logic Blocks (CLBs). 
+    These CLBs were typically modified PAL blocks that included the PAL itself along with macrocells such as flip-flops, multiplexers, and tri-state buffers. 
+    The CPLD functioned as an array of PALs connected by a programmable switch matrix and could be programmed using a hardware description language (HDL) like VHDL.
+
+    \begin{figure}
+        \centering
+        \includegraphics[width=7.0cm]{figures/fpga.png}
+        \caption{Island-style FPGA architecture with 16 CLBs in a grid}
+        \label{fig:fpga}
+        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/FPGA/Field-programmable-gate-array-FPGA/
+    \end{figure}
+
+    The mid-1980s saw the introduction of the homogeneous FPGA, which was built as a grid of CLBs. 
+    Rather than using a central programmable switch matrix as in CPLDs, FPGAs adopted an island style architecture in which each CLB is surrounded on all sides by programmable routing resources. 
+    The first commercially viable FPGA, produced by Xilinx in 1984, featured 16 CLBs arranged in a 4x4 grid. 
+    As FPGA technology advanced, CLBs were redesigned to use lookup tables (LUTs) as programmable memories instead of PAL arrays. 
+    The capacity of an FPGA was often measured by how many logical elements or CLBs it offered, which grew from hundreds to thousands and now to hundreds of thousands of CLBs.
+
+    This brings us to modern day FPGA architectures. 
+    To meet increasing design complexity and performance needs, FPGA vendors introduced heterogeneous FPGAs. 
+    In addition to CLBs, heterogenous FPGAs embedded hard macros like Block RAM (BRAM) and Digital Signal Processing (DSP) slices into the programmable logic fabric, allowing for greater flexibility and scalability for complex designs.
+    This approach allows users to instantiate common subsystems like memories and multipliers directly, without having to recreate them from scratch using CLBs. 
+    Major vendors such as Xilinx and Altera now use heterogeneous island style architectures in their FPGA products 
+    FPGAs continue to become more heterogenous as they offer a wider variety of hard macros within the programmable logic fabric, with the latest now offering embedded CPUs and networking macros in the fabric.
+
+\section{Xilinx 7-Series Architecture}
 
@@
         \includegraphics[width=\textwidth]{figures/hierarchy.png}
-    \caption{Architecture Hierarchy of a Xilinx FPGA}\label{fig:hierarchy}
+        \caption{Architecture Hierarchy of a Xilinx FPGA (Ultrascale)}
+        \label{fig:hierarchy}
     \end{figure*}
 
-% \begin{multicols}{2}
-\section{Xilinx 7-Series Architecture}
-    Before any work can be done on a placer, one must understand the objects being placed and the structure of the medium in which those objects are placed. 
-    Figure~\ref{fig:hierarchy} illustrates the architectural hierarchy of a Xilinx FPGA. 
+    Xilinx introduced the 7-Series architecture in 2010 as an example of a heterogeneous island-style architecture. 
+    Although the latest generation of Xilinx FPGAs is based on the Ultrascale architecture, first released in 2013, this paper focuses on the former 7-Series devices due to their greater accessibility for new users and compatibility with open source tooling.
+
+    Figure~\ref{fig:hierarchy} illustrates the architectural hierarchy of an Ultrascale Xilinx FPGA. 
+    The architecture hierarchy is the same for the 7-Series devices. 
     At its most basic level, an FPGA is a vast physical array of replicated atomic components known as Basic Elements of Logic (BELs). 
@@
 
-    In modern FPGAs like the Xilinx 7-Series, additional BELs like Block RAM (BRAM) and Digital Processing Slices (DSPs) are also available, creating a more heterogeneous programmable logic fabric. 
+    In the FPGAs 7-Series FPGAs, additional BELs like Block RAM (BRAM) and Digital Processing Slices (DSPs) are also available, creating a more heterogeneous programmable logic fabric. 
     These dedicated resources allow designers to implement common digital structures, such as multipliers and memories, without having to recreate them from scratch using only LUTs and flip flops.
@@
 
+\section{The FPGA Design Flow and Toolchain}
+    The FPGA design flow converts a high-level digital design written in an HDL into a configuration that programs the physical FPGA.
+
+    The process begins with design entry, where an engineer describes the functionality of their circuit using a hardware description language (HDL) such as Verilog or VHDL.
+This description abstracts away details of the physical hardware while defining the logic and behavior of the design.
+
+    Next, synthesis translates the HDL code into a netlist that represents the design in terms of basic logic elements, such as lookup tables (LUTs) and flip flops. 
+    During this stage, techniques like instantiation, substitution, inference, optimization, and technology mapping are applied to map the user design to a particular FPGA architecture.
+
+    After synthesis, the placement phase assigns each logic element in the netlist a specific location within the FPGA’s fabric. 
+    This stage includes prepacking, packing, global placement, legalization, and detailed placement. 
+    The objective is to position the elements in a way that minimizes delay, satisfies design constraints, and simplify routing.
+
+    Following placement, the routing phase connects the placed elements by assigning physical wiring resources to nets. 
+    The router determines their paths through the programmable interconnect network while attempting to meet timing and performance constraints.
+
+    The final stage is the bitstream generation, where a vendor-specific tool produces a configuration file that programs the physical FPGA device. 
+    The bitstream specifies the state of every configurable element—logic blocks, routing switches, and hard macros—ensuring the FPGA implements the user's design.
+
+    Throughout the flow, additional tools, such as constraint managers and timing analyzers, are used to enforce design rules and verify that the implementation meets required performance specifications.
+
+    Parallel to this entire design flow are simulators for which the user can write testbenches to verify their design at different stages of implementation.
+    Xilinx's Vivado, for example, provides functional simulation, post-implementation simulation, and post-implementation timing simulation.
+
+    This integrated toolchain, whether provided by FPGA vendors or open source communities, facilitates the transformation from an abstract HDL description to a functioning hardware configuration.
+
 \section{RapidWright API}
+    \begin{figure}
+        \centering
+        \includegraphics[width=6.5cm]{figures/vivado_dcps.png}
+        \caption{Showing how RapidWright can take a design checkpoint at different stages of the Vivado toolchain, make customizations, then feed it back into the toolchain.}
+        \label{fig:device_carry_chain_routing}
+    \end{figure}
 

### 2025-03-28 latex/paper/paper.tex
more design flow

@@
         \includegraphics[width=7.0cm]{figures/fpga.png}
-        \caption{Island-style FPGA architecture with 16 CLBs in a grid}
+        \caption{Homogenous island-style FPGA architecture with 16 CLBs in a grid}
         \label{fig:fpga}
@@
     The first commercially viable FPGA, produced by Xilinx in 1984, featured 16 CLBs arranged in a 4x4 grid. 
-    As FPGA technology advanced, CLBs were redesigned to use lookup tables (LUTs) as programmable memories instead of PAL arrays. 
+    As FPGA technology advanced, CLBs were redesigned to use lookup tables (LUTs) instead of PAL arrays for greater logic density. 
     The capacity of an FPGA was often measured by how many logical elements or CLBs it offered, which grew from hundreds to thousands and now to hundreds of thousands of CLBs.
@@
     This brings us to modern day FPGA architectures.
-    To meet increasing design complexity and performance needs, FPGA vendors introduced heterogeneous FPGAs. 
-    In addition to CLBs, heterogenous FPGAs embedded hard macros like Block RAM (BRAM) and Digital Signal Processing (DSP) slices into the programmable logic fabric, allowing for greater flexibility and scalability for complex designs.
-    This approach allows users to instantiate common subsystems like memories and multipliers directly, without having to recreate them from scratch using CLBs. 
-    Major vendors such as Xilinx and Altera now use heterogeneous island style architectures in their FPGA products 
-    FPGAs continue to become more heterogenous as they offer a wider variety of hard macros within the programmable logic fabric, with the latest now offering embedded CPUs and networking macros in the fabric.
+    To meet the needs of increasingly complex designs, FPGA vendors introduced heterogeneous FPGAs. 
+    In these devices, hard macros such as Block RAM (BRAM) and Digital Signal Processing (DSP) slices are integrated into the programmable logic fabric along with CLBs. 
+    This design enables the direct instantiation of common subsystems like memories and multipliers, without having to implement them solely with CLBs. 
+    Major vendors such as Xilinx and Altera now employ heterogeneous island-style architectures in their devices. 
+    As designs become increasingly large and complex, FPGAs become increasingly heterogenous as they incorporate a wider variety of hard macros, including embedded CPUs and communications blocks, into the fabric.
 
@@
 
-    Xilinx introduced the 7-Series architecture in 2010 as an example of a heterogeneous island-style architecture. 
-    Although the latest generation of Xilinx FPGAs is based on the Ultrascale architecture, first released in 2013, this paper focuses on the former 7-Series devices due to their greater accessibility for new users and compatibility with open source tooling.
+    Xilinx introduced the 7-Series architecture in 2010 which followed the heterogeneous island-style architecture. 
+    Then in 2013 Xilinx unveiled the Ultrascale architecture which made several improvements on the 7-Series architecture.
+    Although the cutting edge of FPGAs follow the Ultrascale architecture, this paper focuses on the former 7-Series devices due to their greater accessibility for beginner users and compatibility with open source tooling.
 
@@
 
-    Take for example, the xc7z020 chip, a low-end Xilinx device from the 7-Series family of FPGAs, commonly found on hobbyist boards like the Digilent Arty-Z7.
+\section{The FPGA Design Flow and Toolchain}
 
-\section{The FPGA Design Flow and Toolchain}
-    The FPGA design flow converts a high-level digital design written in an HDL into a configuration that programs the physical FPGA.
+    \begin{figure}
+        \centering
+        \includegraphics[width=7.0cm]{figures/design_flow.png}
+        \caption{A typical FPGA design and verification flow}
+        \label{fig:design_flow}
+    \end{figure}
 
-    The process begins with design entry, where an engineer describes the functionality of their circuit using a hardware description language (HDL) such as Verilog or VHDL.
+    Figure \ref{fig:design_flow} shows a typical FPGA design and verification flow.
+    The FPGA design flow takes a high-level digital design written in an HDL and generates a configuration file that programs the physical FPGA.
+
+    The process begins with \textbf{design entry}, where an engineer describes the functionality of their circuit using a hardware description language (HDL) such as Verilog or VHDL.
 This description abstracts away details of the physical hardware while defining the logic and behavior of the design.
 
-    Next, synthesis translates the HDL code into a netlist that represents the design in terms of basic logic elements, such as lookup tables (LUTs) and flip flops. 
-    During this stage, techniques like instantiation, substitution, inference, optimization, and technology mapping are applied to map the user design to a particular FPGA architecture.
+    Next, the \textbf{synthesis} tool interprets the HDL code and generates a netlist that represents the design in terms of basic logic elements, such as lookup tables (LUTs) and flip flops. 
+    During this stage, substages like instantiation, substitution, inference, optimization, and technology mapping are applied to map the user design to a particular FPGA architecture.
 
-    After synthesis, the placement phase assigns each logic element in the netlist a specific location within the FPGA’s fabric. 
-    This stage includes prepacking, packing, global placement, legalization, and detailed placement. 
-    The objective is to position the elements in a way that minimizes delay, satisfies design constraints, and simplify routing.
+    After synthesis, the \textbf{placement} tool assigns each logic element in the netlist a specific location within the FPGA’s fabric. 
+    This stage can include substages of prepacking, packing, global placement, legalization, and detailed placement. 
+    The objective is to place the logical cell instances of the netlist onto physical BELs on the device in a way that minimizes delay, satisfies design constraints, and simplifies routing.
 
-    Following placement, the routing phase connects the placed elements by assigning physical wiring resources to nets. 
+    Following placement, the \textbf{routing} tool connects the placed elements by assigning physical wiring resources to the nets in the netlist. 
     The router determines their paths through the programmable interconnect network while attempting to meet timing and performance constraints.
 
-    The final stage is the bitstream generation, where a vendor-specific tool produces a configuration file that programs the physical FPGA device. 
+    Placement and routing are generally referred to together as the \textbf{implementation} stage, as both stages are interdependent and must adhere to the constraints of the same device.
+    State of the art (SOTA) tools will perform routing-aware placement for greater optimization.
+
+    The final stage is the \textbf{bitstream} generation, where a vendor-specific programming tool produces a configuration file or bitstream that programs the physical FPGA device. 
     The bitstream specifies the state of every configurable element—logic blocks, routing switches, and hard macros—ensuring the FPGA implements the user's design.
@@
     Among the logical classes are the BEL, Site, Tile, and ClockRegion objects discussed before. 
-    Among the routing classes are BELPin, SitePin, PIP (Programmable InterConnect Point), Wire, and Node.
+    Among the routing classes are BELPin, SitePin, PIP (programmable interconnect point), Wire, and Node.
 

### 2025-03-29 latex/paper/paper.tex
more paper work

@@
 
+    Here we will briefly outline the evolution of configurable logic architecture as well as the FPGA design flow. 
+    However, readers who are already familiar with basic FPGA architecture and design flow may skip to the next section on Xilinx 7-Series Architecture. 
     Configurable logic devices have undergone significant evolution over the past four decades. 
@@
     Major vendors such as Xilinx and Altera now employ heterogeneous island-style architectures in their devices. 
-    As designs become increasingly large and complex, FPGAs become increasingly heterogenous as they incorporate a wider variety of hard macros, including embedded CPUs and communications blocks, into the fabric.
-
-\section{Xilinx 7-Series Architecture}
-
-    \begin{figure*}
-        \centering
-        \includegraphics[width=\textwidth]{figures/hierarchy.png}
-        \caption{Architecture Hierarchy of a Xilinx FPGA (Ultrascale)}
-        \label{fig:hierarchy}
-    \end{figure*}
-
-    Xilinx introduced the 7-Series architecture in 2010 which followed the heterogeneous island-style architecture. 
-    Then in 2013 Xilinx unveiled the Ultrascale architecture which made several improvements on the 7-Series architecture.
-    Although the cutting edge of FPGAs follow the Ultrascale architecture, this paper focuses on the former 7-Series devices due to their greater accessibility for beginner users and compatibility with open source tooling.
-
-    Figure~\ref{fig:hierarchy} illustrates the architectural hierarchy of an Ultrascale Xilinx FPGA. 
-    The architecture hierarchy is the same for the 7-Series devices. 
-    At its most basic level, an FPGA is a vast physical array of replicated atomic components known as Basic Elements of Logic (BELs). 
-    These BELs, which include programmable lookup tables (LUTs), flip flops (FFs), and programmable interconnects, form the fundamental building blocks for custom digital circuit implementations on FPGAs.
-
-    In the FPGAs 7-Series FPGAs, additional BELs like Block RAM (BRAM) and Digital Processing Slices (DSPs) are also available, creating a more heterogeneous programmable logic fabric. 
-    These dedicated resources allow designers to implement common digital structures, such as multipliers and memories, without having to recreate them from scratch using only LUTs and flip flops.
-
-    The Xilinx architectural hierarchy organizes these BELs into increasingly abstract structures. 
-    \textbf{BELs} are packaged into \textbf{Sites}, which are then embedded into \textbf{Tiles}, which are subsequently consolidated into \textbf{Clock Regions}. 
-    In high-end Xilinx devices, Clock Regions may be further consolidated into \textbf{Super Logic Regions} (SLRs). 
-    However, for the scope of this paper, we focus on Xilinx devices that have only a single SLR.
+    As designs become increasingly large and complex, FPGAs become increasingly heterogenous as they incorporate a wider variety of hard macros into the fabric.
 
@@
 
-    Throughout the flow, additional tools, such as constraint managers and timing analyzers, are used to enforce design rules and verify that the implementation meets required performance specifications.
-
     Parallel to this entire design flow are simulators for which the user can write testbenches to verify their design at different stages of implementation.
@@
     This integrated toolchain, whether provided by FPGA vendors or open source communities, facilitates the transformation from an abstract HDL description to a functioning hardware configuration.
+
+\section{Xilinx 7-Series Architecture}
+
+    \begin{figure*}
+        \centering
+        \includegraphics[width=\textwidth]{figures/hierarchy.png}
+        \caption{Architecture Hierarchy of a Xilinx FPGA (Ultrascale)}
+        \label{fig:hierarchy}
+    \end{figure*}
+
+    Xilinx introduced the 7-Series architecture in 2010 which followed the heterogeneous island-style architecture described previously. 
+    Then in 2013 Xilinx unveiled Ultrascale architecture which kept the same heterogenous island-style architecture but made several optimization over the 7-Series architecture.
+    Although the latest FPGAs follow the Ultrascale architecture, this paper focuses on the 7-Series FPGAs due to their greater accessibility for beginner users and compatibility with open source tooling.
+
+    Figure~\ref{fig:hierarchy} illustrates the architectural hierarchy of an Ultrascale Xilinx FPGA. 
+    The architecture hierarchy is the same for the 7-Series devices. 
+    The only difference between Ultrascale and 7-Series is simply the design of the CLBs (Sites and BELs) themselves. 
+    At its most basic level, an FPGA is a vast physical array of replicated atomic components known as Basic Elements of Logic (BELs). 
+    These BELs, which include LUTs, FFs, BRAMs, DSPs, and programmable interconnects, form the fundamental building blocks for custom digital circuit implementations on FPGAs.
+
+    The Xilinx architectural hierarchy organizes these BELs into increasingly abstract structures. 
+    \textbf{BELs} are packaged into \textbf{Sites}, which are then embedded into \textbf{Tiles}, which are subsequently consolidated into \textbf{Clock Regions}. 
+    In high-end Xilinx devices, Clock Regions may be further consolidated into \textbf{Super Logic Regions} (SLRs). 
+    However, for the scope of this paper, we focus on Xilinx devices that have only a single SLR.
 
@@
         \caption{Showing how RapidWright can take a design checkpoint at different stages of the Vivado toolchain, make customizations, then feed it back into the toolchain.}
-        \label{fig:device_carry_chain_routing}
+        \label{fig:vivado_dcps}
     \end{figure}
 
-    DO WE REALLY NEED THIS SECTION? JUST REFERENCE THE RW DOCUMENTATION.
+    We use the RapidWright API to implement a custom placer for Xilinx 7-Series FPGAs.
+    RapidWright is an open source Java framework that enables netlist implementation manipulation for modern Xilinx FPGAs. 
+    Figure \ref{fig:vivado_dcps} shows how the API can take design checkpoints in the form of .dcp files generated at any stage of the design implementation, allow the user to make custom optimizations, and reintegrate it back into the design flow.
+    This allows the user to substitute any stage in the default FPGA design flow offered by Vivado with user-defined optimizations and implementations to meet complex design requirements. 
+    There are three packages of objects that are most relavent to building our placer: the edif package, design package, and device package.
 
-    The RapidWright API offers a user-friendly toolkit to take pre-implemented designs, that is, designs that have already been placed and/or routed by Vivado, and make custom optimizations to fit complex design criteria.
-    RapidWright has several "BlockPlacers", a general Router, but no general Placer. 
-    We use this API to implement a crude general placer, in our case, a simulated annealer. 
-    (Show what parts of the FPGA toolchain that RapidWright already has and how our placer fits into that toolchain).
+    maybe just list the relevant objects?
 
-    There are three packages of objects that are most relavent to building our placer: the edif package, design package, and device package.
+    \subsection{EDIF Objects}
+    In Vivado, all designs post synthesis have a logical netlist that can be exported as an Electronic Design Interchange Format (EDIF) file or within a .dcp checkpoint file. 
+    This is the result of the synthesis tool parsing the raw Verilog source files, inferring logical behavior, and instantiating primitive cells to emulate the desired logical behavior. 
+    The final stage is often referred to as "technology mapping", as the primitive cells are specific to the FPGA architecture and consequently the FPGA manufacturer.
+
+    The EDIFNetlist is the top level class that contains the netlist and cell libraries. 
+    The EDIFNetlist keeps a reference to the top cell which is wrapped in the EDIFDesign class. 
+    It also maintains a top cell instance reference that is generated when the file is loaded.
+
+    \begin{figure*}
+        \centering
+        \includegraphics[width=\textwidth]{figures/routing_objects.png}
+        \caption{Examples of routing objects in a Xilinx FPGA}
+        \label{fig:routing_objects}
+    \end{figure*}
 
@@
     The most prominent class, aptly named Device, represents a specific product family member (like the xc7z020 device of the 7-series family) and provides a reference to all of the logical and routing resources available on the device.
-    There device classes can be separated into two groups: logical and routing. 
-    Among the logical classes are the BEL, Site, Tile, and ClockRegion objects discussed before. 
-    Among the routing classes are BELPin, SitePin, PIP (programmable interconnect point), Wire, and Node.
+    The device classes can be separated into two main groups: logical and routing. 
+    Among the logical classes are the BEL, Site, Tile, and ClockRegion objects as shown previously in Figure \ref{fig:hierarchy}. 
+    Shown in figure \ref{fig:routing_objects} are the routing classes including BELPin, SitePin, PIP (programmable interconnect point), Wire, and Node.
 
@@
 
-    \subsection{EDIF Package}
-    In Vivado, all designs post synthesis have a logical netlist that can be exported in the Electronic Design Interchange Format (EDIF) netlist format. 
-    The synthesis stage of the FPGA toolchain produces an EDIF file, which is a low-level file containing the technology-mapped netlist. 
-    The EDIFNetlist is the top level class that contains the netlist and cell libraries. 
-    The EDIFNetlist keeps a reference to the top cell which is wrapped in the EDIFDesign class. 
-    It also maintains a top cell instance reference that is generated when the file is loaded.
+    In its current state, RapidWright has several SA-based "BlockPlacers" that can take pre-implemented macros, a general Router, but no general Placer. 
+    To fill this gap, we attempt to build a general Placer that can take a raw EDIF netlist and create a placed design. 
+    As mentioned before, our placer will be divided into three substages: prepacking, packing, and placement.
+    We use this API to implement a crude general placer, in our case, a simulated annealer. 
+    (Show what parts of the FPGA toolchain that RapidWright already has and how our placer fits into that toolchain).
 
@@
     \subsection{PrePacking}
-        In the prepacking stage, we traverse the raw EDIF Netlist to identify recurring cell patterns. In particular, we will specifically look for CARRY chains, DSP cascades, and LUT-FF pairs.
+        In the prepacking stage, we traverse the raw EDIF Netlist to identify recurring cell patterns. 
+        In particular, we will specifically look for CARRY chains, DSP cascades, and LUT-FF pairs as these patterns are most prevalent in all FPGA designs.
         \subsection{CARRY Chains}

### 2025-03-30 latex/paper/paper.tex
more work

@@
     There are three packages of objects that are most relavent to building our placer: the edif package, design package, and device package. 
+    RapidWright reads in a post-synthesis design checkpoint and allows us to traverse the netlist's logical EDIFCells through various function calls. 
+    For each EDIFCell object in the netlist, we create a Cell design object, which we then place onto physical device BEL object. 
+    CellPins and Nets are automatically generated from EDIFHierPortInst and EDIFHierNet upon the Cell's creation from an EDIFHierCellInst.
 
-    maybe just list the relevant objects?
+    These three steps correspond to our placement substages of PrePacking, Packing, and Placement, respectively.
+    Figure \ref{fig:edif_design_device} illustrates the flow between these three substages in relation to the RapidWright object classes.
+
+    \begin{figure*}
+        \centering
+        \includegraphics[width=\textwidth]{figures/edif_design_device.png}
+        \caption{The placement flow illustrating PrePacking into Packing into Placement for a logical netlist with 3 EDIF cells.
+EHCI is short for EDIFHierCellInst.
+EHPI is short for EDIFHierPortInst.}
+        \label{edif_design_device}
+    \end{figure*}
+
+    For a complete understanding of the RapidWright packages, please refer to the short API Overview \url{https://www.rapidwright.io/docs/RapidWright_Overview.html}.
+    Listed below are the most relevant objects to our placer and a short description.
+
+    \begin{figure*}
+        \centering
+        \includegraphics[width=\textwidth]{figures/edif_netlist.png}
+        \caption{Examples of edif netlist objects for a Xilinx FPGA}
+        \label{fig:routing_objects}
+    \end{figure*}
+
+    \begin{figure*}
+        \centering
+        \includegraphics[width=\textwidth]{figures/routing_objects.png}
+        \caption{Examples of device routing objects in a Xilinx FPGA}
+        \label{fig:routing_objects}
+    \end{figure*}
 
     \subsection{EDIF Objects}
+        \begin{itemize}
+            \item EDIFNetlist
+            \item EDIFHierNet
+            \item EDIFHierCellInst
+            \item EDIFHierPortInst
+        \end{itemize}
+    \subsection{Design Objects}
+        \begin{itemize}
+            \item Netlist
+            \item Net
+            \item Cell
+            \item CellPin
+            \item SiteInst
+            \item SitePinInst
+        \end{itemize}
+    \subsection{Device Objects}
+        \begin{itemize}
+            \item BEL
+            \item PIP (programmable interconnect point)
+            \item Site
+            \item SitePin
+            \item Tile
+            \item ClockRegion
+        \end{itemize}
+
     In Vivado, all designs post synthesis have a logical netlist that can be exported as an Electronic Design Interchange Format (EDIF) file or within a .dcp checkpoint file. 
@@
 
-    The EDIFNetlist is the top level class that contains the netlist and cell libraries. 
-    The EDIFNetlist keeps a reference to the top cell which is wrapped in the EDIFDesign class. 
-    It also maintains a top cell instance reference that is generated when the file is loaded.
+    In its current state, RapidWright has several SA-based "BlockPlacers", a general Router, but no general Placer. 
+    The BlockPlacer takes multiple pre-implemented designs (that is, designs that have been placed and routed), and instantiates them as hard macros in the form of ModuleInsts. 
+    ModuleInsts are a collection of SiteInsts whose positional relationship to eachother are fixed. 
+    The BlockPlacer takes these hard ModuleInsts and performs simulated annealing on them to find an optimal placement. 
+    The Router takes the placed design and routes the nets onto the device wires and programmable interconnects.
 
-    \begin{figure*}
-        \centering
-        \includegraphics[width=\textwidth]{figures/routing_objects.png}
-        \caption{Examples of routing objects in a Xilinx FPGA}
-        \label{fig:routing_objects}
-    \end{figure*}
-
-    \subsection{Device Objects}
-    The \textbf{device} package contains the classes that correspond to constructs in the hardware and/or silicon devices. 
-    The most prominent class, aptly named Device, represents a specific product family member (like the xc7z020 device of the 7-series family) and provides a reference to all of the logical and routing resources available on the device.
-    The device classes can be separated into two main groups: logical and routing. 
-    Among the logical classes are the BEL, Site, Tile, and ClockRegion objects as shown previously in Figure \ref{fig:hierarchy}. 
-    Shown in figure \ref{fig:routing_objects} are the routing classes including BELPin, SitePin, PIP (programmable interconnect point), Wire, and Node.
-
-    \subsection{Design Package}
-    The \textbf{design} package is the collection of objects used to describe how a logical netlist maps to the device netlist. 
-    The design is also referred to as the "physical netlist" or "implementation". 
-    It contains all of the primitive logical cell mappings to hardware, specifically the cells to BEL placements and physical net mapping to programmable interconnect or routing.
-
-    In its current state, RapidWright has several SA-based "BlockPlacers" that can take pre-implemented macros, a general Router, but no general Placer. 
     To fill this gap, we attempt to build a general Placer that can take a raw EDIF netlist and create a placed design. 
@@
     (Describe the algorithm, then in context of 7-Series arhictecture). \\
+
     \subsection{PrePacking}

### 2025-03-31 latex/paper/paper.tex
more paperwork

@@
 \usepackage{mathpazo}
+\usepackage{listings}
 % \usepackage{multicol}
@@
             footskip=.25in]{geometry}
+
+\lstset{
+  basicstyle=\ttfamily\small,
+  breaklines=true,
+  columns=fullflexible,
+  frame=single,
+  backgroundcolor=\color{gray!10},
+  captionpos=b
+}
 
@@
     In these devices, hard macros such as Block RAM (BRAM) and Digital Signal Processing (DSP) slices are integrated into the programmable logic fabric along with CLBs. 
-    This design enables the direct instantiation of common subsystems like memories and multipliers, without having to implement them solely with CLBs. 
+    This design enables the direct instantiation of common subsystems like memories and multipliers, without having to recreate them from scratch using CLBs. 
     Major vendors such as Xilinx and Altera now employ heterogeneous island-style architectures in their devices. 
-    As designs become increasingly large and complex, FPGAs become increasingly heterogenous as they incorporate a wider variety of hard macros into the fabric.
+    As designs become increasingly large and complex, FPGAs become increasingly heterogenous by incorporating a wider variety of hard macros into the fabric.
 
@@
         \includegraphics[width=7.0cm]{figures/design_flow.png}
-        \caption{A typical FPGA design and verification flow}
+        \caption{A typical FPGA design and verification workflow}
         \label{fig:design_flow}
@@
 
-    Figure \ref{fig:design_flow} shows a typical FPGA design and verification flow.
+    Figure \ref{fig:design_flow} shows a typical FPGA design and verification workflow.
     The FPGA design flow takes a high-level digital design written in an HDL and generates a configuration file that programs the physical FPGA.
@@
 
-    Next, the \textbf{synthesis} tool interprets the HDL code and generates a netlist that represents the design in terms of basic logic elements, such as lookup tables (LUTs) and flip flops. 
-    During this stage, substages like instantiation, substitution, inference, optimization, and technology mapping are applied to map the user design to a particular FPGA architecture.
+    Next, the \textbf{synthesis} tool interprets the HDL code and generates a netlist that represents the design in terms of primitive cells, such as lookup tables (LUTs) and flip flops. 
+    This is the result of the synthesis tool parsing the raw Verilog source files, inferring logical behavior, and instantiating primitive cells to emulate the desired logical behavior. 
+    The final stage is often referred to as "technology mapping", as the primitive cells are specific to the FPGA architecture and consequently the FPGA manufacturer.
 
     After synthesis, the \textbf{placement} tool assigns each logic element in the netlist a specific location within the FPGA’s fabric. 
-    This stage can include substages of prepacking, packing, global placement, legalization, and detailed placement. 
+    Depending on the placement strategy, this stage can include substages of prepacking, packing, global placement, legalization, and detailed placement. 
+    Our placer will only involve prepacking, packing, and detailed placement for simplicity. 
     The objective is to place the logical cell instances of the netlist onto physical BELs on the device in a way that minimizes delay, satisfies design constraints, and simplifies routing.
@@
 
-    \begin{figure*}
+    \begin{figure*}[!hbt]
         \centering
@@
         \centering
-        \includegraphics[width=6.5cm]{figures/vivado_dcps.png}
-        \caption{Showing how RapidWright can take a design checkpoint at different stages of the Vivado toolchain, make customizations, then feed it back into the toolchain.}
+        \includegraphics[width=6.0cm]{figures/vivado_dcps.png}
+        \caption{Design flow showing how RapidWright can take a design checkpoint, allow user customization, and reintegrate it back into the design flow.}
         \label{fig:vivado_dcps}
     \end{figure}
+
+    \begin{figure*}[!htt]
+        \centering
+        \includegraphics[width=\textwidth]{figures/edif_design_device.png}
+        \caption{The placement flow illustrating PrePacking into Packing into Placement for a logical netlist with 3 EDIF cells.
+EHCI is short for EDIFHierCellInst.
+EHPI is short for EDIFHierPortInst.}
+        \label{fig:edif_design_device}
+    \end{figure*}
 
@@
     This allows the user to substitute any stage in the default FPGA design flow offered by Vivado with user-defined optimizations and implementations to meet complex design requirements. 
-    There are three packages of objects that are most relavent to building our placer: the edif package, design package, and device package. 
+    There are three packages of objects that are most relavent to building our placer: the \textbf{edif} package, \textbf{design} package, and \textbf{device} package. 
+    Figure \ref{fig:routing_objects} shows the device routing objects as shown in the Vivado device viewer while figure \ref{fig:edif_netlist} shows the EDIF netlist objects as shown in the Vivado netlist viewer.
+    Figure \ref{fig:edif_design_device} illustrates the flow between these packages in relation to our PrePacking, Packing, and Placement flow.
+
     RapidWright reads in a post-synthesis design checkpoint and allows us to traverse the netlist's logical EDIFCells through various function calls. 
@@
     These three steps correspond to our placement substages of PrePacking, Packing, and Placement, respectively.
-    Figure \ref{fig:edif_design_device} illustrates the flow between these three substages in relation to the RapidWright object classes.
 
-    \begin{figure*}
+    For a complete understanding of the RapidWright packages, please refer to the short API Overview \url{https://www.rapidwright.io/docs/RapidWright_Overview.html}.
+
+    \begin{figure*}[!htt]
         \centering
-        \includegraphics[width=\textwidth]{figures/edif_design_device.png}
-        \caption{The placement flow illustrating PrePacking into Packing into Placement for a logical netlist with 3 EDIF cells.
-EHCI is short for EDIFHierCellInst.
-EHPI is short for EDIFHierPortInst.}
-        \label{edif_design_device}
+        \includegraphics[width=12.0cm]{figures/edif_netlist.png}
+        \caption{Examples of edif netlist objects for a Xilinx FPGA}
+        \label{fig:edif_netlist}
     \end{figure*}
 
-    For a complete understanding of the RapidWright packages, please refer to the short API Overview \url{https://www.rapidwright.io/docs/RapidWright_Overview.html}.
-    Listed below are the most relevant objects to our placer and a short description.
-
-    \begin{figure*}
-        \centering
-        \includegraphics[width=\textwidth]{figures/edif_netlist.png}
-        \caption{Examples of edif netlist objects for a Xilinx FPGA}
-        \label{fig:routing_objects}
-    \end{figure*}
-
-    \begin{figure*}
+    \begin{figure*}[!htt]
         \centering
@@
 
-    \subsection{EDIF Objects}
-        \begin{itemize}
-            \item EDIFNetlist
-            \item EDIFHierNet
-            \item EDIFHierCellInst
-            \item EDIFHierPortInst
-        \end{itemize}
-    \subsection{Design Objects}
-        \begin{itemize}
-            \item Netlist
-            \item Net
-            \item Cell
-            \item CellPin
-            \item SiteInst
-            \item SitePinInst
-        \end{itemize}
-    \subsection{Device Objects}
-        \begin{itemize}
-            \item BEL
-            \item PIP (programmable interconnect point)
-            \item Site
-            \item SitePin
-            \item Tile
-            \item ClockRegion
-        \end{itemize}
+\section{Placement Implementation}
 
-    In Vivado, all designs post synthesis have a logical netlist that can be exported as an Electronic Design Interchange Format (EDIF) file or within a .dcp checkpoint file. 
-    This is the result of the synthesis tool parsing the raw Verilog source files, inferring logical behavior, and instantiating primitive cells to emulate the desired logical behavior. 
-    The final stage is often referred to as "technology mapping", as the primitive cells are specific to the FPGA architecture and consequently the FPGA manufacturer.
-
-    In its current state, RapidWright has several SA-based "BlockPlacers", a general Router, but no general Placer. 
-    The BlockPlacer takes multiple pre-implemented designs (that is, designs that have been placed and routed), and instantiates them as hard macros in the form of ModuleInsts. 
-    ModuleInsts are a collection of SiteInsts whose positional relationship to eachother are fixed. 
-    The BlockPlacer takes these hard ModuleInsts and performs simulated annealing on them to find an optimal placement. 
-    The Router takes the placed design and routes the nets onto the device wires and programmable interconnects.
-
-    To fill this gap, we attempt to build a general Placer that can take a raw EDIF netlist and create a placed design. 
-    As mentioned before, our placer will be divided into three substages: prepacking, packing, and placement.
-    We use this API to implement a crude general placer, in our case, a simulated annealer. 
-    (Show what parts of the FPGA toolchain that RapidWright already has and how our placer fits into that toolchain).
-
-\section{Placement Implementation}
-    (Describe the algorithm, then in context of 7-Series arhictecture). \\
+    With a basic understanding of FPGA architecture, design placement, and RapidWright, we can begin implementing our SA placer. 
+    Here we outline in detail each substage of our implementation.
 
     \subsection{PrePacking}
-        In the prepacking stage, we traverse the raw EDIF Netlist to identify recurring cell patterns. 
-        In particular, we will specifically look for CARRY chains, DSP cascades, and LUT-FF pairs as these patterns are most prevalent in all FPGA designs.
+        In the prepacking stage, we traverse the raw EDIF Netlist to identify recurring Cell patterns. 
+        We do this because there are certain desirable cell configurations that should be clustered together to exploit some proximity characteristics of the FPGA architecture 
+        There are also some cell configurations that must necessarily be placed in certain ways in relation to each other to ensure legality within the architecture constraints. 
+        First, we traverse the entire EDIF netlist and identify all unique Cell types and group them together via a {\tt HashMap<String, List<EDIFHierCellInst>>}, where String is the name cell type. 
+        Cell type groups can include {\tt CARRY4}, {\tt LUT1-6}, {\tt FF}, {\tt DSP48E1}, {\tt RAMB18E1}, and others. 
+        We specifically look for CARRY chains, DSP cascades, and LUT-FF pairs as these patterns are most prevalent in all FPGA designs.
+
         \subsection{CARRY Chains}
+            To find CARRY chains, we take the CARRY4 group and follow the below pseudocode. 
+            \begin{verbatim}
+                for (EDIFHierCellInst ehci : CARRYCells) {
+                    ehci.getPorts()
+
+                }
+            \end{verbatim}
             Each SLICE Site contains one CARRY4 BEL.
@@
             For SLICEMs, LUT6s can be configured as shallow 32-bit LUTRAMs or "RAMS32".
-    \subsection{Placement}
+    \subsection{Placement: Simulated Annealing}
         Create BEL "fields": CLB, DSP, CLB and DSP Chains, RAMs
@@
         \subsection{Detailed Placement}
+
+\section{Placement Results}
 

### 2025-03-31 java/src/main/java/placer/Main.java

@@ -85,7 +85,7 @@ public class Main {
             for (PlacerAnnealRandom placer : SAPlacers) {
                 System.out.println("\n\nStarting " + placer.getPlacerName() + "... \n\n");
                 placer.makeOutputDirs(placer.getPlacerName());
-                placer.initCoolingSchedule(10000.0d, 0.98d, 400);
+                placer.initCoolingSchedule(10000.0d, 0.98d, 300);
                 placer.run(packedDesign);
             }
 

### 2025-04-02 latex/paper/paper.tex
more paperwork

@@
 \usepackage{mathpazo}
+\usepackage{xcolor}
+\usepackage{float}
 \usepackage{listings}
-% \usepackage{multicol}
-\usepackage[a4paper,
-            bindingoffset=0.2in,
-            left=1in,
-            right=1in,
-            top=1in,
-            bottom=1in,
-            footskip=.25in]{geometry}
-
 \lstset{
@@
   frame=single,
-  backgroundcolor=\color{gray!10},
-  captionpos=b
+  showspaces=false,
+  showtabs=false,
+  keepspaces=false
 }
+
+% \usepackage{multicol}
+\usepackage[a4paper,
+            bindingoffset=0.2in,
+            left=0.5in,
+            right=0.5in,
+            top=0.5in,
+            bottom=0.5in,
+            footskip=.25in]{geometry}
 
@@
 \section{Abstract}
-    fdsafdsafdsa.
+    fdsafdsafdsa. \\ 
+    fdsafdsafdsa. \\ 
+    fdsafdsafdsa. \\ 
+    fdsafdsafdsa. \\ 
+    fdsafdsafdsa. \\ 
+    fdsafdsafdsa. \\ 
+    fdsafdsafdsa. \\ 
+    fdsafdsafdsa. \\ 
+    fdsafdsafdsa. \\ 
+    fdsafdsafdsa. \\ 
+    fdsafdsafdsa. \\ 
+    fdsafdsafdsa. \\ 
+    fdsafdsafdsa. \\
 
@@
     Since the early-to-mid 2000s, the stagnation of single-processor performance relative to the rapid increase in integrated circuit sizes has led to a design productivity gap, where the computational effort for designing complex chips continues to rise.
-    In FPGA CAD flows—which traditionally encompass logic synthesis, placement, routing, and bitstream generation—the placement stage has emerged as one of the most time-consuming processes. 
-    Inefficiencies in placement not only extend design times from hours to days, thereby elevating cost and reducing engineering productivity, but also limit the broader adoption of FPGAs by software engineers who expect compile times akin to those of conventional software compilers.
+    FPGA CAD flows mainly encompass synthesis, placement, and routing; all of which are HP-hard problems, of which placement is one of the most time-consuming processes. 
+    Inefficienct placement strategy not only extends design times from hours to days, thereby elevating cost and reducing engineering productivity, but also limits the broader adoption of FPGAs by software engineers who expect compile times akin to those of conventional software compilers like {\tt gcc}.
 
     For these reasons, FPGA placement remains a critical research effort even today. 
-    In this paper, we study and implement established placement methods by leveraging RapidWright, which is a semi-open-source API that offers backend access to Xilinx's industry-standard FPGA environment, Vivado.
+    In this paper, we study and implement established placement methods. 
+    To do this, we use the RapidWright API, which is a semi-open-source research effort from AMD/Xilinx that enables custom solutions to FPGA design implementations and design tools that are not offered by their industry-standard FPGA environment, Vivado.
     We implement multiple variations of simulated annealing placers for Xilinx's 7-series FPGAs, with an emphasis on minimizing total wirelength while mitigating runtime. 
@@
         \centering
-        \includegraphics[width=8.5cm]{figures/pal_2.png}
+        \includegraphics[width=8.0cm]{figures/pal_2.png}
         \caption{PAL architecture with 5 inputs, 8 programmable AND gates and 4 fixed OR gates}
@@
 
-    Before any work can begin on a placer, it is necessary to understand both the objects to be placed and the medium in which they are placed.
+    Before any work can begin on an FPGA placer, it is necessary to understand both the objects being placed and the medium in which they are placed.
 
@@
     The PAL simplified the PLA by fixing the OR gates, resulting in a fixed-OR, programmable-AND design, which sacrificed some logic flexibility to simplify its manufacture. 
+    Figure \ref{fig:pla} shows one such PLA architecture.
 
@@
         \centering
-        \includegraphics[width=7.5cm]{figures/cpld.png}
+        \includegraphics[width=8.0cm]{figures/cpld.png}
         \caption{CPLD architecture with 4 CLBs (PAL-like blocks)}
@@
     These CLBs were typically modified PAL blocks that included the PAL itself along with macrocells such as flip-flops, multiplexers, and tri-state buffers. 
-    The CPLD functioned as an array of PALs connected by a programmable switch matrix and could be programmed using a hardware description language (HDL) like VHDL.
+    The CPLD functioned as an array of PALs connected by a central programmable switch matrix and could be programmed using a hardware description language (HDL) like VHDL. 
+    Figure \ref{fig:cpld} shows one such CPLD architecture.
 
@@
         \centering
-        \includegraphics[width=7.0cm]{figures/fpga.png}
-        \caption{Homogenous island-style FPGA architecture with 16 CLBs in a grid}
+        \includegraphics[width=8.0cm]{figures/homogenous_fpga.png}
+        \includegraphics[width=8.0cm]{figures/heterogenous_fpga_3.jpg}
+        \caption{
+            \textbf{Top}: A homogeneous island-style FPGA architecture with 16 CLBs in a grid. 
+            \textbf{Bottom}: A heterogeneous island-style FPGA with a mix of CLBs and macrocells.
+        }
         \label{fig:fpga}
+        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/FPGA/Field-programmable-gate-array-FPGA/
         % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/FPGA/Field-programmable-gate-array-FPGA/
@@
     The mid-1980s saw the introduction of the homogeneous FPGA, which was built as a grid of CLBs. 
-    Rather than using a central programmable switch matrix as in CPLDs, FPGAs adopted an island style architecture in which each CLB is surrounded on all sides by programmable routing resources. 
+    Rather than using a central programmable switch matrix as in CPLDs, FPGAs adopted an island style architecture in which each CLB is surrounded on all sides by programmable routing resources, as shown in Figure \ref{fig:homogenous_fpga}.
     The first commercially viable FPGA, produced by Xilinx in 1984, featured 16 CLBs arranged in a 4x4 grid. 
@@
     To meet the needs of increasingly complex designs, FPGA vendors introduced heterogeneous FPGAs. 
-    In these devices, hard macros such as Block RAM (BRAM) and Digital Signal Processing (DSP) slices are integrated into the programmable logic fabric along with CLBs. 
+    In these devices, hard macros such as Block RAM (BRAM) and Digital Signal Processing (DSP) slices are integrated into the programmable logic fabric along with CLBs, like shown in Figure \ref{fig:fpga}.
     This design enables the direct instantiation of common subsystems like memories and multipliers, without having to recreate them from scratch using CLBs. 
@@
 
-    \begin{figure*}[!hbt]
+    \begin{figure*}[t]
         \centering
-        \includegraphics[width=\textwidth]{figures/hierarchy.png}
+        \includegraphics[width=12.0cm]{figures/hierarchy.png}
         \caption{Architecture Hierarchy of a Xilinx FPGA (Ultrascale)}
@@
     Xilinx introduced the 7-Series architecture in 2010 which followed the heterogeneous island-style architecture described previously. 
-    Then in 2013 Xilinx unveiled Ultrascale architecture which kept the same heterogenous island-style architecture but made several optimization over the 7-Series architecture.
+    Then in 2013 Xilinx unveiled Ultrascale architecture which also followed the heterogenous island-style design but made several improvements over the 7-Series architecture.
     Although the latest FPGAs follow the Ultrascale architecture, this paper focuses on the 7-Series FPGAs due to their greater accessibility for beginner users and compatibility with open source tooling.
@@
     The architecture hierarchy is the same for the 7-Series devices. 
-    The only difference between Ultrascale and 7-Series is simply the design of the CLBs (Sites and BELs) themselves. 
     At its most basic level, an FPGA is a vast physical array of replicated atomic components known as Basic Elements of Logic (BELs). 
@@
 \section{RapidWright API}
-    \begin{figure}
+    We use the RapidWright API to implement a custom placer for Xilinx 7-Series FPGAs.
+    RapidWright is an open source Java framework that enables netlist implementation manipulation for modern Xilinx FPGAs.
+
+    \begin{figure}[]
         \centering
-        \includegraphics[width=6.0cm]{figures/vivado_dcps.png}
-        \caption{Design flow showing how RapidWright can take a design checkpoint, allow user customization, and reintegrate it back into the design flow.}
+        \includegraphics[width=8.0cm]{figures/vivado_dcps.png}
+        \caption{Showing how RapidWright can take a design checkpoint from any stage of the design flow, allow user customization, and reintegrate it back into the design flow.}
         \label{fig:vivado_dcps}
@@
 
-    \begin{figure*}[!htt]
+    \begin{figure}[]
         \centering
-        \includegraphics[width=\textwidth]{figures/edif_design_device.png}
-        \caption{The placement flow illustrating PrePacking into Packing into Placement for a logical netlist with 3 EDIF cells.
-EHCI is short for EDIFHierCellInst.
-EHPI is short for EDIFHierPortInst.}
+        \includegraphics[width=8.5cm]{figures/edif_design_device.png}
+        \caption{The placement flow illustrating PrePacking into Packing into Placement for a logical netlist with 3 EDIF cells. EHCI is short for EDIFHierCellInst.}
         \label{fig:edif_design_device}
+    \end{figure}
+
+    Figure \ref{fig:vivado_dcps} shows how the API allows us to take design checkpoints in the form of .dcp files generated at any stage of the design flow, allow the user to make custom optimizations, and reintegrate it back into the design flow.
+    This allows the user to substitute any stage in the default FPGA design flow offered by Vivado with a custom user-defined process to meet complex design constraints. 
+    There are three packages of objects that are most relevant to building our placer: the \textbf{edif} package, \textbf{design} package, and \textbf{device} package. 
+    Figure \ref{fig:edif_objects} shows the EDIF netlist objects and device routing objects as shown in the Vivado netlist viewer and device viewer.
+    Figure \ref{fig:edif_design_device} illustrates the flow between these packages in relation to our PrePacking, Packing, and Placement flow.
+
+    \begin{figure*}[t]
+        \centering
+        \includegraphics[width=14.0cm]{figures/edif_netlist.png}
+        \includegraphics[width=14.0cm]{figures/routing_objects.png}
+        \caption{Examples of edif netlist objects (top) and routing objects (bottom) for a Xilinx FPGA}
+        \label{fig:edif_objects}
     \end{figure*}
-
-    We use the RapidWright API to implement a custom placer for Xilinx 7-Series FPGAs.
-    RapidWright is an open source Java framework that enables netlist implementation manipulation for modern Xilinx FPGAs. 
-    Figure \ref{fig:vivado_dcps} shows how the API can take design checkpoints in the form of .dcp files generated at any stage of the design implementation, allow the user to make custom optimizations, and reintegrate it back into the design flow.
-    This allows the user to substitute any stage in the default FPGA design flow offered by Vivado with user-defined optimizations and implementations to meet complex design requirements. 
-    There are three packages of objects that are most relavent to building our placer: the \textbf{edif} package, \textbf{design} package, and \textbf{device} package. 
-    Figure \ref{fig:routing_objects} shows the device routing objects as shown in the Vivado device viewer while figure \ref{fig:edif_netlist} shows the EDIF netlist objects as shown in the Vivado netlist viewer.
-    Figure \ref{fig:edif_design_device} illustrates the flow between these packages in relation to our PrePacking, Packing, and Placement flow.
 
@@
 
-    For a complete understanding of the RapidWright packages, please refer to the short API Overview \url{https://www.rapidwright.io/docs/RapidWright_Overview.html}.
-
-    \begin{figure*}[!htt]
-        \centering
-        \includegraphics[width=12.0cm]{figures/edif_netlist.png}
-        \caption{Examples of edif netlist objects for a Xilinx FPGA}
-        \label{fig:edif_netlist}
-    \end{figure*}
-
-    \begin{figure*}[!htt]
-        \centering
-        \includegraphics[width=\textwidth]{figures/routing_objects.png}
-        \caption{Examples of device routing objects in a Xilinx FPGA}
-        \label{fig:routing_objects}
-    \end{figure*}
+    For a complete understanding of the RapidWright packages, please refer to the API Overview \url{https://www.rapidwright.io/docs/RapidWright_Overview.html}.
 
@@
 
-    With a basic understanding of FPGA architecture, design placement, and RapidWright, we can begin implementing our SA placer. 
+    With a basic understanding of FPGA architecture, design placement, and RapidWright, we have all the necessary pieces to implement our SA placer. 
     Here we outline in detail each substage of our implementation.
 
+    \begin{figure*}[t]
+        \centering
+        \includegraphics[width=14.0cm]{figures/substages.png}
+        \caption{The data classes populated at each substage: PrepackedDesign, PackedDesign, and PlacedDesign.}
+        \label{fig:substages}
+    \end{figure*}
+
     \subsection{PrePacking}
-        In the prepacking stage, we traverse the raw EDIF Netlist to identify recurring Cell patterns. 
-        We do this because there are certain desirable cell configurations that should be clustered together to exploit some proximity characteristics of the FPGA architecture 
+        In the prepacking stage, we traverse the raw EDIF Netlist to identify common recurring Cell patterns. 
+        We do this because there are certain desirable cell configurations that should be clustered together to exploit some proximity characteristics of the FPGA architecture. 
         There are also some cell configurations that must necessarily be placed in certain ways in relation to each other to ensure legality within the architecture constraints. 
@@
         Cell type groups can include {\tt CARRY4}, {\tt LUT1-6}, {\tt FF}, {\tt DSP48E1}, {\tt RAMB18E1}, and others. 
-        We specifically look for CARRY chains, DSP cascades, and LUT-FF pairs as these patterns are most prevalent in all FPGA designs.
+        We specifically look for CARRY chains, DSP cascades, and LUT-FF pairs as these patterns are common to nearly all FPGA designs.
 
         \subsection{CARRY Chains}
+            \begin{figure}[]
+                \centering
+                \includegraphics[width=5.0cm]{figures/carry_cell_edif.png}
+                \includegraphics[width=3.0cm]{figures/carry_cell_device.png}
+                \caption{
+                    Left: A CARRY4 EDIFCell as seen in the Vivado netlist viewer.
+                    Right: The corresponding CARRY4 Cell as seen in the device viewer.
+                }
+                \label{fig:carry_cell_edif}
+            \end{figure}
+
+            \begin{figure*}[t]
+                \centering
+                \includegraphics[width=\textwidth]{figures/carry_chain_edif.png}
+                \caption{Example of a CARRY4 chain of length 6 as seen in the Vivado netlist viewer.}
+                \label{fig:carry_chain_edif}
+            \end{figure*}
+
             To find CARRY chains, we take the CARRY4 group and follow the below pseudocode.
-            \begin{verbatim}
-                for (EDIFHierCellInst ehci : CARRYCells) {
-                    ehci.getPorts()
 
-                }
-            \end{verbatim}
             Each SLICE Site contains one CARRY4 BEL.
@@
                 \includegraphics[width=3.0cm]{figures/dsp_cascade_routes.png}
-                \caption{A Carry Chain of length 6 spanning 6 Tiles. On the left, the cell-only representation. On the right, cells with routing representation.}
+                \caption{Left: A Carry Chain of length 6 spanning 6 Sites across 6 Tiles. On the left, the cell-only representation. Right: A DSP cascade of length 2 spanning 2 Sites across 1 Tile.}
                 \label{fig:device_carry_chain_routing}
@@
 
-\section{Analytical Placement}
-    Introduce placement with legalization methods. \\
-    Many AP approaches use a Global placement, Legalization, then Detailed placement approach. \\
-    Our SA only considers legal moves via bookkeeping structures, so it has no concept of global placement or legalization. \\
-    It is a detailed placement only approach. \\
-    \begin{itemize}
-        \item Talk about HeAP as an example.
-        \item Global Placement, Legalization, then Detailed Placement approach.
-        \item Global placement using the Bound2Bound net model from SimPL.
-        \item Legalization Step 1: \\
-            Find an area of the device that is overutilized (illegal) for twhich the blocks contained within must spread to a larger area. \\
-            To obtain this overutilized area, adjacent locations (Sites, Tiles, etc.) on the device that are occupied by more than one block (SiteInst, ModuleInst, etc.) are repeatedly clustered together, until all clusters are bordered on all sides by non-overutilized locations.
+% \section{Analytical Placement}
+%     Introduce placement with legalization methods. \\
+%     Many AP approaches use a Global placement, Legalization, then Detailed placement approach. \\
+%     Our SA only considers legal moves via bookkeeping structures, so it has no concept of global placement or legalization. \\
+%     It is a detailed placement only approach. \\
+%     \begin{itemize}
+%         \item Talk about HeAP as an example.
+%         \item Global Placement, Legalization, then Detailed Placement approach.
+%         \item Global placement using the Bound2Bound net model from SimPL.
+%         \item Legalization Step 1: \\
+%             Find an area of the device that is overutilized (illegal) for twhich the blocks contained within must spread to a larger area. \\
+%             To obtain this overutilized area, adjacent locations (Sites, Tiles, etc.) on the device that are occupied by more than one block (SiteInst, ModuleInst, etc.) are repeatedly clustered together, until all clusters are bordered on all sides by non-overutilized locations.
 \\
-            Then, the area (?) is expanded in both the x and y dimensions (?) until it is large enough to accommodate all blocks contained within. \\
-            Specifically, the area is expanded until its "occupance" \(O_A\) divided by its "capacity" \(C_A\) is less than a maximum "utilization factor" $\beta$, where $\beta$ is less than or equal to 1.
+%             Then, the area (?) is expanded in both the x and y dimensions (?) until it is large enough to accommodate all blocks contained within. \\
+%             Specifically, the area is expanded until its "occupance" \(O_A\) divided by its "capacity" \(C_A\) is less than a maximum "utilization factor" $\beta$, where $\beta$ is less than or equal to 1.
 HeAP uses 0.9.
-        \item Legalization Step 2: \\
-            Two cuts are generated: a "source cut" and a "target cut". \\
-            The source cut pertains to the blocks being placed (SiteInsts in home buffer). \\ 
-            The target cut pertains to the area into which the blocks are placed (away buffer). \\
-            The source cut splits the blocks into 2 partitions, while the target cut splits the area into 2 sub-areas, into which the blocks in each partition are spread. \\
-            Two objectives are minimized during this process: The imbalance between the number of blocks in each partition, and the difference in the utilization of each sub-area. \\
-            Utilization defined as the occupancy divided by the capacity of the sub-area. \\
-            \(U_{sub-area} = \frac{O_{sub-area}}{C_{sub-area}}\) \\
-            To generate the source cut, the cells are first sorted by their x or y location, depending on the orientation of the desired cut. \\
-            Once the cells are sorted, the source cut generation is akin to choosing a pivot in a sorted list, where all blocks to the left of the pivot are assigned to the left/bottom partition and all blocks to the right of the pivot are assigned to the right/top partition.
+%         \item Legalization Step 2: \\
+%             Two cuts are generated: a "source cut" and a "target cut". \\
+%             The source cut pertains to the blocks being placed (SiteInsts in home buffer). \\ 
+%             The target cut pertains to the area into which the blocks are placed (away buffer). \\
+%             The source cut splits the blocks into 2 partitions, while the target cut splits the area into 2 sub-areas, into which the blocks in each partition are spread. \\
+%             Two objectives are minimized during this process: The imbalance between the number of blocks in each partition, and the difference in the utilization of each sub-area. \\
+%             Utilization defined as the occupancy divided by the capacity of the sub-area. \\
+%             \(U_{sub-area} = \frac{O_{sub-area}}{C_{sub-area}}\) \\
+%             To generate the source cut, the cells are first sorted by their x or y location, depending on the orientation of the desired cut. \\
+%             Once the cells are sorted, the source cut generation is akin to choosing a pivot in a sorted list, where all blocks to the left of the pivot are assigned to the left/bottom partition and all blocks to the right of the pivot are assigned to the right/top partition.
 \\
-            The target cut is an x or y cut of the area such that all blocks in each partition fit in their respective sub-areas, and such that \( | U_{sub-area_1} 
-             - U_{sub-area_2} | \) is minimized.
-
-    \end{itemize}
+%             The target cut is an x or y cut of the area such that all blocks in each partition fit in their respective sub-areas, and such that \( | U_{sub-area_1} 
+%              - U_{sub-area_2} | \) is minimized.
+%     \end{itemize}
 

### 2025-04-02 java/src/main/java/placer/ImageMaker.java

@@ -58,6 +58,30 @@ public class ImageMaker {
         SITE_TYPE_COLORS.put(SiteTypeEnum.IOB33M, Color.decode("#888888")); // Medium Gray
     }
 
+    // private static final Map<SiteTypeEnum, Color> SITE_TYPE_COLORS = new
+    // HashMap<>();
+    // static {
+    // // Dull empty colors
+    // SITE_TYPE_COLORS.put(SiteTypeEnum.SLICEL, Color.decode("#000088")); // Dull
+    // Blue
+    // SITE_TYPE_COLORS.put(SiteTypeEnum.SLICEM, Color.decode("#880000")); // Dull
+    // Green
+    // SITE_TYPE_COLORS.put(SiteTypeEnum.RAMB18E1, Color.decode("#880000")); // Dull
+    // Red
+    // SITE_TYPE_COLORS.put(SiteTypeEnum.FIFO18E1, Color.decode("#880000")); // Dull
+    // Red
+    // SITE_TYPE_COLORS.put(SiteTypeEnum.RAMB36E1, Color.decode("#880000")); // Dull
+    // Red
+    // SITE_TYPE_COLORS.put(SiteTypeEnum.DSP48E1, Color.decode("#444400")); // Dull
+    // Yellow
+    // SITE_TYPE_COLORS.put(SiteTypeEnum.BUFGCTRL, Color.decode("#888888")); //
+    // Medium Gray
+    // SITE_TYPE_COLORS.put(SiteTypeEnum.IOB33S, Color.decode("#888888")); // Medium
+    // Gray
+    // SITE_TYPE_COLORS.put(SiteTypeEnum.IOB33M, Color.decode("#888888")); // Medium
+    // Gray
+    // }
+
     private static final Map<SiteTypeEnum, Color> PLACEMENT_COLORS = new HashMap<>();
     static {
         // Bright active colors
@@ -70,6 +94,16 @@ public class ImageMaker {
         PLACEMENT_COLORS.put(SiteTypeEnum.BUFGCTRL, Color.decode("#FFFFFF")); // White
         PLACEMENT_COLORS.put(SiteTypeEnum.IOB33S, Color.decode("#FFFFFF")); // White
         PLACEMENT_COLORS.put(SiteTypeEnum.IOB33M, Color.decode("#FFFFFF")); // White
+    }
+
+    public ImageMaker(Device device) throws IOException {
+        this.design = null;
+        this.device = device;
+        this.uniqueSiteTypes = new HashSet<>();
[…]

### 2025-04-02 java/src/main/java/placer/Main.java

@@ -53,6 +53,8 @@ public class Main {
 
             Design design = Design.readCheckpoint(synthesizedDcp);
             Device device = Device.getDevice("xc7z020clg400-1");
+            ImageMaker imdev = new ImageMaker(device);
+            imdev.exportImage(rootDir + "/outputs/device.png");
 
             // Stage 1) Prepacker:
             // works entirely on the EDIFHierCellInst level.

### 2025-04-02 latex/paper/paper.tex
larger filter

@@
 \begin{document}
-\title{MS Technical Paper: \\ Placement Algorithms for Heterogenous FPGAs}
+\title{MS Technical Paper: \\ Placement Algorithms for Heterogeneous FPGAs}
 \author{Brian B Cheng \\ Rutgers University Department of Electrical and Computer Engineering}
@@
     The mid-1980s saw the introduction of the homogeneous FPGA, which was built as a grid of CLBs. 
-    Rather than using a central programmable switch matrix as in CPLDs, FPGAs adopted an island style architecture in which each CLB is surrounded on all sides by programmable routing resources, as shown in Figure \ref{fig:homogenous_fpga}.
+    Rather than using a central programmable switch matrix as in CPLDs, FPGAs adopted an island style architecture in which each CLB is surrounded on all sides by programmable routing resources, as shown in Figure \ref{fig:fpga}.
     The first commercially viable FPGA, produced by Xilinx in 1984, featured 16 CLBs arranged in a 4x4 grid. 
@@
     Major vendors such as Xilinx and Altera now employ heterogeneous island-style architectures in their devices. 
-    As designs become increasingly large and complex, FPGAs become increasingly heterogenous by incorporating a wider variety of hard macros into the fabric.
+    As designs become increasingly large and complex, FPGAs meet the demand by becoming increasingly heterogenous by incorporating a wider variety of hard macros into the fabric.
 
@@
     Xilinx introduced the 7-Series architecture in 2010 which followed the heterogeneous island-style architecture described previously. 
-    Then in 2013 Xilinx unveiled Ultrascale architecture which also followed the heterogenous island-style design but made several improvements over the 7-Series architecture.
+    Then in 2013 Xilinx unveiled the Ultrascale architecture which also followed the heterogeneous island-style design but made several improvements over the 7-Series architecture.
     Although the latest FPGAs follow the Ultrascale architecture, this paper focuses on the 7-Series FPGAs due to their greater accessibility for beginner users and compatibility with open source tooling.
@@
         We do this because there are certain desirable cell configurations that should be clustered together to exploit some proximity characteristics of the FPGA architecture. 
+        Doing so will innately improve wirelength minimization. 
         There are also some cell configurations that must necessarily be placed in certain ways in relation to each other to ensure legality within the architecture constraints. 
@@
             \end{figure}
+
         \subsection{DSP Cascades}

### 2025-04-02 java/src/main/java/placer/PlacerAnnealRandom.java

@@ -290,14 +290,17 @@ public class PlacerAnnealRandom extends Placer {
             Site site = device.getSite(siteTypePrefix + "X" + instX + "Y" + i);
             sites.add(site);
         }
-        if (finalTailInstY - finalAnchorInstY > 16) {
-            System.out.println("buffer size: " + (finalTailInstY - finalAnchorInstY));
-            System.out.println("\tinitAnchor: " + initAnchor.getInstanceY() + ", initTail: " + initTail.getInstanceY());
-            System.out.println("\tfinalAnchor: " + finalAnchorInstY + ", finalTail: " + finalTailInstY);
-            for (Site site : sites) {
-                System.out.println("\t\tSiteInst: " + occupiedSites.get(site.getSiteTypeEnum()).get(site));
-            }
-        }
+        // if (finalTailInstY - finalAnchorInstY > 16) {
+        // System.out.println("buffer size: " + (finalTailInstY - finalAnchorInstY));
+        // System.out.println("\tinitAnchor: " + initAnchor.getInstanceY() + ",
+        // initTail: " + initTail.getInstanceY());
+        // System.out.println("\tfinalAnchor: " + finalAnchorInstY + ", finalTail: " +
+        // finalTailInstY);
+        // for (Site site : sites) {
+        // System.out.println("\t\tSiteInst: " +
+        // occupiedSites.get(site.getSiteTypeEnum()).get(site));
+        // }
+        // }
         return sites;
     } // end findBufferZone()
 

### 2025-04-02 run.sh

@@ -115,6 +115,9 @@ EOL
         fi
     done
 
+    xvlog "$XILINX_VIVADO/data/verilog/src/glbl.v" -L xil_defaultlib -L uvm --incr --relax
+    check_exit_status "xvlog for glbl"
+
     # Read verification files and log
     verif_files=("$DESIGN_DIR"/verif/*.sv)
     for file in "${verif_files[@]}"; do
@@ -127,8 +130,9 @@ EOL
     # Elaboration
     xelab -debug typical -top "tb_$TOP_LEVEL" -snapshot "${TOP_LEVEL}_functional" \
         -timescale 1ns/1ps \
-        -L xpm \
-        $XELAB_TOP_PARAMS
+        -L xil_defaultlib -L uvm -L secureip -L xpm -L simprims_ver -L unisims_ver \
+        $XELAB_TOP_PARAMS \
+        glbl
     # -L xil_defaultlib -L uvm -L secureip -L unisims_ver -L simprims_ver
     check_exit_status "xelab"
 
@@ -158,15 +162,14 @@ if [ "$start_stage" == "place" ] || [ "$start_stage" == "all" ]; then
     gradle run
     check_exit_status "Gradle run"
     cd $PROJ_DIR
-    python3 python/plot_convergence.py
     echo "Java placement executed. Check 'logger.txt' for output."
 fi
 
 # Return to outer dir
 cd $PROJ_DIR
 
-# Render placement video
-if [ "$start_stage" == "video" ] || [ "$start_stage" == "all" ]; then
+# Render placement graphics
+if [ "$start_stage" == "graphics" ] || [ "$start_stage" == "all" ]; then
     cd $PROJ_DIR
     FPS=10
     # INPUT="outputs/placers/PlacerAnnealRandom/graphics/images"

### 2025-04-04 latex/paper/paper.tex
carry chains

@@
 
-    The mid-1980s saw the introduction of the homogeneous FPGA, which was built as a grid of CLBs. 
+    The mid-1980s saw the introduction of homogeneous FPGAs, which were built as a grid of CLBs. 
     Rather than using a central programmable switch matrix as in CPLDs, FPGAs adopted an island style architecture in which each CLB is surrounded on all sides by programmable routing resources, as shown in Figure \ref{fig:fpga}.
@@
     \subsection{PrePacking}
+
         In the prepacking stage, we traverse the raw EDIF Netlist to identify common recurring Cell patterns. 
@@
         There are also some cell configurations that must necessarily be placed in certain ways in relation to each other to ensure legality within the architecture constraints.
-        First, we traverse the entire EDIF netlist and identify all unique Cell types and group them together via a {\tt HashMap<String, List<EDIFHierCellInst>>}, where String is the name cell type. 
-        Cell type groups can include {\tt CARRY4}, {\tt LUT1-6}, {\tt FF}, {\tt DSP48E1}, {\tt RAMB18E1}, and others. 
-        We specifically look for CARRY chains, DSP cascades, and LUT-FF pairs as these patterns are common to nearly all FPGA designs.
+
+        Furthermore, certain cell patterns are specific to certain cell types. 
+        Therefore, we first traverse the entire EDIF netlist and identify all unique cell types and group them together via a {\tt HashMap<String, List<EDIFHierCellInst>>}, where String is the name cell type and {\tt EDIFHierCellInst} is the cell class.
+
+        Cell groups can include {\tt CARRY4}, {\tt LUT1-6}, {\tt FF}, {\tt DSP48E1}, {\tt RAMB18E1}, and others. 
+        We specifically look for CARRY4 chains, DSP48E1 cascades, and LUT-FF pairs as these patterns are common to nearly all FPGA designs.
 
@@
                 \caption{
-                    Left: A CARRY4 EDIFCell as seen in the Vivado netlist viewer.
-                    Right: The corresponding CARRY4 Cell as seen in the device viewer.
+                    \textbf{Left}: A CARRY4 EDIFCell as seen in the Vivado netlist viewer.
+                    \textbf{Right}: The corresponding CARRY4 Cell as seen in the device viewer.
                 }
@@
 
+            Figure \ref{fig:carry_cell_edif} shows an example of a CARRY4 cell while figure \ref{fig:carry_chain_edif} shows an example of a CARRY4 chain of size 6.
+
+            As the name implies, CARRY chains are chains of CARRY cells. 
+            Very often an HDL design will implement many adders, counters, subtractors, or comparators, all operations based on binary addition. 
+            They are so ubiquitous that every that in the 7-Series architecture, every CLB comes embedded with a CARRY4 BEL, a 4-bit carry-lookahead (CLA) adder that can be chained together across CLBs to implement wider adders.
+            On the Ultrascale architecture, the each CLB contains a CARRY8 BEL, an 8-bit CLA adder, allowing for even greater logic density.
+
+            (By now Need to have introduced SLICEL / SLICEM Sites and differentiated them from "CLBs", which are referred to as CLBL or CLBM Tiles)
+            (rip some figures from the 7-series CLB user guide)
+
+            These CARRY4 cells, if chained together, must be placed vertically across CLBs, as each CLB is connected vertically by Carry In (CI)
+            This allows the facilitation of fast adders on the device.
+
+            The CARRY4 
+            We identify CARRY4 chains by taking the CARRY4 entry of the HashMap. 
+            We follow the below pseudocode
+
+            Carry chain cells are primitive elements that are provided with a group of LUTs to enable more efficient programmable arithmetic. 
+            Primarily it provides dedicated paths for the carry logic of simple arithmetic operations (add, subtract, comparisons, equals, etc).
+Implementing these arithmetic operations with raw LUTs would results in an inefficient use of resources and performance would suffer.
+            In this way, the CARRY4 cell is like a macrocell.
+
+            On the device, they must necessarily be arranged consecutively vertically, from bottom to top. 
+            Figure \ref{fig:carry_chain_edif} shows an example of a CARRY4 EDIFCell chain in an EDIFNetlist. 
+            Notice how there are many different types of cells connected to the CARRY4 Cells. 
+            The raw netlist does not tell us the presence of any carry chains or other cell clusterings. 
+            We must manually traverse the netlist to identify all of the carry chains by accessing each carry cell, accessing its Ports, accessing the net on the port, accessing all of the other ports on the net, and checking if those ports belong to another carry cell.
+            We continue to traverse the carry chain until the final carry cell is not connected to another carry cell. 
+            We must traverse in both directions. 
+            After we reach the tail of the current chain, we traverse in the opposite direction (CIN vs COUT) to find the carry chain anchor. 
+            We then remove all the cells in this chain from the set of all carry cells. 
+            We select the next carry cell in the set of all carry cells and repeat the process, until there are no more remaining carry cells in the set of all carry cells.
+
+            Shown in figure \ref{fig:carry_cell_edif}
+
             To find CARRY chains, we take the CARRY4 group and follow the below pseudocode. 
-
             Each SLICE Site contains one CARRY4 BEL. 
@@
             CARRY chains span vertically across multiple SLICE Sites/Tiles. (Show picture)
+
             \begin{figure}
@@
                 \includegraphics[width=3.0cm]{figures/dsp_cascade_routes.png}
-                \caption{Left: A Carry Chain of length 6 spanning 6 Sites across 6 Tiles. On the left, the cell-only representation. Right: A DSP cascade of length 2 spanning 2 Sites across 1 Tile.}
+                \caption{\textbf{Left}: A Carry Chain of length 6 spanning 6 Sites across 6 Tiles. \textbf{Right}: A DSP cascade of length 2 spanning 2 Sites across 1 Tile.}
                 \label{fig:device_carry_chain_routing}
@@
         \subsection{LUT-FF Pairs}
-            Identify unique CE-SR net pairs. All FF cells in a Site must share the same CE and SR nets.
+            Identify unique CE-SR net pairs. 
+            All FF cells in a Site must share the same CE and SR nets.
+
     \subsection{Packing}
-        In the packing stage, we take the identified cell clusters and pack them into SiteInst Design objects which target the Device Site objects.
+        In the packing stage, we take the identified cell clusters and package them into SiteInst Design objects which target the Device Site objects. 
         \subsection{CLB Sites}
@@
             For SLICEMs, LUT6s can be configured as shallow 32-bit LUTRAMs or "RAMS32".
+
     \subsection{Placement: Simulated Annealing} 
-        Create BEL "fields": CLB, DSP, CLB and DSP Chains, RAMs
+        Bookkeeping: keep track of Sites occupied by single SiteInsts and Sites occupied by SiteInst chains. 
+        Create BEL "fields": CLB, DSP, CLB and DSP Chains, RAMs.
 
@@
 \section{Placement Results}
+    \begin{itemize}
+    \item Random Site Selection:
+    \item Midpoint Site Selection:
+    \item Hybrid: 
+        Random Site selection until progress stagnates, then use Midpoint Site selection. 
+        Randomly choose betwee Random and Midpoint Site selection each iteration. 
+    \item Cooling Schedule: 
+        Greedy. 
+        Geometric with modifiable alpha and intial temperature. 
+    \end{itemize}
+
+\section{Future Work}
+    \begin{itemize}
+    \item Different cooling schedules: linear cooling, logarithmic cooling, piecewise cooling.
+    \item Support for more macros: buffer cells, FIFO, clock generators, etc..
+    \item Support Ultrascale architecture. 
+    \item Better packing scheme. Currently pseudorandom packing. 
+    \item BEL-centric placement over Site-centric placement.
+    \item Analytical placement. Global placement, legalization, detailed placement. 
+    \end{itemize}
+
+\section{Conclusion}
+    \begin{itemize}
+        \item 
+    \end{itemize}
 

### 2025-04-06 latex/paper/paper.tex
more carry

@@
 
+\section{Xilinx 7-Series Architecture}
+
+    \begin{figure}[]
+        \centering
+        \includegraphics[width=9.0cm]{figures/hierarchy.png}
+        \caption{Architecture Hierarchy of a Xilinx FPGA (Ultrascale)}
+        \label{fig:hierarchy}
+    \end{figure}
+
+    Xilinx introduced the 7-Series architecture in 2010 which followed the heterogeneous island-style architecture described previously. 
+    Then in 2013 Xilinx unveiled the Ultrascale architecture which also followed the heterogeneous island-style design and made several improvements over the 7-Series architecture. 
+    Although the latest FPGAs follow the Ultrascale architecture, this paper focuses on the 7-Series FPGAs due to their greater accessibility for beginner users and compatibility with open source tooling.
+
+    % Figure~\ref{fig:hierarchy} illustrates the architectural hierarchy of an Ultrascale Xilinx FPGA. 
+    % The architecture hierarchy is the same for the 7-Series devices. 
+    Figure \ref{fig:hierarchy_3} illustrates the architectural hierarchy of a 7-Series Xilinx FPGA. 
+    At its most basic level, an FPGA is a vast physical array of replicated atomic components known as Basic Elements of Logic (BELs). 
+    These BELs, which include LUTs, FFs, BRAMs, DSPs, and programmable interconnects, form the fundamental building blocks for custom digital circuit implementations on FPGAs.
+
+    The Xilinx architectural hierarchy organizes these BELs into increasingly abstract structures. 
+    \textbf{BELs} are packaged into \textbf{Sites}, which are then embedded into \textbf{Tiles}, which are subsequently consolidated into \textbf{Clock Regions}. 
+    In high-end Xilinx devices, Clock Regions may be further consolidated into \textbf{Super Logic Regions} (SLRs). 
+    However, for the scope of this paper, we focus on Xilinx devices that have only a single SLR.
+
+    \begin{figure}[]
+        \centering
+        \includegraphics[width=9.0cm]{figures/hierarchy_3.png}
+        \caption{Tiles, Sites, BELs hierarchy}
+        \label{fig:hierarchy_3}
+    \end{figure}
+
 \section{The FPGA Design Flow and Toolchain}
@@
 
-    Following placement, the \textbf{routing} tool connects the placed elements by assigning physical wiring resources to the nets in the netlist. 
+    Following placement, the \textbf{routing} tool connects the placed elements by assigning physical wiring resources (switchboxes, programmable interconnects) to the nets in the netlist. 
     The router determines their paths through the programmable interconnect network while attempting to meet timing and performance constraints.
@@
     This integrated toolchain, whether provided by FPGA vendors or open source communities, facilitates the transformation from an abstract HDL description to a functioning hardware configuration.
-
-\section{Xilinx 7-Series Architecture}
-
-    \begin{figure*}[t]
-        \centering
-        \includegraphics[width=12.0cm]{figures/hierarchy.png}
-        \caption{Architecture Hierarchy of a Xilinx FPGA (Ultrascale)}
-        \label{fig:hierarchy}
-    \end{figure*}
-
-    Xilinx introduced the 7-Series architecture in 2010 which followed the heterogeneous island-style architecture described previously. 
-    Then in 2013 Xilinx unveiled the Ultrascale architecture which also followed the heterogeneous island-style design but made several improvements over the 7-Series architecture. 
-    Although the latest FPGAs follow the Ultrascale architecture, this paper focuses on the 7-Series FPGAs due to their greater accessibility for beginner users and compatibility with open source tooling.
-
-    Figure~\ref{fig:hierarchy} illustrates the architectural hierarchy of an Ultrascale Xilinx FPGA. 
-    The architecture hierarchy is the same for the 7-Series devices. 
-    At its most basic level, an FPGA is a vast physical array of replicated atomic components known as Basic Elements of Logic (BELs). 
-    These BELs, which include LUTs, FFs, BRAMs, DSPs, and programmable interconnects, form the fundamental building blocks for custom digital circuit implementations on FPGAs.
-
-    The Xilinx architectural hierarchy organizes these BELs into increasingly abstract structures. 
-    \textbf{BELs} are packaged into \textbf{Sites}, which are then embedded into \textbf{Tiles}, which are subsequently consolidated into \textbf{Clock Regions}. 
-    In high-end Xilinx devices, Clock Regions may be further consolidated into \textbf{Super Logic Regions} (SLRs). 
-    However, for the scope of this paper, we focus on Xilinx devices that have only a single SLR.
 
@@
 
-    For a complete understanding of the RapidWright packages, please refer to the API Overview \url{https://www.rapidwright.io/docs/RapidWright_Overview.html}.
+    For a complete understanding of the RapidWright packages, please refer to the API Overview \url{https://www.rapidwright.io/docs/RapidWright_Overview.html} and the API reference.
 
@@
         Furthermore, certain cell patterns are specific to certain cell types. 
-        Therefore, we first traverse the entire EDIF netlist and identify all unique cell types and group them together via a {\tt HashMap<String, List<EDIFHierCellInst>>}, where String is the name cell type and {\tt EDIFHierCellInst} is the cell class.
+        Therefore, we must first traverse the entire EDIF netlist and identify all unique cell types and group them together via a {\tt HashMap<String, List<EDIFHierCellInst>>}, where key String is the name cell type and value {\tt List<EDIFHierCellInst>} is the list of cells of that type.
+        The resulting lists in the hashmap are mutually exclusive.
 
-        Cell groups can include {\tt CARRY4}, {\tt LUT1-6}, {\tt FF}, {\tt DSP48E1}, {\tt RAMB18E1}, and others. 
+        Cell types can include {\tt CARRY4}, {\tt LUT1-6}, {\tt FF}, {\tt DSP48E1}, {\tt RAMB18E1}, and others. 
         We specifically look for CARRY4 chains, DSP48E1 cascades, and LUT-FF pairs as these patterns are common to nearly all FPGA designs. 
+        We will describe each cell pattern in detail.
 
@@
             They are so ubiquitous that every that in the 7-Series architecture, every CLB comes embedded with a CARRY4 BEL, a 4-bit carry-lookahead (CLA) adder that can be chained together across CLBs to implement wider adders.
-            On the Ultrascale architecture, the each CLB contains a CARRY8 BEL, an 8-bit CLA adder, allowing for even greater logic density.
+            % On the Ultrascale architecture, the each CLB contains a CARRY8 BEL, an 8-bit CLA adder, allowing for even greater logic density.
 
@@
 
-            These CARRY4 cells, if chained together, must be placed vertically across CLBs, as each CLB is connected vertically by Carry In (CI)
+            These CARRY4 cells, if chained together, must be placed vertically across CLBs, as each CLB is connected vertically by a Carry In (CI) to Carry Out (COUT) wire between them. 
+            The only way to route the carry net is through this wire, thus why these CLBs must be placed consecutively vertically. 
             This allows the facilitation of fast adders on the device.
@@
     \item Better packing scheme. Currently pseudorandom packing. 
+    \item Inter-Site BEL swapping. 
     \item BEL-centric placement over Site-centric placement.

### 2025-04-12 latex/paper/paper.tex
o

@@
 
-    \begin{figure}[]
+    % \begin{figure}[]
+    %     \centering
+    %     \includegraphics[width=9.0cm]{figures/hierarchy.png}
+    %     \caption{Architecture Hierarchy of a Xilinx FPGA (Ultrascale)}
+    %     \label{fig:hierarchy}
+    % \end{figure}
+
+    \begin{figure*}[]
         \centering
-        \includegraphics[width=9.0cm]{figures/hierarchy.png}
-        \caption{Architecture Hierarchy of a Xilinx FPGA (Ultrascale)}
-        \label{fig:hierarchy}
-    \end{figure}
+        \includegraphics[width=15.0cm]{figures/hierarchy_3.png}
+        \caption{Architecture Hierarchy of a Xilinx FPGA}
+        \label{fig:hierarchy_4}
+    \end{figure*}
 
@@
     % The architecture hierarchy is the same for the 7-Series devices. 
-    Figure \ref{fig:hierarchy_3} illustrates the architectural hierarchy of a 7-Series Xilinx FPGA. 
+    Figure \ref{fig:hierarchy_4} illustrates the architectural hierarchy of a 7-Series Xilinx FPGA. 
     At its most basic level, an FPGA is a vast physical array of replicated atomic components known as Basic Elements of Logic (BELs). 
@@
 
-    \begin{figure}[]
-        \centering
-        \includegraphics[width=9.0cm]{figures/hierarchy_3.png}
-        \caption{Tiles, Sites, BELs hierarchy}
-        \label{fig:hierarchy_3}
-    \end{figure}
+    Note that in the Xilinx architectures, "CLBs" refer to CLB Tiles, each containing two "SLICE" Sites, with each SLICE Site containing 8 LUTs, 8 Flip Flop, and 1 CARRY BEL.
+    These SLICEs facilitate the core reprogrammability of the FPGA.
 
@@
         \centering
-        \includegraphics[width=7.0cm]{figures/design_flow.png}
+        \includegraphics[width=8.0cm]{figures/design_flow.png}
         \caption{A typical FPGA design and verification workflow}
@@
 
-    Figure \ref{fig:design_flow} shows a typical FPGA design and verification workflow.
-    The FPGA design flow takes a high-level digital design written in an HDL and generates a configuration file that programs the physical FPGA.
+    Figure \ref{fig:design_flow} illustrates a typical FPGA design and verification workflow, transforming a high-level digital design into a configuration file that programs the physical FPGA.
 
-    The process begins with \textbf{design entry}, where an engineer describes the functionality of their circuit using a hardware description language (HDL) such as Verilog or VHDL.
-This description abstracts away details of the physical hardware while defining the logic and behavior of the design.
+    The process begins with \textbf{design entry}, where an engineer describes the desired functionality using a hardware description language (HDL) such as Verilog or VHDL. 
+    Engineers may use various coding styles - behavioral, structural, dataflow, or a mix of all three - to capture the logic and behavior of the design while abstracting away the intricacies of the physical hardware architecture.
 
-    Next, the \textbf{synthesis} tool interprets the HDL code and generates a netlist that represents the design in terms of primitive cells, such as lookup tables (LUTs) and flip flops. 
-    This is the result of the synthesis tool parsing the raw Verilog source files, inferring logical behavior, and instantiating primitive cells to emulate the desired logical behavior. 
-    The final stage is often referred to as "technology mapping", as the primitive cells are specific to the FPGA architecture and consequently the FPGA manufacturer.
+    Next, the user runs the \textbf{synthesis} tool, which parses the raw HDL source files to infer logical behavior, perform optimizations, and instantiate primitive cells. 
+    The result is a detailed netlist, typically output as a low-level Verilog or Electronic Design Interchange Format (EDIF) file, where the design is represented purely in structural HDL. 
+    The final stage is often referred to as "technology mapping", as the vocabulary or library of primitive cells must be specific to the FPGA architecture and consequently the FPGA manufacturer.
 
@@
 
-    Placement and routing are generally referred to together as the \textbf{implementation} stage, as both stages are interdependent and must adhere to the constraints of the same device. 
+    Placement and routing are generally referred jointly as the \textbf{implementation} stage, as both stages are interdependent and must adhere to the constraints of the same device. 
     State of the art (SOTA) tools will perform routing-aware placement for greater optimization.
@@
         There are also some cell configurations that must necessarily be placed in certain ways in relation to each other to ensure legality within the architecture constraints.
+
+        \begin{figure}
+            \centering
+            \includegraphics[width=4.0cm]{figures/edif_net.png}
+            \includegraphics[width=4.0cm]{figures/intrasite_route.png}
+            \includegraphics[width=8.0cm]{figures/intersite_route.png}
+            \caption{
+                \textbf{Upper Left}: A LUT-FF cell pair in the netlist.
+                \textbf{Upper Right}: The LUT-FF pair placed within the same Site.
+                \textbf{Bottom}: The LUT-FF pair placed across different Sites.
+            }
+            \label{fig:intrasite_intersite}
+        \end{figure}
+
+        For example, in many HDL designs, LUTs and FFs often come in pairs, as is the case in Figure \ref{fig:intrasite_intersite}
 

### 2025-04-12 latex/paper/paper.tex
need to rethink how to explain arch BEL/Site constraints

@@
 
-    % \begin{figure}[]
-    %     \centering
-    %     \includegraphics[width=9.0cm]{figures/hierarchy.png}
-    %     \caption{Architecture Hierarchy of a Xilinx FPGA (Ultrascale)}
-    %     \label{fig:hierarchy}
-    % \end{figure}
-
     \begin{figure*}[]
@@
 
-    Xilinx introduced the 7-Series architecture in 2010 which followed the heterogeneous island-style architecture described previously. 
-    Then in 2013 Xilinx unveiled the Ultrascale architecture which also followed the heterogeneous island-style design and made several improvements over the 7-Series architecture. 
-    Although the latest FPGAs follow the Ultrascale architecture, this paper focuses on the 7-Series FPGAs due to their greater accessibility for beginner users and compatibility with open source tooling.
+    In this paper we will be working with the Xilinx 7-Series devices which follows the heterogenous island-style architecture. 
+    The 7-Series architecture first released in 2010 and was superceded by the Ultrascale architecture which made several improvements over the 7-Series. 
+    Although the latest FPGAs follow the Ultrascale architecture, this paper focuses on the 7-Series FPGAs due to their greater accessibility to beginner users and compatibility with open source tooling.
 
@@
     With a basic understanding of FPGA architecture, design placement, and RapidWright, we have all the necessary pieces to implement our SA placer. 
-    Here we outline in detail each substage of our implementation.
+    Here we outline in detail each substage of our implementation: PrePacking, Packing, and Placement.
 
@@
 
-        In the prepacking stage, we traverse the raw EDIF Netlist to identify common recurring Cell patterns. 
+        In the prepacking stage, we traverse the raw post-synthesis EDIF Netlist to identify common recurring Cell patterns. 
         We do this because there are certain desirable cell configurations that should be clustered together to exploit some proximity characteristics of the FPGA architecture. 
+        For example, in Figure \ref{fig:intrasite_intersite}
         Doing so will innately improve wirelength minimization. 
@@
             \centering
-            \includegraphics[width=4.0cm]{figures/edif_net.png}
-            \includegraphics[width=4.0cm]{figures/intrasite_route.png}
-            \includegraphics[width=8.0cm]{figures/intersite_route.png}
+            \includegraphics[width=8.0cm]{figures/intersite_intrasite.png}
             \caption{

### 2025-04-12 latex/paper/paper.tex
lutffpair

@@
     In this paper we will be working with the Xilinx 7-Series devices which follows the heterogenous island-style architecture. 
-    The 7-Series architecture first released in 2010 and was superceded by the Ultrascale architecture which made several improvements over the 7-Series. 
+    The 7-Series architecture first released in 2010 and was superceded in 2013 by the Ultrascale architecture which made several improvements over the 7-Series. 
     Although the latest FPGAs follow the Ultrascale architecture, this paper focuses on the 7-Series FPGAs due to their greater accessibility to beginner users and compatibility with open source tooling.
 
-    % Figure~\ref{fig:hierarchy} illustrates the architectural hierarchy of an Ultrascale Xilinx FPGA. 
-    % The architecture hierarchy is the same for the 7-Series devices. 
     Figure \ref{fig:hierarchy_4} illustrates the architectural hierarchy of a 7-Series Xilinx FPGA. 
@@
     Note that in the Xilinx architectures, "CLBs" refer to CLB Tiles, each containing two "SLICE" Sites, with each SLICE Site containing 8 LUTs, 8 Flip Flop, and 1 CARRY BEL. 
-    These SLICEs facilitate the core reprogrammability of the FPGA.
+    These SLICE Sites facilitate the core programmable logic of the FPGA.
 
@@
 \section{Placement Implementation}
-
     With a basic understanding of FPGA architecture, design placement, and RapidWright, we have all the necessary pieces to implement our SA placer. 
@@
     \subsection{PrePacking}
-
         In the prepacking stage, we traverse the raw post-synthesis EDIF Netlist to identify common recurring Cell patterns. 
-        We do this because there are certain desirable cell configurations that should be clustered together to exploit some proximity characteristics of the FPGA architecture. 
-        For example, in Figure \ref{fig:intrasite_intersite}
+        We do this because there are certain desirable cell configurations that should be clustered together to exploit the physical organization of the BELs and Sites of the architecture. 
         Doing so will innately improve wirelength minimization. 
@@
 
-        \begin{figure}
+        \subsection{LUT-FF Pairs}
+            In the vast majority of HDL designs, LUTs and FFs will be synthesized in pairs, as is the case in Figure \ref{fig:lut_ff_pair}.
+            Shown in Figure \ref{fig:intrasite_intersite} are two possible placements for this cell net. 
+            On the right, the cells are placed across different Sites, thus the only way to route the net between the cells is through general inter-site routing. 
+            On the bottom left, the cells are placed within the same Site, taking advantage of the intra-site routing and avoiding costly inter-site routing.
+
+            Identify unique CE-SR net pairs. 
+            All FF cells in a Site must share the same CE and SR nets.
+
+        \begin{figure}[H]
+            \centering
+            \includegraphics[width=8.0cm]{figures/lut_ff_pair.png}
+            \caption{
+                Example of a LUT-FF Cell Pair as shown the Vivado netlist viewer. 
+                FDRE is a type of Flip Flop. 
+            }
+            \label{fig:lut_ff_pair}
+        \end{figure}
+
+        \begin{figure}[H]
             \centering
@@
         \end{figure}
-
-        For example, in many HDL designs, LUTs and FFs often come in pairs, as is the case in Figure \ref{fig:intrasite_intersite}
 
@@
 
-        \subsection{LUT-FF Pairs}
-            Identify unique CE-SR net pairs. 
-            All FF cells in a Site must share the same CE and SR nets.
-
     \subsection{Packing}

### 2025-04-14 latex/paper/paper.tex
more lutff

@@
 
-    Here we will briefly outline the evolution of configurable logic architecture as well as the FPGA design flow. 
+    First we will briefly outline the evolution of configurable logic architecture as well as the FPGA design flow. 
     However, readers who are already familiar with basic FPGA architecture and design flow may skip to the next section on Xilinx 7-Series Architecture. 
@@
         % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/FPGA/Field-programmable-gate-array-FPGA/
-        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/FPGA/Field-programmable-gate-array-FPGA/
     \end{figure}
@@
     Major vendors such as Xilinx and Altera now employ heterogeneous island-style architectures in their devices. 
-    As designs become increasingly large and complex, FPGAs meet the demand by becoming increasingly heterogenous by incorporating a wider variety of hard macros into the fabric.
+    As designs become increasingly large and complex, FPGAs meet the demand by becoming increasingly heterogenous, incorporating a wider variety of hard macros into the fabric.
 
@@
 
-    In this paper we will be working with the Xilinx 7-Series devices which follows the heterogenous island-style architecture. 
-    The 7-Series architecture first released in 2010 and was superceded in 2013 by the Ultrascale architecture which made several improvements over the 7-Series. 
-    Although the latest FPGAs follow the Ultrascale architecture, this paper focuses on the 7-Series FPGAs due to their greater accessibility to beginner users and compatibility with open source tooling.
+    In this paper we will be working with the Xilinx 7-Series architecture which follows a heterogenous island-style design. 
+    The 7-Series architecture first released in 2010 and was then superceded in 2013 by the Ultrascale architecture which offered several improvements over the 7-Series. 
+    Although the latest FPGAs follow the Ultrascale architecture, this paper focuses on the 7-Series FPGAs due to their greater accessibility to hobbyists and compatibility with open source tooling.
 
     Figure \ref{fig:hierarchy_4} illustrates the architectural hierarchy of a 7-Series Xilinx FPGA. 
-    At its most basic level, an FPGA is a vast physical array of replicated atomic components known as Basic Elements of Logic (BELs). 
+    At its most basic level, an FPGA is a vast physical array of atomic components known as Basic Elements of Logic (BELs). 
     These BELs, which include LUTs, FFs, BRAMs, DSPs, and programmable interconnects, form the fundamental building blocks for custom digital circuit implementations on FPGAs.
@@
 
-    Note that in the Xilinx architectures, "CLBs" refer to CLB Tiles, each containing two "SLICE" Sites, with each SLICE Site containing 8 LUTs, 8 Flip Flop, and 1 CARRY BEL. 
-    These SLICE Sites facilitate the core programmable logic of the FPGA.
+    Note that in the Xilinx architectures, "CLBs" refer to CLB Tiles, each containing two "SLICE" Sites, with each SLICE Site containing a collection of BELs - 8 LUTs, 8 Flip Flops, and 1 CARRY4.
+
+    The LUTs and Flip Flops constitute the core of the FPGA programmable logic.
+
+    Flip Flops (FFs) facilitate synchronous (clock-driven or event-driven) registers.
+
+    SLICE Sites come in two flavors - SLICEL (SLICE-Logic) and SLICEM (SLICE-Memory). 
+    In the SLICEL, Lookup Tables (LUTs), as their name implies, are fast shallow read-only memories (ROMs) with 6 input pins and 1 or 2 output pins. 
+    All combinational logic - boolean functions - are mapped onto these LUTs as read-only-memories during synthesis. 
+    Larger boolean functions with more than 6 inputs will be mapped onto multiple cascaded LUTs.
+
+    The SLICEM is similar to the SLICEL, the key difference being that their LUTs can optionally be programmed as RAMs to facilitate "Distributed RAM", an alternative to the larger dedicated Block RAM (BRAM) macro.
+    Xilinx devices usually come a 25\% SLICEM, 75\% SLICEL split amongst all SLICE Sites, where a CLB Tile can contain either two SLICELs or one SLICEL one SLICEM.
+
+    Each SLICE also comes with a shallow 4-bit carry-lookahead adder (CLA) BEL called "CARRY4".
 
 \section{The FPGA Design Flow and Toolchain}
-
     \begin{figure}
@@
 
+    In addition to the FPGA hardware architecture, one should also understand the role that placement plays in the borader FPGA design flow and software toolchain.
+
     Figure \ref{fig:design_flow} illustrates a typical FPGA design and verification workflow, transforming a high-level digital design into a configuration file that programs the physical FPGA.
 
-    The process begins with \textbf{design entry}, where an engineer describes the desired functionality using a hardware description language (HDL) such as Verilog or VHDL. 
+    The process begins with \textbf{design entry}, where an engineer describes the desired functionality of a digital system using a hardware description language (HDL) such as Verilog or VHDL. 
     Engineers may use various coding styles - behavioral, structural, dataflow, or a mix of all three - to capture the logic and behavior of the design while abstracting away the intricacies of the physical hardware architecture.
@@
 
+        Furthermore, certain cell patterns are specific to certain cell types. 
+        Therefore, we must first traverse the entire EDIF netlist and identify all unique cell types and group them together via a {\tt HashMap<String, List<EDIFHierCellInst>>}, where key String is the name cell type and value {\tt List<EDIFHierCellInst>} is the list of cells of that type.
+        The resulting lists in the hashmap are mutually exclusive.
+
+        Cell types can include {\tt CARRY4}, {\tt LUT1-6}, {\tt FF}, {\tt DSP48E1}, {\tt RAMB18E1}, and others. 
+        We specifically look for CARRY4 chains, DSP48E1 cascades, and LUT-FF pairs as these patterns are common to nearly all FPGA designs. 
+        We will describe each cell pattern in detail.
+
         \subsection{LUT-FF Pairs}
+            Very often, engineers model the behavior of a digital system as a collection of Finite State Machines (FSM) as shown in Figure \ref{fig:fsm}. 
+            An FSM at its core is a synchronous sequential circuit comprising of a combinational element and a synchronous memory element.
+            In the 7-Series architecture, combinational elements are synthesized as LUTs, where the truth table of the combinational function is mapped onto the LUT's memory.
+
+            Combinational functions in 7-Series FPGAs are synthesized as Look-Up Tables (LUTs). 
+            The FPGA synthesizer maps the combinational element into LUTs and the memory element onto FFs.
+
+            \begin{figure}[H]
+                \centering
+                \includegraphics[width=8.0cm]{figures/fsm.png}
+                \caption{A Finite State Machine (Moore)}
+                \label{fig:fsm}
+            \end{figure}
+
             In the vast majority of HDL designs, LUTs and FFs will be synthesized in pairs, as is the case in Figure \ref{fig:lut_ff_pair}.
@@
                 \centering
-            \includegraphics[width=8.0cm]{figures/intersite_intrasite.png}
+                \includegraphics[width=9.0cm]{figures/intersite_intrasite_2.png}
                 \caption{
@@
             \end{figure}
-
-        Furthermore, certain cell patterns are specific to certain cell types. 
-        Therefore, we must first traverse the entire EDIF netlist and identify all unique cell types and group them together via a {\tt HashMap<String, List<EDIFHierCellInst>>}, where key String is the name cell type and value {\tt List<EDIFHierCellInst>} is the list of cells of that type.
-        The resulting lists in the hashmap are mutually exclusive.
-
-        Cell types can include {\tt CARRY4}, {\tt LUT1-6}, {\tt FF}, {\tt DSP48E1}, {\tt RAMB18E1}, and others. 
-        We specifically look for CARRY4 chains, DSP48E1 cascades, and LUT-FF pairs as these patterns are common to nearly all FPGA designs. 
-        We will describe each cell pattern in detail.
 
@@
 
+    \begin{figure*}[b]
+        \centering
+        \includegraphics[width=\textwidth]{figures/slicem.png}
+        \caption{SLICEM Site.}
+    \end{figure}
+
+    \begin{figure*}[t]
+        \centering
+        \includegraphics[width=\textwidth]{figures/slicel.png}
+        \caption{SLICEL Site.}
+    \end{figure}
+
 \end{document}

### 2025-04-15 latex/paper/paper.tex
,

@@
 
+    The paper first begins by elaborating on general FPGA architecture and then specifically the Xilinx 7-Series architecture. 
+    Then, the paper will elaborate on the FPGA design flow, then the role that the RapidWright API plays in the design flow. 
+    We explain in detail each of these concepts for a broader audience as they provide much needed context for FPGA placement algorithms as a concept. 
+    However, readers who are already familiar with these concepts can skip directly to the RapidWright API section.
+
 \section{A Brief History on FPGA Architecture}
@@
 
-    First we will briefly outline the evolution of configurable logic architecture as well as the FPGA design flow. 
-    However, readers who are already familiar with basic FPGA architecture and design flow may skip to the next section on Xilinx 7-Series Architecture. 
+    First we will briefly review the evolution of configurable logic architecture as well as general FPGA design flow. 
+    % However, readers who are already familiar with basic FPGA architecture and design flow may skip to the next section on Xilinx 7-Series Architecture. 
     Configurable logic devices have undergone significant evolution over the past four decades. 
@@
 
-    In this paper we will be working with the Xilinx 7-Series architecture which follows a heterogenous island-style design. 
+    In this paper we will be working with the Xilinx 7-Series devices which follows a heterogenous island-style architecture. 
     The 7-Series architecture first released in 2010 and was then superceded in 2013 by the Ultrascale architecture which offered several improvements over the 7-Series. 
@@
 
-    Note that in the Xilinx architectures, "CLBs" refer to CLB Tiles, each containing two "SLICE" Sites, with each SLICE Site containing a collection of BELs - 8 LUTs, 8 Flip Flops, and 1 CARRY4.
+    In the 7-Series architecture, "CLBs" refer to CLB Tiles, each containing two "SLICE" Sites, with each SLICE Site containing a collection of BELs - 8 LUTs, 8 Flip Flops, and 1 CARRY4.
 
@@
     SLICE Sites come in two flavors - SLICEL (SLICE-Logic) and SLICEM (SLICE-Memory). 
-    In the SLICEL, Lookup Tables (LUTs), as their name implies, are fast shallow read-only memories (ROMs) with 6 input pins and 1 or 2 output pins. 
+    In the SLICEL, Lookup Tables (LUTs) are fast shallow read-only memories (ROMs) with 6 input pins and 1 or 2 output pins. 
     All combinational logic - boolean functions - are mapped onto these LUTs as read-only-memories during synthesis. 
@@
 
-    The SLICEM is similar to the SLICEL, the key difference being that their LUTs can optionally be programmed as RAMs to facilitate "Distributed RAM", an alternative to the larger dedicated Block RAM (BRAM) macro.
-    Xilinx devices usually come a 25\% SLICEM, 75\% SLICEL split amongst all SLICE Sites, where a CLB Tile can contain either two SLICELs or one SLICEL one SLICEM.
+    The SLICEM is similar to the SLICEL, with the key difference that their LUTs can optionally be programmed as RAMs to facilitate "Distributed RAM", an alternative to the larger dedicated Block RAM (BRAM) macro.
+
+    Shown in Figure \ref{fig:slicel} and Figure \ref{fig:slicem} are a SLICEL and SLICEM respectively as seen in the Vivado device viewer.
+
+    Xilinx devices usually come a 25\% SLICEM, 75\% SLICEL split amongst all SLICE Sites, where a CLB Tile can contain either two SLICELs or one SLICEL plus one SLICEM.
 
@@
 
-    In addition to the FPGA hardware architecture, one should also understand the role that placement plays in the borader FPGA design flow and software toolchain.
+    In addition ot the FPGA architecture, one should also understand the role that placement plays in the broader FPGA design flow and software toolchain.
 
@@
     Our placer will only involve prepacking, packing, and detailed placement for simplicity. 
-    The objective is to place the logical cell instances of the netlist onto physical BELs on the device in a way that minimizes delay, satisfies design constraints, and simplifies routing.
+    The objective is to place the logical cell instances of the netlist onto physical BELs on the device in a way that minimizes wirelength, simplifies routing, and satisfies timing constraints (signal propagation delay).
 
@@
     Placement and routing are generally referred jointly as the \textbf{implementation} stage, as both stages are interdependent and must adhere to the constraints of the same device. 
-    State of the art (SOTA) tools will perform routing-aware placement for greater optimization.
+    State of the art (SOTA) tools will perform routing-aware placement or placement-fluid routing for greater optimization.
 
-    The final stage is the \textbf{bitstream} generation, where a vendor-specific programming tool produces a configuration file or bitstream that programs the physical FPGA device. 
+    The final stage is the \textbf{bitstream} generation, where a vendor-specific programming tool takes the implemented design and generates a configuration file or bitstream that programs the physical FPGA device.
     The bitstream specifies the state of every configurable element—logic blocks, routing switches, and hard macros—ensuring the FPGA implements the user's design.
@@
 
-\section{RapidWright API}
-    We use the RapidWright API to implement a custom placer for Xilinx 7-Series FPGAs.
+\section{Placement Implementation via RapidWright API}
+
+    %%%
+
+    The RapidWright API offers an additional workflow column in parallel to both the design flow and verification flow. 
     RapidWright is an open source Java framework that enables netlist implementation manipulation for modern Xilinx FPGAs.
@@
 
-\section{Placement Implementation}
     With a basic understanding of FPGA architecture, design placement, and RapidWright, we have all the necessary pieces to implement our SA placer. 
@@
 \section{Conclusion}
-    \begin{itemize}
-        \item 
-    \end{itemize}
+    fdsfdsafdsa
 
@@
 
+\section{Appendix}
+
     \begin{figure*}[b]
@@
         \includegraphics[width=\textwidth]{figures/slicem.png}
-        \caption{SLICEM Site.}
-    \end{figure}
+        \caption{SLICEM Site}
+        \label{slicem}
+    \end{figure*}
 
@@
         \includegraphics[width=\textwidth]{figures/slicel.png}
-        \caption{SLICEL Site.}
-    \end{figure}
+        \caption{SLICEL Site}
+        \label{slicel}
+    \end{figure*}
 

### 2025-04-16 latex/paper/paper.tex
rewrite sa

@@
     Before any work can begin on an FPGA placer, it is necessary to understand both the objects being placed and the medium in which they are placed.
-
-    First we will briefly review the evolution of configurable logic architecture as well as general FPGA design flow. 
-    % However, readers who are already familiar with basic FPGA architecture and design flow may skip to the next section on Xilinx 7-Series Architecture. 
     Configurable logic devices have undergone significant evolution over the past four decades. 
+    We will briefly review the evolution of configurable logic architecture starting in the 1970s and quickly work our way up to modern day FPGA architecture. 
     The journey began with the Programmable Logic Array (PLA) in the early 1970s. 
@@
 
-    \begin{figure*}[]
+    \begin{figure*}[!ht]
         \centering
@@
 
-    In this paper we will be working with the Xilinx 7-Series devices which follows a heterogenous island-style architecture. 
-    The 7-Series architecture first released in 2010 and was then superceded in 2013 by the Ultrascale architecture which offered several improvements over the 7-Series. 
-    Although the latest FPGAs follow the Ultrascale architecture, this paper focuses on the 7-Series FPGAs due to their greater accessibility to hobbyists and compatibility with open source tooling.
+    The Xilinx 7-Series devices, first introduced in 2010, follow a heterogeneous island-style architecture as discussed previously. 
+    Although the 7-Series was later superseded in 2013 by the UltraScale architecture, the 7-Series remains highly relevant due to its accessibility, wide availability, and compatibility with open-source tooling.
+    Representative sub-families include Artix-7, Kintex-7, Virtex-7, and Zynq-7000, each designed with different performance and cost trade-offs but all follow the core 7-Series architecture.
 
-    Figure \ref{fig:hierarchy_4} illustrates the architectural hierarchy of a 7-Series Xilinx FPGA. 
-    At its most basic level, an FPGA is a vast physical array of atomic components known as Basic Elements of Logic (BELs). 
-    These BELs, which include LUTs, FFs, BRAMs, DSPs, and programmable interconnects, form the fundamental building blocks for custom digital circuit implementations on FPGAs.
+    Figure~\ref{fig:hierarchy_4} illustrates a high-level view of the hierarchical organization of a 7-Series FPGA. 
+    At its lowest level, the device consists of a large array of atomic components called \emph{Basic Elements of Logic}~(BELs). 
+    These BELs encompass look-up tables (LUTs), flip-flops (FFs), block RAMs (BRAMs), DSP slices (DSPs), and the configurable interconnect fabric. 
+    They constitute the fundamental building blocks for implementing digital circuits on the FPGA.
 
-    The Xilinx architectural hierarchy organizes these BELs into increasingly abstract structures. 
-    \textbf{BELs} are packaged into \textbf{Sites}, which are then embedded into \textbf{Tiles}, which are subsequently consolidated into \textbf{Clock Regions}. 
-    In high-end Xilinx devices, Clock Regions may be further consolidated into \textbf{Super Logic Regions} (SLRs). 
-    However, for the scope of this paper, we focus on Xilinx devices that have only a single SLR.
+    To manage this complexity, Xilinx organizes these BELs into incrementally abstract structures. 
+    First, \textbf{BELs} are grouped into \textbf{Sites}. 
+    Each Site is embedded into a \textbf{Tile}, and Tiles are further arranged into \textbf{Clock Regions}. 
+    In some high-density devices, multiple Clock Regions may be consolidated into one or more \textbf{Super Logic Regions}~(SLRs). 
+    However, for the scope of this paper, we focus on Xilinx 7-Series devices with a single SLR.
 
-    In the 7-Series architecture, "CLBs" refer to CLB Tiles, each containing two "SLICE" Sites, with each SLICE Site containing a collection of BELs - 8 LUTs, 8 Flip Flops, and 1 CARRY4.
+    In the 7-Series architecture, the term \emph{CLB} (Configurable Logic Block) refers to a \emph{CLB Tile} that contains two \emph{SLICE} Sites. 
+    Each SLICE Site has a set of BELs including eight LUTs, eight FFs, and one \emph{CARRY4} adder. 
+    These SLICEs provide the core logic and arithmetic functionality of the device.
 
-    The LUTs and Flip Flops constitute the core of the FPGA programmable logic.
+    \textbf{SLICE Types.}\quad 
+    Xilinx offers two variants of SLICE Sites: \textbf{SLICEL} and \textbf{SLICEM}.  
+    \begin{itemize}
+    \item \emph{SLICEL (SLICE-Logic):} Each LUT in a SLICEL can implement a 6-input Boolean function (with up to two outputs under certain input-sharing conditions).
+During logic synthesis, larger Boolean expressions exceeding six inputs are automatically split across multiple LUTs.
+    \item \emph{SLICEM (SLICE-Memory):} SLICEM LUTs include all the features of a SLICEL but can also be configured as small distributed RAM (known as \emph{Distributed RAM}) or as shift registers.
+This local memory capability supplements the larger Block RAM (BRAM) cells and is beneficial for designs requiring numerous small RAM blocks.
+    \end{itemize}
 
-    Flip Flops (FFs) facilitate synchronous (clock-driven or event-driven) registers.
-
-    SLICE Sites come in two flavors - SLICEL (SLICE-Logic) and SLICEM (SLICE-Memory). 
-    In the SLICEL, Lookup Tables (LUTs) are fast shallow read-only memories (ROMs) with 6 input pins and 1 or 2 output pins. 
-    All combinational logic - boolean functions - are mapped onto these LUTs as read-only-memories during synthesis. 
-    Larger boolean functions with more than 6 inputs will be mapped onto multiple cascaded LUTs.
-
-    The SLICEM is similar to the SLICEL, with the key difference that their LUTs can optionally be programmed as RAMs to facilitate "Distributed RAM", an alternative to the larger dedicated Block RAM (BRAM) macro.
-
-    Shown in Figure \ref{fig:slicel} and Figure \ref{fig:slicem} are a SLICEL and SLICEM respectively as seen in the Vivado device viewer.
-
-    Xilinx devices usually come a 25\% SLICEM, 75\% SLICEL split amongst all SLICE Sites, where a CLB Tile can contain either two SLICELs or one SLICEL plus one SLICEM.
-
-    Each SLICE also comes with a shallow 4-bit carry-lookahead adder (CLA) BEL called "CARRY4".
-
-\section{The FPGA Design Flow and Toolchain}
-    \begin{figure}
+    \begin{figure}[!ht]
         \centering
-        \includegraphics[width=8.0cm]{figures/design_flow.png}
-        \caption{A typical FPGA design and verification workflow}
-        \label{fig:design_flow}
+        \includegraphics[width=7.5cm]{figures/slicel.png}
+        \caption{A SLICEL Site in the Vivado device viewer}
+        \label{fig:slicel}
     \end{figure}
 
-    In addition ot the FPGA architecture, one should also understand the role that placement plays in the broader FPGA design flow and software toolchain.
+    \begin{figure}[!ht]
+        \centering
+        \includegraphics[width=7.5cm]{figures/slicem.png}
+        \caption{A SLICEM Site in the Vivado device viewer}
+        \label{fig:slicem}
+    \end{figure}
 
-    Figure \ref{fig:design_flow} illustrates a typical FPGA design and verification workflow, transforming a high-level digital design into a configuration file that programs the physical FPGA.
+    In a typical 7-Series device, approximately 75\% of the SLICE Sites are SLICELs and 25\% are SLICEMs. 
+    A single CLB Tile can therefore host either two SLICELs or one SLICEL and one SLICEM.
 
-    The process begins with \textbf{design entry}, where an engineer describes the desired functionality of a digital system using a hardware description language (HDL) such as Verilog or VHDL. 
-    Engineers may use various coding styles - behavioral, structural, dataflow, or a mix of all three - to capture the logic and behavior of the design while abstracting away the intricacies of the physical hardware architecture.
+    \textbf{CARRY4 for Arithmetic.}\quad
+    Very often an HDL design will implement many adders, counters, subtractors, or comparators, all of which are based on binary addition. 
+    They are so ubiquitous that every that in the 7-Series architecture, every SLICE features a \emph{CARRY4} BEL -- a 4-bit carry-lookahead (CLA) adder. 
+    These CARRY4 blocks can be chained vertically across SLICEs for implementing wide adders efficiently.
 
-    Next, the user runs the \textbf{synthesis} tool, which parses the raw HDL source files to infer logical behavior, perform optimizations, and instantiate primitive cells. 
-    The result is a detailed netlist, typically output as a low-level Verilog or Electronic Design Interchange Format (EDIF) file, where the design is represented purely in structural HDL. 
-    The final stage is often referred to as "technology mapping", as the vocabulary or library of primitive cells must be specific to the FPGA architecture and consequently the FPGA manufacturer.
+    \textbf{Summary and Relevance to Placement.}\quad
+    To summarize, Xilinx 7-Series FPGAs are structured into multiple Clock Regions, each divided into Tiles, which in turn contain Sites specialized for logic (CLB), DSP, RAM, or I/O. 
+    Within each CLB Tile, the two SLICE Sites host the BELs (LUTs, FFs, and a CARRY4 adder). 
+    Any placement tool must carefully ahdere to these hierarchy boundaries to ensure legal mappings that minimize wirelength and meet timing requirements.
 
-    After synthesis, the \textbf{placement} tool assigns each logic element in the netlist a specific location within the FPGA’s fabric. 
-    Depending on the placement strategy, this stage can include substages of prepacking, packing, global placement, legalization, and detailed placement. 
-    Our placer will only involve prepacking, packing, and detailed placement for simplicity. 
-    The objective is to place the logical cell instances of the netlist onto physical BELs on the device in a way that minimizes wirelength, simplifies routing, and satisfies timing constraints (signal propagation delay).
+    For more in-depth details about 7-Series FPGAs, refer to the official Xilinx user guides such as:
+    \begin{itemize}
+    \item \emph{7 Series FPGAs Overview} (UG476)
+    \item \emph{7 Series FPGA Configurable Logic Block} (UG474)
+    \item \emph{7 Series Memory Resources} (UG473)
+    \item \emph{7 Series DSP48E1 Slice} (UG479)
+    \end{itemize}
 
-    Following placement, the \textbf{routing} tool connects the placed elements by assigning physical wiring resources (switchboxes, programmable interconnects) to the nets in the netlist. 
-    The router determines their paths through the programmable interconnect network while attempting to meet timing and performance constraints.
+    This architectural context provides the necessary background for understanding how a placement algorithm should account for resource constraints and optimize performance in modern FPGA designs.
 
-    Placement and routing are generally referred jointly as the \textbf{implementation} stage, as both stages are interdependent and must adhere to the constraints of the same device. 
-    State of the art (SOTA) tools will perform routing-aware placement or placement-fluid routing for greater optimization.
+\section{FPGA Design Flow and Toolchain}
+    \label{sec:fpga_flow_toolchain}
 
-    The final stage is the \textbf{bitstream} generation, where a vendor-specific programming tool takes the implemented design and generates a configuration file or bitstream that programs the physical FPGA device.
-    The bitstream specifies the state of every configurable element—logic blocks, routing switches, and hard macros—ensuring the FPGA implements the user's design.
+    Modern FPGA designs require a sophisticated toolchain to bridge the gap between high-level hardware descriptions and the final bitstream used to configure the FPGA. 
+    Figure~\ref{fig:design_flow} illustrates a representative process that converts an abstract Hardware Description Language (HDL) design into a verified configuration file for a target device.
 
-    Parallel to this entire design flow are simulators for which the user can write testbenches to verify their design at different stages of implementation.
-    Xilinx's Vivado, for example, provides functional simulation, post-implementation simulation, and post-implementation timing simulation.
+    \begin{enumerate}
+        \item \textbf{Design Entry:} 
+            An engineer describes the intended functionality of the digital system using a hardware description language (HDL) such as Verilog or VHDL. 
+            During this phase, the coding style can vary (behavioral, structural, dataflow, etc.), but it always aims to capture high-level behavior rather than device-specific details.
 
-    This integrated toolchain, whether provided by FPGA vendors or open source communities, facilitates the transformation from an abstract HDL description to a functioning hardware configuration.
+        \item \textbf{Synthesis:} 
+            The synthesis tool parses the HDL source, performs logical optimizations, and maps the design onto primitive cells that suit the target FPGA technology. 
+            The output is typically a \emph{structural} netlist (e.g., EDIF or structural Verilog) which details how the design’s logic is broken down into LUTs, FFs, and other vendor-specific cells.
 
-\section{Placement Implementation via RapidWright API}
+        \item \textbf{Placement and Routing (Implementation):} 
+            In \emph{placement}, each logical cell from the synthesized netlist is assigned to a physical location on the FPGA fabric. 
+            For instance, LUTs and FFs go into specific \emph{BELs} within the device’s CLB sites, and specialized cells such as DSPs and Block RAMs must be placed in their corresponding tile types. 
+            Next, \emph{routing} determines how signals are physically wired through the FPGA’s configurable interconnect network. 
+            Modern tools often interleave these steps (e.g., placement-aware routing or routing-aware placement) to better meet timing and area objectives.
 
-    %%%
+        \item \textbf{Bitstream Generation:} 
+            After a design is fully placed, routed, and timing-closed, the toolchain produces a final \emph{bitstream} that sets the configuration of every programmable element in the FPGA. 
+            This bitstream can then be loaded onto the device, either through vendor software or via a custom programming interface.
 
-    The RapidWright API offers an additional workflow column in parallel to both the design flow and verification flow. 
-    RapidWright is an open source Java framework that enables netlist implementation manipulation for modern Xilinx FPGAs.
+        \item \textbf{Verification:} 
+            In parallel to the design flow, simulations and testbenches validate correctness of the user's design at multiple abstraction levels. 
+            Engineers may begin with functional or behavioral simulations, then progress to post-synthesis simulations, and finally to post-implementation simulations that incorporate real routing delays.
+            With each higher level of fidelity, computational requirements grow significantly due to increasing complexity and the need to analyze more variables over time. 
+            Ensuring correct functionality and meeting timing closure at the post-implementation stage is crucial before deploying the design to hardware.
+Given the importance of thorough verification, many established companies dedicate one verification engineer for every design engineer.
+
+    \end{enumerate}
+
+    This workflow underscores the critical role of \textbf{placement} in bridging the netlist to a physical realization. 
+    An efficient placement algorithm can drastically reduce compile times and improve design performance, enabling broader adoption of FPGAs in application spaces that require fast design iteration.
 
@@
     \centering
-        \includegraphics[width=8.0cm]{figures/vivado_dcps.png}
-        \caption{Showing how RapidWright can take a design checkpoint from any stage of the design flow, allow user customization, and reintegrate it back into the design flow.}
+    \includegraphics[width=7.5cm]{figures/design_flow.png}
+    \caption{Typical FPGA design and verification workflow.}
+    \label{fig:design_flow.png}
+    \end{figure}
+
+\section{RapidWright API}
+    \label{sec:rapidwright_api}
+
+    \textbf{RapidWright} is an open-source Java framework from AMD/Xilinx that provides direct access to the netlist and device databases used by vendor tools. 
+    This framework positions itself as an additional workflow column, allowing users to intercept or replace parts of the standard design flow with custom optimization stages (see Figure~\ref{fig:vivado_dcps}).
+
+    \begin{itemize}
+    \item \textbf{Design Checkpoints:} 
+        RapidWright leverages \texttt{.dcp} files (design checkpoints) generated at various stages of a Vivado flow. 
+        By importing a checkpoint, engineers can manipulate the netlist, placement, or routing externally, then re-export a modified checkpoint for further processing in the Vivado workflow column.
+
+    \item \textbf{Key Packages:} 
+        RapidWright revolves around three primary data model packages:
+        \begin{enumerate}
+        \item \emph{edif} -- Represents the logical netlist in an abstracted EDIF-like structure.
+        \item \emph{design} -- Contains data structures for the physical implementation (Cells, Nets, Sites, BELs, etc.).
+        \item \emph{device} -- Provides a database of the target FPGA architecture (e.g., Site coordinates, Tile definitions, routing resources).
+        \end{enumerate}
+
+    \item \textbf{Interfacing with the Netlist and Device:} 
+        An engineer can query the netlist to find specific resources (LUTs, FFs, DSPs, etc.) and then map or move them onto device sites. 
+        This level of control is invaluable for research in custom placement, advanced packing techniques, or experimental routing algorithms.
+    \end{itemize}
+
+    \begin{figure}[]
+        \centering
+        \includegraphics[width=7.5cm]{figures/vivado_dcps.png}
+        \caption{RapidWright workflow integrating into the default Vivado design flow.}
         \label{fig:vivado_dcps}
@@
 
-    \begin{figure}[]
+    By exposing these low-level internals, RapidWright allows fine-grained design transformations that go beyond the standard Vivado IDE’s capabilities. 
+    Researchers can prototype new EDA strategies without needing to re-implement an entire FPGA backend from scratch, thus accelerating innovation in placement and routing methodologies.
+
+\section{Simulated Annealing}
+    With a basic understanding of FPGA architecture, design placement, and RapidWright, we have all the necessary pieces to implement our SA placer. 
+    Here we outline in detail each substage of our implementation: PrePacking, Packing, and Placement.
+
+    \begin{figure*}[]
         \centering
-        \includegraphics[width=8.5cm]{figures/edif_design_device.png}
-        \caption{The placement flow illustrating PrePacking into Packing into Placement for a logical netlist with 3 EDIF cells. EHCI is short for EDIFHierCellInst.}
-        \label{fig:edif_design_device}
-    \end{figure}
-
-    Figure \ref{fig:vivado_dcps} shows how the API allows us to take design checkpoints in the form of .dcp files generated at any stage of the design flow, allow the user to make custom optimizations, and reintegrate it back into the design flow.
-    This allows the user to substitute any stage in the default FPGA design flow offered by Vivado with a custom user-defined process to meet complex design constraints. 
-    There are three packages of objects that are most relevant to building our placer: the \textbf{edif} package, \textbf{design} package, and \textbf{device} package. 
-    Figure \ref{fig:edif_objects} shows the EDIF netlist objects and device routing objects as shown in the Vivado netlist viewer and device viewer.
-    Figure \ref{fig:edif_design_device} illustrates the flow between these packages in relation to our PrePacking, Packing, and Placement flow.
-
-    \begin{figure*}[t]
-        \centering
-        \includegraphics[width=14.0cm]{figures/edif_netlist.png}
-        \includegraphics[width=14.0cm]{figures/routing_objects.png}
-        \caption{Examples of edif netlist objects (top) and routing objects (bottom) for a Xilinx FPGA}
-        \label{fig:edif_objects}
+        \includegraphics[width=14.0cm]{figures/substages.png}
+        \caption{The data classes populated at each substage: PrepackedDesign, PackedDesign, and PlacedDesign.}
+        \label{fig:substages}
     \end{figure*}
 
-    RapidWright reads in a post-synthesis design checkpoint and allows us to traverse the netlist's logical EDIFCells through various function calls. 
-    For each EDIFCell object in the netlist, we create a Cell design object, which we then place onto physical device BEL object. 
-    CellPins and Nets are automatically generated from EDIFHierPortInst and EDIFHierNet upon the Cell's creation from an EDIFHierCellInst.
+    \subsection{Prepacking}
+        \label{subsec:prepacking}
 
-    These three steps correspond to our placement substages of PrePacking, Packing, and Placement, respectively.
+        In the context of implementing a custom \emph{Simulated Annealing} (SA) placer, \textbf{prepacking} is the first step in consolidating functionally related cells from the raw post-synthesis netlist into logical clusters that correspond more naturally to the physical FPGA structures.
+        We traverse the netlist (e.g., via RapidWright’s \texttt{edif} and \texttt{design} packages) to identify:
 
-    For a complete understanding of the RapidWright packages, please refer to the API Overview \url{https://www.rapidwright.io/docs/RapidWright_Overview.html} and the API reference.
+        \begin{itemize}
+        \item \textbf{LUT-FF pairs:} 
+            Most synchronous designs implement finite-state machines or pipelined datapaths, mapping their combinational logic into LUTs directly feeding FFs. 
+            Clustering each LUT-FF pair under a single “prepacked object” facilitates usage of dedicated intra-SLICE connections, thereby reducing routing overhead.
 
+        \item \textbf{CARRY4 chains:} 
+            Arithmetic operations (adders, counters, comparators) often chain multiple \emph{CARRY4} cells vertically. 
+            During prepacking, it is crucial to detect these chains so they remain grouped and placed in physically consecutive and vertical SLICEs for placement legality and optimal performance.
+
+        \item \textbf{DSP cascades:} 
+            In designs with large multiply-accumulate structures, DSP48E1 blocks can be used and cascaded along dedicated paths. 
+            Identifying such cascades early helps the placer keep them localized, minimizing high-speed interconnect usage.
+            Like carry chains, each DSP in a DSP cascades must be placed vertically, in physically consecutive and vertical DSP48E1 Sites.
+
+        \item \textbf{Other specialized patterns:} 
+            Various design-specific macros (e.g., BRAM-based FIFOs, shift-register logic) also benefit from being clustered, though these may be addressed in subsequent or more advanced stages.
+        \end{itemize}
+
+        These prepacked “super-cells” or “block groups” more closely match the physical site resources. 
+        This strategy not only reduces the complexity of the placement search space but also helps ensure legality (e.g., shared control signals, shared carry nets). 
+        The subsequent \emph{packing} and \emph{placement} steps can then operate on these higher-level groupings with improved efficiency.
+
+\newpage
+o
+\newpage
+%%% OLD
+
+\section{OLD: Simulated Annealing}
     With a basic understanding of FPGA architecture, design placement, and RapidWright, we have all the necessary pieces to implement our SA placer. 
@@
             As the name implies, CARRY chains are chains of CARRY cells. 
-            Very often an HDL design will implement many adders, counters, subtractors, or comparators, all operations based on binary addition. 
-            They are so ubiquitous that every that in the 7-Series architecture, every CLB comes embedded with a CARRY4 BEL, a 4-bit carry-lookahead (CLA) adder that can be chained together across CLBs to implement wider adders.
             % On the Ultrascale architecture, the each CLB contains a CARRY8 BEL, an 8-bit CLA adder, allowing for even greater logic density.
@@
         \caption{SLICEM Site}
-        \label{slicem}
+        \label{fig:slicem}
     \end{figure*}
@@
         \caption{SLICEL Site}
-        \label{slicel}
+        \label{fig:slicel}
     \end{figure*}

### 2025-04-16 latex/paper/paper.tex
move cell pattern explanation into architecture section

@@
     We explain in detail each of these concepts for a broader audience as they provide much needed context for FPGA placement algorithms as a concept. 
-    However, readers who are already familiar with these concepts can skip directly to the RapidWright API section.
+    However, readers who are already familiar with these concepts can skip directly to the RapidWright API section \ref{sec:rapidwright_api} or to the Simulated Annealing section \ref{sec:simulated_annealing}.
 
@@
     We will briefly review the evolution of configurable logic architecture starting in the 1970s and quickly work our way up to modern day FPGA architecture.
+
+    \textbf{PLA:} \quad
     The journey began with the Programmable Logic Array (PLA) in the early 1970s. 
@@
 
+    \textbf{CPLD:} \quad 
     Later in the same decade came the Complex Programmable Logic Device (CPLD), which took the form of an array of Configurable Logic Blocks (CLBs). 
@@
 
+    \textbf{Homogeneous FPGA:} \quad 
     The mid-1980s saw the introduction of homogeneous FPGAs, which were built as a grid of CLBs. 
@@
 
+    \textbf{Heterogenous FPGA:} \quad 
     This brings us to modern day FPGA architectures. 
@@
     \section{Xilinx 7-Series Architecture}
+    \label{sec:7_series}
 
-    \begin{figure*}[!ht]
+    \begin{figure}[]
         \centering
-        \includegraphics[width=15.0cm]{figures/hierarchy_3.png}
-        \caption{Architecture Hierarchy of a Xilinx FPGA}
+        \includegraphics[width=5.0cm]{figures/hierarchy_5.png}
+        \caption{Overview of Architecture Hierarchy of a Xilinx FPGA}
+        \label{fig:hierarchy_5}
+    \end{figure}
+
+    \begin{figure}[]
+        \centering
+        \includegraphics[width=9.0cm]{figures/hierarchy_4.png}
+        \caption{Detailed Architecture Hierarchy of a Xilinx FPGA}
         \label{fig:hierarchy_4}
-    \end{figure*}
+    \end{figure}
 
@@
 During logic synthesis, larger Boolean expressions exceeding six inputs are automatically split across multiple LUTs.
-    \item \emph{SLICEM (SLICE-Memory):} SLICEM LUTs include all the features of a SLICEL but can also be configured as small distributed RAM (known as \emph{Distributed RAM}) or as shift registers.
+        \item \emph{SLICEM (SLICE-Memory):} The SLICEM include all the features of a SLICEL but its LUTs can also be configured as small distributed RAM (known as \emph{Distributed RAM}) or as shift registers.
 This local memory capability supplements the larger Block RAM (BRAM) cells and is beneficial for designs requiring numerous small RAM blocks.
@@
 
-    \begin{figure}[!ht]
+    \begin{figure}[h]
         \centering
@@
 
-    \textbf{CARRY4 for Arithmetic.}\quad
+    ( Explain cell patterns here )
+
+    \textbf{LUTs} \quad
+
+    \textbf{FFs} \quad
+
+    \textbf{LUT-FF Pairs} \quad
+
+    \textbf{CARRY} \quad
         Very often an HDL design will implement many adders, counters, subtractors, or comparators, all of which are based on binary addition. 
@@
 
-    \textbf{Summary and Relevance to Placement.}\quad
-    To summarize, Xilinx 7-Series FPGAs are structured into multiple Clock Regions, each divided into Tiles, which in turn contain Sites specialized for logic (CLB), DSP, RAM, or I/O. 
-    Within each CLB Tile, the two SLICE Sites host the BELs (LUTs, FFs, and a CARRY4 adder). 
-    Any placement tool must carefully ahdere to these hierarchy boundaries to ensure legal mappings that minimize wirelength and meet timing requirements.
+    \textbf{CARRY Chains} \quad
+
+    \textbf{DSPs} \quad
+
+    \textbf{DSP Cascades} \quad
+
+    \textbf{BRAMs} \quad
+
+    \textbf{Loose Single Cells} \quad
 
@@
 
+    \begin{figure}[]
+    \centering
+    \includegraphics[width=7.5cm]{figures/design_flow.png}
+    \caption{Typical FPGA design and verification workflow.}
+    \label{fig:design_flow}
+    \end{figure}
+
     \begin{enumerate}
@@
             Next, \emph{routing} determines how signals are physically wired through the FPGA’s configurable interconnect network. 
-            Modern tools often interleave these steps (e.g., placement-aware routing or routing-aware placement) to better meet timing and area objectives.
+            Modern tools often interleave these steps (e.g., fluid-placement routing or routing-aware placement) to better meet timing and area objectives.
 
@@
 
-    \begin{figure}[]
-    \centering
-    \includegraphics[width=7.5cm]{figures/design_flow.png}
-    \caption{Typical FPGA design and verification workflow.}
-    \label{fig:design_flow.png}
-    \end{figure}
-
 \section{RapidWright API}
@@
     \textbf{RapidWright} is an open-source Java framework from AMD/Xilinx that provides direct access to the netlist and device databases used by vendor tools. 
-    This framework positions itself as an additional workflow column, allowing users to intercept or replace parts of the standard design flow with custom optimization stages (see Figure~\ref{fig:vivado_dcps}).
+    This framework positions itself as an additional workflow column, allowing users to intercept or replace stages of the standard design flow with custom optimization stages (see Figure~\ref{fig:vivado_dcps}).
 
@@
         An engineer can query the netlist to find specific resources (LUTs, FFs, DSPs, etc.) and then map or move them onto device sites. 
-        This level of control is invaluable for research in custom placement, advanced packing techniques, or experimental routing algorithms.
+        This level of control over backend resources is necessary for research in custom placement, advanced packing techniques, or experimental routing algorithms.
     \end{itemize}
@@
 \section{Simulated Annealing}
+    \label{sec:simulated_annealing}
     With a basic understanding of FPGA architecture, design placement, and RapidWright, we have all the necessary pieces to implement our SA placer. 
     Here we outline in detail each substage of our implementation: PrePacking, Packing, and Placement. 
+    Shown in Figure
+
+    \begin{figure}[]
+        \centering
+        \includegraphics[width=8.0cm]{figures/edif_design_device.png}
+        \caption{}
+    \end{figure}
 
@@
         In the context of implementing a custom \emph{Simulated Annealing} (SA) placer, \textbf{prepacking} is the first step in consolidating functionally related cells from the raw post-synthesis netlist into logical clusters that correspond more naturally to the physical FPGA structures.
-        We traverse the netlist (e.g., via RapidWright’s \texttt{edif} and \texttt{design} packages) to identify:
+        We traverse the raw post-synthesis netlist via RapidWright’s \texttt{edif} packages to identify LUT-FF Pairs, CARRY4 Chains, and DSP Cascades.
 
-        \begin{itemize}
-        \item \textbf{LUT-FF pairs:} 
+        \textbf{LUT-FF pairs:} \quad
             Most synchronous designs implement finite-state machines or pipelined datapaths, mapping their combinational logic into LUTs directly feeding FFs. 
@@
 
-        \item \textbf{CARRY4 chains:} 
+        \textbf{CARRY4 chains:} \quad
             Arithmetic operations (adders, counters, comparators) often chain multiple \emph{CARRY4} cells vertically. 
@@
 
-        \item \textbf{DSP cascades:} 
+        \textbf{DSP cascades:} \quad
             In designs with large multiply-accumulate structures, DSP48E1 blocks can be used and cascaded along dedicated paths. 
@@
 
-        \item \textbf{Other specialized patterns:} 
+        \textbf{Loose or Single Cells} \quad
+            The remaining cells 
             Various design-specific macros (e.g., BRAM-based FIFOs, shift-register logic) also benefit from being clustered, though these may be addressed in subsequent or more advanced stages.
-        \end{itemize}
 
-        These prepacked “super-cells” or “block groups” more closely match the physical site resources. 
         This strategy not only reduces the complexity of the placement search space but also helps ensure legality (e.g., shared control signals, shared carry nets). 
         The subsequent \emph{packing} and \emph{placement} steps can then operate on these higher-level groupings with improved efficiency.
+
+    \subsection{Packing}
+        \label{subsec:packing}
+
+    \subsection{Placement}
+        \label{subsec:placement}
+        Up until now we have only organized the logical \texttt{Cells} into \texttt{SiteInsts}. 
+        This is where simulated annealing actually begins.
 

### 2025-04-18 latex/paper/paper.tex
primitive BEL explanations

@@
     In some high-density devices, multiple Clock Regions may be consolidated into one or more \textbf{Super Logic Regions}~(SLRs). 
-    However, for the scope of this paper, we focus on Xilinx 7-Series devices with a single SLR.
+    However, for the scope of this paper, we focus on Xilinx 7-Series devices with only a single SLR.
 
@@
     Each SLICE Site has a set of BELs including eight LUTs, eight FFs, and one \emph{CARRY4} adder. 
-    These SLICEs provide the core logic and arithmetic functionality of the device.
+    The BELs in these SLICEs facilitate the bulk of the general programmability of the FPGA fabric. 
+    We will explain in detail the motivation and function of these BELs.
+
+    \textbf{LUTs} \quad
+        Combinational logic is universal to all HDL designs. 
+        As the their name suggests, Look‑Up Tables (LUTs) facilitate combinational logic by acting as tiny asynchronously‑accessed ROMs whose contents are fixed when the FPGA is programmed.  
+        For any boolean function, the synthesizer precalculates the output to every possible input combination and stores the resulting truth table into a LUT's static memory. 
+        The inputs are then essentially treated as an address space that maps to a data value space in an asynchronous ROM. 
+        No explicit logic gates like NAND or XNOR are synthesized, contrary to what newcomers might expect from a "Field Programmable Gate Array". 
+        In the 7‑Series devices, one LUT can facilitate any 6-input boolean function, or two 5-input functions, as long as they share the same input signals.  
+        The LUT can also host two independent boolean functions of up to 3 inputs each, even when the inputs are not shared.  
+        Functions requiring more than six unique inputs are decomposed across multiple cascaded LUTs.
+        Figure \ref{fig:lut_synthesis} shows an example of where a LUT is typically synthesized in a design entry.
+
+        \begin{figure}[]
+            \centering
+            \includegraphics[width=9.0cm]{figures/lut_synthesis.png}
+            \caption{LUT synthesis from user design}
+            \label{fig:lut_synthesis}
+        \end{figure}
+
+    \textbf{FFs} \quad
+        FFs are synthesized to facilitate synchronous event-driven signal assignment. 
+        For most FPGA users, in Verilog, this essentially means assignments wrapped in \texttt{always @(posedge clk)} statements. 
+        Figure \ref{fig:ff_synthesis} shows an example of where a flip flop (FF) is typically synthesized. 
+        The cell primitive FDRE is a type of FF and belongs to a family of D Flip Flops (DFFs) with Clock Enable (CE). 
+        \begin{itemize}
+            \item \textbf{FDCE:} DFF with CE and Asynchronous Clear
+            \item \textbf{FDPE:} DFF with CE and Asynchronous Preset
+            \item \textbf{FDSE:} DFF with CE and Synchronous Set
+            \item \textbf{FDRE:} DFF with CE and Synchronous Reset
+        \end{itemize}
+        In a typical HDL design, the vast majority of FFs will be synthesized as FDREs. 
+        A FF BEL may also be used to implement latches, however, since latches are generally known to be bad practice in FPGA design, we will not consider latch synthesis for this paper. 
+        \begin{figure}[]
+            \centering
+            \includegraphics[width=9.0cm]{figures/ff_synthesis.png}
+            \caption{FF synthesis from user design}
+            \label{fig:ff_synthesis}
+        \end{figure}
+
+    \textbf{LUT-FF Pairs} \quad
+
+    \textbf{CARRY} \quad
+        Very often an HDL design will implement many adders, counters, subtractors, or comparators, all of which are based on binary addition. 
+        They are so ubiquitous that every that in the 7-Series architecture, every SLICE features a \emph{CARRY4} BEL -- a 4-bit carry-lookahead (CLA) adder. 
+        These CARRY4 blocks can be chained vertically across SLICEs for implementing wide adders efficiently.
+
+    \textbf{CARRY Chains} \quad
 
@@
     ( Explain cell patterns here )
-
-    \textbf{LUTs} \quad
-
-    \textbf{FFs} \quad
-
-    \textbf{LUT-FF Pairs} \quad
-
-    \textbf{CARRY} \quad
-        Very often an HDL design will implement many adders, counters, subtractors, or comparators, all of which are based on binary addition. 
-        They are so ubiquitous that every that in the 7-Series architecture, every SLICE features a \emph{CARRY4} BEL -- a 4-bit carry-lookahead (CLA) adder. 
-        These CARRY4 blocks can be chained vertically across SLICEs for implementing wide adders efficiently.
-
-    \textbf{CARRY Chains} \quad
 

### 2025-04-20 latex/paper/paper.tex
finish lutff pair

@@
     In the 7-Series architecture, the term \emph{CLB} (Configurable Logic Block) refers to a \emph{CLB Tile} that contains two \emph{SLICE} Sites. 
-    Each SLICE Site has a set of BELs including eight LUTs, eight FFs, and one \emph{CARRY4} adder. 
+    Xilinx offers two variants of SLICE Sites: \textbf{SLICEL} and \textbf{SLICEM}.
+    \begin{itemize}
+        \item Each \textbf{SLICEL} has a set of BELs including eight LUTs, eight FFs, and one   \emph{CARRY4} adder. The LUT BELs in a SLICEL can only host LUT Cells. 
+        \item The \textbf{SLICEM} includes all the features of a SLICEL but its LUT BELs can host both LUT Cells, which are asynchronous ROM elements, or RAM32M Cells, which are synchronous 32-deep RAM elements.
+These cells are also referred to as \emph{Distributed RAM} in the Xilinx documentation.
+        These cells can offer an alternative to the larger, more dedicated 18K-36K RAMB18E1 cells when RAM resources are highly utilized. 
+    \end{itemize}
+    In a typical 7-Series device, approximately 75\% of the SLICE Sites are SLICELs and 25\% are SLICEMs. 
+    A single CLB Tile can therefore host either two SLICELs or one SLICEL and one SLICEM.
+    For simplicity, however, we will only consider SLICELs for general logic and use the dedicated Block RAM (BRAM) Tiles to implement RAM elements.
+
     The BELs in these SLICEs facilitate the bulk of the general programmability of the FPGA fabric. 
     We will explain in detail the motivation and function of these BELs.
+
+    \subsection{7-Series BEL Library}
 
@@
             \includegraphics[width=9.0cm]{figures/lut_synthesis.png}
-            \caption{LUT synthesis from user design}
+            \caption{\textbf{Above:} LUT synthesis from user design}
             \label{fig:lut_synthesis}
@@
         FFs are synthesized to facilitate synchronous event-driven signal assignment. 
-        For most FPGA users, in Verilog, this essentially means assignments wrapped in \texttt{always @(posedge clk)} statements. 
+        For most Verilog users, this generally means signal assignments wrapped in \texttt{always @(posedge clk)} statements. 
         Figure \ref{fig:ff_synthesis} shows an example of where a flip flop (FF) is typically synthesized. 
@@
         \begin{itemize}
-            \item \textbf{FDCE:} DFF with CE and Asynchronous Clear
-            \item \textbf{FDPE:} DFF with CE and Asynchronous Preset
-            \item \textbf{FDSE:} DFF with CE and Synchronous Set
-            \item \textbf{FDRE:} DFF with CE and Synchronous Reset
+            \item \emph{FDCE - DFF with CE and Asynchronous Clear}
+            \item \emph{FDPE - DFF with CE and Asynchronous Preset}
+            \item \emph{FDSE - DFF with CE and Synchronous Set}
+            \item \emph{FDRE - DFF with CE and Synchronous Reset}
         \end{itemize}
-        In a typical HDL design, the vast majority of FFs will be synthesized as FDREs. 
-        A FF BEL may also be used to implement latches, however, since latches are generally known to be bad practice in FPGA design, we will not consider latch synthesis for this paper. 
+        In a typical HDL design, the vast majority of FFs will be synthesized as FDREs with the occasional FDSE, as it is generally good practice to keep FPGA designs synchronous. 
+        A FF BEL may also host a latch cell, however, since they are generally known to be bad practice in FPGA design, we will not consider latches in this paper. 
         \begin{figure}[]
@@
             \includegraphics[width=9.0cm]{figures/ff_synthesis.png}
-            \caption{FF synthesis from user design}
+            \caption{\textbf{Above:} FF synthesis from user design}
             \label{fig:ff_synthesis}
@@
     \textbf{LUT-FF Pairs} \quad
+        FPGA designs are very often modelled as a collection of Finite State Machines (FSM) like shown in Figure \ref{fig:fsm}.
+        \begin{figure}[]
+            \centering
+            \includegraphics[width=8.0cm]{figures/fsm.png}
+            \caption{\textbf{Above:} Finite state machine (Moore)}
+            \label{fig:fsm}
+        \end{figure}
+        Many times a design will also use pipelining like shown in Figure, either to model signal buffers or shift registers, or to split up large combinational logic into time slices to meet timing constraints.
+        These common design structures result in many consecutive sections of combinational logic feeding into a vector of registers. 
+        The synthesizer naturally synthesizes these structures as consecutive pairs of LUTs feeding into FFs as shown in Figure \ref{fig:pipelining}. 
+        Figure \ref{fig:lut_ff_pair} shows an example of a synthesized LUT-FF Pair. 
+        \begin{figure}[]
+            \centering
+            \includegraphics[width=8.0cm]{figures/pipelining.png}
+            \caption{\textbf{Above:} Pipelining synthesized as consecutive LUT-FF pairs}
+            \label{fig:pipelining}
+        \end{figure}
+
+        \begin{figure}[]
+            \centering
+            \includegraphics[width=8.0cm]{figures/lut_ff_pair.png}
+            \caption{A synthesized LUT-FF Pair}
+            \label{fig:lut_ff_pair}
+        \end{figure}
+
+        Shown in Figure \ref{fig:intersite_intrasite} are two possible placements for this cell net. 
+        On the right, the cells are placed across different Sites, thus the only way to route the net between the cells is through general inter-site routing. 
+        On the bottom left, the cells are placed within the same Site, taking advantage of the intra-site routing and avoiding costly inter-site routing.
+
+        \begin{figure}[]
+            \centering
+            \includegraphics[width=8.0cm]{figures/intersite_intrasite_2.png}
+            \caption{Intrasite vs Intersite LUT-FF Placement}
+            \label{fig:intersite_intrasite}
+        \end{figure}
 
     \textbf{CARRY} \quad
-        Very often an HDL design will implement many adders, counters, subtractors, or comparators, all of which are based on binary addition. 
+        An FPGA design will also typically implement many adders, counters, subtractors, or comparators, all of which are based on binary addition. 
         They are so ubiquitous that every that in the 7-Series architecture, every SLICE features a \emph{CARRY4} BEL -- a 4-bit carry-lookahead (CLA) adder. 
@@
         \caption{The data classes populated at each substage: PrepackedDesign, PackedDesign, and PlacedDesign.}
-        \label{fig:substages}
+        \label{}
     \end{figure*}
@@
                 \caption{A Finite State Machine (Moore)}
-                \label{fig:fsm}
             \end{figure}
@@
                 }
-                \label{fig:lut_ff_pair}
             \end{figure}
@@
                 }
-                \label{fig:intrasite_intersite}
             \end{figure}
@@
                 }
-                \label{fig:carry_cell_edif}
             \end{figure}
@@
                 \caption{Example of a CARRY4 chain of length 6 as seen in the Vivado netlist viewer.}
-                \label{fig:carry_chain_edif}
             \end{figure*}
@@
                 \caption{\textbf{Left}: A Carry Chain of length 6 spanning 6 Sites across 6 Tiles. \textbf{Right}: A DSP cascade of length 2 spanning 2 Sites across 1 Tile.}
-                \label{fig:device_carry_chain_routing}
             \end{figure}

### 2025-04-20 latex/paper/paper.tex
carry chains

@@
 
-        Shown in Figure \ref{fig:intersite_intrasite} are two possible placements for this cell net. 
+        Shown in Figure \ref{fig:intersite_intrasite} are two possible placements for a LUT-FF Pair on the physical device. 
         On the right, the cells are placed across different Sites, thus the only way to route the net between the cells is through general inter-site routing. 
-        On the bottom left, the cells are placed within the same Site, taking advantage of the intra-site routing and avoiding costly inter-site routing.
+        On the bottom left, the cells are placed within the same Site in the same lane, taking advantage of the intra-site routing without burdening the general router with additional inter-site routing.
+        This is an important consideration for minimizing wirelength during placement.
 
@@
         These CARRY4 blocks can be chained vertically across SLICEs for implementing wide adders efficiently. 
+        These CARRY4 BELs will also be connected to LUTs and FFs, and should be placed in the same Site whenever possible to minimize wirelength. 
+        \begin{figure*}[]
+            \centering
+            \includegraphics[width=16.0cm]{figures/carry_chain_edif_3.png}
+            \caption{A carry chain of size 3 as shown in the Vivado EDIF netlist viewer}
+            \label{fig:carry_chain_edif}
+        \end{figure*}
 

### 2025-04-21 latex/paper/paper.tex
lots of reorg

@@
-\documentclass[twocolumn]{article}
+\documentclass{article}
 \usepackage{hyperref}
@@
 }
-
-% \usepackage{multicol}
+\usepackage[export]{adjustbox}
+\usepackage{multicol}
+\usepackage{caption}
 \usepackage[a4paper,
@@
 \maketitle
+
+\begin{multicols}{2}
 
@@
 
-\section{A Brief History on FPGA Architecture}
-
-    \begin{figure}
-        \centering
-        \includegraphics[width=8.0cm]{figures/pal_2.png}
-        \caption{PAL architecture with 5 inputs, 8 programmable AND gates and 4 fixed OR gates}
-        \label{fig:pla}
-        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/Programmable-Logic-Devices/Programmable-Array-Logic-PAL/
-        % https://www.naukri.com/code360/library/difference-between-pla-and-pal 
-    \end{figure}
+\section{FPGA Architecture History}
 
@@
     The PAL simplified the PLA by fixing the OR gates, resulting in a fixed-OR, programmable-AND design, which sacrificed some logic flexibility to simplify its manufacture. 
-    Figure \ref{fig:pla} shows one such PLA architecture.
-
-    \begin{figure}
-        \centering
-        \includegraphics[width=8.0cm]{figures/cpld.png}
-        \caption{CPLD architecture with 4 CLBs (PAL-like blocks)}
-        \label{fig:cpld}
-        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/CPLD/Complex-Programmable-Logic-Device-CPLDs/
-    \end{figure}
+    Figure \ref{fig:pal_2} shows one such PAL architecture.
 
@@
     Figure \ref{fig:cpld} shows one such CPLD architecture. 
+    \\
 
-    \begin{figure}
+    {
         \centering
-        \includegraphics[width=8.0cm]{figures/homogenous_fpga.png}
-        \includegraphics[width=8.0cm]{figures/heterogenous_fpga_3.jpg}
-        \caption{
-            \textbf{Top}: A homogeneous island-style FPGA architecture with 16 CLBs in a grid. 
-            \textbf{Bottom}: A heterogeneous island-style FPGA with a mix of CLBs and macrocells.
+        \includegraphics[width=\columnwidth]{figures/pal_2.png}
+        \captionof{figure}{PAL architecture with 5 inputs, 8 programmable AND gates and 4 fixed OR gates}
+        \label{fig:pal_2}
+        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/Programmable-Logic-Devices/Programmable-Array-Logic-PAL/
+        % https://www.naukri.com/code360/library/difference-between-pla-and-pal 
     }
-        \label{fig:fpga}
-        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/FPGA/Field-programmable-gate-array-FPGA/
-    \end{figure}
+
+    \newcolumn
 
@@
     The capacity of an FPGA was often measured by how many logical elements or CLBs it offered, which grew from hundreds to thousands and now to hundreds of thousands of CLBs.
+    \\
+
+    {
+        \centering
+        \includegraphics[width=0.8\columnwidth]{figures/cpld.png}
+        \captionof{figure}{CPLD architecture with 4 CLBs (PAL-like blocks)}
+        \label{fig:cpld}
+        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/CPLD/Complex-Programmable-Logic-Device-CPLDs/
+    }
+    {
+        \centering
+        \includegraphics[width=\columnwidth]{figures/homogenous_fpga.png}
+        \captionof{figure}{A homogeneous island-style FPGA architecture with 16 CLBs in a grid.}
+        \label{fig:fpga}
+        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/FPGA/Field-programmable-gate-array-FPGA/
+    }
+
+    \newpage
+
+    {
+        \centering
+        \includegraphics[width=\columnwidth]{figures/heterogenous_fpga_3.jpg}
+        \captionof{figure}{A heterogeneous island-style FPGA with a mix of CLBs and macrocells.}
+        \label{fig:fpga}
+        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/FPGA/Field-programmable-gate-array-FPGA/
+    }
 
@@
     \label{sec:7_series}
-
-    \begin{figure}[]
-        \centering
-        \includegraphics[width=5.0cm]{figures/hierarchy_5.png}
-        \caption{Overview of Architecture Hierarchy of a Xilinx FPGA}
-        \label{fig:hierarchy_5}
-    \end{figure}
-
-    \begin{figure}[]
-        \centering
-        \includegraphics[width=9.0cm]{figures/hierarchy_4.png}
-        \caption{Detailed Architecture Hierarchy of a Xilinx FPGA}
-        \label{fig:hierarchy_4}
-    \end{figure}
-
     The Xilinx 7-Series devices, first introduced in 2010, follow a heterogeneous island-style architecture as discussed previously. 
@@
     Figure~\ref{fig:hierarchy_4} illustrates a high-level view of the hierarchical organization of a 7-Series FPGA. 
-    At its lowest level, the device consists of a large array of atomic components called \emph{Basic Elements of Logic}~(BELs). 
-    These BELs encompass look-up tables (LUTs), flip-flops (FFs), block RAMs (BRAMs), DSP slices (DSPs), and the configurable interconnect fabric. 
+    At its lowest level, the device consists of a large array of atomic components called \emph{Basic Elements of Logic}~(\textbf{BELs}). 
+    These BELs encompass look-up tables (\textbf{LUTs}), flip-flops (\textbf{FFs}), block RAMs (\textbf{BRAMs}), DSP slices (\textbf{DSPs}), and the configurable interconnect fabric. 
     They constitute the fundamental building blocks for implementing digital circuits on the FPGA.
@@
     However, for the scope of this paper, we focus on Xilinx 7-Series devices with only a single SLR.
+
+    \end{multicols}
+    {
+        \centering
+        \includegraphics[valign=c, width=0.3\columnwidth]{figures/hierarchy_5.png}
+        \includegraphics[valign=c, width=0.6\columnwidth]{figures/hierarchy_4.png}
+        \captionof{figure}{Architecture Hierarchy of a Xilinx FPGA}
+        \label{fig:hierarchy_4}
+    }
+    \vfill
+    \begin{multicols}{2}
 
@@
     A single CLB Tile can therefore host either two SLICELs or one SLICEL and one SLICEM.
-    For simplicity, however, we will only consider SLICELs for general logic and use the dedicated Block RAM (BRAM) Tiles to implement RAM elements.
+    For simplicity, however, we will only consider SLICELs for general logic and use the dedicated RAMB18E1 cells to implement RAM elements.
 
@@
         No explicit logic gates like NAND or XNOR are synthesized, contrary to what newcomers might expect from a "Field Programmable Gate Array".
+
+        {
+            \centering
+            \includegraphics[width=\columnwidth]{figures/lut_synthesis.png}
+            \captionof{figure}{LUT synthesis from user design}
+            \label{fig:lut_synthesis}
+        }
+
         In the 7‑Series devices, one LUT can facilitate any 6-input boolean function, or two 5-input functions, as long as they share the same input signals.  
@@
         Figure \ref{fig:lut_synthesis} shows an example of where a LUT is typically synthesized in a design entry.
-
-        \begin{figure}[]
-            \centering
-            \includegraphics[width=9.0cm]{figures/lut_synthesis.png}
-            \caption{\textbf{Above:} LUT synthesis from user design}
-            \label{fig:lut_synthesis}
-        \end{figure}
 
@@
         In a typical HDL design, the vast majority of FFs will be synthesized as FDREs with the occasional FDSE, as it is generally good practice to keep FPGA designs synchronous. 
-        A FF BEL may also host a latch cell, however, since they are generally known to be bad practice in FPGA design, we will not consider latches in this paper. 
-        \begin{figure}[]
+        A FF BEL may also host a latch cell, however, since they are generally bad practice in FPGA design, we will not consider latches in this paper. 
+        {
             \centering
-            \includegraphics[width=9.0cm]{figures/ff_synthesis.png}
-            \caption{\textbf{Above:} FF synthesis from user design}
+            \includegraphics[width=\columnwidth]{figures/ff_synthesis.png}
+            \captionof{figure}{FF synthesis from user design}
             \label{fig:ff_synthesis}
-        \end{figure}
+        }
 
@@
         FPGA designs are very often modelled as a collection of Finite State Machines (FSM) like shown in Figure \ref{fig:fsm}.
-        \begin{figure}[]
-            \centering
-            \includegraphics[width=8.0cm]{figures/fsm.png}
-            \caption{\textbf{Above:} Finite state machine (Moore)}
-            \label{fig:fsm}
-        \end{figure}
-        Many times a design will also use pipelining like shown in Figure, either to model signal buffers or shift registers, or to split up large combinational logic into time slices to meet timing constraints.
+        Many times a design will also use pipelining, either to model signal buffers or shift registers, or to split up large combinational logic blocks into time slices to meet timing constraints. 
         These common design structures result in many consecutive sections of combinational logic feeding into a vector of registers. 
@@
         Figure \ref{fig:lut_ff_pair} shows an example of a synthesized LUT-FF Pair.
-        \begin{figure}[]
+
+        {
             \centering
-            \includegraphics[width=8.0cm]{figures/pipelining.png}
-            \caption{\textbf{Above:} Pipelining synthesized as consecutive LUT-FF pairs}
+            \includegraphics[width=\columnwidth]{figures/fsm.png}
+            \captionof{figure}{Finite state machine (Moore)}
+            \label{fig:fsm}
+        }
+
+        {
+            \centering
+            \includegraphics[width=0.9\columnwidth]{figures/pipelining.png}
+            \captionof{figure}{Pipelining synthesized as consecutive LUT-FF pairs}
             \label{fig:pipelining}
-        \end{figure}
+        }
 
-        \begin{figure}[]
+        {
             \centering
-            \includegraphics[width=8.0cm]{figures/lut_ff_pair.png}
-            \caption{A synthesized LUT-FF Pair}
+            \includegraphics[width=0.9\columnwidth]{figures/lut_ff_pair.png}
+            \captionof{figure}{A synthesized LUT-FF Pair}
             \label{fig:lut_ff_pair}
-        \end{figure}
+        }
+        \vspace{1.0cm}
 
@@
         On the bottom left, the cells are placed within the same Site in the same lane, taking advantage of the intra-site routing without burdening the general router with additional inter-site routing.
-        This is an important consideration for minimizing wirelength during placement.
+        This is an important consideration to make while minimizing wirelength during placement.
 
-        \begin{figure}[]
+        \vspace{1.0cm}
+        {
             \centering
-            \includegraphics[width=8.0cm]{figures/intersite_intrasite_2.png}
-            \caption{Intrasite vs Intersite LUT-FF Placement}
+            \includegraphics[width=\columnwidth]{figures/intersite_intrasite_2.png}
+            \captionof{figure}{Intrasite vs Intersite LUT-FF Placement}
             \label{fig:intersite_intrasite}
-        \end{figure}
+        }
+        \vspace{1.0cm}
 
@@
         An FPGA design will also typically implement many adders, counters, subtractors, or comparators, all of which are based on binary addition. 
-        They are so ubiquitous that every that in the 7-Series architecture, every SLICE features a \emph{CARRY4} BEL -- a 4-bit carry-lookahead (CLA) adder. 
-        These CARRY4 blocks can be chained vertically across SLICEs for implementing wide adders efficiently. 
-        These CARRY4 BELs will also be connected to LUTs and FFs, and should be placed in the same Site whenever possible to minimize wirelength. 
-        \begin{figure*}[]
-            \centering
-            \includegraphics[width=16.0cm]{figures/carry_chain_edif_3.png}
-            \caption{A carry chain of size 3 as shown in the Vivado EDIF netlist viewer}
-            \label{fig:carry_chain_edif}
-        \end{figure*}
+        They are so ubiquitous that every that in the 7-Series architecture, every SLICE features a \emph{CARRY4} BEL -- a 4-bit carry-lookahead (CLA) adder, a much better alternative to synthesizing adders via LUTs.
 
     \textbf{CARRY Chains} \quad
+        These CARRY4 blocks can be chained across SLICEs to implement wide adders efficiently. 
+        The CARRY4 BELs \emph{must} be chained vertically consecutively across SLICEs as the Carry-In (CI) and Carry-Out (CO) pins can only be routed this way. 
+        These CARRY4 BELs will also be connected to LUTs and FFs, and should be placed in the same Site whenever possible to minimize wirelength. 
+        Shown in figure \ref{fig:carry_cell_group} shows how a CARRY4 chain and associated LUTs and FFs can be placed across SLICEs.
 
-    \textbf{SLICE Types.}\quad 
-    Xilinx offers two variants of SLICE Sites: \textbf{SLICEL} and \textbf{SLICEM}.  
-    \begin{itemize}
-        \item \emph{SLICEL (SLICE-Logic):} Each LUT in a SLICEL can implement a 6-input Boolean function (with up to two outputs under certain input-sharing conditions).
-During logic synthesis, larger Boolean expressions exceeding six inputs are automatically split across multiple LUTs.
-        \item \emph{SLICEM (SLICE-Memory):} The SLICEM include all the features of a SLICEL but its LUTs can also be configured as small distributed RAM (known as \emph{Distributed RAM}) or as shift registers.
-This local memory capability supplements the larger Block RAM (BRAM) cells and is beneficial for designs requiring numerous small RAM blocks.
-    \end{itemize}
+        \end{multicols}
+        % \vfill
+        {
+            \centering
+            \includegraphics[valign=c, width=3.0cm]{figures/adder_synthesis.png}
+            \includegraphics[valign=c, width=14.0cm]{figures/carry_chain_edif_3.png}
+            \captionof{figure}{A synthesized carry chain of size 3 as shown in the Vivado netlist viewer}
+            \label{fig:carry_chain_edif}
+        }
+        \newpage
+        \begin{multicols}{2}
 
-    \begin{figure}[h]
+        {
             \centering
-        \includegraphics[width=7.5cm]{figures/slicel.png}
-        \caption{A SLICEL Site in the Vivado device viewer}
-        \label{fig:slicel}
-    \end{figure}
+            \includegraphics[width=\columnwidth]{figures/carry_cell_group.png}
+            \captionof{figure}{A SLICEL with a CARRY4 cell, 4 LUT cells, and 4 FF cells placed inside as shown in the Vivado device viewer}
+            \label{fig:carry_cell_group}
+        }
 
-    \begin{figure}[h]
+        {
             \centering
-        \includegraphics[width=7.5cm]{figures/slicem.png}
-        \caption{A SLICEM Site in the Vivado device viewer}
-        \label{fig:slicem}
-    \end{figure}
-
-    In a typical 7-Series device, approximately 75\% of the SLICE Sites are SLICELs and 25\% are SLICEMs. 
-    A single CLB Tile can therefore host either two SLICELs or one SLICEL and one SLICEM.
-
-    ( Explain cell patterns here )
+            \includegraphics[width=4.0cm]{figures/carry_chain_placement.png}
+            \includegraphics[width=4.0cm]{figures/carry_chain_routes_3.png}
+            \captionof{figure}{
+                A CARRY4 chain of size 3 placed across 3 SLICEs.
+                \textbf{Left:} Simplified view, \textbf{Right:} As shown in the Vivado device viewer.
+            }
+            \label{fig:carry_cell_group}
+        }
 
@@
 
+\newpage
 \section{FPGA Design Flow and Toolchain}
@@
 
-    \begin{figure}[]
+    {
         \centering
-    \includegraphics[width=7.5cm]{figures/design_flow.png}
-    \caption{Typical FPGA design and verification workflow.}
+        \includegraphics[width=0.9\columnwidth]{figures/design_flow.png}
+        \captionof{figure}{A typical FPGA design and verification workflow.}
         \label{fig:design_flow}
-    \end{figure}
+    }
 
@@
 
-    \begin{figure}[]
+    {
         \centering
-        \includegraphics[width=7.5cm]{figures/vivado_dcps.png}
-        \caption{RapidWright workflow integrating into the default Vivado design flow.}
+        \includegraphics[width=0.8\columnwidth]{figures/vivado_dcps.png}
+        \captionof{figure}{RapidWright workflow integrating into the default Vivado design flow.}
         \label{fig:vivado_dcps}
-    \end{figure}
+    }
 
@@
     Here we outline in detail each substage of our implementation: PrePacking, Packing, and Placement. 
-    Shown in Figure
+    Shown in Figure ....
 
-    \begin{figure}[]
+    {
         \centering
-        \includegraphics[width=8.0cm]{figures/edif_design_device.png}
-        \caption{}
-    \end{figure}
+        \includegraphics[width=0.9\columnwidth]{figures/edif_design_device.png}
+        \captionof{figure}{Our placement workflow}
+        \label{fig:edif_design_device}
+    }
 
-    \begin{figure*}[]
+    \end{multicols}
+    {
         \centering
-        \includegraphics[width=14.0cm]{figures/substages.png}
-        \caption{The data classes populated at each substage: PrepackedDesign, PackedDesign, and PlacedDesign.}
+        \includegraphics[width=0.8\columnwidth]{figures/substages.png}
+        \captionof{figure}{The data classes populated at each substage: PrepackedDesign, PackedDesign, and PlacedDesign.}
         \label{fig:substages}
-    \end{figure*}
+    }
+    \begin{multicols}{2}
 
@@
 
-    \begin{figure*}[t]
-        \centering
-        \includegraphics[width=14.0cm]{figures/substages.png}
-        \caption{The data classes populated at each substage: PrepackedDesign, PackedDesign, and PlacedDesign.}
-        \label{}
-    \end{figure*}
-
     \subsection{PrePacking}
@@
 
-            \begin{figure}[H]
-                \centering
-                \includegraphics[width=8.0cm]{figures/fsm.png}
-                \caption{A Finite State Machine (Moore)}
-            \end{figure}
-
             In the vast majority of HDL designs, LUTs and FFs will be synthesized in pairs, as is the case in Figure \ref{fig:lut_ff_pair}. 
@@
 
-            \begin{figure}[H]
-                \centering
-                \includegraphics[width=8.0cm]{figures/lut_ff_pair.png}
-                \caption{
-                    Example of a LUT-FF Cell Pair as shown the Vivado netlist viewer. 
-                    FDRE is a type of Flip Flop. 
-                }
-            \end{figure}
-
-            \begin{figure}[H]
-                \centering
-                \includegraphics[width=9.0cm]{figures/intersite_intrasite_2.png}
-                \caption{
-                    \textbf{Upper Left}: A LUT-FF cell pair in the netlist.
-                    \textbf{Upper Right}: The LUT-FF pair placed within the same Site.
-                    \textbf{Bottom}: The LUT-FF pair placed across different Sites.
-                }
-            \end{figure}
-
         \subsection{CARRY Chains}
-            \begin{figure}[]
-                \centering
-                \includegraphics[width=5.0cm]{figures/carry_cell_edif.png}
-                \includegraphics[width=3.0cm]{figures/carry_cell_device.png}
-                \caption{
-                    \textbf{Left}: A CARRY4 EDIFCell as seen in the Vivado netlist viewer.
-                    \textbf{Right}: The corresponding CARRY4 Cell as seen in the device viewer.
-                }
-            \end{figure}
-
-            \begin{figure*}[t]
-                \centering
-                \includegraphics[width=\textwidth]{figures/carry_chain_edif.png}
-                \caption{Example of a CARRY4 chain of length 6 as seen in the Vivado netlist viewer.}
-            \end{figure*}
 
@@
             CARRY chains span vertically across multiple SLICE Sites/Tiles. (Show picture)
-
-            \begin{figure}
-                \centering
-                \includegraphics[width=3.0cm]{figures/carry_chain_routes.png}
-                \includegraphics[width=3.0cm]{figures/dsp_cascade_routes.png}
-                \caption{\textbf{Left}: A Carry Chain of length 6 spanning 6 Sites across 6 Tiles. \textbf{Right}: A DSP cascade of length 2 spanning 2 Sites across 1 Tile.}
-            \end{figure}
 
@@
 
+\end{multicols}
+
 \end{document}

### 2025-04-21 latex/paper/paper.tex
dsp cascades

@@
         The CARRY4 BELs \emph{must} be chained vertically consecutively across SLICEs as the Carry-In (CI) and Carry-Out (CO) pins can only be routed this way. 
-        These CARRY4 BELs will also be connected to LUTs and FFs, and should be placed in the same Site whenever possible to minimize wirelength. 
+        These CARRY4 BELs may also be directly connected to LUTs and FFs, and should be placed in the same Site whenever possible to minimize wirelength. 
         Shown in figure \ref{fig:carry_cell_group} shows how a CARRY4 chain and associated LUTs and FFs can be placed across SLICEs.
@@
             \centering
-            \includegraphics[width=4.0cm]{figures/carry_chain_placement.png}
-            \includegraphics[width=4.0cm]{figures/carry_chain_routes_3.png}
+            \includegraphics[valign=c, width=4.0cm]{figures/carry_chain_placement.png}
+            \includegraphics[valign=c, width=4.0cm]{figures/carry_chain_routes_3.png}
             \captionof{figure}{
@@
     \textbf{DSPs} \quad 
+        FPGAs are often used as low latency Digital Signal Processing (DSP) accelerators. 
+        Common DSP subsystems like Finite Impulse Response (FIR) filters, Fast Fourier Transform (FFTs), and convolutional neural nets (CNNs) demand fast large scale multiply-accumulate (MAC) capabilities.
+        The 7-Series arhictecture integrates DSP BELs into the logic fabric called DSP48E1 that can facilitate MAC efficiently.
+
+        {
+            \centering
+            \includegraphics[width=\columnwidth]{figures/dsp_diagram.png}
+            \captionof{figure}{
+                Basic DSP48E1 Slice Functionality
+            }
+            \label{dsp_diagram}
+        }
 
     \textbf{DSP Cascades} \quad 
+        Higher level DSP functions are supported by cascading DSP48E1 slices in a DSP48E1 column.
+        Much like CARRY4 chains, DSP48E1 cascades must necessarily be placed vertically consecutively across DSP48E1 Sites. 
+        They are connected by three 48-bit signal busses: ACOUT to ACIN, BCOUT to BCIN, PCOUT to PCIN.
+        These signal busses run directly between the vertical DSP48E1 slices without burdening the general routing resources. 
+        The ability to cascade this way provides a high-performance and low-power implementation of DSP filter functions.
+
+        {
+            \centering
+            \includegraphics[valign=c, width=3.0cm]{figures/dsp_cascade_placement.png}
+            \includegraphics[valign=c, width=4.0cm]{figures/dsp_cascade_routed.png}
+            \captionof{figure}{
+                A DSP48E1 cascade of size 2 placed across 2 DSP48E1 Sites.
+                \textbf{Left:} Simplified view, \textbf{Right:} As shown in the Vivado device viewer.
+            }
+        }
 

### 2025-04-21 latex/paper/paper.tex
more dsp casc

@@
 
+    \newpage
+    \subsection{CLB SLICEs}
+
     In the 7-Series architecture, the term \emph{CLB} (Configurable Logic Block) refers to a \emph{CLB Tile} that contains two \emph{SLICE} Sites. 
@@
 
-    \subsection{7-Series BEL Library}
-
     \textbf{LUTs} \quad
@@
         }
+        \vspace{1.0cm}
 
@@
 
+    \newpage
+    \subsection{DSP Slices}
+
     \textbf{DSPs} \quad 
@@
         }
+        \vspace{0.5cm}
 
@@
 
+        \vspace{1.0cm}
         {
             \centering
-            \includegraphics[valign=c, width=3.0cm]{figures/dsp_cascade_placement.png}
-            \includegraphics[valign=c, width=4.0cm]{figures/dsp_cascade_routed.png}
+            \includegraphics[valign=c, width=2.5cm]{figures/dsp_cascade_placement.png}
+            \includegraphics[valign=c, width=3.0cm]{figures/dsp_cascade_routed.png}
             \captionof{figure}{
@@
         }
+
+        \end{multicols}
+        {
+            \centering
+            \includegraphics[valign=c, width=4.0cm]{figures/mult_synthesis.png}
+            \includegraphics[valign=c, width=13.0cm]{figures/dsp_synthesis.png}
+            \captionof{figure}{
+                Simple multiplier synthesis from user design.
+            }
+            \label{mult_synthesis}
+        }
+        \begin{multicols}{2}
+
+    \newpage
 

### 2025-04-22 latex/paper/paper.tex
minor edit
minor edits

@@
     Since the early-to-mid 2000s, the stagnation of single-processor performance relative to the rapid increase in integrated circuit sizes has led to a design productivity gap, where the computational effort for designing complex chips continues to rise.
-    FPGA CAD flows mainly encompass synthesis, placement, and routing; all of which are HP-hard problems, of which placement is one of the most time-consuming processes. 
+    FPGA CAD flows mainly encompass synthesis, placement, and routing; all of which are NP-hard problems, of which placement is one of the most time-consuming processes. 
     Inefficienct placement strategy not only extends design times from hours to days, thereby elevating cost and reducing engineering productivity, but also limits the broader adoption of FPGAs by software engineers who expect compile times akin to those of conventional software compilers like {\tt gcc}.
@@
         In a typical HDL design, the vast majority of FFs will be synthesized as FDREs with the occasional FDSE, as it is generally good practice to keep FPGA designs synchronous. 
-        A FF BEL may also host a latch cell, however, since they are generally bad practice in FPGA design, we will not consider latches in this paper. 
+        A FF BEL may also host a LATCH Cell, however, since they are generally bad practice in FPGA design, we will not consider latches in this paper.
+
         {
@@
         The CARRY4 BELs \emph{must} be chained vertically consecutively across SLICEs as the Carry-In (CI) and Carry-Out (CO) pins can only be routed this way. 
-        These CARRY4 BELs may also be directly connected to LUTs and FFs, and should be placed in the same Site whenever possible to minimize wirelength. 
+        A CARRY4 cell may also directly connect to LUTs and FFs, and should be placed in the same Site whenever possible to minimize wirelength. 
         Shown in figure \ref{fig:carry_cell_group} shows how a CARRY4 chain and associated LUTs and FFs can be placed across SLICEs.
@@
         Much like CARRY4 chains, DSP48E1 cascades must necessarily be placed vertically consecutively across DSP48E1 Sites. 
-        They are connected by three 48-bit signal busses: ACOUT to ACIN, BCOUT to BCIN, PCOUT to PCIN.
+        They are connected by three busses: ACOUT to ACIN, BCOUT to BCIN, PCOUT to PCIN.
         These signal busses run directly between the vertical DSP48E1 slices without burdening the general routing resources. 

### 2025-04-22 latex/paper/paper.tex
brams

@@
     }
+    \vspace{0.5cm}
 
@@
     They constitute the fundamental building blocks for implementing digital circuits on the FPGA.
+    Note how the Tile arrangement is columnar, where each column will have only one Tile type.
 
@@
         Figure \ref{fig:ff_synthesis} shows an example of where a flip flop (FF) is typically synthesized. 
-        The cell primitive FDRE is a type of FF and belongs to a family of D Flip Flops (DFFs) with Clock Enable (CE). 
+        The cell primitive \textbf{FDRE} is a type of FF and belongs to a family of D Flip Flops (DFFs) with Clock Enable (CE). 
         \begin{itemize}
-            \item \emph{FDCE - DFF with CE and Asynchronous Clear}
-            \item \emph{FDPE - DFF with CE and Asynchronous Preset}
-            \item \emph{FDSE - DFF with CE and Synchronous Set}
-            \item \emph{FDRE - DFF with CE and Synchronous Reset}
+            \item FDCE - DFF with CE and Asynchronous Clear
+            \item FDPE - DFF with CE and Asynchronous Preset
+            \item FDSE - DFF with CE and Synchronous Set
+            \item \textbf{FDRE} - DFF with CE and Synchronous Reset
         \end{itemize}
@@
         The 7-Series arhictecture integrates DSP BELs into the logic fabric called DSP48E1 that can facilitate MAC efficiently. 
+        The architecture hierarchy for DSPs is simple compared to CLBs and SLICEs. 
+        A DSP48E1 Tile contains two DSP48E1 Sites, each containing one DSP48E1 BEL, which can host a DSP48E1 Cell.
 
@@
 
-    \textbf{BRAMs} \quad
+    \subsection{Block RAM}
+        In addition to SLICEs and DSPs, the 7-Series also offers dedicated Block Random Access Memory (BRAM) BELs. 
+        These BRAMs come in two variants: \textbf{RAMB18E1} and \textbf{RAMB36E1}. 
+        \begin{itemize}
+            \item \textbf{RAMB18E1} - 
+            Has a capacity of 18 Kilobits. 
+            It can be configured as single port RAM with dimensions ranging between (1-bit wide by 16K deep) to (18-bit wide by 1024 deep). 
+            It can also be configured as a (36-bit wide by 512 deep) true simple dual port RAM. 
+            \item \textbf{RAMB36E1} - 
+            Has a capacity of 36 Kilobits. 
+            It can be configured as single port RAM with dimensions ranging between (1-bit wide by 32K deep) to (36-bit wide by 1024 deep). 
+            It can also be configured as a (72-bit wide by 512 deep) simple dual port RAM. 
+        \end{itemize}
 
-    \textbf{Loose Single Cells} \quad
+        For readability, we will abbreviate RAMB18E1 as \textbf{RAM18} and RAMB36E1 as \textbf{RAM36}.
+        One BRAM Tile contains one RAM36 Site and two RAM18 Sites. 
+        The RAM36 Site contains one RAM36 BEL while the RAM18 Site contains one RAM18 BEL.
+        The RAM18 BEL can host one RAM18 Cell, while the RAM36 BEL can host either one RAM36 Cell or two RAM18 Cells.
+
+        Like DSP cascades, BRAMs may also be cascaded together in a column to implement large memories efficiently, with some intermediate signals between them routed intra-Tile without burdening the general routing.
+        However, unlike DSPs, large memories decomposed amongst multiple BRAMs can also be routed together through general routing. 
+        Furthermore, in most design scenarios, large memories will not utilize the intra-Tile signals.
+
+        For simplicity we will not constrain large BRAMs to be cascaded together in consecutive vertical Tiles. 
+        We can essentially treat RAMB18E1 and RAMB36E1 Cells as loose, minimally constrained single cells, in contrast to the highly constrained DSP cascades and CARRY chains.
+
+        {
+            \centering
+            \includegraphics[valign=c, width=3.0cm]{figures/bram.png}
+            \includegraphics[valign=c, width=2.0cm]{figures/bram_tile.png}
+            \captionof{figure}{
+                A BRAM Tile containing two RAMB18E1 Sites and one RAMB36E1 Site.
+                \textbf{Left:} simplified view. \textbf{Right:} as seen in the device viewer, BELs highlighted in white. 
+            }
+            \label{fig:bram}
+        }
+
+    \subsection{Further Documentation}
 

### 2025-04-25 latex/paper/paper.tex
bram synth
prepack

@@
         }
+        \vspace{1.0cm}
 
+        Up to eight (8) FFs can be placed within the same SLICE, but only if they all share a common Clock-Enable (CE) net and a common Set-Reset (SR) net.
+        This is because the SLICE has only one CE pin and one SR pin to interface with general routing. 
+        The CE and SR signals from these pins are broadcast intra-Site to all FFs within.
+
+    \newpage
     \textbf{LUT-FF Pairs} \quad
@@
 
+        Shown in Figure \ref{fig:intersite_intrasite} are two possible placements for a LUT-FF Pair on the physical device. 
+        On the right, the cells are placed across different Sites, thus the only way to route the net between the cells is through general inter-site routing. 
+        On the bottom left, the cells are placed within the same Site in the same lane, taking advantage of the intra-site routing without burdening the general router with additional inter-site routing.
+        This is an important consideration to make while minimizing wirelength during placement.
+
+        \vspace{1.0cm}
         {
@@
         }
+
+        \vfill
+        .
+        \newcolumn
 
@@
 
-        Shown in Figure \ref{fig:intersite_intrasite} are two possible placements for a LUT-FF Pair on the physical device. 
-        On the right, the cells are placed across different Sites, thus the only way to route the net between the cells is through general inter-site routing. 
-        On the bottom left, the cells are placed within the same Site in the same lane, taking advantage of the intra-site routing without burdening the general router with additional inter-site routing.
-        This is an important consideration to make while minimizing wirelength during placement.
-
-        \vspace{1.0cm}
         {
@@
         }
-        \vspace{1.0cm}
 
+    \newpage
     \textbf{CARRY} \quad
@@
         Shown in figure \ref{fig:carry_cell_group} shows how a CARRY4 chain and associated LUTs and FFs can be placed across SLICEs.
-
-        \end{multicols}
-        % \vfill
-        {
-            \centering
-            \includegraphics[valign=c, width=3.0cm]{figures/adder_synthesis.png}
-            \includegraphics[valign=c, width=14.0cm]{figures/carry_chain_edif_3.png}
-            \captionof{figure}{A synthesized carry chain of size 3 as shown in the Vivado netlist viewer}
-            \label{fig:carry_chain_edif}
-        }
-        \newpage
-        \begin{multicols}{2}
 
@@
         }
+
+        \end{multicols}
+        % \vfill
+        {
+            \centering
+            \includegraphics[valign=c, width=3.0cm]{figures/adder_synthesis.png}
+            \includegraphics[valign=c, width=14.0cm]{figures/carry_chain_edif_3.png}
+            \captionof{figure}{A synthesized carry chain of size 3 as shown in the Vivado netlist viewer}
+            \label{fig:carry_chain_edif}
+        }
+        \begin{multicols}{2}
+        \vspace{1.0cm}
+        .
 
@@
             }
-            \label{fig:bram}
+            \label{fig:bram_tile}
+        }
+
+        {
+            \centering
+            \includegraphics[valign=c, width=4.0cm]{figures/bram_synthesis.png}
+            \includegraphics[valign=c, width=\columnwidth]{figures/bram_edif.png}
+            \captionof{figure}{
+                An example of BRAM synthesis (via inference)
+            }
+            \label{fig:bram_synthesis}
         }
@@
             In parallel to the design flow, simulations and testbenches validate correctness of the user's design at multiple abstraction levels. 
-            Engineers may begin with functional or behavioral simulations, then progress to post-synthesis simulations, and finally to post-implementation simulations that incorporate real routing delays.
+            Engineers may begin with behavioral simulations, then progress to post-synthesis simulations, and finally to post-implementation simulations that incorporate estimated routing delays. 
             With each higher level of fidelity, computational requirements grow significantly due to increasing complexity and the need to analyze more variables over time. 
@@
     Here we outline in detail each substage of our implementation: PrePacking, Packing, and Placement. 
-    Shown in Figure ....
+    Shown in Figure \ref{fig:edif_design_device} shows an overview of the placement workflow.
 
@@
 
+        \end{multicols}
+        {
+            \centering
+            \includegraphics[width=\columnwidth]{figures/raw_edif.png}
+            \captionof{figure}{A raw EDIF Netlist with uncountably many cells}
+            \label{fig:raw_edif}
+        }
+        \begin{multicols}{2}
+
+        The first step in our placement flow is \textbf{prepacking}. 
+        Recall from the 7-Series architecture that there are certain cell groups that must adhere to certain placements constraints to ensure legality, and by design, to minimize wirelength. 
+        CARRY4 chains must necessarily be placed vertically and consecutively across a column of SLICEs in ascending order. 
+        Likewise, DSP48E1 cascades must necessarily be placed vertically and consecutively across a column of DSP48E1 Sites in ascending order. 
+        A LUT-FF may be placed freely, but should be placed within the same SLICE within the same lane to minimize wirelength.
+        A group of LUT-FF pairs may be placed in the same SLICE, with the constraint that the FFs must share the same Clock-Enable (CE) and Set-Reset (SR) nets.
+        Clustering LUT-FF pairs like this can reduce the redundancy of having to route the same CE and SR nets to many SLICEs by routing the nets to fewer SLICEs and connecting them to the individual FFs within via intra-Site routing, and at the same time, packing greater logic density over a smaller area of the device.
+
+        The job of the prepacker is to consolidate functionally related cells from the raw post-synthesis netlist into clusters or groups of clusters that correspond naturally to the physical FPGA architecture.
+        We traverse the netlist to detect and identify LUT-FF pairs, CARRY4 chains, and DSP48E1 cascades.
+
+        LUT-FF pairs, CARRY4 chains, and DSP cascades.
+        In this stage, we work solely at the \texttt{edif} level.
+
         In the context of implementing a custom \emph{Simulated Annealing} (SA) placer, \textbf{prepacking} is the first step in consolidating functionally related cells from the raw post-synthesis netlist into logical clusters that correspond more naturally to the physical FPGA structures.
-        We traverse the raw post-synthesis netlist via RapidWright’s \texttt{edif} packages to identify LUT-FF Pairs, CARRY4 Chains, and DSP Cascades.
+        We take a synthesized design checkpoint and open it via RapidWright. 
+        We traverse the raw post-synthesis netlist via RapidWright’s \texttt{edif} package to identify LUT-FF Pairs, CARRY4 Chains, DSP Cascades, and remaining single cells.
+        We cluster these detected cell structures into their respective \texttt{List}s.
+
+        First, we build the vocabulary of primitive cells by identifying the set of all unique cell types in the EDIF Netlist. 
+        We build a \texttt{HashMap}, where the keys are the cell type and the values are \texttt{List}s.
 

### 2025-04-25 java/src/main/java/placer/Main.java

@@ -78,9 +78,9 @@ public class Main {
             // Sites via simulated annealing, analytical, electrostatic placement, etc.
             // works entirely on the SiteInst/Site/Tile level.
             List<PlacerAnnealRandom> SAPlacers = new ArrayList<PlacerAnnealRandom>();
-            SAPlacers.add(new PlacerAnnealRandom(rootDir, design, device, region));
+            // SAPlacers.add(new PlacerAnnealRandom(rootDir, design, device, region));
             SAPlacers.add(new PlacerAnnealHybrid(rootDir, design, device, region));
-            SAPlacers.add(new PlacerAnnealMidpoint(rootDir, design, device, region));
+            // SAPlacers.add(new PlacerAnnealMidpoint(rootDir, design, device, region));
             // SAPlacers.add(new PlacerGreedyRandom(rootDir, design, device, region));
             // SAPlacers.add(new PlacerGreedyMidpoint(rootDir, design, device, region));
 

### 2025-04-25 latex/paper/paper.tex
woof

@@
         The first step in our placement flow is \textbf{prepacking}. 
-        Recall from the 7-Series architecture that there are certain cell groups that must adhere to certain placements constraints to ensure legality, and by design, to minimize wirelength. 
+        Recall from the 7-Series architecture that there are certain cell structures that must adhere to certain placements constraints to ensure legality, and by design, to minimize wirelength. 
+        The job of the prepacker is to consolidate the cells in the EDIF Netlist into clusters or groups of clusters that naturally reflect these placement constraints.
+
         CARRY4 chains must necessarily be placed vertically and consecutively across a column of SLICEs in ascending order. 
         Likewise, DSP48E1 cascades must necessarily be placed vertically and consecutively across a column of DSP48E1 Sites in ascending order. 
-        A LUT-FF may be placed freely, but should be placed within the same SLICE within the same lane to minimize wirelength.
+        A LUT-FF pair may be placed freely, but should be placed in the same lane within the same SLICE to minimize wirelength.
+
+        The raw post-synthesis netlist only tells us the instantiated cells and the nets between them. 
+        It does not report the presence of any cell structures (CARRY4 chains, etc.). 
+        Thus, we must traverse the netlist to detect these cell structures and store that structure information in a class we will call \texttt{PrepackedDesign}.
+
         A group of LUT-FF pairs may be placed in the same SLICE, with the constraint that the FFs must share the same Clock-Enable (CE) and Set-Reset (SR) nets.
         Clustering LUT-FF pairs like this can reduce the redundancy of having to route the same CE and SR nets to many SLICEs by routing the nets to fewer SLICEs and connecting them to the individual FFs within via intra-Site routing, and at the same time, packing greater logic density over a smaller area of the device.
+        Recall that up to eight FFs may be placed in the same SLICE, thus theoretically, up to eight LUT-FF pairs can be placed in the same SLICE.
 
-        The job of the prepacker is to consolidate functionally related cells from the raw post-synthesis netlist into clusters or groups of clusters that correspond naturally to the physical FPGA architecture.
-        We traverse the netlist to detect and identify LUT-FF pairs, CARRY4 chains, and DSP48E1 cascades.
+        {
+            \centering
+            \includegraphics[width=\columnwidth]{figures/slicel.png}
+            \captionof{figure}{A SLICEL Site}
+            \label{fig:slicel}
+        }
 
-        LUT-FF pairs, CARRY4 chains, and DSP cascades.
-        In this stage, we work solely at the \texttt{edif} level.
+        Utilizing all 8 LUT-FF lanes in a SLICE can help reduce device area utilization and minimize wirelength, but \emph{too much} logic density in an area can contribute to general routing congestion.
+        Furthermore, attempting to fill all 8 lanes in a SLICE requires some careful consideration of LUT BEL constraints. 
+        The LUT BELs in a SLICE, as shown in Figure \ref{fig:slicel}, the LUTs appear to be interleaved.
 
-        In the context of implementing a custom \emph{Simulated Annealing} (SA) placer, \textbf{prepacking} is the first step in consolidating functionally related cells from the raw post-synthesis netlist into logical clusters that correspond more naturally to the physical FPGA structures.
-        We take a synthesized design checkpoint and open it via RapidWright. 
-        We traverse the raw post-synthesis netlist via RapidWright’s \texttt{edif} package to identify LUT-FF Pairs, CARRY4 Chains, DSP Cascades, and remaining single cells.
-        We cluster these detected cell structures into their respective \texttt{List}s.
+        Show a block diagram of a LUT6 comprising of two LUT5s.
 
-        First, we build the vocabulary of primitive cells by identifying the set of all unique cell types in the EDIF Netlist. 
-        We build a \texttt{HashMap}, where the keys are the cell type and the values are \texttt{List}s.
-
-        \textbf{LUT-FF pairs:} \quad
-            Most synchronous designs implement finite-state machines or pipelined datapaths, mapping their combinational logic into LUTs directly feeding FFs. 
-            Clustering each LUT-FF pair under a single “prepacked object” facilitates usage of dedicated intra-SLICE connections, thereby reducing routing overhead.
-
-        \textbf{CARRY4 chains:} \quad
-            Arithmetic operations (adders, counters, comparators) often chain multiple \emph{CARRY4} cells vertically. 
-            During prepacking, it is crucial to detect these chains so they remain grouped and placed in physically consecutive and vertical SLICEs for placement legality and optimal performance.
-
-        \textbf{DSP cascades:} \quad
-            In designs with large multiply-accumulate structures, DSP48E1 blocks can be used and cascaded along dedicated paths. 
-            Identifying such cascades early helps the placer keep them localized, minimizing high-speed interconnect usage.
-            Like carry chains, each DSP in a DSP cascades must be placed vertically, in physically consecutive and vertical DSP48E1 Sites.
-
-        \textbf{Loose or Single Cells} \quad
-            The remaining cells 
-            Various design-specific macros (e.g., BRAM-based FIFOs, shift-register logic) also benefit from being clustered, though these may be addressed in subsequent or more advanced stages.
-
-        This strategy not only reduces the complexity of the placement search space but also helps ensure legality (e.g., shared control signals, shared carry nets). 
-        The subsequent \emph{packing} and \emph{placement} steps can then operate on these higher-level groupings with improved efficiency.
+        Recall that a LUT can accommodate one 6-input boolean function or two 5-input boolean functions sharing the same inputs or any two 3-input or less boolean functions regardless of shared inputs.
+        A LUT6 Cell (a LUT with 6-inputs) will actually occupy two lanes in a SLICEL, rendering one of the FF BELs in either lane ineligible for another LUT-FF pair. 
+        The LUT BELs
 
@@
         This is where simulated annealing actually begins.
-
-\newpage
-o
-\newpage
-%%% OLD
-
-\section{OLD: Simulated Annealing}
-    With a basic understanding of FPGA architecture, design placement, and RapidWright, we have all the necessary pieces to implement our SA placer. 
-    Here we outline in detail each substage of our implementation: PrePacking, Packing, and Placement.
-
-    \subsection{PrePacking}
-        In the prepacking stage, we traverse the raw post-synthesis EDIF Netlist to identify common recurring Cell patterns. 
-        We do this because there are certain desirable cell configurations that should be clustered together to exploit the physical organization of the BELs and Sites of the architecture. 
-        Doing so will innately improve wirelength minimization. 
-        There are also some cell configurations that must necessarily be placed in certain ways in relation to each other to ensure legality within the architecture constraints.
-
-        Furthermore, certain cell patterns are specific to certain cell types. 
-        Therefore, we must first traverse the entire EDIF netlist and identify all unique cell types and group them together via a {\tt HashMap<String, List<EDIFHierCellInst>>}, where key String is the name cell type and value {\tt List<EDIFHierCellInst>} is the list of cells of that type.
-        The resulting lists in the hashmap are mutually exclusive.
-
-        Cell types can include {\tt CARRY4}, {\tt LUT1-6}, {\tt FF}, {\tt DSP48E1}, {\tt RAMB18E1}, and others. 
-        We specifically look for CARRY4 chains, DSP48E1 cascades, and LUT-FF pairs as these patterns are common to nearly all FPGA designs. 
-        We will describe each cell pattern in detail.
-
-        \subsection{LUT-FF Pairs}
-            Very often, engineers model the behavior of a digital system as a collection of Finite State Machines (FSM) as shown in Figure \ref{fig:fsm}. 
-            An FSM at its core is a synchronous sequential circuit comprising of a combinational element and a synchronous memory element. 
-            In the 7-Series architecture, combinational elements are synthesized as LUTs, where the truth table of the combinational function is mapped onto the LUT's memory.
-
-            Combinational functions in 7-Series FPGAs are synthesized as Look-Up Tables (LUTs). 
-            The FPGA synthesizer maps the combinational element into LUTs and the memory element onto FFs.
-
-            In the vast majority of HDL designs, LUTs and FFs will be synthesized in pairs, as is the case in Figure \ref{fig:lut_ff_pair}. 
-            Shown in Figure \ref{fig:intrasite_intersite} are two possible placements for this cell net. 
-            On the right, the cells are placed across different Sites, thus the only way to route the net between the cells is through general inter-site routing. 
-            On the bottom left, the cells are placed within the same Site, taking advantage of the intra-site routing and avoiding costly inter-site routing.
-
-            Identify unique CE-SR net pairs. 
-            All FF cells in a Site must share the same CE and SR nets.
-
-        \subsection{CARRY Chains}
-
-            Figure \ref{fig:carry_cell_edif} shows an example of a CARRY4 cell while figure \ref{fig:carry_chain_edif} shows an example of a CARRY4 chain of size 6.
-
-            As the name implies, CARRY chains are chains of CARRY cells. 
-            % On the Ultrascale architecture, the each CLB contains a CARRY8 BEL, an 8-bit CLA adder, allowing for even greater logic density.
-
-            (By now Need to have introduced SLICEL / SLICEM Sites and differentiated them from "CLBs", which are referred to as CLBL or CLBM Tiles)
-            (rip some figures from the 7-series CLB user guide)
-
-            These CARRY4 cells, if chained together, must be placed vertically across CLBs, as each CLB is connected vertically by a Carry In (CI) to Carry Out (COUT) wire between them. 
-            The only way to route the carry net is through this wire, thus why these CLBs must be placed consecutively vertically. 
-            This allows the facilitation of fast adders on the device.
-
-            The CARRY4 
-            We identify CARRY4 chains by taking the CARRY4 entry of the HashMap. 
-            We follow the below pseudocode
-
-            Carry chain cells are primitive elements that are provided with a group of LUTs to enable more efficient programmable arithmetic. 
-            Primarily it provides dedicated paths for the carry logic of simple arithmetic operations (add, subtract, comparisons, equals, etc).
-Implementing these arithmetic operations with raw LUTs would results in an inefficient use of resources and performance would suffer.
-            In this way, the CARRY4 cell is like a macrocell.
-
-            On the device, they must necessarily be arranged consecutively vertically, from bottom to top. 
-            Figure \ref{fig:carry_chain_edif} shows an example of a CARRY4 EDIFCell chain in an EDIFNetlist. 
-            Notice how there are many different types of cells connected to the CARRY4 Cells. 
-            The raw netlist does not tell us the presence of any carry chains or other cell clusterings. 
-            We must manually traverse the netlist to identify all of the carry chains by accessing each carry cell, accessing its Ports, accessing the net on the port, accessing all of the other ports on the net, and checking if those ports belong to another carry cell.
-            We continue to traverse the carry chain until the final carry cell is not connected to another carry cell. 
-            We must traverse in both directions. 
-            After we reach the tail of the current chain, we traverse in the opposite direction (CIN vs COUT) to find the carry chain anchor. 
-            We then remove all the cells in this chain from the set of all carry cells. 
-            We select the next carry cell in the set of all carry cells and repeat the process, until there are no more remaining carry cells in the set of all carry cells.
-
-            Shown in figure \ref{fig:carry_cell_edif}
-
-            To find CARRY chains, we take the CARRY4 group and follow the below pseudocode. 
-            Each SLICE Site contains one CARRY4 BEL. 
-            Each CLB Tile contains two SLICE Sites. 
-            CARRY chains span vertically across multiple SLICE Sites/Tiles. (Show picture)
-
-        \subsection{DSP Cascades}
-            Each DSP Site contains one DSP BEL. 
-            Each DSP Tile contains two DSP Sites. 
-            DSP chains can span vertically across multiple DSP Sites/Tiles.
-
-    \subsection{Packing}
-        In the packing stage, we take the identified cell clusters and package them into SiteInst Design objects which target the Device Site objects. 
-        \subsection{CLB Sites}
-            Can support LUT-FF pairs, loose LUTs, loose FFs, CARRY chains. 
-            Each SLICE has 8 "lanes" of LUT-FFs. 4 LUT5s and 4 LUT6s. 8 FFs. 
-            For SLICEMs, LUT6s can be configured as shallow 32-bit LUTRAMs or "RAMS32".
-
-    \subsection{Placement: Simulated Annealing} 
-        Bookkeeping: keep track of Sites occupied by single SiteInsts and Sites occupied by SiteInst chains. 
-        Create BEL "fields": CLB, DSP, CLB and DSP Chains, RAMs.
-
-        \subsection{Detailed Placement}
 

### 2025-04-28 latex/paper/paper.tex
dedicated netlist explanation

@@
             The synthesis tool parses the HDL source, performs logical optimizations, and maps the design onto primitive cells that suit the target FPGA technology. 
-            The output is typically a \emph{structural} netlist (e.g., EDIF or structural Verilog) which details how the design’s logic is broken down into LUTs, FFs, and other vendor-specific cells.
+            The output is typically a structural \texttt{netlist} (e.g., EDIF or structural Verilog) which details how the design’s logic is broken down into LUTs, FFs, and other vendor-specific cells.
 
@@
 
+\newpage
 \section{RapidWright API}
@@
 
+\newpage
+\section{What is a Netlist?}
+    \label{sec:netlist}
+    In its simplest form, a netlist consists of a list of every component in an electronic design and a list of nets they connect to. 
+    Here, a net, or network, denotes any group of two or more interconnected elements - whether they be transistors, logic gates, or higher-level macrocells. 
+    More precisely, nets attach to component pins, and those pins reside on cells, each of which exposing multiple ports. 
+    In this way, a netlist makes explicit both the individual building blocks of a circuit and the precise wiring that links them.
+
+    In Vivado, a Netlist can be synthesized as a Hierarchical or a Flattened netlist. 
+    Figure ~\ref{fig:hierarchical_design} shows the Verilog code for a simple HDL design using module instance hierarchy. 
+    Figure ~\ref{fig:hier_netlist} shows the design synthesized into a hierarchical netlist, while Figure ~\ref{fig:flat_netlist} shows the same design synthesized into a flat netlist. 
+    In a synthesized Netlist, the leaf cells (lowest-level cells) are necessarily primitive cells from the vendor primitives library such as LUT6, FDRE, CARRY4, DSP48E1, etc..
+
+    \end{multicols}
+    {
+        \raggedright
+        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/top_level.png}
+        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_0.png}
+        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_1.png}
+        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_2.png}
+        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_3.png}
+        \captionof{figure}{A simple HDL design with module hierarchy.}
+        \label{fig:hierarchical_design}
+    }
+    {
+        \centering
+        \includegraphics[valign=c, width=11.5cm]{figures/netlist_synth/hier_netlist.png}
+        \includegraphics[valign=c, width=6cm]{figures/netlist_synth/hier_graph.png}
+        \captionof{figure}{
+            \textbf{Left:} A hierarchical netlist consisting of LUTs and FFs.
+            \textbf{Right:} The cell hierarchy graph.
+        }
+        \label{fig:hier_netlist}
+    }
+    \vspace{0.5cm}
+    {
+        \centering
+        \includegraphics[valign=c, width=10cm]{figures/netlist_synth/flat_netlist.png}
+        \includegraphics[valign=c, width=4cm]{figures/netlist_synth/flat_graph.png}
+        \captionof{figure}{
+            \textbf{Left:} A flattened netlist consisting of LUTs and FFs.
+            \textbf{Right:} The flattened cell hierarchy graph.
+        }
+        \label{fig:flat_netlist}
+    }
+    \begin{multicols}{2}
+
+    RapidWright offers an easy way to access, read, or manipulate this netlist for a given HDL design. 
+    Shown in Figure~\ref{fig:netlist_syn}
+
+    Verilog \texttt{module}s are represented simply as
+
+    \end{multicols}
+    {
+        \centering
+        \includegraphics[width=\columnwidth]{figures/raw_edif.png}
+        \captionof{figure}{A raw EDIF Netlist with uncountably many cells}
+        \label{fig:raw_edif}
+    }
+    \begin{multicols}{2}
+
+    .
+
+\newpage
 \section{Simulated Annealing}
@@
 
-        \end{multicols}
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/raw_edif.png}
-            \captionof{figure}{A raw EDIF Netlist with uncountably many cells}
-            \label{fig:raw_edif}
-        }
-        \begin{multicols}{2}
-
         The first step in our placement flow is \textbf{prepacking}. 
@@
 
-        The raw post-synthesis netlist only tells us the instantiated cells and the nets between them. 
-        It does not report the presence of any cell structures (CARRY4 chains, etc.). 
+        The raw post-synthesis netlist only tells us the list of nets and the cell ports that they connect to. 
+        It does not report the presence of any macro cell structures (CARRY4 chains, etc.). 
         Thus, we must traverse the netlist to detect these cell structures and store that structure information in a class we will call \texttt{PrepackedDesign}.

### 2025-04-28 latex/paper/paper.tex
minor edits

@@
     \begin{itemize}
-        \item Each \textbf{SLICEL} has a set of BELs including eight LUTs, eight FFs, and one   \emph{CARRY4} adder. The LUT BELs in a SLICEL can only host LUT Cells. 
-        \item The \textbf{SLICEM} includes all the features of a SLICEL but its LUT BELs can host both LUT Cells, which are asynchronous ROM elements, or RAM32M Cells, which are synchronous 32-deep RAM elements.
+        \item Each \texttt{SLICEL} has a set of BELs including eight \texttt{LUT}s, eight \texttt{FF}s, and one \texttt{CARRY4} adder.
+The \texttt{LUT} BELs in a \texttt{SLICEL} can only host \texttt{LUT} Cells.
+        \item The \texttt{SLICEM} includes all the features of a \texttt{SLICEL} but its \texttt{LUT} BELs can host both \texttt{LUT} Cells, which are asynchronous ROM elements, or \texttt{RAM32M} Cells, which are synchronous 32-deep RAM elements.
 These cells are also referred to as \emph{Distributed RAM} in the Xilinx documentation.
-        These cells can offer an alternative to the larger, more dedicated 18K-36K RAMB18E1 cells when RAM resources are highly utilized. 
+            These cells can offer an alternative to the larger, more dedicated 18K-36K \texttt{RAMB18E1} cells when RAM resources are highly utilized. 
     \end{itemize}
-    In a typical 7-Series device, approximately 75\% of the SLICE Sites are SLICELs and 25\% are SLICEMs. 
-    A single CLB Tile can therefore host either two SLICELs or one SLICEL and one SLICEM.
-    For simplicity, however, we will only consider SLICELs for general logic and use the dedicated RAMB18E1 cells to implement RAM elements.
+    In a typical 7-Series device, approximately 75\% of the SLICE Sites are \texttt{SLICEL}s and 25\% are \texttt{SLICEM}s. 
+    A single CLB Tile can therefore host either two \texttt{SLICEL}s or one \texttt{SLICEL} and one \texttt{SLICEM}.
+    To simplify the problem space, however, we will only consider \texttt{SLICEL}s for general logic and use the dedicated \texttt{RAMB18E1} cells to implement RAM elements.
 
     The BELs in these SLICEs facilitate the bulk of the general programmability of the FPGA fabric. 
-    We will explain in detail the motivation and function of these BELs.
+    We will explain in detail the function and motivation behind these BELs.
 
@@
         Combinational logic is universal to all HDL designs. 
-        As the their name suggests, Look‑Up Tables (LUTs) facilitate combinational logic by acting as tiny asynchronously‑accessed ROMs whose contents are fixed when the FPGA is programmed.  
-        For any boolean function, the synthesizer precalculates the output to every possible input combination and stores the resulting truth table into a LUT's static memory. 
+        As the their name suggests, a Look‑Up Table (LUT) map an input value to an output value. 
+        LUTs facilitate combinational logic by acting as tiny asynchronously‑accessed ROMs whose contents are fixed when the FPGA is programmed.  
+        For any boolean function, the synthesizer precalculates the boolean output to every possible input combination and stores the resulting truth table into a LUT's static memory. 
         The inputs are then essentially treated as an address space that maps to a data value space in an asynchronous ROM. 
@@
         For most Verilog users, this generally means signal assignments wrapped in \texttt{always @(posedge clk)} statements. 
-        Figure \ref{fig:ff_synthesis} shows an example of where a flip flop (FF) is typically synthesized. 
-        The cell primitive \textbf{FDRE} is a type of FF and belongs to a family of D Flip Flops (DFFs) with Clock Enable (CE). 
+        Figure \ref{fig:ff_synthesis} shows an example of where a FF is typically synthesized. 
+        The cell primitive \textbf{FDRE} is a type of FF and belongs to a family of D Flip Flops (DFFs) with Clock Enable (\texttt{CE}). 
         \begin{itemize}
-            \item FDCE - DFF with CE and Asynchronous Clear
-            \item FDPE - DFF with CE and Asynchronous Preset
-            \item FDSE - DFF with CE and Synchronous Set
-            \item \textbf{FDRE} - DFF with CE and Synchronous Reset
+            \item \texttt{FDCE} - DFF with CE and Asynchronous Clear
+            \item \texttt{FDPE} - DFF with CE and Asynchronous Preset
+            \item \texttt{FDSE} - DFF with CE and Synchronous Set
+            \item \texttt{FDRE} - DFF with CE and Synchronous Reset
         \end{itemize}
@@
         An FPGA design will also typically implement many adders, counters, subtractors, or comparators, all of which are based on binary addition. 
-        They are so ubiquitous that every that in the 7-Series architecture, every SLICE features a \emph{CARRY4} BEL -- a 4-bit carry-lookahead (CLA) adder, a much better alternative to synthesizing adders via LUTs.
+        They are so ubiquitous that every that in the 7-Series architecture, every SLICE features a \texttt{CARRY4} BEL -- a 4-bit carry-lookahead (CLA) adder, a much better alternative to synthesizing adders via LUTs.
 
     \textbf{CARRY Chains} \quad
-        These CARRY4 blocks can be chained across SLICEs to implement wide adders efficiently. 
-        The CARRY4 BELs \emph{must} be chained vertically consecutively across SLICEs as the Carry-In (CI) and Carry-Out (CO) pins can only be routed this way. 
-        A CARRY4 cell may also directly connect to LUTs and FFs, and should be placed in the same Site whenever possible to minimize wirelength. 
-        Shown in figure \ref{fig:carry_cell_group} shows how a CARRY4 chain and associated LUTs and FFs can be placed across SLICEs.
+        These \texttt{CARRY4} blocks can be chained across SLICEs to implement wide adders efficiently. 
+        The \texttt{CARRY4} BELs \emph{must} be chained vertically consecutively across SLICEs as the Carry-In (\texttt{CI}) and Carry-Out (\texttt{CO}) pins can only be routed this way. 
+        A \texttt{CARRY4} cell may also directly connect to LUTs and FFs, and should be placed in the same Site whenever possible to minimize wirelength. 
+        Shown in figure \ref{fig:carry_cell_group} shows how a \texttt{CARRY4} chain and associated LUTs and FFs can be placed across SLICEs.
 
@@
             \includegraphics[width=\columnwidth]{figures/carry_cell_group.png}
-            \captionof{figure}{A SLICEL with a CARRY4 cell, 4 LUT cells, and 4 FF cells placed inside as shown in the Vivado device viewer}
+            \captionof{figure}{A \texttt{SLICEL} with a \texttt{CARRY4} cell, 4 LUT cells, and 4 FF cells placed inside as shown in the Vivado device viewer}
             \label{fig:carry_cell_group}
@@
             \captionof{figure}{
-                A CARRY4 chain of size 3 placed across 3 SLICEs.
+                A \texttt{CARRY4} chain of size 3 placed across 3 SLICEs.
                 \textbf{Left:} Simplified view, \textbf{Right:} As shown in the Vivado device viewer.
@@
             \includegraphics[valign=c, width=14.0cm]{figures/carry_chain_edif_3.png}
-            \captionof{figure}{A synthesized carry chain of size 3 as shown in the Vivado netlist viewer}
+            \captionof{figure}{A \texttt{CARRY4} chain of size 3 as shown in the Vivado netlist viewer}
             \label{fig:carry_chain_edif}
@@
         The architecture hierarchy for DSPs is simple compared to CLBs and SLICEs. 
-        A DSP48E1 Tile contains two DSP48E1 Sites, each containing one DSP48E1 BEL, which can host a DSP48E1 Cell.
+        A \texttt{DSP48E1} Tile contains two \texttt{DSP48E1} Sites, each containing one \texttt{DSP48E1} BEL, which can host a \texttt{DSP48E1} Cell.
 
@@
             \captionof{figure}{
-                Basic DSP48E1 Slice Functionality
+                Basic \texttt{DSP48E1} Slice Functionality
             }
@@
     \textbf{DSP Cascades} \quad 
-        Higher level DSP functions are supported by cascading DSP48E1 slices in a DSP48E1 column.
-        Much like CARRY4 chains, DSP48E1 cascades must necessarily be placed vertically consecutively across DSP48E1 Sites. 
-        They are connected by three busses: ACOUT to ACIN, BCOUT to BCIN, PCOUT to PCIN.
-        These signal busses run directly between the vertical DSP48E1 slices without burdening the general routing resources. 
-        The ability to cascade this way provides a high-performance and low-power implementation of DSP filter functions.
+        Higher level DSP functions are supported by cascading \texttt{DSP48E1} slices in a \texttt{DSP48E1} column.
+        Much like \texttt{CARRY4} chains, \texttt{DSP48E1} cascades must necessarily be placed vertically consecutively across \texttt{DSP48E1} Sites. 
+        They are connected by three busses: \texttt{ACOUT} to \texttt{ACIN}, \texttt{BCOUT} to \texttt{BCIN}, \texttt{PCOUT} to \texttt{PCIN}.
+        These signal busses run directly between the vertical \texttt{DSP48E1} slices without burdening the general routing resources. 
+        The ability to cascade this way provides a high-performance and low-power implementation of digital signal processing systems.
 
@@
             \captionof{figure}{
-                A DSP48E1 cascade of size 2 placed across 2 DSP48E1 Sites.
+                A \texttt{DSP48E1} cascade of size 2 placed across 2 \texttt{DSP48E1} Sites.
                 \textbf{Left:} Simplified view, \textbf{Right:} As shown in the Vivado device viewer.
@@
         \begin{itemize}
-            \item \textbf{RAMB18E1} - 
+            \item \texttt{RAMB18E1} - 
             Has a capacity of 18 Kilobits. 
@@
             It can also be configured as a (36-bit wide by 512 deep) true simple dual port RAM. 
-            \item \textbf{RAMB36E1} - 
+            \item \texttt{RAMB36E1} - 
             Has a capacity of 36 Kilobits. 
@@
 
-        For readability, we will abbreviate RAMB18E1 as \textbf{RAM18} and RAMB36E1 as \textbf{RAM36}.
-        One BRAM Tile contains one RAM36 Site and two RAM18 Sites. 
-        The RAM36 Site contains one RAM36 BEL while the RAM18 Site contains one RAM18 BEL.
-        The RAM18 BEL can host one RAM18 Cell, while the RAM36 BEL can host either one RAM36 Cell or two RAM18 Cells.
+        One BRAM Tile contains one \texttt{RAMB36E1} Site and two \texttt{RAMB18E1} Sites. 
+        The \texttt{RAMB36E1} Site contains one \texttt{RAMB36E1} BEL, which can host one \texttt{RAMB36E1} Cell. 
+        Likewise, the \texttt{RAMB18E1} Site contains one \texttt{RAMB18E1} BEL, which can host one \texttt{RAMB18E1} Cell.
 
@@
 
-        For simplicity we will not constrain large BRAMs to be cascaded together in consecutive vertical Tiles. 
-        We can essentially treat RAMB18E1 and RAMB36E1 Cells as loose, minimally constrained single cells, in contrast to the highly constrained DSP cascades and CARRY chains.
+        To simplify the problem space, we will not constrain large BRAMs to be cascaded together in consecutive vertical Tiles. 
+        We can essentially treat \texttt{RAMB18E1} and \texttt{RAMB36E1} Cells as loose, minimally constrained single cells, in contrast to the highly constrained \texttt{DSP48E1} cascades and \texttt{CARRY4} chains.
 
@@
             \captionof{figure}{
-                A BRAM Tile containing two RAMB18E1 Sites and one RAMB36E1 Site.
+                A BRAM Tile containing two \texttt{RAMB18E1} Sites and one \texttt{RAMB36E1} Site.
                 \textbf{Left:} simplified view. \textbf{Right:} as seen in the device viewer, BELs highlighted in white. 
@@
 
-    This workflow underscores the critical role of \textbf{placement} in bridging the netlist to a physical realization. 
-    An efficient placement algorithm can drastically reduce compile times and improve design performance, enabling broader adoption of FPGAs in application spaces that require fast design iteration.
+    % This workflow underscores the critical role of \textbf{placement} in bridging the netlist to a physical realization. 
+    % An efficient placement algorithm can drastically reduce compile times and improve design performance, enabling broader adoption of FPGAs in application spaces that require fast design iteration.
 
@@
     In its simplest form, a netlist consists of a list of every component in an electronic design and a list of nets they connect to. 
-    Here, a net, or network, denotes any group of two or more interconnected elements - whether they be transistors, logic gates, or higher-level macrocells. 
+    Generally, a net, or network, denotes any group of two or more interconnected elements - whether they be transistors, logic gates, or in our case, logical cells. 
     More precisely, nets attach to component pins, and those pins reside on cells, each of which exposing multiple ports. 
@@
     Figure ~\ref{fig:hier_netlist} shows the design synthesized into a hierarchical netlist, while Figure ~\ref{fig:flat_netlist} shows the same design synthesized into a flat netlist. 
-    In a synthesized Netlist, the leaf cells (lowest-level cells) are necessarily primitive cells from the vendor primitives library such as LUT6, FDRE, CARRY4, DSP48E1, etc..
+    In a synthesized Netlist, the leaf cells (lowest-level cells) are necessarily primitive cells from the vendor primitives library such as \texttt{LUT6}, \texttt{FDRE}, \texttt{CARRY4}, \texttt{DSP48E1}, etc..
 
@@
         Recall from the 7-Series architecture that there are certain cell structures that must adhere to certain placements constraints to ensure legality, and by design, to minimize wirelength. 
-        The job of the prepacker is to consolidate the cells in the EDIF Netlist into clusters or groups of clusters that naturally reflect these placement constraints.
+        The job of the prepacker is to traverse the raw EDIF netlist, detect these cell structures, and consolidate these cells into clusters or groups of clusters that naturally reflect these placement constraints.
 
-        CARRY4 chains must necessarily be placed vertically and consecutively across a column of SLICEs in ascending order. 
-        Likewise, DSP48E1 cascades must necessarily be placed vertically and consecutively across a column of DSP48E1 Sites in ascending order. 
+        Recall that CARRY4 chains must necessarily be placed vertically and consecutively across a column of SLICEs in ascending order. 
+        Likewise, \textttt{DSP48E1} cascades must necessarily be placed vertically and consecutively across a column of DSP48E1 Sites in ascending order. 
         A LUT-FF pair may be placed freely, but should be placed in the same lane within the same SLICE to minimize wirelength.

### 2025-05-01 latex/paper/paper.tex
augh

@@
     \label{sec:netlist}
-    In its simplest form, a netlist consists of a list of every component in an electronic design and a list of nets they connect to. 
+    In its simplest form, a netlist is a list of every component in an electronic design and a list of nets they connect to. 
     Generally, a net, or network, denotes any group of two or more interconnected elements - whether they be transistors, logic gates, or in our case, logical cells. 
@@
     Figure ~\ref{fig:hierarchical_design} shows the Verilog code for a simple HDL design using module instance hierarchy. 
-    Figure ~\ref{fig:hier_netlist} shows the design synthesized into a hierarchical netlist, while Figure ~\ref{fig:flat_netlist} shows the same design synthesized into a flat netlist. 
-    In a synthesized Netlist, the leaf cells (lowest-level cells) are necessarily primitive cells from the vendor primitives library such as \texttt{LUT6}, \texttt{FDRE}, \texttt{CARRY4}, \texttt{DSP48E1}, etc..
+    Figure ~\ref{fig:hier_netlist} shows the design synthesized into a hierarchical netlist. 
+    The synthesizer attempts to construct the module hierarchy as close to the module instantiation hierarchy as defined by the raw user design entry. 
+    Figure ~\ref{fig:flat_netlist} shows the same design synthesized into a flat netlist. 
+    In a synthesized Netlist, the leaf cells (lowest-level cells) are necessarily primitive cells from the architecture primitives library which includes \texttt{LUT6}, \texttt{FDRE}, \texttt{CARRY4}, \texttt{DSP48E1}, etc..
 
@@
 
-    RapidWright offers an easy way to access, read, or manipulate this netlist for a given HDL design. 
-    Shown in Figure~\ref{fig:netlist_syn}
+    \subsection{Netlist Manipulation in RapidWright}
 
-    Verilog \texttt{module}s are represented simply as
+    One of RapidWright's most powerful features is netlist manipulation via its \texttt{edif} package. 
+    A netlist can first be extracted from a \texttt{.dcp} design checkpoint like the following:
+
+    SHOULD I INCLUDE A RAPIDWRIGHT HELLO WORLD EXAMPLE JUST TO GIVE AN IMPRESSION?
+
+    OR CAN I JUST DIRECT THE READER TO THE RAPIDWRIGHT DOCUMENTATION?
 
@@
         Recall that CARRY4 chains must necessarily be placed vertically and consecutively across a column of SLICEs in ascending order. 
-        Likewise, \textttt{DSP48E1} cascades must necessarily be placed vertically and consecutively across a column of DSP48E1 Sites in ascending order. 
+        Likewise, \texttt{DSP48E1} cascades must necessarily be placed vertically and consecutively across a column of DSP48E1 Sites in ascending order. 
         A LUT-FF pair may be placed freely, but should be placed in the same lane within the same SLICE to minimize wirelength.

### 2025-05-03 latex/paper/sections/02-abstract.tex
split tex file into multiple files

--- 2025-04-28 latex/paper/paper.tex
@@
-\documentclass{article}
-\usepackage{hyperref}
-\usepackage{enumitem}
-\usepackage{graphicx}
-\usepackage{amsmath}
-\usepackage{mathpazo}
-\usepackage{xcolor}
-\usepackage{float}
-\usepackage{listings}
-\lstset{
-  basicstyle=\ttfamily\small,
-  breaklines=true,
-  columns=fullflexible,
-  frame=single,
-  showspaces=false,
-  showtabs=false,
-  keepspaces=false
-}
-\usepackage[export]{adjustbox}
-\usepackage{multicol}
-\usepackage{caption}
-\usepackage[a4paper,
-            bindingoffset=0.2in,
-            left=0.5in,
-            right=0.5in,
-            top=0.5in,
-            bottom=0.5in,
-            footskip=.25in]{geometry}
-
-\begin{document}
-\title{MS Technical Paper: \\ Placement Algorithms for Heterogeneous FPGAs}
-\author{Brian B Cheng \\ Rutgers University Department of Electrical and Computer Engineering}
-
-\date{}
-\maketitle
-
-\begin{multicols}{2}
-
-\section{Keywords}
-\begin{itemize}
-    \item FPGA, EDA, Placement, Simulated Annealing, Optimization, RapidWright
-\end{itemize}
-
 \section{Abstract}
@@
 fdsafdsafdsa. \\
-
-\section{Introduction}
-
-    Field-Programmable Gate Arrays (FPGAs) have witnessed rapid growth in capacity and versatility, driving significant advances in computer-aided design (CAD) and electronic design automation (EDA) methodologies.
-    Since the early-to-mid 2000s, the stagnation of single-processor performance relative to the rapid increase in integrated circuit sizes has led to a design productivity gap, where the computational effort for designing complex chips continues to rise.
-    FPGA CAD flows mainly encompass synthesis, placement, and routing; all of which are NP-hard problems, of which placement is one of the most time-consuming processes. 
-    Inefficienct placement strategy not only extends design times from hours to days, thereby elevating cost and reducing engineering productivity, but also limits the broader adoption of FPGAs by software engineers who expect compile times akin to those of conventional software compilers like {\tt gcc}.
-
-    For these reasons, FPGA placement remains a critical research effort even today. 
-    In this paper, we study and implement established placement methods. 
-    To do this, we use the RapidWright API, which is a semi-open-source research effort from AMD/Xilinx that enables custom solutions to FPGA design implementations and design tools that are not offered by their industry-standard FPGA environment, Vivado.
-    We implement multiple variations of simulated annealing placers for Xilinx's 7-series FPGAs, with an emphasis on minimizing total wirelength while mitigating runtime. 
-    Our implementation is organized into three consecutive substages. 
-    The \textbf{prepacking} stage involves traversing a raw EDIF netlist to identify recurring cell patterns—such as CARRY chains, DSP cascades, and LUT-FF pairs—that are critical for efficient mapping and legalization.
-    In the subsequent \textbf{packing} stage, these identified patterns, along with any remaining loose cells, are consolidated into SiteInst objects that encapsulate the FPGA’s discrete resource constraints and architectural nuances.
-    Finally, the \textbf{placement} stage employs a simulated annealing (SA) algorithm to optimally assign SiteInst objects to physical sites, aiming to minimize total wirelength while adhering to the constraints of the 7-series architecture.
-
-    Simulated annealing iteratively swaps placement objects guided by a cost function that decides which swaps should be accepted or rejected. 
-    Hill climbing is permitted by occasionally accepting moves that increase cost, in hope that such swaps may later lead to a better final solution. 
-    SA remains a popular approach in FPGA placement research due to its simplicity and robustness in handling the discrete architectural constraints of FPGA devices. 
-    While SA yields surprisingly good results given relatively simple rules, it is ultimately a heuristic and stochastic approach that explores the vast placement space by making random moves. 
-    Most of these moves will be rejected, meaning that SA must run many iterations, usually hundreds to thousands, to arrive at a desirable solution.
-
-    In the ASIC domain, where placers must handle designs with millions of cells, the SA approach has largely been abandoned in favor of analytical techniques, owing to SA's runtime and poor scalability.
-    Modern FPGA placers have also followed suit, as new legalization strategies allow FPGA placers to leverage traditionally ASIC placement algorithms and adapt them to the discrete constraints of FPGA architectures.
-    While this paper does not present a working analytical placer, it will explore ways to build upon our existing infrastructure (prepacker and packer) to replace SA with AP.
-
-    The paper first begins by elaborating on general FPGA architecture and then specifically the Xilinx 7-Series architecture. 
-    Then, the paper will elaborate on the FPGA design flow, then the role that the RapidWright API plays in the design flow. 
-    We explain in detail each of these concepts for a broader audience as they provide much needed context for FPGA placement algorithms as a concept. 
-    However, readers who are already familiar with these concepts can skip directly to the RapidWright API section \ref{sec:rapidwright_api} or to the Simulated Annealing section \ref{sec:simulated_annealing}.
-
-\section{FPGA Architecture History}
-
-    Before any work can begin on an FPGA placer, it is necessary to understand both the objects being placed and the medium in which they are placed.
-    Configurable logic devices have undergone significant evolution over the past four decades. 
-    We will briefly review the evolution of configurable logic architecture starting in the 1970s and quickly work our way up to modern day FPGA architecture.
-
-    \textbf{PLA:} \quad
-    The journey began with the Programmable Logic Array (PLA) in the early 1970s. 
-    The PLA implemented output logic using a programmable-OR and programmable-AND plane that formed a sum-of-products equation for each output through programmable fuses. 
-    Around the same time, the Programmable Array Logic (PAL) was introduced. 
-    The PAL simplified the PLA by fixing the OR gates, resulting in a fixed-OR, programmable-AND design, which sacrificed some logic flexibility to simplify its manufacture. 
-    Figure \ref{fig:pal_2} shows one such PAL architecture.
-
-    \textbf{CPLD:} \quad 
-    Later in the same decade came the Complex Programmable Logic Device (CPLD), which took the form of an array of Configurable Logic Blocks (CLBs). 
-    These CLBs were typically modified PAL blocks that included the PAL itself along with macrocells such as flip-flops, multiplexers, and tri-state buffers. 
-    The CPLD functioned as an array of PALs connected by a central programmable switch matrix and could be programmed using a hardware description language (HDL) like VHDL. 
-    Figure \ref{fig:cpld} shows one such CPLD architecture. 
-    \\
-
-    {
-        \centering
-        \includegraphics[width=\columnwidth]{figures/pal_2.png}
-        \captionof{figure}{PAL architecture with 5 inputs, 8 programmable AND gates and 4 fixed OR gates}
-        \label{fig:pal_2}
-        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/Programmable-Logic-Devices/Programmable-Array-Logic-PAL/
-        % https://www.naukri.com/code360/library/difference-between-pla-and-pal 
-    }
-
-    \newcolumn
-
-    \textbf{Homogeneous FPGA:} \quad 
-    The mid-1980s saw the introduction of homogeneous FPGAs, which were built as a grid of CLBs. 
-    Rather than using a central programmable switch matrix as in CPLDs, FPGAs adopted an island style architecture in which each CLB is surrounded on all sides by programmable routing resources, as shown in Figure \ref{fig:fpga}.
-    The first commercially viable FPGA, produced by Xilinx in 1984, featured 16 CLBs arranged in a 4x4 grid. 
-    As FPGA technology advanced, CLBs were redesigned to use lookup tables (LUTs) instead of PAL arrays for greater logic density. 
-    The capacity of an FPGA was often measured by how many logical elements or CLBs it offered, which grew from hundreds to thousands and now to hundreds of thousands of CLBs.
-    \\
-
-    {
-        \centering
-        \includegraphics[width=0.8\columnwidth]{figures/cpld.png}
-        \captionof{figure}{CPLD architecture with 4 CLBs (PAL-like blocks)}
-        \label{fig:cpld}
-        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/CPLD/Complex-Programmable-Logic-Device-CPLDs/
-    }
-    {
-        \centering
-        \includegraphics[width=\columnwidth]{figures/homogenous_fpga.png}
-        \captionof{figure}{A homogeneous island-style FPGA architecture with 16 CLBs in a grid.}
-        \label{fig:fpga}
-        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/FPGA/Field-programmable-gate-array-FPGA/
-    }
-
-    \newpage
-
-    {
-        \centering
-        \includegraphics[width=\columnwidth]{figures/heterogenous_fpga_3.jpg}
-        \captionof{figure}{A heterogeneous island-style FPGA with a mix of CLBs and macrocells.}
-        \label{fig:fpga}
-        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/FPGA/Field-programmable-gate-array-FPGA/
-    }
-    \vspace{0.5cm}
-
-    \textbf{Heterogenous FPGA:} \quad 
-    This brings us to modern day FPGA architectures. 
-    To meet the needs of increasingly complex designs, FPGA vendors introduced heterogeneous FPGAs. 
-    In these devices, hard macros such as Block RAM (BRAM) and Digital Signal Processing (DSP) slices are integrated into the programmable logic fabric along with CLBs, like shown in Figure \ref{fig:fpga}.
-    This design enables the direct instantiation of common subsystems like memories and multipliers, without having to recreate them from scratch using CLBs. 
-    Major vendors such as Xilinx and Altera now employ heterogeneous island-style architectures in their devices. 
-    As designs become increasingly large and complex, FPGAs meet the demand by becoming increasingly heterogenous, incorporating a wider variety of hard macros into the fabric.
-
-\section{Xilinx 7-Series Architecture}
-    \label{sec:7_series}
-    The Xilinx 7-Series devices, first introduced in 2010, follow a heterogeneous island-style architecture as discussed previously. 
-    Although the 7-Series was later superseded in 2013 by the UltraScale architecture, the 7-Series remains highly relevant due to its accessibility, wide availability, and compatibility with open-source tooling.
-    Representative sub-families include Artix-7, Kintex-7, Virtex-7, and Zynq-7000, each designed with different performance and cost trade-offs but all follow the core 7-Series architecture.
-
-    Figure~\ref{fig:hierarchy_4} illustrates a high-level view of the hierarchical organization of a 7-Series FPGA. 
-    At its lowest level, the device consists of a large array of atomic components called \emph{Basic Elements of Logic}~(\textbf{BELs}). 
-    These BELs encompass look-up tables (\textbf{LUTs}), flip-flops (\textbf{FFs}), block RAMs (\textbf{BRAMs}), DSP slices (\textbf{DSPs}), and the configurable interconnect fabric. 
-    They constitute the fundamental building blocks for implementing digital circuits on the FPGA.
-    Note how the Tile arrangement is columnar, where each column will have only one Tile type.
-
-    To manage this complexity, Xilinx organizes these BELs into incrementally abstract structures. 
-    First, \textbf{BELs} are grouped into \textbf{Sites}. 
-    Each Site is embedded into a \textbf{Tile}, and Tiles are further arranged into \textbf{Clock Regions}. 
-    In some high-density devices, multiple Clock Regions may be consolidated into one or more \textbf{Super Logic Regions}~(SLRs). 
-    However, for the scope of this paper, we focus on Xilinx 7-Series devices with only a single SLR.
-
-    \end{multicols}
-    {
-        \centering
-        \includegraphics[valign=c, width=0.3\columnwidth]{figures/hierarchy_5.png}
-        \includegraphics[valign=c, width=0.6\columnwidth]{figures/hierarchy_4.png}
-        \captionof{figure}{Architecture Hierarchy of a Xilinx FPGA}
-        \label{fig:hierarchy_4}
-    }
-    \vfill
-    \begin{multicols}{2}
-
-    \newpage
-    \subsection{CLB SLICEs}
-
-    In the 7-Series architecture, the term \emph{CLB} (Configurable Logic Block) refers to a \emph{CLB Tile} that contains two \emph{SLICE} Sites. 
-    Xilinx offers two variants of SLICE Sites: \textbf{SLICEL} and \textbf{SLICEM}.
-    \begin{itemize}
-        \item Each \textbf{SLICEL} has a set of BELs including eight LUTs, eight FFs, and one   \emph{CARRY4} adder. The LUT BELs in a SLICEL can only host LUT Cells. 
-        \item The \textbf{SLICEM} includes all the features of a SLICEL but its LUT BELs can host both LUT Cells, which are asynchronous ROM elements, or RAM32M Cells, which are synchronous 32-deep RAM elements.
-These cells are also referred to as \emph{Distributed RAM} in the Xilinx documentation.
-        These cells can offer an alternative to the larger, more dedicated 18K-36K RAMB18E1 cells when RAM resources are highly utilized. 
-    \end{itemize}
-    In a typical 7-Series device, approximately 75\% of the SLICE Sites are SLICELs and 25\% are SLICEMs. 
-    A single CLB Tile can therefore host either two SLICELs or one SLICEL and one SLICEM.
-    For simplicity, however, we will only consider SLICELs for general logic and use the dedicated RAMB18E1 cells to implement RAM elements.
-
-    The BELs in these SLICEs facilitate the bulk of the general programmability of the FPGA fabric. 
-    We will explain in detail the motivation and function of these BELs.
-
-    \textbf{LUTs} \quad
-        Combinational logic is universal to all HDL designs. 
-        As the their name suggests, Look‑Up Tables (LUTs) facilitate combinational logic by acting as tiny asynchronously‑accessed ROMs whose contents are fixed when the FPGA is programmed.  
-        For any boolean function, the synthesizer precalculates the output to every possible input combination and stores the resulting truth table into a LUT's static memory. 
-        The inputs are then essentially treated as an address space that maps to a data value space in an asynchronous ROM. 
-        No explicit logic gates like NAND or XNOR are synthesized, contrary to what newcomers might expect from a "Field Programmable Gate Array".
-
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/lut_synthesis.png}
-            \captionof{figure}{LUT synthesis from user design}
-            \label{fig:lut_synthesis}
-        }
-        \vspace{1.0cm}
-
-        In the 7‑Series devices, one LUT can facilitate any 6-input boolean function, or two 5-input functions, as long as they share the same input signals.  
-        The LUT can also host two independent boolean functions of up to 3 inputs each, even when the inputs are not shared.  
-        Functions requiring more than six unique inputs are decomposed across multiple cascaded LUTs.
-        Figure \ref{fig:lut_synthesis} shows an example of where a LUT is typically synthesized in a design entry.
-
-    \textbf{FFs} \quad
-        FFs are synthesized to facilitate synchronous event-driven signal assignment. 
-        For most Verilog users, this generally means signal assignments wrapped in \texttt{always @(posedge clk)} statements. 
-        Figure \ref{fig:ff_synthesis} shows an example of where a flip flop (FF) is typically synthesized. 
-        The cell primitive \textbf{FDRE} is a type of FF and belongs to a family of D Flip Flops (DFFs) with Clock Enable (CE). 
-        \begin{itemize}
-            \item FDCE - DFF with CE and Asynchronous Clear
-            \item FDPE - DFF with CE and Asynchronous Preset
-            \item FDSE - DFF with CE and Synchronous Set
-            \item \textbf{FDRE} - DFF with CE and Synchronous Reset
-        \end{itemize}
-        In a typical HDL design, the vast majority of FFs will be synthesized as FDREs with the occasional FDSE, as it is generally good practice to keep FPGA designs synchronous. 
-        A FF BEL may also host a LATCH Cell, however, since they are generally bad practice in FPGA design, we will not consider latches in this paper.
-
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/ff_synthesis.png}
-            \captionof{figure}{FF synthesis from user design}
-            \label{fig:ff_synthesis}
-        }
-        \vspace{1.0cm}
-
-        Up to eight (8) FFs can be placed within the same SLICE, but only if they all share a common Clock-Enable (CE) net and a common Set-Reset (SR) net.
-        This is because the SLICE has only one CE pin and one SR pin to interface with general routing. 
-        The CE and SR signals from these pins are broadcast intra-Site to all FFs within.
-
-    \newpage
-    \textbf{LUT-FF Pairs} \quad
-        FPGA designs are very often modelled as a collection of Finite State Machines (FSM) like shown in Figure \ref{fig:fsm}.
-        Many times a design will also use pipelining, either to model signal buffers or shift registers, or to split up large combinational logic blocks into time slices to meet timing constraints. 
-        These common design structures result in many consecutive sections of combinational logic feeding into a vector of registers. 
-        The synthesizer naturally synthesizes these structures as consecutive pairs of LUTs feeding into FFs as shown in Figure \ref{fig:pipelining}. 
-        Figure \ref{fig:lut_ff_pair} shows an example of a synthesized LUT-FF Pair.
-
-        Shown in Figure \ref{fig:intersite_intrasite} are two possible placements for a LUT-FF Pair on the physical device. 
-        On the right, the cells are placed across different Sites, thus the only way to route the net between the cells is through general inter-site routing. 
-        On the bottom left, the cells are placed within the same Site in the same lane, taking advantage of the intra-site routing without burdening the general router with additional inter-site routing.
-        This is an important consideration to make while minimizing wirelength during placement.
-
-        \vspace{1.0cm}
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/fsm.png}
-            \captionof{figure}{Finite state machine (Moore)}
-            \label{fig:fsm}
-        }
-        \vspace{1.0cm}
-
-        {
-            \centering
-            \includegraphics[width=0.9\columnwidth]{figures/pipelining.png}
-            \captionof{figure}{Pipelining synthesized as consecutive LUT-FF pairs}
-            \label{fig:pipelining}
-        }
-
-        \vfill
-        .
-        \newcolumn
-
-        {
-            \centering
-            \includegraphics[width=0.9\columnwidth]{figures/lut_ff_pair.png}
-            \captionof{figure}{A synthesized LUT-FF Pair}
-            \label{fig:lut_ff_pair}
-        }
-        \vspace{1.0cm}
-
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/intersite_intrasite_2.png}
-            \captionof{figure}{Intrasite vs Intersite LUT-FF Placement}
-            \label{fig:intersite_intrasite}
-        }
-
-    \newpage
-    \textbf{CARRY} \quad
-        An FPGA design will also typically implement many adders, counters, subtractors, or comparators, all of which are based on binary addition. 
-        They are so ubiquitous that every that in the 7-Series architecture, every SLICE features a \emph{CARRY4} BEL -- a 4-bit carry-lookahead (CLA) adder, a much better alternative to synthesizing adders via LUTs.
-
-    \textbf{CARRY Chains} \quad
-        These CARRY4 blocks can be chained across SLICEs to implement wide adders efficiently. 
-        The CARRY4 BELs \emph{must} be chained vertically consecutively across SLICEs as the Carry-In (CI) and Carry-Out (CO) pins can only be routed this way. 
-        A CARRY4 cell may also directly connect to LUTs and FFs, and should be placed in the same Site whenever possible to minimize wirelength. 
-        Shown in figure \ref{fig:carry_cell_group} shows how a CARRY4 chain and associated LUTs and FFs can be placed across SLICEs.
-
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/carry_cell_group.png}
-            \captionof{figure}{A SLICEL with a CARRY4 cell, 4 LUT cells, and 4 FF cells placed inside as shown in the Vivado device viewer}
-            \label{fig:carry_cell_group}
-        }
-
-        {
-            \centering
-            \includegraphics[valign=c, width=4.0cm]{figures/carry_chain_placement.png}
-            \includegraphics[valign=c, width=4.0cm]{figures/carry_chain_routes_3.png}
-            \captionof{figure}{
-                A CARRY4 chain of size 3 placed across 3 SLICEs.
-                \textbf{Left:} Simplified view, \textbf{Right:} As shown in the Vivado device viewer.
-            }
-            \label{fig:carry_cell_group}
-        }
-
-        \end{multicols}
-        % \vfill
-        {
-            \centering
-            \includegraphics[valign=c, width=3.0cm]{figures/adder_synthesis.png}
-            \includegraphics[valign=c, width=14.0cm]{figures/carry_chain_edif_3.png}
-            \captionof{figure}{A synthesized carry chain of size 3 as shown in the Vivado netlist viewer}
-            \label{fig:carry_chain_edif}
-        }
-        \begin{multicols}{2}
-        \vspace{1.0cm}
-        .
-
-    \newpage
-    \subsection{DSP Slices}
-
-    \textbf{DSPs} \quad 
-        FPGAs are often used as low latency Digital Signal Processing (DSP) accelerators. 
-        Common DSP subsystems like Finite Impulse Response (FIR) filters, Fast Fourier Transform (FFTs), and convolutional neural nets (CNNs) demand fast large scale multiply-accumulate (MAC) capabilities.
-        The 7-Series arhictecture integrates DSP BELs into the logic fabric called DSP48E1 that can facilitate MAC efficiently. 
-        The architecture hierarchy for DSPs is simple compared to CLBs and SLICEs. 
-        A DSP48E1 Tile contains two DSP48E1 Sites, each containing one DSP48E1 BEL, which can host a DSP48E1 Cell.
-
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/dsp_diagram.png}
-            \captionof{figure}{
-                Basic DSP48E1 Slice Functionality
-            }
-            \label{dsp_diagram}
-        }
-        \vspace{0.5cm}
-
-    \textbf{DSP Cascades} \quad 
-        Higher level DSP functions are supported by cascading DSP48E1 slices in a DSP48E1 column.
-        Much like CARRY4 chains, DSP48E1 cascades must necessarily be placed vertically consecutively across DSP48E1 Sites. 
-        They are connected by three busses: ACOUT to ACIN, BCOUT to BCIN, PCOUT to PCIN.
-        These signal busses run directly between the vertical DSP48E1 slices without burdening the general routing resources. 
-        The ability to cascade this way provides a high-performance and low-power implementation of DSP filter functions.
-
-        \vspace{1.0cm}
-        {
-            \centering
-            \includegraphics[valign=c, width=2.5cm]{figures/dsp_cascade_placement.png}
-            \includegraphics[valign=c, width=3.0cm]{figures/dsp_cascade_routed.png}
-            \captionof{figure}{
-                A DSP48E1 cascade of size 2 placed across 2 DSP48E1 Sites.
-                \textbf{Left:} Simplified view, \textbf{Right:} As shown in the Vivado device viewer.
-            }
-        }
-
-        \end{multicols}
-        {
-            \centering
-            \includegraphics[valign=c, width=4.0cm]{figures/mult_synthesis.png}
-            \includegraphics[valign=c, width=13.0cm]{figures/dsp_synthesis.png}
-            \captionof{figure}{
-                Simple multiplier synthesis from user design.
-            }
-            \label{mult_synthesis}
-        }
-        \begin{multicols}{2}
-
-    \newpage
-
-    \subsection{Block RAM}
-        In addition to SLICEs and DSPs, the 7-Series also offers dedicated Block Random Access Memory (BRAM) BELs. 
-        These BRAMs come in two variants: \textbf{RAMB18E1} and \textbf{RAMB36E1}. 
-        \begin{itemize}
-            \item \textbf{RAMB18E1} - 
-            Has a capacity of 18 Kilobits. 
-            It can be configured as single port RAM with dimensions ranging between (1-bit wide by 16K deep) to (18-bit wide by 1024 deep). 
-            It can also be configured as a (36-bit wide by 512 deep) true simple dual port RAM. 
-            \item \textbf{RAMB36E1} - 
-            Has a capacity of 36 Kilobits. 
-            It can be configured as single port RAM with dimensions ranging between (1-bit wide by 32K deep) to (36-bit wide by 1024 deep). 
-            It can also be configured as a (72-bit wide by 512 deep) simple dual port RAM. 
-        \end{itemize}
-
-        For readability, we will abbreviate RAMB18E1 as \textbf{RAM18} and RAMB36E1 as \textbf{RAM36}.
-        One BRAM Tile contains one RAM36 Site and two RAM18 Sites. 
-        The RAM36 Site contains one RAM36 BEL while the RAM18 Site contains one RAM18 BEL.
-        The RAM18 BEL can host one RAM18 Cell, while the RAM36 BEL can host either one RAM36 Cell or two RAM18 Cells.
-
-        Like DSP cascades, BRAMs may also be cascaded together in a column to implement large memories efficiently, with some intermediate signals between them routed intra-Tile without burdening the general routing.
-        However, unlike DSPs, large memories decomposed amongst multiple BRAMs can also be routed together through general routing. 
-        Furthermore, in most design scenarios, large memories will not utilize the intra-Tile signals.
-
-        For simplicity we will not constrain large BRAMs to be cascaded together in consecutive vertical Tiles. 
-        We can essentially treat RAMB18E1 and RAMB36E1 Cells as loose, minimally constrained single cells, in contrast to the highly constrained DSP cascades and CARRY chains.
-
-        {
-            \centering
-            \includegraphics[valign=c, width=3.0cm]{figures/bram.png}
-            \includegraphics[valign=c, width=2.0cm]{figures/bram_tile.png}
-            \captionof{figure}{
-                A BRAM Tile containing two RAMB18E1 Sites and one RAMB36E1 Site.
-                \textbf{Left:} simplified view. \textbf{Right:} as seen in the device viewer, BELs highlighted in white. 
-            }
-            \label{fig:bram_tile}
-        }
-
-        {
-            \centering
-            \includegraphics[valign=c, width=4.0cm]{figures/bram_synthesis.png}
-            \includegraphics[valign=c, width=\columnwidth]{figures/bram_edif.png}
-            \captionof{figure}{
-                An example of BRAM synthesis (via inference)
-            }
-            \label{fig:bram_synthesis}
-        }
-
-    \subsection{Further Documentation}
-
-    For more in-depth details about 7-Series FPGAs, refer to the official Xilinx user guides such as:
-    \begin{itemize}
-        \item \emph{7 Series FPGAs Overview} (UG476)
-        \item \emph{7 Series FPGA Configurable Logic Block} (UG474)
-        \item \emph{7 Series Memory Resources} (UG473)
-        \item \emph{7 Series DSP48E1 Slice} (UG479)
-    \end{itemize}
-
-    This architectural context provides the necessary background for understanding how a placement algorithm should account for resource constraints and optimize performance in modern FPGA designs.
-
-\newpage
-\section{FPGA Design Flow and Toolchain}
-    \label{sec:fpga_flow_toolchain}
-
-    Modern FPGA designs require a sophisticated toolchain to bridge the gap between high-level hardware descriptions and the final bitstream used to configure the FPGA. 
-    Figure~\ref{fig:design_flow} illustrates a representative process that converts an abstract Hardware Description Language (HDL) design into a verified configuration file for a target device.
-
-    {
-        \centering
-        \includegraphics[width=0.9\columnwidth]{figures/design_flow.png}
-        \captionof{figure}{A typical FPGA design and verification workflow.}
-        \label{fig:design_flow}
-    }
-
-    \begin{enumerate}
-        \item \textbf{Design Entry:} 
-            An engineer describes the intended functionality of the digital system using a hardware description language (HDL) such as Verilog or VHDL. 
-            During this phase, the coding style can vary (behavioral, structural, dataflow, etc.), but it always aims to capture high-level behavior rather than device-specific details.
-
-        \item \textbf{Synthesis:} 
-            The synthesis tool parses the HDL source, performs logical optimizations, and maps the design onto primitive cells that suit the target FPGA technology. 
-            The output is typically a structural \texttt{netlist} (e.g., EDIF or structural Verilog) which details how the design’s logic is broken down into LUTs, FFs, and other vendor-specific cells.
-
-        \item \textbf{Placement and Routing (Implementation):} 
-            In \emph{placement}, each logical cell from the synthesized netlist is assigned to a physical location on the FPGA fabric. 
-            For instance, LUTs and FFs go into specific \emph{BELs} within the device’s CLB sites, and specialized cells such as DSPs and Block RAMs must be placed in their corresponding tile types. 
-            Next, \emph{routing} determines how signals are physically wired through the FPGA’s configurable interconnect network. 
-            Modern tools often interleave these steps (e.g., fluid-placement routing or routing-aware placement) to better meet timing and area objectives.
-
-        \item \textbf{Bitstream Generation:} 
-            After a design is fully placed, routed, and timing-closed, the toolchain produces a final \emph{bitstream} that sets the configuration of every programmable element in the FPGA. 
-            This bitstream can then be loaded onto the device, either through vendor software or via a custom programming interface.
-
-        \item \textbf{Verification:} 
-            In parallel to the design flow, simulations and testbenches validate correctness of the user's design at multiple abstraction levels. 
-            Engineers may begin with behavioral simulations, then progress to post-synthesis simulations, and finally to post-implementation simulations that incorporate estimated routing delays. 
-            With each higher level of fidelity, computational requirements grow significantly due to increasing complexity and the need to analyze more variables over time. 
-            Ensuring correct functionality and meeting timing closure at the post-implementation stage is crucial before deploying the design to hardware. 
-            Given the importance of thorough verification, many established companies dedicate one verification engineer for every design engineer.
-
-    \end{enumerate}
-
-    This workflow underscores the critical role of \textbf{placement} in bridging the netlist to a physical realization. 
-    An efficient placement algorithm can drastically reduce compile times and improve design performance, enabling broader adoption of FPGAs in application spaces that require fast design iteration.
-
-\newpage
-\section{RapidWright API}
-    \label{sec:rapidwright_api}
-
-    \textbf{RapidWright} is an open-source Java framework from AMD/Xilinx that provides direct access to the netlist and device databases used by vendor tools. 
-    This framework positions itself as an additional workflow column, allowing users to intercept or replace stages of the standard design flow with custom optimization stages (see Figure~\ref{fig:vivado_dcps}).
-
-    \begin{itemize}
-    \item \textbf{Design Checkpoints:} 
-        RapidWright leverages \texttt{.dcp} files (design checkpoints) generated at various stages of a Vivado flow. 
-        By importing a checkpoint, engineers can manipulate the netlist, placement, or routing externally, then re-export a modified checkpoint for further processing in the Vivado workflow column.
-
-    \item \textbf{Key Packages:} 
-        RapidWright revolves around three primary data model packages:
-        \begin{enumerate}
-        \item \emph{edif} -- Represents the logical netlist in an abstracted EDIF-like structure.
-        \item \emph{design} -- Contains data structures for the physical implementation (Cells, Nets, Sites, BELs, etc.).
-        \item \emph{device} -- Provides a database of the target FPGA architecture (e.g., Site coordinates, Tile definitions, routing resources).
-        \end{enumerate}
-
-    \item \textbf{Interfacing with the Netlist and Device:} 
-        An engineer can query the netlist to find specific resources (LUTs, FFs, DSPs, etc.) and then map or move them onto device sites. 
-        This level of control over backend resources is necessary for research in custom placement, advanced packing techniques, or experimental routing algorithms.
-    \end{itemize}
-
-    {
-        \centering
-        \includegraphics[width=0.8\columnwidth]{figures/vivado_dcps.png}
-        \captionof{figure}{RapidWright workflow integrating into the default Vivado design flow.}
-        \label{fig:vivado_dcps}
-    }
-
-    By exposing these low-level internals, RapidWright allows fine-grained design transformations that go beyond the standard Vivado IDE’s capabilities. 
-    Researchers can prototype new EDA strategies without needing to re-implement an entire FPGA backend from scratch, thus accelerating innovation in placement and routing methodologies.
-
-\newpage
-\section{What is a Netlist?}
-    \label{sec:netlist}
-    In its simplest form, a netlist consists of a list of every component in an electronic design and a list of nets they connect to. 
-    Here, a net, or network, denotes any group of two or more interconnected elements - whether they be transistors, logic gates, or higher-level macrocells. 
-    More precisely, nets attach to component pins, and those pins reside on cells, each of which exposing multiple ports. 
-    In this way, a netlist makes explicit both the individual building blocks of a circuit and the precise wiring that links them.
-
-    In Vivado, a Netlist can be synthesized as a Hierarchical or a Flattened netlist. 
-    Figure ~\ref{fig:hierarchical_design} shows the Verilog code for a simple HDL design using module instance hierarchy. 
-    Figure ~\ref{fig:hier_netlist} shows the design synthesized into a hierarchical netlist, while Figure ~\ref{fig:flat_netlist} shows the same design synthesized into a flat netlist. 
-    In a synthesized Netlist, the leaf cells (lowest-level cells) are necessarily primitive cells from the vendor primitives library such as LUT6, FDRE, CARRY4, DSP48E1, etc..
-
-    \end{multicols}
-    {
-        \raggedright
-        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/top_level.png}
-        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_0.png}
-        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_1.png}
-        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_2.png}
-        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_3.png}
-        \captionof{figure}{A simple HDL design with module hierarchy.}
-        \label{fig:hierarchical_design}
-    }
-    {
-        \centering
-        \includegraphics[valign=c, width=11.5cm]{figures/netlist_synth/hier_netlist.png}
-        \includegraphics[valign=c, width=6cm]{figures/netlist_synth/hier_graph.png}
-        \captionof{figure}{
-            \textbf{Left:} A hierarchical netlist consisting of LUTs and FFs.
-            \textbf{Right:} The cell hierarchy graph.
-        }
-        \label{fig:hier_netlist}
-    }
-    \vspace{0.5cm}
-    {
-        \centering
-        \includegraphics[valign=c, width=10cm]{figures/netlist_synth/flat_netlist.png}
-        \includegraphics[valign=c, width=4cm]{figures/netlist_synth/flat_graph.png}
-        \captionof{figure}{
-            \textbf{Left:} A flattened netlist consisting of LUTs and FFs.
-            \textbf{Right:} The flattened cell hierarchy graph.
-        }
-        \label{fig:flat_netlist}
-    }
-    \begin{multicols}{2}
-
-    RapidWright offers an easy way to access, read, or manipulate this netlist for a given HDL design. 
-    Shown in Figure~\ref{fig:netlist_syn}
-
-    Verilog \texttt{module}s are represented simply as
-
-    \end{multicols}
-    {
-        \centering
-        \includegraphics[width=\columnwidth]{figures/raw_edif.png}
-        \captionof{figure}{A raw EDIF Netlist with uncountably many cells}
-        \label{fig:raw_edif}
-    }
-    \begin{multicols}{2}
-
-    .
-
-\newpage
-\section{Simulated Annealing}
-    \label{sec:simulated_annealing}
-    With a basic understanding of FPGA architecture, design placement, and RapidWright, we have all the necessary pieces to implement our SA placer. 
-    Here we outline in detail each substage of our implementation: PrePacking, Packing, and Placement. 
-    Shown in Figure \ref{fig:edif_design_device} shows an overview of the placement workflow.
-
-    {
-        \centering
-        \includegraphics[width=0.9\columnwidth]{figures/edif_design_device.png}
-        \captionof{figure}{Our placement workflow}
-        \label{fig:edif_design_device}
-    }
-
-    \end{multicols}
-    {
-        \centering
-        \includegraphics[width=0.8\columnwidth]{figures/substages.png}
-        \captionof{figure}{The data classes populated at each substage: PrepackedDesign, PackedDesign, and PlacedDesign.}
-        \label{fig:substages}
-    }
-    \begin{multicols}{2}
-
-    \subsection{Prepacking}
-        \label{subsec:prepacking}
-
-        The first step in our placement flow is \textbf{prepacking}. 
-        Recall from the 7-Series architecture that there are certain cell structures that must adhere to certain placements constraints to ensure legality, and by design, to minimize wirelength. 
-        The job of the prepacker is to consolidate the cells in the EDIF Netlist into clusters or groups of clusters that naturally reflect these placement constraints.
-
-        CARRY4 chains must necessarily be placed vertically and consecutively across a column of SLICEs in ascending order. 
-        Likewise, DSP48E1 cascades must necessarily be placed vertically and consecutively across a column of DSP48E1 Sites in ascending order. 
-        A LUT-FF pair may be placed freely, but should be placed in the same lane within the same SLICE to minimize wirelength.
-
-        The raw post-synthesis netlist only tells us the list of nets and the cell ports that they connect to. 
-        It does not report the presence of any macro cell structures (CARRY4 chains, etc.). 
-        Thus, we must traverse the netlist to detect these cell structures and store that structure information in a class we will call \texttt{PrepackedDesign}.
-
-        A group of LUT-FF pairs may be placed in the same SLICE, with the constraint that the FFs must share the same Clock-Enable (CE) and Set-Reset (SR) nets.
-        Clustering LUT-FF pairs like this can reduce the redundancy of having to route the same CE and SR nets to many SLICEs by routing the nets to fewer SLICEs and connecting them to the individual FFs within via intra-Site routing, and at the same time, packing greater logic density over a smaller area of the device.
-        Recall that up to eight FFs may be placed in the same SLICE, thus theoretically, up to eight LUT-FF pairs can be placed in the same SLICE.
-
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/slicel.png}
-            \captionof{figure}{A SLICEL Site}
-            \label{fig:slicel}
-        }
-
-        Utilizing all 8 LUT-FF lanes in a SLICE can help reduce device area utilization and minimize wirelength, but \emph{too much} logic density in an area can contribute to general routing congestion.
-        Furthermore, attempting to fill all 8 lanes in a SLICE requires some careful consideration of LUT BEL constraints. 
-        The LUT BELs in a SLICE, as shown in Figure \ref{fig:slicel}, the LUTs appear to be interleaved.
-
-        Show a block diagram of a LUT6 comprising of two LUT5s.
-
-        Recall that a LUT can accommodate one 6-input boolean function or two 5-input boolean functions sharing the same inputs or any two 3-input or less boolean functions regardless of shared inputs.
-        A LUT6 Cell (a LUT with 6-inputs) will actually occupy two lanes in a SLICEL, rendering one of the FF BELs in either lane ineligible for another LUT-FF pair. 
-        The LUT BELs
-
-    \subsection{Packing}
-        \label{subsec:packing}
-
-    \subsection{Placement}
-        \label{subsec:placement}
-        Up until now we have only organized the logical \texttt{Cells} into \texttt{SiteInsts}. 
-        This is where simulated annealing actually begins.
-
-\section{Placement Results}
-    \begin{itemize}
-    \item Random Site Selection:
-    \item Midpoint Site Selection:
-    \item Hybrid: 
-        Random Site selection until progress stagnates, then use Midpoint Site selection. 
-        Randomly choose betwee Random and Midpoint Site selection each iteration. 
-    \item Cooling Schedule: 
-        Greedy. 
-        Geometric with modifiable alpha and intial temperature. 
-    \end{itemize}
-
-\section{Future Work}
-    \begin{itemize}
-    \item Different cooling schedules: linear cooling, logarithmic cooling, piecewise cooling.
-    \item Support for more macros: buffer cells, FIFO, clock generators, etc..
-    \item Support Ultrascale architecture. 
-    \item Better packing scheme. Currently pseudorandom packing. 
-    \item Inter-Site BEL swapping. 
-    \item BEL-centric placement over Site-centric placement.
-    \item Analytical placement. Global placement, legalization, detailed placement. 
-    \end{itemize}
-
-\section{Conclusion}
-    fdsfdsafdsa
-
-% \section{Analytical Placement}
-%     Introduce placement with legalization methods. \\
-%     Many AP approaches use a Global placement, Legalization, then Detailed placement approach. \\
-%     Our SA only considers legal moves via bookkeeping structures, so it has no concept of global placement or legalization. \\
-%     It is a detailed placement only approach. \\
-%     \begin{itemize}
-%         \item Talk about HeAP as an example.
-%         \item Global Placement, Legalization, then Detailed Placement approach.
-%         \item Global placement using the Bound2Bound net model from SimPL.
-%         \item Legalization Step 1: \\
-%             Find an area of the device that is overutilized (illegal) for twhich the blocks contained within must spread to a larger area. \\
-%             To obtain this overutilized area, adjacent locations (Sites, Tiles, etc.) on the device that are occupied by more than one block (SiteInst, ModuleInst, etc.) are repeatedly clustered together, until all clusters are bordered on all sides by non-overutilized locations.
-\\
-%             Then, the area (?) is expanded in both the x and y dimensions (?) until it is large enough to accommodate all blocks contained within. \\
-%             Specifically, the area is expanded until its "occupance" \(O_A\) divided by its "capacity" \(C_A\) is less than a maximum "utilization factor" $\beta$, where $\beta$ is less than or equal to 1.
-HeAP uses 0.9.
-%         \item Legalization Step 2: \\
-%             Two cuts are generated: a "source cut" and a "target cut". \\
-%             The source cut pertains to the blocks being placed (SiteInsts in home buffer). \\ 
-%             The target cut pertains to the area into which the blocks are placed (away buffer). \\
-%             The source cut splits the blocks into 2 partitions, while the target cut splits the area into 2 sub-areas, into which the blocks in each partition are spread. \\
-%             Two objectives are minimized during this process: The imbalance between the number of blocks in each partition, and the difference in the utilization of each sub-area. \\
-%             Utilization defined as the occupancy divided by the capacity of the sub-area. \\
-%             \(U_{sub-area} = \frac{O_{sub-area}}{C_{sub-area}}\) \\
-%             To generate the source cut, the cells are first sorted by their x or y location, depending on the orientation of the desired cut. \\
-%             Once the cells are sorted, the source cut generation is akin to choosing a pivot in a sorted list, where all blocks to the left of the pivot are assigned to the left/bottom partition and all blocks to the right of the pivot are assigned to the right/top partition.
-\\
-%             The target cut is an x or y cut of the area such that all blocks in each partition fit in their respective sub-areas, and such that \( | U_{sub-area_1} 
-%              - U_{sub-area_2} | \) is minimized.
-%     \end{itemize}
-
-\section{Appendix}
-
-    \begin{figure*}[b]
-        \centering
-        \includegraphics[width=\textwidth]{figures/slicem.png}
-        \caption{SLICEM Site}
-        \label{fig:slicem}
-    \end{figure*}
-
-    \begin{figure*}[t]
-        \centering
-        \includegraphics[width=\textwidth]{figures/slicel.png}
-        \caption{SLICEL Site}
-        \label{fig:slicel}
-    \end{figure*}
-
-\end{multicols}
-
-\end{document}

### 2025-05-03 latex/paper/sections/03-introduction.tex
split tex file into multiple files

--- 2025-04-28 latex/paper/paper.tex
@@
-\documentclass{article}
-\usepackage{hyperref}
-\usepackage{enumitem}
-\usepackage{graphicx}
-\usepackage{amsmath}
-\usepackage{mathpazo}
-\usepackage{xcolor}
-\usepackage{float}
-\usepackage{listings}
-\lstset{
-  basicstyle=\ttfamily\small,
-  breaklines=true,
-  columns=fullflexible,
-  frame=single,
-  showspaces=false,
-  showtabs=false,
-  keepspaces=false
-}
-\usepackage[export]{adjustbox}
-\usepackage{multicol}
-\usepackage{caption}
-\usepackage[a4paper,
-            bindingoffset=0.2in,
-            left=0.5in,
-            right=0.5in,
-            top=0.5in,
-            bottom=0.5in,
-            footskip=.25in]{geometry}
-
-\begin{document}
-\title{MS Technical Paper: \\ Placement Algorithms for Heterogeneous FPGAs}
-\author{Brian B Cheng \\ Rutgers University Department of Electrical and Computer Engineering}
-
-\date{}
-\maketitle
-
-\begin{multicols}{2}
-
-\section{Keywords}
-\begin{itemize}
-    \item FPGA, EDA, Placement, Simulated Annealing, Optimization, RapidWright
-\end{itemize}
-
-\section{Abstract}
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\
-
 \section{Introduction}
@@
 However, readers who are already familiar with these concepts can skip directly to the RapidWright API section \ref{sec:rapidwright_api} or to the Simulated Annealing section \ref{sec:simulated_annealing}.
-
-\section{FPGA Architecture History}
-
-    Before any work can begin on an FPGA placer, it is necessary to understand both the objects being placed and the medium in which they are placed.
-    Configurable logic devices have undergone significant evolution over the past four decades. 
-    We will briefly review the evolution of configurable logic architecture starting in the 1970s and quickly work our way up to modern day FPGA architecture.
-
-    \textbf{PLA:} \quad
-    The journey began with the Programmable Logic Array (PLA) in the early 1970s. 
-    The PLA implemented output logic using a programmable-OR and programmable-AND plane that formed a sum-of-products equation for each output through programmable fuses. 
-    Around the same time, the Programmable Array Logic (PAL) was introduced. 
-    The PAL simplified the PLA by fixing the OR gates, resulting in a fixed-OR, programmable-AND design, which sacrificed some logic flexibility to simplify its manufacture. 
-    Figure \ref{fig:pal_2} shows one such PAL architecture.
-
-    \textbf{CPLD:} \quad 
-    Later in the same decade came the Complex Programmable Logic Device (CPLD), which took the form of an array of Configurable Logic Blocks (CLBs). 
-    These CLBs were typically modified PAL blocks that included the PAL itself along with macrocells such as flip-flops, multiplexers, and tri-state buffers. 
-    The CPLD functioned as an array of PALs connected by a central programmable switch matrix and could be programmed using a hardware description language (HDL) like VHDL. 
-    Figure \ref{fig:cpld} shows one such CPLD architecture. 
-    \\
-
-    {
-        \centering
-        \includegraphics[width=\columnwidth]{figures/pal_2.png}
-        \captionof{figure}{PAL architecture with 5 inputs, 8 programmable AND gates and 4 fixed OR gates}
-        \label{fig:pal_2}
-        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/Programmable-Logic-Devices/Programmable-Array-Logic-PAL/
-        % https://www.naukri.com/code360/library/difference-between-pla-and-pal 
-    }
-
-    \newcolumn
-
-    \textbf{Homogeneous FPGA:} \quad 
-    The mid-1980s saw the introduction of homogeneous FPGAs, which were built as a grid of CLBs. 
-    Rather than using a central programmable switch matrix as in CPLDs, FPGAs adopted an island style architecture in which each CLB is surrounded on all sides by programmable routing resources, as shown in Figure \ref{fig:fpga}.
-    The first commercially viable FPGA, produced by Xilinx in 1984, featured 16 CLBs arranged in a 4x4 grid. 
-    As FPGA technology advanced, CLBs were redesigned to use lookup tables (LUTs) instead of PAL arrays for greater logic density. 
-    The capacity of an FPGA was often measured by how many logical elements or CLBs it offered, which grew from hundreds to thousands and now to hundreds of thousands of CLBs.
-    \\
-
-    {
-        \centering
-        \includegraphics[width=0.8\columnwidth]{figures/cpld.png}
-        \captionof{figure}{CPLD architecture with 4 CLBs (PAL-like blocks)}
-        \label{fig:cpld}
-        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/CPLD/Complex-Programmable-Logic-Device-CPLDs/
-    }
-    {
-        \centering
-        \includegraphics[width=\columnwidth]{figures/homogenous_fpga.png}
-        \captionof{figure}{A homogeneous island-style FPGA architecture with 16 CLBs in a grid.}
-        \label{fig:fpga}
-        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/FPGA/Field-programmable-gate-array-FPGA/
-    }
-
-    \newpage
-
-    {
-        \centering
-        \includegraphics[width=\columnwidth]{figures/heterogenous_fpga_3.jpg}
-        \captionof{figure}{A heterogeneous island-style FPGA with a mix of CLBs and macrocells.}
-        \label{fig:fpga}
-        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/FPGA/Field-programmable-gate-array-FPGA/
-    }
-    \vspace{0.5cm}
-
-    \textbf{Heterogenous FPGA:} \quad 
-    This brings us to modern day FPGA architectures. 
-    To meet the needs of increasingly complex designs, FPGA vendors introduced heterogeneous FPGAs. 
-    In these devices, hard macros such as Block RAM (BRAM) and Digital Signal Processing (DSP) slices are integrated into the programmable logic fabric along with CLBs, like shown in Figure \ref{fig:fpga}.
-    This design enables the direct instantiation of common subsystems like memories and multipliers, without having to recreate them from scratch using CLBs. 
-    Major vendors such as Xilinx and Altera now employ heterogeneous island-style architectures in their devices. 
-    As designs become increasingly large and complex, FPGAs meet the demand by becoming increasingly heterogenous, incorporating a wider variety of hard macros into the fabric.
-
-\section{Xilinx 7-Series Architecture}
-    \label{sec:7_series}
-    The Xilinx 7-Series devices, first introduced in 2010, follow a heterogeneous island-style architecture as discussed previously. 
-    Although the 7-Series was later superseded in 2013 by the UltraScale architecture, the 7-Series remains highly relevant due to its accessibility, wide availability, and compatibility with open-source tooling.
-    Representative sub-families include Artix-7, Kintex-7, Virtex-7, and Zynq-7000, each designed with different performance and cost trade-offs but all follow the core 7-Series architecture.
-
-    Figure~\ref{fig:hierarchy_4} illustrates a high-level view of the hierarchical organization of a 7-Series FPGA. 
-    At its lowest level, the device consists of a large array of atomic components called \emph{Basic Elements of Logic}~(\textbf{BELs}). 
-    These BELs encompass look-up tables (\textbf{LUTs}), flip-flops (\textbf{FFs}), block RAMs (\textbf{BRAMs}), DSP slices (\textbf{DSPs}), and the configurable interconnect fabric. 
-    They constitute the fundamental building blocks for implementing digital circuits on the FPGA.
-    Note how the Tile arrangement is columnar, where each column will have only one Tile type.
-
-    To manage this complexity, Xilinx organizes these BELs into incrementally abstract structures. 
-    First, \textbf{BELs} are grouped into \textbf{Sites}. 
-    Each Site is embedded into a \textbf{Tile}, and Tiles are further arranged into \textbf{Clock Regions}. 
-    In some high-density devices, multiple Clock Regions may be consolidated into one or more \textbf{Super Logic Regions}~(SLRs). 
-    However, for the scope of this paper, we focus on Xilinx 7-Series devices with only a single SLR.
-
-    \end{multicols}
-    {
-        \centering
-        \includegraphics[valign=c, width=0.3\columnwidth]{figures/hierarchy_5.png}
-        \includegraphics[valign=c, width=0.6\columnwidth]{figures/hierarchy_4.png}
-        \captionof{figure}{Architecture Hierarchy of a Xilinx FPGA}
-        \label{fig:hierarchy_4}
-    }
-    \vfill
-    \begin{multicols}{2}
-
-    \newpage
-    \subsection{CLB SLICEs}
-
-    In the 7-Series architecture, the term \emph{CLB} (Configurable Logic Block) refers to a \emph{CLB Tile} that contains two \emph{SLICE} Sites. 
-    Xilinx offers two variants of SLICE Sites: \textbf{SLICEL} and \textbf{SLICEM}.
-    \begin{itemize}
-        \item Each \textbf{SLICEL} has a set of BELs including eight LUTs, eight FFs, and one   \emph{CARRY4} adder. The LUT BELs in a SLICEL can only host LUT Cells. 
-        \item The \textbf{SLICEM} includes all the features of a SLICEL but its LUT BELs can host both LUT Cells, which are asynchronous ROM elements, or RAM32M Cells, which are synchronous 32-deep RAM elements.
-These cells are also referred to as \emph{Distributed RAM} in the Xilinx documentation.
-        These cells can offer an alternative to the larger, more dedicated 18K-36K RAMB18E1 cells when RAM resources are highly utilized. 
-    \end{itemize}
-    In a typical 7-Series device, approximately 75\% of the SLICE Sites are SLICELs and 25\% are SLICEMs. 
-    A single CLB Tile can therefore host either two SLICELs or one SLICEL and one SLICEM.
-    For simplicity, however, we will only consider SLICELs for general logic and use the dedicated RAMB18E1 cells to implement RAM elements.
-
-    The BELs in these SLICEs facilitate the bulk of the general programmability of the FPGA fabric. 
-    We will explain in detail the motivation and function of these BELs.
-
-    \textbf{LUTs} \quad
-        Combinational logic is universal to all HDL designs. 
-        As the their name suggests, Look‑Up Tables (LUTs) facilitate combinational logic by acting as tiny asynchronously‑accessed ROMs whose contents are fixed when the FPGA is programmed.  
-        For any boolean function, the synthesizer precalculates the output to every possible input combination and stores the resulting truth table into a LUT's static memory. 
-        The inputs are then essentially treated as an address space that maps to a data value space in an asynchronous ROM. 
-        No explicit logic gates like NAND or XNOR are synthesized, contrary to what newcomers might expect from a "Field Programmable Gate Array".
-
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/lut_synthesis.png}
-            \captionof{figure}{LUT synthesis from user design}
-            \label{fig:lut_synthesis}
-        }
-        \vspace{1.0cm}
-
-        In the 7‑Series devices, one LUT can facilitate any 6-input boolean function, or two 5-input functions, as long as they share the same input signals.  
-        The LUT can also host two independent boolean functions of up to 3 inputs each, even when the inputs are not shared.  
-        Functions requiring more than six unique inputs are decomposed across multiple cascaded LUTs.
-        Figure \ref{fig:lut_synthesis} shows an example of where a LUT is typically synthesized in a design entry.
-
-    \textbf{FFs} \quad
-        FFs are synthesized to facilitate synchronous event-driven signal assignment. 
-        For most Verilog users, this generally means signal assignments wrapped in \texttt{always @(posedge clk)} statements. 
-        Figure \ref{fig:ff_synthesis} shows an example of where a flip flop (FF) is typically synthesized. 
-        The cell primitive \textbf{FDRE} is a type of FF and belongs to a family of D Flip Flops (DFFs) with Clock Enable (CE). 
-        \begin{itemize}
-            \item FDCE - DFF with CE and Asynchronous Clear
-            \item FDPE - DFF with CE and Asynchronous Preset
-            \item FDSE - DFF with CE and Synchronous Set
-            \item \textbf{FDRE} - DFF with CE and Synchronous Reset
-        \end{itemize}
-        In a typical HDL design, the vast majority of FFs will be synthesized as FDREs with the occasional FDSE, as it is generally good practice to keep FPGA designs synchronous. 
-        A FF BEL may also host a LATCH Cell, however, since they are generally bad practice in FPGA design, we will not consider latches in this paper.
-
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/ff_synthesis.png}
-            \captionof{figure}{FF synthesis from user design}
-            \label{fig:ff_synthesis}
-        }
-        \vspace{1.0cm}
-
-        Up to eight (8) FFs can be placed within the same SLICE, but only if they all share a common Clock-Enable (CE) net and a common Set-Reset (SR) net.
-        This is because the SLICE has only one CE pin and one SR pin to interface with general routing. 
-        The CE and SR signals from these pins are broadcast intra-Site to all FFs within.
-
-    \newpage
-    \textbf{LUT-FF Pairs} \quad
-        FPGA designs are very often modelled as a collection of Finite State Machines (FSM) like shown in Figure \ref{fig:fsm}.
-        Many times a design will also use pipelining, either to model signal buffers or shift registers, or to split up large combinational logic blocks into time slices to meet timing constraints. 
-        These common design structures result in many consecutive sections of combinational logic feeding into a vector of registers. 
-        The synthesizer naturally synthesizes these structures as consecutive pairs of LUTs feeding into FFs as shown in Figure \ref{fig:pipelining}. 
-        Figure \ref{fig:lut_ff_pair} shows an example of a synthesized LUT-FF Pair.
-
-        Shown in Figure \ref{fig:intersite_intrasite} are two possible placements for a LUT-FF Pair on the physical device. 
-        On the right, the cells are placed across different Sites, thus the only way to route the net between the cells is through general inter-site routing. 
-        On the bottom left, the cells are placed within the same Site in the same lane, taking advantage of the intra-site routing without burdening the general router with additional inter-site routing.
-        This is an important consideration to make while minimizing wirelength during placement.
-
-        \vspace{1.0cm}
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/fsm.png}
-            \captionof{figure}{Finite state machine (Moore)}
-            \label{fig:fsm}
-        }
-        \vspace{1.0cm}
-
-        {
-            \centering
-            \includegraphics[width=0.9\columnwidth]{figures/pipelining.png}
-            \captionof{figure}{Pipelining synthesized as consecutive LUT-FF pairs}
-            \label{fig:pipelining}
-        }
-
-        \vfill
-        .
-        \newcolumn
-
-        {
-            \centering
-            \includegraphics[width=0.9\columnwidth]{figures/lut_ff_pair.png}
-            \captionof{figure}{A synthesized LUT-FF Pair}
-            \label{fig:lut_ff_pair}
-        }
-        \vspace{1.0cm}
-
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/intersite_intrasite_2.png}
-            \captionof{figure}{Intrasite vs Intersite LUT-FF Placement}
-            \label{fig:intersite_intrasite}
-        }
-
-    \newpage
-    \textbf{CARRY} \quad
-        An FPGA design will also typically implement many adders, counters, subtractors, or comparators, all of which are based on binary addition. 
-        They are so ubiquitous that every that in the 7-Series architecture, every SLICE features a \emph{CARRY4} BEL -- a 4-bit carry-lookahead (CLA) adder, a much better alternative to synthesizing adders via LUTs.
-
-    \textbf{CARRY Chains} \quad
-        These CARRY4 blocks can be chained across SLICEs to implement wide adders efficiently. 
-        The CARRY4 BELs \emph{must} be chained vertically consecutively across SLICEs as the Carry-In (CI) and Carry-Out (CO) pins can only be routed this way. 
-        A CARRY4 cell may also directly connect to LUTs and FFs, and should be placed in the same Site whenever possible to minimize wirelength. 
-        Shown in figure \ref{fig:carry_cell_group} shows how a CARRY4 chain and associated LUTs and FFs can be placed across SLICEs.
-
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/carry_cell_group.png}
-            \captionof{figure}{A SLICEL with a CARRY4 cell, 4 LUT cells, and 4 FF cells placed inside as shown in the Vivado device viewer}
-            \label{fig:carry_cell_group}
-        }
-
-        {
-            \centering
-            \includegraphics[valign=c, width=4.0cm]{figures/carry_chain_placement.png}
-            \includegraphics[valign=c, width=4.0cm]{figures/carry_chain_routes_3.png}
-            \captionof{figure}{
-                A CARRY4 chain of size 3 placed across 3 SLICEs.
-                \textbf{Left:} Simplified view, \textbf{Right:} As shown in the Vivado device viewer.
-            }
-            \label{fig:carry_cell_group}
-        }
-
-        \end{multicols}
-        % \vfill
-        {
-            \centering
-            \includegraphics[valign=c, width=3.0cm]{figures/adder_synthesis.png}
-            \includegraphics[valign=c, width=14.0cm]{figures/carry_chain_edif_3.png}
-            \captionof{figure}{A synthesized carry chain of size 3 as shown in the Vivado netlist viewer}
-            \label{fig:carry_chain_edif}
-        }
-        \begin{multicols}{2}
-        \vspace{1.0cm}
-        .
-
-    \newpage
-    \subsection{DSP Slices}
-
-    \textbf{DSPs} \quad 
-        FPGAs are often used as low latency Digital Signal Processing (DSP) accelerators. 
-        Common DSP subsystems like Finite Impulse Response (FIR) filters, Fast Fourier Transform (FFTs), and convolutional neural nets (CNNs) demand fast large scale multiply-accumulate (MAC) capabilities.
-        The 7-Series arhictecture integrates DSP BELs into the logic fabric called DSP48E1 that can facilitate MAC efficiently. 
-        The architecture hierarchy for DSPs is simple compared to CLBs and SLICEs. 
-        A DSP48E1 Tile contains two DSP48E1 Sites, each containing one DSP48E1 BEL, which can host a DSP48E1 Cell.
-
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/dsp_diagram.png}
-            \captionof{figure}{
-                Basic DSP48E1 Slice Functionality
-            }
-            \label{dsp_diagram}
-        }
-        \vspace{0.5cm}
-
-    \textbf{DSP Cascades} \quad 
-        Higher level DSP functions are supported by cascading DSP48E1 slices in a DSP48E1 column.
-        Much like CARRY4 chains, DSP48E1 cascades must necessarily be placed vertically consecutively across DSP48E1 Sites. 
-        They are connected by three busses: ACOUT to ACIN, BCOUT to BCIN, PCOUT to PCIN.
-        These signal busses run directly between the vertical DSP48E1 slices without burdening the general routing resources. 
-        The ability to cascade this way provides a high-performance and low-power implementation of DSP filter functions.
-
-        \vspace{1.0cm}
-        {
-            \centering
-            \includegraphics[valign=c, width=2.5cm]{figures/dsp_cascade_placement.png}
-            \includegraphics[valign=c, width=3.0cm]{figures/dsp_cascade_routed.png}
-            \captionof{figure}{
-                A DSP48E1 cascade of size 2 placed across 2 DSP48E1 Sites.
-                \textbf{Left:} Simplified view, \textbf{Right:} As shown in the Vivado device viewer.
-            }
-        }
-
-        \end{multicols}
-        {
-            \centering
-            \includegraphics[valign=c, width=4.0cm]{figures/mult_synthesis.png}
-            \includegraphics[valign=c, width=13.0cm]{figures/dsp_synthesis.png}
-            \captionof{figure}{
-                Simple multiplier synthesis from user design.
-            }
-            \label{mult_synthesis}
-        }
-        \begin{multicols}{2}
-
-    \newpage
-
-    \subsection{Block RAM}
-        In addition to SLICEs and DSPs, the 7-Series also offers dedicated Block Random Access Memory (BRAM) BELs. 
-        These BRAMs come in two variants: \textbf{RAMB18E1} and \textbf{RAMB36E1}. 
-        \begin{itemize}
-            \item \textbf{RAMB18E1} - 
-            Has a capacity of 18 Kilobits. 
-            It can be configured as single port RAM with dimensions ranging between (1-bit wide by 16K deep) to (18-bit wide by 1024 deep). 
-            It can also be configured as a (36-bit wide by 512 deep) true simple dual port RAM. 
-            \item \textbf{RAMB36E1} - 
-            Has a capacity of 36 Kilobits. 
-            It can be configured as single port RAM with dimensions ranging between (1-bit wide by 32K deep) to (36-bit wide by 1024 deep). 
-            It can also be configured as a (72-bit wide by 512 deep) simple dual port RAM. 
-        \end{itemize}
-
-        For readability, we will abbreviate RAMB18E1 as \textbf{RAM18} and RAMB36E1 as \textbf{RAM36}.
-        One BRAM Tile contains one RAM36 Site and two RAM18 Sites. 
-        The RAM36 Site contains one RAM36 BEL while the RAM18 Site contains one RAM18 BEL.
-        The RAM18 BEL can host one RAM18 Cell, while the RAM36 BEL can host either one RAM36 Cell or two RAM18 Cells.
-
-        Like DSP cascades, BRAMs may also be cascaded together in a column to implement large memories efficiently, with some intermediate signals between them routed intra-Tile without burdening the general routing.
-        However, unlike DSPs, large memories decomposed amongst multiple BRAMs can also be routed together through general routing. 
-        Furthermore, in most design scenarios, large memories will not utilize the intra-Tile signals.
-
-        For simplicity we will not constrain large BRAMs to be cascaded together in consecutive vertical Tiles. 
-        We can essentially treat RAMB18E1 and RAMB36E1 Cells as loose, minimally constrained single cells, in contrast to the highly constrained DSP cascades and CARRY chains.
-
-        {
-            \centering
-            \includegraphics[valign=c, width=3.0cm]{figures/bram.png}
-            \includegraphics[valign=c, width=2.0cm]{figures/bram_tile.png}
-            \captionof{figure}{
-                A BRAM Tile containing two RAMB18E1 Sites and one RAMB36E1 Site.
-                \textbf{Left:} simplified view. \textbf{Right:} as seen in the device viewer, BELs highlighted in white. 
-            }
-            \label{fig:bram_tile}
-        }
-
-        {
-            \centering
-            \includegraphics[valign=c, width=4.0cm]{figures/bram_synthesis.png}
-            \includegraphics[valign=c, width=\columnwidth]{figures/bram_edif.png}
-            \captionof{figure}{
-                An example of BRAM synthesis (via inference)
-            }
-            \label{fig:bram_synthesis}
-        }
-
-    \subsection{Further Documentation}
-
-    For more in-depth details about 7-Series FPGAs, refer to the official Xilinx user guides such as:
-    \begin{itemize}
-        \item \emph{7 Series FPGAs Overview} (UG476)
-        \item \emph{7 Series FPGA Configurable Logic Block} (UG474)
-        \item \emph{7 Series Memory Resources} (UG473)
-        \item \emph{7 Series DSP48E1 Slice} (UG479)
-    \end{itemize}
-
-    This architectural context provides the necessary background for understanding how a placement algorithm should account for resource constraints and optimize performance in modern FPGA designs.
-
-\newpage
-\section{FPGA Design Flow and Toolchain}
-    \label{sec:fpga_flow_toolchain}
-
-    Modern FPGA designs require a sophisticated toolchain to bridge the gap between high-level hardware descriptions and the final bitstream used to configure the FPGA. 
-    Figure~\ref{fig:design_flow} illustrates a representative process that converts an abstract Hardware Description Language (HDL) design into a verified configuration file for a target device.
-
-    {
-        \centering
-        \includegraphics[width=0.9\columnwidth]{figures/design_flow.png}
-        \captionof{figure}{A typical FPGA design and verification workflow.}
-        \label{fig:design_flow}
-    }
-
-    \begin{enumerate}
-        \item \textbf{Design Entry:} 
-            An engineer describes the intended functionality of the digital system using a hardware description language (HDL) such as Verilog or VHDL. 
-            During this phase, the coding style can vary (behavioral, structural, dataflow, etc.), but it always aims to capture high-level behavior rather than device-specific details.
-
-        \item \textbf{Synthesis:} 
-            The synthesis tool parses the HDL source, performs logical optimizations, and maps the design onto primitive cells that suit the target FPGA technology. 
-            The output is typically a structural \texttt{netlist} (e.g., EDIF or structural Verilog) which details how the design’s logic is broken down into LUTs, FFs, and other vendor-specific cells.
-
-        \item \textbf{Placement and Routing (Implementation):} 
-            In \emph{placement}, each logical cell from the synthesized netlist is assigned to a physical location on the FPGA fabric. 
-            For instance, LUTs and FFs go into specific \emph{BELs} within the device’s CLB sites, and specialized cells such as DSPs and Block RAMs must be placed in their corresponding tile types. 
-            Next, \emph{routing} determines how signals are physically wired through the FPGA’s configurable interconnect network. 
-            Modern tools often interleave these steps (e.g., fluid-placement routing or routing-aware placement) to better meet timing and area objectives.
-
-        \item \textbf{Bitstream Generation:} 
-            After a design is fully placed, routed, and timing-closed, the toolchain produces a final \emph{bitstream} that sets the configuration of every programmable element in the FPGA. 
-            This bitstream can then be loaded onto the device, either through vendor software or via a custom programming interface.
-
-        \item \textbf{Verification:} 
-            In parallel to the design flow, simulations and testbenches validate correctness of the user's design at multiple abstraction levels. 
-            Engineers may begin with behavioral simulations, then progress to post-synthesis simulations, and finally to post-implementation simulations that incorporate estimated routing delays. 
-            With each higher level of fidelity, computational requirements grow significantly due to increasing complexity and the need to analyze more variables over time. 
-            Ensuring correct functionality and meeting timing closure at the post-implementation stage is crucial before deploying the design to hardware. 
-            Given the importance of thorough verification, many established companies dedicate one verification engineer for every design engineer.
-
-    \end{enumerate}
-
-    This workflow underscores the critical role of \textbf{placement} in bridging the netlist to a physical realization. 
-    An efficient placement algorithm can drastically reduce compile times and improve design performance, enabling broader adoption of FPGAs in application spaces that require fast design iteration.
-
-\newpage
-\section{RapidWright API}
-    \label{sec:rapidwright_api}
-
-    \textbf{RapidWright} is an open-source Java framework from AMD/Xilinx that provides direct access to the netlist and device databases used by vendor tools. 
-    This framework positions itself as an additional workflow column, allowing users to intercept or replace stages of the standard design flow with custom optimization stages (see Figure~\ref{fig:vivado_dcps}).
-
-    \begin{itemize}
-    \item \textbf{Design Checkpoints:} 
-        RapidWright leverages \texttt{.dcp} files (design checkpoints) generated at various stages of a Vivado flow. 
-        By importing a checkpoint, engineers can manipulate the netlist, placement, or routing externally, then re-export a modified checkpoint for further processing in the Vivado workflow column.
-
-    \item \textbf{Key Packages:} 
-        RapidWright revolves around three primary data model packages:
-        \begin{enumerate}
-        \item \emph{edif} -- Represents the logical netlist in an abstracted EDIF-like structure.
-        \item \emph{design} -- Contains data structures for the physical implementation (Cells, Nets, Sites, BELs, etc.).
-        \item \emph{device} -- Provides a database of the target FPGA architecture (e.g., Site coordinates, Tile definitions, routing resources).
-        \end{enumerate}
-
-    \item \textbf{Interfacing with the Netlist and Device:} 
-        An engineer can query the netlist to find specific resources (LUTs, FFs, DSPs, etc.) and then map or move them onto device sites. 
-        This level of control over backend resources is necessary for research in custom placement, advanced packing techniques, or experimental routing algorithms.
-    \end{itemize}
-
-    {
-        \centering
-        \includegraphics[width=0.8\columnwidth]{figures/vivado_dcps.png}
-        \captionof{figure}{RapidWright workflow integrating into the default Vivado design flow.}
-        \label{fig:vivado_dcps}
-    }
-
-    By exposing these low-level internals, RapidWright allows fine-grained design transformations that go beyond the standard Vivado IDE’s capabilities. 
-    Researchers can prototype new EDA strategies without needing to re-implement an entire FPGA backend from scratch, thus accelerating innovation in placement and routing methodologies.
-
-\newpage
-\section{What is a Netlist?}
-    \label{sec:netlist}
-    In its simplest form, a netlist consists of a list of every component in an electronic design and a list of nets they connect to. 
-    Here, a net, or network, denotes any group of two or more interconnected elements - whether they be transistors, logic gates, or higher-level macrocells. 
-    More precisely, nets attach to component pins, and those pins reside on cells, each of which exposing multiple ports. 
-    In this way, a netlist makes explicit both the individual building blocks of a circuit and the precise wiring that links them.
-
-    In Vivado, a Netlist can be synthesized as a Hierarchical or a Flattened netlist. 
-    Figure ~\ref{fig:hierarchical_design} shows the Verilog code for a simple HDL design using module instance hierarchy. 
-    Figure ~\ref{fig:hier_netlist} shows the design synthesized into a hierarchical netlist, while Figure ~\ref{fig:flat_netlist} shows the same design synthesized into a flat netlist. 
-    In a synthesized Netlist, the leaf cells (lowest-level cells) are necessarily primitive cells from the vendor primitives library such as LUT6, FDRE, CARRY4, DSP48E1, etc..
-
-    \end{multicols}
-    {
-        \raggedright
-        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/top_level.png}
-        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_0.png}
-        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_1.png}
-        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_2.png}
-        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_3.png}
-        \captionof{figure}{A simple HDL design with module hierarchy.}
-        \label{fig:hierarchical_design}
-    }
-    {
-        \centering
-        \includegraphics[valign=c, width=11.5cm]{figures/netlist_synth/hier_netlist.png}
-        \includegraphics[valign=c, width=6cm]{figures/netlist_synth/hier_graph.png}
-        \captionof{figure}{
-            \textbf{Left:} A hierarchical netlist consisting of LUTs and FFs.
-            \textbf{Right:} The cell hierarchy graph.
-        }
-        \label{fig:hier_netlist}
-    }
-    \vspace{0.5cm}
-    {
-        \centering
-        \includegraphics[valign=c, width=10cm]{figures/netlist_synth/flat_netlist.png}
-        \includegraphics[valign=c, width=4cm]{figures/netlist_synth/flat_graph.png}
-        \captionof{figure}{
-            \textbf{Left:} A flattened netlist consisting of LUTs and FFs.
-            \textbf{Right:} The flattened cell hierarchy graph.
-        }
-        \label{fig:flat_netlist}
-    }
-    \begin{multicols}{2}
-
-    RapidWright offers an easy way to access, read, or manipulate this netlist for a given HDL design. 
-    Shown in Figure~\ref{fig:netlist_syn}
-
-    Verilog \texttt{module}s are represented simply as
-
-    \end{multicols}
-    {
-        \centering
-        \includegraphics[width=\columnwidth]{figures/raw_edif.png}
-        \captionof{figure}{A raw EDIF Netlist with uncountably many cells}
-        \label{fig:raw_edif}
-    }
-    \begin{multicols}{2}
-
-    .
-
-\newpage
-\section{Simulated Annealing}
-    \label{sec:simulated_annealing}
-    With a basic understanding of FPGA architecture, design placement, and RapidWright, we have all the necessary pieces to implement our SA placer. 
-    Here we outline in detail each substage of our implementation: PrePacking, Packing, and Placement. 
-    Shown in Figure \ref{fig:edif_design_device} shows an overview of the placement workflow.
-
-    {
-        \centering
-        \includegraphics[width=0.9\columnwidth]{figures/edif_design_device.png}
-        \captionof{figure}{Our placement workflow}
-        \label{fig:edif_design_device}
-    }
-
-    \end{multicols}
-    {
-        \centering
-        \includegraphics[width=0.8\columnwidth]{figures/substages.png}
-        \captionof{figure}{The data classes populated at each substage: PrepackedDesign, PackedDesign, and PlacedDesign.}
-        \label{fig:substages}
-    }
-    \begin{multicols}{2}
-
-    \subsection{Prepacking}
-        \label{subsec:prepacking}
-
-        The first step in our placement flow is \textbf{prepacking}. 
-        Recall from the 7-Series architecture that there are certain cell structures that must adhere to certain placements constraints to ensure legality, and by design, to minimize wirelength. 
-        The job of the prepacker is to consolidate the cells in the EDIF Netlist into clusters or groups of clusters that naturally reflect these placement constraints.
-
-        CARRY4 chains must necessarily be placed vertically and consecutively across a column of SLICEs in ascending order. 
-        Likewise, DSP48E1 cascades must necessarily be placed vertically and consecutively across a column of DSP48E1 Sites in ascending order. 
-        A LUT-FF pair may be placed freely, but should be placed in the same lane within the same SLICE to minimize wirelength.
-
-        The raw post-synthesis netlist only tells us the list of nets and the cell ports that they connect to. 
-        It does not report the presence of any macro cell structures (CARRY4 chains, etc.). 
-        Thus, we must traverse the netlist to detect these cell structures and store that structure information in a class we will call \texttt{PrepackedDesign}.
-
-        A group of LUT-FF pairs may be placed in the same SLICE, with the constraint that the FFs must share the same Clock-Enable (CE) and Set-Reset (SR) nets.
-        Clustering LUT-FF pairs like this can reduce the redundancy of having to route the same CE and SR nets to many SLICEs by routing the nets to fewer SLICEs and connecting them to the individual FFs within via intra-Site routing, and at the same time, packing greater logic density over a smaller area of the device.
-        Recall that up to eight FFs may be placed in the same SLICE, thus theoretically, up to eight LUT-FF pairs can be placed in the same SLICE.
-
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/slicel.png}
-            \captionof{figure}{A SLICEL Site}
-            \label{fig:slicel}
-        }
-
-        Utilizing all 8 LUT-FF lanes in a SLICE can help reduce device area utilization and minimize wirelength, but \emph{too much} logic density in an area can contribute to general routing congestion.
-        Furthermore, attempting to fill all 8 lanes in a SLICE requires some careful consideration of LUT BEL constraints. 
-        The LUT BELs in a SLICE, as shown in Figure \ref{fig:slicel}, the LUTs appear to be interleaved.
-
-        Show a block diagram of a LUT6 comprising of two LUT5s.
-
-        Recall that a LUT can accommodate one 6-input boolean function or two 5-input boolean functions sharing the same inputs or any two 3-input or less boolean functions regardless of shared inputs.
-        A LUT6 Cell (a LUT with 6-inputs) will actually occupy two lanes in a SLICEL, rendering one of the FF BELs in either lane ineligible for another LUT-FF pair. 
-        The LUT BELs
-
-    \subsection{Packing}
-        \label{subsec:packing}
-
-    \subsection{Placement}
-        \label{subsec:placement}
-        Up until now we have only organized the logical \texttt{Cells} into \texttt{SiteInsts}. 
-        This is where simulated annealing actually begins.
-
-\section{Placement Results}
-    \begin{itemize}
-    \item Random Site Selection:
-    \item Midpoint Site Selection:
-    \item Hybrid: 
-        Random Site selection until progress stagnates, then use Midpoint Site selection. 
-        Randomly choose betwee Random and Midpoint Site selection each iteration. 
-    \item Cooling Schedule: 
-        Greedy. 
-        Geometric with modifiable alpha and intial temperature. 
-    \end{itemize}
-
-\section{Future Work}
-    \begin{itemize}
-    \item Different cooling schedules: linear cooling, logarithmic cooling, piecewise cooling.
-    \item Support for more macros: buffer cells, FIFO, clock generators, etc..
-    \item Support Ultrascale architecture. 
-    \item Better packing scheme. Currently pseudorandom packing. 
-    \item Inter-Site BEL swapping. 
-    \item BEL-centric placement over Site-centric placement.
-    \item Analytical placement. Global placement, legalization, detailed placement. 
-    \end{itemize}
-
-\section{Conclusion}
-    fdsfdsafdsa
-
-% \section{Analytical Placement}
-%     Introduce placement with legalization methods. \\
-%     Many AP approaches use a Global placement, Legalization, then Detailed placement approach. \\
-%     Our SA only considers legal moves via bookkeeping structures, so it has no concept of global placement or legalization. \\
-%     It is a detailed placement only approach. \\
-%     \begin{itemize}
-%         \item Talk about HeAP as an example.
-%         \item Global Placement, Legalization, then Detailed Placement approach.
-%         \item Global placement using the Bound2Bound net model from SimPL.
-%         \item Legalization Step 1: \\
-%             Find an area of the device that is overutilized (illegal) for twhich the blocks contained within must spread to a larger area. \\
-%             To obtain this overutilized area, adjacent locations (Sites, Tiles, etc.) on the device that are occupied by more than one block (SiteInst, ModuleInst, etc.) are repeatedly clustered together, until all clusters are bordered on all sides by non-overutilized locations.
-\\
-%             Then, the area (?) is expanded in both the x and y dimensions (?) until it is large enough to accommodate all blocks contained within. \\
-%             Specifically, the area is expanded until its "occupance" \(O_A\) divided by its "capacity" \(C_A\) is less than a maximum "utilization factor" $\beta$, where $\beta$ is less than or equal to 1.
-HeAP uses 0.9.
-%         \item Legalization Step 2: \\
-%             Two cuts are generated: a "source cut" and a "target cut". \\
-%             The source cut pertains to the blocks being placed (SiteInsts in home buffer). \\ 
-%             The target cut pertains to the area into which the blocks are placed (away buffer). \\
-%             The source cut splits the blocks into 2 partitions, while the target cut splits the area into 2 sub-areas, into which the blocks in each partition are spread. \\
-%             Two objectives are minimized during this process: The imbalance between the number of blocks in each partition, and the difference in the utilization of each sub-area. \\
-%             Utilization defined as the occupancy divided by the capacity of the sub-area. \\
-%             \(U_{sub-area} = \frac{O_{sub-area}}{C_{sub-area}}\) \\
-%             To generate the source cut, the cells are first sorted by their x or y location, depending on the orientation of the desired cut. \\
-%             Once the cells are sorted, the source cut generation is akin to choosing a pivot in a sorted list, where all blocks to the left of the pivot are assigned to the left/bottom partition and all blocks to the right of the pivot are assigned to the right/top partition.
-\\
-%             The target cut is an x or y cut of the area such that all blocks in each partition fit in their respective sub-areas, and such that \( | U_{sub-area_1} 
-%              - U_{sub-area_2} | \) is minimized.
-%     \end{itemize}
-
-\section{Appendix}
-
-    \begin{figure*}[b]
-        \centering
-        \includegraphics[width=\textwidth]{figures/slicem.png}
-        \caption{SLICEM Site}
-        \label{fig:slicem}
-    \end{figure*}
-
-    \begin{figure*}[t]
-        \centering
-        \includegraphics[width=\textwidth]{figures/slicel.png}
-        \caption{SLICEL Site}
-        \label{fig:slicel}
-    \end{figure*}
-
-\end{multicols}
-
-\end{document}

### 2025-05-03 latex/paper/sections/04-architecture_history.tex
split tex file into multiple files

--- 2025-04-28 latex/paper/paper.tex
@@
-\documentclass{article}
-\usepackage{hyperref}
-\usepackage{enumitem}
-\usepackage{graphicx}
-\usepackage{amsmath}
-\usepackage{mathpazo}
-\usepackage{xcolor}
-\usepackage{float}
-\usepackage{listings}
-\lstset{
-  basicstyle=\ttfamily\small,
-  breaklines=true,
-  columns=fullflexible,
-  frame=single,
-  showspaces=false,
-  showtabs=false,
-  keepspaces=false
-}
-\usepackage[export]{adjustbox}
-\usepackage{multicol}
-\usepackage{caption}
-\usepackage[a4paper,
-            bindingoffset=0.2in,
-            left=0.5in,
-            right=0.5in,
-            top=0.5in,
-            bottom=0.5in,
-            footskip=.25in]{geometry}
-
-\begin{document}
-\title{MS Technical Paper: \\ Placement Algorithms for Heterogeneous FPGAs}
-\author{Brian B Cheng \\ Rutgers University Department of Electrical and Computer Engineering}
-
-\date{}
-\maketitle
-
-\begin{multicols}{2}
-
-\section{Keywords}
-\begin{itemize}
-    \item FPGA, EDA, Placement, Simulated Annealing, Optimization, RapidWright
-\end{itemize}
-
-\section{Abstract}
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\
-
-\section{Introduction}
-
-    Field-Programmable Gate Arrays (FPGAs) have witnessed rapid growth in capacity and versatility, driving significant advances in computer-aided design (CAD) and electronic design automation (EDA) methodologies.
-    Since the early-to-mid 2000s, the stagnation of single-processor performance relative to the rapid increase in integrated circuit sizes has led to a design productivity gap, where the computational effort for designing complex chips continues to rise.
-    FPGA CAD flows mainly encompass synthesis, placement, and routing; all of which are NP-hard problems, of which placement is one of the most time-consuming processes. 
-    Inefficienct placement strategy not only extends design times from hours to days, thereby elevating cost and reducing engineering productivity, but also limits the broader adoption of FPGAs by software engineers who expect compile times akin to those of conventional software compilers like {\tt gcc}.
-
-    For these reasons, FPGA placement remains a critical research effort even today. 
-    In this paper, we study and implement established placement methods. 
-    To do this, we use the RapidWright API, which is a semi-open-source research effort from AMD/Xilinx that enables custom solutions to FPGA design implementations and design tools that are not offered by their industry-standard FPGA environment, Vivado.
-    We implement multiple variations of simulated annealing placers for Xilinx's 7-series FPGAs, with an emphasis on minimizing total wirelength while mitigating runtime. 
-    Our implementation is organized into three consecutive substages. 
-    The \textbf{prepacking} stage involves traversing a raw EDIF netlist to identify recurring cell patterns—such as CARRY chains, DSP cascades, and LUT-FF pairs—that are critical for efficient mapping and legalization.
-    In the subsequent \textbf{packing} stage, these identified patterns, along with any remaining loose cells, are consolidated into SiteInst objects that encapsulate the FPGA’s discrete resource constraints and architectural nuances.
-    Finally, the \textbf{placement} stage employs a simulated annealing (SA) algorithm to optimally assign SiteInst objects to physical sites, aiming to minimize total wirelength while adhering to the constraints of the 7-series architecture.
-
-    Simulated annealing iteratively swaps placement objects guided by a cost function that decides which swaps should be accepted or rejected. 
-    Hill climbing is permitted by occasionally accepting moves that increase cost, in hope that such swaps may later lead to a better final solution. 
-    SA remains a popular approach in FPGA placement research due to its simplicity and robustness in handling the discrete architectural constraints of FPGA devices. 
-    While SA yields surprisingly good results given relatively simple rules, it is ultimately a heuristic and stochastic approach that explores the vast placement space by making random moves. 
-    Most of these moves will be rejected, meaning that SA must run many iterations, usually hundreds to thousands, to arrive at a desirable solution.
-
-    In the ASIC domain, where placers must handle designs with millions of cells, the SA approach has largely been abandoned in favor of analytical techniques, owing to SA's runtime and poor scalability.
-    Modern FPGA placers have also followed suit, as new legalization strategies allow FPGA placers to leverage traditionally ASIC placement algorithms and adapt them to the discrete constraints of FPGA architectures.
-    While this paper does not present a working analytical placer, it will explore ways to build upon our existing infrastructure (prepacker and packer) to replace SA with AP.
-
-    The paper first begins by elaborating on general FPGA architecture and then specifically the Xilinx 7-Series architecture. 
-    Then, the paper will elaborate on the FPGA design flow, then the role that the RapidWright API plays in the design flow. 
-    We explain in detail each of these concepts for a broader audience as they provide much needed context for FPGA placement algorithms as a concept. 
-    However, readers who are already familiar with these concepts can skip directly to the RapidWright API section \ref{sec:rapidwright_api} or to the Simulated Annealing section \ref{sec:simulated_annealing}.
-
 \section{FPGA Architecture History}
-
 Before any work can begin on an FPGA placer, it is necessary to understand both the objects being placed and the medium in which they are placed.
@@
 As designs become increasingly large and complex, FPGAs meet the demand by becoming increasingly heterogenous, incorporating a wider variety of hard macros into the fabric.
-
-\section{Xilinx 7-Series Architecture}
-    \label{sec:7_series}
-    The Xilinx 7-Series devices, first introduced in 2010, follow a heterogeneous island-style architecture as discussed previously. 
-    Although the 7-Series was later superseded in 2013 by the UltraScale architecture, the 7-Series remains highly relevant due to its accessibility, wide availability, and compatibility with open-source tooling.
-    Representative sub-families include Artix-7, Kintex-7, Virtex-7, and Zynq-7000, each designed with different performance and cost trade-offs but all follow the core 7-Series architecture.
-
-    Figure~\ref{fig:hierarchy_4} illustrates a high-level view of the hierarchical organization of a 7-Series FPGA. 
-    At its lowest level, the device consists of a large array of atomic components called \emph{Basic Elements of Logic}~(\textbf{BELs}). 
-    These BELs encompass look-up tables (\textbf{LUTs}), flip-flops (\textbf{FFs}), block RAMs (\textbf{BRAMs}), DSP slices (\textbf{DSPs}), and the configurable interconnect fabric. 
-    They constitute the fundamental building blocks for implementing digital circuits on the FPGA.
-    Note how the Tile arrangement is columnar, where each column will have only one Tile type.
-
-    To manage this complexity, Xilinx organizes these BELs into incrementally abstract structures. 
-    First, \textbf{BELs} are grouped into \textbf{Sites}. 
-    Each Site is embedded into a \textbf{Tile}, and Tiles are further arranged into \textbf{Clock Regions}. 
-    In some high-density devices, multiple Clock Regions may be consolidated into one or more \textbf{Super Logic Regions}~(SLRs). 
-    However, for the scope of this paper, we focus on Xilinx 7-Series devices with only a single SLR.
-
-    \end{multicols}
-    {
-        \centering
-        \includegraphics[valign=c, width=0.3\columnwidth]{figures/hierarchy_5.png}
-        \includegraphics[valign=c, width=0.6\columnwidth]{figures/hierarchy_4.png}
-        \captionof{figure}{Architecture Hierarchy of a Xilinx FPGA}
-        \label{fig:hierarchy_4}
-    }
-    \vfill
-    \begin{multicols}{2}
-
-    \newpage
-    \subsection{CLB SLICEs}
-
-    In the 7-Series architecture, the term \emph{CLB} (Configurable Logic Block) refers to a \emph{CLB Tile} that contains two \emph{SLICE} Sites. 
-    Xilinx offers two variants of SLICE Sites: \textbf{SLICEL} and \textbf{SLICEM}.
-    \begin{itemize}
-        \item Each \textbf{SLICEL} has a set of BELs including eight LUTs, eight FFs, and one   \emph{CARRY4} adder. The LUT BELs in a SLICEL can only host LUT Cells. 
-        \item The \textbf{SLICEM} includes all the features of a SLICEL but its LUT BELs can host both LUT Cells, which are asynchronous ROM elements, or RAM32M Cells, which are synchronous 32-deep RAM elements.
-These cells are also referred to as \emph{Distributed RAM} in the Xilinx documentation.
-        These cells can offer an alternative to the larger, more dedicated 18K-36K RAMB18E1 cells when RAM resources are highly utilized. 
-    \end{itemize}
-    In a typical 7-Series device, approximately 75\% of the SLICE Sites are SLICELs and 25\% are SLICEMs. 
-    A single CLB Tile can therefore host either two SLICELs or one SLICEL and one SLICEM.
-    For simplicity, however, we will only consider SLICELs for general logic and use the dedicated RAMB18E1 cells to implement RAM elements.
-
-    The BELs in these SLICEs facilitate the bulk of the general programmability of the FPGA fabric. 
-    We will explain in detail the motivation and function of these BELs.
-
-    \textbf{LUTs} \quad
-        Combinational logic is universal to all HDL designs. 
-        As the their name suggests, Look‑Up Tables (LUTs) facilitate combinational logic by acting as tiny asynchronously‑accessed ROMs whose contents are fixed when the FPGA is programmed.  
-        For any boolean function, the synthesizer precalculates the output to every possible input combination and stores the resulting truth table into a LUT's static memory. 
-        The inputs are then essentially treated as an address space that maps to a data value space in an asynchronous ROM. 
-        No explicit logic gates like NAND or XNOR are synthesized, contrary to what newcomers might expect from a "Field Programmable Gate Array".
-
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/lut_synthesis.png}
-            \captionof{figure}{LUT synthesis from user design}
-            \label{fig:lut_synthesis}
-        }
-        \vspace{1.0cm}
-
-        In the 7‑Series devices, one LUT can facilitate any 6-input boolean function, or two 5-input functions, as long as they share the same input signals.  
-        The LUT can also host two independent boolean functions of up to 3 inputs each, even when the inputs are not shared.  
-        Functions requiring more than six unique inputs are decomposed across multiple cascaded LUTs.
-        Figure \ref{fig:lut_synthesis} shows an example of where a LUT is typically synthesized in a design entry.
-
-    \textbf{FFs} \quad
-        FFs are synthesized to facilitate synchronous event-driven signal assignment. 
-        For most Verilog users, this generally means signal assignments wrapped in \texttt{always @(posedge clk)} statements. 
-        Figure \ref{fig:ff_synthesis} shows an example of where a flip flop (FF) is typically synthesized. 
-        The cell primitive \textbf{FDRE} is a type of FF and belongs to a family of D Flip Flops (DFFs) with Clock Enable (CE). 
-        \begin{itemize}
-            \item FDCE - DFF with CE and Asynchronous Clear
-            \item FDPE - DFF with CE and Asynchronous Preset
-            \item FDSE - DFF with CE and Synchronous Set
-            \item \textbf{FDRE} - DFF with CE and Synchronous Reset
-        \end{itemize}
-        In a typical HDL design, the vast majority of FFs will be synthesized as FDREs with the occasional FDSE, as it is generally good practice to keep FPGA designs synchronous. 
-        A FF BEL may also host a LATCH Cell, however, since they are generally bad practice in FPGA design, we will not consider latches in this paper.
-
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/ff_synthesis.png}
-            \captionof{figure}{FF synthesis from user design}
-            \label{fig:ff_synthesis}
-        }
-        \vspace{1.0cm}
-
-        Up to eight (8) FFs can be placed within the same SLICE, but only if they all share a common Clock-Enable (CE) net and a common Set-Reset (SR) net.
-        This is because the SLICE has only one CE pin and one SR pin to interface with general routing. 
-        The CE and SR signals from these pins are broadcast intra-Site to all FFs within.
-
-    \newpage
-    \textbf{LUT-FF Pairs} \quad
-        FPGA designs are very often modelled as a collection of Finite State Machines (FSM) like shown in Figure \ref{fig:fsm}.
-        Many times a design will also use pipelining, either to model signal buffers or shift registers, or to split up large combinational logic blocks into time slices to meet timing constraints. 
-        These common design structures result in many consecutive sections of combinational logic feeding into a vector of registers. 
-        The synthesizer naturally synthesizes these structures as consecutive pairs of LUTs feeding into FFs as shown in Figure \ref{fig:pipelining}. 
-        Figure \ref{fig:lut_ff_pair} shows an example of a synthesized LUT-FF Pair.
-
-        Shown in Figure \ref{fig:intersite_intrasite} are two possible placements for a LUT-FF Pair on the physical device. 
-        On the right, the cells are placed across different Sites, thus the only way to route the net between the cells is through general inter-site routing. 
-        On the bottom left, the cells are placed within the same Site in the same lane, taking advantage of the intra-site routing without burdening the general router with additional inter-site routing.
-        This is an important consideration to make while minimizing wirelength during placement.
-
-        \vspace{1.0cm}
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/fsm.png}
-            \captionof{figure}{Finite state machine (Moore)}
-            \label{fig:fsm}
-        }
-        \vspace{1.0cm}
-
-        {
-            \centering
-            \includegraphics[width=0.9\columnwidth]{figures/pipelining.png}
-            \captionof{figure}{Pipelining synthesized as consecutive LUT-FF pairs}
-            \label{fig:pipelining}
-        }
-
-        \vfill
-        .
-        \newcolumn
-
-        {
-            \centering
-            \includegraphics[width=0.9\columnwidth]{figures/lut_ff_pair.png}
-            \captionof{figure}{A synthesized LUT-FF Pair}
-            \label{fig:lut_ff_pair}
-        }
-        \vspace{1.0cm}
-
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/intersite_intrasite_2.png}
-            \captionof{figure}{Intrasite vs Intersite LUT-FF Placement}
-            \label{fig:intersite_intrasite}
-        }
-
-    \newpage
-    \textbf{CARRY} \quad
-        An FPGA design will also typically implement many adders, counters, subtractors, or comparators, all of which are based on binary addition. 
-        They are so ubiquitous that every that in the 7-Series architecture, every SLICE features a \emph{CARRY4} BEL -- a 4-bit carry-lookahead (CLA) adder, a much better alternative to synthesizing adders via LUTs.
-
-    \textbf{CARRY Chains} \quad
-        These CARRY4 blocks can be chained across SLICEs to implement wide adders efficiently. 
-        The CARRY4 BELs \emph{must} be chained vertically consecutively across SLICEs as the Carry-In (CI) and Carry-Out (CO) pins can only be routed this way. 
-        A CARRY4 cell may also directly connect to LUTs and FFs, and should be placed in the same Site whenever possible to minimize wirelength. 
-        Shown in figure \ref{fig:carry_cell_group} shows how a CARRY4 chain and associated LUTs and FFs can be placed across SLICEs.
-
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/carry_cell_group.png}
-            \captionof{figure}{A SLICEL with a CARRY4 cell, 4 LUT cells, and 4 FF cells placed inside as shown in the Vivado device viewer}
-            \label{fig:carry_cell_group}
-        }
-
-        {
-            \centering
-            \includegraphics[valign=c, width=4.0cm]{figures/carry_chain_placement.png}
-            \includegraphics[valign=c, width=4.0cm]{figures/carry_chain_routes_3.png}
-            \captionof{figure}{
-                A CARRY4 chain of size 3 placed across 3 SLICEs.
-                \textbf{Left:} Simplified view, \textbf{Right:} As shown in the Vivado device viewer.
-            }
-            \label{fig:carry_cell_group}
-        }
-
-        \end{multicols}
-        % \vfill
-        {
-            \centering
-            \includegraphics[valign=c, width=3.0cm]{figures/adder_synthesis.png}
-            \includegraphics[valign=c, width=14.0cm]{figures/carry_chain_edif_3.png}
-            \captionof{figure}{A synthesized carry chain of size 3 as shown in the Vivado netlist viewer}
-            \label{fig:carry_chain_edif}
-        }
-        \begin{multicols}{2}
-        \vspace{1.0cm}
-        .
-
-    \newpage
-    \subsection{DSP Slices}
-
-    \textbf{DSPs} \quad 
-        FPGAs are often used as low latency Digital Signal Processing (DSP) accelerators. 
-        Common DSP subsystems like Finite Impulse Response (FIR) filters, Fast Fourier Transform (FFTs), and convolutional neural nets (CNNs) demand fast large scale multiply-accumulate (MAC) capabilities.
-        The 7-Series arhictecture integrates DSP BELs into the logic fabric called DSP48E1 that can facilitate MAC efficiently. 
-        The architecture hierarchy for DSPs is simple compared to CLBs and SLICEs. 
-        A DSP48E1 Tile contains two DSP48E1 Sites, each containing one DSP48E1 BEL, which can host a DSP48E1 Cell.
-
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/dsp_diagram.png}
-            \captionof{figure}{
-                Basic DSP48E1 Slice Functionality
-            }
-            \label{dsp_diagram}
-        }
-        \vspace{0.5cm}
-
-    \textbf{DSP Cascades} \quad 
-        Higher level DSP functions are supported by cascading DSP48E1 slices in a DSP48E1 column.
-        Much like CARRY4 chains, DSP48E1 cascades must necessarily be placed vertically consecutively across DSP48E1 Sites. 
-        They are connected by three busses: ACOUT to ACIN, BCOUT to BCIN, PCOUT to PCIN.
-        These signal busses run directly between the vertical DSP48E1 slices without burdening the general routing resources. 
-        The ability to cascade this way provides a high-performance and low-power implementation of DSP filter functions.
-
-        \vspace{1.0cm}
-        {
-            \centering
-            \includegraphics[valign=c, width=2.5cm]{figures/dsp_cascade_placement.png}
-            \includegraphics[valign=c, width=3.0cm]{figures/dsp_cascade_routed.png}
-            \captionof{figure}{
-                A DSP48E1 cascade of size 2 placed across 2 DSP48E1 Sites.
-                \textbf{Left:} Simplified view, \textbf{Right:} As shown in the Vivado device viewer.
-            }
-        }
-
-        \end{multicols}
-        {
-            \centering
-            \includegraphics[valign=c, width=4.0cm]{figures/mult_synthesis.png}
-            \includegraphics[valign=c, width=13.0cm]{figures/dsp_synthesis.png}
-            \captionof{figure}{
-                Simple multiplier synthesis from user design.
-            }
-            \label{mult_synthesis}
-        }
-        \begin{multicols}{2}
-
-    \newpage
-
-    \subsection{Block RAM}
-        In addition to SLICEs and DSPs, the 7-Series also offers dedicated Block Random Access Memory (BRAM) BELs. 
-        These BRAMs come in two variants: \textbf{RAMB18E1} and \textbf{RAMB36E1}. 
-        \begin{itemize}
-            \item \textbf{RAMB18E1} - 
-            Has a capacity of 18 Kilobits. 
-            It can be configured as single port RAM with dimensions ranging between (1-bit wide by 16K deep) to (18-bit wide by 1024 deep). 
-            It can also be configured as a (36-bit wide by 512 deep) true simple dual port RAM. 
-            \item \textbf{RAMB36E1} - 
-            Has a capacity of 36 Kilobits. 
-            It can be configured as single port RAM with dimensions ranging between (1-bit wide by 32K deep) to (36-bit wide by 1024 deep). 
-            It can also be configured as a (72-bit wide by 512 deep) simple dual port RAM. 
-        \end{itemize}
-
-        For readability, we will abbreviate RAMB18E1 as \textbf{RAM18} and RAMB36E1 as \textbf{RAM36}.
-        One BRAM Tile contains one RAM36 Site and two RAM18 Sites. 
-        The RAM36 Site contains one RAM36 BEL while the RAM18 Site contains one RAM18 BEL.
-        The RAM18 BEL can host one RAM18 Cell, while the RAM36 BEL can host either one RAM36 Cell or two RAM18 Cells.
-
-        Like DSP cascades, BRAMs may also be cascaded together in a column to implement large memories efficiently, with some intermediate signals between them routed intra-Tile without burdening the general routing.
-        However, unlike DSPs, large memories decomposed amongst multiple BRAMs can also be routed together through general routing. 
-        Furthermore, in most design scenarios, large memories will not utilize the intra-Tile signals.
-
-        For simplicity we will not constrain large BRAMs to be cascaded together in consecutive vertical Tiles. 
-        We can essentially treat RAMB18E1 and RAMB36E1 Cells as loose, minimally constrained single cells, in contrast to the highly constrained DSP cascades and CARRY chains.
-
-        {
-            \centering
-            \includegraphics[valign=c, width=3.0cm]{figures/bram.png}
-            \includegraphics[valign=c, width=2.0cm]{figures/bram_tile.png}
-            \captionof{figure}{
-                A BRAM Tile containing two RAMB18E1 Sites and one RAMB36E1 Site.
-                \textbf{Left:} simplified view. \textbf{Right:} as seen in the device viewer, BELs highlighted in white. 
-            }
-            \label{fig:bram_tile}
-        }
-
-        {
-            \centering
-            \includegraphics[valign=c, width=4.0cm]{figures/bram_synthesis.png}
-            \includegraphics[valign=c, width=\columnwidth]{figures/bram_edif.png}
-            \captionof{figure}{
-                An example of BRAM synthesis (via inference)
-            }
-            \label{fig:bram_synthesis}
-        }
-
-    \subsection{Further Documentation}
-
-    For more in-depth details about 7-Series FPGAs, refer to the official Xilinx user guides such as:
-    \begin{itemize}
-        \item \emph{7 Series FPGAs Overview} (UG476)
-        \item \emph{7 Series FPGA Configurable Logic Block} (UG474)
-        \item \emph{7 Series Memory Resources} (UG473)
-        \item \emph{7 Series DSP48E1 Slice} (UG479)
-    \end{itemize}
-
-    This architectural context provides the necessary background for understanding how a placement algorithm should account for resource constraints and optimize performance in modern FPGA designs.
-
-\newpage
-\section{FPGA Design Flow and Toolchain}
-    \label{sec:fpga_flow_toolchain}
-
-    Modern FPGA designs require a sophisticated toolchain to bridge the gap between high-level hardware descriptions and the final bitstream used to configure the FPGA. 
-    Figure~\ref{fig:design_flow} illustrates a representative process that converts an abstract Hardware Description Language (HDL) design into a verified configuration file for a target device.
-
-    {
-        \centering
-        \includegraphics[width=0.9\columnwidth]{figures/design_flow.png}
-        \captionof{figure}{A typical FPGA design and verification workflow.}
-        \label{fig:design_flow}
-    }
-
-    \begin{enumerate}
-        \item \textbf{Design Entry:} 
-            An engineer describes the intended functionality of the digital system using a hardware description language (HDL) such as Verilog or VHDL. 
-            During this phase, the coding style can vary (behavioral, structural, dataflow, etc.), but it always aims to capture high-level behavior rather than device-specific details.
-
-        \item \textbf{Synthesis:} 
-            The synthesis tool parses the HDL source, performs logical optimizations, and maps the design onto primitive cells that suit the target FPGA technology. 
-            The output is typically a structural \texttt{netlist} (e.g., EDIF or structural Verilog) which details how the design’s logic is broken down into LUTs, FFs, and other vendor-specific cells.
-
-        \item \textbf{Placement and Routing (Implementation):} 
-            In \emph{placement}, each logical cell from the synthesized netlist is assigned to a physical location on the FPGA fabric. 
-            For instance, LUTs and FFs go into specific \emph{BELs} within the device’s CLB sites, and specialized cells such as DSPs and Block RAMs must be placed in their corresponding tile types. 
-            Next, \emph{routing} determines how signals are physically wired through the FPGA’s configurable interconnect network. 
-            Modern tools often interleave these steps (e.g., fluid-placement routing or routing-aware placement) to better meet timing and area objectives.
-
-        \item \textbf{Bitstream Generation:} 
-            After a design is fully placed, routed, and timing-closed, the toolchain produces a final \emph{bitstream} that sets the configuration of every programmable element in the FPGA. 
-            This bitstream can then be loaded onto the device, either through vendor software or via a custom programming interface.
-
-        \item \textbf{Verification:} 
-            In parallel to the design flow, simulations and testbenches validate correctness of the user's design at multiple abstraction levels. 
-            Engineers may begin with behavioral simulations, then progress to post-synthesis simulations, and finally to post-implementation simulations that incorporate estimated routing delays. 
-            With each higher level of fidelity, computational requirements grow significantly due to increasing complexity and the need to analyze more variables over time. 
-            Ensuring correct functionality and meeting timing closure at the post-implementation stage is crucial before deploying the design to hardware. 
-            Given the importance of thorough verification, many established companies dedicate one verification engineer for every design engineer.
-
-    \end{enumerate}
-
-    This workflow underscores the critical role of \textbf{placement} in bridging the netlist to a physical realization. 
-    An efficient placement algorithm can drastically reduce compile times and improve design performance, enabling broader adoption of FPGAs in application spaces that require fast design iteration.
-
-\newpage
-\section{RapidWright API}
-    \label{sec:rapidwright_api}
-
-    \textbf{RapidWright} is an open-source Java framework from AMD/Xilinx that provides direct access to the netlist and device databases used by vendor tools. 
-    This framework positions itself as an additional workflow column, allowing users to intercept or replace stages of the standard design flow with custom optimization stages (see Figure~\ref{fig:vivado_dcps}).
-
-    \begin{itemize}
-    \item \textbf{Design Checkpoints:} 
-        RapidWright leverages \texttt{.dcp} files (design checkpoints) generated at various stages of a Vivado flow. 
-        By importing a checkpoint, engineers can manipulate the netlist, placement, or routing externally, then re-export a modified checkpoint for further processing in the Vivado workflow column.
-
-    \item \textbf{Key Packages:} 
-        RapidWright revolves around three primary data model packages:
-        \begin{enumerate}
-        \item \emph{edif} -- Represents the logical netlist in an abstracted EDIF-like structure.
-        \item \emph{design} -- Contains data structures for the physical implementation (Cells, Nets, Sites, BELs, etc.).
-        \item \emph{device} -- Provides a database of the target FPGA architecture (e.g., Site coordinates, Tile definitions, routing resources).
-        \end{enumerate}
-
-    \item \textbf{Interfacing with the Netlist and Device:} 
-        An engineer can query the netlist to find specific resources (LUTs, FFs, DSPs, etc.) and then map or move them onto device sites. 
-        This level of control over backend resources is necessary for research in custom placement, advanced packing techniques, or experimental routing algorithms.
-    \end{itemize}
-
-    {
-        \centering
-        \includegraphics[width=0.8\columnwidth]{figures/vivado_dcps.png}
-        \captionof{figure}{RapidWright workflow integrating into the default Vivado design flow.}
-        \label{fig:vivado_dcps}
-    }
-
-    By exposing these low-level internals, RapidWright allows fine-grained design transformations that go beyond the standard Vivado IDE’s capabilities. 
-    Researchers can prototype new EDA strategies without needing to re-implement an entire FPGA backend from scratch, thus accelerating innovation in placement and routing methodologies.
-
-\newpage
-\section{What is a Netlist?}
-    \label{sec:netlist}
-    In its simplest form, a netlist consists of a list of every component in an electronic design and a list of nets they connect to. 
-    Here, a net, or network, denotes any group of two or more interconnected elements - whether they be transistors, logic gates, or higher-level macrocells. 
-    More precisely, nets attach to component pins, and those pins reside on cells, each of which exposing multiple ports. 
-    In this way, a netlist makes explicit both the individual building blocks of a circuit and the precise wiring that links them.
-
-    In Vivado, a Netlist can be synthesized as a Hierarchical or a Flattened netlist. 
-    Figure ~\ref{fig:hierarchical_design} shows the Verilog code for a simple HDL design using module instance hierarchy. 
-    Figure ~\ref{fig:hier_netlist} shows the design synthesized into a hierarchical netlist, while Figure ~\ref{fig:flat_netlist} shows the same design synthesized into a flat netlist. 
-    In a synthesized Netlist, the leaf cells (lowest-level cells) are necessarily primitive cells from the vendor primitives library such as LUT6, FDRE, CARRY4, DSP48E1, etc..
-
-    \end{multicols}
-    {
-        \raggedright
-        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/top_level.png}
-        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_0.png}
-        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_1.png}
-        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_2.png}
-        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_3.png}
-        \captionof{figure}{A simple HDL design with module hierarchy.}
-        \label{fig:hierarchical_design}
-    }
-    {
-        \centering
-        \includegraphics[valign=c, width=11.5cm]{figures/netlist_synth/hier_netlist.png}
-        \includegraphics[valign=c, width=6cm]{figures/netlist_synth/hier_graph.png}
-        \captionof{figure}{
-            \textbf{Left:} A hierarchical netlist consisting of LUTs and FFs.
-            \textbf{Right:} The cell hierarchy graph.
-        }
-        \label{fig:hier_netlist}
-    }
-    \vspace{0.5cm}
-    {
-        \centering
-        \includegraphics[valign=c, width=10cm]{figures/netlist_synth/flat_netlist.png}
-        \includegraphics[valign=c, width=4cm]{figures/netlist_synth/flat_graph.png}
-        \captionof{figure}{
-            \textbf{Left:} A flattened netlist consisting of LUTs and FFs.
-            \textbf{Right:} The flattened cell hierarchy graph.
-        }
-        \label{fig:flat_netlist}
-    }
-    \begin{multicols}{2}
-
-    RapidWright offers an easy way to access, read, or manipulate this netlist for a given HDL design. 
-    Shown in Figure~\ref{fig:netlist_syn}
-
-    Verilog \texttt{module}s are represented simply as
-
-    \end{multicols}
-    {
-        \centering
-        \includegraphics[width=\columnwidth]{figures/raw_edif.png}
-        \captionof{figure}{A raw EDIF Netlist with uncountably many cells}
-        \label{fig:raw_edif}
-    }
-    \begin{multicols}{2}
-
-    .
-
-\newpage
-\section{Simulated Annealing}
-    \label{sec:simulated_annealing}
-    With a basic understanding of FPGA architecture, design placement, and RapidWright, we have all the necessary pieces to implement our SA placer. 
-    Here we outline in detail each substage of our implementation: PrePacking, Packing, and Placement. 
-    Shown in Figure \ref{fig:edif_design_device} shows an overview of the placement workflow.
-
-    {
-        \centering
-        \includegraphics[width=0.9\columnwidth]{figures/edif_design_device.png}
-        \captionof{figure}{Our placement workflow}
-        \label{fig:edif_design_device}
-    }
-
-    \end{multicols}
-    {
-        \centering
-        \includegraphics[width=0.8\columnwidth]{figures/substages.png}
-        \captionof{figure}{The data classes populated at each substage: PrepackedDesign, PackedDesign, and PlacedDesign.}
-        \label{fig:substages}
-    }
-    \begin{multicols}{2}
-
-    \subsection{Prepacking}
-        \label{subsec:prepacking}
-
-        The first step in our placement flow is \textbf{prepacking}. 
-        Recall from the 7-Series architecture that there are certain cell structures that must adhere to certain placements constraints to ensure legality, and by design, to minimize wirelength. 
-        The job of the prepacker is to consolidate the cells in the EDIF Netlist into clusters or groups of clusters that naturally reflect these placement constraints.
-
-        CARRY4 chains must necessarily be placed vertically and consecutively across a column of SLICEs in ascending order. 
-        Likewise, DSP48E1 cascades must necessarily be placed vertically and consecutively across a column of DSP48E1 Sites in ascending order. 
-        A LUT-FF pair may be placed freely, but should be placed in the same lane within the same SLICE to minimize wirelength.
-
-        The raw post-synthesis netlist only tells us the list of nets and the cell ports that they connect to. 
-        It does not report the presence of any macro cell structures (CARRY4 chains, etc.). 
-        Thus, we must traverse the netlist to detect these cell structures and store that structure information in a class we will call \texttt{PrepackedDesign}.
-
-        A group of LUT-FF pairs may be placed in the same SLICE, with the constraint that the FFs must share the same Clock-Enable (CE) and Set-Reset (SR) nets.
-        Clustering LUT-FF pairs like this can reduce the redundancy of having to route the same CE and SR nets to many SLICEs by routing the nets to fewer SLICEs and connecting them to the individual FFs within via intra-Site routing, and at the same time, packing greater logic density over a smaller area of the device.
-        Recall that up to eight FFs may be placed in the same SLICE, thus theoretically, up to eight LUT-FF pairs can be placed in the same SLICE.
-
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/slicel.png}
-            \captionof{figure}{A SLICEL Site}
-            \label{fig:slicel}
-        }
-
-        Utilizing all 8 LUT-FF lanes in a SLICE can help reduce device area utilization and minimize wirelength, but \emph{too much} logic density in an area can contribute to general routing congestion.
-        Furthermore, attempting to fill all 8 lanes in a SLICE requires some careful consideration of LUT BEL constraints. 
-        The LUT BELs in a SLICE, as shown in Figure \ref{fig:slicel}, the LUTs appear to be interleaved.
-
-        Show a block diagram of a LUT6 comprising of two LUT5s.
-
-        Recall that a LUT can accommodate one 6-input boolean function or two 5-input boolean functions sharing the same inputs or any two 3-input or less boolean functions regardless of shared inputs.
-        A LUT6 Cell (a LUT with 6-inputs) will actually occupy two lanes in a SLICEL, rendering one of the FF BELs in either lane ineligible for another LUT-FF pair. 
-        The LUT BELs
-
-    \subsection{Packing}
-        \label{subsec:packing}
-
-    \subsection{Placement}
-        \label{subsec:placement}
-        Up until now we have only organized the logical \texttt{Cells} into \texttt{SiteInsts}. 
-        This is where simulated annealing actually begins.
-
-\section{Placement Results}
-    \begin{itemize}
-    \item Random Site Selection:
-    \item Midpoint Site Selection:
-    \item Hybrid: 
-        Random Site selection until progress stagnates, then use Midpoint Site selection. 
-        Randomly choose betwee Random and Midpoint Site selection each iteration. 
-    \item Cooling Schedule: 
-        Greedy. 
-        Geometric with modifiable alpha and intial temperature. 
-    \end{itemize}
-
-\section{Future Work}
-    \begin{itemize}
-    \item Different cooling schedules: linear cooling, logarithmic cooling, piecewise cooling.
-    \item Support for more macros: buffer cells, FIFO, clock generators, etc..
-    \item Support Ultrascale architecture. 
-    \item Better packing scheme. Currently pseudorandom packing. 
-    \item Inter-Site BEL swapping. 
-    \item BEL-centric placement over Site-centric placement.
-    \item Analytical placement. Global placement, legalization, detailed placement. 
-    \end{itemize}
-
-\section{Conclusion}
-    fdsfdsafdsa
-
-% \section{Analytical Placement}
-%     Introduce placement with legalization methods. \\
-%     Many AP approaches use a Global placement, Legalization, then Detailed placement approach. \\
-%     Our SA only considers legal moves via bookkeeping structures, so it has no concept of global placement or legalization. \\
-%     It is a detailed placement only approach. \\
-%     \begin{itemize}
-%         \item Talk about HeAP as an example.
-%         \item Global Placement, Legalization, then Detailed Placement approach.
-%         \item Global placement using the Bound2Bound net model from SimPL.
-%         \item Legalization Step 1: \\
-%             Find an area of the device that is overutilized (illegal) for twhich the blocks contained within must spread to a larger area. \\
-%             To obtain this overutilized area, adjacent locations (Sites, Tiles, etc.) on the device that are occupied by more than one block (SiteInst, ModuleInst, etc.) are repeatedly clustered together, until all clusters are bordered on all sides by non-overutilized locations.
-\\
-%             Then, the area (?) is expanded in both the x and y dimensions (?) until it is large enough to accommodate all blocks contained within. \\
-%             Specifically, the area is expanded until its "occupance" \(O_A\) divided by its "capacity" \(C_A\) is less than a maximum "utilization factor" $\beta$, where $\beta$ is less than or equal to 1.
-HeAP uses 0.9.
-%         \item Legalization Step 2: \\
-%             Two cuts are generated: a "source cut" and a "target cut". \\
-%             The source cut pertains to the blocks being placed (SiteInsts in home buffer). \\ 
-%             The target cut pertains to the area into which the blocks are placed (away buffer). \\
-%             The source cut splits the blocks into 2 partitions, while the target cut splits the area into 2 sub-areas, into which the blocks in each partition are spread. \\
-%             Two objectives are minimized during this process: The imbalance between the number of blocks in each partition, and the difference in the utilization of each sub-area. \\
-%             Utilization defined as the occupancy divided by the capacity of the sub-area. \\
-%             \(U_{sub-area} = \frac{O_{sub-area}}{C_{sub-area}}\) \\
-%             To generate the source cut, the cells are first sorted by their x or y location, depending on the orientation of the desired cut. \\
-%             Once the cells are sorted, the source cut generation is akin to choosing a pivot in a sorted list, where all blocks to the left of the pivot are assigned to the left/bottom partition and all blocks to the right of the pivot are assigned to the right/top partition.
-\\
-%             The target cut is an x or y cut of the area such that all blocks in each partition fit in their respective sub-areas, and such that \( | U_{sub-area_1} 
-%              - U_{sub-area_2} | \) is minimized.
-%     \end{itemize}
-
-\section{Appendix}
-
-    \begin{figure*}[b]
-        \centering
-        \includegraphics[width=\textwidth]{figures/slicem.png}
-        \caption{SLICEM Site}
-        \label{fig:slicem}
-    \end{figure*}
-
-    \begin{figure*}[t]
-        \centering
-        \includegraphics[width=\textwidth]{figures/slicel.png}
-        \caption{SLICEL Site}
-        \label{fig:slicel}
-    \end{figure*}
-
-\end{multicols}
-
-\end{document}

### 2025-05-03 latex/paper/sections/05-7series_architecture.tex
split tex file into multiple files

--- 2025-04-28 latex/paper/paper.tex
@@
-\documentclass{article}
-\usepackage{hyperref}
-\usepackage{enumitem}
-\usepackage{graphicx}
-\usepackage{amsmath}
-\usepackage{mathpazo}
-\usepackage{xcolor}
-\usepackage{float}
-\usepackage{listings}
-\lstset{
-  basicstyle=\ttfamily\small,
-  breaklines=true,
-  columns=fullflexible,
-  frame=single,
-  showspaces=false,
-  showtabs=false,
-  keepspaces=false
-}
-\usepackage[export]{adjustbox}
-\usepackage{multicol}
-\usepackage{caption}
-\usepackage[a4paper,
-            bindingoffset=0.2in,
-            left=0.5in,
-            right=0.5in,
-            top=0.5in,
-            bottom=0.5in,
-            footskip=.25in]{geometry}
-
-\begin{document}
-\title{MS Technical Paper: \\ Placement Algorithms for Heterogeneous FPGAs}
-\author{Brian B Cheng \\ Rutgers University Department of Electrical and Computer Engineering}
-
-\date{}
-\maketitle
-
-\begin{multicols}{2}
-
-\section{Keywords}
-\begin{itemize}
-    \item FPGA, EDA, Placement, Simulated Annealing, Optimization, RapidWright
-\end{itemize}
-
-\section{Abstract}
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\
-
-\section{Introduction}
-
-    Field-Programmable Gate Arrays (FPGAs) have witnessed rapid growth in capacity and versatility, driving significant advances in computer-aided design (CAD) and electronic design automation (EDA) methodologies.
-    Since the early-to-mid 2000s, the stagnation of single-processor performance relative to the rapid increase in integrated circuit sizes has led to a design productivity gap, where the computational effort for designing complex chips continues to rise.
-    FPGA CAD flows mainly encompass synthesis, placement, and routing; all of which are NP-hard problems, of which placement is one of the most time-consuming processes. 
-    Inefficienct placement strategy not only extends design times from hours to days, thereby elevating cost and reducing engineering productivity, but also limits the broader adoption of FPGAs by software engineers who expect compile times akin to those of conventional software compilers like {\tt gcc}.
-
-    For these reasons, FPGA placement remains a critical research effort even today. 
-    In this paper, we study and implement established placement methods. 
-    To do this, we use the RapidWright API, which is a semi-open-source research effort from AMD/Xilinx that enables custom solutions to FPGA design implementations and design tools that are not offered by their industry-standard FPGA environment, Vivado.
-    We implement multiple variations of simulated annealing placers for Xilinx's 7-series FPGAs, with an emphasis on minimizing total wirelength while mitigating runtime. 
-    Our implementation is organized into three consecutive substages. 
-    The \textbf{prepacking} stage involves traversing a raw EDIF netlist to identify recurring cell patterns—such as CARRY chains, DSP cascades, and LUT-FF pairs—that are critical for efficient mapping and legalization.
-    In the subsequent \textbf{packing} stage, these identified patterns, along with any remaining loose cells, are consolidated into SiteInst objects that encapsulate the FPGA’s discrete resource constraints and architectural nuances.
-    Finally, the \textbf{placement} stage employs a simulated annealing (SA) algorithm to optimally assign SiteInst objects to physical sites, aiming to minimize total wirelength while adhering to the constraints of the 7-series architecture.
-
-    Simulated annealing iteratively swaps placement objects guided by a cost function that decides which swaps should be accepted or rejected. 
-    Hill climbing is permitted by occasionally accepting moves that increase cost, in hope that such swaps may later lead to a better final solution. 
-    SA remains a popular approach in FPGA placement research due to its simplicity and robustness in handling the discrete architectural constraints of FPGA devices. 
-    While SA yields surprisingly good results given relatively simple rules, it is ultimately a heuristic and stochastic approach that explores the vast placement space by making random moves. 
-    Most of these moves will be rejected, meaning that SA must run many iterations, usually hundreds to thousands, to arrive at a desirable solution.
-
-    In the ASIC domain, where placers must handle designs with millions of cells, the SA approach has largely been abandoned in favor of analytical techniques, owing to SA's runtime and poor scalability.
-    Modern FPGA placers have also followed suit, as new legalization strategies allow FPGA placers to leverage traditionally ASIC placement algorithms and adapt them to the discrete constraints of FPGA architectures.
-    While this paper does not present a working analytical placer, it will explore ways to build upon our existing infrastructure (prepacker and packer) to replace SA with AP.
-
-    The paper first begins by elaborating on general FPGA architecture and then specifically the Xilinx 7-Series architecture. 
-    Then, the paper will elaborate on the FPGA design flow, then the role that the RapidWright API plays in the design flow. 
-    We explain in detail each of these concepts for a broader audience as they provide much needed context for FPGA placement algorithms as a concept. 
-    However, readers who are already familiar with these concepts can skip directly to the RapidWright API section \ref{sec:rapidwright_api} or to the Simulated Annealing section \ref{sec:simulated_annealing}.
-
-\section{FPGA Architecture History}
-
-    Before any work can begin on an FPGA placer, it is necessary to understand both the objects being placed and the medium in which they are placed.
-    Configurable logic devices have undergone significant evolution over the past four decades. 
-    We will briefly review the evolution of configurable logic architecture starting in the 1970s and quickly work our way up to modern day FPGA architecture.
-
-    \textbf{PLA:} \quad
-    The journey began with the Programmable Logic Array (PLA) in the early 1970s. 
-    The PLA implemented output logic using a programmable-OR and programmable-AND plane that formed a sum-of-products equation for each output through programmable fuses. 
-    Around the same time, the Programmable Array Logic (PAL) was introduced. 
-    The PAL simplified the PLA by fixing the OR gates, resulting in a fixed-OR, programmable-AND design, which sacrificed some logic flexibility to simplify its manufacture. 
-    Figure \ref{fig:pal_2} shows one such PAL architecture.
-
-    \textbf{CPLD:} \quad 
-    Later in the same decade came the Complex Programmable Logic Device (CPLD), which took the form of an array of Configurable Logic Blocks (CLBs). 
-    These CLBs were typically modified PAL blocks that included the PAL itself along with macrocells such as flip-flops, multiplexers, and tri-state buffers. 
-    The CPLD functioned as an array of PALs connected by a central programmable switch matrix and could be programmed using a hardware description language (HDL) like VHDL. 
-    Figure \ref{fig:cpld} shows one such CPLD architecture. 
-    \\
-
-    {
-        \centering
-        \includegraphics[width=\columnwidth]{figures/pal_2.png}
-        \captionof{figure}{PAL architecture with 5 inputs, 8 programmable AND gates and 4 fixed OR gates}
-        \label{fig:pal_2}
-        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/Programmable-Logic-Devices/Programmable-Array-Logic-PAL/
-        % https://www.naukri.com/code360/library/difference-between-pla-and-pal 
-    }
-
-    \newcolumn
-
-    \textbf{Homogeneous FPGA:} \quad 
-    The mid-1980s saw the introduction of homogeneous FPGAs, which were built as a grid of CLBs. 
-    Rather than using a central programmable switch matrix as in CPLDs, FPGAs adopted an island style architecture in which each CLB is surrounded on all sides by programmable routing resources, as shown in Figure \ref{fig:fpga}.
-    The first commercially viable FPGA, produced by Xilinx in 1984, featured 16 CLBs arranged in a 4x4 grid. 
-    As FPGA technology advanced, CLBs were redesigned to use lookup tables (LUTs) instead of PAL arrays for greater logic density. 
-    The capacity of an FPGA was often measured by how many logical elements or CLBs it offered, which grew from hundreds to thousands and now to hundreds of thousands of CLBs.
-    \\
-
-    {
-        \centering
-        \includegraphics[width=0.8\columnwidth]{figures/cpld.png}
-        \captionof{figure}{CPLD architecture with 4 CLBs (PAL-like blocks)}
-        \label{fig:cpld}
-        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/CPLD/Complex-Programmable-Logic-Device-CPLDs/
-    }
-    {
-        \centering
-        \includegraphics[width=\columnwidth]{figures/homogenous_fpga.png}
-        \captionof{figure}{A homogeneous island-style FPGA architecture with 16 CLBs in a grid.}
-        \label{fig:fpga}
-        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/FPGA/Field-programmable-gate-array-FPGA/
-    }
-
-    \newpage
-
-    {
-        \centering
-        \includegraphics[width=\columnwidth]{figures/heterogenous_fpga_3.jpg}
-        \captionof{figure}{A heterogeneous island-style FPGA with a mix of CLBs and macrocells.}
-        \label{fig:fpga}
-        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/FPGA/Field-programmable-gate-array-FPGA/
-    }
-    \vspace{0.5cm}
-
-    \textbf{Heterogenous FPGA:} \quad 
-    This brings us to modern day FPGA architectures. 
-    To meet the needs of increasingly complex designs, FPGA vendors introduced heterogeneous FPGAs. 
-    In these devices, hard macros such as Block RAM (BRAM) and Digital Signal Processing (DSP) slices are integrated into the programmable logic fabric along with CLBs, like shown in Figure \ref{fig:fpga}.
-    This design enables the direct instantiation of common subsystems like memories and multipliers, without having to recreate them from scratch using CLBs. 
-    Major vendors such as Xilinx and Altera now employ heterogeneous island-style architectures in their devices. 
-    As designs become increasingly large and complex, FPGAs meet the demand by becoming increasingly heterogenous, incorporating a wider variety of hard macros into the fabric.
-
 \section{Xilinx 7-Series Architecture}
@@
 This architectural context provides the necessary background for understanding how a placement algorithm should account for resource constraints and optimize performance in modern FPGA designs.
-
-\newpage
-\section{FPGA Design Flow and Toolchain}
-    \label{sec:fpga_flow_toolchain}
-
-    Modern FPGA designs require a sophisticated toolchain to bridge the gap between high-level hardware descriptions and the final bitstream used to configure the FPGA. 
-    Figure~\ref{fig:design_flow} illustrates a representative process that converts an abstract Hardware Description Language (HDL) design into a verified configuration file for a target device.
-
-    {
-        \centering
-        \includegraphics[width=0.9\columnwidth]{figures/design_flow.png}
-        \captionof{figure}{A typical FPGA design and verification workflow.}
-        \label{fig:design_flow}
-    }
-
-    \begin{enumerate}
-        \item \textbf{Design Entry:} 
-            An engineer describes the intended functionality of the digital system using a hardware description language (HDL) such as Verilog or VHDL. 
-            During this phase, the coding style can vary (behavioral, structural, dataflow, etc.), but it always aims to capture high-level behavior rather than device-specific details.
-
-        \item \textbf{Synthesis:} 
-            The synthesis tool parses the HDL source, performs logical optimizations, and maps the design onto primitive cells that suit the target FPGA technology. 
-            The output is typically a structural \texttt{netlist} (e.g., EDIF or structural Verilog) which details how the design’s logic is broken down into LUTs, FFs, and other vendor-specific cells.
-
-        \item \textbf{Placement and Routing (Implementation):} 
-            In \emph{placement}, each logical cell from the synthesized netlist is assigned to a physical location on the FPGA fabric. 
-            For instance, LUTs and FFs go into specific \emph{BELs} within the device’s CLB sites, and specialized cells such as DSPs and Block RAMs must be placed in their corresponding tile types. 
-            Next, \emph{routing} determines how signals are physically wired through the FPGA’s configurable interconnect network. 
-            Modern tools often interleave these steps (e.g., fluid-placement routing or routing-aware placement) to better meet timing and area objectives.
-
-        \item \textbf{Bitstream Generation:} 
-            After a design is fully placed, routed, and timing-closed, the toolchain produces a final \emph{bitstream} that sets the configuration of every programmable element in the FPGA. 
-            This bitstream can then be loaded onto the device, either through vendor software or via a custom programming interface.
-
-        \item \textbf{Verification:} 
-            In parallel to the design flow, simulations and testbenches validate correctness of the user's design at multiple abstraction levels. 
-            Engineers may begin with behavioral simulations, then progress to post-synthesis simulations, and finally to post-implementation simulations that incorporate estimated routing delays. 
-            With each higher level of fidelity, computational requirements grow significantly due to increasing complexity and the need to analyze more variables over time. 
-            Ensuring correct functionality and meeting timing closure at the post-implementation stage is crucial before deploying the design to hardware. 
-            Given the importance of thorough verification, many established companies dedicate one verification engineer for every design engineer.
-
-    \end{enumerate}
-
-    % This workflow underscores the critical role of \textbf{placement} in bridging the netlist to a physical realization. 
-    % An efficient placement algorithm can drastically reduce compile times and improve design performance, enabling broader adoption of FPGAs in application spaces that require fast design iteration.
-
-\newpage
-\section{RapidWright API}
-    \label{sec:rapidwright_api}
-
-    \textbf{RapidWright} is an open-source Java framework from AMD/Xilinx that provides direct access to the netlist and device databases used by vendor tools. 
-    This framework positions itself as an additional workflow column, allowing users to intercept or replace stages of the standard design flow with custom optimization stages (see Figure~\ref{fig:vivado_dcps}).
-
-    \begin{itemize}
-    \item \textbf{Design Checkpoints:} 
-        RapidWright leverages \texttt{.dcp} files (design checkpoints) generated at various stages of a Vivado flow. 
-        By importing a checkpoint, engineers can manipulate the netlist, placement, or routing externally, then re-export a modified checkpoint for further processing in the Vivado workflow column.
-
-    \item \textbf{Key Packages:} 
-        RapidWright revolves around three primary data model packages:
-        \begin{enumerate}
-        \item \emph{edif} -- Represents the logical netlist in an abstracted EDIF-like structure.
-        \item \emph{design} -- Contains data structures for the physical implementation (Cells, Nets, Sites, BELs, etc.).
-        \item \emph{device} -- Provides a database of the target FPGA architecture (e.g., Site coordinates, Tile definitions, routing resources).
-        \end{enumerate}
-
-    \item \textbf{Interfacing with the Netlist and Device:} 
-        An engineer can query the netlist to find specific resources (LUTs, FFs, DSPs, etc.) and then map or move them onto device sites. 
-        This level of control over backend resources is necessary for research in custom placement, advanced packing techniques, or experimental routing algorithms.
-    \end{itemize}
-
-    {
-        \centering
-        \includegraphics[width=0.8\columnwidth]{figures/vivado_dcps.png}
-        \captionof{figure}{RapidWright workflow integrating into the default Vivado design flow.}
-        \label{fig:vivado_dcps}
-    }
-
-    By exposing these low-level internals, RapidWright allows fine-grained design transformations that go beyond the standard Vivado IDE’s capabilities. 
-    Researchers can prototype new EDA strategies without needing to re-implement an entire FPGA backend from scratch, thus accelerating innovation in placement and routing methodologies.
-
-\newpage
-\section{What is a Netlist?}
-    \label{sec:netlist}
-    In its simplest form, a netlist consists of a list of every component in an electronic design and a list of nets they connect to. 
-    Generally, a net, or network, denotes any group of two or more interconnected elements - whether they be transistors, logic gates, or in our case, logical cells. 
-    More precisely, nets attach to component pins, and those pins reside on cells, each of which exposing multiple ports. 
-    In this way, a netlist makes explicit both the individual building blocks of a circuit and the precise wiring that links them.
-
-    In Vivado, a Netlist can be synthesized as a Hierarchical or a Flattened netlist. 
-    Figure ~\ref{fig:hierarchical_design} shows the Verilog code for a simple HDL design using module instance hierarchy. 
-    Figure ~\ref{fig:hier_netlist} shows the design synthesized into a hierarchical netlist, while Figure ~\ref{fig:flat_netlist} shows the same design synthesized into a flat netlist. 
-    In a synthesized Netlist, the leaf cells (lowest-level cells) are necessarily primitive cells from the vendor primitives library such as \texttt{LUT6}, \texttt{FDRE}, \texttt{CARRY4}, \texttt{DSP48E1}, etc..
-
-    \end{multicols}
-    {
-        \raggedright
-        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/top_level.png}
-        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_0.png}
-        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_1.png}
-        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_2.png}
-        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_3.png}
-        \captionof{figure}{A simple HDL design with module hierarchy.}
-        \label{fig:hierarchical_design}
-    }
-    {
-        \centering
-        \includegraphics[valign=c, width=11.5cm]{figures/netlist_synth/hier_netlist.png}
-        \includegraphics[valign=c, width=6cm]{figures/netlist_synth/hier_graph.png}
-        \captionof{figure}{
-            \textbf{Left:} A hierarchical netlist consisting of LUTs and FFs.
-            \textbf{Right:} The cell hierarchy graph.
-        }
-        \label{fig:hier_netlist}
-    }
-    \vspace{0.5cm}
-    {
-        \centering
-        \includegraphics[valign=c, width=10cm]{figures/netlist_synth/flat_netlist.png}
-        \includegraphics[valign=c, width=4cm]{figures/netlist_synth/flat_graph.png}
-        \captionof{figure}{
-            \textbf{Left:} A flattened netlist consisting of LUTs and FFs.
-            \textbf{Right:} The flattened cell hierarchy graph.
-        }
-        \label{fig:flat_netlist}
-    }
-    \begin{multicols}{2}
-
-    RapidWright offers an easy way to access, read, or manipulate this netlist for a given HDL design. 
-    Shown in Figure~\ref{fig:netlist_syn}
-
-    Verilog \texttt{module}s are represented simply as
-
-    \end{multicols}
-    {
-        \centering
-        \includegraphics[width=\columnwidth]{figures/raw_edif.png}
-        \captionof{figure}{A raw EDIF Netlist with uncountably many cells}
-        \label{fig:raw_edif}
-    }
-    \begin{multicols}{2}
-
-    .
-
-\newpage
-\section{Simulated Annealing}
-    \label{sec:simulated_annealing}
-    With a basic understanding of FPGA architecture, design placement, and RapidWright, we have all the necessary pieces to implement our SA placer. 
-    Here we outline in detail each substage of our implementation: PrePacking, Packing, and Placement. 
-    Shown in Figure \ref{fig:edif_design_device} shows an overview of the placement workflow.
-
-    {
-        \centering
-        \includegraphics[width=0.9\columnwidth]{figures/edif_design_device.png}
-        \captionof{figure}{Our placement workflow}
-        \label{fig:edif_design_device}
-    }
-
-    \end{multicols}
-    {
-        \centering
-        \includegraphics[width=0.8\columnwidth]{figures/substages.png}
-        \captionof{figure}{The data classes populated at each substage: PrepackedDesign, PackedDesign, and PlacedDesign.}
-        \label{fig:substages}
-    }
-    \begin{multicols}{2}
-
-    \subsection{Prepacking}
-        \label{subsec:prepacking}
-
-        The first step in our placement flow is \textbf{prepacking}. 
-        Recall from the 7-Series architecture that there are certain cell structures that must adhere to certain placements constraints to ensure legality, and by design, to minimize wirelength. 
-        The job of the prepacker is to traverse the raw EDIF netlist, detect these cell structures, and consolidate these cells into clusters or groups of clusters that naturally reflect these placement constraints.
-
-        Recall that CARRY4 chains must necessarily be placed vertically and consecutively across a column of SLICEs in ascending order. 
-        Likewise, \textttt{DSP48E1} cascades must necessarily be placed vertically and consecutively across a column of DSP48E1 Sites in ascending order. 
-        A LUT-FF pair may be placed freely, but should be placed in the same lane within the same SLICE to minimize wirelength.
-
-        The raw post-synthesis netlist only tells us the list of nets and the cell ports that they connect to. 
-        It does not report the presence of any macro cell structures (CARRY4 chains, etc.). 
-        Thus, we must traverse the netlist to detect these cell structures and store that structure information in a class we will call \texttt{PrepackedDesign}.
-
-        A group of LUT-FF pairs may be placed in the same SLICE, with the constraint that the FFs must share the same Clock-Enable (CE) and Set-Reset (SR) nets.
-        Clustering LUT-FF pairs like this can reduce the redundancy of having to route the same CE and SR nets to many SLICEs by routing the nets to fewer SLICEs and connecting them to the individual FFs within via intra-Site routing, and at the same time, packing greater logic density over a smaller area of the device.
-        Recall that up to eight FFs may be placed in the same SLICE, thus theoretically, up to eight LUT-FF pairs can be placed in the same SLICE.
-
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/slicel.png}
-            \captionof{figure}{A SLICEL Site}
-            \label{fig:slicel}
-        }
-
-        Utilizing all 8 LUT-FF lanes in a SLICE can help reduce device area utilization and minimize wirelength, but \emph{too much} logic density in an area can contribute to general routing congestion.
-        Furthermore, attempting to fill all 8 lanes in a SLICE requires some careful consideration of LUT BEL constraints. 
-        The LUT BELs in a SLICE, as shown in Figure \ref{fig:slicel}, the LUTs appear to be interleaved.
-
-        Show a block diagram of a LUT6 comprising of two LUT5s.
-
-        Recall that a LUT can accommodate one 6-input boolean function or two 5-input boolean functions sharing the same inputs or any two 3-input or less boolean functions regardless of shared inputs.
-        A LUT6 Cell (a LUT with 6-inputs) will actually occupy two lanes in a SLICEL, rendering one of the FF BELs in either lane ineligible for another LUT-FF pair. 
-        The LUT BELs
-
-    \subsection{Packing}
-        \label{subsec:packing}
-
-    \subsection{Placement}
-        \label{subsec:placement}
-        Up until now we have only organized the logical \texttt{Cells} into \texttt{SiteInsts}. 
-        This is where simulated annealing actually begins.
-
-\section{Placement Results}
-    \begin{itemize}
-    \item Random Site Selection:
-    \item Midpoint Site Selection:
-    \item Hybrid: 
-        Random Site selection until progress stagnates, then use Midpoint Site selection. 
-        Randomly choose betwee Random and Midpoint Site selection each iteration. 
-    \item Cooling Schedule: 
-        Greedy. 
-        Geometric with modifiable alpha and intial temperature. 
-    \end{itemize}
-
-\section{Future Work}
-    \begin{itemize}
-    \item Different cooling schedules: linear cooling, logarithmic cooling, piecewise cooling.
-    \item Support for more macros: buffer cells, FIFO, clock generators, etc..
-    \item Support Ultrascale architecture. 
-    \item Better packing scheme. Currently pseudorandom packing. 
-    \item Inter-Site BEL swapping. 
-    \item BEL-centric placement over Site-centric placement.
-    \item Analytical placement. Global placement, legalization, detailed placement. 
-    \end{itemize}
-
-\section{Conclusion}
-    fdsfdsafdsa
-
-% \section{Analytical Placement}
-%     Introduce placement with legalization methods. \\
-%     Many AP approaches use a Global placement, Legalization, then Detailed placement approach. \\
-%     Our SA only considers legal moves via bookkeeping structures, so it has no concept of global placement or legalization. \\
-%     It is a detailed placement only approach. \\
-%     \begin{itemize}
-%         \item Talk about HeAP as an example.
-%         \item Global Placement, Legalization, then Detailed Placement approach.
-%         \item Global placement using the Bound2Bound net model from SimPL.
-%         \item Legalization Step 1: \\
-%             Find an area of the device that is overutilized (illegal) for twhich the blocks contained within must spread to a larger area. \\
-%             To obtain this overutilized area, adjacent locations (Sites, Tiles, etc.) on the device that are occupied by more than one block (SiteInst, ModuleInst, etc.) are repeatedly clustered together, until all clusters are bordered on all sides by non-overutilized locations.
-\\
-%             Then, the area (?) is expanded in both the x and y dimensions (?) until it is large enough to accommodate all blocks contained within. \\
-%             Specifically, the area is expanded until its "occupance" \(O_A\) divided by its "capacity" \(C_A\) is less than a maximum "utilization factor" $\beta$, where $\beta$ is less than or equal to 1.
-HeAP uses 0.9.
-%         \item Legalization Step 2: \\
-%             Two cuts are generated: a "source cut" and a "target cut". \\
-%             The source cut pertains to the blocks being placed (SiteInsts in home buffer). \\ 
-%             The target cut pertains to the area into which the blocks are placed (away buffer). \\
-%             The source cut splits the blocks into 2 partitions, while the target cut splits the area into 2 sub-areas, into which the blocks in each partition are spread. \\
-%             Two objectives are minimized during this process: The imbalance between the number of blocks in each partition, and the difference in the utilization of each sub-area. \\
-%             Utilization defined as the occupancy divided by the capacity of the sub-area. \\
-%             \(U_{sub-area} = \frac{O_{sub-area}}{C_{sub-area}}\) \\
-%             To generate the source cut, the cells are first sorted by their x or y location, depending on the orientation of the desired cut. \\
-%             Once the cells are sorted, the source cut generation is akin to choosing a pivot in a sorted list, where all blocks to the left of the pivot are assigned to the left/bottom partition and all blocks to the right of the pivot are assigned to the right/top partition.
-\\
-%             The target cut is an x or y cut of the area such that all blocks in each partition fit in their respective sub-areas, and such that \( | U_{sub-area_1} 
-%              - U_{sub-area_2} | \) is minimized.
-%     \end{itemize}
-
-\section{Appendix}
-
-    \begin{figure*}[b]
-        \centering
-        \includegraphics[width=\textwidth]{figures/slicem.png}
-        \caption{SLICEM Site}
-        \label{fig:slicem}
-    \end{figure*}
-
-    \begin{figure*}[t]
-        \centering
-        \includegraphics[width=\textwidth]{figures/slicel.png}
-        \caption{SLICEL Site}
-        \label{fig:slicel}
-    \end{figure*}
-
-\end{multicols}
-
-\end{document}

### 2025-05-03 latex/paper/sections/06-designflow_toolchain.tex
split tex file into multiple files

--- 2025-04-28 latex/paper/paper.tex
@@
-\documentclass{article}
-\usepackage{hyperref}
-\usepackage{enumitem}
-\usepackage{graphicx}
-\usepackage{amsmath}
-\usepackage{mathpazo}
-\usepackage{xcolor}
-\usepackage{float}
-\usepackage{listings}
-\lstset{
-  basicstyle=\ttfamily\small,
-  breaklines=true,
-  columns=fullflexible,
-  frame=single,
-  showspaces=false,
-  showtabs=false,
-  keepspaces=false
-}
-\usepackage[export]{adjustbox}
-\usepackage{multicol}
-\usepackage{caption}
-\usepackage[a4paper,
-            bindingoffset=0.2in,
-            left=0.5in,
-            right=0.5in,
-            top=0.5in,
-            bottom=0.5in,
-            footskip=.25in]{geometry}
-
-\begin{document}
-\title{MS Technical Paper: \\ Placement Algorithms for Heterogeneous FPGAs}
-\author{Brian B Cheng \\ Rutgers University Department of Electrical and Computer Engineering}
-
-\date{}
-\maketitle
-
-\begin{multicols}{2}
-
-\section{Keywords}
-\begin{itemize}
-    \item FPGA, EDA, Placement, Simulated Annealing, Optimization, RapidWright
-\end{itemize}
-
-\section{Abstract}
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\
-
-\section{Introduction}
-
-    Field-Programmable Gate Arrays (FPGAs) have witnessed rapid growth in capacity and versatility, driving significant advances in computer-aided design (CAD) and electronic design automation (EDA) methodologies.
-    Since the early-to-mid 2000s, the stagnation of single-processor performance relative to the rapid increase in integrated circuit sizes has led to a design productivity gap, where the computational effort for designing complex chips continues to rise.
-    FPGA CAD flows mainly encompass synthesis, placement, and routing; all of which are NP-hard problems, of which placement is one of the most time-consuming processes. 
-    Inefficienct placement strategy not only extends design times from hours to days, thereby elevating cost and reducing engineering productivity, but also limits the broader adoption of FPGAs by software engineers who expect compile times akin to those of conventional software compilers like {\tt gcc}.
-
-    For these reasons, FPGA placement remains a critical research effort even today. 
-    In this paper, we study and implement established placement methods. 
-    To do this, we use the RapidWright API, which is a semi-open-source research effort from AMD/Xilinx that enables custom solutions to FPGA design implementations and design tools that are not offered by their industry-standard FPGA environment, Vivado.
-    We implement multiple variations of simulated annealing placers for Xilinx's 7-series FPGAs, with an emphasis on minimizing total wirelength while mitigating runtime. 
-    Our implementation is organized into three consecutive substages. 
-    The \textbf{prepacking} stage involves traversing a raw EDIF netlist to identify recurring cell patterns—such as CARRY chains, DSP cascades, and LUT-FF pairs—that are critical for efficient mapping and legalization.
-    In the subsequent \textbf{packing} stage, these identified patterns, along with any remaining loose cells, are consolidated into SiteInst objects that encapsulate the FPGA’s discrete resource constraints and architectural nuances.
-    Finally, the \textbf{placement} stage employs a simulated annealing (SA) algorithm to optimally assign SiteInst objects to physical sites, aiming to minimize total wirelength while adhering to the constraints of the 7-series architecture.
-
-    Simulated annealing iteratively swaps placement objects guided by a cost function that decides which swaps should be accepted or rejected. 
-    Hill climbing is permitted by occasionally accepting moves that increase cost, in hope that such swaps may later lead to a better final solution. 
-    SA remains a popular approach in FPGA placement research due to its simplicity and robustness in handling the discrete architectural constraints of FPGA devices. 
-    While SA yields surprisingly good results given relatively simple rules, it is ultimately a heuristic and stochastic approach that explores the vast placement space by making random moves. 
-    Most of these moves will be rejected, meaning that SA must run many iterations, usually hundreds to thousands, to arrive at a desirable solution.
-
-    In the ASIC domain, where placers must handle designs with millions of cells, the SA approach has largely been abandoned in favor of analytical techniques, owing to SA's runtime and poor scalability.
-    Modern FPGA placers have also followed suit, as new legalization strategies allow FPGA placers to leverage traditionally ASIC placement algorithms and adapt them to the discrete constraints of FPGA architectures.
-    While this paper does not present a working analytical placer, it will explore ways to build upon our existing infrastructure (prepacker and packer) to replace SA with AP.
-
-    The paper first begins by elaborating on general FPGA architecture and then specifically the Xilinx 7-Series architecture. 
-    Then, the paper will elaborate on the FPGA design flow, then the role that the RapidWright API plays in the design flow. 
-    We explain in detail each of these concepts for a broader audience as they provide much needed context for FPGA placement algorithms as a concept. 
-    However, readers who are already familiar with these concepts can skip directly to the RapidWright API section \ref{sec:rapidwright_api} or to the Simulated Annealing section \ref{sec:simulated_annealing}.
-
-\section{FPGA Architecture History}
-
-    Before any work can begin on an FPGA placer, it is necessary to understand both the objects being placed and the medium in which they are placed.
-    Configurable logic devices have undergone significant evolution over the past four decades. 
-    We will briefly review the evolution of configurable logic architecture starting in the 1970s and quickly work our way up to modern day FPGA architecture.
-
-    \textbf{PLA:} \quad
-    The journey began with the Programmable Logic Array (PLA) in the early 1970s. 
-    The PLA implemented output logic using a programmable-OR and programmable-AND plane that formed a sum-of-products equation for each output through programmable fuses. 
-    Around the same time, the Programmable Array Logic (PAL) was introduced. 
-    The PAL simplified the PLA by fixing the OR gates, resulting in a fixed-OR, programmable-AND design, which sacrificed some logic flexibility to simplify its manufacture. 
-    Figure \ref{fig:pal_2} shows one such PAL architecture.
-
-    \textbf{CPLD:} \quad 
-    Later in the same decade came the Complex Programmable Logic Device (CPLD), which took the form of an array of Configurable Logic Blocks (CLBs). 
-    These CLBs were typically modified PAL blocks that included the PAL itself along with macrocells such as flip-flops, multiplexers, and tri-state buffers. 
-    The CPLD functioned as an array of PALs connected by a central programmable switch matrix and could be programmed using a hardware description language (HDL) like VHDL. 
-    Figure \ref{fig:cpld} shows one such CPLD architecture. 
-    \\
-
-    {
-        \centering
-        \includegraphics[width=\columnwidth]{figures/pal_2.png}
-        \captionof{figure}{PAL architecture with 5 inputs, 8 programmable AND gates and 4 fixed OR gates}
-        \label{fig:pal_2}
-        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/Programmable-Logic-Devices/Programmable-Array-Logic-PAL/
-        % https://www.naukri.com/code360/library/difference-between-pla-and-pal 
-    }
-
-    \newcolumn
-
-    \textbf{Homogeneous FPGA:} \quad 
-    The mid-1980s saw the introduction of homogeneous FPGAs, which were built as a grid of CLBs. 
-    Rather than using a central programmable switch matrix as in CPLDs, FPGAs adopted an island style architecture in which each CLB is surrounded on all sides by programmable routing resources, as shown in Figure \ref{fig:fpga}.
-    The first commercially viable FPGA, produced by Xilinx in 1984, featured 16 CLBs arranged in a 4x4 grid. 
-    As FPGA technology advanced, CLBs were redesigned to use lookup tables (LUTs) instead of PAL arrays for greater logic density. 
-    The capacity of an FPGA was often measured by how many logical elements or CLBs it offered, which grew from hundreds to thousands and now to hundreds of thousands of CLBs.
-    \\
-
-    {
-        \centering
-        \includegraphics[width=0.8\columnwidth]{figures/cpld.png}
-        \captionof{figure}{CPLD architecture with 4 CLBs (PAL-like blocks)}
-        \label{fig:cpld}
-        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/CPLD/Complex-Programmable-Logic-Device-CPLDs/
-    }
-    {
-        \centering
-        \includegraphics[width=\columnwidth]{figures/homogenous_fpga.png}
-        \captionof{figure}{A homogeneous island-style FPGA architecture with 16 CLBs in a grid.}
-        \label{fig:fpga}
-        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/FPGA/Field-programmable-gate-array-FPGA/
-    }
-
-    \newpage
-
-    {
-        \centering
-        \includegraphics[width=\columnwidth]{figures/heterogenous_fpga_3.jpg}
-        \captionof{figure}{A heterogeneous island-style FPGA with a mix of CLBs and macrocells.}
-        \label{fig:fpga}
-        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/FPGA/Field-programmable-gate-array-FPGA/
-    }
-    \vspace{0.5cm}
-
-    \textbf{Heterogenous FPGA:} \quad 
-    This brings us to modern day FPGA architectures. 
-    To meet the needs of increasingly complex designs, FPGA vendors introduced heterogeneous FPGAs. 
-    In these devices, hard macros such as Block RAM (BRAM) and Digital Signal Processing (DSP) slices are integrated into the programmable logic fabric along with CLBs, like shown in Figure \ref{fig:fpga}.
-    This design enables the direct instantiation of common subsystems like memories and multipliers, without having to recreate them from scratch using CLBs. 
-    Major vendors such as Xilinx and Altera now employ heterogeneous island-style architectures in their devices. 
-    As designs become increasingly large and complex, FPGAs meet the demand by becoming increasingly heterogenous, incorporating a wider variety of hard macros into the fabric.
-
-\section{Xilinx 7-Series Architecture}
-    \label{sec:7_series}
-    The Xilinx 7-Series devices, first introduced in 2010, follow a heterogeneous island-style architecture as discussed previously. 
-    Although the 7-Series was later superseded in 2013 by the UltraScale architecture, the 7-Series remains highly relevant due to its accessibility, wide availability, and compatibility with open-source tooling.
-    Representative sub-families include Artix-7, Kintex-7, Virtex-7, and Zynq-7000, each designed with different performance and cost trade-offs but all follow the core 7-Series architecture.
-
-    Figure~\ref{fig:hierarchy_4} illustrates a high-level view of the hierarchical organization of a 7-Series FPGA. 
-    At its lowest level, the device consists of a large array of atomic components called \emph{Basic Elements of Logic}~(\textbf{BELs}). 
-    These BELs encompass look-up tables (\textbf{LUTs}), flip-flops (\textbf{FFs}), block RAMs (\textbf{BRAMs}), DSP slices (\textbf{DSPs}), and the configurable interconnect fabric. 
-    They constitute the fundamental building blocks for implementing digital circuits on the FPGA.
-    Note how the Tile arrangement is columnar, where each column will have only one Tile type.
-
-    To manage this complexity, Xilinx organizes these BELs into incrementally abstract structures. 
-    First, \textbf{BELs} are grouped into \textbf{Sites}. 
-    Each Site is embedded into a \textbf{Tile}, and Tiles are further arranged into \textbf{Clock Regions}. 
-    In some high-density devices, multiple Clock Regions may be consolidated into one or more \textbf{Super Logic Regions}~(SLRs). 
-    However, for the scope of this paper, we focus on Xilinx 7-Series devices with only a single SLR.
-
-    \end{multicols}
-    {
-        \centering
-        \includegraphics[valign=c, width=0.3\columnwidth]{figures/hierarchy_5.png}
-        \includegraphics[valign=c, width=0.6\columnwidth]{figures/hierarchy_4.png}
-        \captionof{figure}{Architecture Hierarchy of a Xilinx FPGA}
-        \label{fig:hierarchy_4}
-    }
-    \vfill
-    \begin{multicols}{2}
-
-    \newpage
-    \subsection{CLB SLICEs}
-
-    In the 7-Series architecture, the term \emph{CLB} (Configurable Logic Block) refers to a \emph{CLB Tile} that contains two \emph{SLICE} Sites. 
-    Xilinx offers two variants of SLICE Sites: \textbf{SLICEL} and \textbf{SLICEM}.
-    \begin{itemize}
-        \item Each \textbf{SLICEL} has a set of BELs including eight LUTs, eight FFs, and one   \emph{CARRY4} adder. The LUT BELs in a SLICEL can only host LUT Cells. 
-        \item The \textbf{SLICEM} includes all the features of a SLICEL but its LUT BELs can host both LUT Cells, which are asynchronous ROM elements, or RAM32M Cells, which are synchronous 32-deep RAM elements.
-These cells are also referred to as \emph{Distributed RAM} in the Xilinx documentation.
-        These cells can offer an alternative to the larger, more dedicated 18K-36K RAMB18E1 cells when RAM resources are highly utilized. 
-    \end{itemize}
-    In a typical 7-Series device, approximately 75\% of the SLICE Sites are SLICELs and 25\% are SLICEMs. 
-    A single CLB Tile can therefore host either two SLICELs or one SLICEL and one SLICEM.
-    For simplicity, however, we will only consider SLICELs for general logic and use the dedicated RAMB18E1 cells to implement RAM elements.
-
-    The BELs in these SLICEs facilitate the bulk of the general programmability of the FPGA fabric. 
-    We will explain in detail the motivation and function of these BELs.
-
-    \textbf{LUTs} \quad
-        Combinational logic is universal to all HDL designs. 
-        As the their name suggests, Look‑Up Tables (LUTs) facilitate combinational logic by acting as tiny asynchronously‑accessed ROMs whose contents are fixed when the FPGA is programmed.  
-        For any boolean function, the synthesizer precalculates the output to every possible input combination and stores the resulting truth table into a LUT's static memory. 
-        The inputs are then essentially treated as an address space that maps to a data value space in an asynchronous ROM. 
-        No explicit logic gates like NAND or XNOR are synthesized, contrary to what newcomers might expect from a "Field Programmable Gate Array".
-
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/lut_synthesis.png}
-            \captionof{figure}{LUT synthesis from user design}
-            \label{fig:lut_synthesis}
-        }
-        \vspace{1.0cm}
-
-        In the 7‑Series devices, one LUT can facilitate any 6-input boolean function, or two 5-input functions, as long as they share the same input signals.  
-        The LUT can also host two independent boolean functions of up to 3 inputs each, even when the inputs are not shared.  
-        Functions requiring more than six unique inputs are decomposed across multiple cascaded LUTs.
-        Figure \ref{fig:lut_synthesis} shows an example of where a LUT is typically synthesized in a design entry.
-
-    \textbf{FFs} \quad
-        FFs are synthesized to facilitate synchronous event-driven signal assignment. 
-        For most Verilog users, this generally means signal assignments wrapped in \texttt{always @(posedge clk)} statements. 
-        Figure \ref{fig:ff_synthesis} shows an example of where a flip flop (FF) is typically synthesized. 
-        The cell primitive \textbf{FDRE} is a type of FF and belongs to a family of D Flip Flops (DFFs) with Clock Enable (CE). 
-        \begin{itemize}
-            \item FDCE - DFF with CE and Asynchronous Clear
-            \item FDPE - DFF with CE and Asynchronous Preset
-            \item FDSE - DFF with CE and Synchronous Set
-            \item \textbf{FDRE} - DFF with CE and Synchronous Reset
-        \end{itemize}
-        In a typical HDL design, the vast majority of FFs will be synthesized as FDREs with the occasional FDSE, as it is generally good practice to keep FPGA designs synchronous. 
-        A FF BEL may also host a LATCH Cell, however, since they are generally bad practice in FPGA design, we will not consider latches in this paper.
-
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/ff_synthesis.png}
-            \captionof{figure}{FF synthesis from user design}
-            \label{fig:ff_synthesis}
-        }
-        \vspace{1.0cm}
-
-        Up to eight (8) FFs can be placed within the same SLICE, but only if they all share a common Clock-Enable (CE) net and a common Set-Reset (SR) net.
-        This is because the SLICE has only one CE pin and one SR pin to interface with general routing. 
-        The CE and SR signals from these pins are broadcast intra-Site to all FFs within.
-
-    \newpage
-    \textbf{LUT-FF Pairs} \quad
-        FPGA designs are very often modelled as a collection of Finite State Machines (FSM) like shown in Figure \ref{fig:fsm}.
-        Many times a design will also use pipelining, either to model signal buffers or shift registers, or to split up large combinational logic blocks into time slices to meet timing constraints. 
-        These common design structures result in many consecutive sections of combinational logic feeding into a vector of registers. 
-        The synthesizer naturally synthesizes these structures as consecutive pairs of LUTs feeding into FFs as shown in Figure \ref{fig:pipelining}. 
-        Figure \ref{fig:lut_ff_pair} shows an example of a synthesized LUT-FF Pair.
-
-        Shown in Figure \ref{fig:intersite_intrasite} are two possible placements for a LUT-FF Pair on the physical device. 
-        On the right, the cells are placed across different Sites, thus the only way to route the net between the cells is through general inter-site routing. 
-        On the bottom left, the cells are placed within the same Site in the same lane, taking advantage of the intra-site routing without burdening the general router with additional inter-site routing.
-        This is an important consideration to make while minimizing wirelength during placement.
-
-        \vspace{1.0cm}
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/fsm.png}
-            \captionof{figure}{Finite state machine (Moore)}
-            \label{fig:fsm}
-        }
-        \vspace{1.0cm}
-
-        {
-            \centering
-            \includegraphics[width=0.9\columnwidth]{figures/pipelining.png}
-            \captionof{figure}{Pipelining synthesized as consecutive LUT-FF pairs}
-            \label{fig:pipelining}
-        }
-
-        \vfill
-        .
-        \newcolumn
-
-        {
-            \centering
-            \includegraphics[width=0.9\columnwidth]{figures/lut_ff_pair.png}
-            \captionof{figure}{A synthesized LUT-FF Pair}
-            \label{fig:lut_ff_pair}
-        }
-        \vspace{1.0cm}
-
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/intersite_intrasite_2.png}
-            \captionof{figure}{Intrasite vs Intersite LUT-FF Placement}
-            \label{fig:intersite_intrasite}
-        }
-
-    \newpage
-    \textbf{CARRY} \quad
-        An FPGA design will also typically implement many adders, counters, subtractors, or comparators, all of which are based on binary addition. 
-        They are so ubiquitous that every that in the 7-Series architecture, every SLICE features a \emph{CARRY4} BEL -- a 4-bit carry-lookahead (CLA) adder, a much better alternative to synthesizing adders via LUTs.
-
-    \textbf{CARRY Chains} \quad
-        These CARRY4 blocks can be chained across SLICEs to implement wide adders efficiently. 
-        The CARRY4 BELs \emph{must} be chained vertically consecutively across SLICEs as the Carry-In (CI) and Carry-Out (CO) pins can only be routed this way. 
-        A CARRY4 cell may also directly connect to LUTs and FFs, and should be placed in the same Site whenever possible to minimize wirelength. 
-        Shown in figure \ref{fig:carry_cell_group} shows how a CARRY4 chain and associated LUTs and FFs can be placed across SLICEs.
-
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/carry_cell_group.png}
-            \captionof{figure}{A SLICEL with a CARRY4 cell, 4 LUT cells, and 4 FF cells placed inside as shown in the Vivado device viewer}
-            \label{fig:carry_cell_group}
-        }
-
-        {
-            \centering
-            \includegraphics[valign=c, width=4.0cm]{figures/carry_chain_placement.png}
-            \includegraphics[valign=c, width=4.0cm]{figures/carry_chain_routes_3.png}
-            \captionof{figure}{
-                A CARRY4 chain of size 3 placed across 3 SLICEs.
-                \textbf{Left:} Simplified view, \textbf{Right:} As shown in the Vivado device viewer.
-            }
-            \label{fig:carry_cell_group}
-        }
-
-        \end{multicols}
-        % \vfill
-        {
-            \centering
-            \includegraphics[valign=c, width=3.0cm]{figures/adder_synthesis.png}
-            \includegraphics[valign=c, width=14.0cm]{figures/carry_chain_edif_3.png}
-            \captionof{figure}{A synthesized carry chain of size 3 as shown in the Vivado netlist viewer}
-            \label{fig:carry_chain_edif}
-        }
-        \begin{multicols}{2}
-        \vspace{1.0cm}
-        .
-
-    \newpage
-    \subsection{DSP Slices}
-
-    \textbf{DSPs} \quad 
-        FPGAs are often used as low latency Digital Signal Processing (DSP) accelerators. 
-        Common DSP subsystems like Finite Impulse Response (FIR) filters, Fast Fourier Transform (FFTs), and convolutional neural nets (CNNs) demand fast large scale multiply-accumulate (MAC) capabilities.
-        The 7-Series arhictecture integrates DSP BELs into the logic fabric called DSP48E1 that can facilitate MAC efficiently. 
-        The architecture hierarchy for DSPs is simple compared to CLBs and SLICEs. 
-        A DSP48E1 Tile contains two DSP48E1 Sites, each containing one DSP48E1 BEL, which can host a DSP48E1 Cell.
-
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/dsp_diagram.png}
-            \captionof{figure}{
-                Basic DSP48E1 Slice Functionality
-            }
-            \label{dsp_diagram}
-        }
-        \vspace{0.5cm}
-
-    \textbf{DSP Cascades} \quad 
-        Higher level DSP functions are supported by cascading DSP48E1 slices in a DSP48E1 column.
-        Much like CARRY4 chains, DSP48E1 cascades must necessarily be placed vertically consecutively across DSP48E1 Sites. 
-        They are connected by three busses: ACOUT to ACIN, BCOUT to BCIN, PCOUT to PCIN.
-        These signal busses run directly between the vertical DSP48E1 slices without burdening the general routing resources. 
-        The ability to cascade this way provides a high-performance and low-power implementation of DSP filter functions.
-
-        \vspace{1.0cm}
-        {
-            \centering
-            \includegraphics[valign=c, width=2.5cm]{figures/dsp_cascade_placement.png}
-            \includegraphics[valign=c, width=3.0cm]{figures/dsp_cascade_routed.png}
-            \captionof{figure}{
-                A DSP48E1 cascade of size 2 placed across 2 DSP48E1 Sites.
-                \textbf{Left:} Simplified view, \textbf{Right:} As shown in the Vivado device viewer.
-            }
-        }
-
-        \end{multicols}
-        {
-            \centering
-            \includegraphics[valign=c, width=4.0cm]{figures/mult_synthesis.png}
-            \includegraphics[valign=c, width=13.0cm]{figures/dsp_synthesis.png}
-            \captionof{figure}{
-                Simple multiplier synthesis from user design.
-            }
-            \label{mult_synthesis}
-        }
-        \begin{multicols}{2}
-
-    \newpage
-
-    \subsection{Block RAM}
-        In addition to SLICEs and DSPs, the 7-Series also offers dedicated Block Random Access Memory (BRAM) BELs. 
-        These BRAMs come in two variants: \textbf{RAMB18E1} and \textbf{RAMB36E1}. 
-        \begin{itemize}
-            \item \textbf{RAMB18E1} - 
-            Has a capacity of 18 Kilobits. 
-            It can be configured as single port RAM with dimensions ranging between (1-bit wide by 16K deep) to (18-bit wide by 1024 deep). 
-            It can also be configured as a (36-bit wide by 512 deep) true simple dual port RAM. 
-            \item \textbf{RAMB36E1} - 
-            Has a capacity of 36 Kilobits. 
-            It can be configured as single port RAM with dimensions ranging between (1-bit wide by 32K deep) to (36-bit wide by 1024 deep). 
-            It can also be configured as a (72-bit wide by 512 deep) simple dual port RAM. 
-        \end{itemize}
-
-        For readability, we will abbreviate RAMB18E1 as \textbf{RAM18} and RAMB36E1 as \textbf{RAM36}.
-        One BRAM Tile contains one RAM36 Site and two RAM18 Sites. 
-        The RAM36 Site contains one RAM36 BEL while the RAM18 Site contains one RAM18 BEL.
-        The RAM18 BEL can host one RAM18 Cell, while the RAM36 BEL can host either one RAM36 Cell or two RAM18 Cells.
-
-        Like DSP cascades, BRAMs may also be cascaded together in a column to implement large memories efficiently, with some intermediate signals between them routed intra-Tile without burdening the general routing.
-        However, unlike DSPs, large memories decomposed amongst multiple BRAMs can also be routed together through general routing. 
-        Furthermore, in most design scenarios, large memories will not utilize the intra-Tile signals.
-
-        For simplicity we will not constrain large BRAMs to be cascaded together in consecutive vertical Tiles. 
-        We can essentially treat RAMB18E1 and RAMB36E1 Cells as loose, minimally constrained single cells, in contrast to the highly constrained DSP cascades and CARRY chains.
-
-        {
-            \centering
-            \includegraphics[valign=c, width=3.0cm]{figures/bram.png}
-            \includegraphics[valign=c, width=2.0cm]{figures/bram_tile.png}
-            \captionof{figure}{
-                A BRAM Tile containing two RAMB18E1 Sites and one RAMB36E1 Site.
-                \textbf{Left:} simplified view. \textbf{Right:} as seen in the device viewer, BELs highlighted in white. 
-            }
-            \label{fig:bram_tile}
-        }
-
-        {
-            \centering
-            \includegraphics[valign=c, width=4.0cm]{figures/bram_synthesis.png}
-            \includegraphics[valign=c, width=\columnwidth]{figures/bram_edif.png}
-            \captionof{figure}{
-                An example of BRAM synthesis (via inference)
-            }
-            \label{fig:bram_synthesis}
-        }
-
-    \subsection{Further Documentation}
-
-    For more in-depth details about 7-Series FPGAs, refer to the official Xilinx user guides such as:
-    \begin{itemize}
-        \item \emph{7 Series FPGAs Overview} (UG476)
-        \item \emph{7 Series FPGA Configurable Logic Block} (UG474)
-        \item \emph{7 Series Memory Resources} (UG473)
-        \item \emph{7 Series DSP48E1 Slice} (UG479)
-    \end{itemize}
-
-    This architectural context provides the necessary background for understanding how a placement algorithm should account for resource constraints and optimize performance in modern FPGA designs.
-
-\newpage
 \section{FPGA Design Flow and Toolchain}
@@
 
-    This workflow underscores the critical role of \textbf{placement} in bridging the netlist to a physical realization. 
-    An efficient placement algorithm can drastically reduce compile times and improve design performance, enabling broader adoption of FPGAs in application spaces that require fast design iteration.
-
-\newpage
-\section{RapidWright API}
-    \label{sec:rapidwright_api}
-
-    \textbf{RapidWright} is an open-source Java framework from AMD/Xilinx that provides direct access to the netlist and device databases used by vendor tools. 
-    This framework positions itself as an additional workflow column, allowing users to intercept or replace stages of the standard design flow with custom optimization stages (see Figure~\ref{fig:vivado_dcps}).
-
-    \begin{itemize}
-    \item \textbf{Design Checkpoints:} 
-        RapidWright leverages \texttt{.dcp} files (design checkpoints) generated at various stages of a Vivado flow. 
-        By importing a checkpoint, engineers can manipulate the netlist, placement, or routing externally, then re-export a modified checkpoint for further processing in the Vivado workflow column.
-
-    \item \textbf{Key Packages:} 
-        RapidWright revolves around three primary data model packages:
-        \begin{enumerate}
-        \item \emph{edif} -- Represents the logical netlist in an abstracted EDIF-like structure.
-        \item \emph{design} -- Contains data structures for the physical implementation (Cells, Nets, Sites, BELs, etc.).
-        \item \emph{device} -- Provides a database of the target FPGA architecture (e.g., Site coordinates, Tile definitions, routing resources).
-        \end{enumerate}
-
-    \item \textbf{Interfacing with the Netlist and Device:} 
-        An engineer can query the netlist to find specific resources (LUTs, FFs, DSPs, etc.) and then map or move them onto device sites. 
-        This level of control over backend resources is necessary for research in custom placement, advanced packing techniques, or experimental routing algorithms.
-    \end{itemize}
-
-    {
-        \centering
-        \includegraphics[width=0.8\columnwidth]{figures/vivado_dcps.png}
-        \captionof{figure}{RapidWright workflow integrating into the default Vivado design flow.}
-        \label{fig:vivado_dcps}
-    }
-
-    By exposing these low-level internals, RapidWright allows fine-grained design transformations that go beyond the standard Vivado IDE’s capabilities. 
-    Researchers can prototype new EDA strategies without needing to re-implement an entire FPGA backend from scratch, thus accelerating innovation in placement and routing methodologies.
-
-\newpage
-\section{What is a Netlist?}
-    \label{sec:netlist}
-    In its simplest form, a netlist consists of a list of every component in an electronic design and a list of nets they connect to. 
-    Here, a net, or network, denotes any group of two or more interconnected elements - whether they be transistors, logic gates, or higher-level macrocells. 
-    More precisely, nets attach to component pins, and those pins reside on cells, each of which exposing multiple ports. 
-    In this way, a netlist makes explicit both the individual building blocks of a circuit and the precise wiring that links them.
-
-    In Vivado, a Netlist can be synthesized as a Hierarchical or a Flattened netlist. 
-    Figure ~\ref{fig:hierarchical_design} shows the Verilog code for a simple HDL design using module instance hierarchy. 
-    Figure ~\ref{fig:hier_netlist} shows the design synthesized into a hierarchical netlist, while Figure ~\ref{fig:flat_netlist} shows the same design synthesized into a flat netlist. 
-    In a synthesized Netlist, the leaf cells (lowest-level cells) are necessarily primitive cells from the vendor primitives library such as LUT6, FDRE, CARRY4, DSP48E1, etc..
-
-    \end{multicols}
-    {
-        \raggedright
-        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/top_level.png}
-        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_0.png}
-        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_1.png}
-        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_2.png}
-        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_3.png}
-        \captionof{figure}{A simple HDL design with module hierarchy.}
-        \label{fig:hierarchical_design}
-    }
-    {
-        \centering
-        \includegraphics[valign=c, width=11.5cm]{figures/netlist_synth/hier_netlist.png}
-        \includegraphics[valign=c, width=6cm]{figures/netlist_synth/hier_graph.png}
-        \captionof{figure}{
-            \textbf{Left:} A hierarchical netlist consisting of LUTs and FFs.
-            \textbf{Right:} The cell hierarchy graph.
-        }
-        \label{fig:hier_netlist}
-    }
-    \vspace{0.5cm}
-    {
-        \centering
-        \includegraphics[valign=c, width=10cm]{figures/netlist_synth/flat_netlist.png}
-        \includegraphics[valign=c, width=4cm]{figures/netlist_synth/flat_graph.png}
-        \captionof{figure}{
-            \textbf{Left:} A flattened netlist consisting of LUTs and FFs.
-            \textbf{Right:} The flattened cell hierarchy graph.
-        }
-        \label{fig:flat_netlist}
-    }
-    \begin{multicols}{2}
-
-    RapidWright offers an easy way to access, read, or manipulate this netlist for a given HDL design. 
-    Shown in Figure~\ref{fig:netlist_syn}
-
-    Verilog \texttt{module}s are represented simply as
-
-    \end{multicols}
-    {
-        \centering
-        \includegraphics[width=\columnwidth]{figures/raw_edif.png}
-        \captionof{figure}{A raw EDIF Netlist with uncountably many cells}
-        \label{fig:raw_edif}
-    }
-    \begin{multicols}{2}
-
-    .
-
-\newpage
-\section{Simulated Annealing}
-    \label{sec:simulated_annealing}
-    With a basic understanding of FPGA architecture, design placement, and RapidWright, we have all the necessary pieces to implement our SA placer. 
-    Here we outline in detail each substage of our implementation: PrePacking, Packing, and Placement. 
-    Shown in Figure \ref{fig:edif_design_device} shows an overview of the placement workflow.
-
-    {
-        \centering
-        \includegraphics[width=0.9\columnwidth]{figures/edif_design_device.png}
-        \captionof{figure}{Our placement workflow}
-        \label{fig:edif_design_device}
-    }
-
-    \end{multicols}
-    {
-        \centering
-        \includegraphics[width=0.8\columnwidth]{figures/substages.png}
-        \captionof{figure}{The data classes populated at each substage: PrepackedDesign, PackedDesign, and PlacedDesign.}
-        \label{fig:substages}
-    }
-    \begin{multicols}{2}
-
-    \subsection{Prepacking}
-        \label{subsec:prepacking}
-
-        The first step in our placement flow is \textbf{prepacking}. 
-        Recall from the 7-Series architecture that there are certain cell structures that must adhere to certain placements constraints to ensure legality, and by design, to minimize wirelength. 
-        The job of the prepacker is to consolidate the cells in the EDIF Netlist into clusters or groups of clusters that naturally reflect these placement constraints.
-
-        CARRY4 chains must necessarily be placed vertically and consecutively across a column of SLICEs in ascending order. 
-        Likewise, DSP48E1 cascades must necessarily be placed vertically and consecutively across a column of DSP48E1 Sites in ascending order. 
-        A LUT-FF pair may be placed freely, but should be placed in the same lane within the same SLICE to minimize wirelength.
-
-        The raw post-synthesis netlist only tells us the list of nets and the cell ports that they connect to. 
-        It does not report the presence of any macro cell structures (CARRY4 chains, etc.). 
-        Thus, we must traverse the netlist to detect these cell structures and store that structure information in a class we will call \texttt{PrepackedDesign}.
-
-        A group of LUT-FF pairs may be placed in the same SLICE, with the constraint that the FFs must share the same Clock-Enable (CE) and Set-Reset (SR) nets.
-        Clustering LUT-FF pairs like this can reduce the redundancy of having to route the same CE and SR nets to many SLICEs by routing the nets to fewer SLICEs and connecting them to the individual FFs within via intra-Site routing, and at the same time, packing greater logic density over a smaller area of the device.
-        Recall that up to eight FFs may be placed in the same SLICE, thus theoretically, up to eight LUT-FF pairs can be placed in the same SLICE.
-
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/slicel.png}
-            \captionof{figure}{A SLICEL Site}
-            \label{fig:slicel}
-        }
-
-        Utilizing all 8 LUT-FF lanes in a SLICE can help reduce device area utilization and minimize wirelength, but \emph{too much} logic density in an area can contribute to general routing congestion.
-        Furthermore, attempting to fill all 8 lanes in a SLICE requires some careful consideration of LUT BEL constraints. 
-        The LUT BELs in a SLICE, as shown in Figure \ref{fig:slicel}, the LUTs appear to be interleaved.
-
-        Show a block diagram of a LUT6 comprising of two LUT5s.
-
-        Recall that a LUT can accommodate one 6-input boolean function or two 5-input boolean functions sharing the same inputs or any two 3-input or less boolean functions regardless of shared inputs.
-        A LUT6 Cell (a LUT with 6-inputs) will actually occupy two lanes in a SLICEL, rendering one of the FF BELs in either lane ineligible for another LUT-FF pair. 
-        The LUT BELs
-
-    \subsection{Packing}
-        \label{subsec:packing}
-
-    \subsection{Placement}
-        \label{subsec:placement}
-        Up until now we have only organized the logical \texttt{Cells} into \texttt{SiteInsts}. 
-        This is where simulated annealing actually begins.
-
-\section{Placement Results}
-    \begin{itemize}
-    \item Random Site Selection:
-    \item Midpoint Site Selection:
-    \item Hybrid: 
-        Random Site selection until progress stagnates, then use Midpoint Site selection. 
-        Randomly choose betwee Random and Midpoint Site selection each iteration. 
-    \item Cooling Schedule: 
-        Greedy. 
-        Geometric with modifiable alpha and intial temperature. 
-    \end{itemize}
-
-\section{Future Work}
-    \begin{itemize}
-    \item Different cooling schedules: linear cooling, logarithmic cooling, piecewise cooling.
-    \item Support for more macros: buffer cells, FIFO, clock generators, etc..
-    \item Support Ultrascale architecture. 
-    \item Better packing scheme. Currently pseudorandom packing. 
-    \item Inter-Site BEL swapping. 
-    \item BEL-centric placement over Site-centric placement.
-    \item Analytical placement. Global placement, legalization, detailed placement. 
-    \end{itemize}
-
-\section{Conclusion}
-    fdsfdsafdsa
-
-% \section{Analytical Placement}
-%     Introduce placement with legalization methods. \\
-%     Many AP approaches use a Global placement, Legalization, then Detailed placement approach. \\
-%     Our SA only considers legal moves via bookkeeping structures, so it has no concept of global placement or legalization. \\
-%     It is a detailed placement only approach. \\
-%     \begin{itemize}
-%         \item Talk about HeAP as an example.
-%         \item Global Placement, Legalization, then Detailed Placement approach.
-%         \item Global placement using the Bound2Bound net model from SimPL.
-%         \item Legalization Step 1: \\
-%             Find an area of the device that is overutilized (illegal) for twhich the blocks contained within must spread to a larger area. \\
-%             To obtain this overutilized area, adjacent locations (Sites, Tiles, etc.) on the device that are occupied by more than one block (SiteInst, ModuleInst, etc.) are repeatedly clustered together, until all clusters are bordered on all sides by non-overutilized locations.
-\\
-%             Then, the area (?) is expanded in both the x and y dimensions (?) until it is large enough to accommodate all blocks contained within. \\
-%             Specifically, the area is expanded until its "occupance" \(O_A\) divided by its "capacity" \(C_A\) is less than a maximum "utilization factor" $\beta$, where $\beta$ is less than or equal to 1.
-HeAP uses 0.9.
-%         \item Legalization Step 2: \\
-%             Two cuts are generated: a "source cut" and a "target cut". \\
-%             The source cut pertains to the blocks being placed (SiteInsts in home buffer). \\ 
-%             The target cut pertains to the area into which the blocks are placed (away buffer). \\
-%             The source cut splits the blocks into 2 partitions, while the target cut splits the area into 2 sub-areas, into which the blocks in each partition are spread. \\
-%             Two objectives are minimized during this process: The imbalance between the number of blocks in each partition, and the difference in the utilization of each sub-area. \\
-%             Utilization defined as the occupancy divided by the capacity of the sub-area. \\
-%             \(U_{sub-area} = \frac{O_{sub-area}}{C_{sub-area}}\) \\
-%             To generate the source cut, the cells are first sorted by their x or y location, depending on the orientation of the desired cut. \\
-%             Once the cells are sorted, the source cut generation is akin to choosing a pivot in a sorted list, where all blocks to the left of the pivot are assigned to the left/bottom partition and all blocks to the right of the pivot are assigned to the right/top partition.
-\\
-%             The target cut is an x or y cut of the area such that all blocks in each partition fit in their respective sub-areas, and such that \( | U_{sub-area_1} 
-%              - U_{sub-area_2} | \) is minimized.
-%     \end{itemize}
-
-\section{Appendix}
-
-    \begin{figure*}[b]
-        \centering
-        \includegraphics[width=\textwidth]{figures/slicem.png}
-        \caption{SLICEM Site}
-        \label{fig:slicem}
-    \end{figure*}
-
-    \begin{figure*}[t]
-        \centering
-        \includegraphics[width=\textwidth]{figures/slicel.png}
-        \caption{SLICEL Site}
-        \label{fig:slicel}
-    \end{figure*}
-
-\end{multicols}
-
-\end{document}
+% This workflow underscores the critical role of \textbf{placement} in bridging the netlist to a physical realization. 
+% An efficient placement algorithm can drastically reduce compile times and improve design performance, enabling broader adoption of FPGAs in application spaces that require fast design iteration.

### 2025-05-03 latex/paper/sections/07-rapidwright_api.tex
split tex file into multiple files

--- 2025-04-28 latex/paper/paper.tex
@@
-\documentclass{article}
-\usepackage{hyperref}
-\usepackage{enumitem}
-\usepackage{graphicx}
-\usepackage{amsmath}
-\usepackage{mathpazo}
-\usepackage{xcolor}
-\usepackage{float}
-\usepackage{listings}
-\lstset{
-  basicstyle=\ttfamily\small,
-  breaklines=true,
-  columns=fullflexible,
-  frame=single,
-  showspaces=false,
-  showtabs=false,
-  keepspaces=false
-}
-\usepackage[export]{adjustbox}
-\usepackage{multicol}
-\usepackage{caption}
-\usepackage[a4paper,
-            bindingoffset=0.2in,
-            left=0.5in,
-            right=0.5in,
-            top=0.5in,
-            bottom=0.5in,
-            footskip=.25in]{geometry}
-
-\begin{document}
-\title{MS Technical Paper: \\ Placement Algorithms for Heterogeneous FPGAs}
-\author{Brian B Cheng \\ Rutgers University Department of Electrical and Computer Engineering}
-
-\date{}
-\maketitle
-
-\begin{multicols}{2}
-
-\section{Keywords}
-\begin{itemize}
-    \item FPGA, EDA, Placement, Simulated Annealing, Optimization, RapidWright
-\end{itemize}
-
-\section{Abstract}
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\
-
-\section{Introduction}
-
-    Field-Programmable Gate Arrays (FPGAs) have witnessed rapid growth in capacity and versatility, driving significant advances in computer-aided design (CAD) and electronic design automation (EDA) methodologies.
-    Since the early-to-mid 2000s, the stagnation of single-processor performance relative to the rapid increase in integrated circuit sizes has led to a design productivity gap, where the computational effort for designing complex chips continues to rise.
-    FPGA CAD flows mainly encompass synthesis, placement, and routing; all of which are NP-hard problems, of which placement is one of the most time-consuming processes. 
-    Inefficienct placement strategy not only extends design times from hours to days, thereby elevating cost and reducing engineering productivity, but also limits the broader adoption of FPGAs by software engineers who expect compile times akin to those of conventional software compilers like {\tt gcc}.
-
-    For these reasons, FPGA placement remains a critical research effort even today. 
-    In this paper, we study and implement established placement methods. 
-    To do this, we use the RapidWright API, which is a semi-open-source research effort from AMD/Xilinx that enables custom solutions to FPGA design implementations and design tools that are not offered by their industry-standard FPGA environment, Vivado.
-    We implement multiple variations of simulated annealing placers for Xilinx's 7-series FPGAs, with an emphasis on minimizing total wirelength while mitigating runtime. 
-    Our implementation is organized into three consecutive substages. 
-    The \textbf{prepacking} stage involves traversing a raw EDIF netlist to identify recurring cell patterns—such as CARRY chains, DSP cascades, and LUT-FF pairs—that are critical for efficient mapping and legalization.
-    In the subsequent \textbf{packing} stage, these identified patterns, along with any remaining loose cells, are consolidated into SiteInst objects that encapsulate the FPGA’s discrete resource constraints and architectural nuances.
-    Finally, the \textbf{placement} stage employs a simulated annealing (SA) algorithm to optimally assign SiteInst objects to physical sites, aiming to minimize total wirelength while adhering to the constraints of the 7-series architecture.
-
-    Simulated annealing iteratively swaps placement objects guided by a cost function that decides which swaps should be accepted or rejected. 
-    Hill climbing is permitted by occasionally accepting moves that increase cost, in hope that such swaps may later lead to a better final solution. 
-    SA remains a popular approach in FPGA placement research due to its simplicity and robustness in handling the discrete architectural constraints of FPGA devices. 
-    While SA yields surprisingly good results given relatively simple rules, it is ultimately a heuristic and stochastic approach that explores the vast placement space by making random moves. 
-    Most of these moves will be rejected, meaning that SA must run many iterations, usually hundreds to thousands, to arrive at a desirable solution.
-
-    In the ASIC domain, where placers must handle designs with millions of cells, the SA approach has largely been abandoned in favor of analytical techniques, owing to SA's runtime and poor scalability.
-    Modern FPGA placers have also followed suit, as new legalization strategies allow FPGA placers to leverage traditionally ASIC placement algorithms and adapt them to the discrete constraints of FPGA architectures.
-    While this paper does not present a working analytical placer, it will explore ways to build upon our existing infrastructure (prepacker and packer) to replace SA with AP.
-
-    The paper first begins by elaborating on general FPGA architecture and then specifically the Xilinx 7-Series architecture. 
-    Then, the paper will elaborate on the FPGA design flow, then the role that the RapidWright API plays in the design flow. 
-    We explain in detail each of these concepts for a broader audience as they provide much needed context for FPGA placement algorithms as a concept. 
-    However, readers who are already familiar with these concepts can skip directly to the RapidWright API section \ref{sec:rapidwright_api} or to the Simulated Annealing section \ref{sec:simulated_annealing}.
-
-\section{FPGA Architecture History}
-
-    Before any work can begin on an FPGA placer, it is necessary to understand both the objects being placed and the medium in which they are placed.
-    Configurable logic devices have undergone significant evolution over the past four decades. 
-    We will briefly review the evolution of configurable logic architecture starting in the 1970s and quickly work our way up to modern day FPGA architecture.
-
-    \textbf{PLA:} \quad
-    The journey began with the Programmable Logic Array (PLA) in the early 1970s. 
-    The PLA implemented output logic using a programmable-OR and programmable-AND plane that formed a sum-of-products equation for each output through programmable fuses. 
-    Around the same time, the Programmable Array Logic (PAL) was introduced. 
-    The PAL simplified the PLA by fixing the OR gates, resulting in a fixed-OR, programmable-AND design, which sacrificed some logic flexibility to simplify its manufacture. 
-    Figure \ref{fig:pal_2} shows one such PAL architecture.
-
-    \textbf{CPLD:} \quad 
-    Later in the same decade came the Complex Programmable Logic Device (CPLD), which took the form of an array of Configurable Logic Blocks (CLBs). 
-    These CLBs were typically modified PAL blocks that included the PAL itself along with macrocells such as flip-flops, multiplexers, and tri-state buffers. 
-    The CPLD functioned as an array of PALs connected by a central programmable switch matrix and could be programmed using a hardware description language (HDL) like VHDL. 
-    Figure \ref{fig:cpld} shows one such CPLD architecture. 
-    \\
-
-    {
-        \centering
-        \includegraphics[width=\columnwidth]{figures/pal_2.png}
-        \captionof{figure}{PAL architecture with 5 inputs, 8 programmable AND gates and 4 fixed OR gates}
-        \label{fig:pal_2}
-        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/Programmable-Logic-Devices/Programmable-Array-Logic-PAL/
-        % https://www.naukri.com/code360/library/difference-between-pla-and-pal 
-    }
-
-    \newcolumn
-
-    \textbf{Homogeneous FPGA:} \quad 
-    The mid-1980s saw the introduction of homogeneous FPGAs, which were built as a grid of CLBs. 
-    Rather than using a central programmable switch matrix as in CPLDs, FPGAs adopted an island style architecture in which each CLB is surrounded on all sides by programmable routing resources, as shown in Figure \ref{fig:fpga}.
-    The first commercially viable FPGA, produced by Xilinx in 1984, featured 16 CLBs arranged in a 4x4 grid. 
-    As FPGA technology advanced, CLBs were redesigned to use lookup tables (LUTs) instead of PAL arrays for greater logic density. 
-    The capacity of an FPGA was often measured by how many logical elements or CLBs it offered, which grew from hundreds to thousands and now to hundreds of thousands of CLBs.
-    \\
-
-    {
-        \centering
-        \includegraphics[width=0.8\columnwidth]{figures/cpld.png}
-        \captionof{figure}{CPLD architecture with 4 CLBs (PAL-like blocks)}
-        \label{fig:cpld}
-        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/CPLD/Complex-Programmable-Logic-Device-CPLDs/
-    }
-    {
-        \centering
-        \includegraphics[width=\columnwidth]{figures/homogenous_fpga.png}
-        \captionof{figure}{A homogeneous island-style FPGA architecture with 16 CLBs in a grid.}
-        \label{fig:fpga}
-        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/FPGA/Field-programmable-gate-array-FPGA/
-    }
-
-    \newpage
-
-    {
-        \centering
-        \includegraphics[width=\columnwidth]{figures/heterogenous_fpga_3.jpg}
-        \captionof{figure}{A heterogeneous island-style FPGA with a mix of CLBs and macrocells.}
-        \label{fig:fpga}
-        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/FPGA/Field-programmable-gate-array-FPGA/
-    }
-    \vspace{0.5cm}
-
-    \textbf{Heterogenous FPGA:} \quad 
-    This brings us to modern day FPGA architectures. 
-    To meet the needs of increasingly complex designs, FPGA vendors introduced heterogeneous FPGAs. 
-    In these devices, hard macros such as Block RAM (BRAM) and Digital Signal Processing (DSP) slices are integrated into the programmable logic fabric along with CLBs, like shown in Figure \ref{fig:fpga}.
-    This design enables the direct instantiation of common subsystems like memories and multipliers, without having to recreate them from scratch using CLBs. 
-    Major vendors such as Xilinx and Altera now employ heterogeneous island-style architectures in their devices. 
-    As designs become increasingly large and complex, FPGAs meet the demand by becoming increasingly heterogenous, incorporating a wider variety of hard macros into the fabric.
-
-\section{Xilinx 7-Series Architecture}
-    \label{sec:7_series}
-    The Xilinx 7-Series devices, first introduced in 2010, follow a heterogeneous island-style architecture as discussed previously. 
-    Although the 7-Series was later superseded in 2013 by the UltraScale architecture, the 7-Series remains highly relevant due to its accessibility, wide availability, and compatibility with open-source tooling.
-    Representative sub-families include Artix-7, Kintex-7, Virtex-7, and Zynq-7000, each designed with different performance and cost trade-offs but all follow the core 7-Series architecture.
-
-    Figure~\ref{fig:hierarchy_4} illustrates a high-level view of the hierarchical organization of a 7-Series FPGA. 
-    At its lowest level, the device consists of a large array of atomic components called \emph{Basic Elements of Logic}~(\textbf{BELs}). 
-    These BELs encompass look-up tables (\textbf{LUTs}), flip-flops (\textbf{FFs}), block RAMs (\textbf{BRAMs}), DSP slices (\textbf{DSPs}), and the configurable interconnect fabric. 
-    They constitute the fundamental building blocks for implementing digital circuits on the FPGA.
-    Note how the Tile arrangement is columnar, where each column will have only one Tile type.
-
-    To manage this complexity, Xilinx organizes these BELs into incrementally abstract structures. 
-    First, \textbf{BELs} are grouped into \textbf{Sites}. 
-    Each Site is embedded into a \textbf{Tile}, and Tiles are further arranged into \textbf{Clock Regions}. 
-    In some high-density devices, multiple Clock Regions may be consolidated into one or more \textbf{Super Logic Regions}~(SLRs). 
-    However, for the scope of this paper, we focus on Xilinx 7-Series devices with only a single SLR.
-
-    \end{multicols}
-    {
-        \centering
-        \includegraphics[valign=c, width=0.3\columnwidth]{figures/hierarchy_5.png}
-        \includegraphics[valign=c, width=0.6\columnwidth]{figures/hierarchy_4.png}
-        \captionof{figure}{Architecture Hierarchy of a Xilinx FPGA}
-        \label{fig:hierarchy_4}
-    }
-    \vfill
-    \begin{multicols}{2}
-
-    \newpage
-    \subsection{CLB SLICEs}
-
-    In the 7-Series architecture, the term \emph{CLB} (Configurable Logic Block) refers to a \emph{CLB Tile} that contains two \emph{SLICE} Sites. 
-    Xilinx offers two variants of SLICE Sites: \textbf{SLICEL} and \textbf{SLICEM}.
-    \begin{itemize}
-        \item Each \textbf{SLICEL} has a set of BELs including eight LUTs, eight FFs, and one   \emph{CARRY4} adder. The LUT BELs in a SLICEL can only host LUT Cells. 
-        \item The \textbf{SLICEM} includes all the features of a SLICEL but its LUT BELs can host both LUT Cells, which are asynchronous ROM elements, or RAM32M Cells, which are synchronous 32-deep RAM elements.
-These cells are also referred to as \emph{Distributed RAM} in the Xilinx documentation.
-        These cells can offer an alternative to the larger, more dedicated 18K-36K RAMB18E1 cells when RAM resources are highly utilized. 
-    \end{itemize}
-    In a typical 7-Series device, approximately 75\% of the SLICE Sites are SLICELs and 25\% are SLICEMs. 
-    A single CLB Tile can therefore host either two SLICELs or one SLICEL and one SLICEM.
-    For simplicity, however, we will only consider SLICELs for general logic and use the dedicated RAMB18E1 cells to implement RAM elements.
-
-    The BELs in these SLICEs facilitate the bulk of the general programmability of the FPGA fabric. 
-    We will explain in detail the motivation and function of these BELs.
-
-    \textbf{LUTs} \quad
-        Combinational logic is universal to all HDL designs. 
-        As the their name suggests, Look‑Up Tables (LUTs) facilitate combinational logic by acting as tiny asynchronously‑accessed ROMs whose contents are fixed when the FPGA is programmed.  
-        For any boolean function, the synthesizer precalculates the output to every possible input combination and stores the resulting truth table into a LUT's static memory. 
-        The inputs are then essentially treated as an address space that maps to a data value space in an asynchronous ROM. 
-        No explicit logic gates like NAND or XNOR are synthesized, contrary to what newcomers might expect from a "Field Programmable Gate Array".
-
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/lut_synthesis.png}
-            \captionof{figure}{LUT synthesis from user design}
-            \label{fig:lut_synthesis}
-        }
-        \vspace{1.0cm}
-
-        In the 7‑Series devices, one LUT can facilitate any 6-input boolean function, or two 5-input functions, as long as they share the same input signals.  
-        The LUT can also host two independent boolean functions of up to 3 inputs each, even when the inputs are not shared.  
-        Functions requiring more than six unique inputs are decomposed across multiple cascaded LUTs.
-        Figure \ref{fig:lut_synthesis} shows an example of where a LUT is typically synthesized in a design entry.
-
-    \textbf{FFs} \quad
-        FFs are synthesized to facilitate synchronous event-driven signal assignment. 
-        For most Verilog users, this generally means signal assignments wrapped in \texttt{always @(posedge clk)} statements. 
-        Figure \ref{fig:ff_synthesis} shows an example of where a flip flop (FF) is typically synthesized. 
-        The cell primitive \textbf{FDRE} is a type of FF and belongs to a family of D Flip Flops (DFFs) with Clock Enable (CE). 
-        \begin{itemize}
-            \item FDCE - DFF with CE and Asynchronous Clear
-            \item FDPE - DFF with CE and Asynchronous Preset
-            \item FDSE - DFF with CE and Synchronous Set
-            \item \textbf{FDRE} - DFF with CE and Synchronous Reset
-        \end{itemize}
-        In a typical HDL design, the vast majority of FFs will be synthesized as FDREs with the occasional FDSE, as it is generally good practice to keep FPGA designs synchronous. 
-        A FF BEL may also host a LATCH Cell, however, since they are generally bad practice in FPGA design, we will not consider latches in this paper.
-
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/ff_synthesis.png}
-            \captionof{figure}{FF synthesis from user design}
-            \label{fig:ff_synthesis}
-        }
-        \vspace{1.0cm}
-
-        Up to eight (8) FFs can be placed within the same SLICE, but only if they all share a common Clock-Enable (CE) net and a common Set-Reset (SR) net.
-        This is because the SLICE has only one CE pin and one SR pin to interface with general routing. 
-        The CE and SR signals from these pins are broadcast intra-Site to all FFs within.
-
-    \newpage
-    \textbf{LUT-FF Pairs} \quad
-        FPGA designs are very often modelled as a collection of Finite State Machines (FSM) like shown in Figure \ref{fig:fsm}.
-        Many times a design will also use pipelining, either to model signal buffers or shift registers, or to split up large combinational logic blocks into time slices to meet timing constraints. 
-        These common design structures result in many consecutive sections of combinational logic feeding into a vector of registers. 
-        The synthesizer naturally synthesizes these structures as consecutive pairs of LUTs feeding into FFs as shown in Figure \ref{fig:pipelining}. 
-        Figure \ref{fig:lut_ff_pair} shows an example of a synthesized LUT-FF Pair.
-
-        Shown in Figure \ref{fig:intersite_intrasite} are two possible placements for a LUT-FF Pair on the physical device. 
-        On the right, the cells are placed across different Sites, thus the only way to route the net between the cells is through general inter-site routing. 
-        On the bottom left, the cells are placed within the same Site in the same lane, taking advantage of the intra-site routing without burdening the general router with additional inter-site routing.
-        This is an important consideration to make while minimizing wirelength during placement.
-
-        \vspace{1.0cm}
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/fsm.png}
-            \captionof{figure}{Finite state machine (Moore)}
-            \label{fig:fsm}
-        }
-        \vspace{1.0cm}
-
-        {
-            \centering
-            \includegraphics[width=0.9\columnwidth]{figures/pipelining.png}
-            \captionof{figure}{Pipelining synthesized as consecutive LUT-FF pairs}
-            \label{fig:pipelining}
-        }
-
-        \vfill
-        .
-        \newcolumn
-
-        {
-            \centering
-            \includegraphics[width=0.9\columnwidth]{figures/lut_ff_pair.png}
-            \captionof{figure}{A synthesized LUT-FF Pair}
-            \label{fig:lut_ff_pair}
-        }
-        \vspace{1.0cm}
-
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/intersite_intrasite_2.png}
-            \captionof{figure}{Intrasite vs Intersite LUT-FF Placement}
-            \label{fig:intersite_intrasite}
-        }
-
-    \newpage
-    \textbf{CARRY} \quad
-        An FPGA design will also typically implement many adders, counters, subtractors, or comparators, all of which are based on binary addition. 
-        They are so ubiquitous that every that in the 7-Series architecture, every SLICE features a \emph{CARRY4} BEL -- a 4-bit carry-lookahead (CLA) adder, a much better alternative to synthesizing adders via LUTs.
-
-    \textbf{CARRY Chains} \quad
-        These CARRY4 blocks can be chained across SLICEs to implement wide adders efficiently. 
-        The CARRY4 BELs \emph{must} be chained vertically consecutively across SLICEs as the Carry-In (CI) and Carry-Out (CO) pins can only be routed this way. 
-        A CARRY4 cell may also directly connect to LUTs and FFs, and should be placed in the same Site whenever possible to minimize wirelength. 
-        Shown in figure \ref{fig:carry_cell_group} shows how a CARRY4 chain and associated LUTs and FFs can be placed across SLICEs.
-
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/carry_cell_group.png}
-            \captionof{figure}{A SLICEL with a CARRY4 cell, 4 LUT cells, and 4 FF cells placed inside as shown in the Vivado device viewer}
-            \label{fig:carry_cell_group}
-        }
-
-        {
-            \centering
-            \includegraphics[valign=c, width=4.0cm]{figures/carry_chain_placement.png}
-            \includegraphics[valign=c, width=4.0cm]{figures/carry_chain_routes_3.png}
-            \captionof{figure}{
-                A CARRY4 chain of size 3 placed across 3 SLICEs.
-                \textbf{Left:} Simplified view, \textbf{Right:} As shown in the Vivado device viewer.
-            }
-            \label{fig:carry_cell_group}
-        }
-
-        \end{multicols}
-        % \vfill
-        {
-            \centering
-            \includegraphics[valign=c, width=3.0cm]{figures/adder_synthesis.png}
-            \includegraphics[valign=c, width=14.0cm]{figures/carry_chain_edif_3.png}
-            \captionof{figure}{A synthesized carry chain of size 3 as shown in the Vivado netlist viewer}
-            \label{fig:carry_chain_edif}
-        }
-        \begin{multicols}{2}
-        \vspace{1.0cm}
-        .
-
-    \newpage
-    \subsection{DSP Slices}
-
-    \textbf{DSPs} \quad 
-        FPGAs are often used as low latency Digital Signal Processing (DSP) accelerators. 
-        Common DSP subsystems like Finite Impulse Response (FIR) filters, Fast Fourier Transform (FFTs), and convolutional neural nets (CNNs) demand fast large scale multiply-accumulate (MAC) capabilities.
-        The 7-Series arhictecture integrates DSP BELs into the logic fabric called DSP48E1 that can facilitate MAC efficiently. 
-        The architecture hierarchy for DSPs is simple compared to CLBs and SLICEs. 
-        A DSP48E1 Tile contains two DSP48E1 Sites, each containing one DSP48E1 BEL, which can host a DSP48E1 Cell.
-
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/dsp_diagram.png}
-            \captionof{figure}{
-                Basic DSP48E1 Slice Functionality
-            }
-            \label{dsp_diagram}
-        }
-        \vspace{0.5cm}
-
-    \textbf{DSP Cascades} \quad 
-        Higher level DSP functions are supported by cascading DSP48E1 slices in a DSP48E1 column.
-        Much like CARRY4 chains, DSP48E1 cascades must necessarily be placed vertically consecutively across DSP48E1 Sites. 
-        They are connected by three busses: ACOUT to ACIN, BCOUT to BCIN, PCOUT to PCIN.
-        These signal busses run directly between the vertical DSP48E1 slices without burdening the general routing resources. 
-        The ability to cascade this way provides a high-performance and low-power implementation of DSP filter functions.
-
-        \vspace{1.0cm}
-        {
-            \centering
-            \includegraphics[valign=c, width=2.5cm]{figures/dsp_cascade_placement.png}
-            \includegraphics[valign=c, width=3.0cm]{figures/dsp_cascade_routed.png}
-            \captionof{figure}{
-                A DSP48E1 cascade of size 2 placed across 2 DSP48E1 Sites.
-                \textbf{Left:} Simplified view, \textbf{Right:} As shown in the Vivado device viewer.
-            }
-        }
-
-        \end{multicols}
-        {
-            \centering
-            \includegraphics[valign=c, width=4.0cm]{figures/mult_synthesis.png}
-            \includegraphics[valign=c, width=13.0cm]{figures/dsp_synthesis.png}
-            \captionof{figure}{
-                Simple multiplier synthesis from user design.
-            }
-            \label{mult_synthesis}
-        }
-        \begin{multicols}{2}
-
-    \newpage
-
-    \subsection{Block RAM}
-        In addition to SLICEs and DSPs, the 7-Series also offers dedicated Block Random Access Memory (BRAM) BELs. 
-        These BRAMs come in two variants: \textbf{RAMB18E1} and \textbf{RAMB36E1}. 
-        \begin{itemize}
-            \item \textbf{RAMB18E1} - 
-            Has a capacity of 18 Kilobits. 
-            It can be configured as single port RAM with dimensions ranging between (1-bit wide by 16K deep) to (18-bit wide by 1024 deep). 
-            It can also be configured as a (36-bit wide by 512 deep) true simple dual port RAM. 
-            \item \textbf{RAMB36E1} - 
-            Has a capacity of 36 Kilobits. 
-            It can be configured as single port RAM with dimensions ranging between (1-bit wide by 32K deep) to (36-bit wide by 1024 deep). 
-            It can also be configured as a (72-bit wide by 512 deep) simple dual port RAM. 
-        \end{itemize}
-
-        For readability, we will abbreviate RAMB18E1 as \textbf{RAM18} and RAMB36E1 as \textbf{RAM36}.
-        One BRAM Tile contains one RAM36 Site and two RAM18 Sites. 
-        The RAM36 Site contains one RAM36 BEL while the RAM18 Site contains one RAM18 BEL.
-        The RAM18 BEL can host one RAM18 Cell, while the RAM36 BEL can host either one RAM36 Cell or two RAM18 Cells.
-
-        Like DSP cascades, BRAMs may also be cascaded together in a column to implement large memories efficiently, with some intermediate signals between them routed intra-Tile without burdening the general routing.
-        However, unlike DSPs, large memories decomposed amongst multiple BRAMs can also be routed together through general routing. 
-        Furthermore, in most design scenarios, large memories will not utilize the intra-Tile signals.
-
-        For simplicity we will not constrain large BRAMs to be cascaded together in consecutive vertical Tiles. 
-        We can essentially treat RAMB18E1 and RAMB36E1 Cells as loose, minimally constrained single cells, in contrast to the highly constrained DSP cascades and CARRY chains.
-
-        {
-            \centering
-            \includegraphics[valign=c, width=3.0cm]{figures/bram.png}
-            \includegraphics[valign=c, width=2.0cm]{figures/bram_tile.png}
-            \captionof{figure}{
-                A BRAM Tile containing two RAMB18E1 Sites and one RAMB36E1 Site.
-                \textbf{Left:} simplified view. \textbf{Right:} as seen in the device viewer, BELs highlighted in white. 
-            }
-            \label{fig:bram_tile}
-        }
-
-        {
-            \centering
-            \includegraphics[valign=c, width=4.0cm]{figures/bram_synthesis.png}
-            \includegraphics[valign=c, width=\columnwidth]{figures/bram_edif.png}
-            \captionof{figure}{
-                An example of BRAM synthesis (via inference)
-            }
-            \label{fig:bram_synthesis}
-        }
-
-    \subsection{Further Documentation}
-
-    For more in-depth details about 7-Series FPGAs, refer to the official Xilinx user guides such as:
-    \begin{itemize}
-        \item \emph{7 Series FPGAs Overview} (UG476)
-        \item \emph{7 Series FPGA Configurable Logic Block} (UG474)
-        \item \emph{7 Series Memory Resources} (UG473)
-        \item \emph{7 Series DSP48E1 Slice} (UG479)
-    \end{itemize}
-
-    This architectural context provides the necessary background for understanding how a placement algorithm should account for resource constraints and optimize performance in modern FPGA designs.
-
-\newpage
-\section{FPGA Design Flow and Toolchain}
-    \label{sec:fpga_flow_toolchain}
-
-    Modern FPGA designs require a sophisticated toolchain to bridge the gap between high-level hardware descriptions and the final bitstream used to configure the FPGA. 
-    Figure~\ref{fig:design_flow} illustrates a representative process that converts an abstract Hardware Description Language (HDL) design into a verified configuration file for a target device.
-
-    {
-        \centering
-        \includegraphics[width=0.9\columnwidth]{figures/design_flow.png}
-        \captionof{figure}{A typical FPGA design and verification workflow.}
-        \label{fig:design_flow}
-    }
-
-    \begin{enumerate}
-        \item \textbf{Design Entry:} 
-            An engineer describes the intended functionality of the digital system using a hardware description language (HDL) such as Verilog or VHDL. 
-            During this phase, the coding style can vary (behavioral, structural, dataflow, etc.), but it always aims to capture high-level behavior rather than device-specific details.
-
-        \item \textbf{Synthesis:} 
-            The synthesis tool parses the HDL source, performs logical optimizations, and maps the design onto primitive cells that suit the target FPGA technology. 
-            The output is typically a structural \texttt{netlist} (e.g., EDIF or structural Verilog) which details how the design’s logic is broken down into LUTs, FFs, and other vendor-specific cells.
-
-        \item \textbf{Placement and Routing (Implementation):} 
-            In \emph{placement}, each logical cell from the synthesized netlist is assigned to a physical location on the FPGA fabric. 
-            For instance, LUTs and FFs go into specific \emph{BELs} within the device’s CLB sites, and specialized cells such as DSPs and Block RAMs must be placed in their corresponding tile types. 
-            Next, \emph{routing} determines how signals are physically wired through the FPGA’s configurable interconnect network. 
-            Modern tools often interleave these steps (e.g., fluid-placement routing or routing-aware placement) to better meet timing and area objectives.
-
-        \item \textbf{Bitstream Generation:} 
-            After a design is fully placed, routed, and timing-closed, the toolchain produces a final \emph{bitstream} that sets the configuration of every programmable element in the FPGA. 
-            This bitstream can then be loaded onto the device, either through vendor software or via a custom programming interface.
-
-        \item \textbf{Verification:} 
-            In parallel to the design flow, simulations and testbenches validate correctness of the user's design at multiple abstraction levels. 
-            Engineers may begin with behavioral simulations, then progress to post-synthesis simulations, and finally to post-implementation simulations that incorporate estimated routing delays. 
-            With each higher level of fidelity, computational requirements grow significantly due to increasing complexity and the need to analyze more variables over time. 
-            Ensuring correct functionality and meeting timing closure at the post-implementation stage is crucial before deploying the design to hardware. 
-            Given the importance of thorough verification, many established companies dedicate one verification engineer for every design engineer.
-
-    \end{enumerate}
-
-    This workflow underscores the critical role of \textbf{placement} in bridging the netlist to a physical realization. 
-    An efficient placement algorithm can drastically reduce compile times and improve design performance, enabling broader adoption of FPGAs in application spaces that require fast design iteration.
-
-\newpage
 \section{RapidWright API}
@@
 Researchers can prototype new EDA strategies without needing to re-implement an entire FPGA backend from scratch, thus accelerating innovation in placement and routing methodologies.
-
-\newpage
-\section{What is a Netlist?}
-    \label{sec:netlist}
-    In its simplest form, a netlist consists of a list of every component in an electronic design and a list of nets they connect to. 
-    Here, a net, or network, denotes any group of two or more interconnected elements - whether they be transistors, logic gates, or higher-level macrocells. 
-    More precisely, nets attach to component pins, and those pins reside on cells, each of which exposing multiple ports. 
-    In this way, a netlist makes explicit both the individual building blocks of a circuit and the precise wiring that links them.
-
-    In Vivado, a Netlist can be synthesized as a Hierarchical or a Flattened netlist. 
-    Figure ~\ref{fig:hierarchical_design} shows the Verilog code for a simple HDL design using module instance hierarchy. 
-    Figure ~\ref{fig:hier_netlist} shows the design synthesized into a hierarchical netlist, while Figure ~\ref{fig:flat_netlist} shows the same design synthesized into a flat netlist. 
-    In a synthesized Netlist, the leaf cells (lowest-level cells) are necessarily primitive cells from the vendor primitives library such as LUT6, FDRE, CARRY4, DSP48E1, etc..
-
-    \end{multicols}
-    {
-        \raggedright
-        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/top_level.png}
-        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_0.png}
-        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_1.png}
-        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_2.png}
-        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_3.png}
-        \captionof{figure}{A simple HDL design with module hierarchy.}
-        \label{fig:hierarchical_design}
-    }
-    {
-        \centering
-        \includegraphics[valign=c, width=11.5cm]{figures/netlist_synth/hier_netlist.png}
-        \includegraphics[valign=c, width=6cm]{figures/netlist_synth/hier_graph.png}
-        \captionof{figure}{
-            \textbf{Left:} A hierarchical netlist consisting of LUTs and FFs.
-            \textbf{Right:} The cell hierarchy graph.
-        }
-        \label{fig:hier_netlist}
-    }
-    \vspace{0.5cm}
-    {
-        \centering
-        \includegraphics[valign=c, width=10cm]{figures/netlist_synth/flat_netlist.png}
-        \includegraphics[valign=c, width=4cm]{figures/netlist_synth/flat_graph.png}
-        \captionof{figure}{
-            \textbf{Left:} A flattened netlist consisting of LUTs and FFs.
-            \textbf{Right:} The flattened cell hierarchy graph.
-        }
-        \label{fig:flat_netlist}
-    }
-    \begin{multicols}{2}
-
-    RapidWright offers an easy way to access, read, or manipulate this netlist for a given HDL design. 
-    Shown in Figure~\ref{fig:netlist_syn}
-
-    Verilog \texttt{module}s are represented simply as
-
-    \end{multicols}
-    {
-        \centering
-        \includegraphics[width=\columnwidth]{figures/raw_edif.png}
-        \captionof{figure}{A raw EDIF Netlist with uncountably many cells}
-        \label{fig:raw_edif}
-    }
-    \begin{multicols}{2}
-
-    .
-
-\newpage
-\section{Simulated Annealing}
-    \label{sec:simulated_annealing}
-    With a basic understanding of FPGA architecture, design placement, and RapidWright, we have all the necessary pieces to implement our SA placer. 
-    Here we outline in detail each substage of our implementation: PrePacking, Packing, and Placement. 
-    Shown in Figure \ref{fig:edif_design_device} shows an overview of the placement workflow.
-
-    {
-        \centering
-        \includegraphics[width=0.9\columnwidth]{figures/edif_design_device.png}
-        \captionof{figure}{Our placement workflow}
-        \label{fig:edif_design_device}
-    }
-
-    \end{multicols}
-    {
-        \centering
-        \includegraphics[width=0.8\columnwidth]{figures/substages.png}
-        \captionof{figure}{The data classes populated at each substage: PrepackedDesign, PackedDesign, and PlacedDesign.}
-        \label{fig:substages}
-    }
-    \begin{multicols}{2}
-
-    \subsection{Prepacking}
-        \label{subsec:prepacking}
-
-        The first step in our placement flow is \textbf{prepacking}. 
-        Recall from the 7-Series architecture that there are certain cell structures that must adhere to certain placements constraints to ensure legality, and by design, to minimize wirelength. 
-        The job of the prepacker is to consolidate the cells in the EDIF Netlist into clusters or groups of clusters that naturally reflect these placement constraints.
-
-        CARRY4 chains must necessarily be placed vertically and consecutively across a column of SLICEs in ascending order. 
-        Likewise, DSP48E1 cascades must necessarily be placed vertically and consecutively across a column of DSP48E1 Sites in ascending order. 
-        A LUT-FF pair may be placed freely, but should be placed in the same lane within the same SLICE to minimize wirelength.
-
-        The raw post-synthesis netlist only tells us the list of nets and the cell ports that they connect to. 
-        It does not report the presence of any macro cell structures (CARRY4 chains, etc.). 
-        Thus, we must traverse the netlist to detect these cell structures and store that structure information in a class we will call \texttt{PrepackedDesign}.
-
-        A group of LUT-FF pairs may be placed in the same SLICE, with the constraint that the FFs must share the same Clock-Enable (CE) and Set-Reset (SR) nets.
-        Clustering LUT-FF pairs like this can reduce the redundancy of having to route the same CE and SR nets to many SLICEs by routing the nets to fewer SLICEs and connecting them to the individual FFs within via intra-Site routing, and at the same time, packing greater logic density over a smaller area of the device.
-        Recall that up to eight FFs may be placed in the same SLICE, thus theoretically, up to eight LUT-FF pairs can be placed in the same SLICE.
-
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/slicel.png}
-            \captionof{figure}{A SLICEL Site}
-            \label{fig:slicel}
-        }
-
-        Utilizing all 8 LUT-FF lanes in a SLICE can help reduce device area utilization and minimize wirelength, but \emph{too much} logic density in an area can contribute to general routing congestion.
-        Furthermore, attempting to fill all 8 lanes in a SLICE requires some careful consideration of LUT BEL constraints. 
-        The LUT BELs in a SLICE, as shown in Figure \ref{fig:slicel}, the LUTs appear to be interleaved.
-
-        Show a block diagram of a LUT6 comprising of two LUT5s.
-
-        Recall that a LUT can accommodate one 6-input boolean function or two 5-input boolean functions sharing the same inputs or any two 3-input or less boolean functions regardless of shared inputs.
-        A LUT6 Cell (a LUT with 6-inputs) will actually occupy two lanes in a SLICEL, rendering one of the FF BELs in either lane ineligible for another LUT-FF pair. 
-        The LUT BELs
-
-    \subsection{Packing}
-        \label{subsec:packing}
-
-    \subsection{Placement}
-        \label{subsec:placement}
-        Up until now we have only organized the logical \texttt{Cells} into \texttt{SiteInsts}. 
-        This is where simulated annealing actually begins.
-
-\section{Placement Results}
-    \begin{itemize}
-    \item Random Site Selection:
-    \item Midpoint Site Selection:
-    \item Hybrid: 
-        Random Site selection until progress stagnates, then use Midpoint Site selection. 
-        Randomly choose betwee Random and Midpoint Site selection each iteration. 
-    \item Cooling Schedule: 
-        Greedy. 
-        Geometric with modifiable alpha and intial temperature. 
-    \end{itemize}
-
-\section{Future Work}
-    \begin{itemize}
-    \item Different cooling schedules: linear cooling, logarithmic cooling, piecewise cooling.
-    \item Support for more macros: buffer cells, FIFO, clock generators, etc..
-    \item Support Ultrascale architecture. 
-    \item Better packing scheme. Currently pseudorandom packing. 
-    \item Inter-Site BEL swapping. 
-    \item BEL-centric placement over Site-centric placement.
-    \item Analytical placement. Global placement, legalization, detailed placement. 
-    \end{itemize}
-
-\section{Conclusion}
-    fdsfdsafdsa
-
-% \section{Analytical Placement}
-%     Introduce placement with legalization methods. \\
-%     Many AP approaches use a Global placement, Legalization, then Detailed placement approach. \\
-%     Our SA only considers legal moves via bookkeeping structures, so it has no concept of global placement or legalization. \\
-%     It is a detailed placement only approach. \\
-%     \begin{itemize}
-%         \item Talk about HeAP as an example.
-%         \item Global Placement, Legalization, then Detailed Placement approach.
-%         \item Global placement using the Bound2Bound net model from SimPL.
-%         \item Legalization Step 1: \\
-%             Find an area of the device that is overutilized (illegal) for twhich the blocks contained within must spread to a larger area. \\
-%             To obtain this overutilized area, adjacent locations (Sites, Tiles, etc.) on the device that are occupied by more than one block (SiteInst, ModuleInst, etc.) are repeatedly clustered together, until all clusters are bordered on all sides by non-overutilized locations.
-\\
-%             Then, the area (?) is expanded in both the x and y dimensions (?) until it is large enough to accommodate all blocks contained within. \\
-%             Specifically, the area is expanded until its "occupance" \(O_A\) divided by its "capacity" \(C_A\) is less than a maximum "utilization factor" $\beta$, where $\beta$ is less than or equal to 1.
-HeAP uses 0.9.
-%         \item Legalization Step 2: \\
-%             Two cuts are generated: a "source cut" and a "target cut". \\
-%             The source cut pertains to the blocks being placed (SiteInsts in home buffer). \\ 
-%             The target cut pertains to the area into which the blocks are placed (away buffer). \\
-%             The source cut splits the blocks into 2 partitions, while the target cut splits the area into 2 sub-areas, into which the blocks in each partition are spread. \\
-%             Two objectives are minimized during this process: The imbalance between the number of blocks in each partition, and the difference in the utilization of each sub-area. \\
-%             Utilization defined as the occupancy divided by the capacity of the sub-area. \\
-%             \(U_{sub-area} = \frac{O_{sub-area}}{C_{sub-area}}\) \\
-%             To generate the source cut, the cells are first sorted by their x or y location, depending on the orientation of the desired cut. \\
-%             Once the cells are sorted, the source cut generation is akin to choosing a pivot in a sorted list, where all blocks to the left of the pivot are assigned to the left/bottom partition and all blocks to the right of the pivot are assigned to the right/top partition.
-\\
-%             The target cut is an x or y cut of the area such that all blocks in each partition fit in their respective sub-areas, and such that \( | U_{sub-area_1} 
-%              - U_{sub-area_2} | \) is minimized.
-%     \end{itemize}
-
-\section{Appendix}
-
-    \begin{figure*}[b]
-        \centering
-        \includegraphics[width=\textwidth]{figures/slicem.png}
-        \caption{SLICEM Site}
-        \label{fig:slicem}
-    \end{figure*}
-
-    \begin{figure*}[t]
-        \centering
-        \includegraphics[width=\textwidth]{figures/slicel.png}
-        \caption{SLICEL Site}
-        \label{fig:slicel}
-    \end{figure*}
-
-\end{multicols}
-
-\end{document}

### 2025-05-03 latex/paper/sections/08-netlist.tex
split tex file into multiple files

\section{What is a Netlist?}
\label{sec:netlist}
In its most general form, a netlist is a list of every component in an electronic design paired with a list of nets they connect to. 
Depending on the working abstraction level, these components can be transistors, logic gates, macrocells, or increasingly higher-level modules. 
Generally, a net denotes any group of two or more interconnected components.
In an electronics context, a net can be though of as a wire connecting multiple pins between multiple components, with each wire having one voltage source and one or more voltage sinks. 
Thus, if one were to express the netlist as a graph, the graph would consist of nodes (components) connected by hyperedges (wires). 
More precisely, these hyperedges connect the pins between the components, not the components themselves, with each component exposing multiple pins or ports.

In FPGA context, the components are logical cells (\texttt{LUTs}, \texttt{CARRY4s}, etc.) or hierarchical cells (Verilog module instances) with pins connected together by wires. 
In Vivado, a Netlist can be synthesized as a Hierarchical or a Flattened netlist. 
Figure ~\ref{fig:hierarchical_design} shows an example a Verilog design using module instance hierarchy, while figure ~\ref{fig:hier_netlist} shows the same design synthesized into a flattened netlist. 
The synthesizer attempts to construct the module hierarchy as close to the module instantiation hierarchy as defined by the raw user design entry. 
Figure ~\ref{fig:flat_netlist} shows the same design but synthesized into a flat netlist. 
In either synthesized Netlist, the leaf cells, (deepest level cells), must necessarily consist only of primitive cells from the architecture primitives library (\texttt{LUT6}, \texttt{FDRE}, \texttt{CARRY4}, \texttt{DSP48E1}, etc.). 
The netlist can be compiled and exported as a purely structural low-level Verilog file, or an Electrinic Design Interchange Format (EDIF) file, both describing the netlist explicitly as a list of logical cells connected by a list of wires.

\end{multicols}
{
    \raggedright
    \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/top_level.png}
    \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_0.png}
    \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_1.png}
    \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_2.png}
    \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_3.png}
    \captionof{figure}{A simple HDL design with module hierarchy.}
    \label{fig:hierarchical_design}
}
{
    \centering
    \includegraphics[valign=c, width=11.5cm]{figures/netlist_synth/hier_netlist.png}
    \includegraphics[valign=c, width=6cm]{figures/netlist_synth/hier_graph.png}
    \captionof{figure}{
        \textbf{Left:} A hierarchical netlist consisting of LUTs and FFs.
        \textbf{Right:} The cell hierarchy graph.
    }
    \label{fig:hier_netlist}
}
\vspace{0.5cm}
{
    \centering
    \includegraphics[valign=c, width=10cm]{figures/netlist_synth/flat_netlist.png}
    \includegraphics[valign=c, width=4cm]{figures/netlist_synth/flat_graph.png}
    \captionof{figure}{
        \textbf{Left:} A flattened netlist consisting of LUTs and FFs.
        \textbf{Right:} The flattened cell hierarchy graph.
    }
    \label{fig:flat_netlist}
}
\begin{multicols}{2}

\subsection{Netlist Traversal and Manipulation in RapidWright}

One of RapidWright's most powerful features is netlist manipulation via the \texttt{edif} package of classes. A netlist can be easily extracted from a \texttt{.dcp} design checkpoint and traversed like the following:

\begin{lstlisting}[language=Java, caption={Netlist extraction}, label={lst:netlist_extract}]
Design design = Design.readCheckpoint("synth.dcp")
EDIFNetlist = design.getNetlist();
\end{lstlisting}

The \texttt{Netlist} class in RapidWright stores hierarchical information about the \texttt{Design} and can be accessed and manipulated

SHOULD I INCLUDE A RAPIDWRIGHT HELLO WORLD EXAMPLE JUST TO GIVE AN IMPRESSION?

OR CAN I JUST DIRECT THE READER TO THE RAPIDWRIGHT DOCUMENTATION?

### 2025-05-03 latex/paper/sections/08-netlist.tex
rw hello world

@@
 In Vivado, a Netlist can be synthesized as a Hierarchical or a Flattened netlist. 
-Figure ~\ref{fig:hierarchical_design} shows an example a Verilog design using module instance hierarchy, while figure ~\ref{fig:hier_netlist} shows the same design synthesized into a flattened netlist.
-The synthesizer attempts to construct the module hierarchy as close to the module instantiation hierarchy as defined by the raw user design entry. 
-Figure ~\ref{fig:flat_netlist} shows the same design but synthesized into a flat netlist. 
-In either synthesized Netlist, the leaf cells, (deepest level cells), must necessarily consist only of primitive cells from the architecture primitives library (\texttt{LUT6}, \texttt{FDRE}, \texttt{CARRY4}, \texttt{DSP48E1}, etc.).
+Figure ~\ref{fig:hierarchical_design} shows an example a Verilog design using module instance hierarchy, while figure ~\ref{fig:hier_netlist} shows the design synthesized into a hierarchical netlist with \textbf{hierarchical cells} and \textbf{leaf cells}.
+The synthesizer attempts to construct the module hierarchy as close to the module instantiation hierarchy defined by the user design entry. 
+Figure ~\ref{fig:flat_netlist} shows the same design but synthesized into a flattened netlist.
+
+In either synthesized netlist, the \textbf{leaf cells}, (deepest level cells), must necessarily consist only of \textbf{primitive cells} from the architecture's primitive cell library (\texttt{LUT6}, \texttt{FDRE}, \texttt{CARRY4}, \texttt{DSP48E1}, etc.).
 The netlist can be compiled and exported as a purely structural low-level Verilog file, or an Electrinic Design Interchange Format (EDIF) file, both describing the netlist explicitly as a list of logical cells connected by a list of wires.
@@
 
-\begin{lstlisting}[language=Java, caption={Netlist extraction}, label={lst:netlist_extract}]
+\begin{lstlisting}[language=Java, caption={Netlist extraction and traversal}, label={lst:netlist_extract}]
 Design design = Design.readCheckpoint("synth.dcp")
 EDIFNetlist = design.getNetlist();
+
+// Example task:
+// Extract the set of all nets from the design.
+
+// Initialize a new Set:
+Set<EDIFHierNet> netSet = new HashSet<>();
+
+// Access every cell:
+List<EDIFHierCellInst> ehcis = EDIFNetlist.getAllLeafHierCellInstances();
+
+// Traverse cell list:
+for (EDIFHierCellInst ehci : ehcis) {
+    // Access every port on this cell:
+    List<EDIFHierPortInst> ehpis = ehci.getHierPortInsts();
+    for (EDIFHierPortInst ehpi : ehpis) {
+        // Access the net on this port:
+        EDIFHierNet ehn = ehpi.getHierarchicalNet();
+        netSet.add(ehn);
+    }
+}
+
+// Example downstream task:
+// For each net, access the source cells and sink cells.
+for (EDIFHierNet ehn : netSet) {
+    // Access the source ports (there should only be one)
+    List<EDIFHierPortInst> sourcePorts = ehn.getLeafHierPortInsts(true, false);
+    // Access the source cell
+    EDIFHierCellInst sourceCell = sourcePorts.get(0).getHierarchicalInsts();
+
+    // Access the sink ports
+    List<EDIFHierPortInst> sinkPorts = ehn.getLeafHierPortInsts(false, true);
+    // Access the sink cells
+    List<EDIFHierCellInst> sinkCells = new ArrayList<>();
+    for (EDIFHierPortInst ehpi : sinkPorts) {
+        sinkCells.add(ehpi.getHierarchicalInst());
+    }
+}
+
 \end{lstlisting}
@@
 The \texttt{Netlist} class in RapidWright stores hierarchical information about the \texttt{Design} and can be accessed and manipulated
-
-SHOULD I INCLUDE A RAPIDWRIGHT HELLO WORLD EXAMPLE JUST TO GIVE AN IMPRESSION?
-
-OR CAN I JUST DIRECT THE READER TO THE RAPIDWRIGHT DOCUMENTATION?

### 2025-05-04 latex/paper/sections/03-introduction.tex
edit rw helloworld

@@
 FPGA CAD flows mainly encompass synthesis, placement, and routing; all of which are NP-hard problems, of which placement is one of the most time-consuming processes. 
-Inefficienct placement strategy not only extends design times from hours to days, thereby elevating cost and reducing engineering productivity, but also limits the broader adoption of FPGAs by software engineers who expect compile times akin to those of conventional software compilers like {\tt gcc}.
+Inefficienct placement strategy not only extends design times from hours to days, thereby elevating cost and reducing engineering productivity, but also limits the broader adoption of FPGAs by software engineers who expect compile times akin to those of software compilers like {\tt gcc}.
 
@@
 SA remains a popular approach in FPGA placement research due to its simplicity and robustness in handling the discrete architectural constraints of FPGA devices. 
-While SA yields surprisingly good results given relatively simple rules, it is ultimately a heuristic and stochastic approach that explores the vast placement space by making random moves. 
+While SA yields surprisingly good results given relatively simple rules, it is ultimately a heuristic approach that explores the vast placement space by making random moves. 
 Most of these moves will be rejected, meaning that SA must run many iterations, usually hundreds to thousands, to arrive at a desirable solution.

### 2025-05-04 latex/paper/sections/08-netlist.tex
edit rw helloworld

@@
 In an electronics context, a net can be though of as a wire connecting multiple pins between multiple components, with each wire having one voltage source and one or more voltage sinks. 
-Thus, if one were to express the netlist as a graph, the graph would consist of nodes (components) connected by hyperedges (wires). 
+Thus, one could express the netlist as a hypergraph, nodes representing component ports, hyperedges representing wires connecting two or more component ports. 
 More precisely, these hyperedges connect the pins between the components, not the components themselves, with each component exposing multiple pins or ports.
@@
     \centering
-    \includegraphics[valign=c, width=10cm]{figures/netlist_synth/flat_netlist.png}
+    \includegraphics[valign=c, width=13cm]{figures/netlist_synth/flat_netlist.png}
     \includegraphics[valign=c, width=4cm]{figures/netlist_synth/flat_graph.png}
@@
 
-One of RapidWright's most powerful features is netlist manipulation via the \texttt{edif} package of classes.
+\begin{itemize}
+\item
+
+\end{itemize}
+
+One of RapidWright's most powerful features is netlist traversal and manipulation via the \texttt{edif} classes.
 A netlist can be easily extracted from a \texttt{.dcp} design checkpoint and traversed like the following:
 
-\begin{lstlisting}[language=Java, caption={Netlist extraction and traversal}, label={lst:netlist_extract}]
+\begin{lstlisting}[language=java, caption={Netlist extraction and traversal}, label={lst:netlist_extract}]
 Design design = Design.readCheckpoint("synth.dcp")
@@
 // Example task:
-// Extract the set of all nets from the design.
+// Extract the set of all unique nets from the design.
 
 // Initialize a new Set:
-Set<EDIFHierNet> netSet = new HashSet<>();
+Set<EDIFNet> netSet = new HashSet<>();
 
-// Access every cell:
-List<EDIFHierCellInst> ehcis = EDIFNetlist.getAllLeafHierCellInstances();
+// Access all leaf cells
+List<EDIFCellInst> ecis = netlist.getAllLeafCellInstances();
 
-// Traverse cell list:
-for (EDIFHierCellInst ehci : ehcis) {
-    // Access every port on this cell:
-    List<EDIFHierPortInst> ehpis = ehci.getHierPortInsts();
-    for (EDIFHierPortInst ehpi : ehpis) {
-        // Access the net on this port:
-        EDIFHierNet ehn = ehpi.getHierarchicalNet();
-        netSet.add(ehn);
+// Traverse the cell list
+for (EDIFCellInst eci : ecis) {
+    // Access the ports on this cell
+    Collection<EDIFPortInst> epis = eci.getPortInsts();
+    for (EDIFPortInst epi : epis) {
+        // Access the net on this port
+        EDIFNet net = epi.getNet();
+        netSet.add(net);
     }
@@
 
-// Example downstream task:
-// For each net, access the source cells and sink cells.
-for (EDIFHierNet ehn : netSet) {
-    // Access the source ports (there should only be one)
-    List<EDIFHierPortInst> sourcePorts = ehn.getLeafHierPortInsts(true, false);
-    // Access the source cell
-    EDIFHierCellInst sourceCell = sourcePorts.get(0).getHierarchicalInsts();
+// Downstream task:
+// For each unique net, print out the connected cells.
 
-    // Access the sink ports
-    List<EDIFHierPortInst> sinkPorts = ehn.getLeafHierPortInsts(false, true);
-    // Access the sink cells
-    List<EDIFHierCellInst> sinkCells = new ArrayList<>();
-    for (EDIFHierPortInst ehpi : sinkPorts) {
-        sinkCells.add(ehpi.getHierarchicalInst());
+// Traverse the set of nets
+for (EDIFNet net : netSet) {
+    System.out.println("Net: " + net.getName());
+    // Access the ports connected to this net
+    Collection<EDIFPortInst> epis = net.getPortInsts();
+    for (EDIFPortInst epi : epis) {
+        // Access the cell that this port belongs to
+        EDIFCellInst eci = epi.getCellInst();
+        if (eci == null) {
+            // (top_level ports have no associated cell)
+            continue;
+        } else {
+            System.out.println(
+                "\tCell: " + eci.getName() + 
+                ",\tCellType: " + eci.getCellName()
+            );
+        }
     }
@@
 
-The \texttt{Netlist} class in RapidWright stores hierarchical information about the \texttt{Design} and can be accessed and manipulated
+{
+    \centering
+    \includegraphics[valign=c, width=\columnwidth]{figures/traversal.png}
+    \captionof{figure}{\textttt{Netlist} traversal via \texttt{EDIFHierCellInst}, \texttt{EDIFHierPortInst}, and \texttt{EDIFHierNet} objects}
+    \label{fig:traversal}
+}

### 2025-05-07 latex/paper/sections/08-netlist.tex
netlist stuff

@@
-\section{What is a Netlist?}
-\label{sec:netlist}
+\subsection{What is a Netlist?}
+\label{subsec:netlist}
 In its most general form, a netlist is a list of every component in an electronic design paired with a list of nets they connect to. 
-Depending on the working abstraction level, these components can be transistors, logic gates, macrocells, or increasingly higher-level modules. 
+Depending on the abstraction level at hand, these components can be transistors, logic gates, macrocells, or increasingly higher-level modules. 
 Generally, a net denotes any group of two or more interconnected components.
 In an electronics context, a net can be though of as a wire connecting multiple pins between multiple components, with each wire having one voltage source and one or more voltage sinks. 
-Thus, one could express the netlist as a hypergraph, nodes representing component ports, hyperedges representing wires connecting two or more component ports. 
-More precisely, these hyperedges connect the pins between the components, not the components themselves, with each component exposing multiple pins or ports.
+Thus, one could express the netlist as a hypergraph, nodes representing components, hyperedges representing wires connecting two or more component. 
+More precisely, these hyperedges connect the ports between the components, not the components themselves, with each component exposing multiple ports.
 
@@
 In Vivado, a Netlist can be synthesized as a Hierarchical or a Flattened netlist. 
-Figure ~\ref{fig:hierarchical_design} shows an example a Verilog design using module instance hierarchy, while figure ~\ref{fig:hier_netlist} shows the design synthesized into a hierarchical netlist with \textbf{hierarchical cells} and \textbf{leaf cells}.
+Figure ~\ref{fig:hierarchical_design} shows an example a Verilog design with modules instantiated in a hierarchy. 
+Figure ~\ref{fig:hier_netlist} shows the design synthesized into a hierarchical netlist with \textbf{hierarchical cells} and \textbf{leaf cells}. 
 The synthesizer attempts to construct the module hierarchy as close to the module instantiation hierarchy defined by the user design entry. 
@@
 The netlist can be compiled and exported as a purely structural low-level Verilog file, or an Electrinic Design Interchange Format (EDIF) file, both describing the netlist explicitly as a list of logical cells connected by a list of wires.
+
+RapidWright represents the logical netlist objects via the \texttt{edif} classes:
+\begin{itemize}
+    \item \texttt{EDIFNetlist}: The full logical netlist of a \texttt{Design}. 
+    \item \texttt{EDIFNet}: Represents a logical net within an \texttt{EDIFNetlist}.
+    \item \texttt{EDIFHierNet}: Combines an \texttt{EDIFNet} with a full hierarchical instance name to uniquely identify a net in a netlist.
+    \item \texttt{EDIFCell}: Represents a logical cell in an \texttt{EDIFNetlist}. 
+    \item \texttt{EDIFCellInst}: Represents an instance of an \texttt{EDIFCell}. 
+    \item \texttt{EDIFHierCellInst}: An \texttt{EDIFCellInst} with its hierarchy, described by all the \texttt{EDIFCellInst}s that sit above it within the netlist.
+    \item \texttt{EDIFPort}: Represents a port on an \texttt{EDIFCell}. 
+    \item \texttt{EDIFPortInst}: Represents an instance of a port on an \texttt{EDIFCellInst}. 
+    \item \texttt{EDIFHierPortInst}: Combines an \texttt{EDIFHierPortInst} with a full hierarchical instance name to uniquely identify a port instance in a netlist. 
+\end{itemize}
+
+\newpage
 
@@
 }
+\vspace{0.5cm}
 {
@@
 }
+\newpage
 \begin{multicols}{2}
@@
 
-\begin{itemize}
-\item
-
-\end{itemize}
+{
+    \centering
+    \includegraphics[valign=c, width=\columnwidth]{figures/traversal.png}
+    \captionof{figure}{\texttt{Netlist} traversal via \texttt{EDIFHierCellInst}, \texttt{EDIFHierPortInst}, and \texttt{EDIFHierNet} objects}
+    \label{fig:traversal}
+}
+\vspace{0.5cm}
 
@@
 }
-
 \end{lstlisting}
 
-{
-    \centering
-    \includegraphics[valign=c, width=\columnwidth]{figures/traversal.png}
-    \captionof{figure}{\textttt{Netlist} traversal via \texttt{EDIFHierCellInst}, \texttt{EDIFHierPortInst}, and \texttt{EDIFHierNet} objects}
-    \label{fig:traversal}
-}
+\newcolumn

### 2025-05-07 latex/paper/sections/09-packing.tex
netlist stuff

--- 2025-04-28 latex/paper/paper.tex
@@
-\documentclass{article}
-\usepackage{hyperref}
-\usepackage{enumitem}
-\usepackage{graphicx}
-\usepackage{amsmath}
-\usepackage{mathpazo}
-\usepackage{xcolor}
-\usepackage{float}
-\usepackage{listings}
-\lstset{
-  basicstyle=\ttfamily\small,
-  breaklines=true,
-  columns=fullflexible,
-  frame=single,
-  showspaces=false,
-  showtabs=false,
-  keepspaces=false
-}
-\usepackage[export]{adjustbox}
-\usepackage{multicol}
-\usepackage{caption}
-\usepackage[a4paper,
-            bindingoffset=0.2in,
-            left=0.5in,
-            right=0.5in,
-            top=0.5in,
-            bottom=0.5in,
-            footskip=.25in]{geometry}
-
-\begin{document}
-\title{MS Technical Paper: \\ Placement Algorithms for Heterogeneous FPGAs}
-\author{Brian B Cheng \\ Rutgers University Department of Electrical and Computer Engineering}
-
-\date{}
-\maketitle
-
-\begin{multicols}{2}
-
-\section{Keywords}
-\begin{itemize}
-    \item FPGA, EDA, Placement, Simulated Annealing, Optimization, RapidWright
-\end{itemize}
-
-\section{Abstract}
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\
-
-\section{Introduction}
-
-    Field-Programmable Gate Arrays (FPGAs) have witnessed rapid growth in capacity and versatility, driving significant advances in computer-aided design (CAD) and electronic design automation (EDA) methodologies.
-    Since the early-to-mid 2000s, the stagnation of single-processor performance relative to the rapid increase in integrated circuit sizes has led to a design productivity gap, where the computational effort for designing complex chips continues to rise.
-    FPGA CAD flows mainly encompass synthesis, placement, and routing; all of which are NP-hard problems, of which placement is one of the most time-consuming processes. 
-    Inefficienct placement strategy not only extends design times from hours to days, thereby elevating cost and reducing engineering productivity, but also limits the broader adoption of FPGAs by software engineers who expect compile times akin to those of conventional software compilers like {\tt gcc}.
-
-    For these reasons, FPGA placement remains a critical research effort even today. 
-    In this paper, we study and implement established placement methods. 
-    To do this, we use the RapidWright API, which is a semi-open-source research effort from AMD/Xilinx that enables custom solutions to FPGA design implementations and design tools that are not offered by their industry-standard FPGA environment, Vivado.
-    We implement multiple variations of simulated annealing placers for Xilinx's 7-series FPGAs, with an emphasis on minimizing total wirelength while mitigating runtime. 
-    Our implementation is organized into three consecutive substages. 
-    The \textbf{prepacking} stage involves traversing a raw EDIF netlist to identify recurring cell patterns—such as CARRY chains, DSP cascades, and LUT-FF pairs—that are critical for efficient mapping and legalization.
-    In the subsequent \textbf{packing} stage, these identified patterns, along with any remaining loose cells, are consolidated into SiteInst objects that encapsulate the FPGA’s discrete resource constraints and architectural nuances.
-    Finally, the \textbf{placement} stage employs a simulated annealing (SA) algorithm to optimally assign SiteInst objects to physical sites, aiming to minimize total wirelength while adhering to the constraints of the 7-series architecture.
-
-    Simulated annealing iteratively swaps placement objects guided by a cost function that decides which swaps should be accepted or rejected. 
-    Hill climbing is permitted by occasionally accepting moves that increase cost, in hope that such swaps may later lead to a better final solution. 
-    SA remains a popular approach in FPGA placement research due to its simplicity and robustness in handling the discrete architectural constraints of FPGA devices. 
-    While SA yields surprisingly good results given relatively simple rules, it is ultimately a heuristic and stochastic approach that explores the vast placement space by making random moves. 
-    Most of these moves will be rejected, meaning that SA must run many iterations, usually hundreds to thousands, to arrive at a desirable solution.
-
-    In the ASIC domain, where placers must handle designs with millions of cells, the SA approach has largely been abandoned in favor of analytical techniques, owing to SA's runtime and poor scalability.
-    Modern FPGA placers have also followed suit, as new legalization strategies allow FPGA placers to leverage traditionally ASIC placement algorithms and adapt them to the discrete constraints of FPGA architectures.
-    While this paper does not present a working analytical placer, it will explore ways to build upon our existing infrastructure (prepacker and packer) to replace SA with AP.
-
-    The paper first begins by elaborating on general FPGA architecture and then specifically the Xilinx 7-Series architecture. 
-    Then, the paper will elaborate on the FPGA design flow, then the role that the RapidWright API plays in the design flow. 
-    We explain in detail each of these concepts for a broader audience as they provide much needed context for FPGA placement algorithms as a concept. 
-    However, readers who are already familiar with these concepts can skip directly to the RapidWright API section \ref{sec:rapidwright_api} or to the Simulated Annealing section \ref{sec:simulated_annealing}.
-
-\section{FPGA Architecture History}
-
-    Before any work can begin on an FPGA placer, it is necessary to understand both the objects being placed and the medium in which they are placed.
-    Configurable logic devices have undergone significant evolution over the past four decades. 
-    We will briefly review the evolution of configurable logic architecture starting in the 1970s and quickly work our way up to modern day FPGA architecture.
-
-    \textbf{PLA:} \quad
-    The journey began with the Programmable Logic Array (PLA) in the early 1970s. 
-    The PLA implemented output logic using a programmable-OR and programmable-AND plane that formed a sum-of-products equation for each output through programmable fuses. 
-    Around the same time, the Programmable Array Logic (PAL) was introduced. 
-    The PAL simplified the PLA by fixing the OR gates, resulting in a fixed-OR, programmable-AND design, which sacrificed some logic flexibility to simplify its manufacture. 
-    Figure \ref{fig:pal_2} shows one such PAL architecture.
-
-    \textbf{CPLD:} \quad 
-    Later in the same decade came the Complex Programmable Logic Device (CPLD), which took the form of an array of Configurable Logic Blocks (CLBs). 
-    These CLBs were typically modified PAL blocks that included the PAL itself along with macrocells such as flip-flops, multiplexers, and tri-state buffers. 
-    The CPLD functioned as an array of PALs connected by a central programmable switch matrix and could be programmed using a hardware description language (HDL) like VHDL. 
-    Figure \ref{fig:cpld} shows one such CPLD architecture. 
-    \\
-
-    {
-        \centering
-        \includegraphics[width=\columnwidth]{figures/pal_2.png}
-        \captionof{figure}{PAL architecture with 5 inputs, 8 programmable AND gates and 4 fixed OR gates}
-        \label{fig:pal_2}
-        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/Programmable-Logic-Devices/Programmable-Array-Logic-PAL/
-        % https://www.naukri.com/code360/library/difference-between-pla-and-pal 
-    }
-
-    \newcolumn
-
-    \textbf{Homogeneous FPGA:} \quad 
-    The mid-1980s saw the introduction of homogeneous FPGAs, which were built as a grid of CLBs. 
-    Rather than using a central programmable switch matrix as in CPLDs, FPGAs adopted an island style architecture in which each CLB is surrounded on all sides by programmable routing resources, as shown in Figure \ref{fig:fpga}.
-    The first commercially viable FPGA, produced by Xilinx in 1984, featured 16 CLBs arranged in a 4x4 grid. 
-    As FPGA technology advanced, CLBs were redesigned to use lookup tables (LUTs) instead of PAL arrays for greater logic density. 
-    The capacity of an FPGA was often measured by how many logical elements or CLBs it offered, which grew from hundreds to thousands and now to hundreds of thousands of CLBs.
-    \\
-
-    {
-        \centering
-        \includegraphics[width=0.8\columnwidth]{figures/cpld.png}
-        \captionof{figure}{CPLD architecture with 4 CLBs (PAL-like blocks)}
-        \label{fig:cpld}
-        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/CPLD/Complex-Programmable-Logic-Device-CPLDs/
-    }
-    {
-        \centering
-        \includegraphics[width=\columnwidth]{figures/homogenous_fpga.png}
-        \captionof{figure}{A homogeneous island-style FPGA architecture with 16 CLBs in a grid.}
-        \label{fig:fpga}
-        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/FPGA/Field-programmable-gate-array-FPGA/
-    }
-
-    \newpage
-
-    {
-        \centering
-        \includegraphics[width=\columnwidth]{figures/heterogenous_fpga_3.jpg}
-        \captionof{figure}{A heterogeneous island-style FPGA with a mix of CLBs and macrocells.}
-        \label{fig:fpga}
-        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/FPGA/Field-programmable-gate-array-FPGA/
-    }
-    \vspace{0.5cm}
-
-    \textbf{Heterogenous FPGA:} \quad 
-    This brings us to modern day FPGA architectures. 
-    To meet the needs of increasingly complex designs, FPGA vendors introduced heterogeneous FPGAs. 
-    In these devices, hard macros such as Block RAM (BRAM) and Digital Signal Processing (DSP) slices are integrated into the programmable logic fabric along with CLBs, like shown in Figure \ref{fig:fpga}.
-    This design enables the direct instantiation of common subsystems like memories and multipliers, without having to recreate them from scratch using CLBs. 
-    Major vendors such as Xilinx and Altera now employ heterogeneous island-style architectures in their devices. 
-    As designs become increasingly large and complex, FPGAs meet the demand by becoming increasingly heterogenous, incorporating a wider variety of hard macros into the fabric.
-
-\section{Xilinx 7-Series Architecture}
-    \label{sec:7_series}
-    The Xilinx 7-Series devices, first introduced in 2010, follow a heterogeneous island-style architecture as discussed previously. 
-    Although the 7-Series was later superseded in 2013 by the UltraScale architecture, the 7-Series remains highly relevant due to its accessibility, wide availability, and compatibility with open-source tooling.
-    Representative sub-families include Artix-7, Kintex-7, Virtex-7, and Zynq-7000, each designed with different performance and cost trade-offs but all follow the core 7-Series architecture.
-
-    Figure~\ref{fig:hierarchy_4} illustrates a high-level view of the hierarchical organization of a 7-Series FPGA. 
-    At its lowest level, the device consists of a large array of atomic components called \emph{Basic Elements of Logic}~(\textbf{BELs}). 
-    These BELs encompass look-up tables (\textbf{LUTs}), flip-flops (\textbf{FFs}), block RAMs (\textbf{BRAMs}), DSP slices (\textbf{DSPs}), and the configurable interconnect fabric. 
-    They constitute the fundamental building blocks for implementing digital circuits on the FPGA.
-    Note how the Tile arrangement is columnar, where each column will have only one Tile type.
-
-    To manage this complexity, Xilinx organizes these BELs into incrementally abstract structures. 
-    First, \textbf{BELs} are grouped into \textbf{Sites}. 
-    Each Site is embedded into a \textbf{Tile}, and Tiles are further arranged into \textbf{Clock Regions}. 
-    In some high-density devices, multiple Clock Regions may be consolidated into one or more \textbf{Super Logic Regions}~(SLRs). 
-    However, for the scope of this paper, we focus on Xilinx 7-Series devices with only a single SLR.
-
-    \end{multicols}
-    {
-        \centering
-        \includegraphics[valign=c, width=0.3\columnwidth]{figures/hierarchy_5.png}
-        \includegraphics[valign=c, width=0.6\columnwidth]{figures/hierarchy_4.png}
-        \captionof{figure}{Architecture Hierarchy of a Xilinx FPGA}
-        \label{fig:hierarchy_4}
-    }
-    \vfill
-    \begin{multicols}{2}
-
-    \newpage
-    \subsection{CLB SLICEs}
-
-    In the 7-Series architecture, the term \emph{CLB} (Configurable Logic Block) refers to a \emph{CLB Tile} that contains two \emph{SLICE} Sites. 
-    Xilinx offers two variants of SLICE Sites: \textbf{SLICEL} and \textbf{SLICEM}.
-    \begin{itemize}
-        \item Each \textbf{SLICEL} has a set of BELs including eight LUTs, eight FFs, and one   \emph{CARRY4} adder. The LUT BELs in a SLICEL can only host LUT Cells. 
-        \item The \textbf{SLICEM} includes all the features of a SLICEL but its LUT BELs can host both LUT Cells, which are asynchronous ROM elements, or RAM32M Cells, which are synchronous 32-deep RAM elements.
-These cells are also referred to as \emph{Distributed RAM} in the Xilinx documentation.
-        These cells can offer an alternative to the larger, more dedicated 18K-36K RAMB18E1 cells when RAM resources are highly utilized. 
-    \end{itemize}
-    In a typical 7-Series device, approximately 75\% of the SLICE Sites are SLICELs and 25\% are SLICEMs. 
-    A single CLB Tile can therefore host either two SLICELs or one SLICEL and one SLICEM.
-    For simplicity, however, we will only consider SLICELs for general logic and use the dedicated RAMB18E1 cells to implement RAM elements.
-
-    The BELs in these SLICEs facilitate the bulk of the general programmability of the FPGA fabric. 
-    We will explain in detail the motivation and function of these BELs.
-
-    \textbf{LUTs} \quad
-        Combinational logic is universal to all HDL designs. 
-        As the their name suggests, Look‑Up Tables (LUTs) facilitate combinational logic by acting as tiny asynchronously‑accessed ROMs whose contents are fixed when the FPGA is programmed.  
-        For any boolean function, the synthesizer precalculates the output to every possible input combination and stores the resulting truth table into a LUT's static memory. 
-        The inputs are then essentially treated as an address space that maps to a data value space in an asynchronous ROM. 
-        No explicit logic gates like NAND or XNOR are synthesized, contrary to what newcomers might expect from a "Field Programmable Gate Array".
-
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/lut_synthesis.png}
-            \captionof{figure}{LUT synthesis from user design}
-            \label{fig:lut_synthesis}
-        }
-        \vspace{1.0cm}
-
-        In the 7‑Series devices, one LUT can facilitate any 6-input boolean function, or two 5-input functions, as long as they share the same input signals.  
-        The LUT can also host two independent boolean functions of up to 3 inputs each, even when the inputs are not shared.  
-        Functions requiring more than six unique inputs are decomposed across multiple cascaded LUTs.
-        Figure \ref{fig:lut_synthesis} shows an example of where a LUT is typically synthesized in a design entry.
-
-    \textbf{FFs} \quad
-        FFs are synthesized to facilitate synchronous event-driven signal assignment. 
-        For most Verilog users, this generally means signal assignments wrapped in \texttt{always @(posedge clk)} statements. 
-        Figure \ref{fig:ff_synthesis} shows an example of where a flip flop (FF) is typically synthesized. 
-        The cell primitive \textbf{FDRE} is a type of FF and belongs to a family of D Flip Flops (DFFs) with Clock Enable (CE). 
-        \begin{itemize}
-            \item FDCE - DFF with CE and Asynchronous Clear
-            \item FDPE - DFF with CE and Asynchronous Preset
-            \item FDSE - DFF with CE and Synchronous Set
-            \item \textbf{FDRE} - DFF with CE and Synchronous Reset
-        \end{itemize}
-        In a typical HDL design, the vast majority of FFs will be synthesized as FDREs with the occasional FDSE, as it is generally good practice to keep FPGA designs synchronous. 
-        A FF BEL may also host a LATCH Cell, however, since they are generally bad practice in FPGA design, we will not consider latches in this paper.
-
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/ff_synthesis.png}
-            \captionof{figure}{FF synthesis from user design}
-            \label{fig:ff_synthesis}
-        }
-        \vspace{1.0cm}
-
-        Up to eight (8) FFs can be placed within the same SLICE, but only if they all share a common Clock-Enable (CE) net and a common Set-Reset (SR) net.
-        This is because the SLICE has only one CE pin and one SR pin to interface with general routing. 
-        The CE and SR signals from these pins are broadcast intra-Site to all FFs within.
-
-    \newpage
-    \textbf{LUT-FF Pairs} \quad
-        FPGA designs are very often modelled as a collection of Finite State Machines (FSM) like shown in Figure \ref{fig:fsm}.
-        Many times a design will also use pipelining, either to model signal buffers or shift registers, or to split up large combinational logic blocks into time slices to meet timing constraints. 
-        These common design structures result in many consecutive sections of combinational logic feeding into a vector of registers. 
-        The synthesizer naturally synthesizes these structures as consecutive pairs of LUTs feeding into FFs as shown in Figure \ref{fig:pipelining}. 
-        Figure \ref{fig:lut_ff_pair} shows an example of a synthesized LUT-FF Pair.
-
-        Shown in Figure \ref{fig:intersite_intrasite} are two possible placements for a LUT-FF Pair on the physical device. 
-        On the right, the cells are placed across different Sites, thus the only way to route the net between the cells is through general inter-site routing. 
-        On the bottom left, the cells are placed within the same Site in the same lane, taking advantage of the intra-site routing without burdening the general router with additional inter-site routing.
-        This is an important consideration to make while minimizing wirelength during placement.
-
-        \vspace{1.0cm}
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/fsm.png}
-            \captionof{figure}{Finite state machine (Moore)}
-            \label{fig:fsm}
-        }
-        \vspace{1.0cm}
-
-        {
-            \centering
-            \includegraphics[width=0.9\columnwidth]{figures/pipelining.png}
-            \captionof{figure}{Pipelining synthesized as consecutive LUT-FF pairs}
-            \label{fig:pipelining}
-        }
-
-        \vfill
-        .
-        \newcolumn
-
-        {
-            \centering
-            \includegraphics[width=0.9\columnwidth]{figures/lut_ff_pair.png}
-            \captionof{figure}{A synthesized LUT-FF Pair}
-            \label{fig:lut_ff_pair}
-        }
-        \vspace{1.0cm}
-
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/intersite_intrasite_2.png}
-            \captionof{figure}{Intrasite vs Intersite LUT-FF Placement}
-            \label{fig:intersite_intrasite}
-        }
-
-    \newpage
-    \textbf{CARRY} \quad
-        An FPGA design will also typically implement many adders, counters, subtractors, or comparators, all of which are based on binary addition. 
-        They are so ubiquitous that every that in the 7-Series architecture, every SLICE features a \emph{CARRY4} BEL -- a 4-bit carry-lookahead (CLA) adder, a much better alternative to synthesizing adders via LUTs.
-
-    \textbf{CARRY Chains} \quad
-        These CARRY4 blocks can be chained across SLICEs to implement wide adders efficiently. 
-        The CARRY4 BELs \emph{must} be chained vertically consecutively across SLICEs as the Carry-In (CI) and Carry-Out (CO) pins can only be routed this way. 
-        A CARRY4 cell may also directly connect to LUTs and FFs, and should be placed in the same Site whenever possible to minimize wirelength. 
-        Shown in figure \ref{fig:carry_cell_group} shows how a CARRY4 chain and associated LUTs and FFs can be placed across SLICEs.
-
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/carry_cell_group.png}
-            \captionof{figure}{A SLICEL with a CARRY4 cell, 4 LUT cells, and 4 FF cells placed inside as shown in the Vivado device viewer}
-            \label{fig:carry_cell_group}
-        }
-
-        {
-            \centering
-            \includegraphics[valign=c, width=4.0cm]{figures/carry_chain_placement.png}
-            \includegraphics[valign=c, width=4.0cm]{figures/carry_chain_routes_3.png}
-            \captionof{figure}{
-                A CARRY4 chain of size 3 placed across 3 SLICEs.
-                \textbf{Left:} Simplified view, \textbf{Right:} As shown in the Vivado device viewer.
-            }
-            \label{fig:carry_cell_group}
-        }
-
-        \end{multicols}
-        % \vfill
-        {
-            \centering
-            \includegraphics[valign=c, width=3.0cm]{figures/adder_synthesis.png}
-            \includegraphics[valign=c, width=14.0cm]{figures/carry_chain_edif_3.png}
-            \captionof{figure}{A synthesized carry chain of size 3 as shown in the Vivado netlist viewer}
-            \label{fig:carry_chain_edif}
-        }
-        \begin{multicols}{2}
-        \vspace{1.0cm}
-        .
-
-    \newpage
-    \subsection{DSP Slices}
-
-    \textbf{DSPs} \quad 
-        FPGAs are often used as low latency Digital Signal Processing (DSP) accelerators. 
-        Common DSP subsystems like Finite Impulse Response (FIR) filters, Fast Fourier Transform (FFTs), and convolutional neural nets (CNNs) demand fast large scale multiply-accumulate (MAC) capabilities.
-        The 7-Series arhictecture integrates DSP BELs into the logic fabric called DSP48E1 that can facilitate MAC efficiently. 
-        The architecture hierarchy for DSPs is simple compared to CLBs and SLICEs. 
-        A DSP48E1 Tile contains two DSP48E1 Sites, each containing one DSP48E1 BEL, which can host a DSP48E1 Cell.
-
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/dsp_diagram.png}
-            \captionof{figure}{
-                Basic DSP48E1 Slice Functionality
-            }
-            \label{dsp_diagram}
-        }
-        \vspace{0.5cm}
-
-    \textbf{DSP Cascades} \quad 
-        Higher level DSP functions are supported by cascading DSP48E1 slices in a DSP48E1 column.
-        Much like CARRY4 chains, DSP48E1 cascades must necessarily be placed vertically consecutively across DSP48E1 Sites. 
-        They are connected by three busses: ACOUT to ACIN, BCOUT to BCIN, PCOUT to PCIN.
-        These signal busses run directly between the vertical DSP48E1 slices without burdening the general routing resources. 
-        The ability to cascade this way provides a high-performance and low-power implementation of DSP filter functions.
-
-        \vspace{1.0cm}
-        {
-            \centering
-            \includegraphics[valign=c, width=2.5cm]{figures/dsp_cascade_placement.png}
-            \includegraphics[valign=c, width=3.0cm]{figures/dsp_cascade_routed.png}
-            \captionof{figure}{
-                A DSP48E1 cascade of size 2 placed across 2 DSP48E1 Sites.
-                \textbf{Left:} Simplified view, \textbf{Right:} As shown in the Vivado device viewer.
-            }
-        }
-
-        \end{multicols}
-        {
-            \centering
-            \includegraphics[valign=c, width=4.0cm]{figures/mult_synthesis.png}
-            \includegraphics[valign=c, width=13.0cm]{figures/dsp_synthesis.png}
-            \captionof{figure}{
-                Simple multiplier synthesis from user design.
-            }
-            \label{mult_synthesis}
-        }
-        \begin{multicols}{2}
-
-    \newpage
-
-    \subsection{Block RAM}
-        In addition to SLICEs and DSPs, the 7-Series also offers dedicated Block Random Access Memory (BRAM) BELs. 
-        These BRAMs come in two variants: \textbf{RAMB18E1} and \textbf{RAMB36E1}. 
-        \begin{itemize}
-            \item \textbf{RAMB18E1} - 
-            Has a capacity of 18 Kilobits. 
-            It can be configured as single port RAM with dimensions ranging between (1-bit wide by 16K deep) to (18-bit wide by 1024 deep). 
-            It can also be configured as a (36-bit wide by 512 deep) true simple dual port RAM. 
-            \item \textbf{RAMB36E1} - 
-            Has a capacity of 36 Kilobits. 
-            It can be configured as single port RAM with dimensions ranging between (1-bit wide by 32K deep) to (36-bit wide by 1024 deep). 
-            It can also be configured as a (72-bit wide by 512 deep) simple dual port RAM. 
-        \end{itemize}
-
-        For readability, we will abbreviate RAMB18E1 as \textbf{RAM18} and RAMB36E1 as \textbf{RAM36}.
-        One BRAM Tile contains one RAM36 Site and two RAM18 Sites. 
-        The RAM36 Site contains one RAM36 BEL while the RAM18 Site contains one RAM18 BEL.
-        The RAM18 BEL can host one RAM18 Cell, while the RAM36 BEL can host either one RAM36 Cell or two RAM18 Cells.
-
-        Like DSP cascades, BRAMs may also be cascaded together in a column to implement large memories efficiently, with some intermediate signals between them routed intra-Tile without burdening the general routing.
-        However, unlike DSPs, large memories decomposed amongst multiple BRAMs can also be routed together through general routing. 
-        Furthermore, in most design scenarios, large memories will not utilize the intra-Tile signals.
-
-        For simplicity we will not constrain large BRAMs to be cascaded together in consecutive vertical Tiles. 
-        We can essentially treat RAMB18E1 and RAMB36E1 Cells as loose, minimally constrained single cells, in contrast to the highly constrained DSP cascades and CARRY chains.
-
-        {
-            \centering
-            \includegraphics[valign=c, width=3.0cm]{figures/bram.png}
-            \includegraphics[valign=c, width=2.0cm]{figures/bram_tile.png}
-            \captionof{figure}{
-                A BRAM Tile containing two RAMB18E1 Sites and one RAMB36E1 Site.
-                \textbf{Left:} simplified view. \textbf{Right:} as seen in the device viewer, BELs highlighted in white. 
-            }
-            \label{fig:bram_tile}
-        }
-
-        {
-            \centering
-            \includegraphics[valign=c, width=4.0cm]{figures/bram_synthesis.png}
-            \includegraphics[valign=c, width=\columnwidth]{figures/bram_edif.png}
-            \captionof{figure}{
-                An example of BRAM synthesis (via inference)
-            }
-            \label{fig:bram_synthesis}
-        }
-
-    \subsection{Further Documentation}
-
-    For more in-depth details about 7-Series FPGAs, refer to the official Xilinx user guides such as:
-    \begin{itemize}
-        \item \emph{7 Series FPGAs Overview} (UG476)
-        \item \emph{7 Series FPGA Configurable Logic Block} (UG474)
-        \item \emph{7 Series Memory Resources} (UG473)
-        \item \emph{7 Series DSP48E1 Slice} (UG479)
-    \end{itemize}
-
-    This architectural context provides the necessary background for understanding how a placement algorithm should account for resource constraints and optimize performance in modern FPGA designs.
-
-\newpage
-\section{FPGA Design Flow and Toolchain}
-    \label{sec:fpga_flow_toolchain}
-
-    Modern FPGA designs require a sophisticated toolchain to bridge the gap between high-level hardware descriptions and the final bitstream used to configure the FPGA. 
-    Figure~\ref{fig:design_flow} illustrates a representative process that converts an abstract Hardware Description Language (HDL) design into a verified configuration file for a target device.
-
-    {
-        \centering
-        \includegraphics[width=0.9\columnwidth]{figures/design_flow.png}
-        \captionof{figure}{A typical FPGA design and verification workflow.}
-        \label{fig:design_flow}
-    }
-
-    \begin{enumerate}
-        \item \textbf{Design Entry:} 
-            An engineer describes the intended functionality of the digital system using a hardware description language (HDL) such as Verilog or VHDL. 
-            During this phase, the coding style can vary (behavioral, structural, dataflow, etc.), but it always aims to capture high-level behavior rather than device-specific details.
-
-        \item \textbf{Synthesis:} 
-            The synthesis tool parses the HDL source, performs logical optimizations, and maps the design onto primitive cells that suit the target FPGA technology. 
-            The output is typically a structural \texttt{netlist} (e.g., EDIF or structural Verilog) which details how the design’s logic is broken down into LUTs, FFs, and other vendor-specific cells.
-
-        \item \textbf{Placement and Routing (Implementation):} 
-            In \emph{placement}, each logical cell from the synthesized netlist is assigned to a physical location on the FPGA fabric. 
-            For instance, LUTs and FFs go into specific \emph{BELs} within the device’s CLB sites, and specialized cells such as DSPs and Block RAMs must be placed in their corresponding tile types. 
-            Next, \emph{routing} determines how signals are physically wired through the FPGA’s configurable interconnect network. 
-            Modern tools often interleave these steps (e.g., fluid-placement routing or routing-aware placement) to better meet timing and area objectives.
-
-        \item \textbf{Bitstream Generation:} 
-            After a design is fully placed, routed, and timing-closed, the toolchain produces a final \emph{bitstream} that sets the configuration of every programmable element in the FPGA. 
-            This bitstream can then be loaded onto the device, either through vendor software or via a custom programming interface.
-
-        \item \textbf{Verification:} 
-            In parallel to the design flow, simulations and testbenches validate correctness of the user's design at multiple abstraction levels. 
-            Engineers may begin with behavioral simulations, then progress to post-synthesis simulations, and finally to post-implementation simulations that incorporate estimated routing delays. 
-            With each higher level of fidelity, computational requirements grow significantly due to increasing complexity and the need to analyze more variables over time. 
-            Ensuring correct functionality and meeting timing closure at the post-implementation stage is crucial before deploying the design to hardware. 
-            Given the importance of thorough verification, many established companies dedicate one verification engineer for every design engineer.
-
-    \end{enumerate}
-
-    This workflow underscores the critical role of \textbf{placement} in bridging the netlist to a physical realization. 
-    An efficient placement algorithm can drastically reduce compile times and improve design performance, enabling broader adoption of FPGAs in application spaces that require fast design iteration.
-
-\newpage
-\section{RapidWright API}
-    \label{sec:rapidwright_api}
-
-    \textbf{RapidWright} is an open-source Java framework from AMD/Xilinx that provides direct access to the netlist and device databases used by vendor tools. 
-    This framework positions itself as an additional workflow column, allowing users to intercept or replace stages of the standard design flow with custom optimization stages (see Figure~\ref{fig:vivado_dcps}).
-
-    \begin{itemize}
-    \item \textbf{Design Checkpoints:} 
-        RapidWright leverages \texttt{.dcp} files (design checkpoints) generated at various stages of a Vivado flow. 
-        By importing a checkpoint, engineers can manipulate the netlist, placement, or routing externally, then re-export a modified checkpoint for further processing in the Vivado workflow column.
-
-    \item \textbf{Key Packages:} 
-        RapidWright revolves around three primary data model packages:
-        \begin{enumerate}
-        \item \emph{edif} -- Represents the logical netlist in an abstracted EDIF-like structure.
-        \item \emph{design} -- Contains data structures for the physical implementation (Cells, Nets, Sites, BELs, etc.).
-        \item \emph{device} -- Provides a database of the target FPGA architecture (e.g., Site coordinates, Tile definitions, routing resources).
-        \end{enumerate}
-
-    \item \textbf{Interfacing with the Netlist and Device:} 
-        An engineer can query the netlist to find specific resources (LUTs, FFs, DSPs, etc.) and then map or move them onto device sites. 
-        This level of control over backend resources is necessary for research in custom placement, advanced packing techniques, or experimental routing algorithms.
-    \end{itemize}
-
-    {
-        \centering
-        \includegraphics[width=0.8\columnwidth]{figures/vivado_dcps.png}
-        \captionof{figure}{RapidWright workflow integrating into the default Vivado design flow.}
-        \label{fig:vivado_dcps}
-    }
-
-    By exposing these low-level internals, RapidWright allows fine-grained design transformations that go beyond the standard Vivado IDE’s capabilities. 
-    Researchers can prototype new EDA strategies without needing to re-implement an entire FPGA backend from scratch, thus accelerating innovation in placement and routing methodologies.
-
-\newpage
-\section{What is a Netlist?}
-    \label{sec:netlist}
-    In its simplest form, a netlist consists of a list of every component in an electronic design and a list of nets they connect to. 
-    Here, a net, or network, denotes any group of two or more interconnected elements - whether they be transistors, logic gates, or higher-level macrocells. 
-    More precisely, nets attach to component pins, and those pins reside on cells, each of which exposing multiple ports. 
-    In this way, a netlist makes explicit both the individual building blocks of a circuit and the precise wiring that links them.
-
-    In Vivado, a Netlist can be synthesized as a Hierarchical or a Flattened netlist. 
-    Figure ~\ref{fig:hierarchical_design} shows the Verilog code for a simple HDL design using module instance hierarchy. 
-    Figure ~\ref{fig:hier_netlist} shows the design synthesized into a hierarchical netlist, while Figure ~\ref{fig:flat_netlist} shows the same design synthesized into a flat netlist. 
-    In a synthesized Netlist, the leaf cells (lowest-level cells) are necessarily primitive cells from the vendor primitives library such as LUT6, FDRE, CARRY4, DSP48E1, etc..
-
-    \end{multicols}
-    {
-        \raggedright
-        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/top_level.png}
-        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_0.png}
-        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_1.png}
-        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_2.png}
-        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_3.png}
-        \captionof{figure}{A simple HDL design with module hierarchy.}
-        \label{fig:hierarchical_design}
-    }
-    {
-        \centering
-        \includegraphics[valign=c, width=11.5cm]{figures/netlist_synth/hier_netlist.png}
-        \includegraphics[valign=c, width=6cm]{figures/netlist_synth/hier_graph.png}
-        \captionof{figure}{
-            \textbf{Left:} A hierarchical netlist consisting of LUTs and FFs.
-            \textbf{Right:} The cell hierarchy graph.
-        }
-        \label{fig:hier_netlist}
-    }
-    \vspace{0.5cm}
-    {
-        \centering
-        \includegraphics[valign=c, width=10cm]{figures/netlist_synth/flat_netlist.png}
-        \includegraphics[valign=c, width=4cm]{figures/netlist_synth/flat_graph.png}
-        \captionof{figure}{
-            \textbf{Left:} A flattened netlist consisting of LUTs and FFs.
-            \textbf{Right:} The flattened cell hierarchy graph.
-        }
-        \label{fig:flat_netlist}
-    }
-    \begin{multicols}{2}
-
-    RapidWright offers an easy way to access, read, or manipulate this netlist for a given HDL design. 
-    Shown in Figure~\ref{fig:netlist_syn}
-
-    Verilog \texttt{module}s are represented simply as
-
-    \end{multicols}
-    {
-        \centering
-        \includegraphics[width=\columnwidth]{figures/raw_edif.png}
-        \captionof{figure}{A raw EDIF Netlist with uncountably many cells}
-        \label{fig:raw_edif}
-    }
-    \begin{multicols}{2}
-
-    .
-
-\newpage
-\section{Simulated Annealing}
-    \label{sec:simulated_annealing}
-    With a basic understanding of FPGA architecture, design placement, and RapidWright, we have all the necessary pieces to implement our SA placer. 
-    Here we outline in detail each substage of our implementation: PrePacking, Packing, and Placement. 
-    Shown in Figure \ref{fig:edif_design_device} shows an overview of the placement workflow.
-
-    {
-        \centering
-        \includegraphics[width=0.9\columnwidth]{figures/edif_design_device.png}
-        \captionof{figure}{Our placement workflow}
-        \label{fig:edif_design_device}
-    }
-
-    \end{multicols}
-    {
-        \centering
-        \includegraphics[width=0.8\columnwidth]{figures/substages.png}
-        \captionof{figure}{The data classes populated at each substage: PrepackedDesign, PackedDesign, and PlacedDesign.}
-        \label{fig:substages}
-    }
-    \begin{multicols}{2}
-
-    \subsection{Prepacking}
-        \label{subsec:prepacking}
-
-        The first step in our placement flow is \textbf{prepacking}. 
-        Recall from the 7-Series architecture that there are certain cell structures that must adhere to certain placements constraints to ensure legality, and by design, to minimize wirelength. 
-        The job of the prepacker is to consolidate the cells in the EDIF Netlist into clusters or groups of clusters that naturally reflect these placement constraints.
-
-        CARRY4 chains must necessarily be placed vertically and consecutively across a column of SLICEs in ascending order. 
-        Likewise, DSP48E1 cascades must necessarily be placed vertically and consecutively across a column of DSP48E1 Sites in ascending order. 
-        A LUT-FF pair may be placed freely, but should be placed in the same lane within the same SLICE to minimize wirelength.
-
-        The raw post-synthesis netlist only tells us the list of nets and the cell ports that they connect to. 
-        It does not report the presence of any macro cell structures (CARRY4 chains, etc.). 
-        Thus, we must traverse the netlist to detect these cell structures and store that structure information in a class we will call \texttt{PrepackedDesign}.
-
-        A group of LUT-FF pairs may be placed in the same SLICE, with the constraint that the FFs must share the same Clock-Enable (CE) and Set-Reset (SR) nets.
-        Clustering LUT-FF pairs like this can reduce the redundancy of having to route the same CE and SR nets to many SLICEs by routing the nets to fewer SLICEs and connecting them to the individual FFs within via intra-Site routing, and at the same time, packing greater logic density over a smaller area of the device.
-        Recall that up to eight FFs may be placed in the same SLICE, thus theoretically, up to eight LUT-FF pairs can be placed in the same SLICE.
-
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/slicel.png}
-            \captionof{figure}{A SLICEL Site}
-            \label{fig:slicel}
-        }
-
-        Utilizing all 8 LUT-FF lanes in a SLICE can help reduce device area utilization and minimize wirelength, but \emph{too much} logic density in an area can contribute to general routing congestion.
-        Furthermore, attempting to fill all 8 lanes in a SLICE requires some careful consideration of LUT BEL constraints. 
-        The LUT BELs in a SLICE, as shown in Figure \ref{fig:slicel}, the LUTs appear to be interleaved.
-
-        Show a block diagram of a LUT6 comprising of two LUT5s.
-
-        Recall that a LUT can accommodate one 6-input boolean function or two 5-input boolean functions sharing the same inputs or any two 3-input or less boolean functions regardless of shared inputs.
-        A LUT6 Cell (a LUT with 6-inputs) will actually occupy two lanes in a SLICEL, rendering one of the FF BELs in either lane ineligible for another LUT-FF pair. 
-        The LUT BELs
-
-    \subsection{Packing}
+\subsection{Site Packing}
 \label{subsec:packing}
-
-    \subsection{Placement}
-        \label{subsec:placement}
-        Up until now we have only organized the logical \texttt{Cells} into \texttt{SiteInsts}. 
-        This is where simulated annealing actually begins.
-
-\section{Placement Results}
-    \begin{itemize}
-    \item Random Site Selection:
-    \item Midpoint Site Selection:
-    \item Hybrid: 
-        Random Site selection until progress stagnates, then use Midpoint Site selection. 
-        Randomly choose betwee Random and Midpoint Site selection each iteration. 
-    \item Cooling Schedule: 
-        Greedy. 
-        Geometric with modifiable alpha and intial temperature. 
-    \end{itemize}
-
-\section{Future Work}
-    \begin{itemize}
-    \item Different cooling schedules: linear cooling, logarithmic cooling, piecewise cooling.
-    \item Support for more macros: buffer cells, FIFO, clock generators, etc..
-    \item Support Ultrascale architecture. 
-    \item Better packing scheme. Currently pseudorandom packing. 
-    \item Inter-Site BEL swapping. 
-    \item BEL-centric placement over Site-centric placement.
-    \item Analytical placement. Global placement, legalization, detailed placement. 
-    \end{itemize}
-
-\section{Conclusion}
-    fdsfdsafdsa
-
-% \section{Analytical Placement}
-%     Introduce placement with legalization methods. \\
-%     Many AP approaches use a Global placement, Legalization, then Detailed placement approach. \\
-%     Our SA only considers legal moves via bookkeeping structures, so it has no concept of global placement or legalization. \\
-%     It is a detailed placement only approach. \\
-%     \begin{itemize}
-%         \item Talk about HeAP as an example.
-%         \item Global Placement, Legalization, then Detailed Placement approach.
-%         \item Global placement using the Bound2Bound net model from SimPL.
-%         \item Legalization Step 1: \\
-%             Find an area of the device that is overutilized (illegal) for twhich the blocks contained within must spread to a larger area. \\
-%             To obtain this overutilized area, adjacent locations (Sites, Tiles, etc.) on the device that are occupied by more than one block (SiteInst, ModuleInst, etc.) are repeatedly clustered together, until all clusters are bordered on all sides by non-overutilized locations.
-\\
-%             Then, the area (?) is expanded in both the x and y dimensions (?) until it is large enough to accommodate all blocks contained within. \\
-%             Specifically, the area is expanded until its "occupance" \(O_A\) divided by its "capacity" \(C_A\) is less than a maximum "utilization factor" $\beta$, where $\beta$ is less than or equal to 1.
-HeAP uses 0.9.
-%         \item Legalization Step 2: \\
-%             Two cuts are generated: a "source cut" and a "target cut". \\
-%             The source cut pertains to the blocks being placed (SiteInsts in home buffer). \\ 
-%             The target cut pertains to the area into which the blocks are placed (away buffer). \\
-%             The source cut splits the blocks into 2 partitions, while the target cut splits the area into 2 sub-areas, into which the blocks in each partition are spread. \\
-%             Two objectives are minimized during this process: The imbalance between the number of blocks in each partition, and the difference in the utilization of each sub-area. \\
-%             Utilization defined as the occupancy divided by the capacity of the sub-area. \\
-%             \(U_{sub-area} = \frac{O_{sub-area}}{C_{sub-area}}\) \\
-%             To generate the source cut, the cells are first sorted by their x or y location, depending on the orientation of the desired cut. \\
-%             Once the cells are sorted, the source cut generation is akin to choosing a pivot in a sorted list, where all blocks to the left of the pivot are assigned to the left/bottom partition and all blocks to the right of the pivot are assigned to the right/top partition.
-\\
-%             The target cut is an x or y cut of the area such that all blocks in each partition fit in their respective sub-areas, and such that \( | U_{sub-area_1} 
-%              - U_{sub-area_2} | \) is minimized.
-%     \end{itemize}
-
-\section{Appendix}
-
-    \begin{figure*}[b]
-        \centering
-        \includegraphics[width=\textwidth]{figures/slicem.png}
-        \caption{SLICEM Site}
-        \label{fig:slicem}
-    \end{figure*}
-
-    \begin{figure*}[t]
-        \centering
-        \includegraphics[width=\textwidth]{figures/slicel.png}
-        \caption{SLICEL Site}
-        \label{fig:slicel}
-    \end{figure*}
-
-\end{multicols}
-
-\end{document}

### 2025-05-07 latex/paper/sections/10-simulated_annealing.tex
netlist stuff

--- 2025-04-28 latex/paper/paper.tex
@@
-\documentclass{article}
-\usepackage{hyperref}
-\usepackage{enumitem}
-\usepackage{graphicx}
-\usepackage{amsmath}
-\usepackage{mathpazo}
-\usepackage{xcolor}
-\usepackage{float}
-\usepackage{listings}
-\lstset{
-  basicstyle=\ttfamily\small,
-  breaklines=true,
-  columns=fullflexible,
-  frame=single,
-  showspaces=false,
-  showtabs=false,
-  keepspaces=false
-}
-\usepackage[export]{adjustbox}
-\usepackage{multicol}
-\usepackage{caption}
-\usepackage[a4paper,
-            bindingoffset=0.2in,
-            left=0.5in,
-            right=0.5in,
-            top=0.5in,
-            bottom=0.5in,
-            footskip=.25in]{geometry}
-
-\begin{document}
-\title{MS Technical Paper: \\ Placement Algorithms for Heterogeneous FPGAs}
-\author{Brian B Cheng \\ Rutgers University Department of Electrical and Computer Engineering}
-
-\date{}
-\maketitle
-
-\begin{multicols}{2}
-
-\section{Keywords}
-\begin{itemize}
-    \item FPGA, EDA, Placement, Simulated Annealing, Optimization, RapidWright
-\end{itemize}
-
-\section{Abstract}
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\ 
-    fdsafdsafdsa. \\
-
-\section{Introduction}
-
-    Field-Programmable Gate Arrays (FPGAs) have witnessed rapid growth in capacity and versatility, driving significant advances in computer-aided design (CAD) and electronic design automation (EDA) methodologies.
-    Since the early-to-mid 2000s, the stagnation of single-processor performance relative to the rapid increase in integrated circuit sizes has led to a design productivity gap, where the computational effort for designing complex chips continues to rise.
-    FPGA CAD flows mainly encompass synthesis, placement, and routing; all of which are NP-hard problems, of which placement is one of the most time-consuming processes. 
-    Inefficienct placement strategy not only extends design times from hours to days, thereby elevating cost and reducing engineering productivity, but also limits the broader adoption of FPGAs by software engineers who expect compile times akin to those of conventional software compilers like {\tt gcc}.
-
-    For these reasons, FPGA placement remains a critical research effort even today. 
-    In this paper, we study and implement established placement methods. 
-    To do this, we use the RapidWright API, which is a semi-open-source research effort from AMD/Xilinx that enables custom solutions to FPGA design implementations and design tools that are not offered by their industry-standard FPGA environment, Vivado.
-    We implement multiple variations of simulated annealing placers for Xilinx's 7-series FPGAs, with an emphasis on minimizing total wirelength while mitigating runtime. 
-    Our implementation is organized into three consecutive substages. 
-    The \textbf{prepacking} stage involves traversing a raw EDIF netlist to identify recurring cell patterns—such as CARRY chains, DSP cascades, and LUT-FF pairs—that are critical for efficient mapping and legalization.
-    In the subsequent \textbf{packing} stage, these identified patterns, along with any remaining loose cells, are consolidated into SiteInst objects that encapsulate the FPGA’s discrete resource constraints and architectural nuances.
-    Finally, the \textbf{placement} stage employs a simulated annealing (SA) algorithm to optimally assign SiteInst objects to physical sites, aiming to minimize total wirelength while adhering to the constraints of the 7-series architecture.
-
-    Simulated annealing iteratively swaps placement objects guided by a cost function that decides which swaps should be accepted or rejected. 
-    Hill climbing is permitted by occasionally accepting moves that increase cost, in hope that such swaps may later lead to a better final solution. 
-    SA remains a popular approach in FPGA placement research due to its simplicity and robustness in handling the discrete architectural constraints of FPGA devices. 
-    While SA yields surprisingly good results given relatively simple rules, it is ultimately a heuristic and stochastic approach that explores the vast placement space by making random moves. 
-    Most of these moves will be rejected, meaning that SA must run many iterations, usually hundreds to thousands, to arrive at a desirable solution.
-
-    In the ASIC domain, where placers must handle designs with millions of cells, the SA approach has largely been abandoned in favor of analytical techniques, owing to SA's runtime and poor scalability.
-    Modern FPGA placers have also followed suit, as new legalization strategies allow FPGA placers to leverage traditionally ASIC placement algorithms and adapt them to the discrete constraints of FPGA architectures.
-    While this paper does not present a working analytical placer, it will explore ways to build upon our existing infrastructure (prepacker and packer) to replace SA with AP.
-
-    The paper first begins by elaborating on general FPGA architecture and then specifically the Xilinx 7-Series architecture. 
-    Then, the paper will elaborate on the FPGA design flow, then the role that the RapidWright API plays in the design flow. 
-    We explain in detail each of these concepts for a broader audience as they provide much needed context for FPGA placement algorithms as a concept. 
-    However, readers who are already familiar with these concepts can skip directly to the RapidWright API section \ref{sec:rapidwright_api} or to the Simulated Annealing section \ref{sec:simulated_annealing}.
-
-\section{FPGA Architecture History}
-
-    Before any work can begin on an FPGA placer, it is necessary to understand both the objects being placed and the medium in which they are placed.
-    Configurable logic devices have undergone significant evolution over the past four decades. 
-    We will briefly review the evolution of configurable logic architecture starting in the 1970s and quickly work our way up to modern day FPGA architecture.
-
-    \textbf{PLA:} \quad
-    The journey began with the Programmable Logic Array (PLA) in the early 1970s. 
-    The PLA implemented output logic using a programmable-OR and programmable-AND plane that formed a sum-of-products equation for each output through programmable fuses. 
-    Around the same time, the Programmable Array Logic (PAL) was introduced. 
-    The PAL simplified the PLA by fixing the OR gates, resulting in a fixed-OR, programmable-AND design, which sacrificed some logic flexibility to simplify its manufacture. 
-    Figure \ref{fig:pal_2} shows one such PAL architecture.
-
-    \textbf{CPLD:} \quad 
-    Later in the same decade came the Complex Programmable Logic Device (CPLD), which took the form of an array of Configurable Logic Blocks (CLBs). 
-    These CLBs were typically modified PAL blocks that included the PAL itself along with macrocells such as flip-flops, multiplexers, and tri-state buffers. 
-    The CPLD functioned as an array of PALs connected by a central programmable switch matrix and could be programmed using a hardware description language (HDL) like VHDL. 
-    Figure \ref{fig:cpld} shows one such CPLD architecture. 
-    \\
-
-    {
-        \centering
-        \includegraphics[width=\columnwidth]{figures/pal_2.png}
-        \captionof{figure}{PAL architecture with 5 inputs, 8 programmable AND gates and 4 fixed OR gates}
-        \label{fig:pal_2}
-        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/Programmable-Logic-Devices/Programmable-Array-Logic-PAL/
-        % https://www.naukri.com/code360/library/difference-between-pla-and-pal 
-    }
-
-    \newcolumn
-
-    \textbf{Homogeneous FPGA:} \quad 
-    The mid-1980s saw the introduction of homogeneous FPGAs, which were built as a grid of CLBs. 
-    Rather than using a central programmable switch matrix as in CPLDs, FPGAs adopted an island style architecture in which each CLB is surrounded on all sides by programmable routing resources, as shown in Figure \ref{fig:fpga}.
-    The first commercially viable FPGA, produced by Xilinx in 1984, featured 16 CLBs arranged in a 4x4 grid. 
-    As FPGA technology advanced, CLBs were redesigned to use lookup tables (LUTs) instead of PAL arrays for greater logic density. 
-    The capacity of an FPGA was often measured by how many logical elements or CLBs it offered, which grew from hundreds to thousands and now to hundreds of thousands of CLBs.
-    \\
-
-    {
-        \centering
-        \includegraphics[width=0.8\columnwidth]{figures/cpld.png}
-        \captionof{figure}{CPLD architecture with 4 CLBs (PAL-like blocks)}
-        \label{fig:cpld}
-        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/CPLD/Complex-Programmable-Logic-Device-CPLDs/
-    }
-    {
-        \centering
-        \includegraphics[width=\columnwidth]{figures/homogenous_fpga.png}
-        \captionof{figure}{A homogeneous island-style FPGA architecture with 16 CLBs in a grid.}
-        \label{fig:fpga}
-        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/FPGA/Field-programmable-gate-array-FPGA/
-    }
-
-    \newpage
-
-    {
-        \centering
-        \includegraphics[width=\columnwidth]{figures/heterogenous_fpga_3.jpg}
-        \captionof{figure}{A heterogeneous island-style FPGA with a mix of CLBs and macrocells.}
-        \label{fig:fpga}
-        % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/FPGA/Field-programmable-gate-array-FPGA/
-    }
-    \vspace{0.5cm}
-
-    \textbf{Heterogenous FPGA:} \quad 
-    This brings us to modern day FPGA architectures. 
-    To meet the needs of increasingly complex designs, FPGA vendors introduced heterogeneous FPGAs. 
-    In these devices, hard macros such as Block RAM (BRAM) and Digital Signal Processing (DSP) slices are integrated into the programmable logic fabric along with CLBs, like shown in Figure \ref{fig:fpga}.
-    This design enables the direct instantiation of common subsystems like memories and multipliers, without having to recreate them from scratch using CLBs. 
-    Major vendors such as Xilinx and Altera now employ heterogeneous island-style architectures in their devices. 
-    As designs become increasingly large and complex, FPGAs meet the demand by becoming increasingly heterogenous, incorporating a wider variety of hard macros into the fabric.
-
-\section{Xilinx 7-Series Architecture}
-    \label{sec:7_series}
-    The Xilinx 7-Series devices, first introduced in 2010, follow a heterogeneous island-style architecture as discussed previously. 
-    Although the 7-Series was later superseded in 2013 by the UltraScale architecture, the 7-Series remains highly relevant due to its accessibility, wide availability, and compatibility with open-source tooling.
-    Representative sub-families include Artix-7, Kintex-7, Virtex-7, and Zynq-7000, each designed with different performance and cost trade-offs but all follow the core 7-Series architecture.
-
-    Figure~\ref{fig:hierarchy_4} illustrates a high-level view of the hierarchical organization of a 7-Series FPGA. 
-    At its lowest level, the device consists of a large array of atomic components called \emph{Basic Elements of Logic}~(\textbf{BELs}). 
-    These BELs encompass look-up tables (\textbf{LUTs}), flip-flops (\textbf{FFs}), block RAMs (\textbf{BRAMs}), DSP slices (\textbf{DSPs}), and the configurable interconnect fabric. 
-    They constitute the fundamental building blocks for implementing digital circuits on the FPGA.
-    Note how the Tile arrangement is columnar, where each column will have only one Tile type.
-
-    To manage this complexity, Xilinx organizes these BELs into incrementally abstract structures. 
-    First, \textbf{BELs} are grouped into \textbf{Sites}. 
-    Each Site is embedded into a \textbf{Tile}, and Tiles are further arranged into \textbf{Clock Regions}. 
-    In some high-density devices, multiple Clock Regions may be consolidated into one or more \textbf{Super Logic Regions}~(SLRs). 
-    However, for the scope of this paper, we focus on Xilinx 7-Series devices with only a single SLR.
-
-    \end{multicols}
-    {
-        \centering
-        \includegraphics[valign=c, width=0.3\columnwidth]{figures/hierarchy_5.png}
-        \includegraphics[valign=c, width=0.6\columnwidth]{figures/hierarchy_4.png}
-        \captionof{figure}{Architecture Hierarchy of a Xilinx FPGA}
-        \label{fig:hierarchy_4}
-    }
-    \vfill
-    \begin{multicols}{2}
-
-    \newpage
-    \subsection{CLB SLICEs}
-
-    In the 7-Series architecture, the term \emph{CLB} (Configurable Logic Block) refers to a \emph{CLB Tile} that contains two \emph{SLICE} Sites. 
-    Xilinx offers two variants of SLICE Sites: \textbf{SLICEL} and \textbf{SLICEM}.
-    \begin{itemize}
-        \item Each \texttt{SLICEL} has a set of BELs including eight \texttt{LUT}s, eight \texttt{FF}s, and one \texttt{CARRY4} adder.
-The \texttt{LUT} BELs in a \texttt{SLICEL} can only host \texttt{LUT} Cells.
-        \item The \texttt{SLICEM} includes all the features of a \texttt{SLICEL} but its \texttt{LUT} BELs can host both \texttt{LUT} Cells, which are asynchronous ROM elements, or \texttt{RAM32M} Cells, which are synchronous 32-deep RAM elements.
-These cells are also referred to as \emph{Distributed RAM} in the Xilinx documentation.
-            These cells can offer an alternative to the larger, more dedicated 18K-36K \texttt{RAMB18E1} cells when RAM resources are highly utilized. 
-    \end{itemize}
-    In a typical 7-Series device, approximately 75\% of the SLICE Sites are \texttt{SLICEL}s and 25\% are \texttt{SLICEM}s. 
-    A single CLB Tile can therefore host either two \texttt{SLICEL}s or one \texttt{SLICEL} and one \texttt{SLICEM}.
-    To simplify the problem space, however, we will only consider \texttt{SLICEL}s for general logic and use the dedicated \texttt{RAMB18E1} cells to implement RAM elements.
-
-    The BELs in these SLICEs facilitate the bulk of the general programmability of the FPGA fabric. 
-    We will explain in detail the function and motivation behind these BELs.
-
-    \textbf{LUTs} \quad
-        Combinational logic is universal to all HDL designs. 
-        As the their name suggests, a Look‑Up Table (LUT) map an input value to an output value. 
-        LUTs facilitate combinational logic by acting as tiny asynchronously‑accessed ROMs whose contents are fixed when the FPGA is programmed.  
-        For any boolean function, the synthesizer precalculates the boolean output to every possible input combination and stores the resulting truth table into a LUT's static memory. 
-        The inputs are then essentially treated as an address space that maps to a data value space in an asynchronous ROM. 
-        No explicit logic gates like NAND or XNOR are synthesized, contrary to what newcomers might expect from a "Field Programmable Gate Array".
-
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/lut_synthesis.png}
-            \captionof{figure}{LUT synthesis from user design}
-            \label{fig:lut_synthesis}
-        }
-        \vspace{1.0cm}
-
-        In the 7‑Series devices, one LUT can facilitate any 6-input boolean function, or two 5-input functions, as long as they share the same input signals.  
-        The LUT can also host two independent boolean functions of up to 3 inputs each, even when the inputs are not shared.  
-        Functions requiring more than six unique inputs are decomposed across multiple cascaded LUTs.
-        Figure \ref{fig:lut_synthesis} shows an example of where a LUT is typically synthesized in a design entry.
-
-    \textbf{FFs} \quad
-        FFs are synthesized to facilitate synchronous event-driven signal assignment. 
-        For most Verilog users, this generally means signal assignments wrapped in \texttt{always @(posedge clk)} statements. 
-        Figure \ref{fig:ff_synthesis} shows an example of where a FF is typically synthesized. 
-        The cell primitive \textbf{FDRE} is a type of FF and belongs to a family of D Flip Flops (DFFs) with Clock Enable (\texttt{CE}). 
-        \begin{itemize}
-            \item \texttt{FDCE} - DFF with CE and Asynchronous Clear
-            \item \texttt{FDPE} - DFF with CE and Asynchronous Preset
-            \item \texttt{FDSE} - DFF with CE and Synchronous Set
-            \item \texttt{FDRE} - DFF with CE and Synchronous Reset
-        \end{itemize}
-        In a typical HDL design, the vast majority of FFs will be synthesized as FDREs with the occasional FDSE, as it is generally good practice to keep FPGA designs synchronous. 
-        A FF BEL may also host a LATCH Cell, however, since they are generally bad practice in FPGA design, we will not consider latches in this paper.
-
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/ff_synthesis.png}
-            \captionof{figure}{FF synthesis from user design}
-            \label{fig:ff_synthesis}
-        }
-        \vspace{1.0cm}
-
-        Up to eight (8) FFs can be placed within the same SLICE, but only if they all share a common Clock-Enable (CE) net and a common Set-Reset (SR) net.
-        This is because the SLICE has only one CE pin and one SR pin to interface with general routing. 
-        The CE and SR signals from these pins are broadcast intra-Site to all FFs within.
-
-    \newpage
-    \textbf{LUT-FF Pairs} \quad
-        FPGA designs are very often modelled as a collection of Finite State Machines (FSM) like shown in Figure \ref{fig:fsm}.
-        Many times a design will also use pipelining, either to model signal buffers or shift registers, or to split up large combinational logic blocks into time slices to meet timing constraints. 
-        These common design structures result in many consecutive sections of combinational logic feeding into a vector of registers. 
-        The synthesizer naturally synthesizes these structures as consecutive pairs of LUTs feeding into FFs as shown in Figure \ref{fig:pipelining}. 
-        Figure \ref{fig:lut_ff_pair} shows an example of a synthesized LUT-FF Pair.
-
-        Shown in Figure \ref{fig:intersite_intrasite} are two possible placements for a LUT-FF Pair on the physical device. 
-        On the right, the cells are placed across different Sites, thus the only way to route the net between the cells is through general inter-site routing. 
-        On the bottom left, the cells are placed within the same Site in the same lane, taking advantage of the intra-site routing without burdening the general router with additional inter-site routing.
-        This is an important consideration to make while minimizing wirelength during placement.
-
-        \vspace{1.0cm}
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/fsm.png}
-            \captionof{figure}{Finite state machine (Moore)}
-            \label{fig:fsm}
-        }
-        \vspace{1.0cm}
-
-        {
-            \centering
-            \includegraphics[width=0.9\columnwidth]{figures/pipelining.png}
-            \captionof{figure}{Pipelining synthesized as consecutive LUT-FF pairs}
-            \label{fig:pipelining}
-        }
-
-        \vfill
-        .
-        \newcolumn
-
-        {
-            \centering
-            \includegraphics[width=0.9\columnwidth]{figures/lut_ff_pair.png}
-            \captionof{figure}{A synthesized LUT-FF Pair}
-            \label{fig:lut_ff_pair}
-        }
-        \vspace{1.0cm}
-
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/intersite_intrasite_2.png}
-            \captionof{figure}{Intrasite vs Intersite LUT-FF Placement}
-            \label{fig:intersite_intrasite}
-        }
-
-    \newpage
-    \textbf{CARRY} \quad
-        An FPGA design will also typically implement many adders, counters, subtractors, or comparators, all of which are based on binary addition. 
-        They are so ubiquitous that every that in the 7-Series architecture, every SLICE features a \texttt{CARRY4} BEL -- a 4-bit carry-lookahead (CLA) adder, a much better alternative to synthesizing adders via LUTs.
-
-    \textbf{CARRY Chains} \quad
-        These \texttt{CARRY4} blocks can be chained across SLICEs to implement wide adders efficiently. 
-        The \texttt{CARRY4} BELs \emph{must} be chained vertically consecutively across SLICEs as the Carry-In (\texttt{CI}) and Carry-Out (\texttt{CO}) pins can only be routed this way. 
-        A \texttt{CARRY4} cell may also directly connect to LUTs and FFs, and should be placed in the same Site whenever possible to minimize wirelength. 
-        Shown in figure \ref{fig:carry_cell_group} shows how a \texttt{CARRY4} chain and associated LUTs and FFs can be placed across SLICEs.
-
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/carry_cell_group.png}
-            \captionof{figure}{A \texttt{SLICEL} with a \texttt{CARRY4} cell, 4 LUT cells, and 4 FF cells placed inside as shown in the Vivado device viewer}
-            \label{fig:carry_cell_group}
-        }
-
-        {
-            \centering
-            \includegraphics[valign=c, width=4.0cm]{figures/carry_chain_placement.png}
-            \includegraphics[valign=c, width=4.0cm]{figures/carry_chain_routes_3.png}
-            \captionof{figure}{
-                A \texttt{CARRY4} chain of size 3 placed across 3 SLICEs.
-                \textbf{Left:} Simplified view, \textbf{Right:} As shown in the Vivado device viewer.
-            }
-            \label{fig:carry_cell_group}
-        }
-
-        \end{multicols}
-        % \vfill
-        {
-            \centering
-            \includegraphics[valign=c, width=3.0cm]{figures/adder_synthesis.png}
-            \includegraphics[valign=c, width=14.0cm]{figures/carry_chain_edif_3.png}
-            \captionof{figure}{A \texttt{CARRY4} chain of size 3 as shown in the Vivado netlist viewer}
-            \label{fig:carry_chain_edif}
-        }
-        \begin{multicols}{2}
-        \vspace{1.0cm}
-        .
-
-    \newpage
-    \subsection{DSP Slices}
-
-    \textbf{DSPs} \quad 
-        FPGAs are often used as low latency Digital Signal Processing (DSP) accelerators. 
-        Common DSP subsystems like Finite Impulse Response (FIR) filters, Fast Fourier Transform (FFTs), and convolutional neural nets (CNNs) demand fast large scale multiply-accumulate (MAC) capabilities.
-        The 7-Series arhictecture integrates DSP BELs into the logic fabric called DSP48E1 that can facilitate MAC efficiently. 
-        The architecture hierarchy for DSPs is simple compared to CLBs and SLICEs. 
-        A \texttt{DSP48E1} Tile contains two \texttt{DSP48E1} Sites, each containing one \texttt{DSP48E1} BEL, which can host a \texttt{DSP48E1} Cell.
-
-        {
-            \centering
-            \includegraphics[width=\columnwidth]{figures/dsp_diagram.png}
-            \captionof{figure}{
-                Basic \texttt{DSP48E1} Slice Functionality
-            }
-            \label{dsp_diagram}
-        }
-        \vspace{0.5cm}
-
-    \textbf{DSP Cascades} \quad 
-        Higher level DSP functions are supported by cascading \texttt{DSP48E1} slices in a \texttt{DSP48E1} column.
-        Much like \texttt{CARRY4} chains, \texttt{DSP48E1} cascades must necessarily be placed vertically consecutively across \texttt{DSP48E1} Sites. 
-        They are connected by three busses: \texttt{ACOUT} to \texttt{ACIN}, \texttt{BCOUT} to \texttt{BCIN}, \texttt{PCOUT} to \texttt{PCIN}.
-        These signal busses run directly between the vertical \texttt{DSP48E1} slices without burdening the general routing resources. 
-        The ability to cascade this way provides a high-performance and low-power implementation of digital signal processing systems.
-
-        \vspace{1.0cm}
-        {
-            \centering
-            \includegraphics[valign=c, width=2.5cm]{figures/dsp_cascade_placement.png}
-            \includegraphics[valign=c, width=3.0cm]{figures/dsp_cascade_routed.png}
-            \captionof{figure}{
-                A \texttt{DSP48E1} cascade of size 2 placed across 2 \texttt{DSP48E1} Sites.
-                \textbf{Left:} Simplified view, \textbf{Right:} As shown in the Vivado device viewer.
-            }
-        }
-
-        \end{multicols}
-        {
-            \centering
-            \includegraphics[valign=c, width=4.0cm]{figures/mult_synthesis.png}
-            \includegraphics[valign=c, width=13.0cm]{figures/dsp_synthesis.png}
-            \captionof{figure}{
-                Simple multiplier synthesis from user design.
-            }
-            \label{mult_synthesis}
-        }
-        \begin{multicols}{2}
-
-    \newpage
-
-    \subsection{Block RAM}
-        In addition to SLICEs and DSPs, the 7-Series also offers dedicated Block Random Access Memory (BRAM) BELs. 
-        These BRAMs come in two variants: \textbf{RAMB18E1} and \textbf{RAMB36E1}. 
-        \begin{itemize}
-            \item \texttt{RAMB18E1} - 
-            Has a capacity of 18 Kilobits. 
-            It can be configured as single port RAM with dimensions ranging between (1-bit wide by 16K deep) to (18-bit wide by 1024 deep). 
-            It can also be configured as a (36-bit wide by 512 deep) true simple dual port RAM. 
-            \item \texttt{RAMB36E1} - 
-            Has a capacity of 36 Kilobits. 
-            It can be configured as single port RAM with dimensions ranging between (1-bit wide by 32K deep) to (36-bit wide by 1024 deep). 
-            It can also be configured as a (72-bit wide by 512 deep) simple dual port RAM. 
-        \end{itemize}
-
-        One BRAM Tile contains one \texttt{RAMB36E1} Site and two \texttt{RAMB18E1} Sites. 
-        The \texttt{RAMB36E1} Site contains one \texttt{RAMB36E1} BEL, which can host one \texttt{RAMB36E1} Cell. 
-        Likewise, the \texttt{RAMB18E1} Site contains one \texttt{RAMB18E1} BEL, which can host one \texttt{RAMB18E1} Cell.
-
-        Like DSP cascades, BRAMs may also be cascaded together in a column to implement large memories efficiently, with some intermediate signals between them routed intra-Tile without burdening the general routing.
-        However, unlike DSPs, large memories decomposed amongst multiple BRAMs can also be routed together through general routing. 
-        Furthermore, in most design scenarios, large memories will not utilize the intra-Tile signals.
-
-        To simplify the problem space, we will not constrain large BRAMs to be cascaded together in consecutive vertical Tiles. 
-        We can essentially treat \texttt{RAMB18E1} and \texttt{RAMB36E1} Cells as loose, minimally constrained single cells, in contrast to the highly constrained \texttt{DSP48E1} cascades and \texttt{CARRY4} chains.
-
-        {
-            \centering
-            \includegraphics[valign=c, width=3.0cm]{figures/bram.png}
-            \includegraphics[valign=c, width=2.0cm]{figures/bram_tile.png}
-            \captionof{figure}{
-                A BRAM Tile containing two \texttt{RAMB18E1} Sites and one \texttt{RAMB36E1} Site.
-                \textbf{Left:} simplified view. \textbf{Right:} as seen in the device viewer, BELs highlighted in white. 
-            }
-            \label{fig:bram_tile}
-        }
-
-        {
-            \centering
-            \includegraphics[valign=c, width=4.0cm]{figures/bram_synthesis.png}
-            \includegraphics[valign=c, width=\columnwidth]{figures/bram_edif.png}
-            \captionof{figure}{
-                An example of BRAM synthesis (via inference)
-            }
-            \label{fig:bram_synthesis}
-        }
-
-    \subsection{Further Documentation}
-
-    For more in-depth details about 7-Series FPGAs, refer to the official Xilinx user guides such as:
-    \begin{itemize}
-        \item \emph{7 Series FPGAs Overview} (UG476)
-        \item \emph{7 Series FPGA Configurable Logic Block} (UG474)
-        \item \emph{7 Series Memory Resources} (UG473)
-        \item \emph{7 Series DSP48E1 Slice} (UG479)
-    \end{itemize}
-
-    This architectural context provides the necessary background for understanding how a placement algorithm should account for resource constraints and optimize performance in modern FPGA designs.
-
-\newpage
-\section{FPGA Design Flow and Toolchain}
-    \label{sec:fpga_flow_toolchain}
-
-    Modern FPGA designs require a sophisticated toolchain to bridge the gap between high-level hardware descriptions and the final bitstream used to configure the FPGA. 
-    Figure~\ref{fig:design_flow} illustrates a representative process that converts an abstract Hardware Description Language (HDL) design into a verified configuration file for a target device.
-
-    {
-        \centering
-        \includegraphics[width=0.9\columnwidth]{figures/design_flow.png}
-        \captionof{figure}{A typical FPGA design and verification workflow.}
-        \label{fig:design_flow}
-    }
-
-    \begin{enumerate}
-        \item \textbf{Design Entry:} 
-            An engineer describes the intended functionality of the digital system using a hardware description language (HDL) such as Verilog or VHDL. 
-            During this phase, the coding style can vary (behavioral, structural, dataflow, etc.), but it always aims to capture high-level behavior rather than device-specific details.
-
-        \item \textbf{Synthesis:} 
-            The synthesis tool parses the HDL source, performs logical optimizations, and maps the design onto primitive cells that suit the target FPGA technology. 
-            The output is typically a structural \texttt{netlist} (e.g., EDIF or structural Verilog) which details how the design’s logic is broken down into LUTs, FFs, and other vendor-specific cells.
-
-        \item \textbf{Placement and Routing (Implementation):} 
-            In \emph{placement}, each logical cell from the synthesized netlist is assigned to a physical location on the FPGA fabric. 
-            For instance, LUTs and FFs go into specific \emph{BELs} within the device’s CLB sites, and specialized cells such as DSPs and Block RAMs must be placed in their corresponding tile types. 
-            Next, \emph{routing} determines how signals are physically wired through the FPGA’s configurable interconnect network. 
-            Modern tools often interleave these steps (e.g., fluid-placement routing or routing-aware placement) to better meet timing and area objectives.
-
-        \item \textbf{Bitstream Generation:} 
-            After a design is fully placed, routed, and timing-closed, the toolchain produces a final \emph{bitstream} that sets the configuration of every programmable element in the FPGA. 
-            This bitstream can then be loaded onto the device, either through vendor software or via a custom programming interface.
-
-        \item \textbf{Verification:} 
-            In parallel to the design flow, simulations and testbenches validate correctness of the user's design at multiple abstraction levels. 
-            Engineers may begin with behavioral simulations, then progress to post-synthesis simulations, and finally to post-implementation simulations that incorporate estimated routing delays. 
-            With each higher level of fidelity, computational requirements grow significantly due to increasing complexity and the need to analyze more variables over time. 
-            Ensuring correct functionality and meeting timing closure at the post-implementation stage is crucial before deploying the design to hardware. 
-            Given the importance of thorough verification, many established companies dedicate one verification engineer for every design engineer.
-
-    \end{enumerate}
-
-    % This workflow underscores the critical role of \textbf{placement} in bridging the netlist to a physical realization. 
-    % An efficient placement algorithm can drastically reduce compile times and improve design performance, enabling broader adoption of FPGAs in application spaces that require fast design iteration.
-
-\newpage
-\section{RapidWright API}
-    \label{sec:rapidwright_api}
-
-    \textbf{RapidWright} is an open-source Java framework from AMD/Xilinx that provides direct access to the netlist and device databases used by vendor tools. 
-    This framework positions itself as an additional workflow column, allowing users to intercept or replace stages of the standard design flow with custom optimization stages (see Figure~\ref{fig:vivado_dcps}).
-
-    \begin{itemize}
-    \item \textbf{Design Checkpoints:} 
-        RapidWright leverages \texttt{.dcp} files (design checkpoints) generated at various stages of a Vivado flow. 
-        By importing a checkpoint, engineers can manipulate the netlist, placement, or routing externally, then re-export a modified checkpoint for further processing in the Vivado workflow column.
-
-    \item \textbf{Key Packages:} 
-        RapidWright revolves around three primary data model packages:
-        \begin{enumerate}
-        \item \emph{edif} -- Represents the logical netlist in an abstracted EDIF-like structure.
-        \item \emph{design} -- Contains data structures for the physical implementation (Cells, Nets, Sites, BELs, etc.).
-        \item \emph{device} -- Provides a database of the target FPGA architecture (e.g., Site coordinates, Tile definitions, routing resources).
-        \end{enumerate}
-
-    \item \textbf{Interfacing with the Netlist and Device:} 
-        An engineer can query the netlist to find specific resources (LUTs, FFs, DSPs, etc.) and then map or move them onto device sites. 
-        This level of control over backend resources is necessary for research in custom placement, advanced packing techniques, or experimental routing algorithms.
-    \end{itemize}
-
-    {
-        \centering
-        \includegraphics[width=0.8\columnwidth]{figures/vivado_dcps.png}
-        \captionof{figure}{RapidWright workflow integrating into the default Vivado design flow.}
-        \label{fig:vivado_dcps}
-    }
-
-    By exposing these low-level internals, RapidWright allows fine-grained design transformations that go beyond the standard Vivado IDE’s capabilities. 
-    Researchers can prototype new EDA strategies without needing to re-implement an entire FPGA backend from scratch, thus accelerating innovation in placement and routing methodologies.
-
-\newpage
-\section{What is a Netlist?}
-    \label{sec:netlist}
-    In its simplest form, a netlist consists of a list of every component in an electronic design and a list of nets they connect to. 
-    Generally, a net, or network, denotes any group of two or more interconnected elements - whether they be transistors, logic gates, or in our case, logical cells. 
-    More precisely, nets attach to component pins, and those pins reside on cells, each of which exposing multiple ports. 
-    In this way, a netlist makes explicit both the individual building blocks of a circuit and the precise wiring that links them.
-
-    In Vivado, a Netlist can be synthesized as a Hierarchical or a Flattened netlist. 
-    Figure ~\ref{fig:hierarchical_design} shows the Verilog code for a simple HDL design using module instance hierarchy. 
-    Figure ~\ref{fig:hier_netlist} shows the design synthesized into a hierarchical netlist, while Figure ~\ref{fig:flat_netlist} shows the same design synthesized into a flat netlist. 
-    In a synthesized Netlist, the leaf cells (lowest-level cells) are necessarily primitive cells from the vendor primitives library such as \texttt{LUT6}, \texttt{FDRE}, \texttt{CARRY4}, \texttt{DSP48E1}, etc..
-
-    \end{multicols}
-    {
-        \raggedright
-        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/top_level.png}
-        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_0.png}
-        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_1.png}
-        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_2.png}
-        \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_3.png}
-        \captionof{figure}{A simple HDL design with module hierarchy.}
-        \label{fig:hierarchical_design}
-    }
-    {
-        \centering
-        \includegraphics[valign=c, width=11.5cm]{figures/netlist_synth/hier_netlist.png}
-        \includegraphics[valign=c, width=6cm]{figures/netlist_synth/hier_graph.png}
-        \captionof{figure}{
-            \textbf{Left:} A hierarchical netlist consisting of LUTs and FFs.
-            \textbf{Right:} The cell hierarchy graph.
-        }
-        \label{fig:hier_netlist}
-    }
-    \vspace{0.5cm}
-    {
-        \centering
-        \includegraphics[valign=c, width=10cm]{figures/netlist_synth/flat_netlist.png}
-        \includegraphics[valign=c, width=4cm]{figures/netlist_synth/flat_graph.png}
-        \captionof{figure}{
-            \textbf{Left:} A flattened netlist consisting of LUTs and FFs.
-            \textbf{Right:} The flattened cell hierarchy graph.
-        }
-        \label{fig:flat_netlist}
-    }
-    \begin{multicols}{2}
-
-    RapidWright offers an easy way to access, read, or manipulate this netlist for a given HDL design. 
-    Shown in Figure~\ref{fig:netlist_syn}
-
-    Verilog \texttt{module}s are represented simply as
-
-    \end{multicols}
-    {
-        \centering
-        \includegraphics[width=\columnwidth]{figures/raw_edif.png}
-        \captionof{figure}{A raw EDIF Netlist with uncountably many cells}
-        \label{fig:raw_edif}
-    }
-    \begin{multicols}{2}
-
-    .
-
-\newpage
 \section{Simulated Annealing}
@@
     Recall that CARRY4 chains must necessarily be placed vertically and consecutively across a column of SLICEs in ascending order. 
-        Likewise, \textttt{DSP48E1} cascades must necessarily be placed vertically and consecutively across a column of DSP48E1 Sites in ascending order. 
+    Likewise, \texttt{DSP48E1} cascades must necessarily be placed vertically and consecutively across a column of DSP48E1 Sites in ascending order. 
     A LUT-FF pair may be placed freely, but should be placed in the same lane within the same SLICE to minimize wirelength.
@@
     This is where simulated annealing actually begins.
-
-\section{Placement Results}
-    \begin{itemize}
-    \item Random Site Selection:
-    \item Midpoint Site Selection:
-    \item Hybrid: 
-        Random Site selection until progress stagnates, then use Midpoint Site selection. 
-        Randomly choose betwee Random and Midpoint Site selection each iteration. 
-    \item Cooling Schedule: 
-        Greedy. 
-        Geometric with modifiable alpha and intial temperature. 
-    \end{itemize}
-
-\section{Future Work}
-    \begin{itemize}
-    \item Different cooling schedules: linear cooling, logarithmic cooling, piecewise cooling.
-    \item Support for more macros: buffer cells, FIFO, clock generators, etc..
-    \item Support Ultrascale architecture. 
-    \item Better packing scheme. Currently pseudorandom packing. 
-    \item Inter-Site BEL swapping. 
-    \item BEL-centric placement over Site-centric placement.
-    \item Analytical placement. Global placement, legalization, detailed placement. 
-    \end{itemize}
-
-\section{Conclusion}
-    fdsfdsafdsa
-
-% \section{Analytical Placement}
-%     Introduce placement with legalization methods. \\
-%     Many AP approaches use a Global placement, Legalization, then Detailed placement approach. \\
-%     Our SA only considers legal moves via bookkeeping structures, so it has no concept of global placement or legalization. \\
-%     It is a detailed placement only approach. \\
-%     \begin{itemize}
-%         \item Talk about HeAP as an example.
-%         \item Global Placement, Legalization, then Detailed Placement approach.
-%         \item Global placement using the Bound2Bound net model from SimPL.
-%         \item Legalization Step 1: \\
-%             Find an area of the device that is overutilized (illegal) for twhich the blocks contained within must spread to a larger area. \\
-%             To obtain this overutilized area, adjacent locations (Sites, Tiles, etc.) on the device that are occupied by more than one block (SiteInst, ModuleInst, etc.) are repeatedly clustered together, until all clusters are bordered on all sides by non-overutilized locations.
-\\
-%             Then, the area (?) is expanded in both the x and y dimensions (?) until it is large enough to accommodate all blocks contained within. \\
-%             Specifically, the area is expanded until its "occupance" \(O_A\) divided by its "capacity" \(C_A\) is less than a maximum "utilization factor" $\beta$, where $\beta$ is less than or equal to 1.
-HeAP uses 0.9.
-%         \item Legalization Step 2: \\
-%             Two cuts are generated: a "source cut" and a "target cut". \\
-%             The source cut pertains to the blocks being placed (SiteInsts in home buffer). \\ 
-%             The target cut pertains to the area into which the blocks are placed (away buffer). \\
-%             The source cut splits the blocks into 2 partitions, while the target cut splits the area into 2 sub-areas, into which the blocks in each partition are spread. \\
-%             Two objectives are minimized during this process: The imbalance between the number of blocks in each partition, and the difference in the utilization of each sub-area. \\
-%             Utilization defined as the occupancy divided by the capacity of the sub-area. \\
-%             \(U_{sub-area} = \frac{O_{sub-area}}{C_{sub-area}}\) \\
-%             To generate the source cut, the cells are first sorted by their x or y location, depending on the orientation of the desired cut. \\
-%             Once the cells are sorted, the source cut generation is akin to choosing a pivot in a sorted list, where all blocks to the left of the pivot are assigned to the left/bottom partition and all blocks to the right of the pivot are assigned to the right/top partition.
-\\
-%             The target cut is an x or y cut of the area such that all blocks in each partition fit in their respective sub-areas, and such that \( | U_{sub-area_1} 
-%              - U_{sub-area_2} | \) is minimized.
-%     \end{itemize}
-
-\section{Appendix}
-
-    \begin{figure*}[b]
-        \centering
-        \includegraphics[width=\textwidth]{figures/slicem.png}
-        \caption{SLICEM Site}
-        \label{fig:slicem}
-    \end{figure*}
-
-    \begin{figure*}[t]
-        \centering
-        \includegraphics[width=\textwidth]{figures/slicel.png}
-        \caption{SLICEL Site}
-        \label{fig:slicel}
-    \end{figure*}
-
-\end{multicols}
-
-\end{document}

### 2025-05-08 latex/paper/sections/07-rapidwright_api.tex
carry chain traversal

@@
     \begin{enumerate}
-    \item \emph{edif} -- Represents the logical netlist in an abstracted EDIF-like structure.
-    \item \emph{design} -- Contains data structures for the physical implementation (Cells, Nets, Sites, BELs, etc.).
-    \item \emph{device} -- Provides a database of the target FPGA architecture (e.g., Site coordinates, Tile definitions, routing resources).
+    \item \texttt{edif} -- Represents the logical netlist in an abstracted EDIF-like structure.
+    \item \texttt{design} -- Contains data structures for the physical implementation (Cells, Nets, Sites, BELs, etc.).
+    \item \texttt{device} -- Provides a database of the target FPGA architecture (e.g., Site coordinates, Tile definitions, routing resources).
     \end{enumerate}

### 2025-05-08 latex/paper/sections/08-netlist.tex
carry chain traversal

@@
 
-RapidWright represents the logical netlist objects via the \texttt{edif} classes:
-\begin{itemize}
-    \item \texttt{EDIFNetlist}: The full logical netlist of a \texttt{Design}. 
-    \item \texttt{EDIFNet}: Represents a logical net within an \texttt{EDIFNetlist}.
-    \item \texttt{EDIFHierNet}: Combines an \texttt{EDIFNet} with a full hierarchical instance name to uniquely identify a net in a netlist.
-    \item \texttt{EDIFCell}: Represents a logical cell in an \texttt{EDIFNetlist}. 
-    \item \texttt{EDIFCellInst}: Represents an instance of an \texttt{EDIFCell}. 
-    \item \texttt{EDIFHierCellInst}: An \texttt{EDIFCellInst} with its hierarchy, described by all the \texttt{EDIFCellInst}s that sit above it within the netlist.
-    \item \texttt{EDIFPort}: Represents a port on an \texttt{EDIFCell}. 
-    \item \texttt{EDIFPortInst}: Represents an instance of a port on an \texttt{EDIFCellInst}. 
-    \item \texttt{EDIFHierPortInst}: Combines an \texttt{EDIFHierPortInst} with a full hierarchical instance name to uniquely identify a port instance in a netlist. 
-\end{itemize}
-
 \newpage
-
 \end{multicols}
@@
         \textbf{Left:} A hierarchical netlist consisting of LUTs and FFs.
-        \textbf{Right:} The cell hierarchy graph.
+        \textbf{Right:} The cell hierarchy tree.
     }
@@
     \centering
-    \includegraphics[valign=c, width=13cm]{figures/netlist_synth/flat_netlist.png}
+    \includegraphics[valign=c, width=11.5cm]{figures/netlist_synth/flat_netlist.png}
     \includegraphics[valign=c, width=4cm]{figures/netlist_synth/flat_graph.png}
@@
         \textbf{Left:} A flattened netlist consisting of LUTs and FFs.
-        \textbf{Right:} The flattened cell hierarchy graph.
+        \textbf{Right:} The flattened cell hierarchy tree.
     }
@@
 
-{
-    \centering
-    \includegraphics[valign=c, width=\columnwidth]{figures/traversal.png}
-    \captionof{figure}{\texttt{Netlist} traversal via \texttt{EDIFHierCellInst}, \texttt{EDIFHierPortInst}, and \texttt{EDIFHierNet} objects}
-    \label{fig:traversal}
-}
-\vspace{0.5cm}
+RapidWright represents the logical netlist objects via the \texttt{edif} classes:
+\begin{itemize}
+    \item \texttt{EDIFNetlist}: The full logical netlist of a \texttt{Design}. 
+    \item \texttt{EDIFNet}: Represents a logical net within an \texttt{EDIFNetlist}.
+    \item \texttt{EDIFHierNet}: Combines an \texttt{EDIFNet} with a full hierarchical instance name to uniquely identify a net in a netlist.
+    \item \texttt{EDIFCell}: Represents a logical cell in an \texttt{EDIFNetlist}. 
+    \item \texttt{EDIFCellInst}: Represents an instance of an \texttt{EDIFCell}. 
+    \item \texttt{EDIFHierCellInst}: An \texttt{EDIFCellInst} with its hierarchy, described by all the \texttt{EDIFCellInst}s that sit above it within the netlist.
+    \item \texttt{EDIFPort}: Represents a port on an \texttt{EDIFCell}. 
+    \item \texttt{EDIFPortInst}: Represents an instance of a port on an \texttt{EDIFCellInst}. 
+    \item \texttt{EDIFHierPortInst}: Combines an \texttt{EDIFHierPortInst} with a full hierarchical instance name to uniquely identify a port instance in a netlist. 
+\end{itemize}
 
@@
 Design design = Design.readCheckpoint("synth.dcp")
-EDIFNetlist = design.getNetlist();
+EDIFNetlist netlist = design.getNetlist();
 
@@
 
-\newcolumn
+{
+    \centering
+    \includegraphics[valign=c, width=\columnwidth]{figures/traversal.png}
+    \captionof{figure}{\texttt{Netlist} traversal via \texttt{EDIFHierCellInst}, \texttt{EDIFHierPortInst}, and \texttt{EDIFHierNet} objects}
+    \label{fig:traversal}
+}
+\vspace{0.5cm}
+
+\begin{lstlisting}[caption={Code Printout}]
+Net: dout[0]
+	Port: I0, Cell: dout_i_1, CellType: LUT4
+	Port: Q, Cell: m0/dout_reg, CellType: FDRE
+Net: q_1
+	Port: I2, Cell: dout_i_1, CellType: LUT4
+	Port: Q, Cell: m0/q_1_reg, CellType: FDRE
+	Port: I5, Cell: q_1_i_1, CellType: LUT6
+Net: ce
+	Port: I1, Cell: dout_i_1, CellType: LUT4
+	Port: I1, Cell: dout_i_1__0, CellType: LUT6
+	Port: I1, Cell: dout_i_1__1, CellType: LUT6
+	Port: I3, Cell: q_1_i_1, CellType: LUT6
+Net: dout[1]
+	Port: I0, Cell: dout_i_1__0, CellType: LUT6
+	Port: Q, Cell: m1/m2/dout_reg, CellType: FDRE
+Net: dout_i_1__1_n_0
+	Port: O, Cell: dout_i_1__1, CellType: LUT6
+	Port: D, Cell: m1/m3/dout_reg, CellType: FDRE
+Net: clk
+	Port: C, Cell: m0/dout_reg, CellType: FDRE
+	Port: C, Cell: m0/q_1_reg, CellType: FDRE
+	Port: C, Cell: m1/m2/dout_reg, CellType: FDRE
+	Port: C, Cell: m1/m3/dout_reg, CellType: FDRE
+Net: dout[2]
+	Port: I0, Cell: dout_i_1__1, CellType: LUT6
+	Port: Q, Cell: m1/m3/dout_reg, CellType: FDRE
+Net: <const0>
+	Port: G, Cell: GND, CellType: GND
+	Port: R, Cell: m0/dout_reg, CellType: FDRE
+	Port: R, Cell: m0/q_1_reg, CellType: FDRE
+	Port: R, Cell: m1/m2/dout_reg, CellType: FDRE
+	Port: R, Cell: m1/m3/dout_reg, CellType: FDRE
+Net: <const1>
+	Port: P, Cell: VCC, CellType: VCC
+	Port: CE, Cell: m0/dout_reg, CellType: FDRE
+	Port: CE, Cell: m0/q_1_reg, CellType: FDRE
+	Port: CE, Cell: m1/m2/dout_reg, CellType: FDRE
+	Port: CE, Cell: m1/m3/dout_reg, CellType: FDRE
+Net: dout_i_1__0_n_0
+	Port: O, Cell: dout_i_1__0, CellType: LUT6
+	Port: D, Cell: m1/m2/dout_reg, CellType: FDRE
+Net: rst
+	Port: I3, Cell: dout_i_1, CellType: LUT4
+	Port: I5, Cell: dout_i_1__0, CellType: LUT6
+	Port: I5, Cell: dout_i_1__1, CellType: LUT6
+	Port: I4, Cell: q_1_i_1, CellType: LUT6
+Net: dinc
+	Port: I2, Cell: dout_i_1__0, CellType: LUT6
+	Port: I2, Cell: dout_i_1__1, CellType: LUT6
+	Port: I2, Cell: q_1_i_1, CellType: LUT6
+Net: dinb
+	Port: I3, Cell: dout_i_1__0, CellType: LUT6
+	Port: I4, Cell: dout_i_1__1, CellType: LUT6
+	Port: I0, Cell: q_1_i_1, CellType: LUT6
+Net: dina
+	Port: I4, Cell: dout_i_1__0, CellType: LUT6
+	Port: I3, Cell: dout_i_1__1, CellType: LUT6
+	Port: I1, Cell: q_1_i_1, CellType: LUT6
+Net: q_1_i_1_n_0
+	Port: D, Cell: m0/q_1_reg, CellType: FDRE
+	Port: O, Cell: q_1_i_1, CellType: LUT6
+Net: dout_i_1_n_0
+	Port: O, Cell: dout_i_1, CellType: LUT4
+	Port: D, Cell: m0/dout_reg, CellType: FDRE
+\end{lstlisting}

### 2025-05-08 latex/paper/sections/10-simulated_annealing.tex
carry chain traversal

@@
-\section{Simulated Annealing}
+\section{Placement}
 \label{sec:simulated_annealing}
@@
 Here we outline in detail each substage of our implementation: PrePacking, Packing, and Placement. 
-Shown in Figure \ref{fig:edif_design_device} shows an overview of the placement workflow.
+Shown in Figure \ref{fig:edif_design_device} is an overview of the placement workflow. 
+Figure \ref{fig:substages} shows the data structures of RapidWright objects that are populated at each stage.
 
@@
 
-    The raw post-synthesis netlist only tells us the list of nets and the cell ports that they connect to. 
+The raw EDIF netlist only tells us the list of nets and the cell ports that they connect to. 
 It does not report the presence of any macro cell structures (CARRY4 chains, etc.). 
@@
 Utilizing all 8 LUT-FF lanes in a SLICE can help reduce device area utilization and minimize wirelength, but \emph{too much} logic density in an area can contribute to general routing congestion. 
-    Furthermore, attempting to fill all 8 lanes in a SLICE requires some careful consideration of LUT BEL constraints. 
-    The LUT BELs in a SLICE, as shown in Figure \ref{fig:slicel}, the LUTs appear to be interleaved.
+Furthermore, attempting to fill all 8 lanes in a SLICE requires meticulous adherence to conditional LUT constraints. 
+Recall that a LUT can accommodate one 6-input boolean function or two 5-input boolean functions sharing the same inputs or any two 3-input or less boolean functions regardless of shared inputs. 
+A LUT6 Cell (a LUT with 6-inputs) will actually occupy two LUT-FF lanes in a SLICEL, rendering one of the FF BELs in either lane ineligible for another LUT-FF pair. 
+To simplify the problem space, we will only fill up to four LUT-FF lanes in any given SLICE.
 
-    Show a block diagram of a LUT6 comprising of two LUT5s.
+\begin{lstlisting}[language=java, caption={Netlist extraction and traversal}, label={lst:netlist_extract}]
+Design design = Design.readCheckpoint("synth.dcp")
+EDIFNetlist netlist = design.getNetlist();
+List<EDIFCellInst> ecis = netlist.getAllLeafCellInstances();
 
-    Recall that a LUT can accommodate one 6-input boolean function or two 5-input boolean functions sharing the same inputs or any two 3-input or less boolean functions regardless of shared inputs. 
-    A LUT6 Cell (a LUT with 6-inputs) will actually occupy two lanes in a SLICEL, rendering one of the FF BELs in either lane ineligible for another LUT-FF pair. 
-    The LUT BELs
+// Select only the carry cells.
+for (EDIFCellInst eci : ecis) {
+    if (eci.getCellName() == "CARRY4")
+        carryCells.add(eci);
+}
+
+// Find and remove carry chains until the list is empty
+while (!carryCells.isEmpty()) {
+    // Arbitrarily set "currentCell" pointer to a cell in the list
+    EDIFCellInst currentCell = carryCells.get(0);
+
+    // Find this carry chain anchor.
+    // Traverse the Carry-In (CI) to Carry-Out (CO) nets.
+    // Anchor is found when net on the CI Port is Ground.
+    while (true) {
+        // Access the CI port on this cell.
+        EDIFPortInst sinkPort = currentCell.getPortInst("CI");
+        // Access the net on this CI port.
+        EDIFNet net = sinkPort.getNet();
+        if (net.isGND())
+            // Found this chain anchor!
+            break;
+        // Get all ports on this net.
+        List<EDIFPortInst> netPorts = net.getPortInsts();
+        for (EDIFPortInst netPort : netPorts) {
+            // Access the port belonging to another carry cell.
+            EDIFCellInst sourceCell = netPort.getCellInst();
+            if (sourceCell.getCellName() == "CARRY4") {
+                // Move the "currentCell" pointer
+                currentCell = sourceCell;
+            }
+        }
+    }
+
+    // Now we have the chain anchor as currentCell.
+    // Now traverse in the opposite direction to find the chain tail.
+    // Tail is found when the CO Port is null.
+    // Collect the chain cells into an ordred list.
+    List<EDIFCellInst> currentChain = new ArrayList<>();
+    while (true) {
+        EDIFPortInst sourcePort = currentCell.getPortInst("CO[3]");
+        if (sourcePort == null)
+            // Found this chain's tail!
+            break;
+        EDIFNet net = sourcePort.getNet();
+        List<EDIFPortInst> sinkPorts = net.getPortInsts();
+        for (EDIFPortInst sinkPort : sinkPorts) {
+            EDIFCellInst sinkCell = sinkPort.getCellInst();
+            if (sinkCell.getCellName() == "CARRY4") {
+                currentCell = sinkCell;
+                // Add the cell to the chain list.
+                currentChain.add(currentCell);
+            }
+        }
+    }
+
+    // Add currentChain to the list of chains
+    carryChains.add(currentChain);
+}
+\end{lstlisting}
 

### 2025-05-08 latex/paper/sections/05-7series_architecture.tex
rw hello world
finish prepacking carry chains

@@
 
-\vspace{1.0cm}
 {
@@
 }
-\vspace{1.0cm}
-
 {
@@
 }
-
-\vfill
-.
-\newcolumn
-
 {
@@
 }
+{
+    \centering
+    \includegraphics[width=\columnwidth]{figures/slicel.png}
+    \captionof{figure}{A SLICEL Site}
+    \label{fig:slicel}
+}
+{
+    \centering
+    \includegraphics[width=0.5\columnwidth]{figures/lut6.png}
+    \captionof{figure}{A \texttt{LUT6} Cell is hosted using two \texttt{LUT5} BELs.}
+    \label{fig:lut6}
+}
 
-\newpage
+A group of LUT-FF pairs may be placed in the same SLICE, with the constraint that the FFs must share the same Clock-Enable (\texttt{CE}) and Set-Reset (\texttt{SR}) nets.
+Clustering LUT-FF pairs like this can reduce the redundancy of having to route the same CE and SR nets to many SLICEs by routing the nets to fewer SLICEs and connecting them to the individual FFs within via intra-Site routing, and at the same time, packing greater logic density over a smaller area of the device.
+Recall that up to eight FFs may be placed in the same SLICE, thus theoretically, up to eight LUT-FF pairs can be placed in the same SLICE.
+
+Utilizing all 8 LUT-FF lanes in a SLICE can help reduce device area utilization and minimize wirelength, but \emph{too much} logic density in an area can contribute to general routing congestion. 
+Furthermore, attempting to fill all 8 lanes in a SLICE requires meticulous adherence to conditional LUT constraints. 
+Recall that a LUT can accommodate one 6-input boolean function or two 5-input boolean functions sharing the same inputs or any two 3-input or less boolean functions regardless of shared inputs. 
+A \texttt{LUT6} Cell (a LUT with 6-inputs) will actually occupy two LUT-FF lanes in a SLICE, rendering one of the FF BELs in either lane ineligible for another LUT-FF pair. 
+Figure \ref{fig:slicel} hints at this by depicting pairs of LUTs stacked ontop of one another. 
+The LUTs individually can host up to a \texttt{LUT5}, but combined together can host one \texttt{LUT6}, like shown in Figure \ref{fig:lut6}.
+To simplify the problem space, we will only fill up to four LUT-FF lanes in any given SLICE.
+
 \textbf{CARRY} \quad
@@
 \begin{multicols}{2}
-\vspace{1.0cm}
-.
 
-\newpage
 \subsection{DSP Slices}
@@
 \textbf{DSP Cascades} \quad 
-Higher level DSP functions are supported by cascading \texttt{DSP48E1} slices in a \texttt{DSP48E1} column.
+Wider DSP functions are supported by cascading \texttt{DSP48E1} slices in a \texttt{DSP48E1} column.
 Much like \texttt{CARRY4} chains, \texttt{DSP48E1} cascades must necessarily be placed vertically consecutively across \texttt{DSP48E1} Sites. 

### 2025-05-08 latex/paper/sections/08-netlist.tex
finish prepacking carry chains

@@
     \centering
-    \includegraphics[valign=c, width=11.5cm]{figures/netlist_synth/flat_netlist.png}
+    \includegraphics[valign=c, width=13.0cm]{figures/netlist_synth/flat_netlist.png}
     \includegraphics[valign=c, width=4cm]{figures/netlist_synth/flat_graph.png}
@@
     \item \texttt{EDIFPortInst}: Represents an instance of a port on an \texttt{EDIFCellInst}. 
-    \item \texttt{EDIFHierPortInst}: Combines an \texttt{EDIFHierPortInst} with a full hierarchical instance name to uniquely identify a port instance in a netlist. 
+    \item \texttt{EDIFHierPortInst}: Combines an \texttt{EDIFPortInst} with a full hierarchical instance name to uniquely identify a port instance in a netlist. 
 \end{itemize}
 
-One of RapidWright's most powerful features is netlist traversal and manipulation via the \texttt{edif} classes.
-A netlist can be easily extracted from a \texttt{.dcp} design checkpoint and traversed like the following:
+Using these classes and their associated methods, we can traverse the logical netlist (\texttt{EDIFNetlist}) and analyze or manipulate it as we see fit. 
+A netlist can be easily extracted from a \texttt{.dcp} design checkpoint file and traversed like the following example:
 
@@
     \includegraphics[valign=c, width=\columnwidth]{figures/traversal.png}
-    \captionof{figure}{\texttt{Netlist} traversal via \texttt{EDIFHierCellInst}, \texttt{EDIFHierPortInst}, and \texttt{EDIFHierNet} objects}
+    \captionof{figure}{\texttt{Netlist} traversal via the \texttt{EDIFCellInst}, \texttt{EDIFPortInst}, and \texttt{EDIFNet} classes}
     \label{fig:traversal}

### 2025-05-08 latex/paper/sections/10-simulated_annealing.tex
finish prepacking carry chains

--- 2025-05-07 latex/paper/sections/10-simulated_annealing.tex
@@
-\section{Simulated Annealing}
+\section{Placement}
+\end{multicols}
+{
+    \centering
+    \includegraphics[width=0.8\columnwidth]{figures/substages.png}
+    \captionof{figure}{The data classes populated at each substage: \texttt{PrepackedDesign}, \texttt{PackedDesign}, and \texttt{PlacedDesign}.}
+    \label{fig:substages}
+}
+\begin{multicols}{2}
 \label{sec:simulated_annealing}
@@
 Here we outline in detail each substage of our implementation: PrePacking, Packing, and Placement. 
-Shown in Figure \ref{fig:edif_design_device} shows an overview of the placement workflow.
+Shown in Figure \ref{fig:edif_design_device} is an overview of the placement workflow. 
+Figure \ref{fig:substages} shows the data structures of RapidWright objects that are populated at each stage: \texttt{PrepackedDesign}, which is a group of data structures of \texttt{EDIFHierCellInst}s, \texttt{PackedDesign}, which is a group fo data structures around \texttt{SiteInst}s, and finally, \texttt{PlacedDesign}, which is simply captured by the final RapidWright \texttt{Design} object.
 
+\vspace{0.5cm}
 {
@@
 }
-
-\end{multicols}
-{
-    \centering
-    \includegraphics[width=0.8\columnwidth]{figures/substages.png}
-    \captionof{figure}{The data classes populated at each substage: PrepackedDesign, PackedDesign, and PlacedDesign.}
-    \label{fig:substages}
-}
-\begin{multicols}{2}
 
@@
 
-    Recall that CARRY4 chains must necessarily be placed vertically and consecutively across a column of SLICEs in ascending order. 
+Recall that \texttt{CARRY4} chains must necessarily be placed vertically and consecutively across a column of SLICEs in ascending order. 
 Likewise, \texttt{DSP48E1} cascades must necessarily be placed vertically and consecutively across a column of DSP48E1 Sites in ascending order. 
@@
 
-    The raw post-synthesis netlist only tells us the list of nets and the cell ports that they connect to. 
-    It does not report the presence of any macro cell structures (CARRY4 chains, etc.). 
+The raw EDIF netlist only tells us the list of nets and the cell ports that they connect to. 
+It does not report the presence of any multi-cell structures (\texttt{CARRY4} chains, etc.). 
 Thus, we must traverse the netlist to detect these cell structures and store that structure information in a class we will call \texttt{PrepackedDesign}.
 
-    A group of LUT-FF pairs may be placed in the same SLICE, with the constraint that the FFs must share the same Clock-Enable (CE) and Set-Reset (SR) nets.
-    Clustering LUT-FF pairs like this can reduce the redundancy of having to route the same CE and SR nets to many SLICEs by routing the nets to fewer SLICEs and connecting them to the individual FFs within via intra-Site routing, and at the same time, packing greater logic density over a smaller area of the device.
-    Recall that up to eight FFs may be placed in the same SLICE, thus theoretically, up to eight LUT-FF pairs can be placed in the same SLICE.
+The code snippet \ref{lst:carry_chains} shows how one can detect and collect these \texttt{CARRY4} chains using RapidWright. 
+We first collect the cells in the design that are of type \texttt{CARRY4}, then iteratively traverse their Carry-Out (\texttt{CO}) to Carry-In (\texttt{CI}) nets to find incident \texttt{CARRY4} cells.
+Each \texttt{CARRY4} chain has an anchor cell and a tail cell.
+The anchor is found when the \texttt{CI} net connects to Ground (\texttt{GND}), while the tail is found when the \texttt{CO} port is null.
 
+Similarly, \texttt{DSP48E1} cascades can be found and collected by traversing the \texttt{PCOUT} \texttt{ACOUT} and \texttt{BCOUT} nets.
+\texttt{LUT-FF} pairs can be found by inspecting the LUT output (O) net and checking for FF input (DI) ports. 
+We detect and collect these cell structures and consolidate them into our \texttt{PrepackedDesign} object in preparation for the following Packing stage.
+
+\end{multicols}
 {
     \centering
-        \includegraphics[width=\columnwidth]{figures/slicel.png}
-        \captionof{figure}{A SLICEL Site}
-        \label{fig:slicel}
+    \includegraphics[width=0.8\columnwidth]{figures/carry_chain_traversal.png}
+    \captionof{figure}{A netlist with two \texttt{CARRY4} chains, each of size 3}
+    \label{fig:carry_chain_traversal}
+}
+\begin{multicols}{2}
+
+\begin{lstlisting}[caption={Code Printout}]
+Anchor Cell: adder2/sum_reg[3]_i_1, CellType: CARRY4
+	Cell: adder2/sum_reg[7]_i_1, CellType: CARRY4
+	Cell: adder2/sum_reg[11]_i_1, CellType: CARRY4
+Anchor Cell: adder1/sum_reg[3]_i_1, CellType: CARRY4
+	Cell: adder1/sum_reg[7]_i_1, CellType: CARRY4
+	Cell: adder1/sum_reg[8]_i_1, CellType: CARRY4
+\end{lstlisting}
+
+\begin{lstlisting}[language=java, caption={Finding and storing carry chains.}, label={lst:carry_chains}]
+Design design = Design.readCheckpoint("synth.dcp")
+EDIFNetlist netlist = design.getNetlist();
+List<EDIFCellInst> ecis = netlist.getAllLeafCellInstances();
+
+// Select only the carry cells.
+List<EDIFCellInst> carryCells = new ArrayList<>();
+for (EDIFCellInst eci : ecis) {
+    if (eci.getCellName().contains("CARRY4"))
+        carryCells.add(eci);
 }
 
-    Utilizing all 8 LUT-FF lanes in a SLICE can help reduce device area utilization and minimize wirelength, but \emph{too much} logic density in an area can contribute to general routing congestion. 
-    Furthermore, attempting to fill all 8 lanes in a SLICE requires some careful consideration of LUT BEL constraints. 
-    The LUT BELs in a SLICE, as shown in Figure \ref{fig:slicel}, the LUTs appear to be interleaved.
+// Find and remove carry chains until the list is empty
+List<List<EDIFCellInst>> carryChains = new ArrayList<>();
+while (!carryCells.isEmpty()) {
+    // Arbitrarily set "currentCell" pointer to a cell in the list
+    EDIFCellInst currentCell = carryCells.get(0);
 
-    Show a block diagram of a LUT6 comprising of two LUT5s.
+    // Find this carry chain anchor.
+    // Traverse the Carry-In (CI) to Carry-Out (CO) nets.
+    // Anchor is found when net on the CI Port is Ground.
+    while (true) {
+        System.out.println(currentCell);
+        // Access the CI port on this cell.
+        EDIFPortInst sinkPort = currentCell.getPortInst("CI");
+        // Access the net on this CI port.
+        EDIFNet net = sinkPort.getNet();
+        if (net.isGND()) {
+            // Found this chain anchor!
+            break;
+        }
+        // Get all ports on this net.
+        List<EDIFPortInst> netPorts = net.getPortInsts();
+        for (EDIFPortInst netPort : netPorts) {
+            // Access the port belonging to another carry cell.
+            EDIFCellInst sourceCell = netPort.getCellInst();
+            if (sourceCell.getCellName().equals("CARRY4")) {
+                // Move the "currentCell" pointer
+                currentCell = sourceCell;
+                break;
+            }
+        }
+    }
 
-    Recall that a LUT can accommodate one 6-input boolean function or two 5-input boolean functions sharing the same inputs or any two 3-input or less boolean functions regardless of shared inputs. 
-    A LUT6 Cell (a LUT with 6-inputs) will actually occupy two lanes in a SLICEL, rendering one of the FF BELs in either lane ineligible for another LUT-FF pair. 
-    The LUT BELs
+    // Now we have the chain anchor as currentCell.
+    // Now traverse in the opposite direction to find the chain tail.
+    // Tail is found when the CO Port is null.
+    // Collect the chain cells into an ordred list.
+    List<EDIFCellInst> currentChain = new ArrayList<>();
+    currentChain.add(currentCell);
+    while (true) {
+        System.out.println("while 2");
+        EDIFPortInst sourcePort = currentCell.getPortInst("CO[3]");
+        if (sourcePort == null) {
+            // Found this chain's tail!
+            break;
+        }
+        EDIFNet net = sourcePort.getNet();
+        List<EDIFPortInst> netPorts = net.getPortInsts();
+        for (EDIFPortInst netPort : netPorts) {
+            EDIFCellInst sinkCell = netPort.getCellInst();
+            if (netPort.getName().equals("CI") &&
+                    sinkCell.getCellName().equals("CARRY4")) {
+                currentCell = sinkCell;
+                // Add the cell to the chain list.
+                currentChain.add(currentCell);
+                break;
+            }
+        }
+    }
+    System.out.println(currentChain.size());
+    // Add currentChain to the list of chains
+    carryChains.add(currentChain);
+    // Remove currentChain from the list of cells
+    carryCells.removeAll(currentChain);
+} // end while()
+return carryChains;
+\end{lstlisting}
 

### 2025-05-10 latex/paper/sections/05-7series_architecture.tex
start packing

@@
 They constitute the fundamental building blocks for implementing digital circuits on the FPGA.
-Note how the Tile arrangement is columnar, where each column will have only one Tile type.
+Note how the Tile arrangement is columnar such that column consists of only one Tile type.
 
@@
 \end{itemize}
-In a typical HDL design, the vast majority of FFs will be synthesized as FDREs with the occasional FDSE, as it is generally good practice to keep FPGA designs synchronous. 
-A FF BEL may also host a LATCH Cell, however, since they are generally bad practice in FPGA design, we will not consider latches in this paper.
+In a typical HDL design, the vast majority of FFs will be synthesized as \texttt{FDRE}s with the occasional \texttt{FDSE}, as it is generally good practice to keep FPGA designs synchronous. 
+A FF BEL may also host a \texttt{LATCH} Cell, however, since they are generally bad practice in FPGA design, we will not consider latches in this paper.
 
@@
 
-This architectural context provides the necessary background for understanding how a placement algorithm should account for resource constraints and optimize performance in modern FPGA designs.
+This architectural context provides the necessary background for understanding how a placement algorithm should account for resource and placement constraints and optimize wirelength in modern FPGA architectures.

### 2025-05-10 latex/paper/sections/08-netlist.tex
start packing

@@
     \item \texttt{EDIFNetlist}: The full logical netlist of a \texttt{Design}. 
-    \item \texttt{EDIFNet}: Represents a logical net within an \texttt{EDIFNetlist}.
+    \item \texttt{EDIFNet}: A logical net within an \texttt{EDIFNetlist}.
     \item \texttt{EDIFHierNet}: Combines an \texttt{EDIFNet} with a full hierarchical instance name to uniquely identify a net in a netlist.
-    \item \texttt{EDIFCell}: Represents a logical cell in an \texttt{EDIFNetlist}. 
-    \item \texttt{EDIFCellInst}: Represents an instance of an \texttt{EDIFCell}. 
+    \item \texttt{EDIFCell}: A logical cell in an \texttt{EDIFNetlist}. 
+    \item \texttt{EDIFCellInst}: An instance of an \texttt{EDIFCell}. 
     \item \texttt{EDIFHierCellInst}: An \texttt{EDIFCellInst} with its hierarchy, described by all the \texttt{EDIFCellInst}s that sit above it within the netlist.
-    \item \texttt{EDIFPort}: Represents a port on an \texttt{EDIFCell}. 
-    \item \texttt{EDIFPortInst}: Represents an instance of a port on an \texttt{EDIFCellInst}. 
+    \item \texttt{EDIFPort}: A port on an \texttt{EDIFCell}. 
+    \item \texttt{EDIFPortInst}: An instance of a port on an \texttt{EDIFCellInst}. 
     \item \texttt{EDIFHierPortInst}: Combines an \texttt{EDIFPortInst} with a full hierarchical instance name to uniquely identify a port instance in a netlist. 
@@
 Using these classes and their associated methods, we can traverse the logical netlist (\texttt{EDIFNetlist}) and analyze or manipulate it as we see fit. 
-A netlist can be easily extracted from a \texttt{.dcp} design checkpoint file and traversed like the following example:
+A netlist can be easily extracted from a \texttt{.dcp} design checkpoint file and traversed like shown in Listing \ref{lst:netlist_extract}. 
+This is performed on the same design shown in figure \ref{fig:flat_netlist}.
 

### 2025-05-10 latex/paper/sections/10-simulated_annealing.tex
start packing

@@
 Shown in Figure \ref{fig:edif_design_device} is an overview of the placement workflow. 
-Figure \ref{fig:substages} shows the data structures of RapidWright objects that are populated at each stage: \texttt{PrepackedDesign}, which is a group of data structures of \texttt{EDIFHierCellInst}s, \texttt{PackedDesign}, which is a group fo data structures around \texttt{SiteInst}s, and finally, \texttt{PlacedDesign}, which is simply captured by the final RapidWright \texttt{Design} object.
+Figure \ref{fig:substages} shows the data structures of RapidWright objects that are populated at each stage: \texttt{PrepackedDesign}, which is a group of data structures around \texttt{EDIFHierCellInst}s, \texttt{PackedDesign}, which is a group fo data structures around \texttt{SiteInst}s, and finally, \texttt{PlacedDesign}, which is simply captured by the final RapidWright \texttt{Design} object.
 
@@
 The first step in our placement flow is \textbf{prepacking}. 
-Recall from the 7-Series architecture that there are certain cell structures that must adhere to certain placements constraints to ensure legality, and by design, to minimize wirelength. 
-The job of the prepacker is to traverse the raw EDIF netlist, detect these cell structures, and consolidate these cells into clusters or groups of clusters that naturally reflect these placement constraints.
+Recall from the 7-Series architecture that there are certain multi-cell structures that must adhere to certain placements constraints to ensure legality, and by design, to minimize wirelength. 
+The job of the prepacker is to traverse the raw EDIF netlist, detect these multi-cell structures, and consolidate these cells into clusters or groups of clusters that naturally reflect these placement constraints.
 
@@
 It does not report the presence of any multi-cell structures (\texttt{CARRY4} chains, etc.). 
-Thus, we must traverse the netlist to detect these cell structures and store that structure information in a class we will call \texttt{PrepackedDesign}.
+Thus, we must traverse the netlist to detect these multi-cell structures and store that structure information in a class we will call \texttt{PrepackedDesign}.
 
-The code snippet \ref{lst:carry_chains} shows how one can detect and collect these \texttt{CARRY4} chains using RapidWright. 
+The code snippet in \ref{lst:carry_chains} shows how one can detect and collect these \texttt{CARRY4} chains using RapidWright. 
 We first collect the cells in the design that are of type \texttt{CARRY4}, then iteratively traverse their Carry-Out (\texttt{CO}) to Carry-In (\texttt{CI}) nets to find incident \texttt{CARRY4} cells.
@@
 The anchor is found when the \texttt{CI} net connects to Ground (\texttt{GND}), while the tail is found when the \texttt{CO} port is null. 
+We can further detect if there are \texttt{LUT}s or \texttt{FF}s connected to the \texttt{CARRY4} cell and store that information in a data structure we will call \texttt{CarryCellGroup} as defined in figure \ref{fig:substages}.
+This will help us in knowing which cells can be packed together into the same \texttt{Site} in the subsequent stages.
 
@@
 \texttt{LUT-FF} pairs can be found by inspecting the LUT output (O) net and checking for FF input (DI) ports. 
-We detect and collect these cell structures and consolidate them into our \texttt{PrepackedDesign} object in preparation for the following Packing stage.
+We can bucket these LUT-FF pairs by finding the set of unique \texttt{CE} \texttt{SR} net pairs to know which group of LUT-FF pairs can be placed within the same \texttt{Site}. 
+We detect and collect these multi-cellular structures and consolidate them into our \texttt{PrepackedDesign} object in preparation for the following Packing stage.
 
@@
 } // end while()
-return carryChains;
+
+// Print out the carry chains. 
+for (List<EDIFCellInst> chain : chains) {
+    for (int i = 0; i < chain.size(); i++) {
+        EDIFCellInst carry = chain.get(i);
+        if (i == 0) {
+            writer.write("\nAnchor Cell: " + carry.getName() + 
+                ", CellType: " + carry.getCellName());
+        } else {
+            writer.write("\n\tCell: " + carry.getName() + 
+                ", CellType: " + carry.getCellName());
+        }
+    }
+}
 \end{lstlisting}
@@
 \label{subsec:packing}
+Now that we have our \texttt{PrepackedDesign} object keeping track of multi-cell structures on the \texttt{edif} level, we can start packing them into \texttt{SiteInst} objects on the \texttt{design} level.
+A \texttt{SiteInst} object can be created and its BELs populated by existing \texttt{EDIFHierCellInst} objects. 
+A \texttt{SiteInst} is necessarily placed on a specific \texttt{Site} upon creation. 
+Thus, the packer necessarily gives all \texttt{SiteInst}s an initial placement before randomization via simulated annealing in the following placement stage. 
+In packing, will simply assign \texttt{SiteInst}s pseudorandomly onto device \texttt{Site}s on a first-come-first-serve basis by coordinate order (first SiteInst onto \texttt{SLICEL\_X0Y0}, second \texttt{SiteInst} onto \texttt{SLICEL\_X0Y1}, etc.).
+
+\end{multicols}
+\begin{lstlisting}[caption=\texttt{SiteInst} construction and BEL population.]
+Constructor: 
+    SiteInst(String name, Design design, SiteTypeEnum type, Site site)
+
+Usage:
+    SiteInst si = new SiteInst(<name>, <design>, <type>, <site>);
+    si.createCell(<EDIFHierCellInst>, <BEL>);
+\end{lstlisting}
+
+\begin{lstlisting}[language=java, caption={Packing DSPCascades into \texttt{DSP48E1} \texttt{SiteInst}s.}, label={lst:carry_chains}]
+private List<List<SiteInst>> packDSPCascades(List<List<EDIFHierCellInst>> EDIFDSPCascades) throws IOException {
+    List<List<SiteInst>> siteInstCascades = new ArrayList<>();
+    writer.write("\n\nPacking DSP Cascades... (" + EDIFDSPCascades.size() + ")");
+    for (List<EDIFHierCellInst> cascade : EDIFDSPCascades) {
+        List<SiteInst> siteInstCascade = new ArrayList<>();
+        // each DSP tile has 2 DSP sites on the Zynq 7000
+        writer.write("\n\tCascade Size: (" + cascade.size() + "), Cascade Anchor: "
+                + cascade.get(0).getFullHierarchicalInstName());
+        // Select a Site location on the device. 
+        Site anchorSite = selectDSPAnchorSite(cascade.size());
+        SiteTypeEnum siteType = SiteTypeEnum.DSP48E1;
+        for (int i = 0; i < cascade.size(); i++) {
+            writer.write("\n\t\tPlacing DSP: " + cascade.get(i).getFullHierarchicalInstName());
+            Site site = (i == 0) ? anchorSite
+                    : device.getSite("DSP48_X" + anchorSite.getInstanceX() + "Y" + (anchorSite.getInstanceY() + i));
+            // Create the SiteInst object
+            SiteInst si = new SiteInst(cascade.get(i).getFullHierarchicalInstName(), design, siteType, site);
+            // Populate the SiteInst with a DSP Cell. 
+            si.createCell(cascade.get(i), si.getBEL("DSP48E1"));
+            occupiedSites.get(siteType).add(site);
+            availableSites.get(siteType).remove(site);
+            si.routeSite();
+            siteInstCascade.add(si);
+        }
+        siteInstCascades.add(siteInstCascade);
+    }
+    return siteInstCascades;
+} // end packDSPCascades()
+\end{lstlisting}
+
+\begin{lstlisting}[language=java, caption={Packing LUTGroups into \texttt{SLICEL} \texttt{SiteInst}s}]
+protected String[] LUT6_BELS = new String[] { "A6LUT", "B6LUT", "C6LUT", "D6LUT" };
+
+private List<SiteInst> packLUTGroups(List<List<EDIFHierCellInst>> LUTGroups) throws IOException {
+    List<SiteInst> LUTSiteInsts = new ArrayList<>();
+    writer.write("\n\nPacking LUT Groups...");
+    for (List<EDIFHierCellInst> group : LUTGroups) {
+        Site selectedSite = selectCLBSite();
+        SiteInst si = new SiteInst(group.get(0).getFullHierarchicalInstName(), design,
+                selectedSite.getSiteTypeEnum(), selectedSite);
+        for (int i = 0; i < group.size(); i++) {
+            si.createCell(group.get(i), si.getBEL(LUT6_BELS[i]));
+        }
+        si.routeSite();
+        // for whatever reason, this does not activate DUSED PIP for
+        si.addSitePIP(si.getSitePIP("DUSED", "0"));
+        si.addSitePIP(si.getSitePIP("CUSED", "0"));
+        si.addSitePIP(si.getSitePIP("BUSED", "0"));
+        si.addSitePIP(si.getSitePIP("AUSED", "0"));
+        LUTSiteInsts.add(si);
+    }
+    return LUTSiteInsts;
+} // end packLUTGroups()
+\end{lstlisting}
+\begin{multicols}{2}
 
@@
     \label{subsec:placement}
-    Up until now we have only organized the logical \texttt{Cells} into \texttt{SiteInsts}. 
-    This is where simulated annealing actually begins.
+    Up until now we have only organized the logical \texttt{EDIFCellInst}s into \texttt{SiteInst}s. 
+    This is where simulated annealing actually begins where we actually place the \texttt{SiteInst}s onto physical \texttt{Site}s on the \texttt{device} level.

### 2025-05-11 latex/paper/sections/08-netlist.tex
packing

@@
 Thus, one could express the netlist as a hypergraph, nodes representing components, hyperedges representing wires connecting two or more component. 
-More precisely, these hyperedges connect the ports between the components, not the components themselves, with each component exposing multiple ports.
+More precisely, these hyperedges connect the ports on the components, not the components themselves, with each component exposing multiple ports.
 
@@
 // Downstream task:
-// For each unique net, print out the connected cells.
+// For each unique net, print out the incident cells.
 

### 2025-05-11 latex/paper/sections/09-packing.tex
packing

@@
-\subsection{Site Packing}
+\section{Placement}
+\end{multicols}
+{
+    \centering
+    \includegraphics[width=0.8\columnwidth]{figures/substages.png}
+    \captionof{figure}{The data classes populated at each substage: \texttt{PrepackedDesign}, \texttt{PackedDesign}, and \texttt{PlacedDesign}.}
+    \label{fig:substages}
+}
+\begin{multicols}{2}
+\label{sec:simulated_annealing}
+With a basic understanding of FPGA architecture, design placement, and RapidWright, we have all the necessary pieces to implement our SA placer. 
+Here we outline in detail each substage of our implementation: PrePacking, Packing, and Placement. 
+Shown in Figure \ref{fig:edif_design_device} is an overview of the placement workflow. 
+Figure \ref{fig:substages} shows the data structures of RapidWright objects that are populated at each stage: \texttt{PrepackedDesign}, which is a group of data structures around \texttt{EDIFHierCellInst}s, \texttt{PackedDesign}, which is a group fo data structures around \texttt{SiteInst}s, and finally, \texttt{PlacedDesign}, which is simply captured by the final RapidWright \texttt{Design} object.
+
+\vspace{0.5cm}
+{
+    \centering
+    \includegraphics[width=0.9\columnwidth]{figures/edif_design_device.png}
+    \captionof{figure}{Our placement workflow}
+    \label{fig:edif_design_device}
+}
+
+\subsection{Prepacking}
+\label{subsec:prepacking}
+
+The first step in our placement flow is \textbf{prepacking}. 
+Recall from the 7-Series architecture that there are certain multi-cell structures that must adhere to certain placements constraints to ensure legality, and by design, to minimize wirelength. 
+The job of the prepacker is to traverse the raw EDIF netlist, detect these multi-cell structures, and consolidate these cells into clusters or groups of clusters that naturally reflect these placement constraints.
+
+Recall that \texttt{CARRY4} chains must necessarily be placed vertically and consecutively across a column of SLICEs in ascending order. 
+Likewise, \texttt{DSP48E1} cascades must necessarily be placed vertically and consecutively across a column of \texttt{DSP48E1} Sites in ascending order. 
+A LUT-FF pair may be placed freely, but should be placed in the same lane within the same SLICE to minimize wirelength.
+
+The raw EDIF netlist only tells us the list of nets and the cell ports that they connect to. 
+It does not report the presence of any multi-cell structures (\texttt{CARRY4} chains, etc.). 
+Thus, we must traverse the netlist to detect these multi-cell structures and store that structure information in a class we will call \texttt{PrepackedDesign}.
+
+The code snippet in \ref{lst:carry_chains} shows how one can detect and collect these \texttt{CARRY4} chains using RapidWright. 
+We first collect the cells in the design that are of type \texttt{CARRY4}, then iteratively traverse their Carry-Out (\texttt{CO}) to Carry-In (\texttt{CI}) nets to find incident \texttt{CARRY4} cells.
+Each \texttt{CARRY4} chain has an anchor cell and a tail cell where the chain terminates.
+The anchor is found when the \texttt{CI} net connects to Ground (\texttt{GND}), while the tail is found when the \texttt{CO} port is null. 
+We can further detect if there are \texttt{LUT}s or \texttt{FF}s connected to the \texttt{CARRY4} cell and store that information in a data structure we will call \texttt{CarryCellGroup} as defined in figure \ref{fig:substages}.
+This will help us in knowing which cells can be packed together into the same \texttt{Site} in the subsequent stages.
+
+Similarly, \texttt{DSP48E1} cascades can be found and collected by traversing the \texttt{PCOUT} \texttt{ACOUT} and \texttt{BCOUT} nets.
+\texttt{LUT-FF} pairs can be found by inspecting the LUT output (O) net and checking for FF input (DI) ports. 
+We can bucket these LUT-FF pairs by finding the set of unique \texttt{CE} \texttt{SR} net pairs to know which group of LUT-FF pairs can be placed within the same \texttt{Site}.
+
+We detect the presence of these multi-cellular structures and consolidate that information into our \texttt{PrepackedDesign} object in preparation for the following Packing stage.
+
+\end{multicols}
+{
+    \centering
+    \includegraphics[width=0.8\columnwidth]{figures/carry_chain_traversal.png}
+    \captionof{figure}{A netlist with two \texttt{CARRY4} chains, each of size 3}
+    \label{fig:carry_chain_traversal}
+}
+\begin{multicols}{2}
+
+\begin{lstlisting}[caption={Code Printout}]
+Anchor Cell: adder2/sum_reg[3]_i_1, CellType: CARRY4
+	Cell: adder2/sum_reg[7]_i_1, CellType: CARRY4
+	Cell: adder2/sum_reg[11]_i_1, CellType: CARRY4
+Anchor Cell: adder1/sum_reg[3]_i_1, CellType: CARRY4
+	Cell: adder1/sum_reg[7]_i_1, CellType: CARRY4
+	Cell: adder1/sum_reg[8]_i_1, CellType: CARRY4
+\end{lstlisting}
+
+\begin{lstlisting}[language=java, caption={Finding and storing carry chains.}, label={lst:carry_chains}]
+Design design = Design.readCheckpoint("synth.dcp")
+EDIFNetlist netlist = design.getNetlist();
+List<EDIFCellInst> ecis = netlist.getAllLeafCellInstances();
+
+// Select only the carry cells.
+List<EDIFCellInst> carryCells = new ArrayList<>();
+for (EDIFCellInst eci : ecis) {
+    if (eci.getCellName().contains("CARRY4"))
+        carryCells.add(eci);
+}
+
+// Find and remove carry chains until the list is empty
+List<List<EDIFCellInst>> carryChains = new ArrayList<>();
+while (!carryCells.isEmpty()) {
+    // Arbitrarily set "currentCell" pointer to a cell in the list
+    EDIFCellInst currentCell = carryCells.get(0);
+
+    // Find this carry chain anchor.
+    // Traverse the Carry-In (CI) to Carry-Out (CO) nets.
+    // Anchor is found when net on the CI Port is Ground.
+    while (true) {
+        System.out.println(currentCell);
+        // Access the CI port on this cell.
+        EDIFPortInst sinkPort = currentCell.getPortInst("CI");
+        // Access the net on this CI port.
+        EDIFNet net = sinkPort.getNet();
+        if (net.isGND()) {
+            // Found this chain anchor!
+            break;
+        }
+        // Get all ports on this net.
+        List<EDIFPortInst> netPorts = net.getPortInsts();
+        for (EDIFPortInst netPort : netPorts) {
+            // Access the port belonging to another carry cell.
+            EDIFCellInst sourceCell = netPort.getCellInst();
+            if (sourceCell.getCellName().equals("CARRY4")) {
+                // Move the "currentCell" pointer
+                currentCell = sourceCell;
+                break;
+            }
+        }
+    }
+
+    // Now we have the chain anchor as currentCell.
+    // Now traverse in the opposite direction to find the chain tail.
+    // Tail is found when the CO Port is null.
+    // Collect the chain cells into an ordred list.
+    List<EDIFCellInst> currentChain = new ArrayList<>();
+    currentChain.add(currentCell);
+    while (true) {
+        EDIFPortInst sourcePort = currentCell.getPortInst("CO[3]");
+        if (sourcePort == null) {
+            // Found this chain's tail!
+            break;
+        }
+        EDIFNet net = sourcePort.getNet();
+        List<EDIFPortInst> netPorts = net.getPortInsts();
+        for (EDIFPortInst netPort : netPorts) {
+            EDIFCellInst sinkCell = netPort.getCellInst();
+            if (netPort.getName().equals("CI") &&
+                    sinkCell.getCellName().equals("CARRY4")) {
+                currentCell = sinkCell;
+                // Add the cell to the chain list.
+                currentChain.add(currentCell);
+                break;
+            }
+        }
+    }
+    // Add currentChain to the list of chains
+    carryChains.add(currentChain);
+    // Remove currentChain from the list of cells
+    carryCells.removeAll(currentChain);
+} // end while()
+
+// Print out the carry chains. 
+for (List<EDIFCellInst> chain : chains) {
+    for (int i = 0; i < chain.size(); i++) {
+        EDIFCellInst carry = chain.get(i);
+        if (i == 0) {
+            writer.write("\nAnchor Cell: " + carry.getName() + 
+                ", CellType: " + carry.getCellName());
+        } else {
+            writer.write("\n\tCell: " + carry.getName() + 
+                ", CellType: " + carry.getCellName());
+        }
+    }
+}
+\end{lstlisting}
+\newpage
+
+\subsection{Packing}
 \label{subsec:packing}
+Now that we have our \texttt{PrepackedDesign} object keeping track of multi-cell structures on the \texttt{edif} level, we can start packing them into \texttt{SiteInst} objects on the \texttt{design} level.
+Below are some of the most relevant classes from the \texttt{design} package for this task.
+\begin{itemize}
+    \item \texttt{Cell}: A cell corresponds to the leaf cell within the logical netlist \texttt{EDIFCellInst} and provides a mapping to a physical location BEL on the device.
+A cell can be created directly out of an \texttt{EDIFCellInst} to inherit all of its \texttt{edif} properties on the \texttt{design} level.
+    \item \texttt{Net}: Represents the physical net to be routed (both inter-site and intra-site).
+When an \texttt{Cell} is created out of an \texttt{EDIFCellInst}, the \texttt{Net}s are automatically created out of its corresponding \texttt{EDIFNet}s.
+    \item \texttt{SiteInst}: An instance of a \texttt{Site} on the \texttt{Device}.
+Carries the mapping information between the \texttt{BEL}s in a \texttt{Site} and the \texttt{Cell}s assigned to them.
+Also keeps track of the intra-Site routing information within (\texttt{Net}s, \texttt{SitePinInst}s, \texttt{SitePIP}s, etc.).
+    \item \texttt{SitePIP}: Represents the fuses in routing \texttt{BEL}s. 
+    \item \texttt{SitePinInst}: An instance of a \texttt{SitePin} on a \texttt{Site}. These objects serve as the interface between intra-Site routing and general inter-Site routing. 
+\end{itemize}
+\columnbreak
+
+A \texttt{SiteInst} object can be created and its BELs populated by existing \texttt{EDIFHierCellInst} objects. 
+A \texttt{SiteInst} is necessarily placed on a specific \texttt{Site} upon creation. 
+Thus, the packer necessarily gives all \texttt{SiteInst}s an initial placement before randomization via simulated annealing in the following placement stage. 
+In packing, will simply assign \texttt{SiteInst}s pseudorandomly onto device \texttt{Site}s coordinate order as they are created (e.g.
+first SiteInst onto \texttt{SLICEL\_X0Y0}, second \texttt{SiteInst} onto \texttt{SLICEL\_X0Y1}, etc.).
+
+\end{multicols}
+\begin{lstlisting}[caption=\texttt{SiteInst} constructor and methods.]
+SiteInst Constructor: 
+    SiteInst(String name, Design design, SiteTypeEnum type, Site site)
+
+Most relevant SiteInst Methods:
+    createCell(EDIFHierCellInst inst, BEL bel) // Populating the SiteInst BELs with Cells
+    unplace() // Unplacing the SiteInst from its current Site
+    place(Site site) // Placing an unplaced SiteInst onto a Site
+    routeSite() // Attempt to automatically route all intra-Site nets (manual intervention likely required)
+    routeIntraSiteNet(Net net, BELPin src, BELPin snk) // Manually route an intra-Site net
+
+Example:
+    SiteInst si = new SiteInst("mySiteInst", design, SiteTypeEnum.SLICEL, device.getSite("SLICEL_X0Y1"));
+    si.unplace();
+    si.place(device.getSite("SLICEL_X15Y33"));
+    // In Simulated Annealing, SiteInst objects will be unplaced() and placed() many times to converge to an optimal solution.
+
+\end{lstlisting}
+
+\begin{lstlisting}[language=java, caption={Packing DSPCascades into \texttt{DSP48E1} \texttt{SiteInst}s.}, label={lst:carry_chains}]
+private List<List<SiteInst>> packDSPCascades(List<List<EDIFHierCellInst>> EDIFDSPCascades) throws IOException {
+    List<List<SiteInst>> siteInstCascades = new ArrayList<>();
+    writer.write("\n\nPacking DSP Cascades... (" + EDIFDSPCascades.size() + ")");
+    for (List<EDIFHierCellInst> cascade : EDIFDSPCascades) {
+        List<SiteInst> siteInstCascade = new ArrayList<>();
+        // each DSP tile has 2 DSP sites on the Zynq 7000
+        writer.write("\n\tCascade Size: (" + cascade.size() + "), Cascade Anchor: "
+                + cascade.get(0).getFullHierarchicalInstName());
+        // Select a Site location on the device. 
+        Site anchorSite = selectDSPAnchorSite(cascade.size());
+        SiteTypeEnum siteType = SiteTypeEnum.DSP48E1;
+        for (int i = 0; i < cascade.size(); i++) {
+            writer.write("\n\t\tPlacing DSP: " + cascade.get(i).getFullHierarchicalInstName());
+            Site site = (i == 0) ? anchorSite
+                    : device.getSite("DSP48_X" + anchorSite.getInstanceX() + "Y" + (anchorSite.getInstanceY() + i));
+            // Create the SiteInst object
+            SiteInst si = new SiteInst(cascade.get(i).getFullHierarchicalInstName(), design, siteType, site);
+            // Populate the SiteInst with a DSP Cell. 
+            si.createCell(cascade.get(i), si.getBEL("DSP48E1"));
+            occupiedSites.get(siteType).add(site);
+            availableSites.get(siteType).remove(site);
+            si.routeSite();
+            siteInstCascade.add(si);
+        }
+        siteInstCascades.add(siteInstCascade);
+    }
+    return siteInstCascades;
+} // end packDSPCascades()
+\end{lstlisting}
+
+\newpage
+\begin{lstlisting}[language=java, caption={Packing LUTFFPairGroups into \texttt{SLICEL} \texttt{SiteInst}s}]
+private List<SiteInst> packLUTFFPairGroups(
+        Map<Pair<String, String>, LUTFFGroup> LUTFFEnableResetGroups) throws IOException {
+    List<SiteInst> LUTFFSiteInsts = new ArrayList<>();
+    writer.write("\n\nPacking LUT-FF Pair Groups... (" + LUTFFEnableResetGroups.size() + ")");
+    for (Map.Entry<Pair<String, String>, LUTFFGroup> entry : LUTFFEnableResetGroups.entrySet()) {
+        Pair<String, String> netPair = entry.getKey();
+        String s1 = String.format("\n\tCENet: %-50s RNet: %-50s", netPair.key(), netPair.value());
+        writer.write(s1);
+        LUTFFGroup lutffgroup = entry.getValue();
+        for (List<Pair<EDIFHierCellInst, EDIFHierCellInst>> LUTFFPairs : splitIntoGroups(lutffgroup.group(), 4)) {
+            Site selectedSite = selectCLBSite();
+            SiteInst si = new SiteInst(LUTFFPairs.get(0).value().getFullHierarchicalInstName(), design,
+                    selectedSite.getSiteTypeEnum(), selectedSite);
+            for (int i = 0; i < LUTFFPairs.size(); i++) {
+                EDIFHierCellInst ff = LUTFFPairs.get(i).value();
+                EDIFHierCellInst lut = LUTFFPairs.get(i).key();
+                si.createCell(ff, si.getBEL(FF_BELS[i]));
+                if (lut != null)
+                    si.createCell(lut, si.getBEL(LUT6_BELS[i]));
+            }
+            si.routeSite();
+            List<String> lutOPIP = new ArrayList<>(List.of("AUSED", "BUSED", "CUSED", "DUSED"));
+            // DOES NOT SUPPORT STACKED LUTS OR LUT5s!
+            for (int i = 0; i < LUTFFPairs.size(); i++) {
+                EDIFHierCellInst lut = LUTFFPairs.get(i).key();
+                if (lut != null) {
+                    EDIFHierPortInst lutOPort = lut.getPortInst("O");
+                    if (lutOPort.getHierarchicalNet().getLeafHierPortInsts(false, true).size() > 1) {
+                        si.addSitePIP(si.getSitePIP(lutOPIP.get(i), "0"));
+                    }
+                }
+            }
+            rerouteFFClkSrCeNets(si);
+            LUTFFSiteInsts.add(si);
+        }
+    }
+    return LUTFFSiteInsts;
+} // end packLUTFFPairGroups()
+\end{lstlisting}
+
+\begin{lstlisting}[language=java, caption={Packing LUTGroups into \texttt{SLICEL} \texttt{SiteInst}s}]
+protected String[] LUT6_BELS = new String[] { "A6LUT", "B6LUT", "C6LUT", "D6LUT" };
+
+private List<SiteInst> packLUTGroups(List<List<EDIFHierCellInst>> LUTGroups) throws IOException {
+    List<SiteInst> LUTSiteInsts = new ArrayList<>();
+    writer.write("\n\nPacking LUT Groups...");
+    for (List<EDIFHierCellInst> group : LUTGroups) {
+        Site selectedSite = selectCLBSite();
+        SiteInst si = new SiteInst(group.get(0).getFullHierarchicalInstName(), design,
+                selectedSite.getSiteTypeEnum(), selectedSite);
+        for (int i = 0; i < group.size(); i++) {
+            si.createCell(group.get(i), si.getBEL(LUT6_BELS[i]));
+        }
+        si.routeSite();
+        // for whatever reason, this does not activate DUSED PIP for
+        si.addSitePIP(si.getSitePIP("DUSED", "0"));
+        si.addSitePIP(si.getSitePIP("CUSED", "0"));
+        si.addSitePIP(si.getSitePIP("BUSED", "0"));
+        si.addSitePIP(si.getSitePIP("AUSED", "0"));
+        LUTSiteInsts.add(si);
+    }
+    return LUTSiteInsts;
+} // end packLUTGroups()
+\end{lstlisting}
+\begin{multicols}{2}
+
+\subsection{Placement}
+    \label{subsec:placement}
+    Up until now we have only organized the logical \texttt{EDIFCellInst}s into \texttt{SiteInst}s. 
+    This is where simulated annealing actually begins where we actually place the \texttt{SiteInst}s onto physical \texttt{Site}s on the \texttt{device} level.

### 2025-05-12 latex/paper/sections/05-7series_architecture.tex
more packing

@@
 They constitute the fundamental building blocks for implementing digital circuits on the FPGA.
-Note how the Tile arrangement is columnar such that column consists of only one Tile type.
 
@@
 Each Site is embedded into a \textbf{Tile}, and Tiles are further arranged into \textbf{Clock Regions}. 
+Note how the Tile arrangement is columnar such that column consists of only one Tile type. 
 In some high-density devices, multiple Clock Regions may be consolidated into one or more \textbf{Super Logic Regions}~(SLRs). 
@@
 These cells are also referred to as \emph{Distributed RAM} in the Xilinx documentation.
-    These cells can offer an alternative to the larger, more dedicated 18K-36K \texttt{RAMB18E1} cells when RAM resources are highly utilized. 
+    These cells can offer an alternative to the larger, more dedicated 18K-36K \texttt{RAMB18E1} cells when RAM resources are highly utilized or when the design inheritly demands homogeneously distributed local memory.
 \end{itemize}
@@
     \includegraphics[width=\columnwidth]{figures/carry_cell_group.png}
-    \captionof{figure}{A \texttt{SLICEL} with a \texttt{CARRY4} cell, 4 LUT cells, and 4 FF cells placed inside as shown in the Vivado device viewer}
+    \captionof{figure}{A \texttt{SLICEL} populated by a \texttt{CARRY4} cell, 4 LUT cells, and 4 FF cells}
     \label{fig:carry_cell_group}
@@
 
+\newcolumn
 {

### 2025-05-12 latex/paper/sections/08-netlist.tex
more packing

@@
 In either synthesized netlist, the \textbf{leaf cells}, (deepest level cells), must necessarily consist only of \textbf{primitive cells} from the architecture's primitive cell library (\texttt{LUT6}, \texttt{FDRE}, \texttt{CARRY4}, \texttt{DSP48E1}, etc.).
-The netlist can be compiled and exported as a purely structural low-level Verilog file, or an Electrinic Design Interchange Format (EDIF) file, both describing the netlist explicitly as a list of logical cells connected by a list of wires.
+The netlist can be compiled and exported as a purely structural low-level Verilog file, or an Electrinic Design Interchange Format (EDIF) file, both describing the netlist explicitly as a list of logical cell instances connected by a list of wires.
 
@@
 
-\begin{lstlisting}[language=java, caption={Netlist extraction and traversal}, label={lst:netlist_extract}]
+\begin{lstlisting}[language=java, caption={Basic netlist extraction and traversal}, label={lst:netlist_extract}]
 Design design = Design.readCheckpoint("synth.dcp")

### 2025-05-12 latex/paper/sections/09-packing.tex
more packing

@@
 for (EDIFCellInst eci : ecis) {
-    if (eci.getCellName().contains("CARRY4"))
+    if (eci.getCellName().equals("CARRY4"))
         carryCells.add(eci);
@@
 Also keeps track of the intra-Site routing information within (\texttt{Net}s, \texttt{SitePinInst}s, \texttt{SitePIP}s, etc.).
-    \item \texttt{SitePIP}: Represents the fuses in routing \texttt{BEL}s. 
+    \item \texttt{SitePIP}: A Programmable Interconnect Point (PIP) in a \texttt{Site}. Represents the fuses in intra-Site routing \texttt{BEL}s. 
     \item \texttt{SitePinInst}: An instance of a \texttt{SitePin} on a \texttt{Site}. These objects serve as the interface between intra-Site routing and general inter-Site routing. 
@@
 
-A \texttt{SiteInst} object can be created and its BELs populated by existing \texttt{EDIFHierCellInst} objects. 
-A \texttt{SiteInst} is necessarily placed on a specific \texttt{Site} upon creation. 
-Thus, the packer necessarily gives all \texttt{SiteInst}s an initial placement before randomization via simulated annealing in the following placement stage. 
-In packing, will simply assign \texttt{SiteInst}s pseudorandomly onto device \texttt{Site}s coordinate order as they are created (e.g.
-first SiteInst onto \texttt{SLICEL\_X0Y0}, second \texttt{SiteInst} onto \texttt{SLICEL\_X0Y1}, etc.).
+A new \texttt{SiteInst} is created by populating its BELs with existing \texttt{EDIFHierCellInst} objects from the \texttt{EDIFNetlist}, and is immediately placed on a specific \texttt{Site}. 
+The packer therefore assigns an initial placement to every \texttt{SiteInst} before the simulated‐annealing stage randomizes their positions. 
+During this packing phase, we simply pseudorandomly map \texttt{design} \texttt{SiteInst}s onto \texttt{device} \texttt{Site}s in coordinate order as they are generated (e.g., the first \texttt{SiteInst} onto \texttt{SLICEL\_X0Y0}, the second onto \texttt{SLICEL\_X0Y1}, and so on).
+
+One can think of a \texttt{design} \texttt{SiteInst} as a movable square peg and the corresponding \texttt{device} \texttt{Site} as a fixed square hole. 
+Both \texttt{SiteInst}s and \texttt{Site}s come in various shapes or types (\texttt{SLICEL}, \texttt{SLICEM}, \texttt{DSP48E1}, etc.) and placement is only allowed between compatible pairs. 
+For example, a \texttt{RAMB48E1} \texttt{SiteInst} can only be placed on a \texttt{RAMB48E1} \texttt{Site}.
+A \texttt{DSP48E1} \texttt{SiteInst} can only be placed on a \texttt{DSP48E1} \texttt{Site}.
+A \texttt{SLICEL} \texttt{SiteInst} may occupy either a \texttt{SLICEL} or a \texttt{SLICEM} \texttt{Site}, whereas a \texttt{SLICEM} \texttt{SiteInst} can only be placed on a \texttt{SLICEM} \texttt{Site}.
+These constraints must be followed as the \texttt{SiteInst}s move across the device throughout the placement stage to ensure legality.
 
@@
     SiteInst si = new SiteInst("mySiteInst", design, SiteTypeEnum.SLICEL, device.getSite("SLICEL_X0Y1"));
+    si.createCell(someFDRECell, si.getBEL("AFF"));
+    si.routeSite();
     si.unplace();
@@
     // In Simulated Annealing, SiteInst objects will be unplaced() and placed() many times to converge to an optimal solution. 
+    // All Cell-BEL mapping and intra-Site routing is preserved when a SiteInst is moved.
 
@@
 
-\begin{lstlisting}[language=java, caption={Packing DSPCascades into \texttt{DSP48E1} \texttt{SiteInst}s.}, label={lst:carry_chains}]
-private List<List<SiteInst>> packDSPCascades(List<List<EDIFHierCellInst>> EDIFDSPCascades) throws IOException {
-    List<List<SiteInst>> siteInstCascades = new ArrayList<>();
-    writer.write("\n\nPacking DSP Cascades... (" + EDIFDSPCascades.size() + ")");
-    for (List<EDIFHierCellInst> cascade : EDIFDSPCascades) {
-        List<SiteInst> siteInstCascade = new ArrayList<>();
-        // each DSP tile has 2 DSP sites on the Zynq 7000
-        writer.write("\n\tCascade Size: (" + cascade.size() + "), Cascade Anchor: "
-                + cascade.get(0).getFullHierarchicalInstName());
-        // Select a Site location on the device. 
-        Site anchorSite = selectDSPAnchorSite(cascade.size());
-        SiteTypeEnum siteType = SiteTypeEnum.DSP48E1;
-        for (int i = 0; i < cascade.size(); i++) {
-            writer.write("\n\t\tPlacing DSP: " + cascade.get(i).getFullHierarchicalInstName());
-            Site site = (i == 0) ? anchorSite
-                    : device.getSite("DSP48_X" + anchorSite.getInstanceX() + "Y" + (anchorSite.getInstanceY() + i));
-            // Create the SiteInst object
-            SiteInst si = new SiteInst(cascade.get(i).getFullHierarchicalInstName(), design, siteType, site);
-            // Populate the SiteInst with a DSP Cell. 
-            si.createCell(cascade.get(i), si.getBEL("DSP48E1"));
-            occupiedSites.get(siteType).add(site);
-            availableSites.get(siteType).remove(site);
+\begin{lstlisting}[language=java, caption={Packing an individual \texttt{CarryCellGroup} into one \texttt{SLICEL} \texttt{SiteInst}s.}, label={lst:single_carry_chains}]
+protected String[] FF_BELS = new String[] { "AFF", "BFF", "CFF", "DFF" };
+protected String[] LUT6_BELS = new String[] { "A6LUT", "B6LUT", "C6LUT", "D6LUT" };
+
+private void packCarrySite(CarryCellGroup carryCellGroup, SiteInst si) {
+    // potential bug: what guarantees that all of the FFs connected to the CARRY4 all share the same CE and Reset?
+    for (int i = 0; i < 4; i++) {
+        EDIFHierCellInst ff = carryCellGroup.ffs().get(i);
+        if (ff != null)
+            si.createCell(ff, si.getBEL(FF_BELS[i]));
+        EDIFHierCellInst lut = carryCellGroup.luts().get(i);
+        if (lut != null)
+            si.createCell(lut, si.getBEL(LUT6_BELS[i]));
+        // carry site LUTs MUST be placed on LUT6 BELs.
+        // only LUT6/O6 can connect to CARRY4/S0
+    }
+    si.createCell(carryCellGroup.carry(), si.getBEL("CARRY4"));
+    // default intrasite routing
     si.routeSite();
-            siteInstCascade.add(si);
-        }
-        siteInstCascades.add(siteInstCascade);
-    }
-    return siteInstCascades;
-} // end packDSPCascades()
+    // sometimes the default routeSite() is insufficient, so some manual
+    // intervention is required
+    rerouteCarryNets(si);
+    rerouteFFClkSrCeNets(si);
+} // end placeCarrySite()
 \end{lstlisting}
@@
 \newpage
-\begin{lstlisting}[language=java, caption={Packing LUTFFPairGroups into \texttt{SLICEL} \texttt{SiteInst}s}]
-private List<SiteInst> packLUTFFPairGroups(
-        Map<Pair<String, String>, LUTFFGroup> LUTFFEnableResetGroups) throws IOException {
-    List<SiteInst> LUTFFSiteInsts = new ArrayList<>();
-    writer.write("\n\nPacking LUT-FF Pair Groups... (" + LUTFFEnableResetGroups.size() + ")");
-    for (Map.Entry<Pair<String, String>, LUTFFGroup> entry : LUTFFEnableResetGroups.entrySet()) {
-        Pair<String, String> netPair = entry.getKey();
-        String s1 = String.format("\n\tCENet: %-50s RNet: %-50s", netPair.key(), netPair.value());
-        writer.write(s1);
-        LUTFFGroup lutffgroup = entry.getValue();
-        for (List<Pair<EDIFHierCellInst, EDIFHierCellInst>> LUTFFPairs : splitIntoGroups(lutffgroup.group(), 4)) {
-            Site selectedSite = selectCLBSite();
-            SiteInst si = new SiteInst(LUTFFPairs.get(0).value().getFullHierarchicalInstName(), design,
-                    selectedSite.getSiteTypeEnum(), selectedSite);
-            for (int i = 0; i < LUTFFPairs.size(); i++) {
-                EDIFHierCellInst ff = LUTFFPairs.get(i).value();
-                EDIFHierCellInst lut = LUTFFPairs.get(i).key();
-                si.createCell(ff, si.getBEL(FF_BELS[i]));
-                if (lut != null)
-                    si.createCell(lut, si.getBEL(LUT6_BELS[i]));
+\begin{lstlisting}[language=java, caption={Packing \texttt{CarryCellGroup}s into \texttt{SLICEL} \texttt{SiteInst}s.}, label={lst:carry_chains}]
+private List<List<SiteInst>> packCarryChains(List<List<CarryCellGroup>> EDIFCarryChains)
+        throws IOException {
+    List<List<SiteInst>> siteInstChains = new ArrayList<>();
+    writer.write("\n\nPacking carry chains... (" + EDIFCarryChains.size() + ")");
+    for (List<CarryCellGroup> edifChain : EDIFCarryChains) {
+        List<SiteInst> siteInstChain = new ArrayList<>();
+        writer.write("\n\t\tChain Size: (" + edifChain.size() + "), Chain Anchor: "
+                + edifChain.get(0).carry().getFullHierarchicalInstName());
+        Site anchorSite = selectCarryAnchorSite(edifChain.size());
+        SiteTypeEnum selectedSiteType = anchorSite.getSiteTypeEnum();
+        for (int i = 0; i < edifChain.size(); i++) {
+            Site site = (i == 0) ? anchorSite
+                    : device.getSite("SLICE_X" + anchorSite.getInstanceX() + "Y" + (anchorSite.getInstanceY() + i));
+            SiteInst si = new SiteInst(edifChain.get(i).carry().getFullHierarchicalInstName(), design,
+                    selectedSiteType,
+                    site);
+            packCarrySite(edifChain.get(i), si);
+            if (i == 0) { // additional routing logic for anchor site
+                Net CINNet = si.getNetFromSiteWire("CIN");
+                CINNet.removePin(si.getSitePinInst("CIN"));
+                si.addSitePIP(si.getSitePIP("PRECYINIT", "0"));
             }
-            si.routeSite();
-            List<String> lutOPIP = new ArrayList<>(List.of("AUSED", "BUSED", "CUSED", "DUSED"));
-            // DOES NOT SUPPORT STACKED LUTS OR LUT5s!
-            for (int i = 0; i < LUTFFPairs.size(); i++) {
-                EDIFHierCellInst lut = LUTFFPairs.get(i).key();
-                if (lut != null) {
-                    EDIFHierPortInst lutOPort = lut.getPortInst("O");
-                    if (lutOPort.getHierarchicalNet().getLeafHierPortInsts(false, true).size() > 1) {
-                        si.addSitePIP(si.getSitePIP(lutOPIP.get(i), "0"));
+            occupiedSites.get(selectedSiteType).add(site);
+            availableSites.get(selectedSiteType).remove(site);
+            siteInstChain.add(si);
         }
-                }
-            }
-            rerouteFFClkSrCeNets(si);
-            LUTFFSiteInsts.add(si);
-        }
-    }
-    return LUTFFSiteInsts;
-} // end packLUTFFPairGroups()
+        siteInstChains.add(siteInstChain);
+    } // end for (List<EDIFCellInst> chain : EDIFCarryChains)
+    return siteInstChains;
+} // end packCarryChains()
+
 \end{lstlisting}
 
-\begin{lstlisting}[language=java, caption={Packing LUTGroups into \texttt{SLICEL} \texttt{SiteInst}s}]
-protected String[] LUT6_BELS = new String[] { "A6LUT", "B6LUT", "C6LUT", "D6LUT" };
-
-private List<SiteInst> packLUTGroups(List<List<EDIFHierCellInst>> LUTGroups) throws IOException {
-    List<SiteInst> LUTSiteInsts = new ArrayList<>();
-    writer.write("\n\nPacking LUT Groups...");
-    for (List<EDIFHierCellInst> group : LUTGroups) {
-        Site selectedSite = selectCLBSite();
-        SiteInst si = new SiteInst(group.get(0).getFullHierarchicalInstName(), design,
-                selectedSite.getSiteTypeEnum(), selectedSite);
-        for (int i = 0; i < group.size(); i++) {
-            si.createCell(group.get(i), si.getBEL(LUT6_BELS[i]));
-        }
-        si.routeSite();
-        // for whatever reason, this does not activate DUSED PIP for
-        si.addSitePIP(si.getSitePIP("DUSED", "0"));
-        si.addSitePIP(si.getSitePIP("CUSED", "0"));
-        si.addSitePIP(si.getSitePIP("BUSED", "0"));
-        si.addSitePIP(si.getSitePIP("AUSED", "0"));
-        LUTSiteInsts.add(si);
-    }
-    return LUTSiteInsts;
-} // end packLUTGroups()
+\begin{lstlisting}[language=java, caption={Manually rerouting intra-Site nets in a \texttt{SLICEL} containing \texttt{CARRY4}}]
+    protected String[] FF_BELS = new String[] { "AFF", "BFF", "CFF", "DFF" };
+    private void rerouteCarryNets(SiteInst si) {
+        // activate PIPs for CARRY4/COUT
+        si.addSitePIP(si.getSitePIP("COUTUSED", "0"));
+        // undo default CARRY4/DI nets
+        SitePinInst AX = si.getSitePinInst("AX");
+        if (AX != null)
+            si.unrouteIntraSiteNet(AX.getBELPin(), si.getBELPin("ACY0", "AX"));
+        SitePinInst DX = si.getSitePinInst("DX");
+        if (DX != null)
+            si.unrouteIntraSiteNet(DX.getBELPin(), si.getBELPin("DCY0", "DX"));
+        // activate PIPs for CARRY4/DI pins
+        si.addSitePIP(si.getSitePIP("DCY0", "DX"));
+        si.addSitePIP(si.getSitePIP("CCY0", "CX"));
+        si.addSitePIP(si.getSitePIP("BCY0", "BX"));
+        si.addSitePIP(si.getSitePIP("ACY0", "AX"));
+        // remove stray CARRY4/CO nets
+        if (si.getNetFromSiteWire("CARRY4_CO2") != null)
+            design.removeNet(si.getNetFromSiteWire("CARRY4_CO2"));
+        if (si.getNetFromSiteWire("CARRY4_CO1") != null)
+            design.removeNet(si.getNetFromSiteWire("CARRY4_CO1"));
+        if (si.getNetFromSiteWire("CARRY4_CO0") != null)
+            design.removeNet(si.getNetFromSiteWire("CARRY4_CO0"));
+        // add default XOR PIPs for unused FFs
+        for (String FF : FF_BELS)
+            if (si.getCell(FF) == null)
+                si.addSitePIP(si.getSitePIP(FF.charAt(0) + "OUTMUX", "XOR"));
+    } // end rerouteCarryNets()
 \end{lstlisting}
@@
     \label{subsec:placement}
-    Up until now we have only organized the logical \texttt{EDIFCellInst}s into \texttt{SiteInst}s. 
+    Up until now we have only organized the logical \texttt{EDIFHierCellInst}s into \texttt{SiteInst}s. 
     This is where simulated annealing actually begins where we actually place the \texttt{SiteInst}s onto physical \texttt{Site}s on the \texttt{device} level.

### 2025-05-13 latex/paper/sections/08-netlist.tex
more packing

@@
 Figure ~\ref{fig:hierarchical_design} shows an example a Verilog design with modules instantiated in a hierarchy. 
-Figure ~\ref{fig:hier_netlist} shows the design synthesized into a hierarchical netlist with \textbf{hierarchical cells} and \textbf{leaf cells}. 
+Figure ~\ref{fig:hier_netlist} shows the design synthesized into a \textbf{hierarchical netlist} with \textbf{hierarchical cells} and \textbf{leaf cells}. 
 The synthesizer attempts to construct the module hierarchy as close to the module instantiation hierarchy defined by the user design entry. 
-Figure ~\ref{fig:flat_netlist} shows the same design but synthesized into a flattened netlist.
+Figure ~\ref{fig:flat_netlist} shows the same design but synthesized into a \textbf{flattened netlist} containing \textbf{only leaf cells}.
 
 In either synthesized netlist, the \textbf{leaf cells}, (deepest level cells), must necessarily consist only of \textbf{primitive cells} from the architecture's primitive cell library (\texttt{LUT6}, \texttt{FDRE}, \texttt{CARRY4}, \texttt{DSP48E1}, etc.).
-The netlist can be compiled and exported as a purely structural low-level Verilog file, or an Electrinic Design Interchange Format (EDIF) file, both describing the netlist explicitly as a list of logical cell instances connected by a list of wires.
+The netlist can be compiled and exported as a purely structural low-level Verilog file, or an Electrinic Design Interchange Format (EDIF) file, both describing the netlist explicitly as a list of cell instances connected by a list of wires.
 

### 2025-05-13 latex/paper/sections/09-packing.tex
more packing

@@
 \end{itemize}
-\columnbreak
 
-A new \texttt{SiteInst} is created by populating its BELs with existing \texttt{EDIFHierCellInst} objects from the \texttt{EDIFNetlist}, and is immediately placed on a specific \texttt{Site}. 
+A \texttt{SiteInst} object can be created and its \texttt{BEL}s populated with existing \texttt{EDIFHierCellInst} objects from the \texttt{EDIFNetlist}. 
+A \texttt{SiteInst} is placed on a specific \texttt{device} \texttt{Site} upon its creation using the constructor in Listing \ref{lst:siteinst}. 
 The packer therefore assigns an initial placement to every \texttt{SiteInst} before the simulated‐annealing stage randomizes their positions. 
-During this packing phase, we simply pseudorandomly map \texttt{design} \texttt{SiteInst}s onto \texttt{device} \texttt{Site}s in coordinate order as they are generated (e.g., the first \texttt{SiteInst} onto \texttt{SLICEL\_X0Y0}, the second onto \texttt{SLICEL\_X0Y1}, and so on).
+During this packing phase, we simply map \texttt{design} \texttt{SiteInst}s onto \texttt{device} \texttt{Site}s in coordinate order as they are generated (e.g., the first \texttt{SiteInst} onto \texttt{SLICEL\_X0Y0}, the second onto \texttt{SLICEL\_X0Y1}, and so on).
 
@@
 A \texttt{SLICEL} \texttt{SiteInst} may occupy either a \texttt{SLICEL} or a \texttt{SLICEM} \texttt{Site}, whereas a \texttt{SLICEM} \texttt{SiteInst} can only be placed on a \texttt{SLICEM} \texttt{Site}.
-These constraints must be followed as the \texttt{SiteInst}s move across the device throughout the placement stage to ensure legality.
+These constraints must always be followed as the \texttt{SiteInst}s move across the device throughout the placement stage to ensure legality.
+
+Listing \ref{lst:siteinst} shows how to use the \texttt{SiteInst} object and how it ties together \texttt{Site}s, \texttt{Cell}s, \texttt{BEL}s, and intra-Site \texttt{Net}s. 
+The super-specifics of the packing process via RapidWright is too involved for the scope of this paper, but just to give an impression of how the \texttt{design} classes are used, Listing \ref{lst:single_carry_chains}, \ref{lst:carry_chains}, \ref{lst:carry_nets}, and \ref{lst:ff_nets} together show how \texttt{CarryCellGroup}s are packed into \texttt{SLICEL} \texttt{Sites} in our working code.
+Packing \texttt{DSP48E1} cascades, \texttt{RAMB48E1} Sites, and non-\texttt{CARRY4} \texttt{SLICEL} Sites follow a similar process in instantiating \texttt{SiteInst}s and populating their \texttt{BEL}s with \texttt{Cell}s using existing \texttt{EDIFHierCellInst}s extracted from the \texttt{EDIFNetlist}.
 
 \end{multicols}
-\begin{lstlisting}[caption=\texttt{SiteInst} constructor and methods.]
+\begin{lstlisting}[caption={\texttt{SiteInst} constructor and methods.}, label={lst:siteinst}]
 SiteInst Constructor: 
@@
 
-\begin{lstlisting}[language=java, caption={Packing an individual \texttt{CarryCellGroup} into one \texttt{SLICEL} \texttt{SiteInst}s.}, label={lst:single_carry_chains}]
+\begin{lstlisting}[language=java, caption={Packing one \texttt{CarryCellGroup} into one \texttt{SLICEL} \texttt{SiteInst}.}, label={lst:single_carry_chains}]
+// Names of the BELs in the 4 LUT-FF lanes:
 protected String[] FF_BELS = new String[] { "AFF", "BFF", "CFF", "DFF" };
 protected String[] LUT6_BELS = new String[] { "A6LUT", "B6LUT", "C6LUT", "D6LUT" };
-
+// Pack one CarryCellGroup 
 private void packCarrySite(CarryCellGroup carryCellGroup, SiteInst si) {
-    // potential bug: what guarantees that all of the FFs connected to the CARRY4 all share the same CE and Reset?
+    // Iterate through 4 LUT-FF Lanes
     for (int i = 0; i < 4; i++) {
@@
     si.routeSite();
-    // sometimes the default routeSite() is insufficient, so some manual
-    // intervention is required
+    // sometimes the default routeSite() is insufficient, so some manual intervention is required
     rerouteCarryNets(si);
@@
 \begin{lstlisting}[language=java, caption={Packing \texttt{CarryCellGroup}s into \texttt{SLICEL} \texttt{SiteInst}s.}, label={lst:carry_chains}]
+// Pack all CarryCellGroup chains
 private List<List<SiteInst>> packCarryChains(List<List<CarryCellGroup>> EDIFCarryChains)
@@
 } // end packCarryChains()
-
 \end{lstlisting}
 
-\begin{lstlisting}[language=java, caption={Manually rerouting intra-Site nets in a \texttt{SLICEL} containing \texttt{CARRY4}}]
+\begin{lstlisting}[language=java, caption={Manually rerouting intra-Site nets in a \texttt{SLICEL} containing a \texttt{CARRY4} \texttt{Cell}}, label={lst:carry_nets}]
     protected String[] FF_BELS = new String[] { "AFF", "BFF", "CFF", "DFF" };
@@
 \end{lstlisting}
+
+\begin{lstlisting}[language=java, caption={Manually reroute intra-Site nets in a \texttt{SLICEL} containing FF \texttt{Cells}}, label={lst:ff_nets}]
+private void rerouteFFClkSrCeNets(SiteInst si) {
+    si.addSitePIP("CLKINV", "CLK");
+    // activate PIPs for SR and CE pins
+    Net SRNet = si.getNetFromSiteWire("SRUSEDMUX_OUT");
+    Net CENet = si.getNetFromSiteWire("CEUSEDMUX_OUT");
+    // if routeSite() default PIPs are incorrect, deactivate them then activate correct PIP
+    if (SRNet != null) {
+        if (SRNet.isGNDNet()) {
+            if (si.getSitePIPStatus(si.getSitePIP("SRUSEDMUX", "IN")) == SitePIPStatus.ON)
+                for (String FF : FF_BELS)
+                    si.unrouteIntraSiteNet(si.getSitePinInst("SR").getBELPin(), si.getBELPin(FF, "SR"));
+            si.addSitePIP(si.getSitePIP("SRUSEDMUX", "0"));
+        } else {
+            if (si.getSitePIPStatus(si.getSitePIP("SRUSEDMUX", "0")) == SitePIPStatus.ON)
+                for (String FF : FF_BELS)
+                    si.unrouteIntraSiteNet(si.getBELPin("SRUSEDGND", "0"), si.getBELPin(FF, "SR"));
+            si.addSitePIP(si.getSitePIP("SRUSEDMUX", "IN"));
+        }
+    }
+    if (CENet != null) {
+        if (CENet.isVCCNet()) {
+            if (si.getSitePIPStatus(si.getSitePIP("CEUSEDMUX", "IN")) == SitePIPStatus.ON)
+                for (String FF : FF_BELS)
+                    si.unrouteIntraSiteNet(si.getSitePinInst("CE").getBELPin(), si.getBELPin(FF, "CE"));
+            si.addSitePIP(si.getSitePIP("CEUSEDMUX", "1"));
+        } else {
+            if (si.getSitePIPStatus(si.getSitePIP("CEUSEDMUX", "1")) == SitePIPStatus.ON)
+                for (String FF : FF_BELS)
+                    si.unrouteIntraSiteNet(si.getBELPin("CEUSEDGND", "1"), si.getBELPin(FF, "CE"));
+            si.addSitePIP(si.getSitePIP("CEUSEDMUX", "IN"));
+        }
+    }
+} // end rerouteFFSrCeNets()
+\end{lstlisting}
+
 \begin{multicols}{2}

### 2025-05-14 latex/paper/sections/07-rapidwright.tex
packing edits

--- 2025-05-13 latex/paper/sections/08-netlist.tex
@@
+\section{RapidWright API}
+\label{sec:rapidwright_api}
+
+\textbf{RapidWright} is an open-source Java framework from AMD/Xilinx that provides direct access to the netlist and device databases used by vendor tools. 
+This framework positions itself as an additional workflow column, allowing users to intercept or replace stages of the standard design flow with custom optimization stages (see Figure~\ref{fig:vivado_dcps}).
+
+\begin{itemize}
+\item \textbf{Design Checkpoints:} 
+    RapidWright leverages \texttt{.dcp} files (design checkpoints) generated at various stages of a Vivado flow. 
+    By importing a checkpoint, engineers can manipulate the netlist, placement, or routing externally, then re-export a modified checkpoint for further processing in the Vivado workflow column.
+
+\item \textbf{Key Packages:} 
+    RapidWright revolves around three primary data model packages:
+    \begin{enumerate}
+    \item \texttt{edif} -- Represents the logical netlist in an abstracted EDIF-like structure.
+    \item \texttt{design} -- Contains data structures for the physical implementation (Cells, Nets, Sites, BELs, etc.).
+    \item \texttt{device} -- Provides a database of the target FPGA architecture (e.g., Site coordinates, Tile definitions, routing resources).
+    \end{enumerate}
+
+\item \textbf{Interfacing with the Netlist and Device:} 
+    An engineer can query the netlist to find specific resources (LUTs, FFs, DSPs, etc.) and then map or move them onto device sites. 
+    This level of control over backend resources is necessary for research in custom placement, advanced packing techniques, or experimental routing algorithms.
+\end{itemize}
+
+{
+    \centering
+    \includegraphics[width=0.8\columnwidth]{figures/vivado_dcps.png}
+    \captionof{figure}{RapidWright workflow integrating into the default Vivado design flow.}
+    \label{fig:vivado_dcps}
+}
+
+By exposing these low-level internals, RapidWright allows fine-grained design transformations that go beyond the standard Vivado IDE’s capabilities. 
+Researchers can prototype new EDA strategies without needing to re-implement an entire FPGA backend from scratch, thus accelerating innovation in placement and routing methodologies.
+
 \subsection{What is a Netlist?}
@@
 
-\subsection{Netlist Traversal and Manipulation in RapidWright}
+\subsection{Netlist Traversal and Manipulation via RapidWright}
 
@@
 
+In this method, we get the list of all \texttt{EDIFCellInst} from the \texttt{EDIFNetlist} and traverse the list of cells and access their connected nets. 
+Alternatively, we could have also extracted the list of all \texttt{EDIFNet}s directly from the \texttt{design} and traversed the nets to access their connected cells.
+
 \begin{lstlisting}[caption={Code Printout}]
@@
 \end{lstlisting}
+
+\subsection{Placement Flow via RapidWright}
+\label{subsec:placement}
+\end{multicols}
+{
+    \centering
+    \includegraphics[width=0.8\columnwidth]{figures/substages.png}
+    \captionof{figure}{The data classes populated at each substage: \texttt{PrepackedDesign}, \texttt{PackedDesign}, and \texttt{PlacedDesign}.}
+    \label{fig:substages}
+}
+\begin{multicols}{2}
+\label{sec:simulated_annealing}
+With a basic understanding of FPGA architecture, design placement, and RapidWright, we have all the necessary pieces to implement our SA placer. 
+Here we outline in detail each substage of our implementation: PrePacking, Packing, and Placement. 
+Shown in Figure \ref{fig:edif_design_device} is an overview of the placement workflow. 
+Figure \ref{fig:substages} shows the data structures of RapidWright objects that are populated at each stage: \texttt{PrepackedDesign}, which is a group of data structures around \texttt{EDIFHierCellInst}s, \texttt{PackedDesign}, which is a group fo data structures around \texttt{SiteInst}s, and finally, \texttt{PlacedDesign}, which is simply captured by the final RapidWright \texttt{Design} object.
+
+\vspace{0.5cm}
+{
+    \centering
+    \includegraphics[width=0.9\columnwidth]{figures/edif_design_device.png}
+    \captionof{figure}{Our placement workflow}
+    \label{fig:edif_design_device}
+}

### 2025-05-14 latex/paper/sections/08-prepacking.tex
packing edits

--- 2025-05-11 latex/paper/sections/09-packing.tex
@@
-\section{Placement}
-\end{multicols}
-{
-    \centering
-    \includegraphics[width=0.8\columnwidth]{figures/substages.png}
-    \captionof{figure}{The data classes populated at each substage: \texttt{PrepackedDesign}, \texttt{PackedDesign}, and \texttt{PlacedDesign}.}
-    \label{fig:substages}
-}
-\begin{multicols}{2}
-\label{sec:simulated_annealing}
-With a basic understanding of FPGA architecture, design placement, and RapidWright, we have all the necessary pieces to implement our SA placer. 
-Here we outline in detail each substage of our implementation: PrePacking, Packing, and Placement. 
-Shown in Figure \ref{fig:edif_design_device} is an overview of the placement workflow. 
-Figure \ref{fig:substages} shows the data structures of RapidWright objects that are populated at each stage: \texttt{PrepackedDesign}, which is a group of data structures around \texttt{EDIFHierCellInst}s, \texttt{PackedDesign}, which is a group fo data structures around \texttt{SiteInst}s, and finally, \texttt{PlacedDesign}, which is simply captured by the final RapidWright \texttt{Design} object.
-
-\vspace{0.5cm}
-{
-    \centering
-    \includegraphics[width=0.9\columnwidth]{figures/edif_design_device.png}
-    \captionof{figure}{Our placement workflow}
-    \label{fig:edif_design_device}
-}
-
-\subsection{Prepacking}
-\label{subsec:prepacking}
+\section{Prepacking}
+\label{sec:prepacking}
 
@@
 
-We detect the presence of these multi-cellular structures and consolidate that information into our \texttt{PrepackedDesign} object in preparation for the following Packing stage.
+The overall goal of prepacking is to detect the presence of these multi-cellular structures and consolidate that information into our \texttt{PrepackedDesign} object in preparation for the following Packing stage.
 
@@
 
-\begin{lstlisting}[caption={Code Printout}]
-Anchor Cell: adder2/sum_reg[3]_i_1, CellType: CARRY4
-	Cell: adder2/sum_reg[7]_i_1, CellType: CARRY4
-	Cell: adder2/sum_reg[11]_i_1, CellType: CARRY4
-Anchor Cell: adder1/sum_reg[3]_i_1, CellType: CARRY4
-	Cell: adder1/sum_reg[7]_i_1, CellType: CARRY4
-	Cell: adder1/sum_reg[8]_i_1, CellType: CARRY4
-\end{lstlisting}
-
 \begin{lstlisting}[language=java, caption={Finding and storing carry chains.}, label={lst:carry_chains}]
@@
 while (!carryCells.isEmpty()) {
-    // Arbitrarily set "currentCell" pointer to a cell in the list
+    // Set "currentCell" pointer to an arbitrary cell in list
     EDIFCellInst currentCell = carryCells.get(0);
@@
         if (net.isGND()) {
-            // Found this chain anchor!
-            break;
+            break; // Found this chain anchor!
         }
@@
 
-    // Now we have the chain anchor as currentCell.
+    // Now currentCell points at this chain's anchor.
+
     // Now traverse in the opposite direction to find the chain tail.
@@
         if (sourcePort == null) {
-            // Found this chain's tail!
-            break;
+            break; // Found this chain's tail!
         }
@@
 \end{lstlisting}
-\newpage
-
-\subsection{Packing}
-\label{subsec:packing}
-Now that we have our \texttt{PrepackedDesign} object keeping track of multi-cell structures on the \texttt{edif} level, we can start packing them into \texttt{SiteInst} objects on the \texttt{design} level.
-Below are some of the most relevant classes from the \texttt{design} package for this task.
-\begin{itemize}
-    \item \texttt{Cell}: A cell corresponds to the leaf cell within the logical netlist \texttt{EDIFCellInst} and provides a mapping to a physical location BEL on the device.
-A cell can be created directly out of an \texttt{EDIFCellInst} to inherit all of its \texttt{edif} properties on the \texttt{design} level.
-    \item \texttt{Net}: Represents the physical net to be routed (both inter-site and intra-site).
-When an \texttt{Cell} is created out of an \texttt{EDIFCellInst}, the \texttt{Net}s are automatically created out of its corresponding \texttt{EDIFNet}s.
-    \item \texttt{SiteInst}: An instance of a \texttt{Site} on the \texttt{Device}.
-Carries the mapping information between the \texttt{BEL}s in a \texttt{Site} and the \texttt{Cell}s assigned to them.
-Also keeps track of the intra-Site routing information within (\texttt{Net}s, \texttt{SitePinInst}s, \texttt{SitePIP}s, etc.).
-    \item \texttt{SitePIP}: Represents the fuses in routing \texttt{BEL}s. 
-    \item \texttt{SitePinInst}: An instance of a \texttt{SitePin} on a \texttt{Site}. These objects serve as the interface between intra-Site routing and general inter-Site routing. 
-\end{itemize}
-\columnbreak
-
-A \texttt{SiteInst} object can be created and its BELs populated by existing \texttt{EDIFHierCellInst} objects. 
-A \texttt{SiteInst} is necessarily placed on a specific \texttt{Site} upon creation. 
-Thus, the packer necessarily gives all \texttt{SiteInst}s an initial placement before randomization via simulated annealing in the following placement stage. 
-In packing, will simply assign \texttt{SiteInst}s pseudorandomly onto device \texttt{Site}s coordinate order as they are created (e.g.
-first SiteInst onto \texttt{SLICEL\_X0Y0}, second \texttt{SiteInst} onto \texttt{SLICEL\_X0Y1}, etc.).
 
 \end{multicols}
-\begin{lstlisting}[caption=\texttt{SiteInst} constructor and methods.]
-SiteInst Constructor: 
-    SiteInst(String name, Design design, SiteTypeEnum type, Site site)
-
-Most relevant SiteInst Methods:
-    createCell(EDIFHierCellInst inst, BEL bel) // Populating the SiteInst BELs with Cells
-    unplace() // Unplacing the SiteInst from its current Site
-    place(Site site) // Placing an unplaced SiteInst onto a Site
-    routeSite() // Attempt to automatically route all intra-Site nets (manual intervention likely required)
-    routeIntraSiteNet(Net net, BELPin src, BELPin snk) // Manually route an intra-Site net
-
-Example:
-    SiteInst si = new SiteInst("mySiteInst", design, SiteTypeEnum.SLICEL, device.getSite("SLICEL_X0Y1"));
-    si.unplace();
-    si.place(device.getSite("SLICEL_X15Y33"));
-    // In Simulated Annealing, SiteInst objects will be unplaced() and placed() many times to converge to an optimal solution.
-
-\end{lstlisting}
-
-\begin{lstlisting}[language=java, caption={Packing DSPCascades into \texttt{DSP48E1} \texttt{SiteInst}s.}, label={lst:carry_chains}]
-private List<List<SiteInst>> packDSPCascades(List<List<EDIFHierCellInst>> EDIFDSPCascades) throws IOException {
-    List<List<SiteInst>> siteInstCascades = new ArrayList<>();
-    writer.write("\n\nPacking DSP Cascades... (" + EDIFDSPCascades.size() + ")");
-    for (List<EDIFHierCellInst> cascade : EDIFDSPCascades) {
-        List<SiteInst> siteInstCascade = new ArrayList<>();
-        // each DSP tile has 2 DSP sites on the Zynq 7000
-        writer.write("\n\tCascade Size: (" + cascade.size() + "), Cascade Anchor: "
-                + cascade.get(0).getFullHierarchicalInstName());
-        // Select a Site location on the device. 
-        Site anchorSite = selectDSPAnchorSite(cascade.size());
-        SiteTypeEnum siteType = SiteTypeEnum.DSP48E1;
-        for (int i = 0; i < cascade.size(); i++) {
-            writer.write("\n\t\tPlacing DSP: " + cascade.get(i).getFullHierarchicalInstName());
-            Site site = (i == 0) ? anchorSite
-                    : device.getSite("DSP48_X" + anchorSite.getInstanceX() + "Y" + (anchorSite.getInstanceY() + i));
-            // Create the SiteInst object
-            SiteInst si = new SiteInst(cascade.get(i).getFullHierarchicalInstName(), design, siteType, site);
-            // Populate the SiteInst with a DSP Cell. 
-            si.createCell(cascade.get(i), si.getBEL("DSP48E1"));
-            occupiedSites.get(siteType).add(site);
-            availableSites.get(siteType).remove(site);
-            si.routeSite();
-            siteInstCascade.add(si);
-        }
-        siteInstCascades.add(siteInstCascade);
-    }
-    return siteInstCascades;
-} // end packDSPCascades()
-\end{lstlisting}
-
-\newpage
-\begin{lstlisting}[language=java, caption={Packing LUTFFPairGroups into \texttt{SLICEL} \texttt{SiteInst}s}]
-private List<SiteInst> packLUTFFPairGroups(
-        Map<Pair<String, String>, LUTFFGroup> LUTFFEnableResetGroups) throws IOException {
-    List<SiteInst> LUTFFSiteInsts = new ArrayList<>();
-    writer.write("\n\nPacking LUT-FF Pair Groups... (" + LUTFFEnableResetGroups.size() + ")");
-    for (Map.Entry<Pair<String, String>, LUTFFGroup> entry : LUTFFEnableResetGroups.entrySet()) {
-        Pair<String, String> netPair = entry.getKey();
-        String s1 = String.format("\n\tCENet: %-50s RNet: %-50s", netPair.key(), netPair.value());
-        writer.write(s1);
-        LUTFFGroup lutffgroup = entry.getValue();
-        for (List<Pair<EDIFHierCellInst, EDIFHierCellInst>> LUTFFPairs : splitIntoGroups(lutffgroup.group(), 4)) {
-            Site selectedSite = selectCLBSite();
-            SiteInst si = new SiteInst(LUTFFPairs.get(0).value().getFullHierarchicalInstName(), design,
-                    selectedSite.getSiteTypeEnum(), selectedSite);
-            for (int i = 0; i < LUTFFPairs.size(); i++) {
-                EDIFHierCellInst ff = LUTFFPairs.get(i).value();
-                EDIFHierCellInst lut = LUTFFPairs.get(i).key();
-                si.createCell(ff, si.getBEL(FF_BELS[i]));
-                if (lut != null)
-                    si.createCell(lut, si.getBEL(LUT6_BELS[i]));
-            }
-            si.routeSite();
-            List<String> lutOPIP = new ArrayList<>(List.of("AUSED", "BUSED", "CUSED", "DUSED"));
-            // DOES NOT SUPPORT STACKED LUTS OR LUT5s!
-            for (int i = 0; i < LUTFFPairs.size(); i++) {
-                EDIFHierCellInst lut = LUTFFPairs.get(i).key();
-                if (lut != null) {
-                    EDIFHierPortInst lutOPort = lut.getPortInst("O");
-                    if (lutOPort.getHierarchicalNet().getLeafHierPortInsts(false, true).size() > 1) {
-                        si.addSitePIP(si.getSitePIP(lutOPIP.get(i), "0"));
-                    }
-                }
-            }
-            rerouteFFClkSrCeNets(si);
-            LUTFFSiteInsts.add(si);
-        }
-    }
-    return LUTFFSiteInsts;
-} // end packLUTFFPairGroups()
-\end{lstlisting}
-
-\begin{lstlisting}[language=java, caption={Packing LUTGroups into \texttt{SLICEL} \texttt{SiteInst}s}]
-protected String[] LUT6_BELS = new String[] { "A6LUT", "B6LUT", "C6LUT", "D6LUT" };
-
-private List<SiteInst> packLUTGroups(List<List<EDIFHierCellInst>> LUTGroups) throws IOException {
-    List<SiteInst> LUTSiteInsts = new ArrayList<>();
-    writer.write("\n\nPacking LUT Groups...");
-    for (List<EDIFHierCellInst> group : LUTGroups) {
-        Site selectedSite = selectCLBSite();
-        SiteInst si = new SiteInst(group.get(0).getFullHierarchicalInstName(), design,
-                selectedSite.getSiteTypeEnum(), selectedSite);
-        for (int i = 0; i < group.size(); i++) {
-            si.createCell(group.get(i), si.getBEL(LUT6_BELS[i]));
-        }
-        si.routeSite();
-        // for whatever reason, this does not activate DUSED PIP for
-        si.addSitePIP(si.getSitePIP("DUSED", "0"));
-        si.addSitePIP(si.getSitePIP("CUSED", "0"));
-        si.addSitePIP(si.getSitePIP("BUSED", "0"));
-        si.addSitePIP(si.getSitePIP("AUSED", "0"));
-        LUTSiteInsts.add(si);
-    }
-    return LUTSiteInsts;
-} // end packLUTGroups()
+\begin{lstlisting}[caption={Code Printout}]
+Anchor Cell: adder2/sum_reg[3]_i_1, CellType: CARRY4
+	Cell: adder2/sum_reg[7]_i_1, CellType: CARRY4
+	Cell: adder2/sum_reg[11]_i_1, CellType: CARRY4
+Anchor Cell: adder1/sum_reg[3]_i_1, CellType: CARRY4
+	Cell: adder1/sum_reg[7]_i_1, CellType: CARRY4
+	Cell: adder1/sum_reg[8]_i_1, CellType: CARRY4
 \end{lstlisting}
 \begin{multicols}{2}
-
-\subsection{Placement}
-    \label{subsec:placement}
-    Up until now we have only organized the logical \texttt{EDIFCellInst}s into \texttt{SiteInst}s. 
-    This is where simulated annealing actually begins where we actually place the \texttt{SiteInst}s onto physical \texttt{Site}s on the \texttt{device} level.

### 2025-05-14 latex/paper/sections/09-packing.tex
packing edits

@@
-\section{Placement}
-\end{multicols}
-{
-    \centering
-    \includegraphics[width=0.8\columnwidth]{figures/substages.png}
-    \captionof{figure}{The data classes populated at each substage: \texttt{PrepackedDesign}, \texttt{PackedDesign}, and \texttt{PlacedDesign}.}
-    \label{fig:substages}
-}
-\begin{multicols}{2}
-\label{sec:simulated_annealing}
-With a basic understanding of FPGA architecture, design placement, and RapidWright, we have all the necessary pieces to implement our SA placer. 
-Here we outline in detail each substage of our implementation: PrePacking, Packing, and Placement. 
-Shown in Figure \ref{fig:edif_design_device} is an overview of the placement workflow. 
-Figure \ref{fig:substages} shows the data structures of RapidWright objects that are populated at each stage: \texttt{PrepackedDesign}, which is a group of data structures around \texttt{EDIFHierCellInst}s, \texttt{PackedDesign}, which is a group fo data structures around \texttt{SiteInst}s, and finally, \texttt{PlacedDesign}, which is simply captured by the final RapidWright \texttt{Design} object.
-
-\vspace{0.5cm}
-{
-    \centering
-    \includegraphics[width=0.9\columnwidth]{figures/edif_design_device.png}
-    \captionof{figure}{Our placement workflow}
-    \label{fig:edif_design_device}
-}
-
-\subsection{Prepacking}
-\label{subsec:prepacking}
-
-The first step in our placement flow is \textbf{prepacking}. 
-Recall from the 7-Series architecture that there are certain multi-cell structures that must adhere to certain placements constraints to ensure legality, and by design, to minimize wirelength. 
-The job of the prepacker is to traverse the raw EDIF netlist, detect these multi-cell structures, and consolidate these cells into clusters or groups of clusters that naturally reflect these placement constraints.
-
-Recall that \texttt{CARRY4} chains must necessarily be placed vertically and consecutively across a column of SLICEs in ascending order. 
-Likewise, \texttt{DSP48E1} cascades must necessarily be placed vertically and consecutively across a column of \texttt{DSP48E1} Sites in ascending order. 
-A LUT-FF pair may be placed freely, but should be placed in the same lane within the same SLICE to minimize wirelength.
-
-The raw EDIF netlist only tells us the list of nets and the cell ports that they connect to. 
-It does not report the presence of any multi-cell structures (\texttt{CARRY4} chains, etc.). 
-Thus, we must traverse the netlist to detect these multi-cell structures and store that structure information in a class we will call \texttt{PrepackedDesign}.
-
-The code snippet in \ref{lst:carry_chains} shows how one can detect and collect these \texttt{CARRY4} chains using RapidWright. 
-We first collect the cells in the design that are of type \texttt{CARRY4}, then iteratively traverse their Carry-Out (\texttt{CO}) to Carry-In (\texttt{CI}) nets to find incident \texttt{CARRY4} cells.
-Each \texttt{CARRY4} chain has an anchor cell and a tail cell where the chain terminates.
-The anchor is found when the \texttt{CI} net connects to Ground (\texttt{GND}), while the tail is found when the \texttt{CO} port is null. 
-We can further detect if there are \texttt{LUT}s or \texttt{FF}s connected to the \texttt{CARRY4} cell and store that information in a data structure we will call \texttt{CarryCellGroup} as defined in figure \ref{fig:substages}.
-This will help us in knowing which cells can be packed together into the same \texttt{Site} in the subsequent stages.
-
-Similarly, \texttt{DSP48E1} cascades can be found and collected by traversing the \texttt{PCOUT} \texttt{ACOUT} and \texttt{BCOUT} nets.
-\texttt{LUT-FF} pairs can be found by inspecting the LUT output (O) net and checking for FF input (DI) ports. 
-We can bucket these LUT-FF pairs by finding the set of unique \texttt{CE} \texttt{SR} net pairs to know which group of LUT-FF pairs can be placed within the same \texttt{Site}.
-
-We detect the presence of these multi-cellular structures and consolidate that information into our \texttt{PrepackedDesign} object in preparation for the following Packing stage.
-
-\end{multicols}
-{
-    \centering
-    \includegraphics[width=0.8\columnwidth]{figures/carry_chain_traversal.png}
-    \captionof{figure}{A netlist with two \texttt{CARRY4} chains, each of size 3}
-    \label{fig:carry_chain_traversal}
-}
-\begin{multicols}{2}
-
-\begin{lstlisting}[caption={Code Printout}]
-Anchor Cell: adder2/sum_reg[3]_i_1, CellType: CARRY4
-	Cell: adder2/sum_reg[7]_i_1, CellType: CARRY4
-	Cell: adder2/sum_reg[11]_i_1, CellType: CARRY4
-Anchor Cell: adder1/sum_reg[3]_i_1, CellType: CARRY4
-	Cell: adder1/sum_reg[7]_i_1, CellType: CARRY4
-	Cell: adder1/sum_reg[8]_i_1, CellType: CARRY4
-\end{lstlisting}
-
-\begin{lstlisting}[language=java, caption={Finding and storing carry chains.}, label={lst:carry_chains}]
-Design design = Design.readCheckpoint("synth.dcp")
-EDIFNetlist netlist = design.getNetlist();
-List<EDIFCellInst> ecis = netlist.getAllLeafCellInstances();
-
-// Select only the carry cells.
-List<EDIFCellInst> carryCells = new ArrayList<>();
-for (EDIFCellInst eci : ecis) {
-    if (eci.getCellName().equals("CARRY4"))
-        carryCells.add(eci);
-}
-
-// Find and remove carry chains until the list is empty
-List<List<EDIFCellInst>> carryChains = new ArrayList<>();
-while (!carryCells.isEmpty()) {
-    // Arbitrarily set "currentCell" pointer to a cell in the list
-    EDIFCellInst currentCell = carryCells.get(0);
-
-    // Find this carry chain anchor.
-    // Traverse the Carry-In (CI) to Carry-Out (CO) nets.
-    // Anchor is found when net on the CI Port is Ground.
-    while (true) {
-        System.out.println(currentCell);
-        // Access the CI port on this cell.
-        EDIFPortInst sinkPort = currentCell.getPortInst("CI");
-        // Access the net on this CI port.
-        EDIFNet net = sinkPort.getNet();
-        if (net.isGND()) {
-            // Found this chain anchor!
-            break;
-        }
-        // Get all ports on this net.
-        List<EDIFPortInst> netPorts = net.getPortInsts();
-        for (EDIFPortInst netPort : netPorts) {
-            // Access the port belonging to another carry cell.
-            EDIFCellInst sourceCell = netPort.getCellInst();
-            if (sourceCell.getCellName().equals("CARRY4")) {
-                // Move the "currentCell" pointer
-                currentCell = sourceCell;
-                break;
-            }
-        }
-    }
-
-    // Now we have the chain anchor as currentCell.
-    // Now traverse in the opposite direction to find the chain tail.
-    // Tail is found when the CO Port is null.
-    // Collect the chain cells into an ordred list.
-    List<EDIFCellInst> currentChain = new ArrayList<>();
-    currentChain.add(currentCell);
-    while (true) {
-        EDIFPortInst sourcePort = currentCell.getPortInst("CO[3]");
-        if (sourcePort == null) {
-            // Found this chain's tail!
-            break;
-        }
-        EDIFNet net = sourcePort.getNet();
-        List<EDIFPortInst> netPorts = net.getPortInsts();
-        for (EDIFPortInst netPort : netPorts) {
-            EDIFCellInst sinkCell = netPort.getCellInst();
-            if (netPort.getName().equals("CI") &&
-                    sinkCell.getCellName().equals("CARRY4")) {
-                currentCell = sinkCell;
-                // Add the cell to the chain list.
-                currentChain.add(currentCell);
-                break;
-            }
-        }
-    }
-    // Add currentChain to the list of chains
-    carryChains.add(currentChain);
-    // Remove currentChain from the list of cells
-    carryCells.removeAll(currentChain);
-} // end while()
-
-// Print out the carry chains. 
-for (List<EDIFCellInst> chain : chains) {
-    for (int i = 0; i < chain.size(); i++) {
-        EDIFCellInst carry = chain.get(i);
-        if (i == 0) {
-            writer.write("\nAnchor Cell: " + carry.getName() + 
-                ", CellType: " + carry.getCellName());
-        } else {
-            writer.write("\n\tCell: " + carry.getName() + 
-                ", CellType: " + carry.getCellName());
-        }
-    }
-}
-\end{lstlisting}
-\newpage
-
-\subsection{Packing}
-\label{subsec:packing}
+\section{Packing}
+\label{sec:packing}
 Now that we have our \texttt{PrepackedDesign} object keeping track of multi-cell structures on the \texttt{edif} level, we can start packing them into \texttt{SiteInst} objects on the \texttt{design} level.
@@
 
-One can think of a \texttt{design} \texttt{SiteInst} as a movable square peg and the corresponding \texttt{device} \texttt{Site} as a fixed square hole. 
-Both \texttt{SiteInst}s and \texttt{Site}s come in various shapes or types (\texttt{SLICEL}, \texttt{SLICEM}, \texttt{DSP48E1}, etc.) and placement is only allowed between compatible pairs. 
-For example, a \texttt{RAMB48E1} \texttt{SiteInst} can only be placed on a \texttt{RAMB48E1} \texttt{Site}.
-A \texttt{DSP48E1} \texttt{SiteInst} can only be placed on a \texttt{DSP48E1} \texttt{Site}.
-A \texttt{SLICEL} \texttt{SiteInst} may occupy either a \texttt{SLICEL} or a \texttt{SLICEM} \texttt{Site}, whereas a \texttt{SLICEM} \texttt{SiteInst} can only be placed on a \texttt{SLICEM} \texttt{Site}.
-These constraints must always be followed as the \texttt{SiteInst}s move across the device throughout the placement stage to ensure legality.
+% % One can think of a \texttt{design} \texttt{SiteInst} as a movable square peg and the corresponding \texttt{device} \texttt{Site} as a fixed square hole. 
+% % Both \texttt{SiteInst}s and \texttt{Site}s come in various shapes or types (\texttt{SLICEL}, \texttt{SLICEM}, \texttt{DSP48E1}, etc.) and placement is only allowed between compatible pairs. 
+% % For example, a \texttt{RAMB48E1} \texttt{SiteInst} can only be placed on a \texttt{RAMB48E1} \texttt{Site}.
+% % A \texttt{DSP48E1} \texttt{SiteInst} can only be placed on a \texttt{DSP48E1} \texttt{Site}.
+% % A \texttt{SLICEL} \texttt{SiteInst} may occupy either a \texttt{SLICEL} or a \texttt{SLICEM} \texttt{Site}, whereas a \texttt{SLICEM} \texttt{SiteInst} can only be placed on a \texttt{SLICEM} \texttt{Site}.
+% % These constraints must always be followed as the \texttt{SiteInst}s move across the device throughout the placement stage to ensure legality.
 
-Listing \ref{lst:siteinst} shows how to use the \texttt{SiteInst} object and how it ties together \texttt{Site}s, \texttt{Cell}s, \texttt{BEL}s, and intra-Site \texttt{Net}s. 
+Listing \ref{lst:siteinst} shows how to use the \texttt{SiteInst} object and how it ties \texttt{Site}s, \texttt{Cell}s, \texttt{BEL}s, and intra-Site \texttt{Net}s together. 
 The super-specifics of the packing process via RapidWright is too involved for the scope of this paper, but just to give an impression of how the \texttt{design} classes are used, Listing \ref{lst:single_carry_chains}, \ref{lst:carry_chains}, \ref{lst:carry_nets}, and \ref{lst:ff_nets} together show how \texttt{CarryCellGroup}s are packed into \texttt{SLICEL} \texttt{Sites} in our working code.
-Packing \texttt{DSP48E1} cascades, \texttt{RAMB48E1} Sites, and non-\texttt{CARRY4} \texttt{SLICEL} Sites follow a similar process in instantiating \texttt{SiteInst}s and populating their \texttt{BEL}s with \texttt{Cell}s using existing \texttt{EDIFHierCellInst}s extracted from the \texttt{EDIFNetlist}.
+
+Packing the other \texttt{Cell} structures into \texttt{Site}s follow a similar process. 
+For each cell cluster, we create a \texttt{SiteInst} and populate its \texttt{BEL}s with the corresponding \texttt{Cell}s in the cluster, assign it a physical \texttt{Site} on the \texttt{device}.
+We further consolidate these \texttt{SiteInst}s into clusters or lists of \texttt{SiteInst}s for multi-Site structures like \texttt{CarryCellGroup} chains and DSP cascades. 
+For each \texttt{SiteInst}, we assign it a physical \texttt{Site} on the \texttt{device} and route its internal intra-Site nets. 
+The end result is a \texttt{PackedDesign} object which is a group of data structures solely around \texttt{SiteInst}s like shown in figure \ref{fig:substages}.
+At this point we no longer need to worry about individual \texttt{BEL}s, \texttt{Cell}s, or intra-Site \texttt{Nets} as they have now been packaged and taped-off into their respective \texttt{SiteInst}s.
+These \texttt{SiteInst} become the new atomic level components in the following Placement stage.
 
@@
 \begin{multicols}{2}
-
-\subsection{Placement}
-    \label{subsec:placement}
-    Up until now we have only organized the logical \texttt{EDIFHierCellInst}s into \texttt{SiteInst}s. 
-    This is where simulated annealing actually begins where we actually place the \texttt{SiteInst}s onto physical \texttt{Site}s on the \texttt{device} level.

### 2025-05-14 latex/paper/sections/10-placement.tex
packing edits

--- 2025-05-08 latex/paper/sections/10-simulated_annealing.tex
@@
 \section{Placement}
-\label{sec:simulated_annealing}
-With a basic understanding of FPGA architecture, design placement, and RapidWright, we have all the necessary pieces to implement our SA placer. 
-Here we outline in detail each substage of our implementation: PrePacking, Packing, and Placement. 
-Shown in Figure \ref{fig:edif_design_device} is an overview of the placement workflow. 
-Figure \ref{fig:substages} shows the data structures of RapidWright objects that are populated at each stage.
-
-{
-    \centering
-    \includegraphics[width=0.9\columnwidth]{figures/edif_design_device.png}
-    \captionof{figure}{Our placement workflow}
-    \label{fig:edif_design_device}
-}
-
-\end{multicols}
-{
-    \centering
-    \includegraphics[width=0.8\columnwidth]{figures/substages.png}
-    \captionof{figure}{The data classes populated at each substage: PrepackedDesign, PackedDesign, and PlacedDesign.}
-    \label{fig:substages}
-}
-\begin{multicols}{2}
-
-\subsection{Prepacking}
-\label{subsec:prepacking}
-
-The first step in our placement flow is \textbf{prepacking}. 
-Recall from the 7-Series architecture that there are certain cell structures that must adhere to certain placements constraints to ensure legality, and by design, to minimize wirelength. 
-The job of the prepacker is to traverse the raw EDIF netlist, detect these cell structures, and consolidate these cells into clusters or groups of clusters that naturally reflect these placement constraints.
-
-Recall that CARRY4 chains must necessarily be placed vertically and consecutively across a column of SLICEs in ascending order. 
-Likewise, \texttt{DSP48E1} cascades must necessarily be placed vertically and consecutively across a column of DSP48E1 Sites in ascending order. 
-A LUT-FF pair may be placed freely, but should be placed in the same lane within the same SLICE to minimize wirelength.
-
-The raw EDIF netlist only tells us the list of nets and the cell ports that they connect to. 
-It does not report the presence of any macro cell structures (CARRY4 chains, etc.). 
-Thus, we must traverse the netlist to detect these cell structures and store that structure information in a class we will call \texttt{PrepackedDesign}.
-
-A group of LUT-FF pairs may be placed in the same SLICE, with the constraint that the FFs must share the same Clock-Enable (CE) and Set-Reset (SR) nets.
-Clustering LUT-FF pairs like this can reduce the redundancy of having to route the same CE and SR nets to many SLICEs by routing the nets to fewer SLICEs and connecting them to the individual FFs within via intra-Site routing, and at the same time, packing greater logic density over a smaller area of the device.
-Recall that up to eight FFs may be placed in the same SLICE, thus theoretically, up to eight LUT-FF pairs can be placed in the same SLICE.
-
-{
-    \centering
-    \includegraphics[width=\columnwidth]{figures/slicel.png}
-    \captionof{figure}{A SLICEL Site}
-    \label{fig:slicel}
-}
-
-Utilizing all 8 LUT-FF lanes in a SLICE can help reduce device area utilization and minimize wirelength, but \emph{too much} logic density in an area can contribute to general routing congestion. 
-Furthermore, attempting to fill all 8 lanes in a SLICE requires meticulous adherence to conditional LUT constraints. 
-Recall that a LUT can accommodate one 6-input boolean function or two 5-input boolean functions sharing the same inputs or any two 3-input or less boolean functions regardless of shared inputs. 
-A LUT6 Cell (a LUT with 6-inputs) will actually occupy two LUT-FF lanes in a SLICEL, rendering one of the FF BELs in either lane ineligible for another LUT-FF pair. 
-To simplify the problem space, we will only fill up to four LUT-FF lanes in any given SLICE.
-
-\begin{lstlisting}[language=java, caption={Netlist extraction and traversal}, label={lst:netlist_extract}]
-Design design = Design.readCheckpoint("synth.dcp")
-EDIFNetlist netlist = design.getNetlist();
-List<EDIFCellInst> ecis = netlist.getAllLeafCellInstances();
-
-// Select only the carry cells.
-for (EDIFCellInst eci : ecis) {
-    if (eci.getCellName() == "CARRY4")
-        carryCells.add(eci);
-}
-
-// Find and remove carry chains until the list is empty
-while (!carryCells.isEmpty()) {
-    // Arbitrarily set "currentCell" pointer to a cell in the list
-    EDIFCellInst currentCell = carryCells.get(0);
-
-    // Find this carry chain anchor.
-    // Traverse the Carry-In (CI) to Carry-Out (CO) nets.
-    // Anchor is found when net on the CI Port is Ground.
-    while (true) {
-        // Access the CI port on this cell.
-        EDIFPortInst sinkPort = currentCell.getPortInst("CI");
-        // Access the net on this CI port.
-        EDIFNet net = sinkPort.getNet();
-        if (net.isGND())
-            // Found this chain anchor!
-            break;
-        // Get all ports on this net.
-        List<EDIFPortInst> netPorts = net.getPortInsts();
-        for (EDIFPortInst netPort : netPorts) {
-            // Access the port belonging to another carry cell.
-            EDIFCellInst sourceCell = netPort.getCellInst();
-            if (sourceCell.getCellName() == "CARRY4") {
-                // Move the "currentCell" pointer
-                currentCell = sourceCell;
-            }
-        }
-    }
-
-    // Now we have the chain anchor as currentCell.
-    // Now traverse in the opposite direction to find the chain tail.
-    // Tail is found when the CO Port is null.
-    // Collect the chain cells into an ordred list.
-    List<EDIFCellInst> currentChain = new ArrayList<>();
-    while (true) {
-        EDIFPortInst sourcePort = currentCell.getPortInst("CO[3]");
-        if (sourcePort == null)
-            // Found this chain's tail!
-            break;
-        EDIFNet net = sourcePort.getNet();
-        List<EDIFPortInst> sinkPorts = net.getPortInsts();
-        for (EDIFPortInst sinkPort : sinkPorts) {
-            EDIFCellInst sinkCell = sinkPort.getCellInst();
-            if (sinkCell.getCellName() == "CARRY4") {
-                currentCell = sinkCell;
-                // Add the cell to the chain list.
-                currentChain.add(currentCell);
-            }
-        }
-    }
-
-    // Add currentChain to the list of chains
-    carryChains.add(currentChain);
-}
-\end{lstlisting}
-
-\subsection{Packing}
-    \label{subsec:packing}
-
-\subsection{Placement}
-    \label{subsec:placement}
-    Up until now we have only organized the logical \texttt{Cells} into \texttt{SiteInsts}. 
-    This is where simulated annealing actually begins.
+    \label{sec:placement}
+    Up until now we have only organized the logical \texttt{EDIFHierCellInst}s into \texttt{SiteInst}s. 
+    This is where simulated annealing actually begins where we actually place the \texttt{SiteInst}s onto physical \texttt{Site}s on the \texttt{device} level.

### 2025-05-18 latex/paper/sections/10-placement.tex
sim broken again?

@@
 \label{sec:placement}
-    Up until now we have only organized the logical \texttt{EDIFHierCellInst}s into \texttt{SiteInst}s. 
-    This is where simulated annealing actually begins where we actually place the \texttt{SiteInst}s onto physical \texttt{Site}s on the \texttt{device} level.
+Up until now, we have only dealt with technology problems where we consider the problem space and adress the constraints of the hardware architecture. 
+For every constraint, we devise a solution or a simplification that sacrifices some optimization to reduce the complexity of the problem space.
+
+Now that we have dealt with the bulk of the Cell-BEL level constraints and squared them away into self-contained \texttt{SiteInst}s, we can now apply some well-known optimization algorithms to finally address our optimization objective, which is to place these \texttt{SiteInst}s onto the device in a way that minimizes wirelength.
+
+In this paper we implement and present a Simulating Annealing (SA) placer that operates at the \texttt{SiteInst} level.
+
+\subsection{Simulated Annealing}
+\label{subsec:simulated_annealing}
+
+\begin{lstlisting}[language=java, caption={Simulated Annealing pseudocode}, label={lst:sa_pseudocode}]
+
+\end{lstlisting}

### 2025-05-18 latex/paper/sections/12-future_work.tex
sim broken again?

\section{Future Work}
In this paper we have made many simplifications to the problem space just to make the placer easier to implement.
For example, our placer in its current state does not take advantage of \texttt{SLICEL}/\texttt{SLICEM} homogeneity and simply maps all SLICE \texttt{SiteInst}s onto \texttt{SLICEL}s. 
Recall that the SLICE Sites in Xilinx FPGAs typically come in a 75-25\% split between \texttt{SLICEL}s and \texttt{SLICEM}s. 
This means that we have rendered about 25\% of the CLB fabric unusable which will inevitably hurt wirelength minimization during placement since the \texttt{SiteInst}s must be spread over a larger area.

There is also no rule saying we must pack \texttt{Cells} into \texttt{SiteInst}s before placement or that we follow a strict prepacking-packing-placement flow.

The latest editions of Vivado performs BEL-centric placement without necessarily locking \texttt{Cell}s into \texttt{Site}s, allowing a higher granularity of movement of \texttt{Cell}s during placement.

### 2025-05-18 java/src/main/java/placer/Main.java

@@ -78,7 +78,7 @@ public class Main {
             // Sites via simulated annealing, analytical, electrostatic placement, etc.
             // works entirely on the SiteInst/Site/Tile level.
             List<PlacerAnnealRandom> SAPlacers = new ArrayList<PlacerAnnealRandom>();
-            // SAPlacers.add(new PlacerAnnealRandom(rootDir, design, device, region));
+            SAPlacers.add(new PlacerAnnealRandom(rootDir, design, device, region));
             SAPlacers.add(new PlacerAnnealHybrid(rootDir, design, device, region));
             // SAPlacers.add(new PlacerAnnealMidpoint(rootDir, design, device, region));
             // SAPlacers.add(new PlacerGreedyRandom(rootDir, design, device, region));

### 2025-05-19 java/src/main/java/placer/Main.java
routing errors at chain anchors and tails, pin conflicts between vcc and gnd

@@ -25,6 +25,8 @@ import com.xilinx.rapidwright.device.ClockRegion;
 import com.xilinx.rapidwright.device.Tile;
 import com.xilinx.rapidwright.device.TileTypeEnum;
 
+import com.xilinx.rapidwright.design.DesignTools;
+
 import org.ejml.data.DMatrixRMaj;
 import org.ejml.dense.row.CommonOps_DDRM;
 
@@ -87,7 +89,7 @@ public class Main {
             for (PlacerAnnealRandom placer : SAPlacers) {
                 System.out.println("\n\nStarting " + placer.getPlacerName() + "... \n\n");
                 placer.makeOutputDirs(placer.getPlacerName());
-                placer.initCoolingSchedule(10000.0d, 0.98d, 300);
+                placer.initCoolingSchedule(10000.0d, 0.98d, 50);
                 placer.run(packedDesign);
             }
 

### 2025-05-19 java/src/main/java/placer/PackerBasic1.java

@@ -199,6 +199,10 @@ public class PackerBasic1 extends Packer {
                 si.createCell(cascade.get(i), si.getBEL("DSP48E1"));
                 occupiedSites.get(siteType).add(site);
                 availableSites.get(siteType).remove(site);
+                for (int j = 0; j < 48; j++) {
+                    SitePinInst pspi = new SitePinInst(true, "P" + j, si);
+                    si.addPin(pspi);
+                }
                 si.routeSite();
                 siteInstCascade.add(si);
             }
@@ -438,6 +442,36 @@ public class PackerBasic1 extends Packer {
         for (String FF : FF_BELS)
             if (si.getCell(FF) == null)
                 si.addSitePIP(si.getSitePIP(FF.charAt(0) + "OUTMUX", "XOR"));
+        // if CARRY S0-S3 nets pass-through LUT
+        String[] letters = new String[] { "A", "B", "C", "D" };
+        for (String LUT : LUT6_BELS) {
+            String letter = String.valueOf(LUT.charAt(0));
+
+        }
+        for (int i = 0; i < 4; i++) {
+            if (si.getNetFromSiteWire(letters[i] + "6LUT_O6").isGNDNet()) {
+
+            }
+
+            // might need to add another conditional to check if pin/wire is even used
+            if (si.getCell(LUT6_BELS[0]) == null) {
+
+                // Map<String, Net> nets = si.getSiteWireToNetMap();
+                // for (Map.Entry<String, Net> net : nets.entrySet()) {
+                // System.out.println("SiteWire: " + net.getKey() + ", Net: " +
+                // net.getValue().getName());
+                // }
+
+                SitePinInst spi = new SitePinInst(false, letters[i] + "6", si);
+                si.addPin(spi);
+                si.routeIntraSiteNet(
+                        si.getNetFromSiteWire(letters[i] + "6LUT_O6"), // BEL
+                        si.getSitePinInst(letters[i] + "6").getBELPin(),
+                        si.getBELPin("CARRY4", "S" + i));
+            } else {
+                // System.out.println(si.getCell(LUT).getName());
+            }
+        }
     } // end rerouteCarryNets()
 
     private void rerouteFFClkSrCeNets(SiteInst si) {

### 2025-05-21 java/src/main/java/placer/Main.java
just stick to 24 bit. passes sim and route

@@ -27,8 +27,8 @@ import com.xilinx.rapidwright.device.TileTypeEnum;
 
 import com.xilinx.rapidwright.design.DesignTools;
 
-import org.ejml.data.DMatrixRMaj;
-import org.ejml.dense.row.CommonOps_DDRM;
+// import org.ejml.data.DMatrixRMaj;
+// import org.ejml.dense.row.CommonOps_DDRM;
 
 public class Main {
 
@@ -89,7 +89,7 @@ public class Main {
             for (PlacerAnnealRandom placer : SAPlacers) {
                 System.out.println("\n\nStarting " + placer.getPlacerName() + "... \n\n");
                 placer.makeOutputDirs(placer.getPlacerName());
-                placer.initCoolingSchedule(10000.0d, 0.98d, 50);
+                placer.initCoolingSchedule(10000.0d, 0.98d, 300);
                 placer.run(packedDesign);
             }
 
@@ -121,48 +121,48 @@ public class Main {
         System.out.println("Design Cost: " + cost);
     }
 
-    public static void helloEJML() {
-        // 1. Create two 2x2 matrices A and B
-        DMatrixRMaj A = new DMatrixRMaj(new double[][] {
-                { 1, 2 },
-                { 3, 4 }
-        });
-        DMatrixRMaj B = new DMatrixRMaj(new double[][] {
-                { 5, 6 },
-                { 7, 8 }
-        });
-
-        // 2. Matrix addition C = A + B
-        DMatrixRMaj C = new DMatrixRMaj(A.numRows, A.numCols);
-        CommonOps_DDRM.add(A, B, C);
-
-        // 3. Matrix multiplication D = A * B
-        DMatrixRMaj D = new DMatrixRMaj(A.numRows, B.numCols);
-        CommonOps_DDRM.mult(A, B, D);
-
-        // 4. Matrix inversion invA = A^-1
-        DMatrixRMaj invA = new DMatrixRMaj(A.numRows, A.numCols);
-        CommonOps_DDRM.invert(A, invA);
-
-        // 5. Solve Ax = b (for x)
-        // Here, b is a 2x1 column vector
-        DMatrixRMaj b = new DMatrixRMaj(new double[][] { { 1 }, { 2 } });
-        DMatrixRMaj x = new DMatrixRMaj(A.numRows, b.numCols);
-        CommonOps_DDRM.solve(A, b, x);
-
-        // Print results
-        System.out.println("Matrix A:");
[…]

### 2025-05-21 java/src/main/java/placer/PackerBasic1.java

@@ -199,10 +199,10 @@ public class PackerBasic1 extends Packer {
                 si.createCell(cascade.get(i), si.getBEL("DSP48E1"));
                 occupiedSites.get(siteType).add(site);
                 availableSites.get(siteType).remove(site);
-                for (int j = 0; j < 48; j++) {
-                    SitePinInst pspi = new SitePinInst(true, "P" + j, si);
-                    si.addPin(pspi);
-                }
+                // for (int j = 0; j < 48; j++) {
+                // SitePinInst pspi = new SitePinInst(true, "P" + j, si);
+                // si.addPin(pspi);
+                // }
                 si.routeSite();
                 siteInstCascade.add(si);
             }
@@ -448,30 +448,26 @@ public class PackerBasic1 extends Packer {
             String letter = String.valueOf(LUT.charAt(0));
 
         }
-        for (int i = 0; i < 4; i++) {
-            if (si.getNetFromSiteWire(letters[i] + "6LUT_O6").isGNDNet()) {
-
-            }
-
-            // might need to add another conditional to check if pin/wire is even used
-            if (si.getCell(LUT6_BELS[0]) == null) {
-
-                // Map<String, Net> nets = si.getSiteWireToNetMap();
-                // for (Map.Entry<String, Net> net : nets.entrySet()) {
-                // System.out.println("SiteWire: " + net.getKey() + ", Net: " +
-                // net.getValue().getName());
-                // }
-
-                SitePinInst spi = new SitePinInst(false, letters[i] + "6", si);
-                si.addPin(spi);
-                si.routeIntraSiteNet(
-                        si.getNetFromSiteWire(letters[i] + "6LUT_O6"), // BEL
-                        si.getSitePinInst(letters[i] + "6").getBELPin(),
-                        si.getBELPin("CARRY4", "S" + i));
-            } else {
-                // System.out.println(si.getCell(LUT).getName());
-            }
-        }
+        // for (int i = 0; i < 4; i++) {
+        // if (si.getNetFromSiteWire(letters[i] + "6LUT_O6").isGNDNet()) {
+        // }
+        // // might need to add another conditional to check if pin/wire is even used
+        // if (si.getCell(LUT6_BELS[0]) == null) {
+        // // Map<String, Net> nets = si.getSiteWireToNetMap();
+        // // for (Map.Entry<String, Net> net : nets.entrySet()) {
[…]

### 2025-05-22 latex/paper/sections/10-placement.tex
single site movement

--- 2025-05-14 latex/paper/sections/10-placement.tex
@@
 \label{sec:placement}
-    Up until now we have only organized the logical \texttt{EDIFHierCellInst}s into \texttt{SiteInst}s. 
-    This is where simulated annealing actually begins where we actually place the \texttt{SiteInst}s onto physical \texttt{Site}s on the \texttt{device} level.
+So far we have only dealt with technology problems where we consider the problem space and adress the constraints of the hardware architecture. 
+For every constraint, we devise a solution or a simplification to reduce the complexity of the problem space.
+
+Now that we have dealt with the bulk of these constraints and squared them away into self-contained \texttt{SiteInst}s, we can now apply some general optimization algorithms to finally address our optimization objective, which is to place these \texttt{SiteInst}s onto the device while minimizing wirelength.
+
+In this paper we implement and present a Simulating Annealing (SA) placer that operates at the \texttt{SiteInst} level.
+
+\subsection{Simulated Annealing}
+\label{subsec:simulated_annealing}
+
+\begin{lstlisting}[language=java, caption={Simulated Annealing pseudocode}, label={lst:sa_pseudocode}]
+public void placeDesign(PackedDesign packedDesign) {
+    // unplace the pseudorandom packed placement
+    unplaceAllSiteInsts(packedDesign);
+    // place randomly
+    randomInitialPlacement(packedDesign);
+    int moves = 0;
+    while (moves < 300) {
+        move(packedDesign)
+        updateTemperature();
+        moves++;
+    }
+}
+
+private void move(PackedDesign packedDesign) {
+    moveSiteChains(packedDesign.DSPSiteInstCascades);
+    moveSiteChains(packedDesign.CARRYSiteInstChains);
+    moveSingleSite(packedDesign.RAMSiteInsts);
+    moveSingleSite(packedDesign.CLBSiteInsts);
+}
+
+protected void moveSingleSite(List<SiteInst> sites) {
+    for (SiteInst si : sites) {
+        // We refer to the current Site si as the "home site"
+        // We refer to the proposed Site as the "away site"
+
+        // Find the list of Sites connected to the home site
+        List<Site> homeConns = findConnectedSites(si);
+
+        Site homeSite = si.getSite();
+        // Select an away site (random selection, midpoint, centroid, etc.)
+        Site awaySite = proposeSite();
+
+    }
+}
+
+protected void moveSingleSite(List<SiteInst> sites) {
+    for (SiteInst si : sites) {
+        SiteTypeEnum ste = si.getSiteTypeEnum();
+        List<Site> homeConns = findConnectedSites(si, null);
+        Site homeSite = si.getSite();
+        Site awaySite = proposeSite(si, homeConns, true);
+        SiteInst awaySi = occupiedSites.get(ste).get(awaySite);
+        double oldCost = 0;
+        double newCost = 0;
+        if (awaySi != null) {
+            List<Site> awayConns = findConnectedSites(awaySi, null);
+            oldCost += evaluateSite(homeConns, homeSite);
+            oldCost += evaluateSite(awayConns, awaySite);
+            newCost += evaluateSite(homeConns, awaySite);
+            newCost += evaluateSite(awayConns, homeSite);
+        } else {
+            oldCost += evaluateSite(homeConns, homeSite);
+            newCost += evaluateSite(homeConns, awaySite);
+        }
+        if (evaluateMoveAcceptance(oldCost, newCost)) {
+            if (awaySi != null) {
+                unplaceSiteInst(si);
+                unplaceSiteInst(awaySi);
+                placeSiteInst(si, awaySite);
+                placeSiteInst(awaySi, homeSite);
+            } else {
+                unplaceSiteInst(si);
+                placeSiteInst(si, awaySite);
+            }
+        }
+    }
+} // end randomMoveSingleSite()
+\end{lstlisting}
+
+{
+    \centering
+    \includegraphics[width=\columnwidth]{figures/placement/moveSingleSite.png}
+    \captionof{figure}{Single Site movement}
+    \label{fig:moveSingleSite}
+}

### 2025-05-30 latex/paper/sections/09-packing.tex
single site movement
chain swap pseudocode

@@
 Packing the other \texttt{Cell} structures into \texttt{Site}s follow a similar process. 
-For each cell cluster, we create a \texttt{SiteInst} and populate its \texttt{BEL}s with the corresponding \texttt{Cell}s in the cluster, assign it a physical \texttt{Site} on the \texttt{device}.
+For each cell cluster, we create a \texttt{SiteInst} and populate its \texttt{BEL}s with the corresponding \texttt{Cell}s in the cluster and assign it a physical \texttt{Site} on the \texttt{device}.
 We further consolidate these \texttt{SiteInst}s into clusters or lists of \texttt{SiteInst}s for multi-Site structures like \texttt{CarryCellGroup} chains and DSP cascades. 
@@
 These \texttt{SiteInst} become the new atomic level components in the following Placement stage. 
+This process is analogous to the containerization of loose goods in ocean freight forwarding.
 
@@
     SiteInst(String name, Design design, SiteTypeEnum type, Site site)
-
 Most relevant SiteInst Methods:
@@
 \end{lstlisting}
-
 \begin{multicols}{2}

### 2025-05-30 latex/paper/sections/10-placement.tex
site swap
chain swaps
chain swap pseudocode

@@
     \centering
-    \includegraphics[width=\columnwidth]{figures/placement/moveSingleSite.png}
-    \captionof{figure}{Single Site movement}
-    \label{fig:moveSingleSite}
+    \includegraphics[width=\columnwidth]{figures/placement/swapSingleSite.png}
+    \captionof{figure}{Single Site Swap Proposal}
+    \label{fig:swapSingleSite}
 }
+
+\begin{lstlisting}[language=java, caption={Chain Swapping}, label={lst:chain_swap_pseudocode}]
+protected void moveSiteChains(List<List<SiteInst>> chains) {
+for (List<SiteInst> current_chain : chains) {
+    int ch_size = current_chain.size();
+    /*
+
+    1) Draw a vertical window of equal size to the current chain. We will refer to this window as the "home window". 
+    2) Select a new candidate anchor for this chain.
+    2) Project the home window onto the candidate anchor with anchors coinciding. We will refer to this projected window as the "away window".
+    3) Check if there are any resident SiteInsts or SiteInst chains present in the away window. Check if the away window breaks any resident SiteInst chains. 
+    4)  IF (the away window breaks any chains): 
+            extend the window in either direction to 
+            fully accommodate all resident chains. 
+        ELSE: GOTO evaluateMove(). 
+    5) Project the extended away window back onto the original region with the tail of the window coinciding with the tail of the current chain. This projected window becomes the new home window. 
+    7) WHILE (the home window breaks any resident chains):
+        shift the window upwards by one Site. 
+        IF (number of shifts > home window size - ch_size):
+            reject the candidate swap.
+            CONTINUE onto next chain
+    8) evaluateMove(): HPWL Cost evaluation 
+    9) evaluateMoveAcceptance(): hill-climb acceptance with global temperature
+
+    */
+}
+}
+\end{lstlisting}
+
+\newpage
+\end{multicols}
+{
+    \centering
+    \includegraphics[width=0.9\columnwidth]{figures/placement/swapSiteChain.png}
+    \captionof{figure}{Site Chain Swap Proposal}
+    \label{fig:swapChainSite}
+}
+\begin{multicols}{2}
+
+Select a new anchor Site.
+Project a window onto the proposed Site
+If the window anchor breaks a chain, extend the window in its direction until the anchor coincides with the chain anchor..
+If the window tail breaks a chain, extend the window in its direction until the tail coincides with the chain tail.

### 2025-05-31 latex/paper/sections/10-placement.tex
asdf

--- 2025-05-18 latex/paper/sections/10-placement.tex
@@
 \label{sec:placement}
-Up until now, we have only dealt with technology problems where we consider the problem space and adress the constraints of the hardware architecture. 
-For every constraint, we devise a solution or a simplification that sacrifices some optimization to reduce the complexity of the problem space.
+So far we have only dealt with technology problems where we consider the problem space and adress the constraints of the hardware architecture. 
+For every constraint, we devise a solution or a simplification to reduce the complexity of the problem space.
 
-Now that we have dealt with the bulk of the Cell-BEL level constraints and squared them away into self-contained \texttt{SiteInst}s, we can now apply some well-known optimization algorithms to finally address our optimization objective, which is to place these \texttt{SiteInst}s onto the device in a way that minimizes wirelength.
+Now that we have dealt with the bulk of these constraints and squared them away into self-contained \texttt{SiteInst}s, we can now apply some general optimization algorithms to finally address our optimization objective, which is to place these \texttt{SiteInst}s onto the device while minimizing wirelength.
 
-In this paper we implement and present a Simulating Annealing (SA) placer that operates at the \texttt{SiteInst} level.
+In this paper we implement and present a basic Simulating Annealing (SA) placer that operates at the \texttt{SiteInst} level. 
+Shown in Listing \ref{lst:sa_outer} is the simplified pseudocode for the outer-most loop of the algorithm. 
+First, we unplace all of the SiteInsts and then randomly place them across the device.
+Then, we enter the outer-most loop which terminates when the number of passes exceeds a maximum amount, in this case, 300 passes. 
+In each pass, we move() the design, then update the global temperature.
+
+In each move(), we iterate through every \texttt{SiteInst}, propose a move to a new \texttt{Site} placement. 
+If the proposed \texttt{Site} is already occupied by a resident \texttt{SiteInst}, we evaluate a swap. 
+Each move or swap is evaluated by calculating the HPWL cost before and after the movement.
+If the move reduces the HPWL cost, then the movement is accepted outright.
+If the move increases cost, then it can be accepted by chance depending on if the global temperature is high enough to permit the hill-climb.
 
@@
 
-\begin{lstlisting}[language=java, caption={Simulated Annealing pseudocode}, label={lst:sa_pseudocode}]
+\begin{lstlisting}[language=java, caption={SA pseudocode: outer loop}, label={lst:sa_outer}]
+public void placeDesign(PackedDesign packedDesign) {
+    // unplace the pseudorandom packed placement
+    unplaceAllSiteInsts(packedDesign);
+    // place randomly
+    randomInitialPlacement(packedDesign);
+    int passes = 0;
+    while (passes < 300) {
+        move(packedDesign)
+        updateTemperature();
+        passes++;
+    }
+}
 
+private void move(PackedDesign packedDesign) {
+    moveSiteChains(packedDesign.DSPSiteInstCascades);
+    moveSiteChains(packedDesign.CARRYSiteInstChains);
+    moveSingleSite(packedDesign.RAMSiteInsts);
+    moveSingleSite(packedDesign.CLBSiteInsts);
+}
 \end{lstlisting}
+
+\begin{lstlisting}[language=java, caption={Single Site Movement}, label={lst:sa_move_single}]
+protected void moveSingleSite(List<SiteInst> sites) {
+for (SiteInst si : sites) {
+    SiteTypeEnum ste = si.getSiteTypeEnum();
+    List<Site> homeConns = findConnectedSites(si, null);
+    Site homeSite = si.getSite();
+    Site awaySite = proposeSite(si, homeConns, true);
+    SiteInst awaySi = occupiedSites.get(ste).get(awaySite);
+    double oldCost = 0;
+    double newCost = 0;
+    if (awaySi != null) {
+        List<Site> awayConns = findConnectedSites(awaySi, null);
+        oldCost += evaluateSite(homeConns, homeSite);
+        oldCost += evaluateSite(awayConns, awaySite);
+        newCost += evaluateSite(homeConns, awaySite);
+        newCost += evaluateSite(awayConns, homeSite);
+    } else {
+        oldCost += evaluateSite(homeConns, homeSite);
+        newCost += evaluateSite(homeConns, awaySite);
+    }
+    if (evaluateMoveAcceptance(oldCost, newCost)) {
+        if (awaySi != null) {
+            unplaceSiteInst(si);
+            unplaceSiteInst(awaySi);
+            placeSiteInst(si, awaySite);
+            placeSiteInst(awaySi, homeSite);
+        } else {
+            unplaceSiteInst(si);
+            placeSiteInst(si, awaySite);
+        }
+    }
+}
+} // end randomMoveSingleSite()
+\end{lstlisting}
+
+\begin{lstlisting}[language=java, caption={Move Acceptance with Temperature}, label={lst:sa_acceptance}]
+protected boolean evaluateMoveAcceptance(double oldCost, double newCost) {
+    // if the new cost is lower, accept it outright
+    if (newCost < oldCost)
+        return true;
+    // otherwise, evaluate probability to accept higher cost
+    double delta = newCost - oldCost;
+    double acceptanceProbability = 
+        Math.exp(-delta / this.currentTemp);
+    return Math.random() < acceptanceProbability;
+}
+\end{lstlisting}
+
+{
+    \centering
+    \includegraphics[width=\columnwidth]{figures/placement/swapSingleSite.png}
+    \captionof{figure}{Single Site Swap Proposal}
+    \label{fig:swapSingleSite}
+}
+
+\begin{lstlisting}[language=java, caption={Chain Swapping}, label={lst:chain_swap_pseudocode}]
+protected void moveSiteChains(List<List<SiteInst>> chains) {
+for (List<SiteInst> current_chain : chains) {
+    /*
+
+    1) Draw a vertical window of equal size to the current chain. We will refer to this window as the "home window". 
+    2) Select a new candidate anchor for this chain.
+    3) Project the home window onto the candidate anchor with anchors coinciding. We will refer to this projected window as the "away window".
+    4) Check if there are any resident SiteInsts or SiteInst chains present in the away window. Check if the away window breaks any resident SiteInst chains. 
+    5)  IF (the away window breaks any chains): 
+            extend the window in either direction to 
+            fully accommodate all resident chains. 
+        ELSE: GOTO evaluateMove(). 
+    6) Project the extended away window back onto the original region with the tail of the window coinciding with the tail of the current chain. This projected window becomes the new home window. 
+    7) WHILE (the home window breaks any resident chains):
+        shift the window upwards by one Site. 
+        IF (num_shifts > home_win_size - current_chain.size()):
+            reject the candidate swap.
+            CONTINUE onto next chain
+    8) int oldCost = evaluateCost(original placement) 
+    9) int newCost = evaluateCost(proposed placement) 
+    10) evaluateMoveAcceptance(oldCost, newCost, temperature) 
+    11) IF (move accepted):
+            element-wise swap SiteInsts in both windows
+
+    */
+}
+}
+\end{lstlisting}
+
+\newpage
+\end{multicols}
+{
+    \centering
+    \includegraphics[width=0.9\columnwidth]{figures/placement/swapSiteChain.png}
+    \captionof{figure}{Site Chain Swap Proposal}
+    \label{fig:swapChainSite}
+}
+\begin{multicols}{2}
+
+Select a new anchor Site.
+Project a window onto the proposed Site
+If the window anchor breaks a chain, extend the window in its direction until the anchor coincides with the chain anchor..
+If the window tail breaks a chain, extend the window in its direction until the tail coincides with the chain tail.

### 2025-06-02 latex/paper/sections/06-designflow_toolchain.tex
finally start talking about sa

@@
     An engineer describes the intended functionality of the digital system using a hardware description language (HDL) such as Verilog or VHDL. 
-    During this phase, the coding style can vary (behavioral, structural, dataflow, etc.), but it always aims to capture high-level behavior rather than device-specific details.
+    During this phase, the coding style can vary (behavioral, structural, dataflow, etc.), but it generally aims to capture high-level behavior rather than device-specific details.
 
@@
     Given the importance of thorough verification, many established companies dedicate one verification engineer for every design engineer.
-
 \end{enumerate}

### 2025-06-02 latex/paper/sections/10-placement.tex
finally start talking about sa

@@
 \label{sec:placement}
-So far we have only dealt with technology problems where we consider the problem space and adress the constraints of the hardware architecture. 
-For every constraint, we devise a solution or a simplification to reduce the complexity of the problem space.
+So far we have only dealt with the technology where we consider the problem space and adress the constraints of the hardware architecture. 
+For every constraint, we devise a solution to fully address it or a simplification that sacrifices some optimization but reduces the complexity of the problem space.
 
-Now that we have dealt with the bulk of these constraints and squared them away into self-contained \texttt{SiteInst}s, we can now apply some general optimization algorithms to finally address our optimization objective, which is to place these \texttt{SiteInst}s onto the device while minimizing wirelength.
+Now that we have dealt with the bulk of these constraints and squared them away into self-contained \texttt{SiteInst}s, we can now apply some general optimization algorithms to finally address our optimization objective, which is to place these \texttt{SiteInst} objects onto the device while minimizing wirelength.
 
-In this paper we implement and present a basic Simulating Annealing (SA) placer that operates at the \texttt{SiteInst} level. 
-Shown in Listing \ref{lst:sa_outer} is the simplified pseudocode for the outer-most loop of the algorithm. 
+\subsection{Simulated Annealing}
+\label{subsec:simulated_annealing}
+
+In this paper we implement and present a basic Simulating Annealing (SA) placer. 
+As mentioned in the Introduction, SA is a metaheuristic that approximates a global optimum in a large optimization search space. 
+Its name is borrowed from metallurgy and is inspired by the physical process of heating a metal and then slowly lowering the temperature to decrease structural atomic defects, thus minimizing the system energy.
+SA does not guarantee a globally optimal solution, but provides solutions that are often "good enough", especially when an approximate global optimum is more important than finding a precise local optimum in a fixed amount of time.
+SA is typically employed in discrete combinatorial optimization problems like the travelling salesman problem or job-shop scheduling.
+
+The state \texttt{s} of the system and the function \texttt{E(s)} to be minimized is analogous to the internal energy of the system. 
+The goal of SA is to bring the system from some initial state to a state with the minimum possible energy until an energy budget met or until a computational budget is exceeded.
+At each step or iteration, the SA heuristic considers some neighboring state \(s^*\). 
+If \(s^*\) has a lower energy than \(s\), then the state transition from \(s\) to \(s^*\) is accepted outright.
+If \(s^*\) has a higher energy, then it can be probabilistically accepted depending on the current global temperature.
+The global temperature starts at some positive amount then gradually decreases to zero according to a cooling schedule, which can be linear, geometric, logarithmic, or some piecewise combination. 
+The higher the temperature, the higher the likelihood of accepting an energy-increasing move. 
+The selection of cooling schedule parameters is usually fine tuned after empirical experimentation.
+
+%
+%
+%
+%
+
+In the context of our placer, we use SA to minimize wirelength between \texttt{SiteInst}s when placed over the discrete \texttt{Site}s of the FPGA \texttt{device}.
+The state \texttt{s}
+In each iteration of the SA algorithm, a new
+
+%
+%
+%
+%
+
+Shown in Listing \ref{lst:sa_outer} is a simplified pseudocode for the outer-most loop of the SA algorithm. 
 First, we unplace all of the SiteInsts and then randomly place them across the device.
@@
 
-In each move(), we iterate through every \texttt{SiteInst}, propose a move to a new \texttt{Site} placement. 
+In each or move() or "pass", we iterate through every single \texttt{SiteInst} and \texttt{SiteInst} chain, propose a move to a new placement. 
 If the proposed \texttt{Site} is already occupied by a resident \texttt{SiteInst}, we evaluate a swap. 
@@
 If the move increases cost, then it can be accepted by chance depending on if the global temperature is high enough to permit the hill-climb.
-
-\subsection{Simulated Annealing}
-\label{subsec:simulated_annealing}
 
@@
     while (passes < 300) {
+        updateTemperature(passes);
         move(packedDesign)
-        updateTemperature();
         passes++;
@@
 
+\newcolumn
 \begin{lstlisting}[language=java, caption={Single Site Movement}, label={lst:sa_move_single}]
@@
 
-\begin{lstlisting}[language=java, caption={Chain Swapping}, label={lst:chain_swap_pseudocode}]
+% \begin{lstlisting}[language=java, caption={Chain Swapping}, label={lst:chain_swap_pseudocode}]
+% protected void moveSiteChains(List<List<SiteInst>> chains) {
+% for (List<SiteInst> current_chain : chains) {
+%     /*
+% 
+%     1) Draw a vertical window of equal size to the current chain. We will refer to this window as the "home window". 
+%     2) Select a new candidate anchor for this chain.
+%     3) Project the home window onto the candidate anchor with anchors coinciding. We will refer to this projected window as the "away window".
+%     4) Check if there are any resident SiteInsts or SiteInst chains present in the away window. Check if the away window breaks any resident SiteInst chains. 
+%     5)  IF (the away window breaks any chains): 
+%             extend the window in either direction to 
+%             fully accommodate all resident chains. 
+%         ELSE: GOTO evaluateMove(). 
+%     6) Project the extended away window back onto the original region with the tail of the window coinciding with the tail of the current chain. This projected window becomes the new home window. 
+%     7) WHILE (the home window breaks any resident chains):
+%         shift the window upwards by one Site. 
+%         IF (num_shifts > home_win_size - current_chain.size()):
+%             reject the candidate swap.
+%             CONTINUE onto next chain
+%     8) int oldCost = evaluateCost(original placement) 
+%     9) int newCost = evaluateCost(proposed placement) 
+%     10) evaluateMoveAcceptance(oldCost, newCost, temperature) 
+%     11) IF (move accepted):
+%             element-wise swap SiteInsts in both windows
+% 
+%     */
+% }
+% }
+% \end{lstlisting}
+
+\newcolumn
+\begin{lstlisting}[language=Java, caption={Chain Swapping Pseudocode}, label={lst:chain_swap_pseudocode}]
 protected void moveSiteChains(List<List<SiteInst>> chains) {
-for (List<SiteInst> current_chain : chains) {
-    /*
+for (List<SiteInst> currentChain : chains) {
+    int chainSize = currentChain.size();
 
-    1) Draw a vertical window of equal size to the current chain. We will refer to this window as the "home window". 
-    2) Select a new candidate anchor for this chain.
-    3) Project the home window onto the candidate anchor with anchors coinciding. We will refer to this projected window as the "away window".
-    4) Check if there are any resident SiteInsts or SiteInst chains present in the away window. Check if the away window breaks any resident SiteInst chains. 
-    5)  IF (the away window breaks any chains): 
-            extend the window in either direction to 
-            fully accommodate all resident chains. 
-        ELSE: GOTO evaluateMove(). 
-    6) Project the extended away window back onto the original region with the tail of the window coinciding with the tail of the current chain. This projected window becomes the new home window. 
-    7) WHILE (the home window breaks any resident chains):
-        shift the window upwards by one Site. 
-        IF (num_shifts > home_win_size - current_chain.size()):
-            reject the candidate swap.
-            CONTINUE onto next chain
-    8) int oldCost = evaluateCost(original placement) 
-    9) int newCost = evaluateCost(proposed placement) 
-    10) evaluateMoveAcceptance(oldCost, newCost, temperature) 
-    11) IF (move accepted):
-            element-wise swap SiteInsts in both windows
+    // Step 1: Identify home window for this chain
+    Site homeAnchor = currentChain.get(0).getSite();
+    List<Site> homeWindow = getSitesInWindow(homeAnchor, chainSize);
 
-    */
+    // Step 2: Select a candidate away anchor
+    Site awayAnchor = proposeAnchorSite(currentChain, homeWindow, true);
+
+    // Step 3: Determine away window based on awayAnchor and chainSize
+    List<Site> awayWindow = getSitesInWindow(awayAnchor, chainSize);
+
+    // Step 4: Find any resident SiteInst chains in the away window
+    List<List<SiteInst>> residentChainsInAway = 
+        collectChainsInWindow(siteType, awayWindow);
+
+    // Step 5: If any resident chains overlap, extend the away window to include them
+    if (!residentChainsInAway.isEmpty()) {
+        awayWindow = extendWindowToIncludeChains(awayWindow, residentChainsInAway);
+    }
+
+    // Step 6: Map the (possibly extended) away window back onto the original region 
+    //         so that the tail of that window coincides with the tail of the current chain
+    List<Site> candidateHomeWindow = mapAwayToHomeWindow(
+        homeAnchor, awayWindow, chainSize);
+
+    // Step 7: While the candidate home window still overlaps resident chains, shift upward
+    int shifts = 0;
+    while (windowHasOverlap(candidateHomeWindow)) {
+        candidateHomeWindow = shiftWindowUp(candidateHomeWindow);
+        shifts++;
+        if (shifts > (homeWindow.size() - chainSize)) {
+            // Reject this swap attempt and move on to the next chain
+            continue;
+        }
+    }
+
+    // Step 8: Compute cost of the original placement in home window
+    double oldCost = evaluateWindowCost(homeWindow);
+
+    // Step 9: Compute cost of the proposed swap placement
+    double newCost = evaluateWindowCostForSwap(
+        homeWindow, awayWindow, currentChain);
+
+    // Step 10: Decide whether to accept the move based on oldCost, newCost, and temperature
+    if (evaluateMoveAcceptance(oldCost, newCost, currentTemp)) {
+        // Step 11: Perform an element-wise swap of SiteInsts between homeWindow and awayWindow
+        swapChainsBetweenWindows(homeWindow, awayWindow);
+    }
 }

### 2025-06-04 latex/paper/sections/10-placement.tex
work on sa

@@
 
-The state \texttt{s} of the system and the function \texttt{E(s)} to be minimized is analogous to the internal energy of the system. 
+The state \(s\) of the system and the function \(E(s)\) to be minimized is analogous to the internal energy of the system. 
 The goal of SA is to bring the system from some initial state to a state with the minimum possible energy until an energy budget met or until a computational budget is exceeded.
-At each step or iteration, the SA heuristic considers some neighboring state \(s^*\). 
+At each step or iteration, the algorithm considers some neighboring state \(s^*\). 
 If \(s^*\) has a lower energy than \(s\), then the state transition from \(s\) to \(s^*\) is accepted outright.
 If \(s^*\) has a higher energy, then it can be probabilistically accepted depending on the current global temperature.
-The global temperature starts at some positive amount then gradually decreases to zero according to a cooling schedule, which can be linear, geometric, logarithmic, or some piecewise combination. 
+The global temperature starts at some positive amount then gradually decreases to zero according to a cooling schedule
 The higher the temperature, the higher the likelihood of accepting an energy-increasing move. 
-The selection of cooling schedule parameters is usually fine tuned after empirical experimentation.
+This cooling schedule can be linear, geometric, logarithmic, or some piecewise combination. 
+The parameters of the cooling schedule are fine tuned through empirical experimentation.
 
-%
-%
-%
-%
-
-In the context of our placer, we use SA to minimize wirelength between \texttt{SiteInst}s when placed over the discrete \texttt{Site}s of the FPGA \texttt{device}.
-The state \texttt{s}
-In each iteration of the SA algorithm, a new
-
-%
-%
-%
-%
+In the context of our placer, we use SA to minimize wirelength between \texttt{SiteInst}s as we place them over the discrete \texttt{Site}s of the FPGA \texttt{device}.
+The state \texttt{s} represents the current placement of all \texttt{SiteInst}s on the \texttt{device}, while the system energy \(E(s)\) represents the total HPWL of all nets of the placement.
 
@@
 First, we unplace all of the SiteInsts and then randomly place them across the device.
-Then, we enter the outer-most loop which terminates when the number of passes exceeds a maximum amount, in this case, 300 passes. 
+Then, we enter the outer-most loop which terminates when the number of passes exceeds a computational budget, in this case, 300 passes. 
 In each pass, we move() the design, then update the global temperature.
 
-In each or move() or "pass", we iterate through every single \texttt{SiteInst} and \texttt{SiteInst} chain, propose a move to a new placement. 
-If the proposed \texttt{Site} is already occupied by a resident \texttt{SiteInst}, we evaluate a swap. 
-Each move or swap is evaluated by calculating the HPWL cost before and after the movement.
+In each move(), we iterate through four groups of placement objects: \texttt{DSPSiteInstCascades} , \texttt{CARRYSiteInstChains}, \texttt{RAMSiteInst}s, and \texttt{CLBSiteInst}s (single SLICEs that do not belong to CARRY chains).
+For each placement object, we propose a move to a new placement on the device. 
+If the design has for example, 1000 single \texttt{SiteInst}s and 500 multi-\texttt{SiteInst} chains (all mutually exclusive groups), and we set a budget of 300 passes, then the algorithm will consider up to \(300 * (1000 + 500) = 450000\) state transitions.
+The proposed \texttt{Site} or \texttt{Site} chain is conventionally selected randomly, but can also be selected by finding the midpoint or centroid of \texttt{SiteInst}s that share a net with the current object, or even a hybrid of both.
+
+If the proposed \texttt{Site} is already occupied by a resident or existing \texttt{SiteInst}, we evaluate a swap in their placements. 
+Each move or swap is evaluated by calculating the HPWL cost of the placement before and after the movement.
 If the move reduces the HPWL cost, then the movement is accepted outright.
 If the move increases cost, then it can be accepted by chance depending on if the global temperature is high enough to permit the hill-climb.
+
+The temperature decreases with each pass until reaching zero, at which point, the algorithm makes exclusively energy-reducing moves, effectively becoming a greedy algorithm. 
+The hope is that by this point, the placement will crystallize into a global optimum of minimum total HPWL.
 
@@
 
-private void move(PackedDesign packedDesign) {
+private void moveAll(PackedDesign packedDesign) {
     moveSiteChains(packedDesign.DSPSiteInstCascades);
@@
 
-{
-    \centering
-    \includegraphics[width=\columnwidth]{figures/placement/swapSingleSite.png}
-    \captionof{figure}{Single Site Swap Proposal}
-    \label{fig:swapSingleSite}
-}
-
 % \begin{lstlisting}[language=java, caption={Chain Swapping}, label={lst:chain_swap_pseudocode}]
@@
 % \end{lstlisting}
+
+{
+    \centering
+    \includegraphics[width=\columnwidth]{figures/placement/swapSingleSite.png}
+    \captionof{figure}{Single Site Swap Proposal}
+    \label{fig:swapSingleSite}
+}
 

### 2025-06-10 latex/paper/sections/10-placement.tex
more work on sa

@@
 
-In this paper we implement and present a basic Simulating Annealing (SA) placer. 
+In this paper we implement and present a basic Simulated Annealing (SA) placer. 
 As mentioned in the Introduction, SA is a metaheuristic that approximates a global optimum in a large optimization search space. 
-Its name is borrowed from metallurgy and is inspired by the physical process of heating a metal and then slowly lowering the temperature to decrease structural atomic defects, thus minimizing the system energy.
-SA does not guarantee a globally optimal solution, but provides solutions that are often "good enough", especially when an approximate global optimum is more important than finding a precise local optimum in a fixed amount of time.
+SA does not guarantee a globally optimal solution, but provides solutions that are often "good enough", especially when finding an approximate global optimum is more important than finding a precise local optimum in a fixed amount of time.
 SA is typically employed in discrete combinatorial optimization problems like the travelling salesman problem or job-shop scheduling. 
+It is directly inspired by Annealing in metallurgy, which is the physical process of heating a metal above its crystallization temperature, allowing its atoms to migrate through the crystal lattice, and slowly cooling it to allow its atoms to recrystallize into a more desirable structure, such as one that minimizes structural defects
 
 The state \(s\) of the system and the function \(E(s)\) to be minimized is analogous to the internal energy of the system. 
-The goal of SA is to bring the system from some initial state to a state with the minimum possible energy until an energy budget met or until a computational budget is exceeded.
+In metallurgic Annealing, the state \(s\) represents the position of every atom in the metal object at any given snapshot in time, while \(E(s)\) represents the defectiveness in the metal's structure however that may be quantified.
+In the context of our SA placer, the state \(s\) represents the placement of every \texttt{SiteInst} on the \texttt{device} at any given iteration, while \(E(s)\) represents the total wirelength (HPWL) of \texttt{Net}s between all \texttt{SiteInst}s of the current placement.
+The goal of SA is to bring the system from some initial state to a state with lower energy until some stopping criterion is met, whether that be when a computational budget is exceeded, an energy budget is met, or until the energy's rate of change approaches zero.
+
 At each step or iteration, the algorithm considers some neighboring state \(s^*\). 
 If \(s^*\) has a lower energy than \(s\), then the state transition from \(s\) to \(s^*\) is accepted outright.
-If \(s^*\) has a higher energy, then it can be probabilistically accepted depending on the current global temperature.
-The global temperature starts at some positive amount then gradually decreases to zero according to a cooling schedule
+If \(s^*\) has a higher energy, then it can probabilistically be accepted depending on the current global temperature.
 The higher the temperature, the higher the likelihood of accepting an energy-increasing move. 
+The global temperature starts at some positive amount then gradually decreases to zero according to a cooling schedule.
 This cooling schedule can be linear, geometric, logarithmic, or some piecewise combination. 
-The parameters of the cooling schedule are fine tuned through empirical experimentation.
+Its parameters can be fine tuned through empirical experimentation.
 
-In the context of our placer, we use SA to minimize wirelength between \texttt{SiteInst}s as we place them over the discrete \texttt{Site}s of the FPGA \texttt{device}.
-The state \texttt{s} represents the current placement of all \texttt{SiteInst}s on the \texttt{device}, while the system energy \(E(s)\) represents the total HPWL of all nets of the placement.
+Shown in Listing \ref{lst:sa_outer} is a simplified pseudocode for the outer-most loop of our placer's SA algorithm. 
+First, we randomly place all \texttt{SiteInst}s across the \texttt{device}.
+Then, we initialize a geometric cooling schedule with our parameters: an initial temperature, an alpha value, and a maximum number of passes. 
+Then, we enter the outer-most loop which terminates when the number of passes exceeds a computational budget, in this case, 300 passes. 
+In each pass, we move the design, then update the global temperature.
 
-Shown in Listing \ref{lst:sa_outer} is a simplified pseudocode for the outer-most loop of the SA algorithm. 
-First, we unplace all of the SiteInsts and then randomly place them across the device.
-Then, we enter the outer-most loop which terminates when the number of passes exceeds a computational budget, in this case, 300 passes. 
-In each pass, we move() the design, then update the global temperature.
+In each move, we iterate through four mutually exclusive groups of placement objects: \texttt{DSPSiteInstCascades}, \texttt{CARRYSiteInstChains}, \texttt{RAMSiteInst}s, and \texttt{CLBSiteInst}s (single SLICEs that do not belong to CARRY chains).
+For each placement object in the pass, we propose a move to a new placement on the device. 
+If the proposed \texttt{Site} is already occupied by a resident or existing \texttt{SiteInst}, we evaluate a swap in their placements. 
+Each move or swap is evaluated by calculating the HPWL cost of the placement before and after the proposed movement.
+If the design has for example, 50 \texttt{DSPSiteInstCascades}, 100 \texttt{CARRYSiteInstChains}, 75 \texttt{RAMSiteInst}s, and 250 \texttt{CLBSiteInst}s, and we set a budget of 300 passses, then the algorithm will consider up to \(300 * (50 + 100 + 75 + 250) = 142500\) state transitions.
 
-In each move(), we iterate through four groups of placement objects: \texttt{DSPSiteInstCascades} , \texttt{CARRYSiteInstChains}, \texttt{RAMSiteInst}s, and \texttt{CLBSiteInst}s (single SLICEs that do not belong to CARRY chains).
-For each placement object, we propose a move to a new placement on the device. 
-If the design has for example, 1000 single \texttt{SiteInst}s and 500 multi-\texttt{SiteInst} chains (all mutually exclusive groups), and we set a budget of 300 passes, then the algorithm will consider up to \(300 * (1000 + 500) = 450000\) state transitions.
-The proposed \texttt{Site} or \texttt{Site} chain is conventionally selected randomly, but can also be selected by finding the midpoint or centroid of \texttt{SiteInst}s that share a net with the current object, or even a hybrid of both.
+Figure \ref{fig:swapSingleSite} shows how a swap proposal between two \texttt{CLBSiteInst}s or two \texttt{RAMSiteInst}s is evaluated, while Listing \ref{lst:sa_move_single} shows the corresponding Java code.
+The movement of \texttt{SiteInst} chains is considerably more complex because there are more placement constraints to consider, especially when swapping the positions between multiple chains.
+Figure \ref{fig:swapSiteChain} shows an example of how two \texttt{DSPSiteInstCascades} or two \texttt{CARRYSiteInstChains} can be swapped while Listing \ref{lst:chain_swap_pseudocode} shows the corresponding pseudocode.
 
-If the proposed \texttt{Site} is already occupied by a resident or existing \texttt{SiteInst}, we evaluate a swap in their placements. 
-Each move or swap is evaluated by calculating the HPWL cost of the placement before and after the movement.
+The proposed \texttt{Site} or \texttt{Site} chain is usually selected randomly, but can also be selected by finding the midpoint or centroid between \texttt{SiteInst}s that share a net with the current object, or even a hybrid of both.
+
 If the move reduces the HPWL cost, then the movement is accepted outright.
-If the move increases cost, then it can be accepted by chance depending on if the global temperature is high enough to permit the hill-climb.
-
-The temperature decreases with each pass until reaching zero, at which point, the algorithm makes exclusively energy-reducing moves, effectively becoming a greedy algorithm. 
+If the move increases cost, then it can be accepted by chance if the global temperature is high enough to permit the hill-climb.
+The temperature decreases with each pass until reaching zero, at which point, the algorithm makes exclusively energy-decreasing moves, effectively reducing into a greedy algorithm. 
 The hope is that by this point, the placement will crystallize into a global optimum of minimum total HPWL.
@@
     randomInitialPlacement(packedDesign);
+    initCoolingSchedule(10000, 0.95, 300)
     int passes = 0;
@@
 \newcolumn
+{
+    \centering
+    \includegraphics[width=\columnwidth]{figures/placement/swapSingleSite.png}
+    \captionof{figure}{Single Site Swap Proposal}
+    \label{fig:swapSingleSite}
+}
+
 \begin{lstlisting}[language=java, caption={Single Site Movement}, label={lst:sa_move_single}]
@@
 
-\begin{lstlisting}[language=java, caption={Move Acceptance with Temperature}, label={lst:sa_acceptance}]
+\newcolumn
+\begin{lstlisting}[language=java, caption={Cooling Schedule and Move Acceptance with Temperature}, label={lst:sa_acceptance}]
+protected List<Double> coolingSchedule;
+public void initCoolingSchedule(double initialTemp, double alpha, int movesLimit) {
+    // geometric cooling
+    double currentTemp = initialTemp;
+    for (int i = 0; i < movesLimit; i++) {
+        this.coolingSchedule.add(currentTemp);
+        currentTemp *= alpha;
+    }
+}
+
 protected boolean evaluateMoveAcceptance(double oldCost, double newCost) {
@@
 
-% \begin{lstlisting}[language=java, caption={Chain Swapping}, label={lst:chain_swap_pseudocode}]
-% protected void moveSiteChains(List<List<SiteInst>> chains) {
-% for (List<SiteInst> current_chain : chains) {
-%     /*
-% 
-%     1) Draw a vertical window of equal size to the current chain. We will refer to this window as the "home window". 
-%     2) Select a new candidate anchor for this chain.
-%     3) Project the home window onto the candidate anchor with anchors coinciding. We will refer to this projected window as the "away window".
-%     4) Check if there are any resident SiteInsts or SiteInst chains present in the away window. Check if the away window breaks any resident SiteInst chains. 
-%     5)  IF (the away window breaks any chains): 
-%             extend the window in either direction to 
-%             fully accommodate all resident chains. 
-%         ELSE: GOTO evaluateMove(). 
-%     6) Project the extended away window back onto the original region with the tail of the window coinciding with the tail of the current chain. This projected window becomes the new home window. 
-%     7) WHILE (the home window breaks any resident chains):
-%         shift the window upwards by one Site. 
-%         IF (num_shifts > home_win_size - current_chain.size()):
-%             reject the candidate swap.
-%             CONTINUE onto next chain
-%     8) int oldCost = evaluateCost(original placement) 
-%     9) int newCost = evaluateCost(proposed placement) 
-%     10) evaluateMoveAcceptance(oldCost, newCost, temperature) 
-%     11) IF (move accepted):
-%             element-wise swap SiteInsts in both windows
-% 
-%     */
-% }
-% }
-% \end{lstlisting}
-
-{
-    \centering
-    \includegraphics[width=\columnwidth]{figures/placement/swapSingleSite.png}
-    \captionof{figure}{Single Site Swap Proposal}
-    \label{fig:swapSingleSite}
-}
-
-\newcolumn
 \begin{lstlisting}[language=Java, caption={Chain Swapping Pseudocode}, label={lst:chain_swap_pseudocode}]
@@
     \captionof{figure}{Site Chain Swap Proposal}
-    \label{fig:swapChainSite}
+    \label{fig:swapSiteChain}
 }
 \begin{multicols}{2}
-
-Select a new anchor Site.
-Project a window onto the proposed Site
-If the window anchor breaks a chain, extend the window in its direction until the anchor coincides with the chain anchor..
-If the window tail breaks a chain, extend the window in its direction until the tail coincides with the chain tail.

### 2025-06-11 latex/paper/sections/10-placement.tex
start placement results

@@
 \label{sec:placement}
-So far we have only dealt with the technology where we consider the problem space and adress the constraints of the hardware architecture. 
+So far we have only dealt with technology problems where we consider the problem space and adress the constraints of the hardware architecture. 
 For every constraint, we devise a solution to fully address it or a simplification that sacrifices some optimization but reduces the complexity of the problem space.
 
-Now that we have dealt with the bulk of these constraints and squared them away into self-contained \texttt{SiteInst}s, we can now apply some general optimization algorithms to finally address our optimization objective, which is to place these \texttt{SiteInst} objects onto the device while minimizing wirelength.
+Now that we have dealt with the bulk of these constraints and squared them away into self-contained \texttt{SiteInst}s, we can now more easily apply some general optimization algorithms to finally address our optimization objective, which is to place these \texttt{SiteInst} objects onto the device while minimizing wirelength.
 
@@
 SA is typically employed in discrete combinatorial optimization problems like the travelling salesman problem or job-shop scheduling. 
-It is directly inspired by Annealing in metallurgy, which is the physical process of heating a metal above its crystallization temperature, allowing its atoms to migrate through the crystal lattice, and slowly cooling it to allow its atoms to recrystallize into a more desirable structure, such as one that minimizes structural defects
+It is directly inspired by Annealing in metallurgy, which is the physical process of heating a metal above its crystallization temperature, allowing its atoms to migrate through the crystal lattice, and slowly cooling it to allow its atoms to recrystallize into a more desirable structure, such as one that minimizes structural defects.
 
@@
 First, we randomly place all \texttt{SiteInst}s across the \texttt{device}.
-Then, we initialize a geometric cooling schedule with our parameters: an initial temperature, an alpha value, and a maximum number of passes. 
-Then, we enter the outer-most loop which terminates when the number of passes exceeds a computational budget, in this case, 300 passes. 
-In each pass, we move the design, then update the global temperature.
+Then, we initialize a simple geometric cooling schedule: 
+\begin{equation}
+    \label{coolingSchedule}
+    T_{n+1} = \alpha T_n
+\end{equation}
+with initial temperature \(T_0=10000\), cooling rate \(\alpha=0.98\), and \(n = \{0, 1, 2, ..., 300\}\) passes.
 
-In each move, we iterate through four mutually exclusive groups of placement objects: \texttt{DSPSiteInstCascades}, \texttt{CARRYSiteInstChains}, \texttt{RAMSiteInst}s, and \texttt{CLBSiteInst}s (single SLICEs that do not belong to CARRY chains).
+Then, we enter the outer-most loop which terminates when the number of passes exceeds our computational budget, in this case, 300 passes. 
+In each pass, we \texttt{moveAll()} the design, then update the global temperature.
+
+In each \texttt{moveAll()}, we iterate through four mutually exclusive groups of placement objects: \texttt{DSPSiteInstCascades}, \texttt{CARRYSiteInstChains}, \texttt{RAMSiteInst}s, and \texttt{CLBSiteInst}s (single SLICEs that do not belong to CARRY chains).
 For each placement object in the pass, we propose a move to a new placement on the device. 
@@
 
-Figure \ref{fig:swapSingleSite} shows how a swap proposal between two \texttt{CLBSiteInst}s or two \texttt{RAMSiteInst}s is evaluated, while Listing \ref{lst:sa_move_single} shows the corresponding Java code.
-The movement of \texttt{SiteInst} chains is considerably more complex because there are more placement constraints to consider, especially when swapping the positions between multiple chains.
-Figure \ref{fig:swapSiteChain} shows an example of how two \texttt{DSPSiteInstCascades} or two \texttt{CARRYSiteInstChains} can be swapped while Listing \ref{lst:chain_swap_pseudocode} shows the corresponding pseudocode.
+In SA, proposed movements are selected randomly, then accepted or rejected based on legality and cost. 
+Figure \ref{fig:swapSingleSite} shows how a swap proposal between singular \texttt{CLBSiteInst}s or between \texttt{RAMSiteInst}s is evaluated, while Listing \ref{lst:sa_move_single} shows the corresponding Java code.
+On the other hand, the movement of \texttt{SiteInst} chains is considerably more complex because there are more placement constraints to consider, especially when swapping the positions between multiple chains.
+Figure \ref{fig:swapSiteChain} shows a demonstration of how two \texttt{DSPSiteInstCascades} or two \texttt{CARRYSiteInstChains} can be swapped while Listing \ref{lst:chain_swap_pseudocode} shows the corresponding pseudocode.
 
-The proposed \texttt{Site} or \texttt{Site} chain is usually selected randomly, but can also be selected by finding the midpoint or centroid between \texttt{SiteInst}s that share a net with the current object, or even a hybrid of both.
+In more sophisticated placers, the proposed movement can also be selected by finding the centroid between \texttt{SiteInst}s that share a net with the current object, or even a hybrid of random and centroid selection.
+Moves that actively attempt to predict a better placement are often referred to as "directed" moves in literature. 
+Movement approaches that take inspiration from physical interactions like gravity or electrostatics are often referred to as "force directed" moves.
 
@@
 
+\newcolumn
 \begin{lstlisting}[language=java, caption={SA pseudocode: outer loop}, label={lst:sa_outer}]
 public void placeDesign(PackedDesign packedDesign) {
-    // unplace the pseudorandom packed placement
-    unplaceAllSiteInsts(packedDesign);
-    // place randomly
     randomInitialPlacement(packedDesign);
-    initCoolingSchedule(10000, 0.95, 300)
+    // init_temp=10000, alpha=0.98, max_passes=300
+    initCoolingSchedule(10000, 0.98, 300)
     int passes = 0;
@@
 \end{lstlisting}
-
-\newcolumn
-{
-    \centering
-    \includegraphics[width=\columnwidth]{figures/placement/swapSingleSite.png}
-    \captionof{figure}{Single Site Swap Proposal}
-    \label{fig:swapSingleSite}
-}
 
@@
 
-\newcolumn
+{
+    \centering
+    \includegraphics[width=\columnwidth]{figures/placement/swapSingleSite.png}
+    \captionof{figure}{Single Site Swap Proposal}
+    \label{fig:swapSingleSite}
+}
+\vfill
+
 \begin{lstlisting}[language=java, caption={Cooling Schedule and Move Acceptance with Temperature}, label={lst:sa_acceptance}]
@@
 public void initCoolingSchedule(double initialTemp, double alpha, int movesLimit) {
-    // geometric cooling
     double currentTemp = initialTemp;
@@
         this.coolingSchedule.add(currentTemp);
-        currentTemp *= alpha;
+        currentTemp *= alpha; // geometric cooling
     }
@@
 
-    // Step 5: If any resident chains overlap, extend the away window to include them
+    // Step 5: If any resident chains overlap with the away window, extend the away window to fully accomodate them 
     if (!residentChainsInAway.isEmpty()) {
@@
 
-    // Step 6: Map the (possibly extended) away window back onto the original region 
-    //         so that the tail of that window coincides with the tail of the current chain
+    // Step 6: Map the (possibly extended) away window back onto the original region so that the tail of that window coincides with the tail of the current chain
     List<Site> candidateHomeWindow = mapAwayToHomeWindow(
@@
         if (shifts > (homeWindow.size() - chainSize)) {
-            // Reject this swap attempt and move on to the next chain
+            // Reject this swap attempt.
+            // Move on to the next chain.
             continue;

### 2025-06-11 latex/paper/sections/11-placement_results.tex
start placement results

\section{Placement Results}
\label{sec:results}

\end{multicols}

\subsection{Greedy Algorithm (Undirected/Random Moves)}
{
    \centering
    \includegraphics[valign=t, scale=0.25]{figures/results/PlacerGreedyRandom/random_placement.png}
    \includegraphics[valign=t, scale=0.25]{figures/results/PlacerGreedyRandom/00000010.png}
    \includegraphics[valign=t, scale=0.25]{figures/results/PlacerGreedyRandom/00000100.png}
    \includegraphics[valign=t, scale=0.25]{figures/results/PlacerGreedyRandom/00000299.png}
    \captionof{figure}{Snapshots of placement of a 2048th order FIR Filter on a xc7z020 FPGA. \textbf{From left to right:} \textbf{(1)} Initial random placement, \textbf{(2)} After 10 passes, \textbf{(3)} After 100 passes, \textbf{(4)} Final placement after 300 passes.}
    \label{fig:PGRSnapshots}
}
{
    \centering
    \includegraphics[width=0.7\columnwidth]{figures/results/PlacerGreedyRandom/PlacerGreedyRandom_cost_history.png}
    \captionof{figure}{Total HPWL Cost vs number of passes. Final cost: 333013.}
    \label{fig:PGRCurve}
}

\subsection{Greedy Algorithm (Directed/Midpoint Moves)}
{
    \centering
    \includegraphics[valign=t, scale=0.25]{figures/results/PlacerGreedyMidpoint/random_placement.png}
    \includegraphics[valign=t, scale=0.25]{figures/results/PlacerGreedyMidpoint/00000010.png}
    \includegraphics[valign=t, scale=0.25]{figures/results/PlacerGreedyMidpoint/00000100.png}
    \includegraphics[valign=t, scale=0.25]{figures/results/PlacerGreedyMidpoint/00000299.png}
    \captionof{figure}{Snapshots of placement of a 2048th order FIR Filter on a xc7z020 FPGA. \textbf{From left to right:} \textbf{(1)} Initial random placement, \textbf{(2)} After 10 passes, \textbf{(3)} After 100 passes, \textbf{(4)} Final placement after 300 passes.}
    \label{fig:PGMSnapshots}
}
{
    \centering
    \includegraphics[width=0.7\columnwidth]{figures/results/PlacerGreedyMidpoint/PlacerGreedyMidpoint_cost_history.png}
    \captionof{figure}{Total HPWL Cost vs number of passes. Final cost: 590578.}
    \label{fig:PGMCurve}
}

\subsection{Simulated Annealing (Undirected/Random Moves)}
{
    \centering
    \includegraphics[valign=t, scale=0.25]{figures/results/PlacerAnnealRandom/random_placement.png}
    \includegraphics[valign=t, scale=0.25]{figures/results/PlacerAnnealRandom/00000010.png}
    \includegraphics[valign=t, scale=0.25]{figures/results/PlacerAnnealRandom/00000100.png}
    \includegraphics[valign=t, scale=0.25]{figures/results/PlacerAnnealRandom/00000299.png}
    \captionof{figure}{Snapshots of placement of a 2048th order FIR Filter on a xc7z020 FPGA. \textbf{From left to right:} \textbf{(1)} Initial random placement, \textbf{(2)} After 10 passes, \textbf{(3)} After 100 passes, \textbf{(4)} Final placement after 300 passes.}
    \label{fig:PARSnapshots}
}
{
    \centering
    \includegraphics[width=0.7\columnwidth]{figures/results/PlacerAnnealRandom/PlacerAnnealRandom_cost_history.png}
    \captionof{figure}{Total HPWL Cost vs number of passes. Final cost: 346676.}
    \label{fig:PARCurve}
}

\subsection{Simulated Annealing (Directed/Midpoint Moves)}
{
    \centering
    \includegraphics[valign=t, scale=0.25]{figures/results/PlacerAnnealMidpoint/random_placement.png}
    \includegraphics[valign=t, scale=0.25]{figures/results/PlacerAnnealMidpoint/00000010.png}
    \includegraphics[valign=t, scale=0.25]{figures/results/PlacerAnnealMidpoint/00000100.png}
    \includegraphics[valign=t, scale=0.25]{figures/results/PlacerAnnealMidpoint/00000299.png}
    \captionof{figure}{Snapshots of placement of a 2048th order FIR Filter on a xc7z020 FPGA. \textbf{From left to right:} \textbf{(1)} Initial random placement, \textbf{(2)} After 10 passes, \textbf{(3)} After 100 passes, \textbf{(4)} Final placement after 300 passes.}
    \label{fig:PAMSnapshots}
}
{
    \centering
    \includegraphics[width=0.7\columnwidth]{figures/results/PlacerAnnealMidpoint/PlacerAnnealMidpoint_cost_history.png}
    \captionof{figure}{Total HPWL Cost vs number of passes. Final cost: 466855.}
    \label{fig:PAMCurve}
}

\subsection{Simulated Annealing (Hybrid: Random + Midpoint)}
{
    \centering
    \includegraphics[valign=t, scale=0.25]{figures/results/PlacerAnnealHybrid/random_placement.png}
    \includegraphics[valign=t, scale=0.25]{figures/results/PlacerAnnealHybrid/00000010.png}
    \includegraphics[valign=t, scale=0.25]{figures/results/PlacerAnnealHybrid/00000100.png}
    \includegraphics[valign=t, scale=0.25]{figures/results/PlacerAnnealHybrid/00000299.png}
    \captionof{figure}{Snapshots of placement of a 2048th order FIR Filter on a xc7z020 FPGA. \textbf{From left to right:} \textbf{(1)} Initial random placement, \textbf{(2)} After 10 passes, \textbf{(3)} After 100 passes, \textbf{(4)} Final placement after 300 passes.}
    \label{fig:PAHSnapshots}
}
{
    \centering
    \includegraphics[width=0.7\columnwidth]{figures/results/PlacerAnnealHybrid/PlacerAnnealHybrid_cost_history.png}
    \captionof{figure}{Total HPWL Cost vs number of passes. Final cost: 322933.}
    \label{fig:PAHCurve}
}

\begin{multicols}{2}

### 2025-06-11 java/src/main/java/placer/Main.java

@@ -81,10 +81,10 @@ public class Main {
             // works entirely on the SiteInst/Site/Tile level.
             List<PlacerAnnealRandom> SAPlacers = new ArrayList<PlacerAnnealRandom>();
             SAPlacers.add(new PlacerAnnealRandom(rootDir, design, device, region));
+            SAPlacers.add(new PlacerAnnealMidpoint(rootDir, design, device, region));
             SAPlacers.add(new PlacerAnnealHybrid(rootDir, design, device, region));
-            // SAPlacers.add(new PlacerAnnealMidpoint(rootDir, design, device, region));
-            // SAPlacers.add(new PlacerGreedyRandom(rootDir, design, device, region));
-            // SAPlacers.add(new PlacerGreedyMidpoint(rootDir, design, device, region));
+            SAPlacers.add(new PlacerGreedyRandom(rootDir, design, device, region));
+            SAPlacers.add(new PlacerGreedyMidpoint(rootDir, design, device, region));
 
             for (PlacerAnnealRandom placer : SAPlacers) {
                 System.out.println("\n\nStarting " + placer.getPlacerName() + "... \n\n");

### 2025-06-13 python/overlay_convergences.py
placer comparison

@@ -1,0 +1,101 @@
+"""
+Script to combine cost_history.csv files from each placer into a single CSV
+and plot the combined cost histories on a logarithmic y-scale.
+
+Place this script in the `python` subdirectory of your project (e.g.,
+`~/workspace/dev/place-and-route/python`).
+It will look under `outputs/placers/*/printout/cost_history.csv`, merge them on the Iter column,
+save the merged CSV to `combined_cost_history.csv` in the project root,
+and save a log-scale plot to `combined_cost_history.png` in the project root.
+"""
+import sys
+from pathlib import Path
+import pandas as pd
+import matplotlib.pyplot as plt
+
+def main():
+    # Determine directories
+    script_dir = Path(__file__).resolve().parent  # .../project_root/python
+    root_dir = script_dir.parent                  # .../project_root
+    placers_dir = root_dir / "outputs" / "placers"
+
+    if not placers_dir.is_dir():
+        print(f"Error: placers directory not found at {placers_dir}", file=sys.stderr)
+        sys.exit(1)
+
+    # Find all cost_history.csv files under each placer's printout directory
+    csv_files = placers_dir.glob("*/printout/cost_history.csv")
+    dfs = []
+
+    for csv_file in csv_files:
+        placer_name = csv_file.parents[1].name  # two levels up: placers/<PlacerName>/printout
+        try:
+            df = pd.read_csv(csv_file)
+        except Exception as e:
+            print(f"Warning: failed to read {csv_file}: {e}", file=sys.stderr)
+            continue
+
+        # Strip whitespace from column names
+        df.columns = df.columns.str.strip()
+
+        if "Iter" not in df.columns or "Cost" not in df.columns:
+            print(f"Warning: unexpected columns in {csv_file} (found: {list(df.columns)})", file=sys.stderr)
+            continue
+
+        # Keep only Iter and rename Cost column to the placer name
+        df = df[["Iter", "Cost"]].rename(columns={"Cost": placer_name})
+        dfs.append(df)
+
+    if not dfs:
+        print("No valid cost_history.csv files found.", file=sys.stderr)
+        sys.exit(1)
+
+    # Merge all DataFrames on Iter
+    df_merged = dfs[0]
+    for df in dfs[1:]:
+        df_merged = df_merged.merge(df, on="Iter", how="outer")
+
[…]

### 2025-06-13 python/overlay_convergences.py
sort legend by alphabet

@@ -62,40 +62,39 @@ def main():
     df_merged.to_csv(output_csv, index=False)
     print(f"Combined dataset saved to {output_csv}")
 
-    # 1) Linear-scale plot
+    # get placer columns (everything except "Iter") in alphabetical order
+    placer_cols = sorted([c for c in df_merged.columns if c != "Iter"])
+
+    # 1) Linear‐scale plot
     plt.figure(figsize=(12,8))
-    for col in df_merged.columns:
-        if col == "Iter":
-            continue
+    for col in placer_cols:
         plt.plot(df_merged["Iter"], df_merged[col], label=col)
     plt.xlabel('Iteration')
     plt.ylabel('Cost')
     plt.title('Cost History Across Placers (Linear Scale)')
     plt.grid(True, linestyle='--', linewidth=0.5)
-    plt.legend()
+    plt.legend()  # entries are already alphabetical
     plt.tight_layout()
     plot_linear = root_dir / "outputs" / "combined_cost_history_linear.png"
     plt.savefig(plot_linear)
-    print(f"Linear-scale plot saved to {plot_linear}")
+    print(f"Linear‐scale plot saved to {plot_linear}")
 
-    # 2) Log-scale plot
+    # 2) Log‐scale plot
     plt.figure(figsize=(12,8))
-    for col in df_merged.columns:
-        if col == "Iter":
-            continue
+    for col in placer_cols:
         plt.plot(df_merged["Iter"], df_merged[col], label=col)
     plt.yscale('log')
-    # ensure y runs from 1 to the max cost
-    max_cost = df_merged.drop(columns="Iter").max().max()
-    plt.ylim(300000, max_cost)
+    max_cost = df_merged[placer_cols].max().max()
+    plt.ylim(1, max_cost)
     plt.xlabel('Iteration')
     plt.ylabel('Cost (log scale)')
     plt.title('Cost History Across Placers (Log Scale)')
     plt.grid(True, which='both', linestyle='--', linewidth=0.5)
-    plt.legend()
+    plt.legend()  # still alphabetical
     plt.tight_layout()
     plot_log = root_dir / "outputs" / "combined_cost_history_log.png"
     plt.savefig(plot_log)
+    print(f"Log‐scale plot saved to {plot_log}")
 
 if __name__ == "__main__":
     main()

### 2025-06-16 latex/paper/sections/10-placement.tex
reformat results

@@
 In each \texttt{moveAll()}, we iterate through four mutually exclusive groups of placement objects: \texttt{DSPSiteInstCascades}, \texttt{CARRYSiteInstChains}, \texttt{RAMSiteInst}s, and \texttt{CLBSiteInst}s (single SLICEs that do not belong to CARRY chains).
-For each placement object in the pass, we propose a move to a new placement on the device. 
+For each placement object in the pass, we propose a new placement for that object on the device. 
+Each proposed movement represents a proposed state transition of system. 
+If the design has for example, 50 \texttt{DSPSiteInstCascades}, 100 \texttt{CARRYSiteInstChains}, 75 \texttt{RAMSiteInst}s, and 250 \texttt{CLBSiteInst}s, and we set a budget of 300 passses, then the algorithm will consider up to \(300 * (50 + 100 + 75 + 250) = 142500\) state transitions.
 If the proposed \texttt{Site} is already occupied by a resident or existing \texttt{SiteInst}, we evaluate a swap in their placements.
-Each move or swap is evaluated by calculating the HPWL cost of the placement before and after the proposed movement.
-If the design has for example, 50 \texttt{DSPSiteInstCascades}, 100 \texttt{CARRYSiteInstChains}, 75 \texttt{RAMSiteInst}s, and 250 \texttt{CLBSiteInst}s, and we set a budget of 300 passses, then the algorithm will consider up to \(300 * (50 + 100 + 75 + 250) = 142500\) state transitions.
 
@@
 The temperature decreases with each pass until reaching zero, at which point, the algorithm makes exclusively energy-decreasing moves, effectively reducing into a greedy algorithm. 
-The hope is that by this point, the placement will crystallize into a global optimum of minimum total HPWL.
+The hope is that by this point, the placement has already crystallized into an approximate global optimum and the greediness of the algorithm will help the system settle deeper into the energy well faster.
+We can also switch from random selection to centroid selection when the temperature approaches near zero to speed up the convergence further.
 
@@
         updateTemperature(passes);
-        move(packedDesign)
+        moveAll(packedDesign)
         passes++;

### 2025-06-16 latex/paper/sections/11-placement_results.tex
placer comparison
reformat results

@@
 
-\end{multicols}
+Here, we present test results for a basic SA placer as well as test results for four simple variations as listed below. 
+Note that midpoint simply means centroid.
 
-\subsection{Greedy Algorithm (Undirected/Random Moves)}
+\begin{itemize}
+    \item \texttt{PlacerGreedyRandom}: all undirected random moves with greedy acceptance
+    \item \texttt{PlacerGreedyMidpoint}: all directed centroid moves with greedy acceptance
+    \item \texttt{PlacerAnnealRandom}: all undirected random moves with annealing acceptance \textbf{(the basic SA algorithm)}
+    \item \texttt{PlacerAnnealMidpoint}: all directed centroid moves with annealing acceptance
+    \item \texttt{PlacerAnnealHybrid}: 50\% random - 50\% centroid moves with annealing acceptance
+\end{itemize}
+
 {
     \centering
-    \includegraphics[valign=t, scale=0.25]{figures/results/PlacerGreedyRandom/random_placement.png}
-    \includegraphics[valign=t, scale=0.25]{figures/results/PlacerGreedyRandom/00000010.png}
-    \includegraphics[valign=t, scale=0.25]{figures/results/PlacerGreedyRandom/00000100.png}
-    \includegraphics[valign=t, scale=0.25]{figures/results/PlacerGreedyRandom/00000299.png}
-    \captionof{figure}{Snapshots of placement of a 2048th order FIR Filter on a xc7z020 FPGA.
-\textbf{From left to right:} \textbf{(1)} Initial random placement, \textbf{(2)} After 10 passes, \textbf{(3)} After 100 passes, \textbf{(4)} Final placement after 300 passes.}
+    \includegraphics[width=\columnwidth]{figures/results/combined_cost_history_linear.png}
+    \captionof{figure}{All placers overlaid}
+    \label{fig:placers_overlay}
+}
+
+\newcolumn
+{
+    \centering
+    \includegraphics[valign=t, scale=0.13]{figures/results/PlacerGreedyRandom/random_placement.png}
+    \includegraphics[valign=t, scale=0.13]{figures/results/PlacerGreedyRandom/00000010.png}
+    \includegraphics[valign=t, scale=0.13]{figures/results/PlacerGreedyRandom/00000100.png}
+    \includegraphics[valign=t, scale=0.13]{figures/results/PlacerGreedyRandom/00000299.png}
+    \captionof{figure}{PlacerGreedyRandom}
     \label{fig:PGRSnapshots}
 }
+
+\newcolumn
 {
     \centering
-    \includegraphics[width=0.7\columnwidth]{figures/results/PlacerGreedyRandom/PlacerGreedyRandom_cost_history.png}
-    \captionof{figure}{Total HPWL Cost vs number of passes. Final cost: 333013.}
-    \label{fig:PGRCurve}
+    \includegraphics[valign=t, scale=0.13]{figures/results/PlacerGreedyMidpoint/random_placement.png}
+    \includegraphics[valign=t, scale=0.13]{figures/results/PlacerGreedyMidpoint/00000010.png}
+    \includegraphics[valign=t, scale=0.13]{figures/results/PlacerGreedyMidpoint/00000100.png}
+    \includegraphics[valign=t, scale=0.13]{figures/results/PlacerGreedyMidpoint/00000299.png}
+    \captionof{figure}{PlacerGreedyMidpoint}
+    \label{fig:PGMSnapshots}
 }
 
-\subsection{Greedy Algorithm (Directed/Midpoint Moves)}
 {
     \centering
-    \includegraphics[valign=t, scale=0.25]{figures/results/PlacerGreedyMidpoint/random_placement.png}
-    \includegraphics[valign=t, scale=0.25]{figures/results/PlacerGreedyMidpoint/00000010.png}
-    \includegraphics[valign=t, scale=0.25]{figures/results/PlacerGreedyMidpoint/00000100.png}
-    \includegraphics[valign=t, scale=0.25]{figures/results/PlacerGreedyMidpoint/00000299.png}
-    \captionof{figure}{Snapshots of placement of a 2048th order FIR Filter on a xc7z020 FPGA.
-\textbf{From left to right:} \textbf{(1)} Initial random placement, \textbf{(2)} After 10 passes, \textbf{(3)} After 100 passes, \textbf{(4)} Final placement after 300 passes.}
-    \label{fig:PGMSnapshots}
+    \includegraphics[valign=t, scale=0.13]{figures/results/PlacerAnnealRandom/random_placement.png}
+    \includegraphics[valign=t, scale=0.13]{figures/results/PlacerAnnealRandom/00000010.png}
+    \includegraphics[valign=t, scale=0.13]{figures/results/PlacerAnnealRandom/00000100.png}
+    \includegraphics[valign=t, scale=0.13]{figures/results/PlacerAnnealRandom/00000299.png}
+    \captionof{figure}{PlacerAnnealRandom}
+    \label{fig:PARSnapshots}
 }
+
+\newcolumn
 {
     \centering
-    \includegraphics[width=0.7\columnwidth]{figures/results/PlacerGreedyMidpoint/PlacerGreedyMidpoint_cost_history.png}
-    \captionof{figure}{Total HPWL Cost vs number of passes. Final cost: 590578.}
-    \label{fig:PGMCurve}
+    \includegraphics[valign=t, scale=0.13]{figures/results/PlacerAnnealMidpoint/random_placement.png}
+    \includegraphics[valign=t, scale=0.13]{figures/results/PlacerAnnealMidpoint/00000010.png}
+    \includegraphics[valign=t, scale=0.13]{figures/results/PlacerAnnealMidpoint/00000100.png}
+    \includegraphics[valign=t, scale=0.13]{figures/results/PlacerAnnealMidpoint/00000299.png}
+    \captionof{figure}{PlacerAnnealMidpoint}
+    \label{fig:PAMSnapshots}
 }
 
-\subsection{Simulated Annealing (Undirected/Random Moves)}
 {
     \centering
-    \includegraphics[valign=t, scale=0.25]{figures/results/PlacerAnnealRandom/random_placement.png}
-    \includegraphics[valign=t, scale=0.25]{figures/results/PlacerAnnealRandom/00000010.png}
-    \includegraphics[valign=t, scale=0.25]{figures/results/PlacerAnnealRandom/00000100.png}
-    \includegraphics[valign=t, scale=0.25]{figures/results/PlacerAnnealRandom/00000299.png}
-    \captionof{figure}{Snapshots of placement of a 2048th order FIR Filter on a xc7z020 FPGA.
-\textbf{From left to right:} \textbf{(1)} Initial random placement, \textbf{(2)} After 10 passes, \textbf{(3)} After 100 passes, \textbf{(4)} Final placement after 300 passes.}
-    \label{fig:PARSnapshots}
-}
-{
-    \centering
-    \includegraphics[width=0.7\columnwidth]{figures/results/PlacerAnnealRandom/PlacerAnnealRandom_cost_history.png}
-    \captionof{figure}{Total HPWL Cost vs number of passes. Final cost: 346676.}
-    \label{fig:PARCurve}
-}
-
-\subsection{Simulated Annealing (Directed/Midpoint Moves)}
-{
-    \centering
-    \includegraphics[valign=t, scale=0.25]{figures/results/PlacerAnnealMidpoint/random_placement.png}
-    \includegraphics[valign=t, scale=0.25]{figures/results/PlacerAnnealMidpoint/00000010.png}
-    \includegraphics[valign=t, scale=0.25]{figures/results/PlacerAnnealMidpoint/00000100.png}
-    \includegraphics[valign=t, scale=0.25]{figures/results/PlacerAnnealMidpoint/00000299.png}
-    \captionof{figure}{Snapshots of placement of a 2048th order FIR Filter on a xc7z020 FPGA.
-\textbf{From left to right:} \textbf{(1)} Initial random placement, \textbf{(2)} After 10 passes, \textbf{(3)} After 100 passes, \textbf{(4)} Final placement after 300 passes.}
-    \label{fig:PAMSnapshots}
-}
-{
-    \centering
-    \includegraphics[width=0.7\columnwidth]{figures/results/PlacerAnnealMidpoint/PlacerAnnealMidpoint_cost_history.png}
-    \captionof{figure}{Total HPWL Cost vs number of passes. Final cost: 466855.}
-    \label{fig:PAMCurve}
-}
-
-\subsection{Simulated Annealing (Hybrid: Random + Midpoint)}
-{
-    \centering
-    \includegraphics[valign=t, scale=0.25]{figures/results/PlacerAnnealHybrid/random_placement.png}
-    \includegraphics[valign=t, scale=0.25]{figures/results/PlacerAnnealHybrid/00000010.png}
-    \includegraphics[valign=t, scale=0.25]{figures/results/PlacerAnnealHybrid/00000100.png}
-    \includegraphics[valign=t, scale=0.25]{figures/results/PlacerAnnealHybrid/00000299.png}
-    \captionof{figure}{Snapshots of placement of a 2048th order FIR Filter on a xc7z020 FPGA.
-\textbf{From left to right:} \textbf{(1)} Initial random placement, \textbf{(2)} After 10 passes, \textbf{(3)} After 100 passes, \textbf{(4)} Final placement after 300 passes.}
+    \includegraphics[valign=t, scale=0.13]{figures/results/PlacerAnnealHybrid/random_placement.png}
+    \includegraphics[valign=t, scale=0.13]{figures/results/PlacerAnnealHybrid/00000010.png}
+    \includegraphics[valign=t, scale=0.13]{figures/results/PlacerAnnealHybrid/00000100.png}
+    \includegraphics[valign=t, scale=0.13]{figures/results/PlacerAnnealHybrid/00000299.png}
+    \captionof{figure}{PlacerAnnealHybrid}
     \label{fig:PAHSnapshots}
 }
-{
-    \centering
-    \includegraphics[width=0.7\columnwidth]{figures/results/PlacerAnnealHybrid/PlacerAnnealHybrid_cost_history.png}
-    \captionof{figure}{Total HPWL Cost vs number of passes. Final cost: 322933.}
-    \label{fig:PAHCurve}
-}
+% {
+%     \centering
+%     \includegraphics[width=0.7\columnwidth]{figures/results/PlacerAnnealHybrid/PlacerAnnealHybrid_cost_history.png}
+%     \captionof{figure}{Total HPWL Cost vs number of passes. Final cost: 322933.}
+%     \label{fig:PAHCurve}
+% }
 

### 2025-06-16 python/overlay_convergences.py

@@ -66,10 +66,10 @@ def main():
     placer_cols = sorted([c for c in df_merged.columns if c != "Iter"])
 
     # 1) Linear‐scale plot
-    plt.figure(figsize=(12,8))
+    plt.figure(figsize=(6,4))
     for col in placer_cols:
         plt.plot(df_merged["Iter"], df_merged[col], label=col)
-    plt.xlabel('Iteration')
+    plt.xlabel('Passes')
     plt.ylabel('Cost')
     plt.title('Cost History Across Placers (Linear Scale)')
     plt.grid(True, linestyle='--', linewidth=0.5)
@@ -80,13 +80,13 @@ def main():
     print(f"Linear‐scale plot saved to {plot_linear}")
 
     # 2) Log‐scale plot
-    plt.figure(figsize=(12,8))
+    plt.figure(figsize=(6,4))
     for col in placer_cols:
         plt.plot(df_merged["Iter"], df_merged[col], label=col)
     plt.yscale('log')
     max_cost = df_merged[placer_cols].max().max()
     plt.ylim(1, max_cost)
-    plt.xlabel('Iteration')
+    plt.xlabel('Passes')
     plt.ylabel('Cost (log scale)')
     plt.title('Cost History Across Placers (Log Scale)')
     plt.grid(True, which='both', linestyle='--', linewidth=0.5)

### 2025-06-16 latex/paper/sections/11-placement_results.tex
utilization

@@
 
+We will need an HDL design that utilizes a good mixture of LUTs, FFs, BRAMs, and DSPs at scale to demonstrate the robustness and performance of our placers. 
+Any DSP subsystem can serve as a good candidate for such a demonstration. 
+Here, we will perform placement on a 2048th-order FIR Filter which was conveniently created as coursework for a VLSI course. 
+After passing synthesis, the FIR Filter design calls for the following primitive cells:
+\begin{itemize}
+    \item 1693 \texttt{LUT} (1-6)
+    \item 920 \texttt{FDRE} + \texttt{FDSE}
+    \item 282 \texttt{CARRY4}
+    \item 32 \texttt{RAMB48E1}
+    \item 64 \texttt{DSP48E1}
+\end{itemize}
+
+We will perform placement of this design on a \texttt{xc7z020} FPGA, which is a mid-ranged Xilinx device containing the following relevant BELs:
+\begin{itemize}
+    \item 53200 \texttt{LUT} (counting \texttt{LUT5}-\texttt{LUT6} pairs as one \texttt{LUT})
+    \item 106400 \texttt{FF}
+    \item 13300 \texttt{CARRY4}
+    \item 171 \texttt{RAMB18E1}
+    \item 220 \texttt{DSP48E1}
+\end{itemize}
+
+Our expected overall utilization for the FIR Filter on this device will be:
+\begin{itemize}
+    \item LUTs: 3\%
+    \item Registers: 1\%
+    \item Carry: 2\%
+    \item BRAMs: 23\%
+    \item DSPs: 29\%
+\end{itemize}
+
+Shown in figure \ref{fig:placers_overlay} are the HPWL curves over number of passes for each of the five placer variants. 
+The Midpoint-exclusive placers \texttt{PlacerGreedyMidpoint} and \texttt{PlacerAnnealMidpoint} appear
+
 {
@@
 
-\newcolumn
 {
@@
 
-\newcolumn
+\columnbreak
 {
@@
 }
-
+\vfill
 {
@@
 }
-% {
-%     \centering
-%     \includegraphics[width=0.7\columnwidth]{figures/results/PlacerAnnealHybrid/PlacerAnnealHybrid_cost_history.png}
-%     \captionof{figure}{Total HPWL Cost vs number of passes. Final cost: 322933.}
-%     \label{fig:PAHCurve}
-% }
-
-\begin{multicols}{2}

### 2025-06-18 latex/paper/sections/05-7series_architecture.tex
packing edits
start placement results
results discussion

@@
 A single CLB Tile can therefore host either two \texttt{SLICEL}s or one \texttt{SLICEL} and one \texttt{SLICEM}.
-To simplify the problem space, however, we will only consider \texttt{SLICEL}s for general logic and use the dedicated \texttt{RAMB18E1} cells to implement RAM elements.
+However, to simplify the problem space, we will only consider \texttt{SLICEL}s for general logic and use the dedicated \texttt{RAMB18E1} cells to implement RAM elements.
 
@@
 The inputs are then essentially treated as an address space that maps to a data value space in an asynchronous ROM. 
-No explicit logic gates like NAND or XNOR are synthesized, contrary to what newcomers might expect from a "Field Programmable Gate Array".
+No explicit logic gates like NAND or XNOR are synthesized, contrary to what the name \emph{Field Programmable Gate Array} might suggest.
+
+In the 7‑Series devices, one LUT can facilitate any 6-input boolean function, or two 5-input functions, as long as they share the same input signals.  
+The LUT can also host two independent boolean functions of up to 3 inputs each, even when the inputs are not shared.  
+Functions requiring more than six unique inputs are decomposed across multiple cascaded LUTs.
+Figure \ref{fig:lut_synthesis} shows an example of where a LUT is typically synthesized in a design entry.
 
@@
 \vspace{1.0cm}
-
-In the 7‑Series devices, one LUT can facilitate any 6-input boolean function, or two 5-input functions, as long as they share the same input signals.  
-The LUT can also host two independent boolean functions of up to 3 inputs each, even when the inputs are not shared.  
-Functions requiring more than six unique inputs are decomposed across multiple cascaded LUTs.
-Figure \ref{fig:lut_synthesis} shows an example of where a LUT is typically synthesized in a design entry.
 

### 2025-06-18 latex/paper/sections/10-placement.tex
results discussion

@@
 with initial temperature \(T_0=10000\), cooling rate \(\alpha=0.98\), and \(n = \{0, 1, 2, ..., 300\}\) passes.
+Hill-climbing moves can be accepted via the probability function \(P()\): 
+\begin{equation}
+    \label{moveAcceptance}
+    P(s, s^*, T_n) = exp(\frac{E(s) - E(s^*)}{T_n})
+\end{equation}
+where \(E(s^*)\) is the energy of the proposed state, \(E(s)\) is the energy of the current state, and \(T_n\) is the current global temperature at \(n\) passes. 
+If \(P(s, s^*, T_n) < random(0, 1)\), then \(s \leftarrow s^*\).
+Recall that this probability is only evaluated for hill-climbing moves, thus cost difference \(E(s) - E(s^*)\) is always negative.
+In summary, the higher the cost difference and temperature, the higher the probabilty of hill-climb acceptance.
+On the other hand, all hill-descending moves are simply accepted outright.
 
@@
 Moves that actively attempt to predict a better placement are often referred to as "directed" moves in literature. 
-Movement approaches that take inspiration from physical interactions like gravity or electrostatics are often referred to as "force directed" moves.
+Placers that take inspiration from physical interactions like gravity or electrostatics are often referred to as "force directed" moves and are akin to physics simulators modelling interactions between atoms and molecules.
 

### 2025-06-18 latex/paper/sections/11-placement_results.tex
results discussion

@@
 \label{sec:results}
-
-Here, we present test results for a basic SA placer as well as test results for four simple variations as listed below. 
-Note that midpoint simply means centroid.
-
-\begin{itemize}
-    \item \texttt{PlacerGreedyRandom}: all undirected random moves with greedy acceptance
-    \item \texttt{PlacerGreedyMidpoint}: all directed centroid moves with greedy acceptance
-    \item \texttt{PlacerAnnealRandom}: all undirected random moves with annealing acceptance \textbf{(the basic SA algorithm)}
-    \item \texttt{PlacerAnnealMidpoint}: all directed centroid moves with annealing acceptance
-    \item \texttt{PlacerAnnealHybrid}: 50\% random - 50\% centroid moves with annealing acceptance
-\end{itemize}
 
@@
 
-Shown in figure \ref{fig:placers_overlay} are the HPWL curves over number of passes for each of the five placer variants. 
-The Midpoint-exclusive placers \texttt{PlacerGreedyMidpoint} and \texttt{PlacerAnnealMidpoint} appear
+Now, we present test results for a basic SA placer as well as test results for four simple variations as listed below. 
+Note that midpoint simply means centroid.
+
+\begin{itemize}
+    \item \texttt{PlacerGreedyRandom}: all undirected random moves with greedy acceptance (always zero temperature)
+    \item \texttt{PlacerGreedyMidpoint}: all directed centroid moves with greedy acceptance (always zero temperature)
+    \item \texttt{PlacerAnnealRandom}: all undirected random moves with annealing acceptance \textbf{(the basic SA algorithm)}
+    \item \texttt{PlacerAnnealMidpoint}: all directed centroid moves with annealing acceptance
+    \item \texttt{PlacerAnnealHybrid}: \textbf{Initial:} 50\% random - 50\% centroid moves with annealing acceptance. \textbf{At Near-Zero Temperature:} 100\% centroid moves with annealing acceptance.
+\end{itemize}
+
+((Talk about the initial cost.))
+
+Shown in figure \ref{fig:placers_overlay} are the HPWL curves over number of passes for each of the five variants. 
+\texttt{PlacerGreedyMidpoint} (in red) appears to perform the worst by a considerable margin with a final HPWL cost of 590578.
+This is likely due to the lack of randomness in moves resulting in a lack of diversity in movement proposals combined with the greediness of the movement acceptance that prevents the system from escaping local minima.
+This is supported by its HPWL curve (in red) showing a sharp drop in cost that quickly asymptotes just below the 600000 mark, suggesting that the system crystallized very quickly into a local minimum and was unable to climb out of it.
+
+Next is \texttt{PlacerAnnealMidpoint} with a final cost of 466855, which like the previous placer, lacks the diversity of random moves and contributes to faster crystallization into local minima.
+However, this placer likely performed better than the previous as there is still some randomness in the move acceptance evaluation, which gives way to occasional hill-climbing. 
+Keep in mind that centroid moves do not necessarily decrease system cost, as the centroid move will decrease the HPWL cost of the nets on connected to the current \texttt{SiteInst}, but potentially increase the cost of another group of nets should a swap be evaluated, which can potentially result in a total increase in system cost.
 

### 2025-06-20 latex/paper/sections/10-placement.tex
legalization

--- 2025-06-16 latex/paper/sections/10-placement.tex
@@
 In metallurgic Annealing, the state \(s\) represents the position of every atom in the metal object at any given snapshot in time, while \(E(s)\) represents the defectiveness in the metal's structure however that may be quantified.
-In the context of our SA placer, the state \(s\) represents the placement of every \texttt{SiteInst} on the \texttt{device} at any given iteration, while \(E(s)\) represents the total wirelength (HPWL) of \texttt{Net}s between all \texttt{SiteInst}s of the current placement.
+In the context of our SA placer, the state \(s\) represents the placement of every \texttt{SiteInst} on the \texttt{device} at any given iteration, while \(E(s)\) represents the total wirelength of \texttt{Net}s between all \texttt{SiteInst}s of the current placement.
+We will be measuring wirelength using the manhattan distance between the 2-dimensional coordinates of the \texttt{SiteInst}s on the \texttt{device}.
+We will refer to these distances as half perimeter wirelength (HPWL).
 The goal of SA is to bring the system from some initial state to a state with lower energy until some stopping criterion is met, whether that be when a computational budget is exceeded, an energy budget is met, or until the energy's rate of change approaches zero.
@@
 with initial temperature \(T_0=10000\), cooling rate \(\alpha=0.98\), and \(n = \{0, 1, 2, ..., 300\}\) passes.
+Hill-climbing moves can be accepted via the probability function \(P()\): 
+\begin{equation}
+    \label{moveAcceptance}
+    P(s, s^*, T_n) = exp(\frac{E(s) - E(s^*)}{T_n})
+\end{equation}
+where \(E(s^*)\) is the energy of the proposed state \(s^*\), \(E(s)\) is the energy of the current state \(s\), and \(T_n\) is the current global temperature at \(n\) passes. 
+If \(P(s, s^*, T_n) < random(0, 1)\), then \(s \leftarrow s^*\).
+Recall that this probability is only evaluated for hill-climbing moves, thus the cost difference \(E(s) - E(s^*)\) is always negative.
+In summary, a lower the cost difference and a higher the temperature contributes to a higher the probabilty of hill-climbing acceptance.
+All hill-descending moves on the other hand are simply accepted outright.
 
@@
 Moves that actively attempt to predict a better placement are often referred to as "directed" moves in literature. 
-Movement approaches that take inspiration from physical interactions like gravity or electrostatics are often referred to as "force directed" moves.
+Placers that take inspiration from physical interactions like gravity or electrostatics are often referred to as "force directed" moves and are akin to physics simulators.
+
+Directed moves as opposed to undirected moves give rise to the concept of legalization. 
+When using undirected moves as in SA, we can simply keep track of a \texttt{List} or \texttt{HashMap} of occupied \texttt{Site}s and available \texttt{Site}s and select from that data structure via random selection.
+However, for directed moves like centroid selection, the centroid between \texttt{Site}s may not necessarily fall directly on another discrete \texttt{Site} on the \texttt{device}. 
+Even when it does, the centroid \texttt{Site} may be incompatible with the \texttt{SiteInst} or may fail to meet some other placement constraint. 
+In such cases, we must do a neighborhood search around the centroid up to a certain radius to find a \texttt{Site} that satisfies all the constraints of the \texttt{SiteInst} at hand. 
+We will take some inspiration from ray-marching in graphics and do a diamond-shaped spiral search around the centroid where we check the legality of each \texttt{Site} as we step through the spiral and select the first \texttt{Site} that meets all of the placement constraints.
+
+In force directed moves, there is a concept of global placement versus detailed placement, where global placement is driven by forces and detailed placement attempts to legalize the global placement by discretizing it onto the grid of \texttt{Site}s on the device.
+There also exists other strategies to mix packing, placement, and legalization in different ways. (CITE THAT ONE PAPER).
+In this context, our placer follows a straightforward pack-legalize-place strategy.
 

### 2025-06-20 latex/paper/sections/11-placement_results.tex
legalization

@@
 
-Shown in figure \ref{fig:placers_overlay} are the HPWL curves over number of passes for each of the five variants. 
-\texttt{PlacerGreedyMidpoint} (in red) appears to perform the worst by a considerable margin with a final HPWL cost of 590578.
+Shown in figure \ref{fig:placers_overlay} are the total HPWL curves over number of passes for each of the five placers. 
+\texttt{PlacerGreedyMidpoint} (in red) appears to perform the worst by a considerable margin with a final HPWL cost of 590K.
 This is likely due to the lack of randomness in moves resulting in a lack of diversity in movement proposals combined with the greediness of the movement acceptance that prevents the system from escaping local minima.
-This is supported by its HPWL curve (in red) showing a sharp drop in cost that quickly asymptotes just below the 600000 mark, suggesting that the system crystallized very quickly into a local minimum and was unable to climb out of it.
+This is supported by its HPWL curve (in red) showing a sharp drop in cost that quickly asymptotes just below the 600K mark, suggesting that the system crystallized very quickly into a local minimum and was unable to climb out of it.
 
-Next is \texttt{PlacerAnnealMidpoint} with a final cost of 466855, which like the previous placer, lacks the diversity of random moves and contributes to faster crystallization into local minima.
-However, this placer likely performed better than the previous as there is still some randomness in the move acceptance evaluation, which gives way to occasional hill-climbing. 
-Keep in mind that centroid moves do not necessarily decrease system cost, as the centroid move will decrease the HPWL cost of the nets on connected to the current \texttt{SiteInst}, but potentially increase the cost of another group of nets should a swap be evaluated, which can potentially result in a total increase in system cost.
+Next is \texttt{PlacerAnnealMidpoint} with a final cost of 466K.
+Like the previous placer, \texttt{PlacerAnnealMidpoint} lacks the diversity of random moves which contributes to faster crystallization into local minima.
+However, this placer likely performed better than the previous as there is still some randomness in the move acceptance evaluation which gives way to occasional hill-climbing. 
+Keep in mind that centroid moves do not necessarily decrease system cost, as the centroid move will decrease the HPWL cost of the nets on connected to the current \texttt{SiteInst}, but can potentially increase the cost of another group of nets should a swap be evaluated, which can potentially result in a total increase in system cost.
+
+The next three placers performed significantly better than the previous two placers.
+\texttt{PlacerAnnealRandom} and \texttt{PlacerAnnealHybrid} performed quite similarly with final costs of 346K and 322K respectively.
 

### 2025-06-26 latex/paper/sections/10-placement.tex
minor edits

@@
 Recall that this probability is only evaluated for hill-climbing moves, thus the cost difference \(E(s) - E(s^*)\) is always negative.
-In summary, a lower the cost difference and a higher the temperature contributes to a higher the probabilty of hill-climbing acceptance.
+In summary, a lower cost difference and a higher temperature contributes to a higher probabilty of hill-climbing acceptance.
 All hill-descending moves on the other hand are simply accepted outright.
@@
 
-In force directed moves, there is a concept of global placement versus detailed placement, where global placement is driven by forces and detailed placement attempts to legalize the global placement by discretizing it onto the grid of \texttt{Site}s on the device.
+In force directed moves, there is a concept of global placement and detailed placement, where global placement is driven by forces and detailed placement attempts to legalize the global placement by discretizing it onto the grid of \texttt{Site}s on the device.
+In such placers, objects are first placed freely on an analog field in global placement, even allowing objects to overlap with one another. 
+Then, in detailed placement, the object positions are discretized amongst the grid of resources and overlapping objects spread out until there are no overlaps.
+In literature, detailed placement is also referred to as legalization.
+
 There also exists other strategies to mix packing, placement, and legalization in different ways. (CITE THAT ONE PAPER).

### 2025-06-26 latex/paper/sections/11-placement_results.tex
minor edits

@@
 \texttt{PlacerGreedyMidpoint} (in red) appears to perform the worst by a considerable margin with a final HPWL cost of 590K.
-This is likely due to the lack of randomness in moves resulting in a lack of diversity in movement proposals combined with the greediness of the movement acceptance that prevents the system from escaping local minima.
-This is supported by its HPWL curve (in red) showing a sharp drop in cost that quickly asymptotes just below the 600K mark, suggesting that the system crystallized very quickly into a local minimum and was unable to climb out of it.
+This is likely due to the lack of randomness or lack of diversity in move proposals combined with the persistent greediness of the movement acceptance that prevents the system from escaping local minima.
+This conclusion is supported by its HPWL curve (in red) showing a sharp drop in cost that quickly asymptotes just below the 600K mark, suggesting that the system crystallized very quickly into a local minimum and was unable to climb out of it.
 
@@
 However, this placer likely performed better than the previous as there is still some randomness in the move acceptance evaluation which gives way to occasional hill-climbing. 
-Keep in mind that centroid moves do not necessarily decrease system cost, as the centroid move will decrease the HPWL cost of the nets on connected to the current \texttt{SiteInst}, but can potentially increase the cost of another group of nets should a swap be evaluated, which can potentially result in a total increase in system cost.
+Keep in mind that centroid moves do not necessarily decrease system cost. 
+The centroid move can decrease the HPWL cost of the nets on connected to the current \texttt{SiteInst}, but can potentially increase the cost of another group of nets in the event of a swap, which can potentially result in an increase in total system cost.
 
@@
 \texttt{PlacerAnnealRandom} and \texttt{PlacerAnnealHybrid} performed quite similarly with final costs of 346K and 322K respectively.
+
+In between \texttt{AnnealRandom} and \texttt{AnnealHybrid} comes \texttt{PlacerGreedyRandom} with a final cost of 333K.
 

### 2025-06-27 latex/paper/sections/10-placement.tex
legalization + discussion

--- 2025-06-20 latex/paper/sections/10-placement.tex
@@
 Moves that actively attempt to predict a better placement are often referred to as "directed" moves in literature. 
-Placers that take inspiration from physical interactions like gravity or electrostatics are often referred to as "force directed" moves and are akin to physics simulators.
+Placers that take inspiration from physical interactions like gravity or electrostatics are often referred to as "force directed" placers and are akin to physics simulators.
 
@@
 
-In force directed moves, there is a concept of global placement versus detailed placement, where global placement is driven by forces and detailed placement attempts to legalize the global placement by discretizing it onto the grid of \texttt{Site}s on the device.
+In force directed placers, there is a concept of global placement followed by detailed placement
+Global placement places objects freely on an analog plane and is driven by directed movement or analytical solution.
+In global placement, objects do not have to adhere to any hardware constraints and can even overlap with one another.
+Detailed placement takes the global placement and attempts to fit it within the constraints of the hardware. 
+Detailed placement "legalize" the global placement by discretizing the analog positions of the objects onto the discrete grid of resources of the \texttt{device}.
+Overlapping objects are spread out until every \texttt{SiteInst}-\texttt{Cell} is assigned its own unique \texttt{Site}-\texttt{BEL}.
+
 There also exists other strategies to mix packing, placement, and legalization in different ways. (CITE THAT ONE PAPER).
@@
 The hope is that by this point, the placement has already crystallized into an approximate global optimum and the greediness of the algorithm will help the system settle deeper into the energy well faster.
-We can also switch from random selection to centroid selection when the temperature approaches near zero to speed up the convergence further.
+We can also switch from random selection to centroid selection when the temperature approaches near zero to speed up the settlement further.
 

### 2025-06-27 latex/paper/sections/11-placement_results.tex
legalization + discussion

@@
 After passing synthesis, the FIR Filter design calls for the following primitive cells:
+
 \begin{itemize}
-    \item 1693 \texttt{LUT} (1-6)
-    \item 920 \texttt{FDRE} + \texttt{FDSE}
+    \item 1693 \texttt{LUT}
+    \item 920 \texttt{FF}
     \item 282 \texttt{CARRY4}
@@
 \begin{itemize}
-    \item LUTs: 3\%
-    \item Registers: 1\%
+    \item Lookup Tables: 3\%
+    \item Flip Flops: 1\%
     \item Carry: 2\%
@@
 The next three placers performed significantly better than the previous two placers.
-\texttt{PlacerAnnealRandom} and \texttt{PlacerAnnealHybrid} performed quite similarly with final costs of 346K and 322K respectively.
+Our vanilla SA placer, \texttt{PlacerAnnealRandom} is represented by the green curve.
+We can observe that the cost decreases more steadily than the previous two placers, with noticeable spikes in total system cost due to hill-climbing, until settling at the 346K mark as the global temperature cools to zero.
 
-In between \texttt{AnnealRandom} and \texttt{AnnealHybrid} comes \texttt{PlacerGreedyRandom} with a final cost of 333K.
+Next is \texttt{PlacerGreedyRandom} which is represented by the purple curve.
+We can observe a much steeper initial drop in system cost than the vanilla SA placer due to greedy movement acceptance.
+Interestingly, this placer also settled at a lower final system cost of 333K, which is a slight 4\% improvement over vanilla SA.
+We originally expected the greediness to be a detriment to finding the global minimum, but due to the inherent randomness of the algorithm, these expectations are never certain 
+Broader trends will require many more test trials to observe reliably.
+
+The best performing placer was \texttt{PlacerAnnealHybrid}, represented by the blue curve.
+We can observe that \texttt{AnnealHybrid} settles at a similar rate with the vanilla \texttt{AnnealRandom}, but eventually settles at a lower minimum at 322K, outperforming even \texttt{GreedyRandom}.
 

### 2025-07-02 latex/paper/sections/10-placement.tex
lots of reformatting

@@
 First, we randomly place all \texttt{SiteInst}s across the \texttt{device}.
-Then, we initialize a simple geometric cooling schedule: 
+Then, we initialize a simple geometric cooling schedule \ref{equ:coolingSchedule}: 
 \begin{equation}
-    \label{coolingSchedule}
+    \label{equ:coolingSchedule}
     T_{n+1} = \alpha T_n
@@
 with initial temperature \(T_0=10000\), cooling rate \(\alpha=0.98\), and \(n = \{0, 1, 2, ..., 300\}\) passes.
-Hill-climbing moves can be accepted via the probability function \(P()\): 
+Hill-climbing moves can be accepted via the probability function in \ref{equ:moveAcceptance}: 
 \begin{equation}
-    \label{moveAcceptance}
+    \label{equ:moveAcceptance}
     P(s, s^*, T_n) = exp(\frac{E(s) - E(s^*)}{T_n})
@@
 Overlapping objects are spread out until every \texttt{SiteInst}-\texttt{Cell} is assigned its own unique \texttt{Site}-\texttt{BEL}.
+The process of spreading objects somewhat resembles primitive fluid dynamics simulators or thermal diffusion simulators, where packets of energy are spread around an object until each voxel or pixel meets a maximum threshold of energy.
 

### 2025-07-02 latex/paper/sections/11-placement_results.tex
lots of reformatting

@@
 Here, we will perform placement on a 2048th-order FIR Filter which was conveniently created as coursework for a VLSI course. 
-After passing synthesis, the FIR Filter design calls for the following primitive cells:
+After passing synthesis, the FIR Filter design calls for the primitive cells shown in the second column of table \ref{fig:utilization}.
+We will perform placement of this design on a \texttt{xc7z020} FPGA, which is a mid-ranged Xilinx device containing the available BELs shown in the first column of table \ref{fig:utilization}.
+Note that \texttt{LUT5-LUT6} pairs are counted as a single \texttt{LUT}.
+% Also note that the number of \texttt{CARRY4} BELs is equal to the total number of \texttt{SLICEL} and \texttt{SLICEM} Sites combined and that the number of \texttt{LUT}s is equal to four times that amount while the number of \texttt{FF}s is equal to eight times that amount.
 
-\begin{itemize}
-    \item 1693 \texttt{LUT}
-    \item 920 \texttt{FF}
-    \item 282 \texttt{CARRY4}
-    \item 32 \texttt{RAMB48E1}
-    \item 64 \texttt{DSP48E1}
-\end{itemize}
-
-We will perform placement of this design on a \texttt{xc7z020} FPGA, which is a mid-ranged Xilinx device containing the following relevant BELs:
-\begin{itemize}
-    \item 53200 \texttt{LUT} (counting \texttt{LUT5}-\texttt{LUT6} pairs as one \texttt{LUT})
-    \item 106400 \texttt{FF}
-    \item 13300 \texttt{CARRY4}
-    \item 171 \texttt{RAMB18E1}
-    \item 220 \texttt{DSP48E1}
-\end{itemize}
-
-Our expected overall utilization for the FIR Filter on this device will be:
-\begin{itemize}
-    \item Lookup Tables: 3\%
-    \item Flip Flops: 1\%
-    \item Carry: 2\%
-    \item BRAMs: 23\%
-    \item DSPs: 29\%
-\end{itemize}
+\vspace{0.5cm}
+{
+    \centering
+    \includegraphics[width=0.8\columnwidth]{figures/results/utilization.png}
+    \captionof{figure}{Utilization of BELs on a \texttt{xc7z020} FPGA for a FIR filter design}
+    \label{fig:utilization}
+}
+\vspace{0.5cm}
 
@@
 Interestingly, this placer also settled at a lower final system cost of 333K, which is a slight 4\% improvement over vanilla SA.
-We originally expected the greediness to be a detriment to finding the global minimum, but due to the inherent randomness of the algorithm, these expectations are never certain 
+We originally expected the greediness to be a detriment to finding the global minimum, but due to the inherent randomness of the algorithm, these expectations are never certain.
 Broader trends will require many more test trials to observe reliably.
-
-The best performing placer was \texttt{PlacerAnnealHybrid}, represented by the blue curve.
-We can observe that \texttt{AnnealHybrid} settles at a similar rate with the vanilla \texttt{AnnealRandom}, but eventually settles at a lower minimum at 322K, outperforming even \texttt{GreedyRandom}.
 
@@
 
+The best performing placer was \texttt{PlacerAnnealHybrid}, represented by the blue curve.
+We can observe that \texttt{AnnealHybrid} settles at a similar rate with the vanilla \texttt{AnnealRandom}, but eventually settles at a lower minimum at 322K, outperforming even \texttt{GreedyRandom}.
+
+{
+    \centering
+    \includegraphics[width=0.7\columnwidth]{figures/placement/cooling_schedule_10000_98.png}
+    \captionof{figure}{Cooling Schedule: \(T_0=10000\), \(\alpha=0.98\)}
+    \label{fig:cooling_schedule_10000_98}
+}
+
+\end{multicols}
 {
@@
     \includegraphics[valign=t, scale=0.13]{figures/results/PlacerGreedyRandom/00000299.png}
-    \captionof{figure}{PlacerGreedyRandom}
+    \captionof{figure}{\texttt{PlacerGreedyRandom}}
     \label{fig:PGRSnapshots}
@@
 
-\columnbreak
 {
@@
     \includegraphics[valign=t, scale=0.13]{figures/results/PlacerGreedyMidpoint/00000299.png}
-    \captionof{figure}{PlacerGreedyMidpoint}
+    \captionof{figure}{\texttt{PlacerGreedyMidpoint}}
     \label{fig:PGMSnapshots}
 }
-\vfill
+
+\newcolumn
 {
@@
     \includegraphics[valign=t, scale=0.13]{figures/results/PlacerAnnealRandom/00000299.png}
-    \captionof{figure}{PlacerAnnealRandom}
+    \captionof{figure}{\texttt{PlacerAnnealRandom}}
     \label{fig:PARSnapshots}
@@
     \includegraphics[valign=t, scale=0.13]{figures/results/PlacerAnnealMidpoint/00000299.png}
-    \captionof{figure}{PlacerAnnealMidpoint}
+    \captionof{figure}{\texttt{PlacerAnnealMidpoint}}
     \label{fig:PAMSnapshots}
 }
-\vfill
+
 {
@@
     \includegraphics[valign=t, scale=0.13]{figures/results/PlacerAnnealHybrid/00000299.png}
-    \captionof{figure}{PlacerAnnealHybrid}
+    \captionof{figure}{\texttt{PlacerAnnealHybrid}}
     \label{fig:PAHSnapshots}
 }
+
+\begin{multicols}{2}

### 2025-07-02 python/cooling_schedule.py

@@ -1,0 +1,46 @@
+"""
+plot_numbers.py
+Read a file of newline-delimited numbers and save a line-chart PNG.
+
+Usage:
+    python plot_numbers.py <input.txt> [output.png]
+"""
+import sys
+from pathlib import Path
+import matplotlib.pyplot as plt
+
+def main() -> None:
+    # ---- parse arguments ---------------------------------------------------
+    if len(sys.argv) not in (2, 3):
+        print("Usage: python plot_numbers.py <input.txt> [output.png]", file=sys.stderr)
+        sys.exit(1)
+
+    in_path  = Path(sys.argv[1])
+    out_path = Path(sys.argv[2]) if len(sys.argv) == 3 else in_path.with_suffix(".png")
+
+    # ---- read data ---------------------------------------------------------
+    try:
+        with in_path.open() as f:
+            values = [float(line.strip()) for line in f if line.strip()]
+    except FileNotFoundError:
+        sys.exit(f"Error: {in_path} not found.")
+    except ValueError as e:
+        sys.exit(f"Error parsing numbers: {e}")
+
+    if not values:
+        sys.exit("Error: no numeric data found.")
+
+    # ---- plot --------------------------------------------------------------
+    plt.figure(figsize=(4, 2.5))
+    plt.plot(range(len(values)), values, linewidth=1.5)
+    plt.title(in_path.name)
+    plt.xlabel("Passes")
+    plt.ylabel("Temperature")
+    plt.grid(True)
+    plt.tight_layout()
+    plt.savefig(out_path, dpi=150)
+    plt.close()
+    print(f"Saved plot to {out_path}")
+
+if __name__ == "__main__":
+    main()

### 2025-07-03 latex/paper/sections/10-placement.tex
more reorg

@@
 
-In more sophisticated placers, the proposed movement can also be selected by finding the centroid between \texttt{SiteInst}s that share a net with the current object, or even a hybrid of random and centroid selection.
-Moves that actively attempt to predict a better placement are often referred to as "directed" moves in literature. 
-Placers that take inspiration from physical interactions like gravity or electrostatics are often referred to as "force directed" placers and are akin to physics simulators.
-
-Directed moves as opposed to undirected moves give rise to the concept of legalization. 
-When using undirected moves as in SA, we can simply keep track of a \texttt{List} or \texttt{HashMap} of occupied \texttt{Site}s and available \texttt{Site}s and select from that data structure via random selection.
-However, for directed moves like centroid selection, the centroid between \texttt{Site}s may not necessarily fall directly on another discrete \texttt{Site} on the \texttt{device}. 
-Even when it does, the centroid \texttt{Site} may be incompatible with the \texttt{SiteInst} or may fail to meet some other placement constraint. 
-In such cases, we must do a neighborhood search around the centroid up to a certain radius to find a \texttt{Site} that satisfies all the constraints of the \texttt{SiteInst} at hand. 
-We will take some inspiration from ray-marching in graphics and do a diamond-shaped spiral search around the centroid where we check the legality of each \texttt{Site} as we step through the spiral and select the first \texttt{Site} that meets all of the placement constraints.
-
-In force directed placers, there is a concept of global placement followed by detailed placement
-Global placement places objects freely on an analog plane and is driven by directed movement or analytical solution.
-In global placement, objects do not have to adhere to any hardware constraints and can even overlap with one another.
-Detailed placement takes the global placement and attempts to fit it within the constraints of the hardware. 
-Detailed placement "legalize" the global placement by discretizing the analog positions of the objects onto the discrete grid of resources of the \texttt{device}.
-Overlapping objects are spread out until every \texttt{SiteInst}-\texttt{Cell} is assigned its own unique \texttt{Site}-\texttt{BEL}.
-The process of spreading objects somewhat resembles primitive fluid dynamics simulators or thermal diffusion simulators, where packets of energy are spread around an object until each voxel or pixel meets a maximum threshold of energy.
-
-There also exists other strategies to mix packing, placement, and legalization in different ways. (CITE THAT ONE PAPER).
-In this context, our placer follows a straightforward pack-legalize-place strategy.
-
 If the move reduces the HPWL cost, then the movement is accepted outright.
@@
 
-\newcolumn
 \begin{lstlisting}[language=java, caption={SA pseudocode: outer loop}, label={lst:sa_outer}]
@@
     \centering
-    \includegraphics[width=\columnwidth]{figures/placement/swapSingleSite.png}
+    \includegraphics[width=0.9\columnwidth]{figures/placement/swapSingleSite.png}
     \captionof{figure}{Single Site Swap Proposal}
@@
 }
-\vfill
 
-\begin{lstlisting}[language=java, caption={Cooling Schedule and Move Acceptance with Temperature}, label={lst:sa_acceptance}]
+\begin{lstlisting}[language=java, caption={Cooling Schedule and Move Acceptance}, label={lst:sa_acceptance}]
 protected List<Double> coolingSchedule;
@@
 \begin{multicols}{2}
+
+\subsection{Variations}
+\label{subsec:variations}
+
+In more sophisticated placers, the proposed movement can also be selected by finding the centroid between \texttt{SiteInst}s that share a net with the current object, or even a hybrid of random and centroid selection.
+We will implement variations on the vanilla SA algorithm that use centroid selection. 
+Moves that actively attempt to predict a better placement are often referred to as "directed" moves in literature. 
+Placers that take inspiration from physical interactions like gravity or electrostatics are often referred to as "force directed" placers and are akin to physics simulators.
+
+Directed moves as opposed to undirected moves give rise to the concept of legalization. 
+When using undirected moves as in SA, we can simply keep track of a \texttt{List} or \texttt{HashMap} of occupied \texttt{Site}s and available \texttt{Site}s and select from that data structure via random selection.
+However, for directed moves like centroid selection, the centroid between \texttt{Site}s may not necessarily fall directly on another discrete \texttt{Site} on the \texttt{device}. 
+Even when it does, the centroid \texttt{Site} may be incompatible with the \texttt{SiteInst} or may fail to meet some other placement constraint. 
+In such cases, we must do a neighborhood search around the centroid up to a certain radius to find a \texttt{Site} that satisfies all the constraints of the \texttt{SiteInst} at hand. 
+We will take some inspiration from ray-marching in graphics and do a diamond-shaped spiral search around the centroid where we check the legality of each \texttt{Site} as we step through the spiral and select the first \texttt{Site} that meets all of the placement constraints.
+
+In force directed placers, there is a concept of global placement followed by detailed placement.
+Global placement places objects freely on an analog plane and is driven purely by directed movement and analytical solution.
+In global placement stage, only non-hardware related objectives are considered. 
+Objects do not have to adhere to any hardware constraints and can even overlap with one another.
+Afterwards, detailed placement "legalizes" the global placement by discretizing the analog positions of the objects onto the discrete grid of resources of the \texttt{device}.
+Overlapping objects are iteratively spread out until every \texttt{SiteInst}-\texttt{Cell} is assigned its own unique \texttt{Site}-\texttt{BEL}.
+The spreading process somewhat resembles primitive fluid dynamics simulators or thermal diffusion simulators, where packets of energy are spread through a medium until each voxel or pixel of the medium falls below a maximum threshold of energy.
+There also exists other strategies to mix packing, placement, and legalization in different orders. 
+In this context, our placer follows a very straightforward pack-legalize-place strategy with SA.

### 2025-07-03 latex/paper/sections/11-placement_results.tex
more reorg

@@
 Any DSP subsystem can serve as a good candidate for such a demonstration. 
-Here, we will perform placement on a 2048th-order FIR Filter which was conveniently created as coursework for a VLSI course. 
+Here, we will perform placement on a 2048-order FIR Filter which was conveniently created as coursework for a VLSI course. 
 After passing synthesis, the FIR Filter design calls for the primitive cells shown in the second column of table \ref{fig:utilization}.
@@
 \begin{itemize}
-    \item \texttt{PlacerGreedyRandom}: all undirected random moves with greedy acceptance (always zero temperature)
-    \item \texttt{PlacerGreedyMidpoint}: all directed centroid moves with greedy acceptance (always zero temperature)
+    \item \texttt{PlacerGreedyRandom}: all undirected random moves with greedy acceptance
+    \item \texttt{PlacerGreedyMidpoint}: all directed centroid moves with greedy acceptance
     \item \texttt{PlacerAnnealRandom}: all undirected random moves with annealing acceptance \textbf{(the basic SA algorithm)}
@@
 \end{itemize}
-
-((Talk about the initial cost.))
 
@@
 
-The next three placers performed significantly better than the previous two placers.
-Our vanilla SA placer, \texttt{PlacerAnnealRandom} is represented by the green curve.
+The next three placers performed better than the previous two placers by decent margin.
+The vanilla SA placer, \texttt{PlacerAnnealRandom}, is represented by the green curve.
 We can observe that the cost decreases more steadily than the previous two placers, with noticeable spikes in total system cost due to hill-climbing, until settling at the 346K mark as the global temperature cools to zero.
@@
     \centering
-    \includegraphics[width=\columnwidth]{figures/results/combined_cost_history_linear.png}
+    \includegraphics[width=0.9\columnwidth]{figures/results/combined_cost_history_linear.png}
     \captionof{figure}{All placers overlaid}
@@
 The best performing placer was \texttt{PlacerAnnealHybrid}, represented by the blue curve.
-We can observe that \texttt{AnnealHybrid} settles at a similar rate with the vanilla \texttt{AnnealRandom}, but eventually settles at a lower minimum at 322K, outperforming even \texttt{GreedyRandom}.
+Starting at zero passes, the placer randomly selects between random selection and centroid selection, allowing the placer to make on average more cost-reducing moves, while still maintaining some diversity of moves for hill-climbing.
+Then, at near-zero temperature, the placer makes exclusively centroid moves with annealing acceptance to speed up the final crystallization into the local minimum. 
+With centroid moves and annealing acceptance, hill-climbing can only be achieved via swaps.
+We can observe that \texttt{AnnealHybrid} settles at a similar rate with the vanilla \texttt{AnnealRandom}, but eventually settles at a lower minimum at 322K, slightly outperforming \texttt{GreedyRandom}.
 

### 2025-07-11 latex/paper/sections/10-placement.tex
snapshot discussion

@@
 In force directed placers, there is a concept of global placement followed by detailed placement.
-Global placement places objects freely on an analog plane and is driven purely by directed movement and analytical solution.
+Global placement places object instances freely on an analog plane and is driven purely by directed movement and analytical solution.
 In global placement stage, only non-hardware related objectives are considered. 
-Objects do not have to adhere to any hardware constraints and can even overlap with one another.
-Afterwards, detailed placement "legalizes" the global placement by discretizing the analog positions of the objects onto the discrete grid of resources of the \texttt{device}.
-Overlapping objects are iteratively spread out until every \texttt{SiteInst}-\texttt{Cell} is assigned its own unique \texttt{Site}-\texttt{BEL}.
+Object instance placements do not have to adhere to any hardware constraints and can even overlap with one another.
+Afterwards, detailed placement "legalizes" the global placement by discretizing the analog positions of the instances onto the discrete grid of resources of the \texttt{device}.
+Overlapping instances are iteratively spread out until every \texttt{SiteInst}-\texttt{Cell} is assigned its own unique \texttt{Site}-\texttt{BEL}.
 The spreading process somewhat resembles primitive fluid dynamics simulators or thermal diffusion simulators, where packets of energy are spread through a medium until each voxel or pixel of the medium falls below a maximum threshold of energy.

### 2025-07-11 latex/paper/sections/11-placement_results.tex
snapshot discussion

@@
 Interestingly, this placer also settled at a lower final system cost of 333K, which is a slight 4\% improvement over vanilla SA.
-We originally expected the greediness to be a detriment to finding the global minimum, but due to the inherent randomness of the algorithm, these expectations are never certain.
-Broader trends will require many more test trials to observe reliably.
+We originally expected the greediness to be a detriment to finding the global minimum, but due to the inherent randomness of the algorithm, these broader trends will require many more test trials to reliably observe.
 
@@
 \begin{multicols}{2}
+
+    Figures \ref{fig:PGRSnapshots}, \ref{fig:PGMSnapshots}, \ref{fig:PARSnapshots}, \ref{fig:PAMSnapshots}, and \ref{fig:PAHSnapshots} show snapshots of the placement progress on the physical \texttt{device} after 0, 10, 100, and 300 passes for each placer variant.
+
+    Please zoom into the document to render the figures in greater detail.
+    All black areas represent purely routing \texttt{Site}s that contain no logic resources. 
+    All gray or dull-colored boxes represent \texttt{Site}s that are unoccupied by \texttt{SiteInst}s, while brightly filled boxes represent \texttt{Site}s that are occupied by the supported library of \texttt{SiteInst}s.
+    Cyan boxes represent occupied \texttt{SLICEL}s, magenta boxes represent occupied \texttt{RAMB18E1}s, and yellow boxes represent occupied \texttt{DSP48E1}s.
+
+    The tangle of red lines represent the nets between the \texttt{Site}s, which are yet to be routed at this stage. 
+    Brighter red lines represent nets of higher total HPWL, while duller red lines represent nets of lower total HPWL.
+
+    Notice there are two brightly red nets in every figure that have a particularly high fan-out:
+    one sourced from the near-center, and another sourced from the top right of the device.
+
+    The singular white box in the near-center of all of the figures represents an occupied \texttt{BUFGCTRL} \texttt{Site}, which is a clock buffer used to route clock trees. 
+    Since clock nets tend to have very high fanout, they can use up a significant portion of the routing resource if they are routed through general routing.
+    Modern FPGAs like the 7-Series have dedicated clock routes which can efficiently distribute clock signals to all logical \texttt{Site}s without burdening the general routing resources.
+    These special clock routes are accessed by routing the off-chip crystal oscillator clock to one of these clock buffers which connects to the clock tree network.
+    It is always best practice to use them to improve the routability of the design, and simultaneously, to reduce clock skew and improve timing. 
+    Our simple FIR Filter design only operates on a single clock domain, so only one clock buffer \texttt{Site} is used. 
+    An HDL design operating on many clock domains will use more clock buffers \texttt{Site}s or a dedicated clock generator such as a phase locked loop (PLL) \texttt{Site} or mixed-mode clock manager (MCMM) \texttt{Site}.
+
+    The right and left edges of the FPGA device is where most of the general purpose IO (GPIO) is located. 
+    In the top right of the device is where we source our reset signal. 
+    Like the clock net, the reset net also has high fanout and is supplied to all synchronous resources.
+    Unlike clock signals, reset signals are not subject to sensitive timing constraints as they are not active on most clock cycles and can tolerate higher skew.
+    Thus, they are simply routed via general routing.
+
+    Observe how \texttt{PlacerGreedyMidpoint} \ref{fig:PGMSnapshots} and \texttt{PlacerAnnealMidpoint} \ref{fig:PAMSnapshots} both show almost not shrinkage in profile after the first 10 iterations, likely owing to the inability to hill-climb.

### 2025-07-13 latex/paper/sections/11-placement_results.tex
start future work

@@
 
-\vspace{0.5cm}
+\vspace{0.25cm}
 {
@@
 }
-\vspace{0.5cm}
+\vspace{0.25cm}
 
@@
     \centering
-    \includegraphics[width=0.9\columnwidth]{figures/results/combined_cost_history_linear.png}
+    \includegraphics[width=\columnwidth]{figures/results/combined_cost_history_linear.png}
     \captionof{figure}{All placers overlaid}
@@
     \centering
-    \includegraphics[width=0.7\columnwidth]{figures/placement/cooling_schedule_10000_98.png}
+    \includegraphics[width=0.6\columnwidth]{figures/placement/cooling_schedule_10000_98.png}
     \captionof{figure}{Cooling Schedule: \(T_0=10000\), \(\alpha=0.98\)}
@@
 
+\newpage
+Figures \ref{fig:PGRSnapshots}, \ref{fig:PGMSnapshots}, \ref{fig:PARSnapshots}, \ref{fig:PAMSnapshots}, and \ref{fig:PAHSnapshots} show snapshots of the placement progress on the physical \texttt{device} after 0, 10, 100, and 300 passes for each placer variant.
+
+Please zoom into the document to render the figures in greater detail.
+All black areas represent purely routing \texttt{Site}s that contain no logic resources. 
+All gray or dull-colored boxes represent \texttt{Site}s that are unoccupied by \texttt{SiteInst}s, while brightly filled boxes represent \texttt{Site}s that are occupied by the supported library of \texttt{SiteInst}s.
+Cyan boxes represent occupied \texttt{SLICEL}s, magenta boxes represent occupied \texttt{RAMB18E1}s, and yellow boxes represent occupied \texttt{DSP48E1}s.
+
+The tangle of red lines represent the nets between the \texttt{Site}s, which are yet to be routed at this stage. 
+Brighter red lines represent nets of higher total HPWL, while duller red lines represent nets of lower total HPWL.
+
+Notice there are two brightly red nets in every figure that have a particularly high fan-out:
+one sourced from the near-center, and another sourced from the top right of the device.
+
+The singular white box in the near-center of all of the figures represents an occupied \texttt{BUFGCTRL} \texttt{Site}, which is a clock buffer used to route clock trees. 
+Since clock nets tend to have very high fanout, they can use up a significant portion of the routing resources if they are routed through general routing.
+Modern FPGAs like the 7-Series have dedicated clock routes which can efficiently distribute clock signals to all logical \texttt{Site}s without burdening the general routing resources.
+These special clock routes are accessed by routing the raw clock signal from the crystal oscillator IO port to one these on-chip clock buffers which connects to the on-chip clock tree network.
+It is always best practice to use them to improve the routability of the design, and simultaneously, to reduce clock skew and improve timing. 
+Our simple FIR Filter design only operates on a single clock domain, so only one clock buffer \texttt{Site} is used. 
+An HDL design operating on many clock domains will use more clock buffers or a dedicated clock generator such as a phase locked loop (PLL) or mixed-mode clock manager (MCMM).
+
+The right and left edges of the FPGA device is where most of the general purpose I/O (GPIO) is located. 
+In the top right of the device is where we source our reset signal. 
+Like the clock net, the reset net also has high fanout and is supplied to all synchronous elements.
+Unlike clock signals, reset signals are not subject to sensitive timing constraints as they are not active on most clock cycles and can tolerate higher skew.
+Thus, they are simply routed via general routing.
+
+Observe how \texttt{PlacerGreedyMidpoint} \ref{fig:PGMSnapshots} and \texttt{PlacerAnnealMidpoint} \ref{fig:PAMSnapshots} both show almost not shrinkage in profile after the first 10 iterations, likely owing to the inability to hill-climb, while the other three placers show incremental shrinkage.
+
+Also note the sparsity of the \texttt{DSP48E1} (yellow) and \texttt{RAMB18E1} (magenta) \texttt{Site} columns compared to the large grid of \texttt{SLICEL}s (cyan).
+
+\newpage
 \end{multicols}
@@
 \begin{multicols}{2}
-
-    Figures \ref{fig:PGRSnapshots}, \ref{fig:PGMSnapshots}, \ref{fig:PARSnapshots}, \ref{fig:PAMSnapshots}, and \ref{fig:PAHSnapshots} show snapshots of the placement progress on the physical \texttt{device} after 0, 10, 100, and 300 passes for each placer variant.
-
-    Please zoom into the document to render the figures in greater detail.
-    All black areas represent purely routing \texttt{Site}s that contain no logic resources. 
-    All gray or dull-colored boxes represent \texttt{Site}s that are unoccupied by \texttt{SiteInst}s, while brightly filled boxes represent \texttt{Site}s that are occupied by the supported library of \texttt{SiteInst}s.
-    Cyan boxes represent occupied \texttt{SLICEL}s, magenta boxes represent occupied \texttt{RAMB18E1}s, and yellow boxes represent occupied \texttt{DSP48E1}s.
-
-    The tangle of red lines represent the nets between the \texttt{Site}s, which are yet to be routed at this stage. 
-    Brighter red lines represent nets of higher total HPWL, while duller red lines represent nets of lower total HPWL.
-
-    Notice there are two brightly red nets in every figure that have a particularly high fan-out:
-    one sourced from the near-center, and another sourced from the top right of the device.
-
-    The singular white box in the near-center of all of the figures represents an occupied \texttt{BUFGCTRL} \texttt{Site}, which is a clock buffer used to route clock trees. 
-    Since clock nets tend to have very high fanout, they can use up a significant portion of the routing resource if they are routed through general routing.
-    Modern FPGAs like the 7-Series have dedicated clock routes which can efficiently distribute clock signals to all logical \texttt{Site}s without burdening the general routing resources.
-    These special clock routes are accessed by routing the off-chip crystal oscillator clock to one of these clock buffers which connects to the clock tree network.
-    It is always best practice to use them to improve the routability of the design, and simultaneously, to reduce clock skew and improve timing. 
-    Our simple FIR Filter design only operates on a single clock domain, so only one clock buffer \texttt{Site} is used. 
-    An HDL design operating on many clock domains will use more clock buffers \texttt{Site}s or a dedicated clock generator such as a phase locked loop (PLL) \texttt{Site} or mixed-mode clock manager (MCMM) \texttt{Site}.
-
-    The right and left edges of the FPGA device is where most of the general purpose IO (GPIO) is located. 
-    In the top right of the device is where we source our reset signal. 
-    Like the clock net, the reset net also has high fanout and is supplied to all synchronous resources.
-    Unlike clock signals, reset signals are not subject to sensitive timing constraints as they are not active on most clock cycles and can tolerate higher skew.
-    Thus, they are simply routed via general routing.
-
-    Observe how \texttt{PlacerGreedyMidpoint} \ref{fig:PGMSnapshots} and \texttt{PlacerAnnealMidpoint} \ref{fig:PAMSnapshots} both show almost not shrinkage in profile after the first 10 iterations, likely owing to the inability to hill-climb.

### 2025-07-13 latex/paper/sections/12-future_work.tex
start future work

@@
 \section{Future Work}
+
+\subsection{Expand Library of Primitives}
 In this paper we have made many simplifications to the problem space just to make the placer easier to implement.
@@
 
+\subsection{Robustness}
+In its current state, the prepacker and packer struggle with data busses larger than 24-bits, especially with DSP functions like addition and multiplication.
+
+\subsection{Improvements to Packing}
 There is also no rule saying we must pack \texttt{Cells} into \texttt{SiteInst}s before placement or that we follow a strict prepacking-packing-placement flow. 
+The latest editions of Vivado performs BEL-centric placement without necessarily locking \texttt{Cell}s into \texttt{Site}s, allowing a higher granularity of movement of \texttt{Cell}s during placement.
 
-The latest editions of Vivado performs BEL-centric placement without necessarily locking \texttt{Cell}s into \texttt{Site}s, allowing a higher granularity of movement of \texttt{Cell}s during placement.
+\subsection{Improvements to SA}
+
+\subsection{Force-Directed and Analytical Placement}
+
+In more sophisticated placers, the proposed movement can also be selected by finding the centroid between \texttt{SiteInst}s that share a net with the current object, or even a hybrid of random and centroid selection.
+We will implement variations on the vanilla SA algorithm that use centroid selection. 
+Moves that actively attempt to predict a better placement are often referred to as "directed" moves in literature. 
+Placers that take inspiration from physical interactions like gravity or electrostatics are often referred to as "force directed" placers and are akin to physics simulators.
+
+Directed moves as opposed to undirected moves give rise to the concept of legalization. 
+When using undirected moves as in SA, we can simply keep track of a \texttt{List} or \texttt{HashMap} of occupied \texttt{Site}s and available \texttt{Site}s and select from that data structure via random selection.
+However, for directed moves like centroid selection, the centroid between \texttt{Site}s may not necessarily fall directly on another discrete \texttt{Site} on the \texttt{device}. 
+Even when it does, the centroid \texttt{Site} may be incompatible with the \texttt{SiteInst} or may fail to meet some other placement constraint. 
+In such cases, we must do a neighborhood search around the centroid up to a certain radius to find a \texttt{Site} that satisfies all the constraints of the \texttt{SiteInst} at hand. 
+We will take some inspiration from ray-marching in graphics and do a diamond-shaped spiral search around the centroid where we check the legality of each \texttt{Site} as we step through the spiral and select the first \texttt{Site} that meets all of the placement constraints.
+
+In force directed placers, there is a concept of global placement followed by detailed placement.
+Global placement places object instances freely on an analog plane and is driven purely by directed movement and analytical solution.
+In global placement stage, only non-hardware related objectives are considered. 
+Object instance placements do not have to adhere to any hardware constraints and can even overlap with one another.
+Afterwards, detailed placement "legalizes" the global placement by discretizing the analog positions of the instances onto the discrete grid of resources of the \texttt{device}.
+Overlapping instances are iteratively spread out until every \texttt{SiteInst}-\texttt{Cell} is assigned its own unique \texttt{Site}-\texttt{BEL}.
+The spreading process somewhat resembles primitive fluid dynamics simulators or thermal diffusion simulators, where packets of energy are spread through a medium until each voxel or pixel of the medium falls below a maximum threshold of energy.
+There also exists other strategies to mix packing, placement, and legalization in different orders, such as those studied in \cite{ExplicitPacking}. 
+In this context, our placer follows a very straightforward pack-legalize-place strategy with SA.

### 2025-07-13 latex/paper/sections/10-placement.tex
start future work
citations

--- 2025-07-03 latex/paper/sections/10-placement.tex
@@
 \begin{multicols}{2}
-
-\subsection{Variations}
-\label{subsec:variations}
-
-In more sophisticated placers, the proposed movement can also be selected by finding the centroid between \texttt{SiteInst}s that share a net with the current object, or even a hybrid of random and centroid selection.
-We will implement variations on the vanilla SA algorithm that use centroid selection. 
-Moves that actively attempt to predict a better placement are often referred to as "directed" moves in literature. 
-Placers that take inspiration from physical interactions like gravity or electrostatics are often referred to as "force directed" placers and are akin to physics simulators.
-
-Directed moves as opposed to undirected moves give rise to the concept of legalization. 
-When using undirected moves as in SA, we can simply keep track of a \texttt{List} or \texttt{HashMap} of occupied \texttt{Site}s and available \texttt{Site}s and select from that data structure via random selection.
-However, for directed moves like centroid selection, the centroid between \texttt{Site}s may not necessarily fall directly on another discrete \texttt{Site} on the \texttt{device}. 
-Even when it does, the centroid \texttt{Site} may be incompatible with the \texttt{SiteInst} or may fail to meet some other placement constraint. 
-In such cases, we must do a neighborhood search around the centroid up to a certain radius to find a \texttt{Site} that satisfies all the constraints of the \texttt{SiteInst} at hand. 
-We will take some inspiration from ray-marching in graphics and do a diamond-shaped spiral search around the centroid where we check the legality of each \texttt{Site} as we step through the spiral and select the first \texttt{Site} that meets all of the placement constraints.
-
-In force directed placers, there is a concept of global placement followed by detailed placement.
-Global placement places objects freely on an analog plane and is driven purely by directed movement and analytical solution.
-In global placement stage, only non-hardware related objectives are considered. 
-Objects do not have to adhere to any hardware constraints and can even overlap with one another.
-Afterwards, detailed placement "legalizes" the global placement by discretizing the analog positions of the objects onto the discrete grid of resources of the \texttt{device}.
-Overlapping objects are iteratively spread out until every \texttt{SiteInst}-\texttt{Cell} is assigned its own unique \texttt{Site}-\texttt{BEL}.
-The spreading process somewhat resembles primitive fluid dynamics simulators or thermal diffusion simulators, where packets of energy are spread through a medium until each voxel or pixel of the medium falls below a maximum threshold of energy.
-There also exists other strategies to mix packing, placement, and legalization in different orders. 
-In this context, our placer follows a very straightforward pack-legalize-place strategy with SA.

### 2025-07-17 latex/paper/sections/11-placement_results.tex
more test trials

@@
 
-Also note the sparsity of the \texttt{DSP48E1} (yellow) and \texttt{RAMB18E1} (magenta) \texttt{Site} columns compared to the large grid of \texttt{SLICEL}s (cyan).
+Also note the sparsity of the \texttt{DSP48E1} (yellow) and \texttt{RAMB18E1} (magenta) \texttt{Site} columns compared to the more uniform grid of \texttt{SLICEL}s (cyan).
 

### 2025-07-17 java/src/main/java/placer/Main.java

@@ -5,6 +5,7 @@ import java.util.logging.FileHandler;
 import java.util.logging.SimpleFormatter;
 import java.util.logging.Level;
 
+import java.io.FileWriter;
 import java.io.IOException;
 
 import java.util.ArrayList;
@@ -79,90 +80,82 @@ public class Main {
             // takes the packedDesign and figures out an optimal mapping of SiteInsts onto
             // Sites via simulated annealing, analytical, electrostatic placement, etc.
             // works entirely on the SiteInst/Site/Tile level.
-            List<PlacerAnnealRandom> SAPlacers = new ArrayList<PlacerAnnealRandom>();
-            SAPlacers.add(new PlacerAnnealRandom(rootDir, design, device, region));
-            SAPlacers.add(new PlacerAnnealMidpoint(rootDir, design, device, region));
-            SAPlacers.add(new PlacerAnnealHybrid(rootDir, design, device, region));
-            SAPlacers.add(new PlacerGreedyRandom(rootDir, design, device, region));
-            SAPlacers.add(new PlacerGreedyMidpoint(rootDir, design, device, region));
+            // List<PlacerAnnealRandom> SAPlacers = new ArrayList<PlacerAnnealRandom>();
+            // SAPlacers.add(new PlacerAnnealRandom(rootDir, design, device, region));
+            // SAPlacers.add(new PlacerAnnealMidpoint(rootDir, design, device, region));
+            // SAPlacers.add(new PlacerAnnealHybrid(rootDir, design, device, region));
+            // SAPlacers.add(new PlacerGreedyRandom(rootDir, design, device, region));
+            // SAPlacers.add(new PlacerGreedyMidpoint(rootDir, design, device, region));
 
-            for (PlacerAnnealRandom placer : SAPlacers) {
-                System.out.println("\n\nStarting " + placer.getPlacerName() + "... \n\n");
-                placer.makeOutputDirs(placer.getPlacerName());
-                placer.initCoolingSchedule(10000.0d, 0.98d, 300);
-                placer.run(packedDesign);
+            // for (PlacerAnnealRandom placer : SAPlacers) {
+            // System.out.println("\n\nStarting " + placer.getPlacerName() + "... \n\n");
+            // placer.makeOutputDirs(placer.getPlacerName());
+            // placer.initCoolingSchedule(10000.0d, 0.98d, 300);
+            // placer.run(packedDesign);
+            // }
+
[…]

### 2025-07-17 java/src/main/java/placer/Placer.java

@@ -413,11 +413,19 @@ public abstract class Placer {
     public void exportCostHistory(String fileName) throws IOException {
         FileWriter csv = new FileWriter(fileName);
         csv.write("Iter, Cost");
-        for (int i = 0; i < costHistory.size(); i++) {
-            csv.write("\n" + i + ", " + costHistory.get(i));
+        for (int i = 0; i < this.costHistory.size(); i++) {
+            csv.write("\n" + i + ", " + this.costHistory.get(i));
         }
         if (csv != null)
             csv.close();
+    }
+
+    public List<Double> getCostHistory() {
+        return this.costHistory;
+    }
+
+    public List<Double> getCoolingSchedule() {
+        return this.getCoolingSchedule();
     }
 
     public void printSiteInstPlacements(PackedDesign packedDesign) throws IOException {

### 2025-07-17 python/overlay_convergences.py

@@ -1,100 +1,97 @@
-"""
-Script to combine cost_history.csv files from each placer into a single CSV
-and plot the combined cost histories on a logarithmic y-scale.
-
-Place this script in the `python` subdirectory of your project (e.g.,
-`~/workspace/dev/place-and-route/python`).
-It will look under `outputs/placers/*/printout/cost_history.csv`, merge them on the Iter column,
-save the merged CSV to `combined_cost_history.csv` in the project root,
-and save a log-scale plot to `combined_cost_history.png` in the project root.
-"""
-import sys
-from pathlib import Path
 import pandas as pd
+from pathlib import Path
 import matplotlib.pyplot as plt
+import re
+
+
+def parse_label(label):
+    match = re.match(r'PlacerAnnealRandom_(\d+)_(\d+)', label)
+    if match:
+        temp = int(match.group(1))
+        rate = int(match.group(2))
+        return temp, rate
+    return None, None
+
+
+def cooling_to_brightness(cooling_rate, min_rate=90, max_rate=98):
+    return (cooling_rate - min_rate) / (max_rate - min_rate)
+
+
+def plot_group(temp_group, temp_to_color, df, output_path, title_suffix=""):
+    plt.figure(figsize=(12, 8))
+    for label in temp_group:
+        temp, rate = parse_label(label)
+        base_rgb = temp_to_color[temp]
+        alpha = 0.4 + 0.6 * (1 - cooling_to_brightness(rate))
+        y = pd.to_numeric(df[label], errors='coerce').dropna().values
+        x = range(len(y))
+        plt.plot(x, y, color=base_rgb, label=label, alpha=alpha, linewidth=1.5)
+
+    plt.xlabel("Iteration")
+    plt.ylabel("Cost")
+    plt.title(f"Cost History of SA Placers {title_suffix}")
+    plt.legend(fontsize="small", bbox_to_anchor=(1.05, 1), loc="upper left")
+    plt.grid(True)
+    plt.tight_layout()
+    plt.savefig(output_path)
+    plt.close()
+
 
 def main():
-    # Determine directories
-    script_dir = Path(__file__).resolve().parent  # .../project_root/python
-    root_dir = script_dir.parent                  # .../project_root
-    placers_dir = root_dir / "outputs" / "placers"
+    script_dir = Path(__file__).resolve().parent
+    root_dir = script_dir.parent
 
-    if not placers_dir.is_dir():
[…]

### 2025-07-21 latex/paper/sections/11-placement_results.tex
cooling schedules

--- 2025-07-13 latex/paper/sections/11-placement_results.tex
@@
 }
+\vspace{1.0cm}
 
-\newpage
 Figures \ref{fig:PGRSnapshots}, \ref{fig:PGMSnapshots}, \ref{fig:PARSnapshots}, \ref{fig:PAMSnapshots}, and \ref{fig:PAHSnapshots} show snapshots of the placement progress on the physical \texttt{device} after 0, 10, 100, and 300 passes for each placer variant.
@@
 Brighter red lines represent nets of higher total HPWL, while duller red lines represent nets of lower total HPWL.
-
-Notice there are two brightly red nets in every figure that have a particularly high fan-out:
+Notice there are two brightly red nets in every figure that have a particularly high fanout:
 one sourced from the near-center, and another sourced from the top right of the device.
 
-The singular white box in the near-center of all of the figures represents an occupied \texttt{BUFGCTRL} \texttt{Site}, which is a clock buffer used to route clock trees. 
-Since clock nets tend to have very high fanout, they can use up a significant portion of the routing resources if they are routed through general routing.
-Modern FPGAs like the 7-Series have dedicated clock routes which can efficiently distribute clock signals to all logical \texttt{Site}s without burdening the general routing resources.
+The singular white box in the near-center of all of the figures represents an occupied \texttt{BUFGCTRL} \texttt{Site}, which is a clock buffer used to route clock nets. 
+Since clock nets tend to have very high fanout, they can use up an unnecessarily large portion of routing resources if they are routed through general routing.
+Modern FPGAs like the 7-Series have dedicated clock tree routes which can evenly distribute clock signals to all logical \texttt{Site}s without burdening the general routing resources.
 These special clock routes are accessed by routing the raw clock signal from the crystal oscillator IO port to one these on-chip clock buffers which connects to the on-chip clock tree network.
@@
 
-Also note the sparsity of the \texttt{DSP48E1} (yellow) and \texttt{RAMB18E1} (magenta) \texttt{Site} columns compared to the large grid of \texttt{SLICEL}s (cyan).
+Also note the sparsity of the \texttt{DSP48E1} (yellow) and \texttt{RAMB18E1} (magenta) \texttt{Site} columns compared to the more uniform grid of \texttt{SLICEL}s (cyan), which poses a computational challenge for the legalization process in directed placement.
 
@@
 \begin{multicols}{2}
+    In addition to variations on the basic SA algorithm, we also test various cooling schedules on the basic SA algorithm.
+    Shown in figure \ref{fig:VaryCoolingSchedules} are the cost histories for various cooling schedules, grouped three different initial temperatures \texttt{(5000, 10000, and 15000)}. 
+    For each initial temperature group, we test four different cooling rates \texttt{(0.92, 0.94, 0.96, and 0.98)}, with the darkness of each curve corresponding with a slower cooling rate.
+\end{multicols}
+
+{
+    \centering
+    \includegraphics[valign=t, scale=0.55]{figures/results/cost_history_5000.png}
+    \includegraphics[valign=t, scale=0.55]{figures/results/cost_history_10000.png}
+    \includegraphics[valign=t, scale=0.55]{figures/results/cost_history_15000.png}
+    \includegraphics[valign=t, scale=0.55]{figures/results/combined_cost_history_cooling.png}
+    \captionof{figure}{\texttt{PlacerAnnealRandom} with various cooling schedules.}
+    \label{fig:VaryCoolingSchedules}
+}
+
+\begin{multicols}{2}

### 2025-07-21 latex/paper/sections/12-future_work.tex
citations
cooling schedules

--- 2025-05-18 latex/paper/sections/12-future_work.tex
@@
 This means that we have rendered about 25\% of the CLB fabric unusable which will inevitably hurt wirelength minimization during placement since the \texttt{SiteInst}s must be spread over a larger area.
+Enabling \texttt{SLICEL}-\texttt{SLICEM} homogeneity can lead do greater logic density and consequently less total HPWL, but can make the packing process more complex and may contribute to higher routing congestion.
 
+\subsection{Robustness}
+In its current state, the prepacker and packer struggle with data busses larger than 24-bits, especially with DSP functions like multiplication. 
+In such designs, the synthesizer will synthesize \texttt{CARRY4} chains with particular \texttt{EDIFHierPortInst} configurations that are currently not supported by our packer which eventually lead to errors or failures in the subsequent routing stage.
+Further work is required to resolve these routing constraints.
+
+\subsection{Improvements to Packing}
 There is also no rule saying we must pack \texttt{Cells} into \texttt{SiteInst}s before placement or that we follow a strict prepacking-packing-placement flow.
 
+Our current pack-place approach is a \texttt{Site}-centric approach and resembles that of Xilinx ISE, which is the predecessor to the Vivado design suite.
 The latest editions of Vivado performs BEL-centric placement without necessarily locking \texttt{Cell}s into \texttt{Site}s, allowing a higher granularity of movement of \texttt{Cell}s during placement.
+
+{
+    \centering
+    \includegraphics[width=\columnwidth]{figures/future_work/legalization.png}
+    \captionof{figure}{Representative FPGA placement and packing flows. Figure taken from Wuxi et al. (2019), page 1 \cite{ExplicitPacking}}
+}
+\vspace{0.25cm}
+
+\subsection{Improvements to SA}
+
+\subsection{Force-Directed and Analytical Placement}

### 2025-07-21 python/overlay_convergences.py

@@ -18,7 +18,7 @@ def cooling_to_brightness(cooling_rate, min_rate=90, max_rate=98):
 
 
 def plot_group(temp_group, temp_to_color, df, output_path, title_suffix=""):
-    plt.figure(figsize=(12, 8))
+    plt.figure(figsize=(6, 4))
     for label in temp_group:
         temp, rate = parse_label(label)
         base_rgb = temp_to_color[temp]
@@ -29,8 +29,9 @@ def plot_group(temp_group, temp_to_color, df, output_path, title_suffix=""):
 
     plt.xlabel("Iteration")
     plt.ylabel("Cost")
+    plt.ylim(300000, 1000000)
     plt.title(f"Cost History of SA Placers {title_suffix}")
-    plt.legend(fontsize="small", bbox_to_anchor=(1.05, 1), loc="upper left")
+    plt.legend(fontsize="small")
     plt.grid(True)
     plt.tight_layout()
     plt.savefig(output_path)
@@ -52,7 +53,7 @@ def main():
     temp_to_labels = {}
 
     # --- Combined plot ---
-    plt.figure(figsize=(12, 8))
+    plt.figure(figsize=(6, 4))
 
     for label in labels:
         if label.strip().lower() in ("iteration", "unnamed: 0"):
@@ -79,8 +80,8 @@ def main():
 
     plt.xlabel("Iteration")
     plt.ylabel("Cost")
-    plt.title("Cost History of SA Placers")
-    plt.legend(fontsize="small", bbox_to_anchor=(1.05, 1), loc="upper left")
+    plt.title("Combined Cost History of all SA Placers")
+    plt.legend(fontsize="small")
     plt.grid(True)
     plt.tight_layout()
     plt.savefig(root_dir / "outputs" / "combined_cost_history_plot.png")

### 2025-07-22 latex/paper/sections/11-placement_results.tex
wider cooling rate range

@@
     In addition to variations on the basic SA algorithm, we also test various cooling schedules on the basic SA algorithm.
-    Shown in figure \ref{fig:VaryCoolingSchedules} are the cost histories for various cooling schedules, grouped three different initial temperatures \texttt{(5000, 10000, and 15000)}. 
-    For each initial temperature group, we test four different cooling rates \texttt{(0.92, 0.94, 0.96, and 0.98)}, with the darkness of each curve corresponding with a slower cooling rate.
+    Shown in figure \ref{fig:VaryCoolingSchedules} are the cost histories for various cooling schedules, grouped into three different initial temperatures \texttt{(5000, 10000, and 15000)}. 
+    For each initial temperature group, we test five different cooling rates \texttt{(0.79, 0.84, 0.89, 0.94, and 0.99)}, with the darkness of each curve corresponding with a faster cooling rate.
+    The curves suggest that a lower initial temperature and faster cooling rate both contribute to a faster convergence to the global minimum, with generally small differences of less than 10\% in final HPWL cost.
+
 \end{multicols}

### 2025-07-22 latex/paper/sections/12-future_work.tex
wider cooling rate range

--- 2025-05-18 latex/paper/sections/12-future_work.tex
@@
 \section{Future Work}
+
+\subsection{Improve Robustness}
 In this paper we have made many simplifications to the problem space just to make the placer easier to implement.
@@
 This means that we have rendered about 25\% of the CLB fabric unusable which will inevitably hurt wirelength minimization during placement since the \texttt{SiteInst}s must be spread over a larger area.
+Enabling \texttt{SLICEL}-\texttt{SLICEM} homogeneity can lead do greater logic density and consequently less total HPWL, but can make the packing process more complex and may contribute to higher routing congestion.
+
+We can also add packing support for other Xilinx primitives such as the shift-register primitive \texttt{SRLx}, distributed RAM primitives \texttt{RAMSx}, or even \texttt{LATCH} primitives as discussed in \ref{sec:7_series}.
+Adding support for additional primitives and macros will allow our placer to handle a wider range of HDL designs and will require deeper consideration of hardware constraints to ensure robustness.
+
+In its current state, the prepacker and packer struggle to handle signals larger than 24-bits, especially when involved in DSP functions like addition and multiplication. 
+In such designs, the Vivado synthesizer may synthesize long \texttt{CARRY4} chains with particular \texttt{EDIFHierPortInst} configurations that are currently not handled by our packer and eventually to failures in the subsequent routing stage.
+Further work is required to resolve these constraints in the packer.
+
+\subsection{Variations on Packing}
+Our current placer follows a \texttt{Site}-centric approach that resembles that of Xilinx ISE, which is the predecessor to the Vivado design suite that was primarily used in the 90s and 2000s.
+The more modern Vivado now performs BEL-centric placement without necessarily locking \texttt{Cell}s into \texttt{Site}s, allowing for a higher granularity of movement of \texttt{Cell}s. 
+Enabling BEL-centric movement as opposed to Site-centric movement in our placer can improve HPWL minimization, but will add much more complexity to the packing and placement process to ensure robustness with respect to hardware constraints.
 
 There is also no rule saying we must pack \texttt{Cells} into \texttt{SiteInst}s before placement or that we follow a strict prepacking-packing-placement flow. 
+Future work can 
+{
+    \centering
+    \includegraphics[width=\columnwidth]{figures/future_work/legalization.png}
+    \captionof{figure}{Representative FPGA placement and packing flows. Figure taken from Wuxi et al. (2019), page 1 \cite{ExplicitPacking}}
+}
+\vspace{0.25cm}
 
-The latest editions of Vivado performs BEL-centric placement without necessarily locking \texttt{Cell}s into \texttt{Site}s, allowing a higher granularity of movement of \texttt{Cell}s during placement.
+\subsection{Add Hard Macro Support}
+
+\subsection{Improvements to SA}
+
+\subsection{Force-Directed and Analytical Placement}

### 2025-07-22 java/src/main/java/placer/Main.java

@@ -99,9 +99,9 @@ public class Main {
 
             List<PlacerAnnealRandom> SAPlacers = new ArrayList<PlacerAnnealRandom>();
             for (int i = 0; i < 3; i++) {
-                for (int j = 0; j < 4; j++) {
+                for (int j = 0; j < 5; j++) {
                     int initialTemp = 5000 + (i * 5000);
-                    Double coolingRate = 0.92 + (j * 0.02);
+                    Double coolingRate = 0.79 + (j * 0.05);
                     PlacerAnnealRandom placer = new PlacerAnnealRandom(rootDir, design, device, region);
                     SAPlacers.add(placer);
                     String fullName = placer.getPlacerName() + "_" + initialTemp + "_"

### 2025-07-22 python/overlay_convergences.py

@@ -18,18 +18,19 @@ def cooling_to_brightness(cooling_rate, min_rate=90, max_rate=98):
 
 
 def plot_group(temp_group, temp_to_color, df, output_path, title_suffix=""):
-    plt.figure(figsize=(6, 4))
+    plt.figure(figsize=(6, 5))
     for label in temp_group:
         temp, rate = parse_label(label)
         base_rgb = temp_to_color[temp]
-        alpha = 0.4 + 0.6 * (1 - cooling_to_brightness(rate))
+        alpha = 0.2 + 0.3 * (1 - cooling_to_brightness(rate))
         y = pd.to_numeric(df[label], errors='coerce').dropna().values
         x = range(len(y))
         plt.plot(x, y, color=base_rgb, label=label, alpha=alpha, linewidth=1.5)
 
-    plt.xlabel("Iteration")
+    plt.xlabel("Passes")
     plt.ylabel("Cost")
-    plt.ylim(300000, 1000000)
+    # plt.yscale("log")
+    plt.ylim(3e5, 1e6)
     plt.title(f"Cost History of SA Placers {title_suffix}")
     plt.legend(fontsize="small")
     plt.grid(True)
@@ -53,10 +54,12 @@ def main():
     temp_to_labels = {}
 
     # --- Combined plot ---
-    plt.figure(figsize=(6, 4))
+    plt.figure(figsize=(6, 5))
+    # plt.yscale("log")
+    plt.ylim(3e5, 1e6)
 
     for label in labels:
-        if label.strip().lower() in ("iteration", "unnamed: 0"):
+        if label.strip().lower() in ("passes", "unnamed: 0"):
             continue
 
         temp, rate = parse_label(label)
@@ -73,18 +76,18 @@ def main():
         temp_to_labels.setdefault(temp, []).append(label)
 
         base_rgb = temp_to_color[temp]
-        alpha = 0.4 + 0.6 * (1 - cooling_to_brightness(rate))
+        alpha = 0.2 + 0.3 * (1 - cooling_to_brightness(rate))
         y = pd.to_numeric(df[label], errors='coerce').dropna().values
         x = range(len(y))
         plt.plot(x, y, color=base_rgb, label=label, alpha=alpha, linewidth=1.5)
 
-    plt.xlabel("Iteration")
+    plt.xlabel("Passes")
     plt.ylabel("Cost")
     plt.title("Combined Cost History of all SA Placers")
-    plt.legend(fontsize="small")
[…]

### 2025-07-25 java/src/main/java/placer/Main.java
hybrid runs

@@ -97,13 +97,34 @@ public class Main {
             List<String> placerNames = new ArrayList<>();
             List<List<Double>> combinedCostHistory = new ArrayList<>();
 
-            List<PlacerAnnealRandom> SAPlacers = new ArrayList<PlacerAnnealRandom>();
-            for (int i = 0; i < 3; i++) {
+            List<PlacerAnnealRandom> PARPlacers = new ArrayList<PlacerAnnealRandom>();
+            for (int i = 0; i < 2; i++) {
                 for (int j = 0; j < 5; j++) {
-                    int initialTemp = 5000 + (i * 5000);
-                    Double coolingRate = 0.79 + (j * 0.05);
+                    int initialTemp = 10000 + (i * 10000);
+                    Double coolingRate = 0.84 + (j * 0.03);
                     PlacerAnnealRandom placer = new PlacerAnnealRandom(rootDir, design, device, region);
-                    SAPlacers.add(placer);
+                    PARPlacers.add(placer);
+                    String fullName = placer.getPlacerName() + "_" + initialTemp + "_"
+                            + (int) Math.round(coolingRate * 100);
+                    System.out.println("========================================");
+                    System.out.println("STARTING: " + fullName + " ... ");
+                    System.out.println("========================================");
+                    placer.makeOutputDirs(fullName);
+                    placer.initCoolingSchedule(initialTemp, coolingRate, 300);
+                    placer.run(packedDesign);
+
+                    placerNames.add(fullName);
+                    combinedCostHistory.add(placer.getCostHistory());
+                }
+            }
+
+            List<PlacerAnnealHybrid> PAHPlacers = new ArrayList<PlacerAnnealHybrid>();
+            for (int i = 0; i < 2; i++) {
+                for (int j = 0; j < 5; j++) {
+                    int initialTemp = 10000 + (i * 10000);
+                    Double coolingRate = 0.84 + (j * 0.03);
+                    PlacerAnnealHybrid placer = new PlacerAnnealHybrid(rootDir, design, device, region);
+                    PAHPlacers.add(placer);
                     String fullName = placer.getPlacerName() + "_" + initialTemp + "_"
                             + (int) Math.round(coolingRate * 100);
                     System.out.println("========================================");

### 2025-07-25 python/overlay_convergences.py

@@ -5,7 +5,8 @@ import re
 
 
 def parse_label(label):
-    match = re.match(r'PlacerAnnealRandom_(\d+)_(\d+)', label)
+    # match either "PlacerAnnealHybrid" or "PlacerAnnealRandom" followed by numbers
+    match = re.match(r'PlacerAnneal(?:Hybrid|Random)_(\d+)_(\d+)', label)
     if match:
         temp = int(match.group(1))
         rate = int(match.group(2))
@@ -13,7 +14,7 @@ def parse_label(label):
     return None, None
 
 
-def cooling_to_brightness(cooling_rate, min_rate=90, max_rate=98):
+def cooling_to_brightness(cooling_rate, min_rate=84, max_rate=99):
     return (cooling_rate - min_rate) / (max_rate - min_rate)
 
 
@@ -22,7 +23,7 @@ def plot_group(temp_group, temp_to_color, df, output_path, title_suffix=""):
     for label in temp_group:
         temp, rate = parse_label(label)
         base_rgb = temp_to_color[temp]
-        alpha = 0.2 + 0.3 * (1 - cooling_to_brightness(rate))
+        alpha = 0.2 + 0.6 * (1 - cooling_to_brightness(rate))
         y = pd.to_numeric(df[label], errors='coerce').dropna().values
         x = range(len(y))
         plt.plot(x, y, color=base_rgb, label=label, alpha=alpha, linewidth=1.5)
@@ -30,7 +31,7 @@ def plot_group(temp_group, temp_to_color, df, output_path, title_suffix=""):
     plt.xlabel("Passes")
     plt.ylabel("Cost")
     # plt.yscale("log")
-    plt.ylim(3e5, 1e6)
+    plt.ylim(2.5e5, 1e6)
     plt.title(f"Cost History of SA Placers {title_suffix}")
     plt.legend(fontsize="small")
     plt.grid(True)
@@ -56,7 +57,7 @@ def main():
     # --- Combined plot ---
     plt.figure(figsize=(6, 5))
     # plt.yscale("log")
-    plt.ylim(3e5, 1e6)
+    plt.ylim(2.5e5, 1e6)
 
     for label in labels:
         if label.strip().lower() in ("passes", "unnamed: 0"):
@@ -76,7 +77,7 @@ def main():
         temp_to_labels.setdefault(temp, []).append(label)
 
         base_rgb = temp_to_color[temp]
-        alpha = 0.2 + 0.3 * (1 - cooling_to_brightness(rate))
[…]

### 2025-07-25 run.sh

@@ -178,7 +178,7 @@ if [ "$start_stage" == "graphics" ] || [ "$start_stage" == "all" ]; then
 
     # Loop over all items in PLACERS_DIR
     for placer_dir in "$PLACERS_DIR"/*; do
-        python3 python/plot_convergence.py "$placer_dir"
+        # python3 python/plot_convergence.py "$placer_dir"
         # Only proceed if it's a directory
         if [ -d "$placer_dir" ]; then
             images_path="$placer_dir/graphics/images"

### 2025-07-28 latex/paper/sections/10-placement.tex
cooling schedules
start talking about ap

--- 2025-07-11 latex/paper/sections/10-placement.tex
@@
 We will refer to these distances as half perimeter wirelength (HPWL).
-The goal of SA is to bring the system from some initial state to a state with lower energy until some stopping criterion is met, whether that be when a computational budget is exceeded, an energy budget is met, or until the energy's rate of change approaches zero.
+The purpose of SA is to bring the system from some initial state to a state with lower energy until some stopping criterion is met, whether that be when a computational budget is exceeded, an energy budget is met, or until the energy's rate of change approaches zero.
 
@@
 
+\newcolumn
 \begin{lstlisting}[language=java, caption={Cooling Schedule and Move Acceptance}, label={lst:sa_acceptance}]
@@
 
-\subsection{Variations}
-\label{subsec:variations}
-
 In more sophisticated placers, the proposed movement can also be selected by finding the centroid between \texttt{SiteInst}s that share a net with the current object, or even a hybrid of random and centroid selection.
@@
 Object instance placements do not have to adhere to any hardware constraints and can even overlap with one another.
+
 Afterwards, detailed placement "legalizes" the global placement by discretizing the analog positions of the instances onto the discrete grid of resources of the \texttt{device}.
@@
 The spreading process somewhat resembles primitive fluid dynamics simulators or thermal diffusion simulators, where packets of energy are spread through a medium until each voxel or pixel of the medium falls below a maximum threshold of energy.
-There also exists other strategies to mix packing, placement, and legalization in different orders. 
+There also exists other strategies to mix packing, placement, and legalization in different orders, such as those studied in \cite{ExplicitPacking}. 
 In this context, our placer follows a very straightforward pack-legalize-place strategy with SA.

### 2025-07-28 latex/paper/sections/11-placement_results.tex
hybrid runs
start talking about ap

@@
 
-Shown in figure \ref{fig:placers_overlay} are the total HPWL curves over number of passes for each of the five placers. 
+Shown in figure \ref{fig:placers_overlay} are the HPWL curves for each of the five placers. 
 \texttt{PlacerGreedyMidpoint} (in red) appears to perform the worst by a considerable margin with a final HPWL cost of 590K.
@@
     For each initial temperature group, we test five different cooling rates \texttt{(0.79, 0.84, 0.89, 0.94, and 0.99)}, with the darkness of each curve corresponding with a faster cooling rate.
-    The curves suggest that a lower initial temperature and faster cooling rate both contribute to a faster convergence to the global minimum, with generally small differences of less than 10\% in final HPWL cost.
 

### 2025-07-28 latex/paper/sections/12-future_work.tex
start talking about ap

@@
 In its current state, the prepacker and packer struggle to handle signals larger than 24-bits, especially when involved in DSP functions like addition and multiplication. 
-In such designs, the Vivado synthesizer may synthesize long \texttt{CARRY4} chains with particular \texttt{EDIFHierPortInst} configurations that are currently not handled by our packer and eventually to failures in the subsequent routing stage.
+In such designs, the Vivado synthesizer may synthesize long \texttt{CARRY4} chains with particular \texttt{EDIFHierPortInst} configurations which are currently not handled by our packer and lead to failures in the subsequent routing stage.
 Further work is required to resolve these constraints in the packer.
 
+\subsection{Force-Directed and Analytical Placement}
+As mentioned previously, SA has largely been abandoned in favor of AP in SOTA placers, owing to SA's poor scalability and runtime.
+Our placer follows a basic prepack-pack-place flow, where the final placement stage is driven by simulated annealing. 
+In future work, we can retrofit this flow by replacing SA with AP in the final placement stage while fully reusing the prepacker and packer in its current state. 
+We can still treat \texttt{SiteInst}s as the atomic objects to be placed and use analytical solvers to minimize the total HPWL of the nets between them.
+
+The following papers can serve as foundational references for building an AP placer or insight into the SOTA placers.
+\begin{itemize}
+    \item \emph{Analytical minimization of half-permieter wirelength} - \textbf{Kennings and Markov (2000)} \cite{AP_2000}.
+    \item \emph{SimPL: an algorithm for placing VLSI circuits} - \textbf{Kim et al. (2013)} \cite{SimPL}.
+    \item \emph{Analytical placement for heterogeneous FPGAs} - \textbf{Gort et al. (2012)} \cite{AP_2012}.
+    \item \emph{Multi-Electrostatic FPGA Placement Considering SLICEL-SLICEM Heterogeneity, Clock Feasibility, and Timing Optimization} - \textbf{Jing et al. (2023)} \cite{MultiElectrostatic}.
+    \item \emph{OpenPARF 3.0: Robust Multi-Electrostatics Based FPGA Macro Placement Considering Cascaded Macros Groups and Fence Regions} - \textbf{Jing et al. (2024)} \cite{OpenPARF}.
+\end{itemize}
+
 \subsection{Variations on Packing}
-Our current placer follows a \texttt{Site}-centric approach that resembles that of Xilinx ISE, which is the predecessor to the Vivado design suite that was primarily used in the 90s and 2000s.
-The more modern Vivado now performs BEL-centric placement without necessarily locking \texttt{Cell}s into \texttt{Site}s, allowing for a higher granularity of movement of \texttt{Cell}s. 
-Enabling BEL-centric movement as opposed to Site-centric movement in our placer can improve HPWL minimization, but will add much more complexity to the packing and placement process to ensure robustness with respect to hardware constraints.
+Our current placer follows a \textbf{Site-centric} approach that resembles that of the old Xilinx ISE - the predecessor to Vivado used in the 90s and 2000s.
+These days, Vivado performs \textbf{BEL-centric} placement without necessarily locking \texttt{Cell}s into \texttt{Site}s, allowing for a higher granularity of movement of \texttt{Cell}s. 
+Enabling BEL-centric moveme
+nt as opposed to Site-centric movement can improve HPWL minimization, but will add much more complexity to the packing and placement process to ensure robustness with respect to hardware constraints.
 
-There is also no rule saying we must pack \texttt{Cells} into \texttt{SiteInst}s before placement or that we follow a strict prepacking-packing-placement flow. 
-Future work can 
+There is no constraint that forces us to pack \texttt{Cells} into \texttt{SiteInst}s before placement or to follow a strict prepacking-packing-placement flow. 
+After implementing a force-directed or analytical placer, we can begin to explore different packing and placement ordering like those studied in Wuxi et al. (2019) \cite{ExplicitPacking}.
+
+\begin{equation}
+    \boldsymbol{\Phi} (\vec{x}, \vec{y}) = \sum_{i, j} w_{i, j} ( |x_i - x_j| + |y_i - y_j| )
+\end{equation}
+
 {
@@
 \subsection{Add Hard Macro Support}
-
-\subsection{Improvements to SA}
-
-\subsection{Force-Directed and Analytical Placement}

### 2025-08-08 latex/paper/sections/04-architecture_history.tex
more packing
start talking about ap
more ap

@@
 The mid-1980s saw the introduction of homogeneous FPGAs, which were built as a grid of CLBs. 
-Rather than using a central programmable switch matrix as in CPLDs, FPGAs adopted an island style architecture in which each CLB is surrounded on all sides by programmable routing resources, as shown in Figure \ref{fig:fpga}.
+Rather than using a central programmable switch matrix as in CPLDs, FPGAs adopted an island style architecture in which each CLB is surrounded on all sides by programmable routing resources, as shown in Figure \ref{fig:homogen_fpga}.
 The first commercially viable FPGA, produced by Xilinx in 1984, featured 16 CLBs arranged in a 4x4 grid. 
@@
     \captionof{figure}{A homogeneous island-style FPGA architecture with 16 CLBs in a grid.}
-    \label{fig:fpga}
+    \label{fig:homogen_fpga}
     % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/FPGA/Field-programmable-gate-array-FPGA/
@@
     \captionof{figure}{A heterogeneous island-style FPGA with a mix of CLBs and macrocells.}
-    \label{fig:fpga}
+    \label{fig:heterogen_fpga}
     % https://www.electronics-tutorial.net/Programmable-Logic-Device-Architectures/FPGA/Field-programmable-gate-array-FPGA/
 }
-\vspace{0.5cm}
+\vspace{0.25cm}
 
@@
 To meet the needs of increasingly complex designs, FPGA vendors introduced heterogeneous FPGAs. 
-In these devices, hard macros such as Block RAM (BRAM) and Digital Signal Processing (DSP) slices are integrated into the programmable logic fabric along with CLBs, like shown in Figure \ref{fig:fpga}.
+In these devices, hard macros such as Block RAM (BRAM) and Digital Signal Processing (DSP) slices are integrated into the programmable logic fabric along with CLBs, like shown in Figure \ref{fig:heterogen_fpga}.
 This design enables the direct instantiation of common subsystems like memories and multipliers, without having to recreate them from scratch using CLBs. 
-Major vendors such as Xilinx and Altera now employ heterogeneous island-style architectures in their devices. 
+Major vendors such as Xilinx (AMD) and Altera (Intel) now employ heterogeneous island-style architectures in their devices. 
 As designs become increasingly large and complex, FPGAs meet the demand by becoming increasingly heterogenous, incorporating a wider variety of hard macros into the fabric.

### 2025-08-08 latex/paper/sections/09-packing.tex
more reorg
more ap

@@
 Listing \ref{lst:siteinst} shows how to use the \texttt{SiteInst} object and how it ties \texttt{Site}s, \texttt{Cell}s, \texttt{BEL}s, and intra-Site \texttt{Net}s together. 
-The super-specifics of the packing process via RapidWright is too involved for the scope of this paper, but just to give an impression of how the \texttt{design} classes are used, Listing \ref{lst:single_carry_chains}, \ref{lst:carry_chains}, \ref{lst:carry_nets}, and \ref{lst:ff_nets} together show how \texttt{CarryCellGroup}s are packed into \texttt{SLICEL} \texttt{Sites} in our working code.
+The super-specifics of the packing process via RapidWright is too involved for the scope of this paper, but just to give an impression of how the \texttt{design} classes are used, Listing \ref{lst:single_carry_chains}, \ref{lst:carry_cell_group_chains}, \ref{lst:carry_nets}, and \ref{lst:ff_nets} together show how \texttt{CarryCellGroup}s are packed into \texttt{SLICEL} \texttt{Sites} in our working code.
 
@@
 The end result is a \texttt{PackedDesign} object which is a group of data structures solely around \texttt{SiteInst}s like shown in figure \ref{fig:substages}.
+The process is analogous to containerization of loose goods in ocean freight forwarding. 
 At this point we no longer need to worry about individual \texttt{BEL}s, \texttt{Cell}s, or intra-Site \texttt{Nets} as they have now been packaged and taped-off into their respective \texttt{SiteInst}s.
 These \texttt{SiteInst} become the new atomic level components in the following Placement stage.
-This process is analogous to the containerization of loose goods in ocean freight forwarding.
 
@@
 \newpage
-\begin{lstlisting}[language=java, caption={Packing \texttt{CarryCellGroup}s into \texttt{SLICEL} \texttt{SiteInst}s.}, label={lst:carry_chains}]
+\begin{lstlisting}[language=java, caption={Packing \texttt{CarryCellGroup}s into \texttt{SLICEL} \texttt{SiteInst}s.}, label={lst:carry_cell_group_chains}]
 // Pack all CarryCellGroup chains

### 2025-08-08 latex/paper/sections/11-placement_results.tex
more ap

@@
 
-Also note the sparsity of the \texttt{DSP48E1} (yellow) and \texttt{RAMB18E1} (magenta) \texttt{Site} columns compared to the more uniform grid of \texttt{SLICEL}s (cyan), which poses a computational challenge for the legalization process in directed placement.
+Also note the sparsity of the \texttt{DSP48E1} (yellow) and \texttt{RAMB18E1} (magenta) \texttt{Site} columns compared to the more uniform grid of \texttt{SLICEL}s (cyan), which poses a computational challenge for the legalization process in directed and analytical placement.
 

### 2025-08-08 latex/paper/sections/12-future_work.tex
more ap

@@
 In this paper we have made many simplifications to the problem space just to make the placer easier to implement.
-For example, our placer in its current state does not take advantage of \texttt{SLICEL}/\texttt{SLICEM} homogeneity and simply maps all SLICE \texttt{SiteInst}s onto \texttt{SLICEL}s. 
+For example, our placer in its current state does not take advantage of \texttt{SLICEL}/\texttt{SLICEM} heterogeneity and simply maps all SLICE \texttt{SiteInst}s onto \texttt{SLICEL}s. 
 Recall that the SLICE Sites in Xilinx FPGAs typically come in a 75-25\% split between \texttt{SLICEL}s and \texttt{SLICEM}s. 
 This means that we have rendered about 25\% of the CLB fabric unusable which will inevitably hurt wirelength minimization during placement since the \texttt{SiteInst}s must be spread over a larger area.
-Enabling \texttt{SLICEL}-\texttt{SLICEM} homogeneity can lead do greater logic density and consequently less total HPWL, but can make the packing process more complex and may contribute to higher routing congestion.
+Enabling \texttt{SLICEL}-\texttt{SLICEM} heterogeneity can lead do greater logic density and consequently less total HPWL, but can make the packing process more complex and may contribute to higher routing congestion.
 
@@
 \subsection{Force-Directed and Analytical Placement}
-As mentioned previously, SA has largely been abandoned in favor of AP in SOTA placers, owing to SA's poor scalability and runtime.
-Our placer follows a basic prepack-pack-place flow, where the final placement stage is driven by simulated annealing. 
-In future work, we can retrofit this flow by replacing SA with AP in the final placement stage while fully reusing the prepacker and packer in its current state. 
-We can still treat \texttt{SiteInst}s as the atomic objects to be placed and use analytical solvers to minimize the total HPWL of the nets between them.
+As discussed earlier, simulated annealing (SA) has largely been phased out in state-of-the-art (SOTA) placers due to its poor scalability and long runtimes. 
+Our current placer follows a straightforward \emph{prepack–pack–place} flow, with the final placement stage driven by SA. 
+In future work, this stage can be retrofitted to use analytical placement (AP) while reusing the prepacker and packer. 
+Here, \texttt{SiteInst}s remain the atomic placement objects, and analytical solvers can be applied to minimize the total half-perimeter wirelength (HPWL) of the connecting nets.
 
-The following papers can serve as foundational references for building an AP placer or insight into the SOTA placers.
+The following works provide a foundation for understanding AP, ranging from early formulations to modern SOTA methods:
 \begin{itemize}
-    \item \emph{Analytical minimization of half-permieter wirelength} - \textbf{Kennings and Markov (2000)} \cite{AP_2000}.
-    \item \emph{SimPL: an algorithm for placing VLSI circuits} - \textbf{Kim et al. (2013)} \cite{SimPL}.
-    \item \emph{Analytical placement for heterogeneous FPGAs} - \textbf{Gort et al. (2012)} \cite{AP_2012}.
-    \item \emph{Multi-Electrostatic FPGA Placement Considering SLICEL-SLICEM Heterogeneity, Clock Feasibility, and Timing Optimization} - \textbf{Jing et al. (2023)} \cite{MultiElectrostatic}.
-    \item \emph{OpenPARF 3.0: Robust Multi-Electrostatics Based FPGA Macro Placement Considering Cascaded Macros Groups and Fence Regions} - \textbf{Jing et al. (2024)} \cite{OpenPARF}.
+    \item \emph{Analytical minimization of half-perimeter wirelength} — \textbf{Kennings and Markov (2000)} \cite{AP_2000}.
+    \item \emph{Analytical placement for heterogeneous FPGAs} — \textbf{Gort et al. (2012)} \cite{AP_2012}.
+    \item \emph{SimPL: An algorithm for placing VLSI circuits} — \textbf{Kim et al. (2013)} \cite{SimPL}.
+    \item \emph{Multi-Electrostatic FPGA Placement Considering SLICEL–SLICEM Heterogeneity, Clock Feasibility, and Timing Optimization} — \textbf{Jing et al. (2023)} \cite{MultiElectrostatic}.
+    \item \emph{OpenPARF 3.0: Robust Multi-Electrostatics Based FPGA Macro Placement Considering Cascaded Macro Groups and Fence Regions} — \textbf{Jing et al. (2024)} \cite{OpenPARF}.
 \end{itemize}
+
+As noted in Section~\ref{sec:placement}, replacing SA with AP requires splitting the placement stage into two substages:  
+(1) \emph{global placement}, which determines continuous target positions using analytical optimization, and  
+(2) \emph{detailed placement} (legalization), which snaps objects to valid \texttt{Site}s while adhering to hardware constraints.  
+Global placement can be performed using any number of off-the-shelf (OTS) analytical solvers.
+
+Unlike SA, which relies primarily on heuristics and probabilistic moves, AP begins with a formal mathematical formulation that is compatible with analytic optimization. 
+As introduced in Section~\ref{subsec:netlist}, a circuit netlist can be naturally modeled as a \emph{hypergraph}
+\[
+    G_{H} = (V_{H}, E_{H}),
+\]
+where \(V_{H}\) is the set of vertices and \(E_{H}\) is the set of hyperedges, each of which may connect more than two vertices. 
+In the FPGA context, \(V_{H}\) corresponds to the set of \texttt{SiteInst}s placed on the \texttt{device}, while \(E_{H}\) corresponds to the set of \texttt{Net}s interconnecting them.
+
+We can then reduce this hypergraph into a graph by converting all multi-pin hyperedges into a set of 2-pin edges.
+
+Most AP approaches then minimize the weighted sum of half perimeter wirelengths (HPWL) of these 2-pin edges, with some approaches using the sum of linear distances \ref{equ:Manhattan} and others using the sum of squared distances \ref{equ:Euclidian}.
+
+\begin{equation}
+    \boldsymbol{\Phi} (\vec{x}, \vec{y}) = \sum_{i,j} w_{i,j} \left( |x_i - x_j| + |y_i - y_j| \right)
+    \label{equ:Manhattan}
+\end{equation}
+
+\begin{equation}
+    \boldsymbol{\Phi} (\vec{x}, \vec{y}) = \sum_{i,j} w_{i,j} \left[ (x_i - x_j)^2 + (y_i - y_j)^2 \right]
+    \label{equ:Euclidian}
+\end{equation}
+
+% use these formulations
+% HPWL_AP_2000
+% HPWL_AP_2012
 
@@
 
-\begin{equation}
-    \boldsymbol{\Phi} (\vec{x}, \vec{y}) = \sum_{i, j} w_{i, j} ( |x_i - x_j| + |y_i - y_j| )
-\end{equation}
-
 {

### 2025-08-18 latex/paper/sections/12-future_work.tex
more ap

@@
 \begin{itemize}
-    \item \emph{Analytical minimization of half-perimeter wirelength} — \textbf{Kennings and Markov (2000)} \cite{AP_2000}.
-    \item \emph{Analytical placement for heterogeneous FPGAs} — \textbf{Gort et al. (2012)} \cite{AP_2012}.
-    \item \emph{SimPL: An algorithm for placing VLSI circuits} — \textbf{Kim et al. (2013)} \cite{SimPL}.
-    \item \emph{Multi-Electrostatic FPGA Placement Considering SLICEL–SLICEM Heterogeneity, Clock Feasibility, and Timing Optimization} — \textbf{Jing et al. (2023)} \cite{MultiElectrostatic}.
-    \item \emph{OpenPARF 3.0: Robust Multi-Electrostatics Based FPGA Macro Placement Considering Cascaded Macro Groups and Fence Regions} — \textbf{Jing et al. (2024)} \cite{OpenPARF}.
+    \item \emph{Analytical minimization of half-perimeter wirelength} - \textbf{Kennings and Markov (2000)} \cite{AP_2000}.
+    \item \emph{Kraftwerk2—A Fast Force-Directed Quadratic Placement Approach Using an Accurate Net Model} - \textbf{Spindler et al. (2008)} \cite{kraftwerk2}
+    \item \emph{Analytical placement for heterogeneous FPGAs} - \textbf{Gort et al. (2012)} \cite{AP_2012}.
+    \item \emph{SimPL: An algorithm for placing VLSI circuits} - \textbf{Kim et al. (2013)} \cite{SimPL}.
+    \item \emph{Multi-Electrostatic FPGA Placement Considering SLICEL–SLICEM Heterogeneity, Clock Feasibility, and Timing Optimization} - \textbf{Jing et al. (2023)} \cite{MultiElectrostatic}.
+    \item \emph{OpenPARF 3.0: Robust Multi-Electrostatics Based FPGA Macro Placement Considering Cascaded Macro Groups and Fence Regions} - \textbf{Jing et al. (2024)} \cite{OpenPARF}.
 \end{itemize}
@@
 
-Unlike SA, which relies primarily on heuristics and probabilistic moves, AP begins with a formal mathematical formulation that is compatible with analytic optimization. 
-As introduced in Section~\ref{subsec:netlist}, a circuit netlist can be naturally modeled as a \emph{hypergraph}
-\[
-    G_{H} = (V_{H}, E_{H}),
-\]
+Unlike SA, which primarily relies on heuristics and probabilistic moves, AP begins by contextualizing the problem as a collection of math expressions that are compatible with OTS solvers.
+As briefly introduced in Section~\ref{subsec:netlist}, a circuit netlist can be naturally modeled as a weighted hypergraph
+\begin{equation}
+    G_{H} = (V_{H}, E_{H})
+    \label{equ:hypergraph}
+\end{equation}
 where \(V_{H}\) is the set of vertices and \(E_{H}\) is the set of hyperedges, each of which may connect more than two vertices. 
-In the FPGA context, \(V_{H}\) corresponds to the set of \texttt{SiteInst}s placed on the \texttt{device}, while \(E_{H}\) corresponds to the set of \texttt{Net}s interconnecting them.
+In FPGA context, \(V_{H}\) corresponds to the set of \texttt{SiteInst}s placed on the \texttt{device}, while \(E_{H}\) corresponds to the set of \texttt{Net}s connecting them.
+Each hyperedge is incident to one voltage source and one or many voltage sinks, defining the directionality of each hyperedge.
+We can then reduce this hypergraph into a graph using one of various net models - star net model, clique net model, Bound2Bound net model, among others.
 
-We can then reduce this hypergraph into a graph by converting all multi-pin hyperedges into a set of 2-pin edges.
+The star model transforms the hyperedge into a set of 2-pin edges connecting the incident sink pins to the source pin.
+The clique net model transforms the hyperdedge into a local fully connected network of 2-pin edges.
+We could preserve the directionality of the edges for possible future optimizations, but since HPWL minimization is not directly affected by the directionality of nets, we can just reduce the directed hypergraph into an undirected graph.
+Figure \ref{fig:net_model} shows an example of a hyperedge reduction using the star model and clique model.
 
-Most AP approaches then minimize the weighted sum of half perimeter wirelengths (HPWL) of these 2-pin edges, with some approaches using the sum of linear distances \ref{equ:Manhattan} and others using the sum of squared distances \ref{equ:Euclidian}.
+\vspace{1.0cm}
+{
+    \centering
+    \includegraphics[width=\columnwidth]{figures/future_work/net_model.png}
+    \captionof{figure}{Hypergraph reduction via Star and Clique net models.}
+    \label{fig:net_model}
+}
+
+After the hypergraph reduction, most AP approaches then minimize a HPWL cost function using either the sum of linear distances \ref{equ:Manhattan} or the sum of squared distances \ref{equ:Euclidian}.
 
 \begin{equation}
-    \boldsymbol{\Phi} (\vec{x}, \vec{y}) = \sum_{i,j} w_{i,j} \left( |x_i - x_j| + |y_i - y_j| \right)
+    \boldsymbol{\Phi} (\boldsymbol{x}, \boldsymbol{y}) = \sum_{i,j} w_{i,j} \left( |x_i - x_j| + |y_i - y_j| \right)
     \label{equ:Manhattan}
@@
 \begin{equation}
-    \boldsymbol{\Phi} (\vec{x}, \vec{y}) = \sum_{i,j} w_{i,j} \left[ (x_i - x_j)^2 + (y_i - y_j)^2 \right]
+    \boldsymbol{\Phi} (\boldsymbol{x}, \boldsymbol{y}) = \sum_{i,j} w_{i,j} \left[ (x_i - x_j)^2 + (y_i - y_j)^2 \right]
     \label{equ:Euclidian}
+\end{equation}
+
+Here, module placements in the \(x\) and \(y\) directions are captured by the placement vectors \( \boldsymbol{x} = (x_1, x_2, ..., x_n) \) and \( \boldsymbol{y} = (y_1, y_2, ..., y_n) \).
+Each 2-pin edge has a weight \( w_{i,j} \) and is calculated depending on the net model used.
+
+\begin{equation}
+    w_{i,j} = \frac{1}{|x_i - x_j|}
+    \label{equ:weight}
+\end{equation}
+
+\begin{equation}
+    w_{i,j} = \frac{1}{(p-1) |x_i - x_j|}
+    \label{equ:weight_linearized}
+\end{equation}
+
+\begin{equation}
+    \boldsymbol{\Phi} (\boldsymbol{x}) = \frac{1}{2} \boldsymbol{x^T} \boldsymbol{Q_x} \boldsymbol{x} + \boldsymbol{c_x^T} \boldsymbol{x} + const.
+    \label{equ:quadratic}
 \end{equation}
@@
 \subsection{Variations on Packing}
-Our current placer follows a \textbf{Site-centric} approach that resembles that of the old Xilinx ISE - the predecessor to Vivado used in the 90s and 2000s.
+Our current placer follows a \textbf{Site-centric} approach that resembles that of the old Xilinx ISE - the predecessor to Vivado used in the 2000s.
 These days, Vivado performs \textbf{BEL-centric} placement without necessarily locking \texttt{Cell}s into \texttt{Site}s, allowing for a higher granularity of movement of \texttt{Cell}s. 

### 2025-08-19 latex/paper/sections/12-future_work.tex
need to review AP_2012

@@
 As noted in Section~\ref{sec:placement}, replacing SA with AP requires splitting the placement stage into two substages:  
-(1) \emph{global placement}, which determines continuous target positions using analytical optimization, and  
-(2) \emph{detailed placement} (legalization), which snaps objects to valid \texttt{Site}s while adhering to hardware constraints.  
+\emph{global placement}, which determines continuous target positions using analytical optimization, and  
+\emph{detailed placement} (legalization), which snaps objects to valid \texttt{site}s while adhering to hardware constraints.  
 Global placement can be performed using any number of off-the-shelf (OTS) analytical solvers.
@@
 The star model transforms the hyperedge into a set of 2-pin edges connecting the incident sink pins to the source pin.
-The clique net model transforms the hyperdedge into a local fully connected network of 2-pin edges.
+The clique net model transforms the hyperdedge into a local fully connected network of 2-pin edges between the vertices incident to the original hyperedge.
+This can lead to higher quality results but requires a quadratic growth in edges.
 We could preserve the directionality of the edges for possible future optimizations, but since HPWL minimization is not directly affected by the directionality of nets, we can just reduce the directed hypergraph into an undirected graph.
 Figure \ref{fig:net_model} shows an example of a hyperedge reduction using the star model and clique model.
+The weight of each edge corresponds inversely to the wirelength of the net it represents.
 
@@
 Here, module placements in the \(x\) and \(y\) directions are captured by the placement vectors \( \boldsymbol{x} = (x_1, x_2, ..., x_n) \) and \( \boldsymbol{y} = (y_1, y_2, ..., y_n) \).
-Each 2-pin edge has a weight \( w_{i,j} \) and is calculated depending on the net model used.
+The weight can be calculated depending on the net model used.
+Placers using the star net model typically assign the weight as the inverse of the HPWL of the net, as shown in (\ref{equ:weight}).
+Placers using the clique model or the Bound2Bound (B2B) model include a \(p-1\) term in the denominator as shown in (\ref{equ:weight_linearized}), where \(p\) is the number of pins on the original hyperedge, to de-emphasize high-fanout nets like the clock and reset nets.
 
@@
 \begin{equation}
-    \boldsymbol{\Phi} (\boldsymbol{x}) = \frac{1}{2} \boldsymbol{x^T} \boldsymbol{Q_x} \boldsymbol{x} + \boldsymbol{c_x^T} \boldsymbol{x} + const.
+    \boldsymbol{\Phi} (\boldsymbol{x}) = \frac{1}{2} \boldsymbol{x}^T \boldsymbol{Q_x} \boldsymbol{x} + \boldsymbol{c_x}^T \boldsymbol{x} + const.
     \label{equ:quadratic}
 \end{equation}
-
-% use these formulations
-% HPWL_AP_2000
-% HPWL_AP_2012
 
@@
 Enabling BEL-centric moveme
-nt as opposed to Site-centric movement can improve HPWL minimization, but will add much more complexity to the packing and placement process to ensure robustness with respect to hardware constraints.
+nt as opposed to Site-centric movement can improve HPWL minimization, but will add much more complexity to the packing and placement process to ensure robustness with respect to hardware constraints, particularly with intra-\texttt{Site} routing.
 
@@
 \subsection{Add Hard Macro Support}
+fdsa
+
+\subsection{Parallelization}
+fdsa

### 2025-08-20 latex/paper/sections/12-future_work.tex
review kennings and simpl

--- 2025-08-08 latex/paper/sections/12-future_work.tex
@@
 \subsection{Force-Directed and Analytical Placement}
-As discussed earlier, simulated annealing (SA) has largely been phased out in state-of-the-art (SOTA) placers due to its poor scalability and long runtimes. 
+As discussed earlier, simulated annealing (SA) has largely been phased out in state-of-the-art (SOTA) placers due to poor scalability and long runtimes. 
 Our current placer follows a straightforward \emph{prepack–pack–place} flow, with the final placement stage driven by SA. 
@@
     \item \emph{Analytical minimization of half-perimeter wirelength} — \textbf{Kennings and Markov (2000)} \cite{AP_2000}.
+    \item \emph{Kraftwerk2—A Fast Force-Directed Quadratic Placement Approach Using an Accurate Net Model} — \textbf{Spindler et al. (2008)} \cite{kraftwerk2}.
     \item \emph{Analytical placement for heterogeneous FPGAs} — \textbf{Gort et al. (2012)} \cite{AP_2012}.
@@
 As noted in Section~\ref{sec:placement}, replacing SA with AP requires splitting the placement stage into two substages:  
-(1) \emph{global placement}, which determines continuous target positions using analytical optimization, and  
-(2) \emph{detailed placement} (legalization), which snaps objects to valid \texttt{Site}s while adhering to hardware constraints.  
+\emph{global placement}, which determines continuous target positions using analytical optimization, and  
+\emph{detailed placement} (legalization), which snaps objects to valid \texttt{site}s while adhering to hardware constraints.  
 Global placement can be performed using any number of off-the-shelf (OTS) analytical solvers.
 
-Unlike SA, which relies primarily on heuristics and probabilistic moves, AP begins with a formal mathematical formulation that is compatible with analytic optimization. 
-As introduced in Section~\ref{subsec:netlist}, a circuit netlist can be naturally modeled as a \emph{hypergraph}
-\[
-    G_{H} = (V_{H}, E_{H}),
-\]
+Unlike SA, which primarily relies on heuristics and probabilistic moves, AP begins by contextualizing the problem as a collection of mathematical expressions compatible with convex optimization solvers.
+As briefly introduced in Section~\ref{subsec:netlist}, a circuit netlist can be naturally modeled as a hypergraph
+\begin{equation}
+    G_{H} = (V_{H}, E_{H})
+    \label{equ:hypergraph}
+\end{equation}
 where \(V_{H}\) is the set of vertices and \(E_{H}\) is the set of hyperedges, each of which may connect more than two vertices. 
-In the FPGA context, \(V_{H}\) corresponds to the set of \texttt{SiteInst}s placed on the \texttt{device}, while \(E_{H}\) corresponds to the set of \texttt{Net}s interconnecting them.
+In FPGA context, \(V_{H}\) corresponds to the set of \texttt{SiteInst}s placed on the \texttt{device}, while \(E_{H}\) corresponds to the set of \texttt{Net}s connecting them.
 
-We can then reduce this hypergraph into a graph by converting all multi-pin hyperedges into a set of 2-pin edges.
+This hypergraph is reduced into a graph using one of several net models:
+\begin{itemize}
+    \item \textbf{Star model:} classically, introduce a virtual star node and connect it to all pins of the net.
+However, in electronics, we can simply use the voltage source node as the star node and avoid adding virtual nodes.
+    \item \textbf{Clique model:} connect all pins of the net to each other (quadratic growth in edges). 
+    \item \textbf{Bound2Bound (B2B) model:} connect only the pins at the geometric boundary (min/max in $x$ or $y$) to each other, and connect boundary pins to all interior pins.
+With inverse-distance weights, this model matches HPWL exactly at the point of weight calculation.
+\end{itemize}
 
-Most AP approaches then minimize the weighted sum of half perimeter wirelengths (HPWL) of these 2-pin edges, with some approaches using the sum of linear distances \ref{equ:Manhattan} and others using the sum of squared distances \ref{equ:Euclidian}.
+Since HPWL is direction-agnostic, directionality of FPGA nets (source/sinks) is typically ignored in reduction.
+Figure \ref{fig:net_model} shows an example of hyperedge reduction using the star and clique models.
+
+\vspace{1.0cm}
+{
+    \centering
+    \includegraphics[width=\columnwidth]{figures/future_work/net_model.png}
+    \captionof{figure}{Hypergraph reduction via Star and Clique net models.}
+    \label{fig:net_model}
+}
+
+After reduction, AP minimizes a wirelength objective, typically either a linear HPWL formulation \eqref{equ:Manhattan} or a surrogate quadratic formulation \eqref{equ:Euclidian}:
 
 \begin{equation}
-    \boldsymbol{\Phi} (\vec{x}, \vec{y}) = \sum_{i,j} w_{i,j} \left( |x_i - x_j| + |y_i - y_j| \right)
+    \boldsymbol{\Phi} (\boldsymbol{x}, \boldsymbol{y}) = \sum_{i,j} w_{i,j} \left( |x_i - x_j| + |y_i - y_j| \right)
     \label{equ:Manhattan}
@@
 \begin{equation}
-    \boldsymbol{\Phi} (\vec{x}, \vec{y}) = \sum_{i,j} w_{i,j} \left[ (x_i - x_j)^2 + (y_i - y_j)^2 \right]
+    \boldsymbol{\Phi} (\boldsymbol{x}, \boldsymbol{y}) = \sum_{i,j} w_{i,j} \left[ (x_i - x_j)^2 + (y_i - y_j)^2 \right]
     \label{equ:Euclidian}
@@
 
-% use these formulations
-% HPWL_AP_2000
-% HPWL_AP_2012
+Here, module placements in the $x$ and $y$ directions are captured by the placement vectors \( \boldsymbol{x} = (x_1, x_2, ..., x_n) \) and \( \boldsymbol{y} = (y_1, y_2, ..., y_n) \).
+In both formulations, the objective function is separable with $x$ and $y$.
+The weights \(w_{i,j}\) depend on both the net model and the objective:
+
+\begin{itemize}
+    \item \textbf{Star/Clique models + Quadratic objective:}  
+    Assign constant weights proportional to $1$ or $1/(p-1)$ (where $p$ is the pin count) to prevent overemphasis of high-fanout nets.
+
+    \item \textbf{Star/Clique models + Linear HPWL:}  
+    Require iterative reweighting (e.g., $w_{i,j}=1/|x_i-x_j|$) so that the quadratic surrogate approximates the absolute value.
+
+    \item \textbf{Bound2Bound + Linear HPWL:}  
+    Assign weights as inverse distances, e.g.
+    \begin{equation}
+        w_{i,j} = \frac{1}{(p-1)\,|x_i - x_j|}
+        \label{equ:weight_linearized}
+    \end{equation}
+    This makes the quadratic surrogate equal the HPWL for each net at the evaluation point.
+\end{itemize}
+
+The squared wirelength objective yields a unique minimizer and requires a single linear solve per iteration. 
+However, it tends to over-emphasize long wires relative to short ones and tends to worsen routing demand compared to linear metrics \cite{AP_2000}. 
+The linear wirelength objective more closely reflects how HPWL is measured and better correlates with routability. 
+However, it is non-differentiable, so analytical placers typically handle it by iterated quadratic approximations (e.g., GORDIAN-L reweights edge coefficients each iteration) or by smooth regularizations that enable Newton-type solvers.
+
+For ease of implementation, it would be best to first use the squared objective before exploring the linear objective.
+We can use HeAP \cite{AP_2012}, which uses the squared objective along with the B2B net model, as an instructive reference.
+
+We can cast the squared formulation into quadratic form, which can be written compactly as
+\begin{equation}
+    \boldsymbol{\Phi} (\boldsymbol{x}) 
+    = \tfrac{1}{2}\,\boldsymbol{x}^T \boldsymbol{Q_x}\,\boldsymbol{x} + \boldsymbol{c_x}^T \boldsymbol{x} + \text{const},
+    \label{equ:quadratic}
+\end{equation}
+where $\boldsymbol{Q_x}$ encodes 2-pin connections between movable cells and $\boldsymbol{c_x}$ accounts for fixed terminals.  
+Solving $\boldsymbol{Q_x}\boldsymbol{x} = -\boldsymbol{c_x}$ yields the optimal continuous $x$ coordinates, and the same applies independently for $y$.
 
 \subsection{Variations on Packing}
-Our current placer follows a \textbf{Site-centric} approach that resembles that of the old Xilinx ISE - the predecessor to Vivado used in the 90s and 2000s.
+Our current placer follows a \textbf{Site-centric} approach that resembles that of the old Xilinx ISE - the predecessor to Vivado used in the 2000s.
 These days, Vivado performs \textbf{BEL-centric} placement without necessarily locking \texttt{Cell}s into \texttt{Site}s, allowing for a higher granularity of movement of \texttt{Cell}s. 
 Enabling BEL-centric moveme
-nt as opposed to Site-centric movement can improve HPWL minimization, but will add much more complexity to the packing and placement process to ensure robustness with respect to hardware constraints.
+nt as opposed to Site-centric movement can improve HPWL minimization, but will add much more complexity to the packing and placement process to ensure robustness with respect to hardware constraints, particularly with intra-\texttt{Site} routing.
 

### 2025-08-27 latex/paper/sections/12-future_work.tex
bound2bound diagram
weights explanation

@@
 
-\subsection{Force-Directed and Analytical Placement}
+\subsection{Analytical Placement}
 As discussed earlier, simulated annealing (SA) has largely been phased out in state-of-the-art (SOTA) placers due to poor scalability and long runtimes. 
@@
 
-This hypergraph is reduced into a graph using one of several net models:
+This hypergraph can be reduced into a graph using one of several net models:
 \begin{itemize}
@@
     \includegraphics[width=\columnwidth]{figures/future_work/net_model.png}
-    \captionof{figure}{Hypergraph reduction via Star and Clique net models.}
+    \captionof{figure}{Hyperedge reduction via Star and Clique net models.}
     \label{fig:net_model}
+}
+{
+    \centering
+    \includegraphics[width=\columnwidth]{figures/future_work/bound2bound.png}
+    \captionof{figure}{Hyperedge reduction via Bound2Bound net model.}
+    \label{fig:bound2bound}
 }
@@
 
-Here, module placements in the $x$ and $y$ directions are captured by the placement vectors \( \boldsymbol{x} = (x_1, x_2, ..., x_n) \) and \( \boldsymbol{y} = (y_1, y_2, ..., y_n) \).
+Here, module placements in the $x$ and $y$ directions are captured by the placement vectors \( \boldsymbol{x} = (x_1, x_2, ..., x_n) \) and \( \boldsymbol{y} = (y_1, y_2, ..., y_n) \), where $x_n$ and $y_n$ correspond to the coordinates of the $n$th module instance.
 In both formulations, the objective function is separable with $x$ and $y$.
-The weights \(w_{i,j}\) depend on both the net model and the objective:
+The weight \(w_{i,j}\) is the weight of the 2-pin edge connecting nodes $i$ and $j$. 
+Its value depends on both the net model and the objective:
 
 \begin{itemize}
-    \item \textbf{Star/Clique models + Quadratic objective:}  
-    Assign constant weights proportional to $1$ or $1/(p-1)$ (where $p$ is the pin count) to prevent overemphasis of high-fanout nets.
+    \item \textbf{Star/Clique models with a quadratic (squared-distance) objective:}  
+    Use constant edge weights. In the star model, each reduced 2-pin edge is given unit weight $w_{ij}=1$.  
+    In the clique model, each pin–pin edge is given weight $w_{ij}=1/(p-1)$, where $p$ is the pin count of the original hyperedge from which this edge was produced.  
+    This scaling prevents large-fanout nets from dominating the cost and makes star and clique models equivalent in the quadratic formulation.
 
-    \item \textbf{Star/Clique models + Linear HPWL:}  
-    Require iterative reweighting (e.g., $w_{i,j}=1/|x_i-x_j|$) so that the quadratic surrogate approximates the absolute value.
+    \item \textbf{Star/Clique models with a linear HPWL objective (via quadratic surrogate):}  
+    Use iterative reweighting. With the star model, each edge weight is updated as $w_{ij}=1/|x_i-x_j|$. With the clique model, the weight is $1/((p-1)|x_i-x_j|)$.  
+    This reweighting makes the quadratic form act as a surrogate for $|x_i-x_j|$, approximating the true HPWL.
 
-    \item \textbf{Bound2Bound + Linear HPWL:}  
-    Assign weights as inverse distances, e.g.
+    \item \textbf{Bound2Bound model with a linear HPWL objective:}  
+    Connect only boundary–boundary and boundary–interior pin pairs, and assign weights as inverse distances:  
     \begin{equation}
-        w_{i,j} = \frac{1}{(p-1)\,|x_i - x_j|}
+        w_{i,j} = \frac{2}{(p-1)\,|x_i - x_j|}.
         \label{equ:weight_linearized}
     \end{equation}
-    This makes the quadratic surrogate equal the HPWL for each net at the evaluation point.
+    With this assignment, the quadratic cost of a net equals its HPWL (in each dimension) at the evaluation point, making the Bound2Bound model HPWL-accurate.
 \end{itemize}
@@
 The linear wirelength objective more closely reflects how HPWL is measured and better correlates with routability. 
-However, it is non-differentiable, so analytical placers typically handle it by iterated quadratic approximations (e.g., GORDIAN-L reweights edge coefficients each iteration) or by smooth regularizations that enable Newton-type solvers.
+However, it is non-differentiable, so analytical placers typically handle it by iterated quadratic approximations (e.g., GORDIAN-L reweights edge coefficients each iteration) or by smooth regularizations that enable Newton-type solvers \cite{AP_2000}.
 
-For ease of implementation, it would be best to first use the squared objective before exploring the linear objective.
-We can use HeAP \cite{AP_2012}, which uses the squared objective along with the B2B net model, as an instructive reference.
-
-We can cast the squared formulation into quadratic form, which can be written compactly as
+For ease of implementation, it would be best to first implement the squared objective before exploring the linear objective.
+We can use the FPGA placer HeAP (2012) \cite{AP_2012}, which uses the squared objective along with the B2B net model, as an instructive reference.
+We can cast the squared formulation into quadratic form. 
+In the x-direction this can be written conveniently as
 \begin{equation}
@@
 Solving $\boldsymbol{Q_x}\boldsymbol{x} = -\boldsymbol{c_x}$ yields the optimal continuous $x$ coordinates, and the same applies independently for $y$.
+This quadratic form lends itself nicely to some OTS quadratic solvers.
 

### 2025-08-28 latex/paper/sections/12-future_work.tex
review kraftwerk

@@
 where \(V_{H}\) is the set of vertices and \(E_{H}\) is the set of hyperedges, each of which may connect more than two vertices. 
-In FPGA context, \(V_{H}\) corresponds to the set of \texttt{SiteInst}s placed on the \texttt{device}, while \(E_{H}\) corresponds to the set of \texttt{Net}s connecting them.
+In RapidWright context, \(V_{H}\) corresponds to the set of \texttt{SiteInst}s placed on the \texttt{device}, while \(E_{H}\) corresponds to the set of \texttt{Net}s connecting them.
+More accurately, $V_H$ corresponds to the set of pins (\texttt{SitePinInst}, or SPI for short), not the \texttt{SiteInst}s themselves.
+The distinction is usually only important in VLSI context, when the area of a module or macro instance is large enough that the distance between its pins becomes significant \cite{kraftwerk2}.
+However, in the context of FPGAs where \texttt{Site}s are relatively uniform size and whose areas are small compared to the area of the \texttt{device}, all \texttt{SitePinInst}s belonging to the same \texttt{SiteInst} can share the same location coordinates.
+
+(SHOW THE NOTATION)
+
+Modules 
+$m \in M$
+
+Pins
+$p \in P$
+
+Nets (hyperedges)
+$n \in N$
+
+$e \in {\mathcal{E}}_n$
+
+$e = (p, q)$
 
@@
     \item \textbf{Star model:} classically, introduce a virtual star node and connect it to all pins of the net.
-However, in electronics, we can simply use the voltage source node as the star node and avoid adding virtual nodes.
+However, in electronics, we can simply use the voltage source node as the star pipipinnnnode and avoid adding virtual nodes.
     \item \textbf{Clique model:} connect all pins of the net to each other (quadratic growth in edges). 
@@
 Since HPWL is direction-agnostic, directionality of FPGA nets (source/sinks) is typically ignored in reduction. 
-Figure \ref{fig:net_model} shows an example of hyperedge reduction using the star and clique models.
+Figure \ref{fig:net_model} shows an example of hyperedge reduction using the star and clique models while \ref{fig:bound2bound} shows the B2B reduction.
 
@@
 \begin{equation}
-    \boldsymbol{\Phi} (\boldsymbol{x}, \boldsymbol{y}) = \sum_{i,j} w_{i,j} \left( |x_i - x_j| + |y_i - y_j| \right)
+    \boldsymbol{\Phi} (\boldsymbol{x}, \boldsymbol{y}) = \sum_{i,j} w_{ij} \left( |x_i - x_j| + |y_i - y_j| \right)
     \label{equ:Manhattan}
@@
 \begin{equation}
-    \boldsymbol{\Phi} (\boldsymbol{x}, \boldsymbol{y}) = \sum_{i,j} w_{i,j} \left[ (x_i - x_j)^2 + (y_i - y_j)^2 \right]
+    \boldsymbol{\Phi} (\boldsymbol{x}, \boldsymbol{y}) = \sum_{i,j} w_{ij} \left[ (x_i - x_j)^2 + (y_i - y_j)^2 \right]
     \label{equ:Euclidian}
@@
 In both formulations, the objective function is separable with $x$ and $y$.
-The weight \(w_{i,j}\) is the weight of the 2-pin edge connecting nodes $i$ and $j$. 
-Its value depends on both the net model and the objective:
+The weight \(w_{ij}\) is the weight of the edge connecting pins $i$ and $j$. 
+Its value depends on both the selected net model and objective:
 
@@
     \item \textbf{Star/Clique models with a quadratic (squared-distance) objective:}  
-    Use constant edge weights. In the star model, each reduced 2-pin edge is given unit weight $w_{ij}=1$.  
+    Use constant edge weights. In the star model, each pin-pin edge is given unit weight $w_{ij}=1$.  
     In the clique model, each pin–pin edge is given weight $w_{ij}=1/(p-1)$, where $p$ is the pin count of the original hyperedge from which this edge was produced.  
@@
     \begin{equation}
-        w_{i,j} = \frac{2}{(p-1)\,|x_i - x_j|}.
+        w_{ij} = \frac{2}{(p-1)\,|x_i - x_j|}.
         \label{equ:weight_linearized}
@@
 We can use the FPGA placer HeAP (2012) \cite{AP_2012}, which uses the squared objective along with the B2B net model, as an instructive reference.
-We can cast the squared formulation into quadratic form. 
-In the x-direction this can be written conveniently as
+
+We can cast the squared formulation (\ref{equ:Euclidian}) into standard quadratic form. 
+The derivation is shown in Kraftwerk2 (2008) \cite{kraftwerk2}.
+Take the original squared objective in the $x$ direction (\ref{equ:sq_obj_x}) and expand each term (\ref{equ:w_expand}).
+\begin{equation}
+    {\boldsymbol{\Phi}}_x (\boldsymbol{x}) = \sum_{i, j} w_{ij} \left( x_i - x_j \right)^2
+    \label{equ:sq_obj_x}
+\end{equation}
+
+\begin{equation}
+    w_{ij}(x_i - x_j)^2 = w_{ij} \left(x_i^2 - 2 x_i x_j + x_j^2 \right)
+    \label{equ:w_expand}
+\end{equation}
+
+In the x-direction this can be written as
 \begin{equation}
     \boldsymbol{\Phi} (\boldsymbol{x}) 
-    = \tfrac{1}{2}\,\boldsymbol{x}^T \boldsymbol{Q_x}\,\boldsymbol{x} + \boldsymbol{c_x}^T \boldsymbol{x} + \text{const},
+    = \tfrac{1}{2}\,\boldsymbol{x}^T \boldsymbol{Q}_x\,\boldsymbol{x} + \boldsymbol{c}_x^T \boldsymbol{x} + \text{const},
     \label{equ:quadratic}

### 2025-09-03 latex/paper/sections/12-future_work.tex
wrap it up bro

@@
 
-In its current state, the prepacker and packer struggle to handle signals larger than 24-bits, especially when involved in DSP functions like addition and multiplication. 
-In such designs, the Vivado synthesizer may synthesize long \texttt{CARRY4} chains with particular \texttt{EDIFHierPortInst} configurations which are currently not handled by our packer and lead to failures in the subsequent routing stage.
+In its current state, the prepacker and packer encounter problems when handling signals larger than 24-bits, especially when involved in DSP functions like addition and multiplication. 
+In such designs, the Vivado synthesizer may synthesize long \texttt{CARRY4} chains with specific \texttt{EDIFHierPortInst} configurations which are currently not handled by our packer and lead to failures in the subsequent routing stage.
 Further work is required to resolve these constraints in the packer.
@@
 
-(SHOW THE NOTATION)
-
-Modules 
-$m \in M$
-
-Pins
-$p \in P$
-
-Nets (hyperedges)
-$n \in N$
-
-$e \in {\mathcal{E}}_n$
-
-$e = (p, q)$
-
 This hypergraph can be reduced into a graph using one of several net models:
@@
     \item \textbf{Star model:} classically, introduce a virtual star node and connect it to all pins of the net.
-However, in electronics, we can simply use the voltage source node as the star pipipinnnnode and avoid adding virtual nodes.
+However, in electronics, we can simply use the voltage source node as the star node and avoid adding virtual nodes.
     \item \textbf{Clique model:} connect all pins of the net to each other (quadratic growth in edges). 
@@
 \end{itemize}
+
+% The VLSI placer Kraftwerk2 (2008) \cite{kraftwerk2} uses the following notation.
+% Integrated circuits consist of a set of modules (set $M$), the modules have pins (set $P$), and the pins are connected by nets (set $N$).
+% The nets are modeled by 2-pin connections 
+% After hypergraph reduction using one of the net models, this results in the representation of each net $n \in N$ by a set $\mathcal{E}_n$ of 2-pin connections. 
+% A single 2-pin connection $e = (p, q)$ connects pins $p$ and $q$.
+% Each pin, $p$ for example, is located at coordinates $(x_p, y_p)$.
 
@@
 
-After reduction, AP minimizes a wirelength objective, typically either a linear HPWL formulation \eqref{equ:Manhattan} or a surrogate quadratic formulation \eqref{equ:Euclidian}:
+After reduction comes global placement, where AP minimizes a wirelength objective, typically either a linear HPWL formulation \eqref{equ:Manhattan} or a quadratic approximation \eqref{equ:Euclidian}:
 
@@
 
-Here, module placements in the $x$ and $y$ directions are captured by the placement vectors \( \boldsymbol{x} = (x_1, x_2, ..., x_n) \) and \( \boldsymbol{y} = (y_1, y_2, ..., y_n) \), where $x_n$ and $y_n$ correspond to the coordinates of the $n$th module instance.
+with $i, j \in \{1, 2, ..., M\}$, where $M$ is the total number of modules in the design.
+Module placements in the $x$ and $y$ directions are captured by the placement vectors \( \boldsymbol{x} = (x_1, x_2, ..., x_n) \) and \( \boldsymbol{y} = (y_1, y_2, ..., y_n) \), where $x_n$ and $y_n$ correspond to the coordinates of the $n$th module instance.
+Edges connecting pins $i$ and $j$ have weights $w_{ij}$ and are assigned values depending on the selected net model and objective.
+
 In both formulations, the objective function is separable with $x$ and $y$.
-The weight \(w_{ij}\) is the weight of the edge connecting pins $i$ and $j$. 
-Its value depends on both the selected net model and objective:
+The squared objective is strongly convex and yields a unique minimizer while requiring a single linear solve per iteration. 
+However, it tends to over-emphasize long wires relative to short ones and tends to worsen routing demand compared to linear metrics \cite{AP_2000}. 
+The linear wirelength objective more closely reflects how HPWL is measured and better correlates with routability. 
+However, it is non-differentiable, so analytical placers typically handle it by iterated quadratic approximations or by smooth regularizations that enable Newton-type solvers \cite{AP_2000}. 
+For ease of implementation, it would be best to first implement the squared objective before exploring the linear objective.
 
-\begin{itemize}
-    \item \textbf{Star/Clique models with a quadratic (squared-distance) objective:}  
-    Use constant edge weights. In the star model, each pin-pin edge is given unit weight $w_{ij}=1$.  
-    In the clique model, each pin–pin edge is given weight $w_{ij}=1/(p-1)$, where $p$ is the pin count of the original hyperedge from which this edge was produced.  
-    This scaling prevents large-fanout nets from dominating the cost and makes star and clique models equivalent in the quadratic formulation.
+We can use the FPGA placer HeAP (2012) \cite{AP_2012}, which uses the squared objective along with the B2B net model, as an instructive reference.
+In HeAP, edges are weighted inversely with the distances between between pins and the number of pins on the original hyperedge from which the edge was produced.
 
-    \item \textbf{Star/Clique models with a linear HPWL objective (via quadratic surrogate):}  
-    Use iterative reweighting. With the star model, each edge weight is updated as $w_{ij}=1/|x_i-x_j|$. With the clique model, the weight is $1/((p-1)|x_i-x_j|)$.  
-    This reweighting makes the quadratic form act as a surrogate for $|x_i-x_j|$, approximating the true HPWL.
-
-    \item \textbf{Bound2Bound model with a linear HPWL objective:}  
-    Connect only boundary–boundary and boundary–interior pin pairs, and assign weights as inverse distances:  
 \begin{equation}
@@
 \end{equation}
-    With this assignment, the quadratic cost of a net equals its HPWL (in each dimension) at the evaluation point, making the Bound2Bound model HPWL-accurate.
-\end{itemize}
 
-The squared wirelength objective yields a unique minimizer and requires a single linear solve per iteration. 
-However, it tends to over-emphasize long wires relative to short ones and tends to worsen routing demand compared to linear metrics \cite{AP_2000}. 
-The linear wirelength objective more closely reflects how HPWL is measured and better correlates with routability. 
-However, it is non-differentiable, so analytical placers typically handle it by iterated quadratic approximations (e.g., GORDIAN-L reweights edge coefficients each iteration) or by smooth regularizations that enable Newton-type solvers \cite{AP_2000}.
-
-For ease of implementation, it would be best to first implement the squared objective before exploring the linear objective.
-We can use the FPGA placer HeAP (2012) \cite{AP_2012}, which uses the squared objective along with the B2B net model, as an instructive reference.
-
-We can cast the squared formulation (\ref{equ:Euclidian}) into standard quadratic form. 
-The derivation is shown in Kraftwerk2 (2008) \cite{kraftwerk2}.
-Take the original squared objective in the $x$ direction (\ref{equ:sq_obj_x}) and expand each term (\ref{equ:w_expand}).
-\begin{equation}
-    {\boldsymbol{\Phi}}_x (\boldsymbol{x}) = \sum_{i, j} w_{ij} \left( x_i - x_j \right)^2
-    \label{equ:sq_obj_x}
-\end{equation}
-
-\begin{equation}
-    w_{ij}(x_i - x_j)^2 = w_{ij} \left(x_i^2 - 2 x_i x_j + x_j^2 \right)
-    \label{equ:w_expand}
-\end{equation}
-
-In the x-direction this can be written as
+The squared objective \ref{equ:Euclidian} can be recast into matrix-vector form.
+In the x-direction this can be written as:
 \begin{equation}
@@
 \end{equation}
-where $\boldsymbol{Q_x}$ encodes 2-pin connections between movable cells and $\boldsymbol{c_x}$ accounts for fixed terminals.  
-Solving $\boldsymbol{Q_x}\boldsymbol{x} = -\boldsymbol{c_x}$ yields the optimal continuous $x$ coordinates, and the same applies independently for $y$.
-This quadratic form lends itself nicely to some OTS quadratic solvers.
+
+The connectivity matrix $\boldsymbol{Q_x}$ encodes 2-pin connections between movable instances while $\boldsymbol{c_x}$ encodes 2-pin connections between movable instances and fixed instances   
+Movable instances are for example \texttt{SLICEL}s, DSP, and RAM, while fixed instances can be IO pads and clock buffers.
+We distinguish fixed instances because their placement cannot be optimized and can be thought of as constraints encoded into the objective.
+For a design with $M$ pins belonging to movable instances and $F$ pins belonging to fixed instances, $\boldsymbol{Q}_x \in {\mathbb{R}}^{M \times M}$ and $\boldsymbol{c}_x \in {\mathbb{R}}^F$.
+
+Since the objective is strongly convex, we can solve for the global minimum cost by setting the partial derivatives to zero.
+
+\begin{equation}
+    \frac{\partial}{\partial x} \boldsymbol{\Phi} (\boldsymbol{x}) = 0 \Longleftrightarrow \boldsymbol{Q}_x \boldsymbol{x} = - \boldsymbol{c}_x
+\end{equation}
+
+Solving this linear system yields the optimal $x$ coordinate placements, and the same applies independently for $y$.
+We can use any number of OTS solver frameworks for this step, potentially \texttt{cvxpy} in Python or a Java-native package like \texttt{EMJ}.
+Note that this is only global placement, which is not legal and contains many constraint violations and placement overlaps.
+
+Following global placement is detailed placement (legalization) which is where the real challenge begins.
+Most placers have a custom method for doing so, tailored to specific optimizations or cost objectives.
+We can continue using HeAP as a reference, which performs recursive partitioning to transform the solution provided by the solver into a legal overlap-free placement.
+The discrete nature of the arrangement of \texttt{Site}s on the \texttt{device} and all of the other hardware constraints will make implementation very difficult.
+Just like in our SA placer, the most challenging parts to implement were the prepacker and packer where we deal with placement constraints, not the placer itself.
+
+% The first step of legalization is to identify areas or \texttt{Site}s on the \texttt{device} that are overutilized.
+% If we take the global placement containing the ($x$, $y$) coordinates of all instances, overlay it onto the grid of resources of the \texttt{device}, and snap each instance to the closest legal \texttt{Site}, any \texttt{Site} occupied by more than one instance can be considered overutilized.
+% Overutilized \texttt{Site}s that are adjacent to one another are repeatedly clustered together until all clusters are bordered on all sides by non-overutilized \texttt{Site}s, giving us so-called "overutilized areas".
+% Next, each area is expanded in both $x$ and $y$ dimensions until it is large enough to accomodate all instances within.
+% 
+% In the second step, we generate two cuts: source cut and a target cut.
+% The source cut pertains to the instances being placed and the target cut pertains to the area of \texttt{Site}s into which the instances are placed.
+% This is a similar to our SA placer when making random moves and swaps, where we have an original "home" \texttt{Site} and a proposed "away" \texttt{Site} that we accept or reject based on cost and temperature, except now we are dealing with clusters of \texttt{Site}s which are non-uniform in shape.
+% The source cut splits the instances into two partitions, while the target cut splits the area into two sub-areas, into which the instances in each partition are spread.
+% The target cut is a horizontal $x$ or vertical $y$ cut of the area such that all blocks in each partition fit in their respective sub-areas.
+% 
+% Next, the instances in sub-areas are spread to distribute them evenly.
+% The discrete nature of FPGA \texttt{Site} locations will make this process difficult.
+
+% We know the squared objective function \ref{equ:Euclidian} to be strongly convex and quadratic, so it would be beneficial to be able to re-express it in matrix-vector form to easily access gradient and Hessian information about the system and to make it compatible with OTS solvers.
+% With some algebraic manipulation, we can cast sum of squares formulation (\ref{equ:Euclidian}) into quadratic matrix form.
+% Take the original objective in the x-direction:
+% 
+% \begin{equation}
+%     {\boldsymbol{\Phi}}_x (\boldsymbol{x}) = \sum_{i, j} w_{ij} \left( x_i - x_j \right)^2
+%     \label{equ:sq_obj_x}
+% \end{equation}
+% 
+% Expanding the summation:
+% 
+% \begin{equation}
+%     {\boldsymbol{\Phi}}_x (\boldsymbol{x}) = \sum_{i, j} w_{ij} x_i^2 + \sum_{i, j} w_{ij} x_j^2 - 2 \sum_{i, j} w_{ij} x_i x_j
+%     \label{equ:w_expand}
+% \end{equation}
+% 
+% A net connecting node $i$ with $j$ is the same as connecting node $j$ with $i$, so $w_{ij} = w_{ji}$.
+% This means the weight matrix $W$ with elements $w_{ij}$ is symmetric.
+% Using this symmetry, we can group the first two terms in \ref{equ:w_expand}:
+% \begin{equation}
+%     {\boldsymbol{\Phi}}_x (\boldsymbol{x}) = 2 \sum_{i, j} w_{ij} x_i^2  - 2 \sum_{i, j} w_{ij} x_i x_j
+%     \label{equ:w_grouped} 
+% \end{equation}
+
+% by constructing the \emph{graph Laplacian matrix} out of the undirected weighted graph $G = (V, E)$ after hypergraph reduction. 
+% 
+% The Laplacian matrix \(L\) can be constructed with:
+% 
+% \begin{equation}
+%     L = BWB^T
+% \end{equation}
+% 
+% where $B$ is the \emph{incidence matrix} and $W$ is the diagonal \emph{weight matrix}.
+% The $|v| \times |e|$ incidence matrix $B$ of graph $G$ has elements $b_{ve}$, where $v$ denotes the vertex index and $e$ denotes the edge index.
+% Each edge $e$ connects a head vertex and a tail vertex, denoted as $e = (v_{head}, v_{tail})$.
+% 
+% \begin{equation}
+%     b_{ve} = 
+%     \begin{cases}
+%         +1, & \text{if } v \text{ is the head (source) of edge e} \\
+%         -1, & \text{if } v \text{ is the tail (sink) of edge e} \\ 
+%         0, & \text{otherwise}.
+%     \end{cases}
+% \end{equation}
+% 
+% The assignment of which vertex is the head and which is the tail is arbitrary and does not correlate with any directionality of the edges.
+% But for consistent bookkeeping purposes, we can assign the lower-indexed vertex as the head and the higher-indexed vertex as the tail like so: 
+% 
+% \begin{equation}
+%     b_{ve} = 
+%     \begin{cases}
+%         +1, & \text{if } v = v_i, \\
+%         -1, & \text{if } v = v_j, \\ 
+%         0, & \text{otherwise}.
+%     \end{cases}
+% \end{equation}
 
@@
 \subsection{Add Hard Macro Support}
-fdsa
-
-\subsection{Parallelization}
-fdsa
+As mentioned before, our SA placer is Site-centric.
+An interesting direction of study would be to expand our placer to support Module-Site-centric mixed placement, where Modules are hard macros comprising multiple Sites.
+This of course would also add complexity to the prepacker and packer, as now we are dealing with non-uniform mixed-size placement objects.
+The RapidWright repository has two implementations of purely Module-centric placers, called \texttt{BlockPlacer} and \texttt{BlockPlacer2}, though it currently lacks a purely Site-centric placer or a mixed Module-Site-centric placer.
+The corresponding paper by Lavin et. al. (2010) \cite{HardBlock} can serve as a useful reference for this direction.

### 2025-09-03 latex/paper/sections/12-future_work.tex
finish the abstract

@@
 \emph{detailed placement} (legalization), which snaps objects to valid \texttt{site}s while adhering to hardware constraints.  
-Global placement can be performed using any number of off-the-shelf (OTS) analytical solvers.
+Global placement can be performed using any number of off-the-shelf (OTS) analytical solvers, but detailed placement requires custom implementation tailored to the hardware.
+We repeatedly loop through global placement into detailed placement until HPWL minimization slows to a stop.
 

### 2025-09-05 latex/paper/sections/02-abstract.tex
readme graphics

@@
 \section{Abstract}
-fdsafdsafdsa. \\ 
-fdsafdsafdsa. \\ 
-fdsafdsafdsa. \\ 
-fdsafdsafdsa. \\ 
-fdsafdsafdsa. \\ 
-fdsafdsafdsa. \\ 
-fdsafdsafdsa. \\ 
-fdsafdsafdsa. \\ 
-fdsafdsafdsa. \\ 
-fdsafdsafdsa. \\ 
-fdsafdsafdsa. \\ 
-fdsafdsafdsa. \\ 
-fdsafdsafdsa. \\
+Placement remains one of the most computationally demanding stages of FPGA computer-aided design (CAD) flows, directly influencing routability, timing, and design productivity. 
+This paper presents a modular placement framework targeting Xilinx 7-Series devices, implemented using the RapidWright API. 
+The flow first clusters multi-cell structures—such as CARRY4 chains, DSP48E1 cascades, and LUT–FF pairs—into prepacked groups, then maps them to physical sites before applying a simulated annealing (SA) placer.
+The SA engine minimizes half-perimeter wirelength (HPWL) using a geometric cooling schedule, with support for both random and directed moves. 
+The result is a legal, wirelength-optimized placement that forms a foundation for experimenting with alternative cost functions and analytical or hybrid placement strategies.

### 2025-09-05 latex/paper/sections/03-introduction.tex
readme graphics

@@
 The \textbf{prepacking} stage involves traversing a raw EDIF netlist to identify recurring cell patterns—such as CARRY chains, DSP cascades, and LUT-FF pairs—that are critical for efficient mapping and legalization.
-In the subsequent \textbf{packing} stage, these identified patterns, along with any remaining loose cells, are consolidated into SiteInst objects that encapsulate the FPGA’s discrete resource constraints and architectural nuances.
-Finally, the \textbf{placement} stage employs a simulated annealing (SA) algorithm to optimally assign SiteInst objects to physical sites, aiming to minimize total wirelength while adhering to the constraints of the 7-series architecture.
+In the subsequent \textbf{packing} stage, these identified patterns, along with any remaining loose cells, are consolidated into \texttt{SiteInst} objects that encapsulate the FPGA’s discrete resource constraints and architectural nuances.
+Finally, the \textbf{placement} stage employs a simulated annealing (SA) algorithm to optimally assign \texttt{SiteInst} objects to physical \texttt{Sites}, aiming to minimize total wirelength while respecting the constraints of the 7-series architecture.
 

### 2025-09-05 latex/paper/sections/12-future_work.tex
readme graphics

@@
 For a design with $M$ pins belonging to movable instances and $F$ pins belonging to fixed instances, $\boldsymbol{Q}_x \in {\mathbb{R}}^{M \times M}$ and $\boldsymbol{c}_x \in {\mathbb{R}}^F$.
+We can construct $\boldsymbol{Q}_x$ and $\boldsymbol{c}_x$ formally by taking the graph Laplacian matrix $L_x$ of our netlist reduced into a graph and partitioning the matrix to separate M-M from M-F weights to form $\boldsymbol{Q}_x$ and $\boldsymbol{c}_x$.
+However, we can simply follow a shortcut outlined in SimPL \cite{SimPL}.
+If we take our netlist reduced into a graph
 

### 2025-09-05 latex/paper/sections/13-conclusion.tex
readme graphics

\section{Conclusion}

This work has demonstrated a complete placement flow for heterogeneous FPGAs built upon the RapidWright framework, culminating in a functional simulated annealing placer. 
By systematically prepacking multi-cell structures, packing them into SiteInst containers, and applying an SA-driven optimization, we produce placements that respect the discrete architectural constraints of Xilinx 7-Series devices while reducing wirelength. 
Although SA is computationally expensive compared to modern analytical placers, its simplicity, flexibility, and robustness make it an excellent foundation for research and education. 
The infrastructure we have developed can be extended in several directions: integrating timing-driven cost functions, incorporating legalization-aware directed moves, and experimenting with analytical or hybrid placement algorithms. 
Future work will focus on replacing or augmenting the SA stage with an analytical solver to combine the quality of wirelength minimization with the scalability and runtime performance needed for larger industrial designs.

### 2025-09-11 latex/paper/sections/03-introduction.tex
basic edits

@@
 FPGA CAD flows mainly encompass synthesis, placement, and routing; all of which are NP-hard problems, of which placement is one of the most time-consuming processes. 
-Inefficienct placement strategy not only extends design times from hours to days, thereby elevating cost and reducing engineering productivity, but also limits the broader adoption of FPGAs by software engineers who expect compile times akin to those of software compilers like {\tt gcc}.
+Inefficient placement strategy not only extends design times from hours to days, thereby elevating cost and reducing engineering productivity, but also limits the broader adoption of FPGAs by software engineers who expect compile times akin to those of software compilers like {\tt gcc}.
 
@@
 In the subsequent \textbf{packing} stage, these identified patterns, along with any remaining loose cells, are consolidated into \texttt{SiteInst} objects that encapsulate the FPGA’s discrete resource constraints and architectural nuances.
-Finally, the \textbf{placement} stage employs a simulated annealing (SA) algorithm to optimally assign \texttt{SiteInst} objects to physical \texttt{Sites}, aiming to minimize total wirelength while respecting the constraints of the 7-series architecture.
+Finally, the \textbf{placement} stage employs a simulated annealing (SA) algorithm to optimally assign \texttt{SiteInst} objects to physical \texttt{Sites}, aiming to minimize total wirelength while the constraints of the 7-series architecture.
 

### 2025-09-11 latex/paper/sections/12-future_work.tex
basic edits

--- 2025-09-03 latex/paper/sections/12-future_work.tex
@@
 
-with $i, j \in \{1, 2, ..., M\}$, where $M$ is the total number of modules in the design.
+with $i, j \in \{1, 2, ..., N\}$, where $N$ is the total number of modules in the design.
 Module placements in the $x$ and $y$ directions are captured by the placement vectors \( \boldsymbol{x} = (x_1, x_2, ..., x_n) \) and \( \boldsymbol{y} = (y_1, y_2, ..., y_n) \), where $x_n$ and $y_n$ correspond to the coordinates of the $n$th module instance.
@@
 For a design with $M$ pins belonging to movable instances and $F$ pins belonging to fixed instances, $\boldsymbol{Q}_x \in {\mathbb{R}}^{M \times M}$ and $\boldsymbol{c}_x \in {\mathbb{R}}^F$.
+We can construct $\boldsymbol{Q}_x$ and $\boldsymbol{c}_x$ formally by taking the graph Laplacian matrix of our undirected weighted graph (with M-M and M-F connections included), then partitioning the matrix to separate M-M from M-F weights to form $\boldsymbol{Q}_x$ and $\boldsymbol{c}_x$.
+Alternatively, we can simply follow a shortcut outlined in SimPL \cite{SimPL}, where we construct $Q_x$ by setting the diagonal elements $w_{ii}$ as the sum of net weights of all connections to movable module pin $i$.
+The off-diagonal elements $w_{ij}$ are the negative sum of net weights of connections between movable module pins $i$ and $j$, resulting in a symmetric matrix.
+The vector $\boldsymbol{c}_x$ has entry $-c_i$ in the $i$th row, where $c_i$ is the weight of the connection between fixed pin $i$ and and movable pin $i$, most of which will be zero.
+The SimPL paper offers a small example of this construction for a toy netlist \cite{SimPL}.
 

### 2025-10-23 latex/paper/sections/02-abstract.tex
update vivado version, edits

@@
 This paper presents a modular placement framework targeting Xilinx 7-Series devices, implemented using the RapidWright API. 
-The flow first clusters multi-cell structures—such as CARRY4 chains, DSP48E1 cascades, and LUT–FF pairs—into prepacked groups, then maps them to physical sites before applying a simulated annealing (SA) placer.
+The flow first clusters multi-cell structures (such as CARRY4 chains, DSP48E1 cascades, and LUT–FF pairs) into prepacked groups, then maps them to physical sites before applying a simulated annealing (SA) placer.
 The SA engine minimizes half-perimeter wirelength (HPWL) using a geometric cooling schedule, with support for both random and directed moves. 

### 2025-10-23 run.sh

@@ -3,10 +3,10 @@
 DESIGN="fir_filter"
 TOP_LEVEL="top_level"
 
-export XILINX_VIVADO=/home/bcheng/workspace/tools/Xilinx/Vivado/2023.2
+export XILINX_VIVADO=/home/bcheng/workspace/tools/Xilinx/2025.1.1/Vivado
 export PATH="$PATH:$XILINX_VIVADO/bin"
 
-export JAVA_HOME=/home/bcheng/workspace/tools/Xilinx/Vivado/2023.2/tps/lnx64/jre17.0.7_7
+export JAVA_HOME=/home/bcheng/workspace/tools/Xilinx/2025.1.1/Vivado/tps/lnx64/jre21.0.1_12
 export PATH="$JAVA_HOME/bin:$PATH"
 
 export RAPIDWRIGHT_PATH=/home/bcheng/workspace/tools/RapidWright
